### 2018-10-29 main.tex
updated

\documentclass[]{scrartcl}

%opening
\usepackage[round]{natbib}
\usepackage{listings}
% \usepackage{minted}
\usepackage{color}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{dsfont}
\usepackage{pgf}
\usepackage{subcaption}
\usepackage{tikz}
\usepackage{hyperref}
\usepackage{dsfont}
\usetikzlibrary{bayesnet}
\newcommand{\iden}{\ensuremath{\mathbb{I}}}  %identity func
\newcommand{\real}{\ensuremath{\mathbb{R}}} %real numbers
\newcommand{\nat}{\ensuremath{\mathbb{N}}}  %natural numbers
\newcommand{\ints}{\ensuremath{\mathbb{Z}}}
\newcommand{\expt}{\ensuremath{\mathbb{E}}}
\newcommand{\X}{\ensuremath{\mathbb{X}}}
\newcommand{\Z}{\ensuremath{\mathfrak{Z}}}
\newcommand{\y}{\textbf{y}} 
\newcommand{\x}{\textbf{x}} 
\newcommand{\asto}{\overset{a.s.}{\to}}

\newcommand{\om}{\textit{OpenMalaria}}
\DeclareMathOperator*{\argmax}{arg\,\!max}

\title{Notes on Open Malaria}
\author{Bradley Gram-Hansen}

\begin{document}

\maketitle

\begin{abstract}

\end{abstract}

% \input{mognotes/mixofgauss}

\section{Simulation Options}

The \om simulator is designed for simulating individual scenarios. 
In order to simulate a study covering variations in several factors, one 
would have to  design a fully \textit{factorial} experiment.

\begin{quote}
	\begin{itemize}
		\item \textbf{\textit{Sweep}} : A sweep is one of more covarying factors, and set of all \textit{arms} of this sweep.
		\item \textbf{\textit{Arm}}: Each sweep must have one or more values for each of its factors. An \textit{arm} 
		is one of these combinations of values - a configuration assigning a value to each factor on the sweep. 
		\item \textbf{\textit{Experiment}}: Largely synonymous with a study, an \textit{experiment} is the complete
		description of all sweeps and arms used, together with the scenarios generated and results produced. Each scenario is generated by choosing
		one arm for each sweep in the experiment. 
		\item \textbf{\textit{Full factorial design}} : A \textit{full factorial experiment} is one where all possible scenarios are generated: all
		combinations of one arm per sweep are used.
		\item \textbf{\textit{Seed}}: Each time a scenario is simulated, the random number generated must be started from some \textit{seed}.
	\end{itemize}
\end{quote}

\end{document}

Since OpenMalaria simulations are stochastic, they use a number of random seeds (say 2-50) 
to provide estimates of the contributions of random variation to the results.

### 2018-10-29 code/openmalaria/openmalaria-schema-39.0/model/Clinical/CaseManagementCommon.cpp
Code

@@ -1,0 +1,94 @@
+/* This file is part of OpenMalaria.
+ * 
+ * Copyright (C) 2005-2015 Swiss Tropical and Public Health Institute
+ * Copyright (C) 2005-2015 Liverpool School Of Tropical Medicine
+ * 
+ * OpenMalaria is free software; you can redistribute it and/or modify
+ * it under the terms of the GNU General Public License as published by
+ * the Free Software Foundation; either version 2 of the License, or (at
+ * your option) any later version.
+ * 
+ * This program is distributed in the hope that it will be useful, but
+ * WITHOUT ANY WARRANTY; without even the implied warranty of
+ * MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the GNU
+ * General Public License for more details.
+ * 
+ * You should have received a copy of the GNU General Public License
+ * along with this program; if not, write to the Free Software
+ * Foundation, Inc., 51 Franklin Street, Fifth Floor, Boston, MA 02110-1301, USA.
+ */
+
+#include "Clinical/CaseManagementCommon.h"
+#include "util/checkpoint_containers.h"
+#include "util/ModelOptions.h"
+
+namespace OM { namespace Clinical {
+
+bool indirectMortBugfix;
+
+SimTime healthSystemMemory( SimTime::never() );
+
+//log odds ratio of case-fatality in community compared to hospital
+double oddsRatioThreshold;
+
+util::AgeGroupInterpolator caseFatalityRate;
+util::AgeGroupInterpolator pSequelaeInpatient;
+
+///@brief Infant death summaries (checkpointed).
+///
+/// Declared 'extern' in header.
+//@{
+vector<int> infantDeaths;
+vector<int> infantIntervalsAtRisk;
+//@}
+
+/// Non-malaria mortality in under 1year olds.
+/// Set by init ()
+double nonMalariaMortality;
+
+
+void initCMCommon( const OM::Parameters& parameters, SimTime hsMemory ){
+    indirectMortBugfix = util::ModelOptions::option (util::INDIRECT_MORTALITY_FIX);
+    healthSystemMemory = hsMemory;
+    oddsRatioThreshold = exp( parameters[Parameters::LOG_ODDS_RATIO_CF_COMMUNITY] );
+    infantDeaths.resize(SimTime::stepsPerYear());
+    infantIntervalsAtRisk.resize(SimTime::stepsPerYear());
[…]

### 2018-10-29 code/openmalaria/openmalaria-schema-39.0/model/Clinical/DecisionTree5Day.h

@@ -1,0 +1,58 @@
+/* This file is part of OpenMalaria.
+ * 
+ * Copyright (C) 2005-2015 Swiss Tropical and Public Health Institute
+ * Copyright (C) 2005-2015 Liverpool School Of Tropical Medicine
+ * 
+ * OpenMalaria is free software; you can redistribute it and/or modify
+ * it under the terms of the GNU General Public License as published by
+ * the Free Software Foundation; either version 2 of the License, or (at
+ * your option) any later version.
+ * 
+ * This program is distributed in the hope that it will be useful, but
+ * WITHOUT ANY WARRANTY; without even the implied warranty of
+ * MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the GNU
+ * General Public License for more details.
+ * 
+ * You should have received a copy of the GNU General Public License
+ * along with this program; if not, write to the Free Software
+ * Foundation, Inc., 51 Franklin Street, Fifth Floor, Boston, MA 02110-1301, USA.
+ */
+
+#ifndef Hmod_DecisionTree5Day
+#define Hmod_DecisionTree5Day
+
+#include "Clinical/CM5DayCommon.h"
+
+namespace OM {
+namespace Clinical {
+
+/**
+ * This models case management with 5-day time steps, optionally including
+ * PK/PD modeling for uncomplicated cases.
+ * 
+ * Uncomplicated cases: access, otherwise known as "seeking any type of
+ * treatment", is determined by a fixed-function decision, which may be
+ * modified by a treatment-seeking factor. Treatment decisions (type of
+ * treatment, use of diagnostics, effectiveness) are determined by a
+ * programmable decision tree.
+ * 
+ * Severe cases: all decisions and outcomes are calculated via a fixed-function
+ * probability tree, using the same logic for handling severe cases as has long
+ * been used.
+ */
+class DecisionTree5Day : public CM5DayCommon
+{
+public:
+    /** Load health system data from initial data or an intervention's data (both from XML).
+     * (Re)loads all data affected by this healthSystem element. */
+    static void setHealthSystem (const scnXml::HSDT5Day& hsDescription);
+    
+    DecisionTree5Day (double tSF) : CM5DayCommon(tSF) {}
+    
+protected:
+    virtual void uncomplicatedEvent(Human& human, Episode::State pgState);
+};
+
+}
+}
+#endif

### 2018-10-29 code/openmalaria/openmalaria-schema-39.0/model/Clinical/ESCaseManagement.cpp

@@ -1,0 +1,65 @@
+/* This file is part of OpenMalaria.
+ * 
+ * Copyright (C) 2005-2015 Swiss Tropical and Public Health Institute
+ * Copyright (C) 2005-2015 Liverpool School Of Tropical Medicine
+ * 
+ * OpenMalaria is free software; you can redistribute it and/or modify
+ * it under the terms of the GNU General Public License as published by
+ * the Free Software Foundation; either version 2 of the License, or (at
+ * your option) any later version.
+ * 
+ * This program is distributed in the hope that it will be useful, but
+ * WITHOUT ANY WARRANTY; without even the implied warranty of
+ * MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the GNU
+ * General Public License for more details.
+ * 
+ * You should have received a copy of the GNU General Public License
+ * along with this program; if not, write to the Free Software
+ * Foundation, Inc., 51 Franklin Street, Fifth Floor, Boston, MA 02110-1301, USA.
+ */
+
+#include "Clinical/ESCaseManagement.h"
+#include "Clinical/CMDecisionTree.h"
+#include "Clinical/EventScheduler.h"
+#include "util/errors.h"
+#include "util/ModelOptions.h"
+
+#include <set>
+#include <sstream>
+#include <boost/format.hpp>
+#include <boost/assign/std/vector.hpp> // for 'operator+=()'
+
+namespace OM { namespace Clinical {
+    using namespace OM::util;
+    using namespace boost::assign;
+    using boost::format;
+
+
+// -----  ESCaseManagement  -----
+
+const CMDecisionTree *escm_uncomplicated = 0,
+    *escm_complicated = 0;
+
+void ESCaseManagement::setHealthSystem(const scnXml::HSEventScheduler& esData){
+    escm_uncomplicated = &CMDecisionTree::create( esData.getUncomplicated(), true );
+    escm_complicated = &CMDecisionTree::create( esData.getComplicated(), false );
+    
+    // Calling our parent class like this is messy. Changing this would require
+    // moving change-of-health-system handling into ClinicalModel.
+    ClinicalEventScheduler::setParameters( esData );
+}
+
+CMDTOut ESCaseManagement::execute (
+        const CMHostData& hostData
+) {
+    assert (hostData.pgState & Episode::SICK);
+    //TODO: should we remove any existing prescriptions?
[…]

### 2018-10-29 code/openmalaria/openmalaria-schema-39.0/model/Host/NeonatalMortality.h

@@ -1,0 +1,50 @@
+/* This file is part of OpenMalaria.
+ * 
+ * Copyright (C) 2005-2015 Swiss Tropical and Public Health Institute
+ * Copyright (C) 2005-2015 Liverpool School Of Tropical Medicine
+ * 
+ * OpenMalaria is free software; you can redistribute it and/or modify
+ * it under the terms of the GNU General Public License as published by
+ * the Free Software Foundation; either version 2 of the License, or (at
+ * your option) any later version.
+ * 
+ * This program is distributed in the hope that it will be useful, but
+ * WITHOUT ANY WARRANTY; without even the implied warranty of
+ * MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the GNU
+ * General Public License for more details.
+ * 
+ * You should have received a copy of the GNU General Public License
+ * along with this program; if not, write to the Free Software
+ * Foundation, Inc., 51 Franklin Street, Fifth Floor, Boston, MA 02110-1301, USA.
+ */
+
+#ifndef Hmod_NeonatalMortality
+#define Hmod_NeonatalMortality
+
+#include "Global.h"
+
+namespace scnXml {
+    class Clinical;
+}
+namespace OM {
+    class Population;
+namespace Host {
+
+class NeonatalMortality {
+public:
+  /// Initialisation
+  static void init( const scnXml::Clinical& clinical );
+  
+  static void staticCheckpoint (istream& stream);
+  static void staticCheckpoint (ostream& stream);
+  
+  /** Called for each birth; returns true if infant dies due to mother's
+   * infection. */
+  static bool eventNeonatalMortality();
+  
+  /** Calculate risk of a neonatal mortality based on humans 20-25 years old. */
+  static void update (const Population& population);
+};
+
+} }
+#endif

### 2018-10-29 code/openmalaria/openmalaria-schema-39.0/model/Parameters.cpp

@@ -1,0 +1,53 @@
+/* This file is part of OpenMalaria.
+ * 
+ * Copyright (C) 2005-2015 Swiss Tropical and Public Health Institute
+ * Copyright (C) 2005-2015 Liverpool School Of Tropical Medicine
+ * 
+ * OpenMalaria is free software; you can redistribute it and/or modify
+ * it under the terms of the GNU General Public License as published by
+ * the Free Software Foundation; either version 2 of the License, or (at
+ * your option) any later version.
+ * 
+ * This program is distributed in the hope that it will be useful, but
+ * WITHOUT ANY WARRANTY; without even the implied warranty of
+ * MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the GNU
+ * General Public License for more details.
+ * 
+ * You should have received a copy of the GNU General Public License
+ * along with this program; if not, write to the Free Software
+ * Foundation, Inc., 51 Franklin Street, Fifth Floor, Boston, MA 02110-1301, USA.
+ */
+
+#include "Parameters.h"
+#include "util/errors.h"
+#include <schema/scenario.h>
+
+#include <boost/format.hpp>
+
+namespace OM {
+    using boost::format;
+    
+// Initialization functions:
+
+Parameters::Parameters( const scnXml::Parameters& parameters ){
+    // set parameters
+    const scnXml::Parameters::ParameterSequence& paramSeq = parameters.getParameter();
+    for(scnXml::Parameters::ParameterConstIterator it = paramSeq.begin(); it != paramSeq.end(); ++it) {
+        int i = it->getNumber();
+        if (i < 0 || i >= MAX)
+            continue;   // ignore the parameter; no real point in making this an error
+        Parameter parameter = static_cast<Parameter>(i);
+        if( !parameterValues.insert( make_pair( parameter, it->getValue() ) ).second )
+            throw util::xml_scenario_error( (format("parameter with index %1% described twice") %parameter).str() );
+    }
+}
+
+
+double Parameters::operator[]( Parameter parameter )const{
+    std::map<Parameter, double>::const_iterator it = parameterValues.find( parameter );
+    if( it == parameterValues.end() )
+        throw util::xml_scenario_error( (format("parameter %1% required but not described") %parameter).str() );
+    return it->second;
+}
+
+}

### 2018-10-29 code/openmalaria/openmalaria-schema-39.0/model/PkPd/Drug/LSTMDrugConversion.h

@@ -1,0 +1,97 @@
+/* This file is part of OpenMalaria.
+ * 
+ * Copyright (C) 2005-2015 Swiss Tropical and Public Health Institute
+ * Copyright (C) 2005-2015 Liverpool School Of Tropical Medicine
+ * 
+ * OpenMalaria is free software; you can redistribute it and/or modify
+ * it under the terms of the GNU General Public License as published by
+ * the Free Software Foundation; either version 2 of the License, or (at
+ * your option) any later version.
+ * 
+ * This program is distributed in the hope that it will be useful, but
+ * WITHOUT ANY WARRANTY; without even the implied warranty of
+ * MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the GNU
+ * General Public License for more details.
+ * 
+ * You should have received a copy of the GNU General Public License
+ * along with this program; if not, write to the Free Software
+ * Foundation, Inc., 51 Franklin Street, Fifth Floor, Boston, MA 02110-1301, USA.
+ */
+
+#ifndef Hmod_LSTMDrugConversion
+#define Hmod_LSTMDrugConversion
+
+#include "Global.h"
+#include "PkPd/Drug/LSTMDrug.h"
+#include "PkPd/Drug/LSTMDrugType.h"
+
+using namespace std;
+
+namespace OM {
+
+namespace PkPd {
+struct DoseParams;
+}
+
+namespace PkPd {
+
+struct Params_convFactor;
+
+/** A class holding pkpd drug use info.
+ * 
+ * Conversion model: two 1-compartment models with conversion.
+ *
+ * Each human has an instance for each type of drug present in their blood. */
+class LSTMDrugConversion : public LSTMDrug {
+public:
+    /** Create a new instance. */
+    LSTMDrugConversion (const LSTMDrugType& parent, const LSTMDrugType& metabolite);
+    
+    virtual size_t getIndex() const;
+    virtual double getConcentration(size_t index) const;
+    
+    virtual void medicate (double time, double qty, double bodyMass);
+    
+    virtual double calculateDrugFactor(uint32_t genotype, double body_mass) const;
+    virtual void updateConcentration (double body_mass);
+    double getMetaboliteConcentration() const;
+    double getParentConcentration() const;
+    
+protected:
+    virtual void checkpoint (istream& stream);
+    virtual void checkpoint (ostream& stream);
+    
+    double calculateFactor(const Params_convFactor& p, double duration) const;
+    
[…]

### 2018-10-29 code/openmalaria/openmalaria-schema-39.0/model/PkPd/Drug/LSTMDrugType.h

@@ -1,0 +1,207 @@
+/* This file is part of OpenMalaria.
+ * 
+ * Copyright (C) 2005-2015 Swiss Tropical and Public Health Institute
+ * Copyright (C) 2005-2015 Liverpool School Of Tropical Medicine
+ * 
+ * OpenMalaria is free software; you can redistribute it and/or modify
+ * it under the terms of the GNU General Public License as published by
+ * the Free Software Foundation; either version 2 of the License, or (at
+ * your option) any later version.
+ * 
+ * This program is distributed in the hope that it will be useful, but
+ * WITHOUT ANY WARRANTY; without even the implied warranty of
+ * MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the GNU
+ * General Public License for more details.
+ * 
+ * You should have received a copy of the GNU General Public License
+ * along with this program; if not, write to the Free Software
+ * Foundation, Inc., 51 Franklin Street, Fifth Floor, Boston, MA 02110-1301, USA.
+ */
+
+#ifndef Hmod_LSTMDrugType
+#define Hmod_LSTMDrugType
+
+#include "Global.h"
+#include "util/sampler.h"
+
+#include <string>
+#include <deque>
+#include <vector>
+#include <set>
+#include <cassert>
+#include <memory>
+#include <map>
+#include <boost/unordered_set.hpp>
+#include <boost/ptr_container/ptr_vector.hpp>
+
+using namespace std;
+
+namespace scnXml{
+    class Drugs;
+    class PKPDDrug;
+    class Phenotype;
+}
+namespace OM {
+namespace PkPd {
+using util::LognormalSampler;
+
+class LSTMDrugType;
+class LSTMDrug;
+
+/** Drug PD parameters (specified per phenotype), as well as applicable
+ * functions to calculate drug factors and concentrations.
+ * 
+ * Parasite genetics is specified elsewhere; we simply map genotype to
+ * phenotype as well as describing phenotypes here. */
+class LSTMDrugPD {
+public:
+    /// Construct
+    LSTMDrugPD( const scnXml::Phenotype& phenotype );
+    
+    /** Calculate a survival factor induced by a drug already in the blood.
+     * It is expected that no drug doses are taken over the period for which
+     * this function calculates a drug factor.
+     * 
+     * @param neg_elim_rate -k (sampled)
+     * @param C0 Concentration of drug in blood at start of period. Will be
+     *  updated to correct concentration at end of period. Units: mg/l
[…]

### 2018-10-29 code/openmalaria/openmalaria-schema-39.0/model/Simulator.h

@@ -1,0 +1,83 @@
+/* This file is part of OpenMalaria.
+ * 
+ * Copyright (C) 2005-2015 Swiss Tropical and Public Health Institute
+ * Copyright (C) 2005-2015 Liverpool School Of Tropical Medicine
+ * 
+ * OpenMalaria is free software; you can redistribute it and/or modify
+ * it under the terms of the GNU General Public License as published by
+ * the Free Software Foundation; either version 2 of the License, or (at
+ * your option) any later version.
+ * 
+ * This program is distributed in the hope that it will be useful, but
+ * WITHOUT ANY WARRANTY; without even the implied warranty of
+ * MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the GNU
+ * General Public License for more details.
+ * 
+ * You should have received a copy of the GNU General Public License
+ * along with this program; if not, write to the Free Software
+ * Foundation, Inc., 51 Franklin Street, Fifth Floor, Boston, MA 02110-1301, USA.
+ */
+
+#ifndef Hmod_Simulator
+#define Hmod_Simulator
+
+#include "Global.h"
+#include "Population.h"
+#include "Transmission/TransmissionModel.h"
+#include "util/BoincWrapper.h"
+using namespace std;
+
+namespace scnXml{
+    class Monitoring;
+    class Scenario;
+}
+namespace OM {
+namespace interventions{
+    class InterventionManager;
+}
+    
+//! Main simulation class
+class Simulator{
+public: 
+    //!  Inititalise all step specific constants and variables.
+    Simulator( util::Checksum ck, const scnXml::Scenario& scenario );
+    
+    //! Entry point to simulation.
+    void start(const scnXml::Monitoring& monitoring);
+    
+    /// Return true when this simulation started by loading a checkpoint
+    inline static bool isCheckpoint(){ return startedFromCheckpoint; }
+    
+private:
+    /** @brief checkpointing functions
+    *
+    * readCheckpoint/writeCheckpoint prepare to read/write the file,
+    * and read/write read and write the actual data. */
+    //@{
+    void writeCheckpoint();
+    void readCheckpoint();
+    
+    void checkpoint (istream& stream, int checkpointNum);
+    void checkpoint (ostream& stream, int checkpointNum);
+    //@}
+    
+    // Data
+    SimTime simPeriodEnd;
+    SimTime totalSimDuration;
+    int phase;  // only need be a class member because value is checkpointed
+    
[…]

### 2018-10-29 code/openmalaria/openmalaria-schema-39.0/model/Transmission/Anopheles/EmergenceModel.h

@@ -1,0 +1,204 @@
+/* This file is part of OpenMalaria.
+ * 
+ * Copyright (C) 2005-2015 Swiss Tropical and Public Health Institute
+ * Copyright (C) 2005-2015 Liverpool School Of Tropical Medicine
+ * 
+ * OpenMalaria is free software; you can redistribute it and/or modify
+ * it under the terms of the GNU General Public License as published by
+ * the Free Software Foundation; either version 2 of the License, or (at
+ * your option) any later version.
+ * 
+ * This program is distributed in the hope that it will be useful, but
+ * WITHOUT ANY WARRANTY; without even the implied warranty of
+ * MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the GNU
+ * General Public License for more details.
+ * 
+ * You should have received a copy of the GNU General Public License
+ * along with this program; if not, write to the Free Software
+ * Foundation, Inc., 51 Franklin Street, Fifth Floor, Boston, MA 02110-1301, USA.
+ */
+
+#ifndef Hmod_Anopheles_EmergenceModel
+#define Hmod_Anopheles_EmergenceModel
+
+#include "Global.h"
+#include "schema/interventions.h"
+#include "util/SimpleDecayingValue.h"
+#include "util/checkpoint_containers.h"
+#include "util/vecDay.h"
+
+namespace OM {
+namespace Transmission {
+namespace Anopheles {
+
+using namespace std;
+using util::vecDay;
+
+// forward declare to avoid circular dependency:
+class MosqTransmission;
+
+/** Part of vector anopheles model, giving emergence of adult mosquitoes from
+ * water bodies.
+ * 
+ * This is an abstract class (i.e. interface). The following implementations
+ * exist: FixedEmergence.
+ */
+class EmergenceModel
+{
+public:
+    ///@brief Initialisation and destruction
+    //@{
+    EmergenceModel();
+    virtual ~EmergenceModel() {}
+    
+    /** Called to initialise life-cycle parameters from XML data.
+     * 
+     * Default implementation throws an assertion: only call if the life cycle
+     * model is used. */
+    virtual void initLifeCycle(const scnXml::LifeCycle& lcData){
+        assert (false);
+    }
+    
+    /** Called by initialise function to init variables directly related to EIR
+     * 
+     * @param anoph Data from XML
+     * @param initialisationEIR In/out parameter: TransmissionModel::initialisationEIR
[…]

### 2018-10-29 code/openmalaria/openmalaria-schema-39.0/model/Transmission/PerHost.cpp

@@ -1,0 +1,155 @@
+/* This file is part of OpenMalaria.
+ * 
+ * Copyright (C) 2005-2015 Swiss Tropical and Public Health Institute
+ * Copyright (C) 2005-2015 Liverpool School Of Tropical Medicine
+ * 
+ * OpenMalaria is free software; you can redistribute it and/or modify
+ * it under the terms of the GNU General Public License as published by
+ * the Free Software Foundation; either version 2 of the License, or (at
+ * your option) any later version.
+ * 
+ * This program is distributed in the hope that it will be useful, but
+ * WITHOUT ANY WARRANTY; without even the implied warranty of
+ * MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the GNU
+ * General Public License for more details.
+ * 
+ * You should have received a copy of the GNU General Public License
+ * along with this program; if not, write to the Free Software
+ * Foundation, Inc., 51 Franklin Street, Fifth Floor, Boston, MA 02110-1301, USA.
+ */
+
+#include "Transmission/PerHost.h"
+#include "Transmission/VectorModel.h"
+#include "Transmission/Anopheles/PerHostAnoph.h"
+#include "interventions/InterventionManager.hpp"
+#include "util/errors.h"
+#include "util/checkpoint.h"
+
+namespace OM {
+namespace Transmission {
+using namespace OM::util;
+using Anopheles::PerHostAnophParams;
+
+// -----  PerHost static  -----
+
+AgeGroupInterpolator PerHost::relAvailAge;
+
+void PerHost::init ( const scnXml::AgeGroupValues& availabilityToMosquitoes ) {
+    relAvailAge.set( availabilityToMosquitoes, "availabilityToMosquitoes" );
+}
+
+// -----  PerHost non-static -----
+
+PerHost::PerHost () :
+        outsideTransmission(false),
+        _relativeAvailabilityHet(numeric_limits<double>::signaling_NaN())
+{
+}
+void PerHost::initialise (double availabilityFactor) {
+    _relativeAvailabilityHet = availabilityFactor;
+    VectorModel* vTM = dynamic_cast<VectorModel*> (&sim::transmission());
+    if (vTM != 0) {
+        species.resize (vTM->numSpecies);
+        for(size_t i = 0; i < vTM->numSpecies; ++i)
+            species[i].initialise (vTM->species[i].getHumanBaseParams(), availabilityFactor);
+    }
+}
+
[…]

### 2018-10-29 code/openmalaria/openmalaria-schema-39.0/model/Transmission/PerHost.h

@@ -1,0 +1,291 @@
+/* This file is part of OpenMalaria.
+ * 
+ * Copyright (C) 2005-2015 Swiss Tropical and Public Health Institute
+ * Copyright (C) 2005-2015 Liverpool School Of Tropical Medicine
+ * 
+ * OpenMalaria is free software; you can redistribute it and/or modify
+ * it under the terms of the GNU General Public License as published by
+ * the Free Software Foundation; either version 2 of the License, or (at
+ * your option) any later version.
+ * 
+ * This program is distributed in the hope that it will be useful, but
+ * WITHOUT ANY WARRANTY; without even the implied warranty of
+ * MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the GNU
+ * General Public License for more details.
+ * 
+ * You should have received a copy of the GNU General Public License
+ * along with this program; if not, write to the Free Software
+ * Foundation, Inc., 51 Franklin Street, Fifth Floor, Boston, MA 02110-1301, USA.
+ */
+
+#ifndef Hmod_PerHost
+#define Hmod_PerHost
+
+#include "Global.h"
+#include "Transmission/Anopheles/PerHostAnoph.h"
+#include "interventions/Interfaces.hpp"
+#include "util/AgeGroupInterpolation.h"
+#include "util/DecayFunction.h"
+#include "util/checkpoint_containers.h"
+#include <boost/ptr_container/ptr_list.hpp>
+
+namespace OM {
+namespace Transmission {
+
+using Anopheles::PerHostAnophParams;
+using Anopheles::PerHostAnoph;
+using util::AgeGroupInterpolator;
+using util::DecayFunction;
+using util::DecayFuncHet;
+
+class HumanVectorInterventionComponent;
+
+/**
+ * A base class for interventions affecting human-vector interaction.
+ * 
+ * The constructor should initialise the data to represent an intervention
+ * deployed at this time (sim::now()).
+ * 
+ * redeploy() should reset the intervention to a freshly deployed state. If
+ * necessary, PerHost::deployComponent can be updated to make it create a new
+ * instance instead of calling redeploy.
+ */
+class PerHostInterventionData {
+public:
+    virtual ~PerHostInterventionData() {}
+    
+    /** Deploy an intervention. */
+    virtual void redeploy( const HumanVectorInterventionComponent& params ) =0;
+    
+    /** Per time step update. Used by ITNs to update hole decay. */
+    virtual void update(Host::Human& human) =0;
+    
[…]

### 2018-10-29 code/openmalaria/openmalaria-schema-39.0/model/Transmission/VectorModel.cpp

@@ -1,0 +1,579 @@
+/* This file is part of OpenMalaria.
+ * 
+ * Copyright (C) 2005-2015 Swiss Tropical and Public Health Institute
+ * Copyright (C) 2005-2015 Liverpool School Of Tropical Medicine
+ * 
+ * OpenMalaria is free software; you can redistribute it and/or modify
+ * it under the terms of the GNU General Public License as published by
+ * the Free Software Foundation; either version 2 of the License, or (at
+ * your option) any later version.
+ * 
+ * This program is distributed in the hope that it will be useful, but
+ * WITHOUT ANY WARRANTY; without even the implied warranty of
+ * MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the GNU
+ * General Public License for more details.
+ * 
+ * You should have received a copy of the GNU General Public License
+ * along with this program; if not, write to the Free Software
+ * Foundation, Inc., 51 Franklin Street, Fifth Floor, Boston, MA 02110-1301, USA.
+ */
+
+#include "Transmission/VectorModel.h"
+#include "Population.h"
+#include "Host/Human.h"
+#include "WithinHost/WHInterface.h"
+#include "WithinHost/Genotypes.h"
+#include "Monitoring/Continuous.h"
+#include "util/vectors.h"
+#include "util/ModelOptions.h"
+#include "util/SpeciesIndexChecker.h"
+
+#include <fstream>
+#include <map>
+#include <cmath>
+#include <set>
+
+namespace OM {
+namespace Transmission {
+using namespace OM::util;
+using Anopheles::PerHostAnophParams;
+
+void VectorModel::ctsCbN_v0 (ostream& stream) {
+    for(size_t i = 0; i < numSpecies; ++i)
+        stream << '\t' << species[i].getLastN_v0();
+}
+void VectorModel::ctsCbP_A (ostream& stream) {
+    for(size_t i = 0; i < numSpecies; ++i)
+        stream << '\t' << species[i].getLastVecStat(Anopheles::PA);
+}
+void VectorModel::ctsCbP_df (ostream& stream) {
+    for(size_t i = 0; i < numSpecies; ++i)
+        stream << '\t' << species[i].getLastVecStat(Anopheles::PDF);
+}
+void VectorModel::ctsCbP_dif (ostream& stream) {
+    for(size_t i = 0; i < numSpecies; ++i)
[…]

### 2018-10-29 code/openmalaria/openmalaria-schema-39.0/model/WithinHost/Diagnostic.cpp

@@ -1,0 +1,197 @@
+/* This file is part of OpenMalaria.
+ * 
+ * Copyright (C) 2005-2015 Swiss Tropical and Public Health Institute
+ * Copyright (C) 2005-2015 Liverpool School Of Tropical Medicine
+ * 
+ * OpenMalaria is free software; you can redistribute it and/or modify
+ * it under the terms of the GNU General Public License as published by
+ * the Free Software Foundation; either version 2 of the License, or (at
+ * your option) any later version.
+ * 
+ * This program is distributed in the hope that it will be useful, but
+ * WITHOUT ANY WARRANTY; without even the implied warranty of
+ * MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the GNU
+ * General Public License for more details.
+ * 
+ * You should have received a copy of the GNU General Public License
+ * along with this program; if not, write to the Free Software
+ * Foundation, Inc., 51 Franklin Street, Fifth Floor, Boston, MA 02110-1301, USA.
+ */
+
+#include "WithinHost/Diagnostic.h"
+#include "Parameters.h"
+#include "util/random.h"
+#include "util/errors.h"
+#include "util/ModelOptions.h"
+#include "util/CommandLine.h"
+#include "schema/scenario.h"
+#include <limits>
+#include <boost/ptr_container/ptr_map.hpp>
+
+namespace OM { namespace WithinHost {
+
+// ———  Diagnostic (non-static)  ———
+
+Diagnostic::Diagnostic( const Parameters& parameters, const scnXml::Diagnostic& elt ){
+    dens_lim = numeric_limits<double>::quiet_NaN();
+    specificity = numeric_limits<double>::quiet_NaN();
+    
+    if( util::ModelOptions::option( util::VIVAX_SIMPLE_MODEL ) ){
+        return; // both parameters NaN (set above)
+    }
+    
+    if( elt.getDeterministic().present() ){
+        dens_lim = elt.getDeterministic().get().getMinDensity();
+    }else if( elt.getStochastic().present() ){
+        dens_lim = elt.getStochastic().get().getDens_50();
+        if( dens_lim == 0.0 ){
+            // The equation used for stochastic diagnostics breaks down when
+            // dens=dens_lim=0 and for other cases the deterministic model is
+            // the same when dens_lim=0.
+        }else{
+            specificity = elt.getStochastic().get().getSpecificity();
+            if( specificity < 0.0 || specificity > 1.0 ){
+                throw util::xml_scenario_error(
[…]

### 2018-10-29 code/openmalaria/openmalaria-schema-39.0/model/WithinHost/Infection/PennyInfection.h

@@ -1,0 +1,93 @@
+/* This file is part of OpenMalaria.
+ * 
+ * Copyright (C) 2005-2015 Swiss Tropical and Public Health Institute
+ * Copyright (C) 2005-2015 Liverpool School Of Tropical Medicine
+ * 
+ * OpenMalaria is free software; you can redistribute it and/or modify
+ * it under the terms of the GNU General Public License as published by
+ * the Free Software Foundation; either version 2 of the License, or (at
+ * your option) any later version.
+ * 
+ * This program is distributed in the hope that it will be useful, but
+ * WITHOUT ANY WARRANTY; without even the implied warranty of
+ * MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the GNU
+ * General Public License for more details.
+ * 
+ * You should have received a copy of the GNU General Public License
+ * along with this program; if not, write to the Free Software
+ * Foundation, Inc., 51 Franklin Street, Fifth Floor, Boston, MA 02110-1301, USA.
+ */
+
+/*This model refers to the paper:
+ *PENNY et al  (2011). The potential effects of blood stage vaccines on the within-host dynamics of Plasmodium Falciparum */
+
+#ifndef Hmod_PENNYINFECTION_H
+#define Hmod_PENNYINFECTION_H
+
+#include "WithinHost/Infection/CommonInfection.h"
+
+class PennyInfectionSuite;
+
+namespace OM { namespace WithinHost {
+
+class PennyInfection : public CommonInfection {
+public:
+    /// Static initialization (happens once)
+    static void init();
+    
+    /// Constructor
+    PennyInfection(uint32_t protID);
+    /// Resume from a checkpoint
+    PennyInfection (istream& stream);
+    /// Destructor
+    virtual ~PennyInfection () {};
+    
+    virtual bool updateDensity( double survivalFactor, SimTime bsAge, double );
+    
+    /** Get the density of sequestered parasites. */
+    inline double seqDensity(int ageDays){
+        size_t todayV = mod_nn(ageDays, delta_V);
+        return seqDensities[todayV];
+    }
+
+protected:
+    virtual void checkpoint (ostream& stream);
+
+private:
+    // function to obtain the summation component of variant specific immunity
+    double getVariantSpecificSummation(int ageDays);
+    // function to obtain the summation component of clonal immunity
+    double getClonalSummation(int ageDays);
+    
+    // delta_C := delay to clonal antibody response (days) (value 7.2038 round to 7)  
+    static const int delta_C=7;
[…]

### 2018-10-29 code/openmalaria/openmalaria-schema-39.0/model/WithinHost/Pathogenesis/PathogenesisModel.h

@@ -1,0 +1,91 @@
+/* This file is part of OpenMalaria.
+ * 
+ * Copyright (C) 2005-2015 Swiss Tropical and Public Health Institute
+ * Copyright (C) 2005-2015 Liverpool School Of Tropical Medicine
+ * 
+ * OpenMalaria is free software; you can redistribute it and/or modify
+ * it under the terms of the GNU General Public License as published by
+ * the Free Software Foundation; either version 2 of the License, or (at
+ * your option) any later version.
+ * 
+ * This program is distributed in the hope that it will be useful, but
+ * WITHOUT ANY WARRANTY; without even the implied warranty of
+ * MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the GNU
+ * General Public License for more details.
+ * 
+ * You should have received a copy of the GNU General Public License
+ * along with this program; if not, write to the Free Software
+ * Foundation, Inc., 51 Franklin Street, Fifth Floor, Boston, MA 02110-1301, USA.
+ */
+
+#ifndef Hmod_PathogenesisModel
+#define Hmod_PathogenesisModel
+
+#include "Global.h"
+#include "Parameters.h"
+#include "WithinHost/Pathogenesis/State.h"
+
+namespace scnXml{
+    class HSESNMF;
+    class Clinical;
+}
+
+namespace OM {
+    namespace Host { class Human; }
+    namespace WithinHost { namespace Pathogenesis {
+
+/*! PathogenesisModel abstract base class.
+ *
+ * Previously named MorbidityModel and PresentationModel. */
+class PathogenesisModel {
+public:
+    /// Calls static init on correct PathogenesisModel.
+    static void init( const Parameters& parameters, const scnXml::Clinical& clinical, bool nmfOnly );
+
+    /** Create a sub-class instance, dependant on global options.
+    *
+    * @param cF = Comorbidity factor (currently set in Human). */
+    static PathogenesisModel* createPathogenesisModel(double cF);
+    
+    
+    virtual ~PathogenesisModel() {}
+
+    /** Determines the health of the individual based on his/her parasitemia.
+     *
+     * May introduce severe or uncomplicated cases of malaria, as well as non-
+     * malaria fevers. */
+    StatePair determineState( Host::Human& human, double ageYears,
+            double timeStepMaxDensity, double endDensity, bool isDoomed );
+    
+    /** For Vivax: determine the chance of a NMF and sample, returning either
[…]

### 2018-10-29 code/openmalaria/openmalaria-schema-39.0/model/WithinHost/Pathogenesis/Submodels.cpp

@@ -1,0 +1,127 @@
+/* This file is part of OpenMalaria.
+ * 
+ * Copyright (C) 2005-2015 Swiss Tropical and Public Health Institute
+ * Copyright (C) 2005-2015 Liverpool School Of Tropical Medicine
+ * 
+ * OpenMalaria is free software; you can redistribute it and/or modify
+ * it under the terms of the GNU General Public License as published by
+ * the Free Software Foundation; either version 2 of the License, or (at
+ * your option) any later version.
+ * 
+ * This program is distributed in the hope that it will be useful, but
+ * WITHOUT ANY WARRANTY; without even the implied warranty of
+ * MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the GNU
+ * General Public License for more details.
+ * 
+ * You should have received a copy of the GNU General Public License
+ * along with this program; if not, write to the Free Software
+ * Foundation, Inc., 51 Franklin Street, Fifth Floor, Boston, MA 02110-1301, USA.
+ */
+
+#include "WithinHost/Pathogenesis/Submodels.h"
+#include "Parameters.h"
+#include "mon/reporting.h"
+
+#include <cmath>
+using namespace std;
+
+namespace OM {
+namespace WithinHost {
+namespace Pathogenesis {
+
+// ———  Müller presentation model  ———
+
+// Müller model parameters
+double rateMultiplier_31;
+double densityExponent_32;
+
+void MuellerPathogenesis::init( const Parameters& parameters ){
+    rateMultiplier_31 = parameters[Parameters::MUELLER_RATE_MULTIPLIER]
+        * SimTime::yearsPerStep();
+    densityExponent_32 = parameters[Parameters::MUELLER_DENSITY_EXPONENT];
+}
+
+double MuellerPathogenesis::getPEpisode(double, double totalDensity) {
+    double incidenceDensity = rateMultiplier_31
+            * (pow(totalDensity, densityExponent_32));
+    return 1.0 - exp(-incidenceDensity);
+}
+
+
+// ———  Pyrogenic threshold model  ———
+
+//Pyrogenic threshold at birth (Y*0)
+double initPyroThres;
+// Ystar2: critical value in determining increase in pyrogenic threshold
+double Ystar2_13;
+double Ystar1_26;
+
[…]

### 2018-10-29 code/openmalaria/openmalaria-schema-39.0/model/WithinHost/Pathogenesis/Submodels.h

@@ -1,0 +1,78 @@
+/* This file is part of OpenMalaria.
+ * 
+ * Copyright (C) 2005-2015 Swiss Tropical and Public Health Institute
+ * Copyright (C) 2005-2015 Liverpool School Of Tropical Medicine
+ * 
+ * OpenMalaria is free software; you can redistribute it and/or modify
+ * it under the terms of the GNU General Public License as published by
+ * the Free Software Foundation; either version 2 of the License, or (at
+ * your option) any later version.
+ * 
+ * This program is distributed in the hope that it will be useful, but
+ * WITHOUT ANY WARRANTY; without even the implied warranty of
+ * MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the GNU
+ * General Public License for more details.
+ * 
+ * You should have received a copy of the GNU General Public License
+ * along with this program; if not, write to the Free Software
+ * Foundation, Inc., 51 Franklin Street, Fifth Floor, Boston, MA 02110-1301, USA.
+ */
+
+#ifndef Hmod_WH_Path_Submodels
+#define Hmod_WH_Path_Submodels
+
+#include "Global.h"
+#include "WithinHost/Pathogenesis/PathogenesisModel.h"
+
+using namespace std;
+
+namespace OM { namespace WithinHost { namespace Pathogenesis {
+
+/// Müller presentation model.
+class MuellerPathogenesis : public PathogenesisModel {
+public:
+    MuellerPathogenesis(double cF) :
+        PathogenesisModel(cF) {}
+    ~MuellerPathogenesis() {}
+    
+    virtual double getPEpisode(double timeStepMaxDensity, double totalDensity);
+
+    /// Read parameters
+    static void init( const Parameters& parameters );
+};
+
+/// Pyrogenic threshold presentation model.
+class PyrogenPathogenesis : public PathogenesisModel {
+protected:
+    /// Critical density for fever (clinical episodes)
+    double _pyrogenThres;
+    /// Determine the current pyrogenic threshold.
+    virtual void updatePyrogenThres(double totalDensity);
+
+public:
+    PyrogenPathogenesis(double cF);
+    virtual ~PyrogenPathogenesis() {}
+    virtual void summarize (const Host::Human& human);
+    virtual double getPEpisode(double timeStepMaxDensity, double totalDensity);
+    
+    /// Read parameters from XML
+    static void init( const OM::Parameters& parameters );
+    
[…]

### 2018-10-29 code/openmalaria/openmalaria-schema-39.0/model/util/SpeciesIndexChecker.h

@@ -1,0 +1,51 @@
+/* This file is part of OpenMalaria.
+ * 
+ * Copyright (C) 2005-2015 Swiss Tropical and Public Health Institute
+ * Copyright (C) 2005-2015 Liverpool School Of Tropical Medicine
+ * 
+ * OpenMalaria is free software; you can redistribute it and/or modify
+ * it under the terms of the GNU General Public License as published by
+ * the Free Software Foundation; either version 2 of the License, or (at
+ * your option) any later version.
+ * 
+ * This program is distributed in the hope that it will be useful, but
+ * WITHOUT ANY WARRANTY; without even the implied warranty of
+ * MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the GNU
+ * General Public License for more details.
+ * 
+ * You should have received a copy of the GNU General Public License
+ * along with this program; if not, write to the Free Software
+ * Foundation, Inc., 51 Franklin Street, Fifth Floor, Boston, MA 02110-1301, USA.
+ */
+
+#ifndef OM_UTIL_SPECIES_INDEX_CHECKER
+#define OM_UTIL_SPECIES_INDEX_CHECKER
+
+#include "Global.h"
+
+#include <set>
+
+namespace OM { namespace util {
+using std::set;
+
+/** Checks keys exist and that all are used. Fails with an xml_scenario_error
+ * and a suitable error message. */
+class SpeciesIndexChecker{
+public:
+    SpeciesIndexChecker( const string& intervName, const map<string,size_t>& indices ) :
+        _intervName( intervName ), _indices( indices ){}
+    
+    /** Return the index in speciesIndex of mosquito, throwing if not found. */
+    size_t getIndex( string species );
+    
+    /** Throw if some species was missed. */
+    void checkNoneMissed() const;
+    
+private:
+    const string& _intervName;
+    const map<string,size_t>& _indices;
+    set<string> found;
+};
+
+} }
+#endif

### 2018-10-29 code/openmalaria/openmalaria-schema-39.0/model/util/checkpoint_containers.h

@@ -1,0 +1,134 @@
+/* This file is part of OpenMalaria.
+ * 
+ * Copyright (C) 2005-2015 Swiss Tropical and Public Health Institute
+ * Copyright (C) 2005-2015 Liverpool School Of Tropical Medicine
+ * 
+ * OpenMalaria is free software; you can redistribute it and/or modify
+ * it under the terms of the GNU General Public License as published by
+ * the Free Software Foundation; either version 2 of the License, or (at
+ * your option) any later version.
+ * 
+ * This program is distributed in the hope that it will be useful, but
+ * WITHOUT ANY WARRANTY; without even the implied warranty of
+ * MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the GNU
+ * General Public License for more details.
+ * 
+ * You should have received a copy of the GNU General Public License
+ * along with this program; if not, write to the Free Software
+ * Foundation, Inc., 51 Franklin Street, Fifth Floor, Boston, MA 02110-1301, USA.
+ */
+
+#ifndef OM_util_checkpoint_containers
+#define OM_util_checkpoint_containers
+
+#ifndef Hmod_Global
+#error "Please include Global.h first."
+// otherwise "using ..." declaration in Global.h won't work
+#endif
+
+/** Provides some extra functions. See checkpoint.h. */
+namespace OM {
+namespace util {
+namespace checkpoint {
+
+    ///@brief Operator& for stl containers
+    //@{
+    template<class U, class V, class S>
+    void operator& (pair<U,V> x, S& stream) {
+        x.first & stream;
+        x.second & stream;
+    }
+    
+    template<class T>
+    void operator& (vector<T>& x, ostream& stream) {
+        x.size() & stream;
+        foreach (T& y, x) {
+            y & stream;
+        }
+    }
+    template<class T>
+    void operator& (vector<T>& x, istream& stream) {
+        size_t l;
+        l & stream;
+        validateListSize (l);
+        x.resize (l);
+        foreach (T& y, x) {
+            y & stream;
+        }
+    }
+    /// Version of above taking an element to initialize each element from.
+    template<class T>
+    void checkpoint (vector<T>& x, istream& stream, T templateInstance) {
+        size_t l;
+        l & stream;
+        validateListSize (l);
+        x.resize (l, templateInstance);
+        foreach (T& y, x) {
+            y & stream;
+        }
+    }
+    
+    template<class T>
+    void operator& (list<T> x, ostream& stream) {
[…]

### 2018-10-29 code/openmalaria/openmalaria-schema-39.0/model/util/vectors.h

@@ -1,0 +1,198 @@
+/* This file is part of OpenMalaria.
+ * 
+ * Copyright (C) 2005-2015 Swiss Tropical and Public Health Institute
+ * Copyright (C) 2005-2015 Liverpool School Of Tropical Medicine
+ * 
+ * OpenMalaria is free software; you can redistribute it and/or modify
+ * it under the terms of the GNU General Public License as published by
+ * the Free Software Foundation; either version 2 of the License, or (at
+ * your option) any later version.
+ * 
+ * This program is distributed in the hope that it will be useful, but
+ * WITHOUT ANY WARRANTY; without even the implied warranty of
+ * MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the GNU
+ * General Public License for more details.
+ * 
+ * You should have received a copy of the GNU General Public License
+ * along with this program; if not, write to the Free Software
+ * Foundation, Inc., 51 Franklin Street, Fifth Floor, Boston, MA 02110-1301, USA.
+ */
+
+#ifndef Hmod_UtilVectors
+#define Hmod_UtilVectors
+
+#include "Global.h"
+#include "util/checkpoint_containers.h"
+#include <gsl/gsl_vector.h>
+#include "util/checkpoint_containers.h"
+
+#ifndef M_PI
+#define M_PI 3.14159265358979323846
+#endif
+
+namespace OM {
+namespace util {
+
+/** Various utilities acting on vectors. */
+namespace vectors {
+
+  ///@brief Basic operations on std::vector
+  //@{
+  /// Scale all elements of a vector by a in-situ
+  void scale (vector<double>& vec, double a);
+  
+  /// Return sum of all elements
+  double sum (const vector<double>& vec);
+  
+  /// Return sum of all elements
+  double sum (const gsl_vector *vec);
+  
+  /// Return arithmetric mean
+  inline double mean (const vector<double>& vec){
+    return sum(vec) / vec.size();
+  }
+  /// Return arithmetric mean
+  inline double mean (const gsl_vector *vec){
+    return sum(vec) / vec->size;
+  }
+  
+  /// Add one vector into another (x += y)
+  void addTo (vector<double>& x, vector<double>& y);
+  //@}
+  
+  
+  ///@brief Comparissons on std::vector
+  //@{
+  /** Return true if, approximately, a == b
+   *
+   * In detail: true when (fabs(a-b) <= max(fabs(a),fabs(b)) * lim_fact). */
[…]

### 2018-10-29 code/openmalaria/openmalaria-schema-39.0/unittest/AgeGroupInterpolationSuite.h

@@ -1,0 +1,108 @@
+/*
+ This file is part of OpenMalaria.
+ 
+ Copyright (C) 2005-2014 Swiss Tropical and Public Health Institute
+ Copyright (C) 2005-2014 Liverpool School Of Tropical Medicine
+ 
+ OpenMalaria is free software; you can redistribute it and/or modify
+ it under the terms of the GNU General Public License as published by
+ the Free Software Foundation; either version 2 of the License, or (at
+ your option) any later version.
+ 
+ This program is distributed in the hope that it will be useful, but
+ WITHOUT ANY WARRANTY; without even the implied warranty of
+ MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the GNU
+ General Public License for more details.
+ 
+ You should have received a copy of the GNU General Public License
+ along with this program; if not, write to the Free Software
+ Foundation, Inc., 51 Franklin Street, Fifth Floor, Boston, MA 02110-1301, USA.
+*/
+
+#ifndef Hmod_AgeGroupInterpolationSuite
+#define Hmod_AgeGroupInterpolationSuite
+
+#include <cxxtest/TestSuite.h>
+#include "UnittestUtil.h"
+#include "ExtraAsserts.h"
+#include "util/AgeGroupInterpolation.h"
+
+using util::AgeGroupInterpolator;
+
+class AgeGroupInterpolationSuite : public CxxTest::TestSuite
+{
+public:
+    AgeGroupInterpolationSuite () {
+        // Set maximum age to 90 years:
+        UnittestUtil::initTime( 5 );
+        agvElt = unique_ptr<scnXml::AgeGroupValues>(new scnXml::AgeGroupValues());
+        scnXml::AgeGroupValues::GroupSequence& seq = agvElt->getGroup();
+        seq.resize( dataLen, scnXml::Group(0.0,0.0) );
+        for( size_t i = 0; i < dataLen; ++i ){
+            seq[ i ].setLowerbound( stdLbounds[ i ] );
+            seq[ i ].setValue( stdValues[ i ] );
+        }
+    }
+    ~AgeGroupInterpolationSuite () {
+    }
+    
+    void setUp () {
+    }
+    void tearDown () {
+    }
+    
+    void testDummy() {
+        AgeGroupInterpolator o;
+        TS_ASSERT_THROWS( o.eval(5.7), const std::exception& );
+    }
+    
+    void testPiecewiseConst () {
[…]

### 2018-10-29 code/openmalaria/openmalaria-schema-39.0/unittest/DecayFunctionSuite.h

@@ -1,0 +1,138 @@
+/*
+ This file is part of OpenMalaria.
+ 
+ Copyright (C) 2005-2014 Swiss Tropical and Public Health Institute
+ Copyright (C) 2005-2014 Liverpool School Of Tropical Medicine
+ 
+ OpenMalaria is free software; you can redistribute it and/or modify
+ it under the terms of the GNU General Public License as published by
+ the Free Software Foundation; either version 2 of the License, or (at
+ your option) any later version.
+ 
+ This program is distributed in the hope that it will be useful, but
+ WITHOUT ANY WARRANTY; without even the implied warranty of
+ MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the GNU
+ General Public License for more details.
+ 
+ You should have received a copy of the GNU General Public License
+ along with this program; if not, write to the Free Software
+ Foundation, Inc., 51 Franklin Street, Fifth Floor, Boston, MA 02110-1301, USA.
+*/
+
+#ifndef Hmod_DecayFunctionSuite
+#define Hmod_DecayFunctionSuite
+
+#include <cxxtest/TestSuite.h>
+#include "UnittestUtil.h"
+#include "ExtraAsserts.h"
+#include "util/DecayFunction.h"
+
+using ::OM::util::DecayFunction;
+using ::OM::util::DecayFuncHet;
+
+class DecayFunctionSuite : public CxxTest::TestSuite
+{
+public:
+    DecayFunctionSuite () :
+        dfElt( "" )
+    {
+        dfElt.setL( "10y" );
+        dfElt.setK( 1.6 );
+    }
+    
+    void setUp() {
+        UnittestUtil::initTime(5);
+    }
+    
+    void testBad() {
+        dfElt.setFunction( "unknown" );
+        string errMsg = "decay function type unknown of DecayFunctionSuite unrecognized";
+        TS_ASSERT_THROWS_EQUALS( df = DecayFunction::makeObject( dfElt, "DecayFunctionSuite" ),
+                          const std::runtime_error &e, e.what(), errMsg );
+    }
+    
+    void testConstant () {
+        dfElt.setFunction( "constant" );
[…]

### 2018-10-29 code/openmalaria/openmalaria-schema-39.0/unittest/EmpiricalInfectionSuite.h

@@ -1,0 +1,119 @@
+/*
+ This file is part of OpenMalaria.
+ 
+ Copyright (C) 2005-2014 Swiss Tropical and Public Health Institute
+ Copyright (C) 2005-2014 Liverpool School Of Tropical Medicine
+ 
+ OpenMalaria is free software; you can redistribute it and/or modify
+ it under the terms of the GNU General Public License as published by
+ the Free Software Foundation; either version 2 of the License, or (at
+ your option) any later version.
+ 
+ This program is distributed in the hope that it will be useful, but
+ WITHOUT ANY WARRANTY; without even the implied warranty of
+ MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the GNU
+ General Public License for more details.
+ 
+ You should have received a copy of the GNU General Public License
+ along with this program; if not, write to the Free Software
+ Foundation, Inc., 51 Franklin Street, Fifth Floor, Boston, MA 02110-1301, USA.
+*/
+
+#ifndef Hmod_EmpiricalInfectionSuite
+#define Hmod_EmpiricalInfectionSuite
+
+#include <cxxtest/TestSuite.h>
+#include "UnittestUtil.h"
+#include "ExtraAsserts.h"
+#include "WithinHost/Infection/EmpiricalInfection.h"
+#include "WithinHost/CommonWithinHost.h"
+#include "util/random.h"
+#include <limits>
+
+using namespace OM::WithinHost;
+
+class EmpiricalInfectionSuite : public CxxTest::TestSuite
+{
+public:
+    void setUp () {
+        UnittestUtil::initTime(1);
+        UnittestUtil::Infection_init_latentP_and_NaN ();
+        EmpiricalInfection::init();
+        util::random::seed (83);	// seed is unimportant, but must be fixed
+        // pkpdID (1st value) isn't important since we're not using drug model here:
+        infection = CommonWithinHost::createInfection( 0xFFFFFFFF );
+        for( SimTime d = sim::ts1(), end = sim::ts1() + SimTime::fromDays(15); d < end; d += SimTime::oneDay() ){
+            // blood stage starts 15 days after creation
+            UnittestUtil::incrTime( SimTime::oneDay() );
+            infection->update( 1.0, d, numeric_limits<double>::quiet_NaN() );
+        }
+    }
+    void tearDown () {
+        delete infection;
+    }
+
[…]

### 2018-10-29 code/openmalaria/openmalaria-schema-39.0/unittest/LSTMPkPdSuite.h

@@ -1,0 +1,100 @@
+/*
+ This file is part of OpenMalaria.
+ 
+ Copyright (C) 2005-2014 Swiss Tropical and Public Health Institute
+ Copyright (C) 2005-2014 Liverpool School Of Tropical Medicine
+ 
+ OpenMalaria is free software; you can redistribute it and/or modify
+ it under the terms of the GNU General Public License as published by
+ the Free Software Foundation; either version 2 of the License, or (at
+ your option) any later version.
+ 
+ This program is distributed in the hope that it will be useful, but
+ WITHOUT ANY WARRANTY; without even the implied warranty of
+ MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the GNU
+ General Public License for more details.
+ 
+ You should have received a copy of the GNU General Public License
+ along with this program; if not, write to the Free Software
+ Foundation, Inc., 51 Franklin Street, Fifth Floor, Boston, MA 02110-1301, USA.
+*/
+// Unittest for the LSTM drug model
+
+#ifndef Hmod_LSTMPkPdSuite
+#define Hmod_LSTMPkPdSuite
+
+#include <cxxtest/TestSuite.h>
+#include "PkPd/LSTMModel.h"
+#include "UnittestUtil.h"
+#include "ExtraAsserts.h"
+#include <limits>
+
+using namespace OM;
+using namespace OM::PkPd;
+
+/// There is probably little value in this unit-test now that PkPdComplianceSuite exists
+class LSTMPkPdSuite : public CxxTest::TestSuite
+{
+public:
+    LSTMPkPdSuite() :
+            proxy(0)
+    {
+        genotype = 0;                // 0 should work; we definitely don't want random allocation
+        // This is what a previously-used weight distribution gave us,
+        // and is good enough for testing purposes:
+        massAt21 = 55.4993;
+    }
+    
+    void setUp () {
+        UnittestUtil::initTime(1);
+	UnittestUtil::PkPdSuiteSetup();
+	proxy = new LSTMModel ();
+        MQ_index = LSTMDrugType::findDrug( "MQ" );
+    }
+    void tearDown () {
+	delete proxy;
+        LSTMDrugType::clear();
+    }
+    
+    void testNone () {
[…]

### 2018-10-29 code/openmalaria/openmalaria-schema-39.0/util/approxEqual.py

@@ -1,0 +1,126 @@
+#!/usr/bin/env python2
+# -*- coding: utf-8 -*-
+"""
+This file is part of OpenMalaria.
+
+Copyright (C) 2005-2014 Swiss Tropical Institute
+
+OpenMalaria is free software; you can redistribute it and/or modify
+it under the terms of the GNU General Public License as published by
+the Free Software Foundation; either version 2 of the License, or (at
+your option) any later version.
+
+This program is distributed in the hope that it will be useful, but
+WITHOUT ANY WARRANTY; without even the implied warranty of
+MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the GNU
+General Public License for more details.
+
+You should have received a copy of the GNU General Public License
+along with this program; if not, write to the Free Software
+Foundation, Inc., 51 Franklin Street, Fifth Floor, Boston, MA 02110-1301, USA.
+"""
+
+import math
+import unittest
+
+# Could use math.isinf, but it's not present in python 2.5
+def isinf (a):
+    return (a-a) != 0
+def isnan (a):
+    return a != a
+
+# Object to test whether two values are equal to a certain precision
+class ApproxEqual (object):
+    def __init__(self, relPrec, absPrec):
+        self.relPrecision = relPrec
+        self.absPrecision = absPrec
+        self.sumRelDiffs = 0.0
+    
+    # Careful with NaN, +/- inf and 0 values! Note: inf == inf
+    # Check a and b are approximately equal. Return true if:
+    #   a equals b to at least log10(relPrecision) significant figures
+    #   or at least log10(absPrecision) decimal places.
+    # This should work for small and large values, when one is zero, and when
+    # either is infinite or an NaN.
+    # 
+    # What it is limited by, is testing of small values with relative precision
+    # when it should also allow for large values being rounded to zero.
+    # Possibly considering the case where a or b is zero explicitly would help, but
+    # might lead to confusing outcomes.
+    def __call__ (self,a,b):
+        if isinf(a) or isinf(b):
+            relDiff = 1e10000 * 0.0 # NaN
+        else:
+            tolerance = self.relPrecision * max(math.fabs(a),math.fabs(b))
+            tolerance = max(tolerance, self.absPrecision)
+            relDiff = math.fabs(a-b) / tolerance
[…]

### 2018-10-29 code/openmalaria/openmalaria-schema-39.0/util/assimilateForDBImport.py

@@ -1,0 +1,80 @@
+#!/usr/bin/env python2
+# -*- coding: utf-8 -*-
+"""
+This file is part of OpenMalaria.
+
+Copyright (C) 2005-2014 Swiss Tropical Institute
+
+OpenMalaria is free software; you can redistribute it and/or modify
+it under the terms of the GNU General Public License as published by
+the Free Software Foundation; either version 2 of the License, or (at
+your option) any later version.
+
+This program is distributed in the hope that it will be useful, but
+WITHOUT ANY WARRANTY; without even the implied warranty of
+MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the GNU
+General Public License for more details.
+
+You should have received a copy of the GNU General Public License
+along with this program; if not, write to the Free Software
+Foundation, Inc., 51 Franklin Street, Fifth Floor, Boston, MA 02110-1301, USA.
+"""
+
+import string
+import os.path
+import sys
+from optparse import OptionParser
+
+def readSwArmIds(fileName):
+    """Read scenarios.csv file of all arm ids for each sweep per scenario."""
+    handle=open(fileName)
+    
+    # get headers, skipping file name
+    headers=handle.readline().strip().split(',')[1:]
+    ret=dict()
+    for line in handle:
+        csi=line.strip().split(',')
+        ret[csi[0]] = csi[1:]
+    return headers,ret
+
+def main(args):
+    parser = OptionParser(usage="Usage: %prog [options] SCENARIOS.CSV RESULTS_DIR OUTPUT_FILE",
+            description="""Given a .csv file associating scenario file names
+            with DB scenario IDs (sce_id), SCENARIOS.CSV, and a directory of results,
+            RESULTS_DIR, where results have the name of the scenario file but
+            with .xml substituted for .txt, a tab-separated file of combined
+            outputs, OUTPUT_FILE, is written, where each line contains sce_id,
+            survey, group and measure identifiers, and a value.""",version="%prog 0.1")
+    
+    (options, others) = parser.parse_args(args=args[1:])
+    if len(others)!=3 or not os.path.isdir(others[1]):
+        parser.print_usage()
+        return 1
+    
+    sweeps,swArmIds = readSwArmIds(others[0])
+    sce_id_ind = next((i for i in xrange(len(sweeps)) if sweeps[i] == 'sce_id'), None)
+    if sce_id_ind is None:
[…]

### 2018-10-29 code/openmalaria/openmalaria-schema-39.0/util/compareCtsout.py

@@ -1,0 +1,136 @@
+#!/usr/bin/env python2
+# -*- coding: utf-8 -*-
+"""
+This file is part of OpenMalaria.
+
+Copyright (C) 2005-2014 Swiss Tropical Institute
+
+OpenMalaria is free software; you can redistribute it and/or modify
+it under the terms of the GNU General Public License as published by
+the Free Software Foundation; either version 2 of the License, or (at
+your option) any later version.
+
+This program is distributed in the hope that it will be useful, but
+WITHOUT ANY WARRANTY; without even the implied warranty of
+MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the GNU
+General Public License for more details.
+
+You should have received a copy of the GNU General Public License
+along with this program; if not, write to the Free Software
+Foundation, Inc., 51 Franklin Street, Fifth Floor, Boston, MA 02110-1301, USA.
+"""
+
+import sys
+import string
+from optparse import OptionParser
+from approxEqual import ApproxSame
+
+REL_PRECISION=1e-6
+ABS_PRECISION=1e-6
+
+
+def charEqual (fn1,fn2):
+    MAX=10*1024
+    f1 = open(fn1,'r')
+    f2 = open(fn2,'r')
+    while True:
+        s1 = f1.read(MAX)
+        s2 = f2.read(MAX)
+        if (len(s1)==0) or (len(s2)==0):
+            # end of one or both files; equal if it's the end of both
+            return len(s1) == len(s2)
+        if s1 != s2:
+            return False
+
+def readSums(fn):
+    """return list of titles and list of column sums"""
+    f=open(fn,'r')
+    titles=[s.strip() for s in f.readline().split('\t')]
+    if titles==['##','##']:
+        # first line is auxilliary header...
+        titles=[s.strip() for s in f.readline().split('\t')]
+    cols=[0.0 for x in range(len(titles))]
+    for line in f:
+        i=0
+        for item in line.split('\t'):
+            cols[i] += float(item.strip())
+            i+=1
+    return titles,cols
+
+def sumsApproxEq (fn1,fn2):
+    """ return true if approx equal, false otherwise
+    also prints about unequal stuff"""
+    t1,vec1=readSums(fn1)
+    t2,vec2=readSums(fn2)
+    approxSame = ApproxSame(REL_PRECISION,ABS_PRECISION)
+    if t1 != t2:
+        print "\033[1;31mColumns not equal:"
+        print t1
+        print t2
[…]

### 2018-10-29 code/openmalaria/openmalaria-schema-39.0/util/update-header.py

@@ -1,0 +1,119 @@
+#!/usr/bin/env python2
+# -*- coding: utf-8 -*-
+"""
+This file is part of OpenMalaria.
+
+Copyright (C) 2005-2014 Swiss Tropical Institute
+
+OpenMalaria is free software; you can redistribute it and/or modify
+it under the terms of the GNU General Public License as published by
+the Free Software Foundation; either version 2 of the License, or (at
+your option) any later version.
+
+This program is distributed in the hope that it will be useful, but
+WITHOUT ANY WARRANTY; without even the implied warranty of
+MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the GNU
+General Public License for more details.
+
+You should have received a copy of the GNU General Public License
+along with this program; if not, write to the Free Software
+Foundation, Inc., 51 Franklin Street, Fifth Floor, Boston, MA 02110-1301, USA.
+"""
+"""
+Find and replace likely copyright headers in .cpp/.h files.
+Customise the directories to traverse below.
+"""
+
+import os
+
+#excludedir = ["..\\Lib"]
+
+oldheaders=set()
+nonheaders=set()
+
+def update_source(filename, copyright):
+    global oldheaders
+    global nonheaders
+    
+    utfstr = chr(0xef)+chr(0xbb)+chr(0xbf)
+    fdata = file(filename,"r+").read()
+    isUTF = False
+    if (fdata.startswith(utfstr)):
+        isUTF = True
+        fdata = fdata[3:]
+    
+    i=0
+    line_comment=False
+    block_comment=False
+    prev_slash=False # or prev_star
+    want_block=True # false once we have a block comment or set of line comments
+    for c in fdata:
+        i+=1
+        if block_comment:
+            # use prev_slash as prev_star instead
+            if prev_slash and c=='/':
+                block_comment=False
+                prev_slash=False
+                want_block=False
+                # or: break to stop now (don't eat any new-lines)
+            elif c=='*':
+                prev_slash=True
+            else:
+                prev_slash=False
+        elif c in [' ','\t','\f','\v']:
+            pass
+        elif c in ['\r','\n']:
+            line_comment=False # end of line
+        elif want_block and c=='/':
+            if prev_slash:
+                line_comment=True
+                prev_slash=False
+            else:
+                prev_slash=True
+        elif want_block and c=='*' and prev_slash:
+            prev_slash=False
+            block_comment=True
[…]

### 2018-11-05 documents/main.tex
added more notes

\documentclass[]{scrartcl}

%opening
\usepackage[round]{natbib}
\usepackage{listings}
% \usepackage{minted}
\usepackage{color}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{dsfont}
\usepackage{pgf}
\usepackage{subcaption}
\usepackage{tikz}
\usepackage{hyperref}
\usepackage{dsfont}
\usetikzlibrary{bayesnet}
\newcommand{\iden}{\ensuremath{\mathbb{I}}}  %identity func
\newcommand{\real}{\ensuremath{\mathbb{R}}} %real numbers
\newcommand{\nat}{\ensuremath{\mathbb{N}}}  %natural numbers
\newcommand{\ints}{\ensuremath{\mathbb{Z}}}
\newcommand{\expt}{\ensuremath{\mathbb{E}}}
\newcommand{\X}{\ensuremath{\mathbb{X}}}
\newcommand{\Z}{\ensuremath{\mathfrak{Z}}}
\newcommand{\y}{\textbf{y}} 
\newcommand{\x}{\textbf{x}} 
\newcommand{\asto}{\overset{a.s.}{\to}}
\renewcommand{\labelitemii}{$\star$}
\newcommand{\om}{\textit{OpenMalaria}}
\DeclareMathOperator*{\argmax}{arg\,\!max}

\title{Notes on Open Malaria}
\author{Bradley Gram-Hansen}

\begin{document}

\maketitle

\begin{abstract}

\end{abstract}

% \input{mognotes/mixofgauss}

\section{Simulation Options}

The \om simulator is designed for simulating individual scenarios. 
In order to simulate a study covering variations in several factors, one 
would have to  design a fully \textit{factorial} experiment.

\begin{quote}
	\begin{itemize}
		\item \textbf{\textit{Sweep}} : A sweep is one of more covarying factors, and set of all \textit{arms} of this sweep.
		\item \textbf{\textit{Arm}}: Each sweep must have one or more values for each of its factors. An \textit{arm} 
		is one of these combinations of values - a configuration assigning a value to each factor on the sweep. 
		\item \textbf{\textit{Experiment}}: Largely synonymous with a study, an \textit{experiment} is the complete
		description of all sweeps and arms used, together with the scenarios generated and results produced. Each scenario is generated by choosing
		one arm for each sweep in the experiment. 
		\item \textbf{\textit{Full factorial design}} : A \textit{full factorial experiment} is one where all possible scenarios are generated: all
		combinations of one arm per sweep are used.
		\item \textbf{\textit{Seed}}: Each time a scenario is simulated, the random number generated must be started from some \textit{seed}.
	\end{itemize}
\end{quote}

Since OpenMalaria simulations are stochastic, they use a number of random seeds (say 2-50) 
to provide estimates of the contributions of random variation to the results.

\section{XML files}

Within the open malaria simulator \textit{XML} files should be seen as the program, specifying a model over all parameters. Well kind of, it should actually be treated as a prior to the simulator, providing knowledge regarding how a model is specified in terms of the initial parameters. This then interacts directly with the simulator.

\section{Key features of the simulator}

\begin{itemize}
\item Seasonality of transmission
\item Access to treatment an inherent part of model
\item Addressing uncertainty: ensemble of 14 model variants

\end{itemize}
\section{How to Use Open Malaria}

\paragraph*{Objectives}
\begin{itemize}
	\item Understand OpenMalaria workflow to design an experiment
	\item Understand the important features of OpenMalaria
	\item Understand, the topic of choice, the model assumptions and how interventions interact 
	\item Understand what information is needed to parameterize OpenMalaria and where to find it
\end{itemize}

\subsection{Seasonality of transmissions}
The Plasmodium falciparum entomological inoculation rate (Pf EIR) is a measure of exposure to infectious mosquitoes. 
When describing transmissions, the simulator is driven by the input value of annual average EIR, the average number of infectious bites a person receives each year.
The seasonality is specifically specified by 12 monthly values for each vector species. Abs values or proportion for each month.

In epidemiology, a \textbf{disease vector} is any agent that carries and transmits an infectious pathogen into another living organism; most agents regarded as vectors are organisms, such as intermediate parasites or microbes, but it could be an inanimate medium of infection such as dust particles.

\section{How does Malaria spread?}
\begin{itemize}
\item Deterministic model of malaria transmission based on the mosquito feeding cycle
\item Linked to models of interventions
\item Driven by EIR
\item For input EIR values, you may need to work backwards to determine pre-experiment level
\end{itemize}

\begin{figure}
\includegraphics[width=\textwidth]{images/malariacycle.png}
\caption{The mosquito feeding cycle. Malaria is spread through this cycle.}
\end{figure}

\section{Modelling different treatments}

Stochastic simulation models of malaria based on the simulation of infections in humans. Tracks health status of individuals at discrete time points

\begin{figure}
\includegraphics[width=\textwidth]{images/malariatreatment.png}
\end{figure}

\begin{figure}
\includegraphics[width=\textwidth]{images/accesstotreatment.png}
\end{figure}
\section{Scenario transmission}

OpenMalaria simulations of malaria transmissions require a specification of:
\begin{itemize}
	\item The level and seasonality of exposure (measured by the Entomological Inoculation Rate, EIR) to malaria at the start of the simulation
	\item The model for transmission from mosquitoes to humans and the dynamics of malaria parasite cycle within humans.
	\item The model for malaria transmission from humans to 
	\item The entomological model. There are two different variants of the entomological model:
	\begin{itemize}
    \item The "Non-vector" variant does not consider mosquito dynamics and hence uses fixed or periodically (seasonally) varying vectorial capacity. It is appropriate for modeling situations with only interventions such as chemotherapy or vaccines that only act on humans.
	\item The "Vector" variant comprises discrete-time population models that simulate how many mosquitoes belong in each of several categories at each time. The models assume that the infectious (sporozoite positive) mosquitoes act to distribute infections at random to the human population (with human exposure proportionate to availability). Entomological interventions modify the vectorial capacity and require the "vector" transmission model variant. The simulations that include non-periodic changes in the vectorial capacity use a seasonally forced version of the difference equation model for vector dynamics.
\end{itemize}
\end{itemize}
\end{document}

Tom notes

Always write down an abstract and introduction when you are starting a new project.

Write down ideas as formally as possible.

Things that happen first - the less applied part:

As you mathematically formulate in code, you write down the mathematics that describes that and the problem formulation. 
Try and write down a joint of the model. That way you can determine if the problem is doable. Must deal with these things first.

Things to look into:

Are there Multi-modes?
Is the problem high-dimensional?
What is the dimensionality of problems?

### 2018-11-06 documents/main.tex
adding more notes

@@
 \end{itemize}
-\end{document}
 
-Tom notes
+\section{Tom stuff}
 
-Always write down an abstract and introduction when you are starting a new project.
+\paragraph*{Tom notes}
 
-Write down ideas as formally as possible.
-
-Things that happen first - the less applied part:
-
-As you mathematically formulate in code, you write down the mathematics that describes that and the problem formulation. 
-Try and write down a joint of the model. That way you can determine if the problem is doable. Must deal with these things first.
+\begin{itemize}
+	\item Always write down an abstract and introduction when you are starting a new project. 
+	\item Write down ideas as formally as possible. 
+	\item Things that happen first - the less applied part: 
+	\begin{itemize}
+		\item As you mathematically formulate in code, you write down the mathematics that describes that and the problem formulation. 
+         \item Try and write down a joint of the model. That way you can determine if the problem is doable. Must deal with these things first. 
+\end{itemize}
+\end{itemize}
 
@@
 
-Are there Multi-modes?
-Is the problem high-dimensional?
-What is the dimensionality of problems?
+\begin{itemize}
+\item Are there Multi-modes?
+\item Is the problem high-dimensional?
+\item What is the dimensionality of problems?
+\end{itemize}
+
+\section{Gunes Notes}
+\begin{itemize}
+	\item ABC - talk about this in the context to MCMMC
+	\item The forward simulator represents some likelihood function that can be tractable or intractable. 
+	\item Within forward simulators the randomness comes from rejection sampling statements , but conditioning within the simulators is hard
+this is why we need to use the prob prog system . 
+\end{itemize}
+
+\end{document}

### 2018-11-07 documents/main.tex
updated

@@
 
+\subsection{Approximate Bayesian Computation}
+
+When dealing with a constrained set of distributions we can perform exact Bayesian inference, that is we can use statistically correct inference algorithms to find the true posterior $p(x | D) = \frac{p(D|x)p(x)}{p(D)}$, not just heuristic results.
+If we choose the a conjugate-pair of prior $p(x)$ and likelihood function $p(x|D)$ then we can also analytically exactly calculate the true posterior.
+However, in general most real world problems, i.e.
+physically simulations, cannot necessarily be constrained to this family of problems, as the likelihood is typically intractable, or computationally expensive to compute.
+This occurs because the \textit{forward simulator} emulates the likelihood function
+due to the ingrained stochasticity with the model, which makes it impossible to write down a likelihood function.
+Nonetheless, we can still apply a series of approximation techniques, such as variational inference that enable one to approximate the true posterior.
+Although such algorithms typically have no guarantees, they are typically lower bounded.
+
 \section{Gunes Notes}

### 2018-11-22 documents/main.tex
updated
udpated

@@
 % \input{mognotes/mixofgauss}
+
+\section{What is Open Malria?}
+Open malaria is for simulating malaria epidemology and the impacts on that epidemiology of interventions against malaria. 
+ It is based on microsimulations of Plasmodium falciparum malaria in humans, originally developed for simulating malaria vaccines.
+ These models simulate the dynamics of malaria parasitaemia in the course of an infection, of transmission, of immunity, and of the processes leading to illness and death.
+The system is set up to simulate malaria in village, or district size human populations
+
+The original models were designed for simulating the impact of interventions acting on humans(for example, chemotherapy, or vaccines).
+OpenMalaria has subsequently been extended to include simulation of the dynamics of malaria in the mosquito vector, and of interventions that act on mosquitoes.
+It also offers alternative model structures for many of the components of the human models.
+The models allow for variations among humans in their exposure to mosquitoes, and in their responses to the parasite.
+
+\section{Facts about Malaria}
+
+\begin{itemize}
+\item Endemic in 99 countries, 219 million cases, 660,000 deaths 
+\item 90\% in Africa
+\item Mortality rates fell by 26\% around the world in the past decade
 
@@
 
+\subsection{Applied Interpretable Inference}
+
 \section{Gunes Notes}

### 2018-11-27 documents/main.tex
updated

@@
 \item Mortality rates fell by 26\% around the world in the past decade
-
+\end{itemize}
+\section{How can Open malaria help in the real world?}
+\begin{figure}[h!]
+\includegraphics[width=\textwidth]{images/research_process.png}
+\caption{Where OM fits into the real world}
+\end{figure}
 \section{Simulation Options}
@@
 \end{figure}
+
+\section{Running a simulation}
+
+\text{Run ./openMalaria --scenario MyScenario.xml}
 \section{Scenario transmission}
@@
 
+\section{The phases of the simulator}
+\begin{itemize}
+\item the XML document is read, model constants are set and some variables initialised
+\item an initial human population is created
+\item a warm-up period of one human life span is run
+\item there may be some extra simulations to adjust the mosquito population to the pre-specified Transmission situation.
+\item the intervention phase starts.
+\end{itemize}
+During these different phases the feedback from infections in humans to those in mosquitoes is handled in different ways
+
+\section{Gathering Data}
+
+The next step is to gather the necessary data to include in the scenario file to describe the situation in your study area as closely as possible to reality.
+There are more or less 90 separate pieces of data that need to go into the model.
+As many inputs as possible should come from studies in your particular study area, but realistically the overall percentage of inputs that come from site-specific data will be rather modest (about 30\%).
+A sizeable number of inputs will come from previously published literature at the district, national or regional level through a literature review.
+There will be a number of inputs that will remain the same as the initial model parameterization, depending on which model(s) you choose, particularly data on the vector life cycle and drug action.
+It is helpful to document the source of the data you choose for each parameter and if there are multiple sources why you chose the value that will go into the model.
+
+\section{Outputs}
+The simulation results of the outputs you have chosen from Step #2 are produced in the form of two separate .txt files (ctsout.txt and output.txt) that can be imported into your program of choice for analysis (i.e.
+Excel, R, STATA, etc)
 \section{Tom stuff}

### 2018-12-14 documents/main.tex
updated

@@
 \section{What is Open Malria?}
-Open malaria is for simulating malaria epidemology and the impacts on that epidemiology of interventions against malaria. 
+Open malaria is for simulating malaria epidemiology and the impacts on that epidemiology of interventions against malaria. 
  It is based on microsimulations of Plasmodium falciparum malaria in humans, originally developed for simulating malaria vaccines.
@@
 \end{itemize}
+
+\subsection{How does Malaria spread?}
+\begin{itemize}
+\item Deterministic model of malaria transmission based on the mosquito feeding cycle
+\item Linked to models of interventions
+\item Driven by EIR
+\item For input EIR values, you may need to work backwards to determine pre-experiment level
+\end{itemize}
+
+\begin{figure}[ht!]
+\includegraphics[width=\textwidth]{images/malariacycle.png}
+\caption{The mosquito feeding cycle. Malaria is spread through this cycle.}
+\end{figure}
+
 \section{How can Open malaria help in the real world?}
+
+It helps by enabling researchers to propose experiments, based on a number of observations that are collected on the field.
+
 \begin{figure}[h!]
@@
 \end{itemize}
+
+\subsection{Modelling different treatments}
+
+Stochastic simulation models of malaria based on the simulation of infections in humans. Tracks health status of individuals at discrete time points
+
+\begin{figure}
+\includegraphics[width=\textwidth]{images/malariatreatment.png}
+\end{figure}
+
+\begin{figure}
+\includegraphics[width=\textwidth]{images/accesstotreatment.png}
+\end{figure}
+
 \section{How to Use Open Malaria}
@@
 
-\section{How does Malaria spread?}
-\begin{itemize}
-\item Deterministic model of malaria transmission based on the mosquito feeding cycle
-\item Linked to models of interventions
-\item Driven by EIR
-\item For input EIR values, you may need to work backwards to determine pre-experiment level
-\end{itemize}
-
-\begin{figure}
-\includegraphics[width=\textwidth]{images/malariacycle.png}
-\caption{The mosquito feeding cycle. Malaria is spread through this cycle.}
-\end{figure}
-
-\section{Modelling different treatments}
-
-Stochastic simulation models of malaria based on the simulation of infections in humans. Tracks health status of individuals at discrete time points
-
-\begin{figure}
-\includegraphics[width=\textwidth]{images/malariatreatment.png}
-\end{figure}
-
-\begin{figure}
-\includegraphics[width=\textwidth]{images/accesstotreatment.png}
-\end{figure}
-
-\section{Running a simulation}
-
-\text{Run ./openMalaria --scenario MyScenario.xml}
-\section{Scenario transmission}
+\subsection{Scenario transmission}
 
@@
 
-\section{The phases of the simulator}
+\subsection{Running a simulation}
+Within the same directory structure we add the following files:
+
 \begin{itemize}
-\item the XML document is read, model constants are set and some variables initialised
+\item densities.csv
+\item scenario\_xx.xsd
+\end{itemize}
+Then there must exist at least on file <MyScenario.xml>
+
+Then to run we use the following command:
+\text{./openMalaria --scenario MyScenario.xml}
+
+To ensure that it is possible to execute the openMalaria object, it must be within 
+the execution path.
+
+\subsection{The phases of the simulator}
+\begin{itemize}
+\item the XML document is read, model constants are set and some variables initialized
 \item an initial human population is created
@@
 \section{Outputs}
-The simulation results of the outputs you have chosen from Step #2 are produced in the form of two separate .txt files (ctsout.txt and output.txt) that can be imported into your program of choice for analysis (i.e.
+The simulation results of the outputs you have chosen from Step 2 are produced in the form of two separate .txt files (ctsout.txt and output.txt) that can be imported into your program of choice for analysis (i.e.
 Excel, R, STATA, etc)
@@
 \begin{itemize}
-	\item ABC - talk about this in the context to MCMMC
+	\item ABC - talk about this in the context to MCMC
 	\item The forward simulator represents some likelihood function that can be tractable or intractable. 

### 2019-03-11 paper/main.tex
added abstract. Intro to do

\documentclass{article}

% if you need to pass options to natbib, use, e.g.:
%     \PassOptionsToPackage{numbers, compress}{natbib}
% before loading neurips_2018

% ready for submission
% \usepackage{neurips_2018}

% to compile a preprint version, e.g., for submission to arXiv, add add the
% [preprint] option:
%     \usepackage[preprint]{neurips_2018}

% to compile a camera-ready version, add the [final] option, e.g.:
     \usepackage[final]{neurips_2019}

% to avoid loading the natbib package, add option nonatbib:
%     \usepackage[nonatbib]{neurips_2018}

\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
\usepackage{hyperref}       % hyperlinks
\usepackage{url}            % simple URL typesetting
\usepackage{booktabs}       % professional-quality tables
\usepackage{amsfonts}       % blackboard math symbols
\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{microtype}      % microtypography

\title{Applied Interpretable Inference: Connecting Probabilistic Programming with Epidemology Simulators}

% The \author macro works with any number of authors. There are two commands
% used to separate the names and addresses of multiple authors: \And and \AND.
%
% Using \And between authors leaves it to LaTeX to determine where to break the
% lines. Using \AND forces a line break at that point. So, if LaTeX puts 3 of 4
% authors names on the first line, and the last on the second line, try using
% \AND instead of \And before the third author name.

\author{%
  David S.~Hippocampus\thanks{Use footnote for providing further information
    about author (webpage, alternative address)---\emph{not} for acknowledging
    funding agencies.} \\
  Department of Computer Science\\
  Cranberry-Lemon University\\
  Pittsburgh, PA 15213 \\
  \texttt{hippo@cs.cranberry-lemon.edu} \\
  % examples of more authors
  % \And
  % Coauthor \\
  % Affiliation \\
  % Address \\
  % \texttt{email} \\
  % \AND
  % Coauthor \\
  % Affiliation \\
  % Address \\
  % \texttt{email} \\
  % \And
  % Coauthor \\
  % Affiliation \\
  % Address \\
  % \texttt{email} \\
  % \And
  % Coauthor \\
  % Affiliation \\
  % Address \\
  % \texttt{email} \\
}

\begin{document}
% \nipsfinalcopy is no longer used

\maketitle

\begin{abstract}

A goal of probabilistic programming is to couple simulators, with inference. This is 
because stochastic simulators are used prominently in many industrial settings,
do not require one to construct hand-crafted joint distributions as they implicitly 
define a joint distribution of the program and encode learnt structures 
directly. This makes simulators powerful tools and much of machine learning (ML) and 
Artificial Intelligence (AI)
can be seen as trying to emulate such simulators from a purely data-driven approach.
However, in the 
ML/AI setting, although we can often infer outcomes, we have little understanding about what 
in the data led to the outputted inferences. 
This makes it challenging to deploy ML/AI systems into the wild, especially in health-related and safety-critical domains, such
as epidemiology, as we lose \emph{interpretability}. 
In this work, we explain how to design ML/AI systems that combine
probabilistic programming systems (PPSs) and epidemiology simulators, to extract
fully interpretable posterior structures, enabling policy makers 
and practitioners to make interpretable inferences. 
In particular, we demonstrate this for the Malaria disease and show how we can 
perform interpretable inference in such settings.
\end{abstract}

\section{Introduction}

To do.

\begin{itemize}

\item Outline the objectives
\item State the importance of interpretability in the inference of the simulator
\item Our system enables one to understand and interpret the most common paths in a program (simulation).
\item In highlighting this a practitioner can then go back to the field and explore that parameter/s in more depth.
\item This facilitates understanding and the construction of more detailed models and data collection strategies

\end{itemize}

\section{Related Work}
\label{sec:related}

\section{Epidemiology Simulators and Probabilistic Programming}
\label{sec:background}
This will kind of be like a background section.

\section{Methodology}
 TODO

\section{Results}

TO ADD

% \subsubsection*{Acknowledgments}

% Use unnumbered third level headings for the acknowledgments. All acknowledgments
% go at the end of the paper. Do not include acknowledgments in the anonymized
% submission, only in the final paper.

\section*{References}

References follow the acknowledgments. Use unnumbered first-level heading for
the references. Any choice of citation style is acceptable as long as you are
consistent. It is permissible to reduce the font size to \verb+small+ (9 point)
when listing the references. {\bf Remember that you can use more than eight
  pages as long as the additional pages contain \emph{only} cited references.}
\medskip

\small

[1] Alexander, J.A.\ \& Mozer, M.C.\ (1995) Template-based algorithms for
connectionist rule extraction. In G.\ Tesauro, D.S.\ Touretzky and T.K.\ Leen
(eds.), {\it Advances in Neural Information Processing Systems 7},
pp.\ 609--616. Cambridge, MA: MIT Press.

[2] Bower, J.M.\ \& Beeman, D.\ (1995) {\it The Book of GENESIS: Exploring
  Realistic Neural Models with the GEneral NEural SImulation System.}  New York:
TELOS/Springer--Verlag.

[3] Hasselmo, M.E., Schnell, E.\ \& Barkai, E.\ (1995) Dynamics of learning and
recall at excitatory recurrent synapses and cholinergic modulation in rat
hippocampal region CA3. {\it Journal of Neuroscience} {\bf 15}(7):5249-5262.

\end{document}

### 2019-03-13 paper/main.tex
updated;

@@
 
-\section{Related Work}
+% taken from Toms thesis.
+
+Stochastic simulators are common across a number of domains, statistical physics~\cite{landau_binder_2014}, financial modeling~\cite{jackel2002monte},
+weather prediction~\cite{evensen1994sequential}, epidemiology~\cite{smith2008towards} and many others. 
+An advantage of using stochastic simulators is that they provide a level of interpretability not found in modern deep learning settings, as they directly incorporate model structure from carefully reasoned observations and experiments.
+Probabilistic programming systems~(PPSs) are purpose built systems for making inferences and probabilistic 
+modeling accessible. 
+In particular, PPSs allow probabilistic models to be represented in the form of a generative model through the
+probabilistic programming language~(PPL), which enables one to write statements that enable conditioning on data~\cite{gordon2014probabilistic,Goodman08church:a}. 
+Thus, it seems only natural to connect simulators with PPSs, as simulators explicitly define a generative model, in the language in which they are written, and the nature of the PPSs enables us to perform inference in these simulators, by conditioning on observations, which are fed in as input to the simulators.
+In doing this we can infer things about stochastic input parameters and other variables sampled during the program's forward execution, whilst also providing full interpretability in the inference results, which is absolutely necessary in safety-critical domains.
+By exploiting the tools that we develop in this paper, one is able to connect epidemiology simulators with \texttt{Pyprob}~\cite{le-2016-inference} to extract fully interpretable posterior structures allowing practitioners to interpret which input parameters have the largest affect on the epidemic, in particular, malaria.
+Over five-hundred-thousand people die from malaria each year, mostly children under five years of age, with 90 per cent of malaria cases occurring in Sub-Saharan Africa.
+An estimated 100-300 million people suffer from malaria each year~\cite{world2016world}.
+Thus, by understanding the prominence of input parameters to epidemiology simulators, guides decision making processes relating to effective treatment and prevention strategies. 
 \label{sec:related}
@@
 
-References follow the acknowledgments. Use unnumbered first-level heading for
-the references. Any choice of citation style is acceptable as long as you are
-consistent. It is permissible to reduce the font size to \verb+small+ (9 point)
-when listing the references. {\bf Remember that you can use more than eight
-  pages as long as the additional pages contain \emph{only} cited references.}
-\medskip
-
-\small
-
-[1] Alexander, J.A.\ \& Mozer, M.C.\ (1995) Template-based algorithms for
-connectionist rule extraction. In G.\ Tesauro, D.S.\ Touretzky and T.K.\ Leen
-(eds.), {\it Advances in Neural Information Processing Systems 7},
-pp.\ 609--616. Cambridge, MA: MIT Press.
-
-[2] Bower, J.M.\ \& Beeman, D.\ (1995) {\it The Book of GENESIS: Exploring
-  Realistic Neural Models with the GEneral NEural SImulation System.}  New York:
-TELOS/Springer--Verlag.
-
-[3] Hasselmo, M.E., Schnell, E.\ \& Barkai, E.\ (1995) Dynamics of learning and
-recall at excitatory recurrent synapses and cholinergic modulation in rat
-hippocampal region CA3. {\it Journal of Neuroscience} {\bf 15}(7):5249-5262.
+\bibliographystyle{plain}
+\bibliography{refs}
 

### 2019-03-13 paper/main.tex
updated

@@
 In doing this we can infer things about stochastic input parameters and other variables sampled during the program's forward execution, whilst also providing full interpretability in the inference results, which is absolutely necessary in safety-critical domains.
-By exploiting the tools that we develop in this paper, one is able to connect epidemiology simulators with \texttt{Pyprob}~\cite{le-2016-inference} to extract fully interpretable posterior structures allowing practitioners to interpret which input parameters have the largest affect on the epidemic, in particular, malaria.
+By exploiting the tools that we develop in this paper, one is able to connect epidemiology simulators with \texttt{Pyprob}~\cite{le-2016-inference} to extract fully interpretable posterior structures allowing practitioners to interpret which input parameters have the largest affect on the epidemic, in our case malaria.
 Over five-hundred-thousand people die from malaria each year, mostly children under five years of age, with 90 per cent of malaria cases occurring in Sub-Saharan Africa.

### 2019-04-21 paper/main.tex
updated paper

@@
 \usepackage{url}            % simple URL typesetting
+\usepackage{graphicx}
 \usepackage{booktabs}       % professional-quality tables
@@
 \usepackage{microtype}      % microtypography
-
-\title{Applied Interpretable Inference: Connecting Probabilistic Programming with Epidemology Simulators}
+\usepackage{listings}
+\title{Applied Interpretable Inference: Connecting Probabilistic Programming with Epidemiology Simulators}
 
@@
 \author{%
-  David S.~Hippocampus\thanks{Use footnote for providing further information
-    about author (webpage, alternative address)---\emph{not} for acknowledging
-    funding agencies.} \\
-  Department of Computer Science\\
-  Cranberry-Lemon University\\
-  Pittsburgh, PA 15213 \\
-  \texttt{hippo@cs.cranberry-lemon.edu} \\
+Paper ID: }
   % examples of more authors
@@
   % \texttt{email} \\
-}
-
+%}
 \begin{document}
@@
 
+Simulators encode several years, if not decades, of research and development and so by their very nature are structurally complicated.
+Since the development happens over several decades, usually without documentation, understanding a code base written with legacy libraries and software is incredibly difficult and an almost impossible task for a non-expert, or expert who has not had access to the previous years of development.
+Our method of hijacking simulators builds on the work of~\cite{baydin2018efficient} and extends the framework to encapsulate a more diverse range of simulators, in particular epidemiology simulators.
+Our framework provides a simple solution and makes it easy to transform arbitrary stochastic simulators into probabilistic programs, regardless of the complexity of the simulator and the language that the simulator is written in\footnote{The framework currently supports 9 popular languages, but there is nothing to stop this being extended to any language of interest.}.
+In addition to this, we can understand the structure of programs in simulators for both decoding \emph{black-box structures} and for posterior inference.
+This enables two things.
+1) It enables software developers and researchers to understand complicated code bases.
+2) It provides interpretable inference results that provide policy makers with predictions that are truly interpretable, in the sense that the end-user of the inference results understands what physical events led to inference outcome.
+This is critical in several domains, and is particularly important in the medical domain.
+
+\subsection{Connecting a Stochastic Simulator to a Probabilistic Programming System}
+
+In order to hijack the stochastic simulator we are only required to override the location of the stochastic primitives, i.e.
+the operations that generate the stochasticity and randomness within the simulator.
+We will walk through how this is done for EMOD and OpenMalaria,  you will see that it is procedurally identical, however, given the sheer number of simulators it maybe slightly different procedurally.
+To our knowledge it shouldn't be, but we have not been able view of all simulators.
+
+\subsection{Extending Probabilistic Programming Protocols}
+
+\begin{figure}[h!]
+	\centering
+	\includegraphics[width=\textwidth]{plots/ewan_25_pop_10.pdf}
+	\caption{Add something. Population 10}
+\end{figure}
+\begin{figure}[h!]
+	\centering
+	\includegraphics[width=\textwidth]{plots/ewan_25_pop_100.pdf}
+	\caption{Add something. Population 100}
+\end{figure}
+TODO
+
 \section{Results}

### 2019-04-23 paper/main.tex
updated

@@
 This is critical in several domains, and is particularly important in the medical domain.
+We provide examples of both in Section~\textbf{TO ADD REF}.
 
@@
 We will walk through how this is done for EMOD and OpenMalaria,  you will see that it is procedurally identical, however, given the sheer number of simulators it maybe slightly different procedurally.
-To our knowledge it shouldn't be, but we have not been able view of all simulators.
+To our knowledge it shouldn't be, but we have not been able view of all stochastic simulators.
+The first step is to build a containerized environment, such as Docker\textbf{ADD citation} and Singularity, which enables the simulators to be
+run on any device, independent of different hardware architectures removing the requirement for hardware specific machines.
+We provide examples of building a containerized environment in the Supplement \textbf{ADD ref}.
+This environment only has to be built once, which then enables the simulator to be run simultaneously on multiple machines.
+The second step
 

### 2019-04-23 paper/icml_ai4good/main.tex
add icml workshop template and update with text

--- 2019-04-23 paper/main.tex
@@
+%%%%%%%% ICML 2019 EXAMPLE LATEX SUBMISSION FILE %%%%%%%%%%%%%%%%%
+
 \documentclass{article}
 
-% if you need to pass options to natbib, use, e.g.:
-%     \PassOptionsToPackage{numbers, compress}{natbib}
-% before loading neurips_2018
+% Recommended, but optional, packages for figures and better typesetting:
+\usepackage{microtype}
+\usepackage{graphicx}
+\usepackage{subfigure}
+\usepackage{booktabs} % for professional tables
 
-% ready for submission
-% \usepackage{neurips_2018}
+% hyperref makes hyperlinks in the resulting PDF.
+% If your build breaks (sometimes temporarily if a hyperlink spans a page)
+% please comment out the following usepackage line and replace
+% \usepackage{icml2019} with \usepackage[nohyperref]{icml2019} above.
+\usepackage{hyperref}
 
-% to compile a preprint version, e.g., for submission to arXiv, add add the
-% [preprint] option:
-%     \usepackage[preprint]{neurips_2018}
+% Attempt to make hyperref and algorithmic work together better:
+\newcommand{\theHalgorithm}{\arabic{algorithm}}
 
-% to compile a camera-ready version, add the [final] option, e.g.:
-     \usepackage[final]{neurips_2019}
+% Use the following line for the initial blind version submitted for review:
+\usepackage{icml2019}
 
-% to avoid loading the natbib package, add option nonatbib:
-%     \usepackage[nonatbib]{neurips_2018}
+% If accepted, instead use the following line for the camera-ready submission:
+%\usepackage[accepted]{icml2019}
 
-\usepackage[utf8]{inputenc} % allow utf-8 input
-\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
-\usepackage{hyperref}       % hyperlinks
-\usepackage{url}            % simple URL typesetting
-\usepackage{graphicx}
-\usepackage{booktabs}       % professional-quality tables
-\usepackage{amsfonts}       % blackboard math symbols
-\usepackage{nicefrac}       % compact symbols for 1/2, etc.
-\usepackage{microtype}      % microtypography
-\usepackage{listings}
-\title{Applied Interpretable Inference: Connecting Probabilistic Programming with Epidemiology Simulators}
+% The \icmltitle you define below is probably too long as a header.
+% Therefore, a short form for the running title is supplied here:
+\icmltitlerunning{Submission and Formatting Instructions for ICML 2019}
 
-% The \author macro works with any number of authors. There are two commands
-% used to separate the names and addresses of multiple authors: \And and \AND.
-%
-% Using \And between authors leaves it to LaTeX to determine where to break the
-% lines. Using \AND forces a line break at that point. So, if LaTeX puts 3 of 4
-% authors names on the first line, and the last on the second line, try using
-% \AND instead of \And before the third author name.
+\begin{document}
 
-\author{%
-Paper ID: }
-  % examples of more authors
-  % \And
-  % Coauthor \\
-  % Affiliation \\
-  % Address \\
-  % \texttt{email} \\
-  % \AND
-  % Coauthor \\
-  % Affiliation \\
-  % Address \\
-  % \texttt{email} \\
-  % \And
-  % Coauthor \\
-  % Affiliation \\
-  % Address \\
-  % \texttt{email} \\
-  % \And
-  % Coauthor \\
-  % Affiliation \\
-  % Address \\
-  % \texttt{email} \\
-%}
-\begin{document}
-% \nipsfinalcopy is no longer used
+\twocolumn[
+\icmltitle{Submission and Formatting Instructions for \\
+           International Conference on Machine Learning (ICML 2019)}
 
-\maketitle
+% It is OKAY to include author information, even for blind
+% submissions: the style file will automatically remove it for you
+% unless you've provided the [accepted] option to the icml2019
+% package.
+
+% List of affiliations: The first argument should be a (short)
+% identifier you will use later to specify author affiliations
+% Academic affiliations should list Department, University, City, Region, Country
+% Industry affiliations should list Company, City, Region, Country
+
+% You can specify symbols, otherwise they are numbered in order.
+% Ideally, you should not use this facility. Affiliations will be numbered
+% in order of appearance and this is the preferred way.
+\icmlsetsymbol{equal}{*}
+
+\begin{icmlauthorlist}
+\icmlauthor{Aeiau Zzzz}{equal,to}
+\icmlauthor{Bauiu C.~Yyyy}{equal,to,goo}
+\icmlauthor{Cieua Vvvvv}{goo}
+\icmlauthor{Iaesut Saoeu}{ed}
+\icmlauthor{Fiuea Rrrr}{to}
+\icmlauthor{Tateu H.~Yasehe}{ed,to,goo}
+\icmlauthor{Aaoeu Iasoh}{goo}
+\icmlauthor{Buiui Eueu}{ed}
+\icmlauthor{Aeuia Zzzz}{ed}
+\icmlauthor{Bieea C.~Yyyy}{to,goo}
+\icmlauthor{Teoau Xxxx}{ed}
+\icmlauthor{Eee Pppp}{ed}
+\end{icmlauthorlist}
+
+\icmlaffiliation{to}{Department of Computation, University of Torontoland, Torontoland, Canada}
+\icmlaffiliation{goo}{Googol ShallowMind, New London, Michigan, USA}
+\icmlaffiliation{ed}{School of Computation, University of Edenborrow, Edenborrow, United Kingdom}
+
+\icmlcorrespondingauthor{Cieua Vvvvv}{c.vvvvv@googol.com}
+\icmlcorrespondingauthor{Eee Pppp}{ep@eden.co.uk}
+
+% You may provide any keywords that you
+% find helpful for describing your paper; these are used to populate
+% the "keywords" metadata in the PDF but will not be shown in the document
+\icmlkeywords{Machine Learning, ICML}
+
+\vskip 0.3in
+]
+
+% this must go after the closing bracket ] following \twocolumn[ ...
+
+% This command actually creates the footnote in the first column
+% listing the affiliations and the copyright notice.
+% The command takes one argument, which is text to display at the start of the footnote.
+% The \icmlEqualContribution command is standard text for equal contribution.
+% Remove it (just {}) if you do not need this facility.
+
+%\printAffiliationsAndNotice{}  % leave blank if no need to mention equal contribution
+\printAffiliationsAndNotice{\icmlEqualContribution} % otherwise use the standard text.
 
@@
 
-\begin{figure}[h!]
+\begin{figure*}[h!]
 	\centering
-	\includegraphics[width=\textwidth]{plots/ewan_25_pop_10.pdf}
+	\includegraphics[width=\textwidth]{../plots/ewan_25_pop_10.pdf}
 	\caption{Add something. Population 10}
-\end{figure}
-\begin{figure}[h!]
+\end{figure*}
+\begin{figure*}[h!]
 	\centering
-	\includegraphics[width=\textwidth]{plots/ewan_25_pop_100.pdf}
+	\includegraphics[width=\textwidth]{../plots/ewan_25_pop_100.pdf}
 	\caption{Add something. Population 100}
-\end{figure}
+\end{figure*}
 TODO
@@
 
-\bibliographystyle{plain}
+\bibliographystyle{icml2019}
 \bibliography{refs}
@@
 \end{document}
+
+% This document was modified from the file originally made available by
+% Pat Langley and Andrea Danyluk for ICML-2K. This version was created
+% by Iain Murray in 2018, and modified by Alexandre Bouchard in
+% 2019. Previous contributors include Dan Roy, Lise Getoor and Tobias
+% Scheffer, which was slightly modified from the 2010 version by
+% Thorsten Joachims & Johannes Fuernkranz, slightly modified from the
+% 2009 version by Kiri Wagstaff and Sam Roweis's 2008 version, which is
+% slightly modified from Prasad Tadepalli's 2007 version which is a
+% lightly changed version of the previous year's version by Andrew
+% Moore, which was in turn edited from those of Kristian Kersting and
+% Codrina Lauth. Alex Smola contributed to the algorithmic style files.

### 2019-04-23 paper/icml_ai4good/main.tex
update

@@
 \usepackage{icml2019}
-
+\newcommand{\bg}[1]{~{{[{\it \textcolor{red}{{\bf BG:} #1}}]}}}
+\newcommand{\cs}[1]{~{{[{\it \textcolor{red}{{\bf CS:} #1}}]}}}
+\newcommand{\ag}[1]{~{{[{\it \textcolor{red}{{\bf AG:} #1}}]}}}
 % If accepted, instead use the following line for the camera-ready submission:
@@
 \section{Methodology}
- TODO
+
+\subsection{Probabilistic Programming}
+Probabilistic programming~\cite{gordon2014probabilistic,staton2016semantics,kozen1979semantics} is the coming together of multiple disciplines; computer science, statistics and machine learning. 
+It attempts to unify the three fields by creating automated frameworks in which one can easily extrapolate their observed data and perform sophisticated statistical analysis with advance inference procedures such as
+Markov Chain Monte Carlo (MCMC), Variational Inference~(VI) and Deep Neural Networks~(DNN) in an accessible way, i.e. without the need of expert knowledge in implementing and using such procedures.
+Users can then exploit these algorithms on an array of models
+within the probabilistic programming language~(PPL), using the language itself to create models that represent complicated distribution objects. 
+PPLs typically extend existing programming languages such as python, c++, go and so forth, which means that users do not have to learn
+a completely new set of operating semantics.  
+Our PPL PyProb~\cite{le-2016-inference,baydin2018efficient} belongs to the family of universal PPLs 
+that allow the expression of unrestricted probability models in a Turing complete fashion~\cite{wood2014new,goodman2012church,landau_binder_2014,siddharth2017learning,bingham2019pyro}.
+
+\subsection{Stochastic Simulators}
 
 Simulators encode several years, if not decades, of research and development and so by their very nature are structurally complicated.
-Since the development happens over several decades, usually without documentation, understanding a code base written with legacy libraries and software is incredibly difficult and an almost impossible task for a non-expert, or expert who has not had access to the previous years of development.
-Our method of hijacking simulators builds on the work of~\cite{baydin2018efficient} and extends the framework to encapsulate a more diverse range of simulators, in particular epidemiology simulators.
+Within the context of epidemiology simulators we can model treatment, infection and prevention cycle under a variety of different environmental conditions, enabling predictions to be made given a series of field observations \bg{Make sure that I do not state they can already do what PP can do, i.e.
+the ability to update and infer the input parameters continually conditioned on new observations.}.
+However, since the development happens over several decades, usually without comprehensive documentation, understanding a code base written with legacy libraries and software is incredibly difficult and an almost impossible task for a non-expert, or expert who has not had access to the previous years of development.
+Our method of hijacking simulators builds on the work of~\cite{baydin2018efficient} and extends the framework to encapsulate a more diverse range of simulators.
 Our framework provides a simple solution and makes it easy to transform arbitrary stochastic simulators into probabilistic programs, regardless of the complexity of the simulator and the language that the simulator is written in\footnote{The framework currently supports 9 popular languages, but there is nothing to stop this being extended to any language of interest.}.
@@
 This is critical in several domains, and is particularly important in the medical domain.
-We provide examples of both in Section~\textbf{TO ADD REF}.
+We provide examples of former in next Section~\textbf{TO ADD REF} and discuss the latter in ~\bg{Link to conclusion / future work}.
 
+\section{Case Study}
 \subsection{Connecting a Stochastic Simulator to a Probabilistic Programming System}
@@
 the operations that generate the stochasticity and randomness within the simulator.
-We will walk through how this is done for EMOD and OpenMalaria,  you will see that it is procedurally identical, however, given the sheer number of simulators it maybe slightly different procedurally.
-To our knowledge it shouldn't be, but we have not been able view of all stochastic simulators.
-The first step is to build a containerized environment, such as Docker\textbf{ADD citation} and Singularity, which enables the simulators to be
-run on any device, independent of different hardware architectures removing the requirement for hardware specific machines.
-We provide examples of building a containerized environment in the Supplement \textbf{ADD ref}.
-This environment only has to be built once, which then enables the simulator to be run simultaneously on multiple machines.
-The second step
+We will walk through how this is done for EMOD and OpenMalaria, you will see that it is procedurally identical.
+However, given the sheer number of simulators it maybe slightly different procedurally. 
+To our knowledge it shouldn't be, but we have not been able view of all stochastic simulators we cannot make this guarantee with certainty. 
+The first step is to build a containerized environment, such as Docker~\cite{merkel2014docker} and Singularity~\cite{kurtzer2017singularity}, which enables the simulators to be run on any device, independent of different hardware architectures, removing the requirement for hardware specific machines.
+This greatly improves the scope of applicability of our system, as this environment only has to be built once, which then enables the simulator to be run simultaneously on machines of different architectures machine.
+The second step requires one to
 
-\subsection{Extending Probabilistic Programming Protocols}
+% \subsection{Extending Probabilistic Programming Protocols}
 

### 2019-04-23 paper/icml_ai4good/main.tex
updated
will sort out table later

@@
 \usepackage{booktabs} % for professional tables
+\usepackage{tabularx}
 
@@
 \section{Case Study}
-\subsection{Connecting a Stochastic Simulator to a Probabilistic Programming System}
+\subsection{Connecting a Epidemiology Simulators to a Probabilistic Programming System}
 
@@
 
-\section{Results}
-
-TO ADD
-
+% Please add the following required packages to your document preamble:
+% \usepackage{booktabs}
+% Please add the following required packages to your document preamble:
+% \usepackage{booktabs}
+% \usepackage{graphicx}
+% Please add the following required packages to your document preamble:
+% \usepackage{graphicx}
+\begin{table*}[b!]
+\footnotesize
+\setlength{\tabcolsep}{1mm}
+\caption{Addresses}
+\label{table:addresses}
+\def\arraystretch{1.25}
+% \setlength{\tabcolsep}{1mm}
+\begin{tabularx}{\textwidth}{@{}lX@{}} 
+\toprule
+Address & Physical event                                                                                                                            \midrule
+1 & [forward()+0x204; OM::Simulator::start(scnXml::Monitoring const)+0x28a; OM::Population::createInitialHumans()+0x94; OM::Population::newHuman(OM::SimTime)+0x5c; OM::Host::Human::Human(OM::SimTime)+0x12b; OM::WithinHost::WHInterface::createWithinHostModel(double)+0x99; OM::WithinHost::DescriptiveWithinHostModel::DescriptiveWithinHostModel(double)+0x3a; OM::WithinHost::WHFalciparum::WHFalciparum(double)+0xe6; OM::util::random::gauss(double, double)+0xb4]\_\_Normal\\
+2 & [forward()+0x204; OM::Simulator::start(scnXml::Monitoring const\&)+0x468; OM::Population::update1(OM::SimTime)+0x1e9; OM::Population::newHuman(OM::SimTime)+0x5c; OM::Host::Human::Human(OM::SimTime)+0x12b; OM::WithinHost::WHInterface::createWithinHostModel(double)+0x99; OM::WithinHost::DescriptiveWithinHostModel::DescriptiveWithinHostModel(double)+0x3a; OM::WithinHost::WHFalciparum::WHFalciparum(double)+0xe6; OM::util::random::gauss(double, double)+0xb4]\_\_Normal\\
+3 & [forward()+0x204; OM::Simulator::start(scnXml::Monitoring const\&)+0x468; OM::Population::update1(OM::SimTime)+0xff; OM::Host::Human::update(bool)+0x22c; OM::Host::InfectionIncidenceModel::numNewInfections(OM::Host::Human const\&, double)+0x2a3; OM::util::random::poisson(double)+0x13f]\_\_Poisson\\
+4 & [forward()+0x204; OM::Simulator::start(scnXml::Monitoring const\&)+0x468; OM::Population::update1(OM::SimTime)+0xff; OM::Host::Human::update(bool)+0x2bc; OM::Clinical::ClinicalModel::update(OM::Host::Human\&, double, bool)+0x96; OM::Host::NeonatalMortality::eventNeonatalMortality()+0x9; OM::util::random::uniform\_01()+0xc0]\_\_Uniform\\
+5 & [forward()+0x204; OM::Simulator::start(scnXml::Monitoring const\&)+0x468; OM::Population::update1(OM::SimTime)+0xff; OM::Host::Human::update(bool)+0x2bc; OM::Clinical::ClinicalModel::update(OM::Host::Human\&, double, bool)+0x96; OM::Host::NeonatalMortality::eventNeonatalMortality()+0x9; OM::util::random::uniform\_01()+0xc0]\_\_Uniform\\
+6 & [forward()+0x204; OM::Simulator::start(scnXml::Monitoring const\&)+0x468; OM::Population::update1(OM::SimTime)+0xff; OM::Host::Human::update(bool)+0x280; OM::WithinHost::DescriptiveWithinHostModel::update(int, std::vector \textgreater{}\&, double, double)+0x3a2; OM::WithinHost::DescriptiveInfection::determineDensities(double, double, double, double\&, double, double)+0x303; OM::util::random::uniform\_01()+0xc0]\_\_Uniform\\
+7 & [forward()+0x204; OM::Simulator::start(scnXml::Monitoring const\&)+0x468; OM::Population::update1(OM::SimTime)+0xff; OM::Host::Human::update(bool)+0x2bc; OM::Clinical::ClinicalModel::update(OM::Host::Human\&, double, bool)+0xe8; OM::Clinical::CM5DayCommon::doClinicalUpdate(OM::Host::Human\&, double)+0x67; OM::WithinHost::WHFalciparum::determineMorbidity(OM::Host::Human\&, double, bool)+0x73; OM::WithinHost::Pathogenesis::PathogenesisModel::determineState(OM::Host::Human\&, double, double, double, bool)+0x143; OM::util::random::bernoulli(double)+0x47; OM::util::random::uniform\_01()+0xc0]\_\_Uniform\\
+8 & [forward()+0x204; OM::Simulator::start(scnXml::Monitoring const\&)+0x468; OM::Population::update1(OM::SimTime)+0xff; OM::Host::Human::update(bool)+0x2bc; OM::Clinical::ClinicalModel::update(OM::Host::Human\&, double, bool)+0xe8; OM::Clinical::CM5DayCommon::doClinicalUpdate(OM::Host::Human\&, double)+0x67; OM::WithinHost::WHFalciparum::determineMorbidity(OM::Host::Human\&, double, bool)+0x73; OM::WithinHost::Pathogenesis::PathogenesisModel::determineState(OM::Host::Human\&, double, double, double, bool)+0x162; OM::util::random::bernoulli(double)+0x47; OM::util::random::uniform\_01()+0xc0]\_\_Uniform\\
+9 & [forward()+0x204; OM::Simulator::start(scnXml::Monitoring const\&)+0x468; OM::Population::update1(OM::SimTime)+0xff; OM::Host::Human::update(bool)+0x2bc; OM::Clinical::ClinicalModel::update(OM::Host::Human\&, double, bool)+0xe8; OM::Clinical::CM5DayCommon::doClinicalUpdate(OM::Host::Human\&, double)+0x67; OM::WithinHost::WHFalciparum::determineMorbidity(OM::Host::Human\&, double, bool)+0x73; OM::WithinHost::Pathogenesis::PathogenesisModel::determineState(OM::Host::Human\&, double, double, double, bool)+0x1c7; OM::util::random::bernoulli(double)+0x47; OM::util::random::uniform\_01()+0xc0]\_\_Uniform\\
+10 & [forward()+0x204; OM::Simulator::start(scnXml::Monitoring const\&)+0x468; OM::Population::update1(OM::SimTime)+0xff; OM::Host::Human::update(bool)+0x2bc; OM::Clinical::ClinicalModel::update(OM::Host::Human\&, double, bool)+0xe8; OM::Clinical::CM5DayCommon::doClinicalUpdate(OM::Host::Human\&, double)+0xe3; OM::Clinical::ImmediateOutcomes::uncomplicatedEvent(OM::Host::Human\&, OM::Clinical::Episode::State)+0x87; OM::util::random::uniform\_01()+0xc0]\_\_Uniform\\
+11 & [forward()+0x204; OM::Simulator::start(scnXml::Monitoring const\&)+0x468; OM::Population::update1(OM::SimTime)+0xff; OM::Host::Human::update(bool)+0x2bc; OM::Clinical::ClinicalModel::update(OM::Host::Human\&, double, bool)+0xe8; OM::Clinical::CM5DayCommon::doClinicalUpdate(OM::Host::Human\&, double)+0xad; OM::Clinical::CM5DayCommon::severeMalaria(OM::Host::Human\&, OM::Clinical::Episode::State, double, int\&)+0x4b1; OM::util::random::uniform\_01()+0xc0]\_\_Uniform\\
+\bottomrule                                                                                                                                   
+\end{tabularx}
+ \end{table*}
 % \subsubsection*{Acknowledgments}

### 2019-04-24 paper/icml_ai4good/main.tex
started intro

--- 2019-04-23 paper/icml_ai4good/main.tex
@@
 \section{Introduction}
+Ending the epidemics of AIDS, tuberculosis, malaria and other infectious diseases by 2030 is a key target by the Good Health & Well-Being section of the UN Sustainable Development Goals.
+Despite decades of substantial international efforts, malaria still annually kills about a quarter of a million children under the age of 5 in Africa alone.
 
-To do.
+Malaria can be routinely cured thanks to a variety of medical interventions. 
+Issues are further aggravated due to the emergence of parasite resistance to multiple drugs, including artemisinin, or even insecticides.
 
-\begin{itemize}
+To reach the WHO’s target of reducing malaria incidence and mortality rates by at least 90\% by 2030, policymakers increasingly turn to evidence-based methods.
+However, the collection of large quantities of field data is often infeasible due to logistic, economic or political constraints, or outright impossible if novel methods are to be assessed a priori.
+Hence in silico simulation of malaria epidemiology plays an increasingly crucial role in evidence-based policymaking.
+%Hence epidemiology simulators fine-tuned on real data play a crucial role in determining policy implications.
 
-\item Outline the objectives
-\item State the importance of interpretability in the inference of the simulator
-\item Our system enables one to understand and interpret the most common paths in a program (simulation).
-\item In highlighting this a practitioner can then go back to the field and explore that parameter/s in more depth.
-\item This facilitates understanding and the construction of more detailed models and data collection strategies
+Even in the absence of political, economic and social constraints, Malaria epidemiology is governed by a complex set of drivers, few of which can be understood in isolation.
+These include within-host dynamics, population-specific traits and even local geography.
+Some progress in modeling specific components has been made, such as the parasite life-cycle, transmission vector dynamics and the inter-dependencies between infection rates and vicinity to open water.
+However, comprehensive modeling of all of these components in a regional context remains challenging, and is greatly impeded by the lack of available field data.
 
-\end{itemize}
+Studying the implications of a diverse set of anti-malaria
 
-% taken from Toms thesis.
+All of the above limitations require contemporary malaria epidemiology simulators to employ stochastic modeling techniques, which negotiate data scarcity and model uncertainty through the use of appropriate prior distributions.
+Simulator calibration then equates to calibrating the
 
-Stochastic simulators are common across a number of domains, statistical physics~\cite{landau_binder_2014}, financial modeling~\cite{jackel2002monte},
-weather prediction~\cite{evensen1994sequential}, epidemiology~\cite{smith2008towards} and many others. 
-An advantage of using stochastic simulators is that they provide a level of interpretability not found in modern deep learning settings, as they directly incorporate model structure from carefully reasoned observations and experiments.
-Probabilistic programming systems~(PPSs) are purpose built systems for making inferences and probabilistic 
-modeling accessible. 
-In particular, PPSs allow probabilistic models to be represented in the form of a generative model through the
-probabilistic programming language~(PPL), which enables one to write statements that enable conditioning on data~\cite{gordon2014probabilistic,Goodman08church:a}. 
-Thus, it seems only natural to connect simulators with PPSs, as simulators explicitly define a generative model, in the language in which they are written, and the nature of the PPSs enables us to perform inference in these simulators, by conditioning on observations, which are fed in as input to the simulators.
-In doing this we can infer things about stochastic input parameters and other variables sampled during the program's forward execution, whilst also providing full interpretability in the inference results, which is absolutely necessary in safety-critical domains.
-By exploiting the tools that we develop in this paper, one is able to connect epidemiology simulators with \texttt{Pyprob}~\cite{le-2016-inference} to extract fully interpretable posterior structures allowing practitioners to interpret which input parameters have the largest affect on the epidemic, in our case malaria.
+, and, above all, the lack of available field data for calibration,
+
+%parasite life-cycle, transmission vectors, population dynamics and even local geographic circumstances, with policy measures ranging from distribution of mosquito nets to experimental vaccines, requires such simulators to be stochastic.
+Simulation outcomes thus need to be evaluated on a population of subjects.
+
+<Section about prob prog inference, with reference to SHERPA>
+
+In this paper, we for the first time demonstrate the applicability of trace introspection, and hence universal probabilistic inference, to population-based simulators.
+In particular, we discuss how the leading malaria epidemiology simulators EMOD[] and OpenMalaria[] can be interfaced with <TODO>.
+We argue that our work is likely to have significant social impact on many livelihoods affected by Malaria and discuss opportunities for future research.
+
+- issues of public health are complex and
+
+%To do. 
+%
+%\begin{itemize}
+%
+%\item Outline the objectives
+%\item State the importance of interpretability in the inference of the simulator
+%\item Our system enables one to understand and interpret the most common paths in a program (simulation).
+%\item In highlighting this a practitioner can then go back to the field and explore that parameter/s in more depth.
+%\item This facilitates understanding and the construction of more detailed models and data collection strategies
+%
+%
+%\end{itemize}
+%
+%% taken from Toms thesis. 
+%
+%Stochastic simulators are common across a number of domains, statistical physics~\cite{landau_binder_2014}, financial modeling~\cite{jackel2002monte},
+%weather prediction~\cite{evensen1994sequential}, epidemiology~\cite{smith2008towards} and many others. 
+%An advantage of using stochastic simulators is that they provide a level of interpretability not found in modern deep learning settings, as they directly incorporate model structure from carefully reasoned observations and experiments.
+%Probabilistic programming systems~(PPSs) are purpose built systems for making inferences and probabilistic 
+%modeling accessible. 
+%In particular, PPSs allow probabilistic models to be represented in the form of a generative model through the
+%probabilistic programming language~(PPL), which enables one to write statements that enable conditioning on data~\cite{gordon2014probabilistic,Goodman08church:a}. 
+%Thus, it seems only natural to connect simulators with PPSs, as simulators explicitly define a generative model, in the language in which they are written, and the nature of the PPSs enables us to perform inference in these simulators, by conditioning on observations, which are fed in as input to the simulators.
+%In doing this we can infer things about stochastic input parameters and other variables sampled during the program's forward execution, whilst also providing full interpretability in the inference results, which is absolutely necessary in safety-critical domains.
+%By exploiting the tools that we develop in this paper, one is able to connect epidemiology simulators with \texttt{Pyprob}~\cite{le-2016-inference} to extract fully interpretable posterior structures allowing practitioners to interpret which input parameters have the largest affect on the epidemic, in our case malaria.
 Over five-hundred-thousand people die from malaria each year, mostly children under five years of age, with 90 per cent of malaria cases occurring in Sub-Saharan Africa.
 An estimated 100-300 million people suffer from malaria each year~\cite{world2016world}.
-Thus, by understanding the prominence of input parameters to epidemiology simulators, guides decision making processes relating to effective treatment and prevention strategies. 
-\label{sec:related}
+%Thus, by understanding the prominence of input parameters to epidemiology simulators, guides decision making processes relating to effective treatment and prevention strategies. 
+%\label{sec:related}
 
@@
 
-\section{Results}
-
-TO ADD
-
+% Please add the following required packages to your document preamble:
+% \usepackage{booktabs}
+% Please add the following required packages to your document preamble:
+% \usepackage{booktabs}
+% \usepackage{graphicx}
+% Please add the following required packages to your document preamble:
+% \usepackage{graphicx}
+%\begin{table*}[b!]
+%\footnotesize
+%\setlength{\tabcolsep}{1mm}
+%\caption{Addresses}
+%\label{table:addresses}
+%\def\arraystretch{1.25}
+%% \setlength{\tabcolsep}{1mm}
+%\begin{tabularx}{\textwidth}{@{}lX@{}} 
+%\toprule
+%Address & Physical event                                                                                                                            \midrule
+%1 & [forward()+0x204; OM::Simulator::start(scnXml::Monitoring const)+0x28a; OM::Population::createInitialHumans()+0x94; OM::Population::newHuman(OM::SimTime)+0x5c; OM::Host::Human::Human(OM::SimTime)+0x12b; OM::WithinHost::WHInterface::createWithinHostModel(double)+0x99; OM::WithinHost::DescriptiveWithinHostModel::DescriptiveWithinHostModel(double)+0x3a; OM::WithinHost::WHFalciparum::WHFalciparum(double)+0xe6; OM::util::random::gauss(double, double)+0xb4]\_\_Normal\\
+%2 & [forward()+0x204; OM::Simulator::start(scnXml::Monitoring const\&)+0x468; OM::Population::update1(OM::SimTime)+0x1e9; OM::Population::newHuman(OM::SimTime)+0x5c; OM::Host::Human::Human(OM::SimTime)+0x12b; OM::WithinHost::WHInterface::createWithinHostModel(double)+0x99; OM::WithinHost::DescriptiveWithinHostModel::DescriptiveWithinHostModel(double)+0x3a; OM::WithinHost::WHFalciparum::WHFalciparum(double)+0xe6; OM::util::random::gauss(double, double)+0xb4]\_\_Normal\\
+%3 & [forward()+0x204; OM::Simulator::start(scnXml::Monitoring const\&)+0x468; OM::Population::update1(OM::SimTime)+0xff; OM::Host::Human::update(bool)+0x22c; OM::Host::InfectionIncidenceModel::numNewInfections(OM::Host::Human const\&, double)+0x2a3; OM::util::random::poisson(double)+0x13f]\_\_Poisson\\
+%4 & [forward()+0x204; OM::Simulator::start(scnXml::Monitoring const\&)+0x468; OM::Population::update1(OM::SimTime)+0xff; OM::Host::Human::update(bool)+0x2bc; OM::Clinical::ClinicalModel::update(OM::Host::Human\&, double, bool)+0x96; OM::Host::NeonatalMortality::eventNeonatalMortality()+0x9; OM::util::random::uniform\_01()+0xc0]\_\_Uniform\\
+%5 & [forward()+0x204; OM::Simulator::start(scnXml::Monitoring const\&)+0x468; OM::Population::update1(OM::SimTime)+0xff; OM::Host::Human::update(bool)+0x2bc; OM::Clinical::ClinicalModel::update(OM::Host::Human\&, double, bool)+0x96; OM::Host::NeonatalMortality::eventNeonatalMortality()+0x9; OM::util::random::uniform\_01()+0xc0]\_\_Uniform\\
+%6 & [forward()+0x204; OM::Simulator::start(scnXml::Monitoring const\&)+0x468; OM::Population::update1(OM::SimTime)+0xff; OM::Host::Human::update(bool)+0x280; OM::WithinHost::DescriptiveWithinHostModel::update(int, std::vector \textgreater{}\&, double, double)+0x3a2; OM::WithinHost::DescriptiveInfection::determineDensities(double, double, double, double\&, double, double)+0x303; OM::util::random::uniform\_01()+0xc0]\_\_Uniform\\
+%7 & [forward()+0x204; OM::Simulator::start(scnXml::Monitoring const\&)+0x468; OM::Population::update1(OM::SimTime)+0xff; OM::Host::Human::update(bool)+0x2bc; OM::Clinical::ClinicalModel::update(OM::Host::Human\&, double, bool)+0xe8; OM::Clinical::CM5DayCommon::doClinicalUpdate(OM::Host::Human\&, double)+0x67; OM::WithinHost::WHFalciparum::determineMorbidity(OM::Host::Human\&, double, bool)+0x73; OM::WithinHost::Pathogenesis::PathogenesisModel::determineState(OM::Host::Human\&, double, double, double, bool)+0x143; OM::util::random::bernoulli(double)+0x47; OM::util::random::uniform\_01()+0xc0]\_\_Uniform\\
+%8 & [forward()+0x204; OM::Simulator::start(scnXml::Monitoring const\&)+0x468; OM::Population::update1(OM::SimTime)+0xff; OM::Host::Human::update(bool)+0x2bc; OM::Clinical::ClinicalModel::update(OM::Host::Human\&, double, bool)+0xe8; OM::Clinical::CM5DayCommon::doClinicalUpdate(OM::Host::Human\&, double)+0x67; OM::WithinHost::WHFalciparum::determineMorbidity(OM::Host::Human\&, double, bool)+0x73; OM::WithinHost::Pathogenesis::PathogenesisModel::determineState(OM::Host::Human\&, double, double, double, bool)+0x162; OM::util::random::bernoulli(double)+0x47; OM::util::random::uniform\_01()+0xc0]\_\_Uniform\\
+%9 & [forward()+0x204; OM::Simulator::start(scnXml::Monitoring const\&)+0x468; OM::Population::update1(OM::SimTime)+0xff; OM::Host::Human::update(bool)+0x2bc; OM::Clinical::ClinicalModel::update(OM::Host::Human\&, double, bool)+0xe8; OM::Clinical::CM5DayCommon::doClinicalUpdate(OM::Host::Human\&, double)+0x67; OM::WithinHost::WHFalciparum::determineMorbidity(OM::Host::Human\&, double, bool)+0x73; OM::WithinHost::Pathogenesis::PathogenesisModel::determineState(OM::Host::Human\&, double, double, double, bool)+0x1c7; OM::util::random::bernoulli(double)+0x47; OM::util::random::uniform\_01()+0xc0]\_\_Uniform\\
+%10 & [forward()+0x204; OM::Simulator::start(scnXml::Monitoring const\&)+0x468; OM::Population::update1(OM::SimTime)+0xff; OM::Host::Human::update(bool)+0x2bc; OM::Clinical::ClinicalModel::update(OM::Host::Human\&, double, bool)+0xe8; OM::Clinical::CM5DayCommon::doClinicalUpdate(OM::Host::Human\&, double)+0xe3; OM::Clinical::ImmediateOutcomes::uncomplicatedEvent(OM::Host::Human\&, OM::Clinical::Episode::State)+0x87; OM::util::random::uniform\_01()+0xc0]\_\_Uniform\\
+%11 & [forward()+0x204; OM::Simulator::start(scnXml::Monitoring const\&)+0x468; OM::Population::update1(OM::SimTime)+0xff; OM::Host::Human::update(bool)+0x2bc; OM::Clinical::ClinicalModel::update(OM::Host::Human\&, double, bool)+0xe8; OM::Clinical::CM5DayCommon::doClinicalUpdate(OM::Host::Human\&, double)+0xad; OM::Clinical::CM5DayCommon::severeMalaria(OM::Host::Human\&, OM::Clinical::Episode::State, double, int\&)+0x4b1; OM::util::random::uniform\_01()+0xc0]\_\_Uniform\\
+%\bottomrule                                                                                                                                   
+%\end{tabularx}
+% \end{table*}
 % \subsubsection*{Acknowledgments}

### 2019-04-24 paper/icml_ai4good/main.tex
next intro update

@@
 \section{Introduction}
-Ending the epidemics of AIDS, tuberculosis, malaria and other infectious diseases by 2030 is a key target by the Good Health & Well-Being section of the UN Sustainable Development Goals.
-Despite decades of substantial international efforts, malaria still annually kills about a quarter of a million children under the age of 5 in Africa alone.
+Ending the epidemics of AIDS, tuberculosis, malaria and other infectious diseases by 2030 is a key target within the Good Health \& Well-Being section of the UN Sustainable Development Goals\cite{}.
 
-Malaria can be routinely cured thanks to a variety of medical interventions. 
-Issues are further aggravated due to the emergence of parasite resistance to multiple drugs, including artemisinin, or even insecticides.
+In contrast to AIDS, Malaria can be routinely cured thanks to the disposability of a variety of medical interventions. 
+Issues are further aggravated due to the increasing emergence of parasite resistance to multiple drugs, including artemisinin, or even insecticides.
+Hence, despite decades of substantial international efforts, initial hopes of regional elimination or even global eradication have not been met.
+So difficult is Malaria control that, even today, about a quarter of a million children under the age of 5 succumb to it every year in Africa alone.
 
@@
 Some progress in modeling specific components has been made, such as the parasite life-cycle, transmission vector dynamics and the inter-dependencies between infection rates and vicinity to open water.
-However, comprehensive modeling of all of these components in a regional context remains challenging, and is greatly impeded by the lack of available field data.
+However, comprehensive modeling of all of these components in a regional context remains challenging, and is greatly impeded by the lack of available field data for regional calibration.
+Simulational complexity is further aggravated by considering feedback effects from a large arsenal of anti-malaria interventions, ranging from distribution of mosquito nets, indoor residual spraying and mass drug administration to age-specific vaccinations.
 
-Studying the implications of a diverse set of anti-malaria
+All of the above limitations require contemporary malaria epidemiology simulators to employ stochastic modeling techniques, which negotiate data scarcity and model uncertainty through the use of appropriate prior distributions that can be tuned even from small samples of field data.
+The population dynamics of groups of typically several thousand hosts are then initialised by sampling from the initial distributionsm, using initial random seeds.
+Forward simulation then proceeds by rolling out simulation dynamics over time.
+Policymakers then take interest in comparing simulation outcomes of cherry-picked scenarios and interventions with those of suitably chosen control groups.
 
-All of the above limitations require contemporary malaria epidemiology simulators to employ stochastic modeling techniques, which negotiate data scarcity and model uncertainty through the use of appropriate prior distributions.
-Simulator calibration then equates to calibrating the
+Stochastic epidemiology simulators have already proven to be an invaluable tool to policymakers with policy implications that are often counterintuitive yet far-ranging.
+For example, it has been shown that mass vaccination may be largely ineffective in regions of large transmission rates, but may play a crucial role in areas of low transmittivity.
 
-, and, above all, the lack of available field data for calibration,
+However valuable forward simulations of epidemiology scenarios are, the holy grail of goal-driven policymaking lies ultimately in the inverse problem: Given a desired outcome, we wish to find the optimal respective intervention policy given a specific regional context.
+Due to computational constraints, even an approximate search over the whole simulator input space usually proves intractable.
 
-%parasite life-cycle, transmission vectors, population dynamics and even local geographic circumstances, with policy measures ranging from distribution of mosquito nets to experimental vaccines, requires such simulators to be stochastic.
+% , and, above all, the lack of available field data for calibration,
+
+% parasite life-cycle, transmission vectors, population dynamics and even local geographic circumstances, with policy measures ranging from distribution of mosquito nets to experimental vaccines, requires such simulators to be stochastic.
 Simulation outcomes thus need to be evaluated on a population of subjects.

### 2019-04-24 paper/icml_ai4good/main.tex
updated

--- 2019-04-24 paper/icml_ai4good/main.tex
@@
 \section{Introduction}
-Ending the epidemics of AIDS, tuberculosis, malaria and other infectious diseases by 2030 is a key target by the Good Health & Well-Being section of the UN Sustainable Development Goals.
+Ending the epidemics of AIDS, tuberculosis, malaria and other infectious diseases by 2030 is a key target by the Good Health \& Well-Being section of the UN Sustainable Development Goals.
 Despite decades of substantial international efforts, malaria still annually kills about a quarter of a million children under the age of 5 in Africa alone.
@@
 % Please add the following required packages to your document preamble:
-% \usepackage{graphicx}
-%\begin{table*}[b!]
-%\footnotesize
-%\setlength{\tabcolsep}{1mm}
-%\caption{Addresses}
-%\label{table:addresses}
-%\def\arraystretch{1.25}
-%% \setlength{\tabcolsep}{1mm}
-%\begin{tabularx}{\textwidth}{@{}lX@{}} 
-%\toprule
-%Address & Physical event                                                                                                                            \midrule
-%1 & [forward()+0x204; OM::Simulator::start(scnXml::Monitoring const)+0x28a; OM::Population::createInitialHumans()+0x94; OM::Population::newHuman(OM::SimTime)+0x5c; OM::Host::Human::Human(OM::SimTime)+0x12b; OM::WithinHost::WHInterface::createWithinHostModel(double)+0x99; OM::WithinHost::DescriptiveWithinHostModel::DescriptiveWithinHostModel(double)+0x3a; OM::WithinHost::WHFalciparum::WHFalciparum(double)+0xe6; OM::util::random::gauss(double, double)+0xb4]\_\_Normal\\
-%2 & [forward()+0x204; OM::Simulator::start(scnXml::Monitoring const\&)+0x468; OM::Population::update1(OM::SimTime)+0x1e9; OM::Population::newHuman(OM::SimTime)+0x5c; OM::Host::Human::Human(OM::SimTime)+0x12b; OM::WithinHost::WHInterface::createWithinHostModel(double)+0x99; OM::WithinHost::DescriptiveWithinHostModel::DescriptiveWithinHostModel(double)+0x3a; OM::WithinHost::WHFalciparum::WHFalciparum(double)+0xe6; OM::util::random::gauss(double, double)+0xb4]\_\_Normal\\
-%3 & [forward()+0x204; OM::Simulator::start(scnXml::Monitoring const\&)+0x468; OM::Population::update1(OM::SimTime)+0xff; OM::Host::Human::update(bool)+0x22c; OM::Host::InfectionIncidenceModel::numNewInfections(OM::Host::Human const\&, double)+0x2a3; OM::util::random::poisson(double)+0x13f]\_\_Poisson\\
-%4 & [forward()+0x204; OM::Simulator::start(scnXml::Monitoring const\&)+0x468; OM::Population::update1(OM::SimTime)+0xff; OM::Host::Human::update(bool)+0x2bc; OM::Clinical::ClinicalModel::update(OM::Host::Human\&, double, bool)+0x96; OM::Host::NeonatalMortality::eventNeonatalMortality()+0x9; OM::util::random::uniform\_01()+0xc0]\_\_Uniform\\
-%5 & [forward()+0x204; OM::Simulator::start(scnXml::Monitoring const\&)+0x468; OM::Population::update1(OM::SimTime)+0xff; OM::Host::Human::update(bool)+0x2bc; OM::Clinical::ClinicalModel::update(OM::Host::Human\&, double, bool)+0x96; OM::Host::NeonatalMortality::eventNeonatalMortality()+0x9; OM::util::random::uniform\_01()+0xc0]\_\_Uniform\\
-%6 & [forward()+0x204; OM::Simulator::start(scnXml::Monitoring const\&)+0x468; OM::Population::update1(OM::SimTime)+0xff; OM::Host::Human::update(bool)+0x280; OM::WithinHost::DescriptiveWithinHostModel::update(int, std::vector \textgreater{}\&, double, double)+0x3a2; OM::WithinHost::DescriptiveInfection::determineDensities(double, double, double, double\&, double, double)+0x303; OM::util::random::uniform\_01()+0xc0]\_\_Uniform\\
-%7 & [forward()+0x204; OM::Simulator::start(scnXml::Monitoring const\&)+0x468; OM::Population::update1(OM::SimTime)+0xff; OM::Host::Human::update(bool)+0x2bc; OM::Clinical::ClinicalModel::update(OM::Host::Human\&, double, bool)+0xe8; OM::Clinical::CM5DayCommon::doClinicalUpdate(OM::Host::Human\&, double)+0x67; OM::WithinHost::WHFalciparum::determineMorbidity(OM::Host::Human\&, double, bool)+0x73; OM::WithinHost::Pathogenesis::PathogenesisModel::determineState(OM::Host::Human\&, double, double, double, bool)+0x143; OM::util::random::bernoulli(double)+0x47; OM::util::random::uniform\_01()+0xc0]\_\_Uniform\\
-%8 & [forward()+0x204; OM::Simulator::start(scnXml::Monitoring const\&)+0x468; OM::Population::update1(OM::SimTime)+0xff; OM::Host::Human::update(bool)+0x2bc; OM::Clinical::ClinicalModel::update(OM::Host::Human\&, double, bool)+0xe8; OM::Clinical::CM5DayCommon::doClinicalUpdate(OM::Host::Human\&, double)+0x67; OM::WithinHost::WHFalciparum::determineMorbidity(OM::Host::Human\&, double, bool)+0x73; OM::WithinHost::Pathogenesis::PathogenesisModel::determineState(OM::Host::Human\&, double, double, double, bool)+0x162; OM::util::random::bernoulli(double)+0x47; OM::util::random::uniform\_01()+0xc0]\_\_Uniform\\
-%9 & [forward()+0x204; OM::Simulator::start(scnXml::Monitoring const\&)+0x468; OM::Population::update1(OM::SimTime)+0xff; OM::Host::Human::update(bool)+0x2bc; OM::Clinical::ClinicalModel::update(OM::Host::Human\&, double, bool)+0xe8; OM::Clinical::CM5DayCommon::doClinicalUpdate(OM::Host::Human\&, double)+0x67; OM::WithinHost::WHFalciparum::determineMorbidity(OM::Host::Human\&, double, bool)+0x73; OM::WithinHost::Pathogenesis::PathogenesisModel::determineState(OM::Host::Human\&, double, double, double, bool)+0x1c7; OM::util::random::bernoulli(double)+0x47; OM::util::random::uniform\_01()+0xc0]\_\_Uniform\\
-%10 & [forward()+0x204; OM::Simulator::start(scnXml::Monitoring const\&)+0x468; OM::Population::update1(OM::SimTime)+0xff; OM::Host::Human::update(bool)+0x2bc; OM::Clinical::ClinicalModel::update(OM::Host::Human\&, double, bool)+0xe8; OM::Clinical::CM5DayCommon::doClinicalUpdate(OM::Host::Human\&, double)+0xe3; OM::Clinical::ImmediateOutcomes::uncomplicatedEvent(OM::Host::Human\&, OM::Clinical::Episode::State)+0x87; OM::util::random::uniform\_01()+0xc0]\_\_Uniform\\
-%11 & [forward()+0x204; OM::Simulator::start(scnXml::Monitoring const\&)+0x468; OM::Population::update1(OM::SimTime)+0xff; OM::Host::Human::update(bool)+0x2bc; OM::Clinical::ClinicalModel::update(OM::Host::Human\&, double, bool)+0xe8; OM::Clinical::CM5DayCommon::doClinicalUpdate(OM::Host::Human\&, double)+0xad; OM::Clinical::CM5DayCommon::severeMalaria(OM::Host::Human\&, OM::Clinical::Episode::State, double, int\&)+0x4b1; OM::util::random::uniform\_01()+0xc0]\_\_Uniform\\
-%\bottomrule                                                                                                                                   
-%\end{tabularx}
+% 
+% \begin{table*}[b!]
+% \footnotesize
+% \setlength{\tabcolsep}{1mm}
+% \caption{Addresses}
+% \label{table:addresses}
+% \def\arraystretch{1.25}
+% % \setlength{\tabcolsep}{1mm}
+% \begin{tabularx}{\textwidth}{@{}lX@{}} 
+% \toprule
+% Address & Physical event                                                                                                                            \midrule
+%
+% 
+% 
+% \bottomrule                                                                                                                                   
+% \end{tabularx}
 % \end{table*}
+
+  \begin{table*}[b!]
+  \footnotesize
+  \setlength{\tabcolsep}{1mm}
+  \caption{Addresses }
+  \label{table:addresses}
+  \def\arraystretch{1.25}
+  % \setlength{\tabcolsep}{1mm}
+  \begin{tabularx}{\textwidth}{@{}lX@{}} 
+    \toprule
+    Address ID & Full address \\
+    \midrule
+  A1 & [forward()+0x204; OM::Simulator::start(scnXml::Monitoring const)+0x28a;
+
+  OM::Population::createInitialHumans()+0x94;
+
+  OM::Population::newHuman(OM::SimTime)+0x5c;
+
+  OM::Host::Human::Human(OM::SimTime)+0x12b;
+
+  OM::WithinHost::WHInterface::createWithinHostModel(double)+0x99;
+
+  OM::WithinHost::DescriptiveWithinHostModel::DescriptiveWithinHostModel(double)+0x3a;
+
+  OM::WithinHost::WHFalciparum::WHFalciparum(double)+0xe6;
+
+  OM::util::random::gauss(double, double)+0xb4]\_\_Normal\\
+  A2 & [forward()+0x204; 
+  OM::Simulator::start(scnXml::Monitoring const\&)+0x468;
+
+   OM::Population::update1(OM::SimTime)+0x1e9;
+
+   OM::Population::newHuman(OM::SimTime)+0x5c;
+
+   OM::Host::Human::Human(OM::SimTime)+0x12b;
+
+   OM::WithinHost::WHInterface::createWithinHostModel(double)+0x99;
+
+   OM::WithinHost::DescriptiveWithinHostModel::DescriptiveWithinHostModel(double)+0x3a;
+
+   OM::WithinHost::WHFalciparum::WHFalciparum(double)+0xe6;
+
+   OM::util::random::gauss(double, double)+0xb4]\_\_Normal \\
+  A3 & [forward()+0x204; OM::Simulator::start(scnXml::Monitoring const\&)+0x468;
+
+   OM::Population::update1(OM::SimTime)+0xff;
+
+   OM::Host::Human::update(bool)+0x22c;
+
+   OM::Host::InfectionIncidenceModel::numNewInfections(OM::Host::Human const\&, double)+0x2a3;
+
+   OM::util::random::poisson(double)+0x13f]\_\_Poisson\\
+  A4 & [forward()+0x204; OM::Simulator::start(scnXml::Monitoring const\&)+0x468;
+
+  OM::Population::update1(OM::SimTime)+0xff; OM::Host::Human::update(bool)+0x2bc;
+
+   OM::Clinical::ClinicalModel::update(OM::Host::Human\&, double, bool)+0x96;
+
+   OM::Host::NeonatalMortality::eventNeonatalMortality()+0x9;
+
+    OM::util::random::uniform\_01()+0xc0]\_\_Uniform\\
+  A5 & [forward()+0x204; OM::Simulator::start(scnXml::Monitoring const\&)+0x468;
+
+   OM::Population::update1(OM::SimTime)+0xff;
+
+    OM::Host::Human::update(bool)+0x2bc;
+
+    OM::Clinical::ClinicalModel::update(OM::Host::Human\&, double, bool)+0x96;
+
+    OM::Host::NeonatalMortality::eventNeonatalMortality()+0x9;
+
+    OM::util::random::uniform\_01()+0xc0]\_\_Uniform\\
+  A6 & [forward()+0x204; OM::Simulator::start(scnXml::Monitoring const\&)+0x468;
+
+   OM::Population::update1(OM::SimTime)+0xff;
+
+   OM::Host::Human::update(bool)+0x280;
+
+   OM::WithinHost::DescriptiveWithinHostModel::update(int, std::vector \textgreater{}\&, double, double)+0x3a2;
+
+   OM::WithinHost::DescriptiveInfection::determineDensities(double, double, double, double\&, double, double)+0x303;
+
+   OM::util::random::uniform\_01()+0xc0]\_\_Uniform\\
+   A7 & [forward()+0x204; OM::Simulator::start(scnXml::Monitoring const\&)+0x468;
+   ı
+	OM::Population::update1(OM::SimTime)+0xff;
+
+	OM::Host::Human::update(bool)+0x2bc;
+
+	OM::Clinical::ClinicalModel::update(OM::Host::Human\&, double, bool)+0xe8;
+
+	OM::Clinical::CM5DayCommon::doClinicalUpdate(OM::Host::Human\&, double)+0x67;
+
+	OM::WithinHost::WHFalciparum::determineMorbidity(OM::Host::Human\&, double, bool)+0x73;
+ı
+	OM::WithinHost::Pathogenesis::PathogenesisModel::determineState(OM::Host::Human\&, double, double, double, bool)+0x143;
+
+	OM::util::random::bernoulli(double)+0x47;
+
+	OM::util::random::uniform\_01()+0xc0]\_\_Uniform\\
+	A8 & [forward()+0x204; OM::Simulator::start(scnXml::Monitoring const\&)+0x468;
+
+	OM::Population::update1(OM::SimTime)+0xff; OM::Host::Human::update(bool)+0x2bc;
+
+	OM::Clinical::ClinicalModel::update(OM::Host::Human\&, double, bool)+0xe8;
+
+	OM::Clinical::CM5DayCommon::doClinicalUpdate(OM::Host::Human\&, double)+0x67;
+
+	OM::WithinHost::WHFalciparum::determineMorbidity(OM::Host::Human\&, double, bool)+0x73;
+
+	OM::WithinHost::Pathogenesis::PathogenesisModel::determineState(OM::Host::Human\&, double, double, double, bool)+0x162;
+
+	OM::util::random::bernoulli(double)+0x47;
+
+	OM::util::random::uniform\_01()+0xc0]\_\_Uniform\\
+A9 & [forward()+0x204; OM::Simulator::start(scnXml::Monitoring const\&)+0x468;
+ OM::Population::update1(OM::SimTime)+0xff; OM::Host::Human::update(bool)+0x2bc;
+
+ OM::Clinical::ClinicalModel::update(OM::Host::Human\&, double, bool)+0xe8;
+
+  OM::Clinical::CM5DayCommon::doClinicalUpdate(OM::Host::Human\&, double)+0x67;
+
+  OM::WithinHost::WHFalciparum::determineMorbidity(OM::Host::Human\&, double, bool)+0x73;
+
+  OM::WithinHost::Pathogenesis::PathogenesisModel::determineState(OM::Host::Human\&, double, double, double, bool)+0x1c7;
+
+  OM::util::random::bernoulli(double)+0x47; OM::util::random::uniform\_01()+0xc0]\_\_Uniform\\
+\bottomrule
+  \end{tabularx}
+  \end{table*}
 % \subsubsection*{Acknowledgments}
@@
 % submission, only in the final paper.
+  \begin{table*}[ht!]
+  \footnotesize
+  \setlength{\tabcolsep}{1mm}
+  \caption{Addresses .}
+  \label{table:addresses}
+  \def\arraystretch{1.25}
+  % \setlength{\tabcolsep}{1mm}
+  \begin{tabularx}{\textwidth}{@{}lX@{}} 
+    \toprule
+    Address ID & Full address \\
+    \midrule
+A10 & [forward()+0x204; OM::Simulator::start(scnXml::Monitoring const\&)+0x468;
 
+OM::Population::update1(OM::SimTime)+0xff; OM::Host::Human::update(bool)+0x2bc;
+
+OM::Clinical::ClinicalModel::update(OM::Host::Human\&, double, bool)+0xe8;
+
+OM::Clinical::CM5DayCommon::doClinicalUpdate(OM::Host::Human\&, double)+0xe3;
+
+OM::Clinical::ImmediateOutcomes::uncomplicatedEvent(OM::Host::Human\&, OM::Clinical::Episode::State)+0x87;
+
+OM::util::random::uniform\_01()+0xc0]\_\_Uniform\\
+A11 & [forward()+0x204; OM::Simulator::start(scnXml::Monitoring const\&)+0x468;
+OM::Population::update1(OM::SimTime)+0xff; OM::Host::Human::update(bool)+0x2bc;
+
+OM::Clinical::ClinicalModel::update(OM::Host::Human\&, double, bool)+0xe8;
+
+OM::Clinical::CM5DayCommon::doClinicalUpdate(OM::Host::Human\&, double)+0xad;
+
+OM::Clinical::CM5DayCommon::severeMalaria(OM::Host::Human\&, OM::Clinical::Episode::State, double, int\&)+0x4b1;
+OM::util::random::uniform\_01()+0xc0]\_\_Uniform\\
+    \bottomrule
+  \end{tabularx}
+  \end{table*}
 \section*{References}

### 2019-04-24 paper/icml_ai4good/main.tex

@@
 \section{Introduction}
+
 Ending the epidemics of AIDS, tuberculosis, malaria and other infectious diseases by 2030 is a key target by the Good Health \& Well-Being section of the UN Sustainable Development Goals. 
 Despite decades of substantial international efforts, malaria still annually kills about a quarter of a million children under the age of 5 in Africa alone.
+Ending the epidemics of AIDS, tuberculosis, malaria and other infectious diseases by 2030 is a key target within the Good Health \& Well-Being section of the UN Sustainable Development Goals\cite{}.
 
-Malaria can be routinely cured thanks to a variety of medical interventions. 
-Issues are further aggravated due to the emergence of parasite resistance to multiple drugs, including artemisinin, or even insecticides.
+In contrast to AIDS, Malaria can be routinely cured thanks to the disposability of a variety of medical interventions. 
+Issues are further aggravated due to the increasing emergence of parasite resistance to multiple drugs, including artemisinin, or even insecticides. 
+Hence, despite decades of substantial international efforts, initial hopes of regional elimination or even global eradication have not been met.
+ So difficult is Malaria control that, even today, about a quarter of a million children under the age of 5 succumb to it every year in Africa alone.
 
@@
 Some progress in modeling specific components has been made, such as the parasite life-cycle, transmission vector dynamics and the inter-dependencies between infection rates and vicinity to open water.
-However, comprehensive modeling of all of these components in a regional context remains challenging, and is greatly impeded by the lack of available field data.
+However, comprehensive modeling of all of these components in a regional context remains challenging, and is greatly impeded by the lack of available field data for regional calibration.
+Simulation complexity is further aggravated by considering feedback effects from a large arsenal of anti-malaria interventions, ranging from distribution of mosquito nets, indoor residual spraying and mass drug administration to age-specific vaccinations.
 
-Studying the implications of a diverse set of anti-malaria
+All of the above limitations require contemporary malaria epidemiology simulators to employ stochastic modeling techniques, which negotiate data scarcity and model uncertainty through the use of appropriate prior distributions that can be tuned even from small samples of field data.
+The population dynamics of groups of typically several thousand hosts are then initialised by sampling from the initial distributions, using initial random seeds.
+Forward simulation then proceeds by rolling out simulation dynamics over time.
+Policymakers then take interest in comparing simulation outcomes of cherry-picked scenarios and interventions with those of suitably chosen control groups.
 
-All of the above limitations require contemporary malaria epidemiology simulators to employ stochastic modeling techniques, which negotiate data scarcity and model uncertainty through the use of appropriate prior distributions.
-Simulator calibration then equates to calibrating the
+Stochastic epidemiology simulators have already proven to be an invaluable tool to policymakers with policy implications that are often counterintuitive yet far-ranging.
+For example, it has been shown that mass vaccination may be largely ineffective in regions of large transmission rates, but may play a crucial role in areas of low transmission.
 
-, and, above all, the lack of available field data for calibration,
+However valuable forward simulations of epidemiology scenarios are, the holy grail of goal-driven policymaking lies ultimately in the inverse problem: Given a desired outcome, we wish to find the optimal respective intervention policy given a specific regional context.
+Due to computational constraints, even an approximate search over the whole simulator input space usually proves intractable.
 
-%parasite life-cycle, transmission vectors, population dynamics and even local geographic circumstances, with policy measures ranging from distribution of mosquito nets to experimental vaccines, requires such simulators to be stochastic.
+% , and, above all, the lack of available field data for calibration,
+
+% parasite life-cycle, transmission vectors, population dynamics and even local geographic circumstances, with policy measures ranging from distribution of mosquito nets to experimental vaccines, requires such simulators to be stochastic.
 Simulation outcomes thus need to be evaluated on a population of subjects.
@@
 Our method of hijacking simulators builds on the work of~\cite{baydin2018efficient} and extends the framework to encapsulate a more diverse range of simulators. 
-Our framework provides a simple solution and makes it easy to transform arbitrary stochastic simulators into probabilistic programs, regardless of the complexity of the simulator and the language that the simulator is written in\footnote{The framework currently supports 9 popular languages, but there is nothing to stop this being extended to any language of interest.}.
+Our framework provides a simple solution and makes it easy to transform arbitrary stochastic simulators into probabilistic programs, regardless of the complexity of the simulator and the language that the simulator is written in\footnote{The framework currently supports 9 popular languages, but there is nothing to stop this being extended to any language of interest.
 In addition to this, we can understand the structure of programs in simulators for both decoding \emph{black-box structures} and for posterior inference. 
-This enables two things.
-1) It enables software developers and researchers to understand complicated code bases.
+This enables two things. 1) It enables software developers and researchers to understand complicated code bases. 
 2) It provides interpretable inference results that provide policy makers with predictions that are truly interpretable, in the sense that the end-user of the inference results understands what physical events led to inference outcome.
@@
 This greatly improves the scope of applicability of our system, as this environment only has to be built once, which then enables the simulator to be run simultaneously on machines of different architectures machine.
-The second step requires one to
+The second step requires one to override the stochastic primitives in the language, for example, the \emph{Uniform} and \emph{Normal} draws. 
+A stochastic simulator may have a variety of other distribution and sampling primitives, but the underlying methodology is the same.
 

### 2019-04-24 paper/icml_ai4good/main.tex
next update - sorry had an unexpected delay

--- 2019-04-24 paper/icml_ai4good/main.tex
@@
 All of the above limitations require contemporary malaria epidemiology simulators to employ stochastic modeling techniques, which negotiate data scarcity and model uncertainty through the use of appropriate prior distributions that can be tuned even from small samples of field data.
-The population dynamics of groups of typically several thousand hosts are then initialised by sampling from the initial distributionsm, using initial random seeds.
-Forward simulation then proceeds by rolling out simulation dynamics over time.
-Policymakers then take interest in comparing simulation outcomes of cherry-picked scenarios and interventions with those of suitably chosen control groups.
+The population dynamics of groups of typically several thousand hosts are then initialised by sampling from the initial distributions, using initial random seeds.
+Forward simulation then proceeds by rolling out population dynamics over time.
+Simulation outcomes of cherry-picked scenarios and interventions of interest are then compared to suitably designed control runs.
 
@@
 
-However valuable forward simulations of epidemiology scenarios are, the holy grail of goal-driven policymaking lies ultimately in the inverse problem: Given a desired outcome, we wish to find the optimal respective intervention policy given a specific regional context.
+However valuable forward simulations of epidemiology scenarios are, the holy grail of goal-driven policymaking ultimately lies in solving the inverse problem: Given a desired outcome and regional context, identify the series of interventions that will best reach the desired target after a given amount of time.
 Due to computational constraints, even an approximate search over the whole simulator input space usually proves intractable.
+An experience-guided search over the simulation output of a small set of manually-designed scenarios of interest therefore long seemed the only feasible way of investigating implications of policy interventions.
+
+Recent work by \citep{Baydin} introduces a fundamental paradigm change that, for the first time, allows to tackle this inverse problem in a way that .
 
@@
 
-<Section about prob prog inference, with reference to SHERPA>
+% <Section about prob prog inference, with reference to SHERPA>
 

### 2019-04-24 paper/icml_ai4good/main.tex
updated

--- 2019-04-24 paper/icml_ai4good/main.tex
@@
 \usepackage{tabularx}
-
 % hyperref makes hyperlinks in the resulting PDF.
@@
 However, since the development happens over several decades, usually without comprehensive documentation, understanding a code base written with legacy libraries and software is incredibly difficult and an almost impossible task for a non-expert, or expert who has not had access to the previous years of development.
-Our method of hijacking simulators builds on the work of~\cite{baydin2018efficient} and extends the framework to encapsulate a more diverse range of simulators. 
-Our framework provides a simple solution and makes it easy to transform arbitrary stochastic simulators into probabilistic programs, regardless of the complexity of the simulator and the language that the simulator is written in\footnote{The framework currently supports 9 popular languages, but there is nothing to stop this being extended to any language of interest.
+Our method of overriding simulators builds on the work of~\cite{baydin2018efficient} and extends the framework to encapsulate a more diverse range of simulators. 
+Our framework provides a simple solution and makes it easy to transform arbitrary stochastic simulators into probabilistic programs, regardless of the complexity of the simulator and the language that the simulator is written in\footnote{The framework currently supports 9 popular languages, but there is nothing to stop this being extended to any language of interest.}
 In addition to this, we can understand the structure of programs in simulators for both decoding \emph{black-box structures} and for posterior inference. 
@@
 
-In order to hijack the stochastic simulator we are only required to override the location of the stochastic primitives, i.e.
+In order to overriding the stochastic simulator we are only required to override the location of the stochastic primitives, i.e.
 the operations that generate the stochasticity and randomness within the simulator.
@@
 % \subsection{Extending Probabilistic Programming Protocols}
-
 \begin{figure*}[h!]
@@
    \includegraphics[width=\textwidth]{../plots/ewan_25_pop_10.pdf}
-	\caption{Add something. Population 10}
+   \bigbreak 
+   \includegraphics[width=\textwidth]{../plots/ewan_25_pop_100.pdf}
+   % \label{fig:Ng2}
+
+\caption{In overriding the stochasticity in the simulator we are able to control all random variables in the program,
+	 from which we can understand the physical events that unfold. }
 \end{figure*}
-\begin{figure*}[h!]
-	\centering
-	\includegraphics[width=\textwidth]{../plots/ewan_25_pop_100.pdf}
-	\caption{Add something. Population 100}
-\end{figure*}
+
 TODO
@@
 
-  \begin{table*}[b!]
+  \begin{table*}[h!]
   \footnotesize
@@
   OM::util::random::gauss(double, double)+0xb4]\_\_Normal\\
-  A2 & [forward()+0x204; 
-  OM::Simulator::start(scnXml::Monitoring const\&)+0x468;
+  % A2 & [forward()+0x204; 
+  % OM::Simulator::start(scnXml::Monitoring const\&)+0x468;
 
-   OM::Population::update1(OM::SimTime)+0x1e9;
+  %  OM::Population::update1(OM::SimTime)+0x1e9;
 
-   OM::Population::newHuman(OM::SimTime)+0x5c;
+  %  OM::Population::newHuman(OM::SimTime)+0x5c;
 
-   OM::Host::Human::Human(OM::SimTime)+0x12b;
+  %  OM::Host::Human::Human(OM::SimTime)+0x12b;
 
-   OM::WithinHost::WHInterface::createWithinHostModel(double)+0x99;
+  %  OM::WithinHost::WHInterface::createWithinHostModel(double)+0x99;
 
-   OM::WithinHost::DescriptiveWithinHostModel::DescriptiveWithinHostModel(double)+0x3a;
+  %  OM::WithinHost::DescriptiveWithinHostModel::DescriptiveWithinHostModel(double)+0x3a;
 
-   OM::WithinHost::WHFalciparum::WHFalciparum(double)+0xe6;
+  %  OM::WithinHost::WHFalciparum::WHFalciparum(double)+0xe6;
 
-   OM::util::random::gauss(double, double)+0xb4]\_\_Normal \\
-  A3 & [forward()+0x204; OM::Simulator::start(scnXml::Monitoring const\&)+0x468;
+  %  OM::util::random::gauss(double, double)+0xb4]\_\_Normal \\
+  % A3 & [forward()+0x204; OM::Simulator::start(scnXml::Monitoring const\&)+0x468;
 
-   OM::Population::update1(OM::SimTime)+0xff;
+  %  OM::Population::update1(OM::SimTime)+0xff;
 
-   OM::Host::Human::update(bool)+0x22c;
+  %  OM::Host::Human::update(bool)+0x22c;
 
-   OM::Host::InfectionIncidenceModel::numNewInfections(OM::Host::Human const\&, double)+0x2a3;
+  %  OM::Host::InfectionIncidenceModel::numNewInfections(OM::Host::Human const\&, double)+0x2a3;
 
-   OM::util::random::poisson(double)+0x13f]\_\_Poisson\\
-  A4 & [forward()+0x204; OM::Simulator::start(scnXml::Monitoring const\&)+0x468;
+  %  OM::util::random::poisson(double)+0x13f]\_\_Poisson\\
+  % A4 & [forward()+0x204; OM::Simulator::start(scnXml::Monitoring const\&)+0x468;
 
-  OM::Population::update1(OM::SimTime)+0xff; OM::Host::Human::update(bool)+0x2bc;
+  % OM::Population::update1(OM::SimTime)+0xff; OM::Host::Human::update(bool)+0x2bc;
 
-   OM::Clinical::ClinicalModel::update(OM::Host::Human\&, double, bool)+0x96;
+  %  OM::Clinical::ClinicalModel::update(OM::Host::Human\&, double, bool)+0x96;
 
-   OM::Host::NeonatalMortality::eventNeonatalMortality()+0x9;
+  %  OM::Host::NeonatalMortality::eventNeonatalMortality()+0x9;
 
-    OM::util::random::uniform\_01()+0xc0]\_\_Uniform\\
-  A5 & [forward()+0x204; OM::Simulator::start(scnXml::Monitoring const\&)+0x468;
+    % OM::util::random::uniform\_01()+0xc0]\_\_Uniform\\
+    $\vdots$ & $\vdots$ \\
+%   A5 & [forward()+0x204; OM::Simulator::start(scnXml::Monitoring const\&)+0x468;
 
-   OM::Population::update1(OM::SimTime)+0xff;
+%    OM::Population::update1(OM::SimTime)+0xff;
 
-    OM::Host::Human::update(bool)+0x2bc;
+%     OM::Host::Human::update(bool)+0x2bc;
 
-    OM::Clinical::ClinicalModel::update(OM::Host::Human\&, double, bool)+0x96;
+%     OM::Clinical::ClinicalModel::update(OM::Host::Human\&, double, bool)+0x96;
 
-    OM::Host::NeonatalMortality::eventNeonatalMortality()+0x9;
+%     OM::Host::NeonatalMortality::eventNeonatalMortality()+0x9;
 
-    OM::util::random::uniform\_01()+0xc0]\_\_Uniform\\
-  A6 & [forward()+0x204; OM::Simulator::start(scnXml::Monitoring const\&)+0x468;
+%     OM::util::random::uniform\_01()+0xc0]\_\_Uniform\\
+%   A6 & [forward()+0x204; OM::Simulator::start(scnXml::Monitoring const\&)+0x468;
 
-   OM::Population::update1(OM::SimTime)+0xff;
+%    OM::Population::update1(OM::SimTime)+0xff;
 
-   OM::Host::Human::update(bool)+0x280;
+%    OM::Host::Human::update(bool)+0x280;
 
-   OM::WithinHost::DescriptiveWithinHostModel::update(int, std::vector \textgreater{}\&, double, double)+0x3a2;
+%    OM::WithinHost::DescriptiveWithinHostModel::update(int, std::vector \textgreater{}\&, double, double)+0x3a2;
 
-   OM::WithinHost::DescriptiveInfection::determineDensities(double, double, double, double\&, double, double)+0x303;
+%    OM::WithinHost::DescriptiveInfection::determineDensities(double, double, double, double\&, double, double)+0x303;
 
-   OM::util::random::uniform\_01()+0xc0]\_\_Uniform\\
-   A7 & [forward()+0x204; OM::Simulator::start(scnXml::Monitoring const\&)+0x468;
-   ı
-	OM::Population::update1(OM::SimTime)+0xff;
+%    OM::util::random::uniform\_01()+0xc0]\_\_Uniform\\
+%    A7 & [forward()+0x204; OM::Simulator::start(scnXml::Monitoring const\&)+0x468;
+%    ı
+% 	OM::Population::update1(OM::SimTime)+0xff;
 
-	OM::Host::Human::update(bool)+0x2bc;
+% 	OM::Host::Human::update(bool)+0x2bc;
 
-	OM::Clinical::ClinicalModel::update(OM::Host::Human\&, double, bool)+0xe8;
+% 	OM::Clinical::ClinicalModel::update(OM::Host::Human\&, double, bool)+0xe8;
 
-	OM::Clinical::CM5DayCommon::doClinicalUpdate(OM::Host::Human\&, double)+0x67;
+% 	OM::Clinical::CM5DayCommon::doClinicalUpdate(OM::Host::Human\&, double)+0x67;
 
-	OM::WithinHost::WHFalciparum::determineMorbidity(OM::Host::Human\&, double, bool)+0x73;
-ı
-	OM::WithinHost::Pathogenesis::PathogenesisModel::determineState(OM::Host::Human\&, double, double, double, bool)+0x143;
+% 	OM::WithinHost::WHFalciparum::determineMorbidity(OM::Host::Human\&, double, bool)+0x73;
+% ı
+% 	OM::WithinHost::Pathogenesis::PathogenesisModel::determineState(OM::Host::Human\&, double, double, double, bool)+0x143;
 
-	OM::util::random::bernoulli(double)+0x47;
+% 	OM::util::random::bernoulli(double)+0x47;
 
-	OM::util::random::uniform\_01()+0xc0]\_\_Uniform\\
-	A8 & [forward()+0x204; OM::Simulator::start(scnXml::Monitoring const\&)+0x468;
+% 	OM::util::random::uniform\_01()+0xc0]\_\_Uniform\\
+% 	A8 & [forward()+0x204; OM::Simulator::start(scnXml::Monitoring const\&)+0x468;
 
-	OM::Population::update1(OM::SimTime)+0xff; OM::Host::Human::update(bool)+0x2bc;
+% 	OM::Population::update1(OM::SimTime)+0xff; OM::Host::Human::update(bool)+0x2bc;
 
-	OM::Clinical::ClinicalModel::update(OM::Host::Human\&, double, bool)+0xe8;
+% 	OM::Clinical::ClinicalModel::update(OM::Host::Human\&, double, bool)+0xe8;
 
-	OM::Clinical::CM5DayCommon::doClinicalUpdate(OM::Host::Human\&, double)+0x67;
+% 	OM::Clinical::CM5DayCommon::doClinicalUpdate(OM::Host::Human\&, double)+0x67;
 
-	OM::WithinHost::WHFalciparum::determineMorbidity(OM::Host::Human\&, double, bool)+0x73;
+% 	OM::WithinHost::WHFalciparum::determineMorbidity(OM::Host::Human\&, double, bool)+0x73;
 
-	OM::WithinHost::Pathogenesis::PathogenesisModel::determineState(OM::Host::Human\&, double, double, double, bool)+0x162;
+% 	OM::WithinHost::Pathogenesis::PathogenesisModel::determineState(OM::Host::Human\&, double, double, double, bool)+0x162;
 
-	OM::util::random::bernoulli(double)+0x47;
+	% OM::util::random::bernoulli(double)+0x47;
 
-	OM::util::random::uniform\_01()+0xc0]\_\_Uniform\\
-A9 & [forward()+0x204; OM::Simulator::start(scnXml::Monitoring const\&)+0x468;
- OM::Population::update1(OM::SimTime)+0xff; OM::Host::Human::update(bool)+0x2bc;
+% 	OM::util::random::uniform\_01()+0xc0]\_\_Uniform\\
+% A9 & [forward()+0x204; OM::Simulator::start(scnXml::Monitoring const\&)+0x468;
+%  OM::Population::update1(OM::SimTime)+0xff; OM::Host::Human::update(bool)+0x2bc;
 
- OM::Clinical::ClinicalModel::update(OM::Host::Human\&, double, bool)+0xe8;
+%  OM::Clinical::ClinicalModel::update(OM::Host::Human\&, double, bool)+0xe8;
 
-  OM::Clinical::CM5DayCommon::doClinicalUpdate(OM::Host::Human\&, double)+0x67;
+%   OM::Clinical::CM5DayCommon::doClinicalUpdate(OM::Host::Human\&, double)+0x67;
 
-  OM::WithinHost::WHFalciparum::determineMorbidity(OM::Host::Human\&, double, bool)+0x73;
+%   OM::WithinHost::WHFalciparum::determineMorbidity(OM::Host::Human\&, double, bool)+0x73;
 
-  OM::WithinHost::Pathogenesis::PathogenesisModel::determineState(OM::Host::Human\&, double, double, double, bool)+0x1c7;
+%   OM::WithinHost::Pathogenesis::PathogenesisModel::determineState(OM::Host::Human\&, double, double, double, bool)+0x1c7;
 
-  OM::util::random::bernoulli(double)+0x47; OM::util::random::uniform\_01()+0xc0]\_\_Uniform\\
-\bottomrule
-  \end{tabularx}
-  \end{table*}
-% \subsubsection*{Acknowledgments}
-
-% Use unnumbered third level headings for the acknowledgments. All acknowledgments
-% go at the end of the paper. Do not include acknowledgments in the anonymized
-% submission, only in the final paper.
-  \begin{table*}[ht!]
-  \footnotesize
-  \setlength{\tabcolsep}{1mm}
-  \caption{Addresses .}
-  \label{table:addresses}
-  \def\arraystretch{1.25}
-  % \setlength{\tabcolsep}{1mm}
-  \begin{tabularx}{\textwidth}{@{}lX@{}} 
-    \toprule
-    Address ID & Full address \\
-    \midrule
-A10 & [forward()+0x204; OM::Simulator::start(scnXml::Monitoring const\&)+0x468;
-
-OM::Population::update1(OM::SimTime)+0xff; OM::Host::Human::update(bool)+0x2bc;
-
-OM::Clinical::ClinicalModel::update(OM::Host::Human\&, double, bool)+0xe8;
-
-OM::Clinical::CM5DayCommon::doClinicalUpdate(OM::Host::Human\&, double)+0xe3;
-
-OM::Clinical::ImmediateOutcomes::uncomplicatedEvent(OM::Host::Human\&, OM::Clinical::Episode::State)+0x87;
-
-OM::util::random::uniform\_01()+0xc0]\_\_Uniform\\
+%   OM::util::random::bernoulli(double)+0x47; OM::util::random::uniform\_01()+0xc0]\_\_Uniform\\
 A11 & [forward()+0x204; OM::Simulator::start(scnXml::Monitoring const\&)+0x468;
@@
   \end{table*}
-\section*{References}
+% \subsubsection*{Acknowledgments}
+
+% Use unnumbered third level headings for the acknowledgments. All acknowledgments
+% go at the end of the paper. Do not include acknowledgments in the anonymized
+% submission, only in the final paper.
+  % \begin{table*}[ht!]
+%   \footnotesize
+%   \setlength{\tabcolsep}{1mm}
+%   \caption{Addresses .}
+%   \label{table:addresses}
+%   \def\arraystretch{1.25}
+%   % \setlength{\tabcolsep}{1mm}
+%   \begin{tabularx}{\textwidth}{@{}lX@{}} 
+%     \toprule
+%     Address ID & Full address \\
+%     \midrule
+% A10 & [forward()+0x204; OM::Simulator::start(scnXml::Monitoring const\&)+0x468;
+
+% OM::Population::update1(OM::SimTime)+0xff; OM::Host::Human::update(bool)+0x2bc;
+
+% OM::Clinical::ClinicalModel::update(OM::Host::Human\&, double, bool)+0xe8;
+
+% OM::Clinical::CM5DayCommon::doClinicalUpdate(OM::Host::Human\&, double)+0xe3;
+
+% OM::Clinical::ImmediateOutcomes::uncomplicatedEvent(OM::Host::Human\&, OM::Clinical::Episode::State)+0x87;
+
+% OM::util::random::uniform\_01()+0xc0]\_\_Uniform\\
+% A11 & [forward()+0x204; OM::Simulator::start(scnXml::Monitoring const\&)+0x468;
+% OM::Population::update1(OM::SimTime)+0xff; OM::Host::Human::update(bool)+0x2bc;
+
+% OM::Clinical::ClinicalModel::update(OM::Host::Human\&, double, bool)+0xe8;
+
+% OM::Clinical::CM5DayCommon::doClinicalUpdate(OM::Host::Human\&, double)+0xad;
+
+% OM::Clinical::CM5DayCommon::severeMalaria(OM::Host::Human\&, OM::Clinical::Episode::State, double, int\&)+0x4b1;
+% OM::util::random::uniform\_01()+0xc0]\_\_Uniform\\
+%     \bottomrule
+%   \end{tabularx}
+%   \end{table*}
+% \section*{References}
+\newpage\phantom{blabla}\newpage\phantom{blabla}
 

### 2019-04-24 paper/icml_ai4good/main.tex
intro is finished, adding references now

--- 2019-04-24 paper/icml_ai4good/main.tex
@@
 
-Recent work by \citep{Baydin} introduces a fundamental paradigm change that, for the first time, allows to tackle this inverse problem in a way that .
+Recent work by \citep{Baydin} introduces a fundamental paradigm change that allows to tackle the inverse problem directly.
+Using a novel software tool named \textit{pyprob}, \citep{Baydin} extract program execution traces from state-of-the-art particle decay simulation engine SHERPA\cite{}.
+Using latest techniques from universal probabilistic programming inference, the sampled execution traces are then used in order to characterise the posterior distribution of the simulator output.
+From these, regions in simulator input space can be associated with specified outcomes.
+
+SHERPA, however, is different from leading epidemiology simulators in that it does not simulate population dynamics, but instead models individual decays at a time.
+In this paper, we for the first time demonstrate the applicability of \textit{pyprob}, and hence probabilistic programming inference, to population-based epidemiology simulators.
+In particular, we discuss how the leading malaria epidemiology simulators EMOD\cite{} and OpenMalaria\cite{} can be interfaced with \textit{pyprob}.
+We argue that our work is likely to have significant social impact on many livelihoods affected by Malaria and discuss opportunities for future research.
 
@@
 
-In this paper, we for the first time demonstrate the applicability of trace introspection, and hence universal probabilistic inference, to population-based simulators.
-In particular, we discuss how the leading malaria epidemiology simulators EMOD[] and OpenMalaria[] can be interfaced with <TODO>.
-We argue that our work is likely to have significant social impact on many livelihoods affected by Malaria and discuss opportunities for future research.
-
-- issues of public health are complex and
+%- issues of public health are complex and
 

### 2019-04-24 paper/icml_ai4good/main.tex

--- 2019-04-24 paper/icml_ai4good/main.tex
@@
 The population dynamics of groups of typically several thousand hosts are then initialised by sampling from the initial distributions, using initial random seeds.
-Forward simulation then proceeds by rolling out simulation dynamics over time.
-Policymakers then take interest in comparing simulation outcomes of cherry-picked scenarios and interventions with those of suitably chosen control groups.
+Forward simulation then proceeds by rolling out population dynamics over time.
+Simulation outcomes of cherry-picked scenarios and interventions of interest are then compared to suitably designed control runs.
 
 Stochastic epidemiology simulators have already proven to be an invaluable tool to policymakers with policy implications that are often counterintuitive yet far-ranging.
-For example, it has been shown that mass vaccination may be largely ineffective in regions of large transmission rates, but may play a crucial role in areas of low transmission.
+For example, it has been shown that mass vaccination may be largely ineffective in regions of large transmission rates, but may play a crucial role in areas of low transmittivity.
 
-However valuable forward simulations of epidemiology scenarios are, the holy grail of goal-driven policymaking lies ultimately in the inverse problem: Given a desired outcome, we wish to find the optimal respective intervention policy given a specific regional context.
+However valuable forward simulations of epidemiology scenarios are, the holy grail of goal-driven policymaking ultimately lies in solving the inverse problem: Given a desired outcome and regional context, identify the series of interventions that will best reach the desired target after a given amount of time.
 Due to computational constraints, even an approximate search over the whole simulator input space usually proves intractable.
+An experience-guided search over the simulation output of a small set of manually-designed scenarios of interest therefore long seemed the only feasible way of investigating implications of policy interventions.
+
+Recent work by \citep{Baydin} introduces a fundamental paradigm change that allows to tackle the inverse problem directly.
+Using a novel software tool named \textit{pyprob}, \citep{Baydin} extract program execution traces from state-of-the-art particle decay simulation engine SHERPA\cite{}.
+Using latest techniques from universal probabilistic programming inference, the sampled execution traces are then used in order to characterise the posterior distribution of the simulator output.
+From these, regions in simulator input space can be associated with specified outcomes.
+
+SHERPA, however, is different from leading epidemiology simulators in that it does not simulate population dynamics, but instead models individual decays at a time.
+In this paper, we for the first time demonstrate the applicability of \textit{pyprob}, and hence probabilistic programming inference, to population-based epidemiology simulators.
+In particular, we discuss how the leading malaria epidemiology simulators EMOD\cite{} and OpenMalaria\cite{} can be interfaced with \textit{pyprob}.
+We argue that our work is likely to have significant social impact on many livelihoods affected by Malaria and discuss opportunities for future research.
 
@@
 
-<Section about prob prog inference, with reference to SHERPA>
+% <Section about prob prog inference, with reference to SHERPA>
 
-In this paper, we for the first time demonstrate the applicability of trace introspection, and hence universal probabilistic inference, to population-based simulators.
-In particular, we discuss how the leading malaria epidemiology simulators EMOD[] and OpenMalaria[] can be interfaced with <TODO>.
-We argue that our work is likely to have significant social impact on many livelihoods affected by Malaria and discuss opportunities for future research.
-
-- issues of public health are complex and
+%- issues of public health are complex and
 

### 2019-04-24 paper/icml_ai4good/main.tex
intro milestone 1 - focussing on getting EMOD to work now

--- 2019-04-24 paper/icml_ai4good/main.tex
@@
 \section{Introduction}
-Ending the epidemics of AIDS, tuberculosis, malaria and other infectious diseases by 2030 is a key target within the Good Health \& Well-Being section of the UN Sustainable Development Goals\cite{}.
+Ending the epidemics of AIDS, tuberculosis, malaria and other infectious diseases by 2030 is a key target within the Good Health \& Well-Being section of the UN Sustainable Development Goals \cite{refugees_2030_nodate}\cite{un_sustainable_2018}.
 
 In contrast to AIDS, Malaria can be routinely cured thanks to the disposability of a variety of medical interventions.
-Issues are further aggravated due to the increasing emergence of parasite resistance to multiple drugs, including artemisinin, or even insecticides.
-Hence, despite decades of substantial international efforts, initial hopes of regional elimination or even global eradication have not been met.
-So difficult is Malaria control that, even today, about a quarter of a million children under the age of 5 succumb to it every year in Africa alone.
+However, both strands of parasite and associated mosquito host are ever-evolving and transmission vectors are hard to control.
+These circumstances are further aggravated due to the increasing emergence of parasite resistance to multiple drugs, including artemisinin, and even mosquito resistance to insecticides.
+In addition, many of the countries most affected by Malaria are less developed countries, in which logistic and issues of socio-economic stability often impede effective access to treatment.
+Hence, despite decades of substantial international efforts, initial hopes of regional elimination or even global eradication have largely not been met.
+So difficult is Malaria control that, even today, about a quarter of a million children under the age of 5 succumb to it every year in Africa alone\cite{un_sustainable_2018}.
 
-To reach the WHO’s target of reducing malaria incidence and mortality rates by at least 90\% by 2030, policymakers increasingly turn to evidence-based methods.
+To reach the WHO’s target of reducing malaria incidence and mortality rates by at least 90\% by 2030, policymakers increasingly turn to evidence-based methods\cite{who_global_2015}.
 However, the collection of large quantities of field data is often infeasible due to logistic, economic or political constraints, or outright impossible if novel methods are to be assessed a priori.
-Hence in silico simulation of malaria epidemiology plays an increasingly crucial role in evidence-based policymaking.
+Hence in silico simulation of malaria epidemiology plays an increasingly crucial role in evidence-based policymaking \cite{mwendera_challenges_2019}.
 %Hence epidemiology simulators fine-tuned on real data play a crucial role in determining policy implications.
@@
 However, comprehensive modeling of all of these components in a regional context remains challenging, and is greatly impeded by the lack of available field data for regional calibration.
-Simulational complexity is further aggravated by considering feedback effects from a large arsenal of anti-malaria interventions, ranging from distribution of mosquito nets, indoor residual spraying and mass drug administration to age-specific vaccinations.
+Simulational complexity is further aggravated by considering feedback effects from a large arsenal of anti-malaria interventions, ranging from distribution of mosquito nets, indoor residual spraying and mass drug administration to age-specific vaccinations \cite{smith_ensemble_2012}.
 
@@
 Stochastic epidemiology simulators have already proven to be an invaluable tool to policymakers with policy implications that are often counterintuitive yet far-ranging.
-For example, it has been shown that mass vaccination may be largely ineffective in regions of large transmission rates, but may play a crucial role in areas of low transmittivity.
+For example, it has been shown that mass vaccination may be largely ineffective in regions of large transmission rates, but may play a crucial role in areas of low transmittivity \cite{smith_ensemble_2012}.
 
@@
 
-Recent work by \citep{Baydin} introduces a fundamental paradigm change that allows to tackle the inverse problem directly.
-Using a novel software tool named \textit{pyprob}, \citep{Baydin} extract program execution traces from state-of-the-art particle decay simulation engine SHERPA\cite{}.
+Recent work by \citep{gunes_baydin_efficient_2018} introduces a fundamental paradigm change that allows to tackle the inverse problem directly.
+Using a novel software tool named \textit{pyprob}, \citep{gunes_baydin_efficient_2018} extract program execution traces from state-of-the-art particle decay simulation engine SHERPA \cite{gleisberg_event_2008}.
 Using latest techniques from universal probabilistic programming inference, the sampled execution traces are then used in order to characterise the posterior distribution of the simulator output.
@@
 In this paper, we for the first time demonstrate the applicability of \textit{pyprob}, and hence probabilistic programming inference, to population-based epidemiology simulators.
-In particular, we discuss how the leading malaria epidemiology simulators EMOD\cite{} and OpenMalaria\cite{} can be interfaced with \textit{pyprob}.
+In particular, we discuss how the leading open-source malaria epidemiology simulators EMOD \cite{bershteyn_implementation_2018} and OpenMalaria \cite{smith_towards_2008}\cite{chitnis_periodically-forced_2012} can be interfaced with \textit{pyprob}.
 We argue that our work is likely to have significant social impact on many livelihoods affected by Malaria and discuss opportunities for future research.
@@
 
-\section*{References}
+%\section*{References}
 
 \bibliographystyle{icml2019}
-\bibliography{refs}
+\bibliography{malaria,refs}
 

### 2019-04-24 paper/icml_ai4good/main.tex
updated

--- 2019-04-24 paper/icml_ai4good/main.tex
@@
 % Therefore, a short form for the running title is supplied here:
-\icmltitlerunning{Submission and Formatting Instructions for ICML 2019}
+\icmltitlerunning{Understanding the Black-box Structure of Epidemiology Simulators for Policy Decision Making}
 
@@
 \twocolumn[
-\icmltitle{Submission and Formatting Instructions for \\
-           International Conference on Machine Learning (ICML 2019)}
+\icmltitle{Understanding the Black-box Structure of Epidemiology Simulators\\
+			 for Policy Decision Making}
 
@@
 probabilistic programming systems (PPSs) and epidemiology simulators, to extract
-fully interpretable posterior structures, enabling policy makers 
+fully interpretable structures, enabling policy makers 
 and practitioners to make interpretable inferences. 
-In particular, we demonstrate this for the Malaria disease and show how we can 
-perform interpretable inference in such settings.
+In particular, we demonstrate this for the Malaria disease in two commonly used simulators; EMOD and OpenMalaria.
 \end{abstract}
@@
 
-In contrast to AIDS, Malaria can be routinely cured thanks to the disposability of a variety of medical interventions. 
-Issues are further aggravated due to the increasing emergence of parasite resistance to multiple drugs, including artemisinin, or even insecticides. 
-Hence, despite decades of substantial international efforts, initial hopes of regional elimination or even global eradication have not been met.
- So difficult is Malaria control that, even today, about a quarter of a million children under the age of 5 succumb to it every year in Africa alone.
+%maybe cut some of this paragraph out
+% In contrast to AIDS, Malaria can be routinely cured thanks to the disposability of a variety of medical interventions. 
+% Issues are further aggravated due to the increasing emergence of parasite resistance to multiple drugs, including artemisinin, or even insecticides. 
+% Hence, despite decades of substantial international efforts, initial hopes of regional elimination or even global eradication have not been met.
+%  So difficult is Malaria control that, even today, about a quarter of a million children under the age of 5 succumb to it every year in Africa alone.
 
@@
 %Hence epidemiology simulators fine-tuned on real data play a crucial role in determining policy implications.
-
 Even in the absence of political, economic and social constraints, Malaria epidemiology is governed by a complex set of drivers, few of which can be understood in isolation.
@@
 Some progress in modeling specific components has been made, such as the parasite life-cycle, transmission vector dynamics and the inter-dependencies between infection rates and vicinity to open water.
-However, comprehensive modeling of all of these components in a regional context remains challenging, and is greatly impeded by the lack of available field data for regional calibration.
-Simulation complexity is further aggravated by considering feedback effects from a large arsenal of anti-malaria interventions, ranging from distribution of mosquito nets, indoor residual spraying and mass drug administration to age-specific vaccinations.
+However, comprehensive modeling of all of these components in a regional context remains challenging, and is greatly impeded by the lack of available field data for regional calibration, hence the need for comprehensive simulators.
+% Simulation complexity is further aggravated by considering feedback effects from a large arsenal of anti-malaria interventions, ranging from distribution of mosquito nets, indoor residual spraying and mass drug administration to age-specific vaccinations.
+Stochastic epidemiology simulators have already proven to be an invaluable tool to policymakers with policy implications that are often counterintuitive yet far-ranging. 
+For example, it has been shown that mass vaccination may be largely ineffective in regions of large transmission rates, but may play a crucial role in areas of low transmittivity.
 
-All of the above limitations require contemporary malaria epidemiology simulators to employ stochastic modeling techniques, which negotiate data scarcity and model uncertainty through the use of appropriate prior distributions that can be tuned even from small samples of field data.
+However, since many of these simulators are stochastic in nature and therefore act as black-boxes, understanding the structure of these complex simulators is challenging and poses several challenges for policy makers.
+Therefore, sophisticated tools are required that enable us to not only digest the results of the simulation, but provide pathways to perform complex inferences. 
+To this end, we extend the work of \citep{baydin2018efficient} to enabling any epidemiology simulator to be transformed into probabilistic model that can be used as input to a probabilistic programming system enabling trace paths to be extracted from the simulator on both a per human level and a population level, from which we can extract trace probabilities allowing policy makers to easily see what actions had the biggest impact on the final output of the simulator.
+% maybe talk about this framework enables a pathway to inference etc..
+
+% All of the above limitations require contemporary malaria epidemiology simulators to employ stochastic modeling techniques, which negotiate data scarcity and model uncertainty through the use of appropriate prior distributions that can be tuned even from small samples of field data.
 The population dynamics of groups of typically several thousand hosts are then initialised by sampling from the initial distributions, using initial random seeds.
@@
 
-Stochastic epidemiology simulators have already proven to be an invaluable tool to policymakers with policy implications that are often counterintuitive yet far-ranging.
-For example, it has been shown that mass vaccination may be largely ineffective in regions of large transmission rates, but may play a crucial role in areas of low transmittivity.
-
-However valuable forward simulations of epidemiology scenarios are, the holy grail of goal-driven policymaking ultimately lies in solving the inverse problem: Given a desired outcome and regional context, identify the series of interventions that will best reach the desired target after a given amount of time.
+% This is great, if we had inference implemented right now.
+% However valuable forward simulations of epidemiology scenarios are, the holy grail of goal-driven policymaking ultimately lies in solving the inverse problem: Given a desired outcome and regional context, identify the series of interventions that will best reach the desired target after a given amount of time.
 Due to computational constraints, even an approximate search over the whole simulator input space usually proves intractable.
 An experience-guided search over the simulation output of a small set of manually-designed scenarios of interest therefore long seemed the only feasible way of investigating implications of policy interventions.
-
-Recent work by \citep{Baydin} introduces a fundamental paradigm change that allows to tackle the inverse problem directly.
-Using a novel software tool named \textit{pyprob}, \citep{Baydin} extract program execution traces from state-of-the-art particle decay simulation engine SHERPA\cite{}.
-Using latest techniques from universal probabilistic programming inference, the sampled execution traces are then used in order to characterise the posterior distribution of the simulator output.
-From these, regions in simulator input space can be associated with specified outcomes.
-
-SHERPA, however, is different from leading epidemiology simulators in that it does not simulate population dynamics, but instead models individual decays at a time.
-In this paper, we for the first time demonstrate the applicability of \textit{pyprob}, and hence probabilistic programming inference, to population-based epidemiology simulators.
-In particular, we discuss how the leading malaria epidemiology simulators EMOD\cite{} and OpenMalaria\cite{} can be interfaced with \textit{pyprob}.
-We argue that our work is likely to have significant social impact on many livelihoods affected by Malaria and discuss opportunities for future research.
 
@@
 \label{sec:background}
-This will kind of be like a background section.
 
-\section{Methodology}
+Recent work by \citep{Baydin} introduces a fundamental paradigm change that allows to tackle the inverse problem directly. 
+Using a novel software tool named \textit{pyprob}, \citep{Baydin} extract program execution traces from state-of-the-art particle decay simulation engine SHERPA\cite{}.
+ Using latest techniques from universal probabilistic programming inference, the sampled execution traces are then used in order to characterise the posterior distribution of the simulator output. 
+ From these, regions in simulator input space can be associated with specified outcomes.
+
+SHERPA, however, is different from leading epidemiology simulators in that it does not simulate population dynamics, but instead models individual decays at a time.
+ In this paper, we for the first time demonstrate the applicability of \textit{pyprob}, and hence probabilistic programming inference, to population-based epidemiology simulators. 
+ In particular, we discuss how the leading malaria epidemiology simulators EMOD\cite{} and OpenMalaria\cite{} can be interfaced with \textit{pyprob}.
+  We argue that our work is likely to have significant social impact on many livelihoods affected by Malaria and discuss opportunities for future research.
 
@@
 % \section*{References}
-\newpage\phantom{blabla}\newpage\phantom{blabla}
+% \newpage\phantom{blabla}\newpage\phantom{blabla}
 

### 2019-04-24 paper/icml_ai4good/main.tex
updated

@@
 \usepackage{tabularx}
+
 % hyperref makes hyperlinks in the resulting PDF.
@@
 
-% The \icmltitle you define below is probably too long as a header.
+% The \icmltitle you define below is probably too long as aer.
 % Therefore, a short form for the running title is supplied here:
@@
 \section{Introduction}
+Ending the epidemics of AIDS, tuberculosis, malaria and other infectious diseases by 2030 is a key target within the Good Health \& Well-Being section of the UN Sustainable Development Goals \cite{refugees_2030_nodate}\cite{un_sustainable_2018}.
 
@@
 Hence in silico simulation of malaria epidemiology plays an increasingly crucial role in evidence-based policymaking.
+
 %Hence epidemiology simulators fine-tuned on real data play a crucial role in determining policy implications.
@@
 Therefore, sophisticated tools are required that enable us to not only digest the results of the simulation, but provide pathways to perform complex inferences. 
-To this end, we extend the work of \citep{baydin2018efficient} to enabling any epidemiology simulator to be transformed into probabilistic model that can be used as input to a probabilistic programming system enabling trace paths to be extracted from the simulator on both a per human level and a population level, from which we can extract trace probabilities allowing policy makers to easily see what actions had the biggest impact on the final output of the simulator.
+To this end, we extend the work of \citep{baydin2018efficient} to enable multiple stochastic epidemiology simulators to be transformed into a probabilistic model that can be used as input to a probabilistic programming system, enabling trace paths to be extracted from the simulator on both a per human level and a population level, from which we can extract trace probabilities allowing policy makers to determine what actions have the biggest impact on the final output of the simulator and thus can shape policy decisions accordingly.
+ We argue that our work is likely to have significant social impact on many livelihoods affected by Malaria and discuss opportunities for future research.
+
+% Recent work by \citep{Baydin} introduces a fundamental paradigm change that allows to tackle the inverse problem directly. 
+% Using a novel software tool named \textit{pyprob}, \citep{Baydin} extract program execution traces from state-of-the-art particle decay simulation engine SHERPA\cite{}.
+%  Using latest techniques from universal probabilistic programming inference, the sampled execution traces are then used in order to characterise the posterior distribution of the simulator output. 
+%  From these, regions in simulator input space can be associated with specified outcomes.
+
+% SHERPA, however, is different from leading epidemiology simulators in that it does not simulate population dynamics, but instead models individual decays with different sampling strategies.
+Thus, the existing work of \cite{baydin2018efficient} must be extended to encapsulate these different strategies.
+%  In this paper, we for the first time demonstrate the applicability of \textit{pyprob}, and hence probabilistic programming inference, to population-based epidemiology simulators. 
+%  In particular, we discuss how the leading malaria epidemiology simulators EMOD\cite{} and OpenMalaria\cite{} can be interfaced with \textit{pyprob}.
+%   We argue that our work is likely to have significant social impact on many livelihoods affected by Malaria and discuss opportunities for future research.
+
 % maybe talk about this framework enables a pathway to inference etc..
@@
 
-Recent work by \citep{Baydin} introduces a fundamental paradigm change that allows to tackle the inverse problem directly. 
-Using a novel software tool named \textit{pyprob}, \citep{Baydin} extract program execution traces from state-of-the-art particle decay simulation engine SHERPA\cite{}.
- Using latest techniques from universal probabilistic programming inference, the sampled execution traces are then used in order to characterise the posterior distribution of the simulator output. 
- From these, regions in simulator input space can be associated with specified outcomes.
-
-SHERPA, however, is different from leading epidemiology simulators in that it does not simulate population dynamics, but instead models individual decays at a time.
- In this paper, we for the first time demonstrate the applicability of \textit{pyprob}, and hence probabilistic programming inference, to population-based epidemiology simulators. 
- In particular, we discuss how the leading malaria epidemiology simulators EMOD\cite{} and OpenMalaria\cite{} can be interfaced with \textit{pyprob}.
-  We argue that our work is likely to have significant social impact on many livelihoods affected by Malaria and discuss opportunities for future research.
-
 \subsection{Probabilistic Programming}
@@
 However, since the development happens over several decades, usually without comprehensive documentation, understanding a code base written with legacy libraries and software is incredibly difficult and an almost impossible task for a non-expert, or expert who has not had access to the previous years of development.
-Our method of overriding simulators builds on the work of~\cite{baydin2018efficient} and extends the framework to encapsulate a more diverse range of simulators. 
-Our framework provides a simple solution and makes it easy to transform arbitrary stochastic simulators into probabilistic programs, regardless of the complexity of the simulator and the language that the simulator is written in\footnote{The framework currently supports 9 popular languages, but there is nothing to stop this being extended to any language of interest.}
-In addition to this, we can understand the structure of programs in simulators for both decoding \emph{black-box structures} and for posterior inference. 
-This enables two things. 1) It enables software developers and researchers to understand complicated code bases. 
-2) It provides interpretable inference results that provide policy makers with predictions that are truly interpretable, in the sense that the end-user of the inference results understands what physical events led to inference outcome.
-This is critical in several domains, and is particularly important in the medical domain. 
-We provide examples of former in next Section~\textbf{TO ADD REF} and discuss the latter in ~\bg{Link to conclusion / future work}.
+Our method of hijacking simulators builds on the work of~\cite{baydin2018efficient} and extends the framework to encapsulate a more diverse range of simulators.
+Our framework provides a simple solution and makes it easy to transform arbitrary stochastic simulators into probabilistic programs, regardless of the complexity of the simulator and the language that the simulator is written in\footnote{The framework currently supports 9 popular languages, but there is nothing to stop this being extended to any language of interest.}.
+In addition to this, we can understand the structure of programs in simulators for both decoding \emph{black-box structures} and for posterior inference. This enables two things. 
+1) It enables software developers and policy making to understand complicated code bases and dissect outputs of simulators.
+ 2) It provides interpretable inference results that provide policy makers with predictions that are truly interpretable, in the sense that the end-user of the inference results understands what physical events led to the inference outcome.
+ This is critical in several domains, especially in epidemiology. 
+ We provide examples of former in next Section~\ref{sec:casestudy} and discuss the latter in ~\bg{Link to conclusion / future work}.
+
+ \subsubsection*{Epidemiology Simulators}
+ \paragraph*{OpenMalaria} Write a small paragraph on openmalaria, 
+ What is openmalaria? Open Malaria is an open-source simulator funded by the Gates Foundation to model different  aspects of the malaria prevention and treatment cycle for varying population dynamics.
+ Models in OpenMalaria are specified via a scenario file, in the form of an XML document, which specifies the specifics of a particular treatment, or prevention strategy, climatic factors, the demographics of the population and so forth.
+In addition to this, field observations can be encoded inside of, or externally to the scenario file.
+The scenario file encodes all input and output parameters, some of which we wish to simulate to try and infer, others remain fixed.
+
+ \paragraph{EMOD} Do the same for EMOD as OM
 
 \section{Case Study}
+\label{sec:casestudy}
 \subsection{Connecting a Epidemiology Simulators to a Probabilistic Programming System}
 
-In order to overriding the stochastic simulator we are only required to override the location of the stochastic primitives, i.e.
+In order to hijack the stochastic simulator we are only required to override the location of the stochastic primitives, i.e.
 the operations that generate the stochasticity and randomness within the simulator.
@@
 This greatly improves the scope of applicability of our system, as this environment only has to be built once, which then enables the simulator to be run simultaneously on machines of different architectures machine.
-The second step requires one to override the stochastic primitives in the language, for example, the \emph{Uniform} and \emph{Normal} draws. 
-A stochastic simulator may have a variety of other distribution and sampling primitives, but the underlying methodology is the same.
+The second step requires one to override the stochastic primitives, which requires re-writing aspects of the sampling procedures inside the simulator. 
+This requires zero domain-expertise and only requires one to find where the sampling calls are originated, this can be done with any basic editor search function.
+ Such information is typically saved in $\texttt{RANDOM.<filetype>}$.  We now demonstrate this for a normal primitive. 
+\bg{Add code snippet here showing original code then modified code}
+The third step is creating the protocols, which enable communication between the simulator and probabilistic programming system. 
+For this we extend the existing bringing framework introduced in~\cite{baydin2018efficient}, which is comprised of the Probabilistic Programming eXecution~(PPX) protocols and a light-weight c++ interface, however this light-weight interface is dependent on whatever the file type of the $\texttt{RANDOM.<filetype>}$ file is.
+In doing this we enable a larger class of sampling procedures to be overridden.
+The PPX protocols in particular take advantage of flat buffer protocols \bg{add cite to google fb}, which require a flatbuffers script to be specified, see here for an example of a flat buffers script \bg{reference PPX github page}.
+Once this is specified the flat buffers script is compiled and automatically constructs the message passing framework between the simulator and the probabilistic programming system.
+The next steps requires modifying the $\texttt{main}$ function in $\texttt{main.cpp}$, \bg{add more on this later}
+The final step is to create a light-weight interface.
+As both the EMOD and OpenMalaria are solely based on c++ we can extend an existing interface called pyprob\_cpp\footnote{Add github link here} to account for the additional sampling procedures.
+Once these steps are complete, the containerized environment will now contain everything required to pass models through to the simulators, encode the simulator into a probabilistic program, which can then be evaluated in the probabilistic programming system.
+We provide examples of this output in Figure~\ref{fig:plotewan}.
 
 % \subsection{Extending Probabilistic Programming Protocols}
+
 \begin{figure*}[h!]
@@
 	\includegraphics[width=\textwidth]{../plots/ewan_25_pop_10.pdf}
-   \bigbreak 
+	\centering
 	\includegraphics[width=\textwidth]{../plots/ewan_25_pop_100.pdf}
-   % \label{fig:Ng2}
-
-\caption{In overriding the stochasticity in the simulator we are able to control all random variables in the program,
-	 from which we can understand the physical events that unfold. }
+	\caption{Add something. Population 100}
+	\label{fig:plotewan}
 \end{figure*}
-
 TODO
@@
 % Please add the following required packages to your document preamble:
-% 
-% \begin{table*}[b!]
-% \footnotesize
-% \setlength{\tabcolsep}{1mm}
-% \caption{Addresses}
-% \label{table:addresses}
-% \def\arraystretch{1.25}
-% % \setlength{\tabcolsep}{1mm}
-% \begin{tabularx}{\textwidth}{@{}lX@{}} 
-% \toprule
-% Address & Physical event                                                                                                                            \midrule
-%
-% 
-% 
-% \bottomrule                                                                                                                                   
-% \end{tabularx}
-% \end{table*}
-
+% \usepackage{graphicx}
   \begin{table*}[h!]
@@
 A11 & [forward()+0x204; OM::Simulator::start(scnXml::Monitoring const\&)+0x468;
+
 OM::Population::update1(OM::SimTime)+0xff; OM::Host::Human::update(bool)+0x2bc;
@@
   \end{table*}
-% \subsubsection*{Acknowledgments}
-
-% Use unnumbered third level headings for the acknowledgments. All acknowledgments
-% go at the end of the paper. Do not include acknowledgments in the anonymized
-% submission, only in the final paper.
-  % \begin{table*}[ht!]
-%   \footnotesize
-%   \setlength{\tabcolsep}{1mm}
-%   \caption{Addresses .}
-%   \label{table:addresses}
-%   \def\arraystretch{1.25}
-%   % \setlength{\tabcolsep}{1mm}
-%   \begin{tabularx}{\textwidth}{@{}lX@{}} 
-%     \toprule
-%     Address ID & Full address \\
-%     \midrule
-% A10 & [forward()+0x204; OM::Simulator::start(scnXml::Monitoring const\&)+0x468;
-
-% OM::Population::update1(OM::SimTime)+0xff; OM::Host::Human::update(bool)+0x2bc;
-
-% OM::Clinical::ClinicalModel::update(OM::Host::Human\&, double, bool)+0xe8;
-
-% OM::Clinical::CM5DayCommon::doClinicalUpdate(OM::Host::Human\&, double)+0xe3;
-
-% OM::Clinical::ImmediateOutcomes::uncomplicatedEvent(OM::Host::Human\&, OM::Clinical::Episode::State)+0x87;
-
-% OM::util::random::uniform\_01()+0xc0]\_\_Uniform\\
-% A11 & [forward()+0x204; OM::Simulator::start(scnXml::Monitoring const\&)+0x468;
-% OM::Population::update1(OM::SimTime)+0xff; OM::Host::Human::update(bool)+0x2bc;
-
-% OM::Clinical::ClinicalModel::update(OM::Host::Human\&, double, bool)+0xe8;
-
-% OM::Clinical::CM5DayCommon::doClinicalUpdate(OM::Host::Human\&, double)+0xad;
-
-% OM::Clinical::CM5DayCommon::severeMalaria(OM::Host::Human\&, OM::Clinical::Episode::State, double, int\&)+0x4b1;
-% OM::util::random::uniform\_01()+0xc0]\_\_Uniform\\
-%     \bottomrule
-%   \end{tabularx}
-%   \end{table*}
-% \section*{References}
-% \newpage\phantom{blabla}\newpage\phantom{blabla}
 
 \bibliographystyle{icml2019}
-\bibliography{refs}
+\bibliography{malaria,refs}
 

### 2019-04-25 paper/icml_ai4good/main.tex
sort out table

@@
   A1 & [forward()+0x204; OM::Simulator::start(scnXml::Monitoring const)+0x28a;
-
   OM::Population::createInitialHumans()+0x94;
-
   OM::Population::newHuman(OM::SimTime)+0x5c;
@@
   OM::util::random::gauss(double, double)+0xb4]\_\_Normal\\
-  % A2 & [forward()+0x204; 
-  % OM::Simulator::start(scnXml::Monitoring const\&)+0x468;
 
-  %  OM::Population::update1(OM::SimTime)+0x1e9;
+   $\vdots$ & $\vdots$ \\
+  A5 & [forward()+0x204; OM::Simulator::start(scnXml::Monitoring const\&)+0x468;
 
-  %  OM::Population::newHuman(OM::SimTime)+0x5c;
+   OM::Population::update1(OM::SimTime)+0xff;
 
-  %  OM::Host::Human::Human(OM::SimTime)+0x12b;
+    OM::Host::Human::update(bool)+0x2bc;
 
-  %  OM::WithinHost::WHInterface::createWithinHostModel(double)+0x99;
+    OM::Clinical::ClinicalModel::update(OM::Host::Human\&, double, bool)+0x96;
 
-  %  OM::WithinHost::DescriptiveWithinHostModel::DescriptiveWithinHostModel(double)+0x3a;
+    OM::Host::NeonatalMortality::eventNeonatalMortality()+0x9;
 
-  %  OM::WithinHost::WHFalciparum::WHFalciparum(double)+0xe6;
+    OM::util::random::uniform\_01()+0xc0]\_\_Uniform\\
 
-  %  OM::util::random::gauss(double, double)+0xb4]\_\_Normal \\
-  % A3 & [forward()+0x204; OM::Simulator::start(scnXml::Monitoring const\&)+0x468;
-
-  %  OM::Population::update1(OM::SimTime)+0xff;
-
-  %  OM::Host::Human::update(bool)+0x22c;
-
-  %  OM::Host::InfectionIncidenceModel::numNewInfections(OM::Host::Human const\&, double)+0x2a3;
-
-  %  OM::util::random::poisson(double)+0x13f]\_\_Poisson\\
-  % A4 & [forward()+0x204; OM::Simulator::start(scnXml::Monitoring const\&)+0x468;
-
-  % OM::Population::update1(OM::SimTime)+0xff; OM::Host::Human::update(bool)+0x2bc;
-
-  %  OM::Clinical::ClinicalModel::update(OM::Host::Human\&, double, bool)+0x96;
-
-  %  OM::Host::NeonatalMortality::eventNeonatalMortality()+0x9;
-
-    % OM::util::random::uniform\_01()+0xc0]\_\_Uniform\\
-    $\vdots$ & $\vdots$ \\
-%   A5 & [forward()+0x204; OM::Simulator::start(scnXml::Monitoring const\&)+0x468;
-
-%    OM::Population::update1(OM::SimTime)+0xff;
-
-%     OM::Host::Human::update(bool)+0x2bc;
-
-%     OM::Clinical::ClinicalModel::update(OM::Host::Human\&, double, bool)+0x96;
-
-%     OM::Host::NeonatalMortality::eventNeonatalMortality()+0x9;
-
-%     OM::util::random::uniform\_01()+0xc0]\_\_Uniform\\
 %   A6 & [forward()+0x204; OM::Simulator::start(scnXml::Monitoring const\&)+0x468;
@@
 %   OM::util::random::bernoulli(double)+0x47; OM::util::random::uniform\_01()+0xc0]\_\_Uniform\\
-A11 & [forward()+0x204; OM::Simulator::start(scnXml::Monitoring const\&)+0x468;
+ $\vdots$ & $\vdots$ \\
 
-OM::Population::update1(OM::SimTime)+0xff; OM::Host::Human::update(bool)+0x2bc;
+% A11 & [forward()+0x204; OM::Simulator::start(scnXml::Monitoring const\&)+0x468;
 
-OM::Clinical::ClinicalModel::update(OM::Host::Human\&, double, bool)+0xe8;
+% OM::Population::update1(OM::SimTime)+0xff; OM::Host::Human::update(bool)+0x2bc;
 
-OM::Clinical::CM5DayCommon::doClinicalUpdate(OM::Host::Human\&, double)+0xad;
+% OM::Clinical::ClinicalModel::update(OM::Host::Human\&, double, bool)+0xe8;
 
-OM::Clinical::CM5DayCommon::severeMalaria(OM::Host::Human\&, OM::Clinical::Episode::State, double, int\&)+0x4b1;
-OM::util::random::uniform\_01()+0xc0]\_\_Uniform\\
+% OM::Clinical::CM5DayCommon::doClinicalUpdate(OM::Host::Human\&, double)+0xad;
+
+% OM::Clinical::CM5DayCommon::severeMalaria(OM::Host::Human\&, OM::Clinical::Episode::State, double, int\&)+0x4b1;
+
+% OM::util::random::uniform\_01()+0xc0]\_\_Uniform\\
+
 \bottomrule

### 2019-04-25 paper/icml_ai4good/main.tex
sort out table
minor
mimor

@@
 \usepackage{tabularx}
-
+\usepackage{listings}
 % hyperref makes hyperlinks in the resulting PDF.
@@
 
+\lstset{language=C++,
+                basicstyle=\footnotesize,
+                keywordstyle=\color{blue}\ttfamily,
+                stringstyle=\color{red}\ttfamily,
+                commentstyle=\color{green}\ttfamily,
+                morecomment=[l][\color{magenta}]{\#},
+                breaklines = true
+}
 \begin{document}
@@
 This requires zero domain-expertise and only requires one to find where the sampling calls are originated, this can be done with any basic editor search function.
- Such information is typically saved in $\texttt{RANDOM.<filetype>}$.  We now demonstrate this for a normal primitive. 
-\bg{Add code snippet here showing original code then modified code}
+ Such information is typically saved in $\texttt{RANDOM.<filetype>}$.  We now demonstrate this for a uniform primitive.
+
+\begin{tabular}{|p{3.6cm}|p{3.6cm}|}
+Before  &  After  \\
+\begin{lstlisting}
+#include <gsl/gsl_cdf.h>
+#include <gsl/gsl_rng.h>
+#include <gsl/gsl_randist.h>
+//preamble
+double random::uniform_01 () {
+    double result =
+# ifdef 
+OM_RANDOM_USE_BOOST
+       rng_uniform01 ();
+# else
+ gsl_rng_uniform (rng.gsl_generator);
+# endif
+    return result;
+\end{lstlisting}&
+\begin{lstlisting}
+#include <pyprob_cpp.h>
+#include "xtensor/xadapt.hpp"
+// preamble
+double random::uniform_01 () {
+auto uniform = pyprob_cpp::distributions::Uniform(0,1);
+return pyprob_cpp::sample(uniform)(0);
+}
+\end{lstlisting}
+\end{tabular}
+
 The third step is creating the protocols, which enable communication between the simulator and probabilistic programming system. 
@@
 In doing this we enable a larger class of sampling procedures to be overridden.
-The PPX protocols in particular take advantage of flat buffer protocols \bg{add cite to google fb}, which require a flatbuffers script to be specified, see here for an example of a flat buffers script \bg{reference PPX github page}.
+The PPX protocols in particular take advantage of flat buffer protocols\footnote{https://github.com/google/flatbuffers}, which require a flatbuffers script to be specified, see here for an example of a flat buffers script\footnote{https://github.com/probprog/ppx/blob/master/ppx.fbs}.
 Once this is specified the flat buffers script is compiled and automatically constructs the message passing framework between the simulator and the probabilistic programming system.
@@
 The final step is to create a light-weight interface.
-As both the EMOD and OpenMalaria are solely based on c++ we can extend an existing interface called pyprob\_cpp\footnote{Add github link here} to account for the additional sampling procedures.
+As both the EMOD and OpenMalaria are solely based on c++ we can extend an existing interface called pyprob\_cpp\footnote{https://github.com/probprog/pyprob_cpp} to account for the additional sampling procedures.
 Once these steps are complete, the containerized environment will now contain everything required to pass models through to the simulators, encode the simulator into a probabilistic program, which can then be evaluated in the probabilistic programming system.
@@
 \end{figure*}
-TODO
 
-% Please add the following required packages to your document preamble:
-% \usepackage{booktabs}
-% Please add the following required packages to your document preamble:
-% \usepackage{booktabs}
-% \usepackage{graphicx}
-% Please add the following required packages to your document preamble:
-% \usepackage{graphicx}
  \begin{table*}[h!]
@@
   \setlength{\tabcolsep}{1mm}
-  \caption{Addresses }
+  \caption{Addresses TODO }
   \label{table:addresses}
@@
   A1 & [forward()+0x204; OM::Simulator::start(scnXml::Monitoring const)+0x28a;
+
   OM::Population::createInitialHumans()+0x94;
+
   OM::Population::newHuman(OM::SimTime)+0x5c;
@@
 
-%   A6 & [forward()+0x204; OM::Simulator::start(scnXml::Monitoring const\&)+0x468;
-
-%    OM::Population::update1(OM::SimTime)+0xff;
-
-%    OM::Host::Human::update(bool)+0x280;
-
-%    OM::WithinHost::DescriptiveWithinHostModel::update(int, std::vector \textgreater{}\&, double, double)+0x3a2;
-
-%    OM::WithinHost::DescriptiveInfection::determineDensities(double, double, double, double\&, double, double)+0x303;
-
-%    OM::util::random::uniform\_01()+0xc0]\_\_Uniform\\
-%    A7 & [forward()+0x204; OM::Simulator::start(scnXml::Monitoring const\&)+0x468;
-%    ı
-% 	OM::Population::update1(OM::SimTime)+0xff;
-
-% 	OM::Host::Human::update(bool)+0x2bc;
-
-% 	OM::Clinical::ClinicalModel::update(OM::Host::Human\&, double, bool)+0xe8;
-
-% 	OM::Clinical::CM5DayCommon::doClinicalUpdate(OM::Host::Human\&, double)+0x67;
-
-% 	OM::WithinHost::WHFalciparum::determineMorbidity(OM::Host::Human\&, double, bool)+0x73;
-% ı
-% 	OM::WithinHost::Pathogenesis::PathogenesisModel::determineState(OM::Host::Human\&, double, double, double, bool)+0x143;
-
-% 	OM::util::random::bernoulli(double)+0x47;
-
-% 	OM::util::random::uniform\_01()+0xc0]\_\_Uniform\\
-% 	A8 & [forward()+0x204; OM::Simulator::start(scnXml::Monitoring const\&)+0x468;
-
-% 	OM::Population::update1(OM::SimTime)+0xff; OM::Host::Human::update(bool)+0x2bc;
-
-% 	OM::Clinical::ClinicalModel::update(OM::Host::Human\&, double, bool)+0xe8;
-
-% 	OM::Clinical::CM5DayCommon::doClinicalUpdate(OM::Host::Human\&, double)+0x67;
-
-% 	OM::WithinHost::WHFalciparum::determineMorbidity(OM::Host::Human\&, double, bool)+0x73;
-
-% 	OM::WithinHost::Pathogenesis::PathogenesisModel::determineState(OM::Host::Human\&, double, double, double, bool)+0x162;
-
-	% OM::util::random::bernoulli(double)+0x47;
-
-% 	OM::util::random::uniform\_01()+0xc0]\_\_Uniform\\
-% A9 & [forward()+0x204; OM::Simulator::start(scnXml::Monitoring const\&)+0x468;
-%  OM::Population::update1(OM::SimTime)+0xff; OM::Host::Human::update(bool)+0x2bc;
-
-%  OM::Clinical::ClinicalModel::update(OM::Host::Human\&, double, bool)+0xe8;
-
-%   OM::Clinical::CM5DayCommon::doClinicalUpdate(OM::Host::Human\&, double)+0x67;
-
-%   OM::WithinHost::WHFalciparum::determineMorbidity(OM::Host::Human\&, double, bool)+0x73;
-
-%   OM::WithinHost::Pathogenesis::PathogenesisModel::determineState(OM::Host::Human\&, double, double, double, bool)+0x1c7;
-
-%   OM::util::random::bernoulli(double)+0x47; OM::util::random::uniform\_01()+0xc0]\_\_Uniform\\
  $\vdots$ & $\vdots$ \\
-
-% A11 & [forward()+0x204; OM::Simulator::start(scnXml::Monitoring const\&)+0x468;
-
-% OM::Population::update1(OM::SimTime)+0xff; OM::Host::Human::update(bool)+0x2bc;
-
-% OM::Clinical::ClinicalModel::update(OM::Host::Human\&, double, bool)+0xe8;
-
-% OM::Clinical::CM5DayCommon::doClinicalUpdate(OM::Host::Human\&, double)+0xad;
-
-% OM::Clinical::CM5DayCommon::severeMalaria(OM::Host::Human\&, OM::Clinical::Episode::State, double, int\&)+0x4b1;
-
-% OM::util::random::uniform\_01()+0xc0]\_\_Uniform\\
 

### 2019-04-25 paper/icml_ai4good/main.tex
updated- need to solve the table issue

@@
 \usepackage{listings}
+\usepackage{pgfplots}
+\usepackage{tikz}
+
 % hyperref makes hyperlinks in the resulting PDF.
@@
 %\usepackage[accepted]{icml2019}
-
+\usetikzlibrary{arrows,chains,matrix,positioning,scopes}
+%
+\makeatletter
+\tikzset{join/.code=\tikzset{after node path={%
+\ifx\tikzchainprevious\pgfutil@empty\else(\tikzchainprevious)%
+edge[every join]#1(\tikzchaincurrent)\fi}}}
+\makeatother
+%
+\tikzset{>=stealth',every on chain/.append style={join},
+         every join/.style={->}}
+\tikzstyle{labeled}=[execute at begin node=$\scriptstyle,
+   execute at end node=$]
 % The \icmltitle you define below is probably too long as aer.
@@
 }
+\pgfplotsset{width=9.5cm,compat=1.5}
 \begin{document}
@@
 
-\begin{tabular}{|p{3.6cm}|p{3.6cm}|}
+\begin{tabular}{p{3.6cm}|p{3.6cm}}
 Before  &  After  \\
+\midrule
 \begin{lstlisting}
@@
     return result;
+    }
 \end{lstlisting}&
@@
 In doing this we enable a larger class of sampling procedures to be overridden.
-The PPX protocols in particular take advantage of flat buffer protocols\footnote{https://github.com/google/flatbuffers}, which require a flatbuffers script to be specified, see here for an example of a flat buffers script\footnote{https://github.com/probprog/ppx/blob/master/ppx.fbs}.
+The PPX protocols in particular take advantage of flat buffer protocols\footnote{\url{https://github.com/google/flatbuffers}}, which require a flatbuffers script to be specified, see here for an example of a flat buffers script\footnote{https://github.com/probprog/ppx/blob/master/ppx.fbs}.
 Once this is specified the flat buffers script is compiled and automatically constructs the message passing framework between the simulator and the probabilistic programming system.
@@
 The final step is to create a light-weight interface.
-As both the EMOD and OpenMalaria are solely based on c++ we can extend an existing interface called pyprob\_cpp\footnote{https://github.com/probprog/pyprob_cpp} to account for the additional sampling procedures.
+As both the EMOD and OpenMalaria are solely based on c++ we can extend an existing interface called pyprob\_cpp\footnote{\url{https://github.com/probprog/pyprob_cpp}} to account for the additional sampling procedures.
 Once these steps are complete, the containerized environment will now contain everything required to pass models through to the simulators, encode the simulator into a probabilistic program, which can then be evaluated in the probabilistic programming system.
@@
 % \subsection{Extending Probabilistic Programming Protocols}
+% \begin{tikzpicture}
+% \begin{axis}[
+%     ybar,
+%     enlargelimits=0.15,
+%     legend style={at={(0.5,-0.15)},
+%       anchor=north,legend columns=-1},
+%       xlabel = {Address},
+%     % ylabel={Probabilities},
+%     symbolic x coords={A1,A2,A3,A4,A5,A6,A7,A8,A9,A10,A11},
+%     xtick=data,
+%     nodes near coords,
+%     nodes near coords align={vertical},
+%     ]
+% % \addplot coordinates {(tool8,7) (tool9,9) (tool10,4)};
+% \addplot coordinates {(A1,0.2) (A2,0.5) (A3,0.7) (A4,0.3) (A5,0.4) (A6,0.4) (A7,0.6) (A8,0.7) (A9,0.9) (A10,0.12) (A11,0.23)};
+% % \addplot coordinates {(A1,0.2) (A2,0.5) (A3,0.7) (A4,0.9)};
+% % \addplot coordinates {(A1,0.1) (A2,0.1) (A3,0.1) (A4,0.3)};
+% % \legend{used,understood,not understood}
+% \end{axis}
+% \caption{Trace probabilities}
+% \end{tikzpicture}
+\begin{tikzpicture}
+  \matrix (m) [matrix of math nodes, row sep=3em, column sep=3em]
+    { Start & A1  & A2  & A3  & A4 \\
+      Start & A1 & A2 & A3 & A4 \\ };
+  { [start chain] \chainin (m-1-1);
+    \chainin (m-1-2);
+    { [start branch=A] \chainin (m-2-2)
+        [join={node[right,labeled] {0.90}}];}
+    \chainin (m-1-3) [join={node[above,labeled] {\varphi}}];
+    { [start branch=B] \chainin (m-2-3)
+        [join={node[right,labeled] {\eta_2}}];}
+    \chainin (m-1-4) [join={node[above,labeled] {\psi}}];
+    { [start branch=C] \chainin (m-2-4)
+        [join={node[right,labeled] {\eta_3}}];}
+    \chainin (m-1-5); }
+  { [start chain] \chainin (m-2-1);
+    \chainin (m-2-2);
+    \chainin (m-2-3) [join={node[above,labeled] {\varphi'}}];
+    \chainin (m-2-4) [join={node[above,labeled] {\psi'}}];
+    \chainin (m-2-5); }
+\end{tikzpicture}
 

### 2019-04-25 paper/icml_ai4good/main.tex
updated

--- 2019-04-25 paper/icml_ai4good/main.tex
@@
 fully interpretable structures, enabling policy makers 
-and practitioners to make interpretable inferences. 
+and practitioners to make decisions from simulator outputs that can be fully resoned on.
+ % interpretable inferences. 
 In particular, we demonstrate this for the Malaria disease in two commonly used simulators; EMOD and OpenMalaria.
@@
 Ending the epidemics of AIDS, tuberculosis, malaria and other infectious diseases by 2030 is a key target within the Good Health \& Well-Being section of the UN Sustainable Development Goals \cite{refugees_2030_nodate}\cite{un_sustainable_2018}.
-
-Ending the epidemics of AIDS, tuberculosis, malaria and other infectious diseases by 2030 is a key target by the Good Health \& Well-Being section of the UN Sustainable Development Goals. 
-Despite decades of substantial international efforts, malaria still annually kills about a quarter of a million children under the age of 5 in Africa alone.
-Ending the epidemics of AIDS, tuberculosis, malaria and other infectious diseases by 2030 is a key target within the Good Health \& Well-Being section of the UN Sustainable Development Goals\cite{}.
+However, despite decades of substantial international efforts, these diseases kill hundreads of million people a year.
+In particular, malaria still annually kills about a quarter of a million children under the age of 5 in Africa alone.
 
@@
 Stochastic epidemiology simulators have already proven to be an invaluable tool to policymakers with policy implications that are often counterintuitive yet far-ranging. 
-For example, it has been shown that mass vaccination may be largely ineffective in regions of large transmission rates, but may play a crucial role in areas of low transmittivity.
+For example, it has been shown that mass vaccination may be largely ineffective in regions of large transmission rates, but may play a crucial role in areas of low transmittivity. \bg{citation?}
 
-However, since many of these simulators are stochastic in nature and therefore act as black-boxes, understanding the structure of these complex simulators is challenging and poses several challenges for policy makers.
+However, since many of these simulators are stochastic in nature and therefore act as black-boxes, understanding the structure of these complex simulators is challenging and poses several challenges for policy makers who want to interpret the decisions being inferred.
 Therefore, sophisticated tools are required that enable us to not only digest the results of the simulation, but provide pathways to perform complex inferences. 
-To this end, we extend the work of \citep{baydin2018efficient} to enable multiple stochastic epidemiology simulators to be transformed into a probabilistic model that can be used as input to a probabilistic programming system, enabling trace paths to be extracted from the simulator on both a per human level and a population level, from which we can extract trace probabilities allowing policy makers to determine what actions have the biggest impact on the final output of the simulator and thus can shape policy decisions accordingly.
+To this end, we extend the work of \citep{baydin2018efficient} to enable multiple stochastic epidemiology simulators to be transformed into a probabilistic model that can be used as input to a probabilistic programming system, enabling trace paths to be extracted from the simulator on both a per human level and a population level, from which we can extract trace probabilities allowing policy makers to determine what actions have the biggest impact on differing populations dynamics and thus shape policy decisions accordingly.
  We argue that our work is likely to have significant social impact on many livelihoods affected by Malaria and discuss opportunities for future research.
@@
 It attempts to unify the three fields by creating automated frameworks in which one can easily extrapolate their observed data and perform sophisticated statistical analysis with advance inference procedures such as
-Markov Chain Monte Carlo (MCMC), Variational Inference~(VI) and Deep Neural Networks~(DNN) in an accessible way, i.e. without the need of expert knowledge in implementing and using such procedures.
-Users can then exploit these algorithms on an array of models
-within the probabilistic programming language~(PPL), using the language itself to create models that represent complicated distribution objects. 
-PPLs typically extend existing programming languages such as python, c++, go and so forth, which means that users do not have to learn
-a completely new set of operating semantics.  
-Our PPL PyProb~\cite{le-2016-inference,baydin2018efficient} belongs to the family of universal PPLs 
+Markov Chain Monte Carlo (MCMC), Variational Inference~(VI) and Deep Neural Networks~(DNN) in an accessible way, i.e.
+without the need of expert knowledge in implementing and using such procedures.
+In addition to this, probabilistic programming systems~(PPSs) provide tools that enable the source code of a program to be throughly analysed in an automated fashion.
+Users can then exploit these algorithms and tools on an array of models
+within the PPS, using the language itself to create models that represent complicated distribution objects which are then analysed by the PPS. 
+PPSs are extensions, or reductions, of existing programming languages such as python, c++, go and so forth, which means that users do not have to learn
+a completely new set of operating semantics and can write programs in the same vain as they are accustomed to.  
+The PPS that we use, PyProb~\cite{le-2016-inference,baydin2018efficient}, belongs to the family of universal PPSs 
 that allow the expression of unrestricted probability models in a Turing complete fashion~\cite{wood2014new,goodman2012church,landau_binder_2014,siddharth2017learning,bingham2019pyro}.
@@
 Simulators encode several years, if not decades, of research and development and so by their very nature are structurally complicated. 
-Within the context of epidemiology simulators we can model treatment, infection and prevention cycle under a variety of different environmental conditions, enabling predictions to be made given a series of field observations \bg{Make sure that I do not state they can already do what PP can do, i.e.
+Within the context of epidemiology simulators we can model the treatment, infection and prevention cycle under a variety of different environmental conditions, enabling predictions to be made on a set of field observations \bg{Make sure that I do not state they can already do what PP can do, i.e.
 the ability to update and infer the input parameters continually conditioned on new observations.}.
-However, since the development happens over several decades, usually without comprehensive documentation, understanding a code base written with legacy libraries and software is incredibly difficult and an almost impossible task for a non-expert, or expert who has not had access to the previous years of development.
-Our method of hijacking simulators builds on the work of~\cite{baydin2018efficient} and extends the framework to encapsulate a more diverse range of simulators.
-Our framework provides a simple solution and makes it easy to transform arbitrary stochastic simulators into probabilistic programs, regardless of the complexity of the simulator and the language that the simulator is written in\footnote{The framework currently supports 9 popular languages, but there is nothing to stop this being extended to any language of interest.}.
-In addition to this, we can understand the structure of programs in simulators for both decoding \emph{black-box structures} and for posterior inference. This enables two things. 
-1) It enables software developers and policy making to understand complicated code bases and dissect outputs of simulators.
+However, since the development happens over several years, or decades, usually without comprehensive documentation, understanding a code base written with legacy libraries and software is incredibly difficult and an almost impossible task for a non-expert, or expert who has not had access to the previous years of development.
+Our method of overriding simulators builds on the work of~\cite{baydin2018efficient} and extends the framework to encapsulate a more diverse range of simulators, in particular population based simulators such as those found in epidemiology.
+Our framework provides a simple solution to do this and makes it easy to transform arbitrary stochastic simulators into probabilistic programs, regardless of the complexity of the simulator and the language that the simulator is written in.
+In particular, the framework currently supports 13 popular languages, but there is nothing to stop this being extended to any language of interest. 
+In addition to this, we can understand the structure of simulators as programs for both decoding \emph{black-box structures} and for posterior inference. This enables two things. 
+1) It enables software developers and policy makers to understand complicated code bases and dissect outputs of simulators for guiding policy decision making, as we shall see later in Section~\ref{sec:casestudy}.
  2) It provides interpretable inference results that provide policy makers with predictions that are truly interpretable, in the sense that the end-user of the inference results understands what physical events led to the inference outcome.
  This is critical in several domains, especially in epidemiology. 
- We provide examples of former in next Section~\ref{sec:casestudy} and discuss the latter in ~\bg{Link to conclusion / future work}.
+ We provide examples of former in next Section~\ref{sec:casestudy} and discuss the latter in future work as extensions to the the current framework.
 
  \subsubsection*{Epidemiology Simulators}
- \paragraph*{OpenMalaria} Write a small paragraph on openmalaria, 
- What is openmalaria? Open Malaria is an open-source simulator funded by the Gates Foundation to model different  aspects of the malaria prevention and treatment cycle for varying population dynamics.
+ \paragraph*{OpenMalaria} 
+Open Malaria is an open-source simulator funded by the Gates Foundation to model different  aspects of the malaria prevention and treatment cycle for varying population dynamics.
  Models in OpenMalaria are specified via a scenario file, in the form of an XML document, which specifies the specifics of a particular treatment, or prevention strategy, climatic factors, the demographics of the population and so forth.
@@
 
+In both these well used simulators users are accustomed to writing their models as either XML, or JSON files and so it is important that any system built around this framework enables the models to be inputted in the same format, without requiring the users to learn how to use a new set of protocols.
+
 \section{Case Study}
@@
 
-In order to hijack the stochastic simulator we are only required to override the location of the stochastic primitives, i.e.
-the operations that generate the stochasticity and randomness within the simulator.
-We will walk through how this is done for EMOD and OpenMalaria, you will see that it is procedurally identical.
-However, given the sheer number of simulators it maybe slightly different procedurally. 
-To our knowledge it shouldn't be, but we have not been able view of all stochastic simulators we cannot make this guarantee with certainty. 
-The first step is to build a containerized environment, such as Docker~\cite{merkel2014docker} and Singularity~\cite{kurtzer2017singularity}, which enables the simulators to be run on any device, independent of different hardware architectures, removing the requirement for hardware specific machines.
-This greatly improves the scope of applicability of our system, as this environment only has to be built once, which then enables the simulator to be run simultaneously on machines of different architectures machine.
+In order to overhaul the stochastic simulators we are only required to override the location of the stochastic primitives, i.e.
+the base operations that generate the stochasticity and randomness within the simulator.
+We will walk through how this is done for a standard epidemiology simulator.
+
+The initial step first requires one to build a containerized environment, such as Docker~\cite{merkel2014docker} and Singularity~\cite{kurtzer2017singularity}, which enables the simulators to be run on any device, independent of different hardware architectures, removing the requirement for hardware specific machines.
+This greatly improves the scope of applicability of our system and the simulator, as this environment only has to be built once, which then enables the simulator to be run simultaneously on machines of different architectures machine.
+This step will be slightly different for individual simulators, but the overall idea is the same.
+We cannot provide a link to an example of a containerized environment for openmalaria and system in this submission due to anonymity\footbote{We shall provide a link in the camera ready submission}.
 The second step requires one to override the stochastic primitives, which requires re-writing aspects of the sampling procedures inside the simulator. 
 This requires zero domain-expertise and only requires one to find where the sampling calls are originated, this can be done with any basic editor search function.
- Such information is typically saved in $\texttt{RANDOM.<filetype>}$.  We now demonstrate this for a uniform primitive.
+ Such information is typically saved in $\texttt{RANDOM.<filetype>}$, although it may vary slightly between simulators. We now demonstrate this for a uniform primitive.
 
-\begin{tabular}{|p{3.6cm}|p{3.6cm}|}
+\begin{tabular}{p{3.6cm}|p{3.6cm}}
+\footnotesize
 Before  &  After  \\
+\midrule
 \begin{lstlisting}
@@
 
-The third step is creating the protocols, which enable communication between the simulator and probabilistic programming system. 
-For this we extend the existing bringing framework introduced in~\cite{baydin2018efficient}, which is comprised of the Probabilistic Programming eXecution~(PPX) protocols and a light-weight c++ interface, however this light-weight interface is dependent on whatever the file type of the $\texttt{RANDOM.<filetype>}$ file is.
-In doing this we enable a larger class of sampling procedures to be overridden.
-The PPX protocols in particular take advantage of flat buffer protocols\footnote{https://github.com/google/flatbuffers}, which require a flatbuffers script to be specified, see here for an example of a flat buffers script\footnote{https://github.com/probprog/ppx/blob/master/ppx.fbs}.
+As can be seen, writing the new sampling call is cleaner and syntactically simpler. 
+The third step is generating the protocols, which enable communication between the simulator and probabilistic programming system. 
+For this we extend the existing bridging framework introduced in~\cite{baydin2018efficient}, which is comprised of the Probabilistic Programming eXecution~(PPX) protocols and a light-weight C ++ interface, however the language of this light-weight interface is dependent on whatever the file type of the $\texttt{RANDOM.<filetype>}$ file is.
+For both OpenMalaria and EMOD, the file is written in C ++.
+In using these protocols a larger class of sampling procedures can be overridden.
+The PPX protocols work by making use of the flatbuffer protocols\footnote{\url{https://github.com/google/flatbuffers}}, which require a flatbuffers script to be specified, see here for an example of a flatbuffers script\footnote{https://github.com/probprog/ppx/blob/master/ppx.fbs}.
 Once this is specified the flat buffers script is compiled and automatically constructs the message passing framework between the simulator and the probabilistic programming system.
-The next steps requires modifying the $\texttt{main}$ function in $\texttt{main.cpp}$, \bg{add more on this later}
+The flatbuffers script is simple, clean and intuitive, making it easy for a non-expert to produced.
+The next steps requires modifying the $\texttt{main}$ function in $\texttt{main.cpp}$ and defining an additional forward function which runs one execution of the simulator to enable message passing between the simulator and PPS\footnote{A link to an example will be provided after the review period to preserve anonymity}.
 The final step is to create a light-weight interface.
-As both the EMOD and OpenMalaria are solely based on c++ we can extend an existing interface called pyprob\_cpp\footnote{https://github.com/probprog/pyprob_cpp} to account for the additional sampling procedures.
-Once these steps are complete, the containerized environment will now contain everything required to pass models through to the simulators, encode the simulator into a probabilistic program, which can then be evaluated in the probabilistic programming system.
-We provide examples of this output in Figure~\ref{fig:plotewan}.
+As both the EMOD and OpenMalaria are solely based on C ++ we can extend an existing interface called pyprob\_cpp\footnote{\url{https://github.com/probprog/pyprob_cpp}} to account for the additional sampling procedures.
+Once these steps are complete, the containerized environment will now contain everything required to pass models through to the simulators, encode the simulator into a probabilistic program, which can then be evaluated in the PPS.
+We provide examples of the generated trace plots from the connection between the simulator and the PPS in Figure~\ref{fig:plotewan}.
 
-% \subsection{Extending Probabilistic Programming Protocols}
+\subsection{Extending Probabilistic Programming Protocols}
+\begin{tikzpicture}
+\begin{axis}[
+    ybar,
+    enlargelimits=0.15,
+    legend style={at={(0.5,-0.15)},
+      anchor=north,legend columns=-1},
+      xlabel = {Address},
+    ylabel={Probabilities},
+    symbolic x coords={Trace 1, Trace 2, Trace 3}},
+    xtick=data,
+    nodes near coords,
+    nodes near coords align={vertical},
+    ]
+% \addplot coordinates {(tool8,7) (tool9,9) (tool10,4)};
+\addplot coordinates {(Trace 1,0.2) (Trace 2,0.5) (Trace 3,0.23)};
+% \addplot coordinates {(A1,0.2) (A2,0.5) (A3,0.7) (A4,0.9)};
+% \addplot coordinates {(A1,0.1) (A2,0.1) (A3,0.1) (A4,0.3)};
+% \legend{used,understood,not understood}
+\end{axis}
+\caption{Trace probabilities. Trace 1 corresponds to A1,A1,A1,...,A4,... Trace 2 corresponds to A1,A1,A1,...A2,... These still need correctly calculating.}
+\end{tikzpicture}
 

### 2019-04-25 paper/icml_ai4good/main.tex
updated

@@
 }
-\pgfplotsset{width=9.5cm,compat=1.5}
+\pgfplotsset{width=7cm,compat=1.8}
 \begin{document}
@@
 This step will be slightly different for individual simulators, but the overall idea is the same.
-We cannot provide a link to an example of a containerized environment for openmalaria and system in this submission due to anonymity\footbote{We shall provide a link in the camera ready submission}.
+We cannot provide a link to an example of a containerized environment for OpenMalaria and system in this submission due to anonymity\footnote{We shall provide a link in the camera ready submission}.
 The second step requires one to override the stochastic primitives, which requires re-writing aspects of the sampling procedures inside the simulator. 
@@
 
+\begin{table}[h!]
+\footnotesize
 \begin{tabular}{p{3.6cm}|p{3.6cm}}
-\footnotesize
 Before  &  After  \\
@@
 \end{tabular} 
-
+\end{table}
 As can be seen, writing the new sampling call is cleaner and syntactically simpler. 
@@
 \subsection{Extending Probabilistic Programming Protocols}
+\begin{figure}
 \begin{tikzpicture}
@@
     ylabel={Probabilities},
-    symbolic x coords={Trace 1, Trace 2, Trace 3}},
+    symbolic x coords={Trace 1, Trace 2, Trace 3},
     xtick=data,
@@
 \end{axis}
-\caption{Trace probabilities. Trace 1 corresponds to A1,A1,A1,...,A4,... Trace 2 corresponds to A1,A1,A1,...A2,... These still need correctly calculating.}
 \end{tikzpicture}
-
+\caption{Trace probabilities. Trace 1 corresponds to A1,A1,A4,... 
+Trace 2 corresponds to A1,A1,A1,...A2,... These still need correctly calculating.}
+\end{figure}
 \begin{figure*}[h!]
@@
 
+\bg{Add comments about how this trace structure helps policy makers.}
 \bibliographystyle{icml2019}

### 2019-04-26 paper/icml_ai4good/main.tex
updated

@@
 
+% \usepgfplotslibrary{external}
 % Attempt to make hyperref and algorithmic work together better:
@@
 % Therefore, a short form for the running title is supplied here:
-\icmltitlerunning{Understanding the Black-box Structure of Epidemiology Simulators for Policy Decision Making}
+\icmltitlerunning{Hijacking Simulators for the Greater Good}
 
@@
 \twocolumn[
-\icmltitle{Understanding the Black-box Structure of Epidemiology Simulators\\
-			 for Policy Decision Making}
+\icmltitle{Hijacking Simulators for the Greater Good}
 
@@
 fully interpretable structures, enabling policy makers 
-and practitioners to make decisions from simulator outputs that can be fully resoned on.
+and practitioners to make decisions from simulator outputs that can be fully reasoned on.
  % interpretable inferences. 
@@
 \subsection{Extending Probabilistic Programming Protocols}
+
+\bg{Add performance plot showing memory consumptions of pyprob and our new system.}
+% \begin{tikzpicture}
+% \begin{axis}[xlabel={Population size},ylabel={memory consumption (MB)},
+% legend style={at={(0.5,-0.15)},xmin=0,xmax=700,ymin=0,ymax=70, axis lines=center]
+
+% \addplot[pyprob,domain=0:1000, color=blue,]{x};
+
+% \addplot[improved,domain=0:1000, color=red,]{10};
+
+% \legend{pyprob, improved}
+% \end{axis}
+% \end{tikzpicture}
+
 \begin{figure}

### 2019-04-26 paper/icml_ai4good/main.tex
starting to rework the introduction

@@
 % Therefore, a short form for the running title is supplied here:
-\icmltitlerunning{Hijacking Simulators for the Greater Good}
+%\icmltitlerunning{Hijacking Simulators for the Greater Good}
+\icmltitlerunning{Automatically Converting Epidemiology Simulations
+to Probabilistic Programs}
 
@@
 \twocolumn[
-\icmltitle{Hijacking Simulators for the Greater Good}
+%\icmltitle{Hijacking Simulators for the Greater Good}
+\icmltitle{Automatically Converting Epidemiology Simulations
+	to Probabilistic Programs}
 
@@
 \section{Introduction}
-Ending the epidemics of AIDS, tuberculosis, malaria and other infectious diseases by 2030 is a key target within the Good Health \& Well-Being section of the UN Sustainable Development Goals \cite{refugees_2030_nodate}\cite{un_sustainable_2018}.
+Ending the epidemics of AIDS, tuberculosis, malaria and other infectious diseases by 2030 is a key target within the Good Health \& Well-Being section of the UN Sustainable Development Goals \cite{refugees_2030_nodate,un_sustainable_2018}.
 However, despite decades of substantial international efforts, these diseases kill hundreads of million people a year.
-In particular, malaria still annually kills about a quarter of a million children under the age of 5 in Africa alone.
+For example, malaria still annually kills about a quarter of a million children under the age of 5 in Africa alone.
 
@@
 
-To reach the WHO’s target of reducing malaria incidence and mortality rates by at least 90\% by 2030, policymakers increasingly turn to evidence-based methods. 
-However, the collection of large quantities of field data is often infeasible due to logistic, economic or political constraints, or outright impossible if novel methods are to be assessed a priori. 
-Hence in silico simulation of malaria epidemiology plays an increasingly crucial role in evidence-based policymaking.
+To reach the WHO’s target of reducing malaria incidence and mortality rates by at least 90\% by 2030, policymakers increasingly turn to evidence-based methods and computational simulations CITATIONS. 
+These simulations allow policymakers to infer critical information like
+INSERT and make predictions about the impacts of policies, allowing
+more effective decisions to be made in combating the disease at a
+macro-level and ultimately saving lives.
+For example, it has been shown that mass vaccination may be largely ineffective in regions of large transmission rates, but may play a crucial role in areas of low transition. \bg{citation?}
 
-%Hence epidemiology simulators fine-tuned on real data play a crucial role in determining policy implications.
-Even in the absence of political, economic and social constraints, Malaria epidemiology is governed by a complex set of drivers, few of which can be understood in isolation.
+However, malaria epidemiology is governed by a complex set of drivers, 
+few of which can be understood in isolation.
 These include within-host dynamics, population-specific traits and even local geography.
-Some progress in modeling specific components has been made, such as the parasite life-cycle, transmission vector dynamics and the inter-dependencies between infection rates and vicinity to open water.
-However, comprehensive modeling of all of these components in a regional context remains challenging, and is greatly impeded by the lack of available field data for regional calibration, hence the need for comprehensive simulators.
+Comprehensive modeling of all of these components remains challenging, particular in a regional-specific context. Therefore simulators 
+need to be both comprehensive and powerful, while it can
+be highly non-trivial to both interpret their outputs and draw desired
+inferences when coupling them with observed data.
+
+%However, the collection of large quantities of field data is often infeasible due to logistic, economic or political constraints, or outright impossible if novel methods are to be assessed a priori. 
+%Hence in silico simulation of malaria epidemiology plays an increasingly crucial role in evidence-based policymaking.
+%Even in the absence of political, economic and social constraints, 
+%Malaria epidemiology is governed by a complex set of drivers, few of which can be understood in isolation.
+%These include within-host dynamics, population-specific traits and even local geography.
+%Some progress in modeling specific components has been made, such as the parasite life-cycle, transmission vector dynamics and the inter-dependencies between infection rates and vicinity to open water.
+%However, comprehensive modeling of all of these components in a regional context remains challenging, and is greatly impeded by the lack of available field data for regional calibration, hence the need for comprehensive simulators.
 % Simulation complexity is further aggravated by considering feedback effects from a large arsenal of anti-malaria interventions, ranging from distribution of mosquito nets, indoor residual spraying and mass drug administration to age-specific vaccinations.
-Stochastic epidemiology simulators have already proven to be an invaluable tool to policymakers with policy implications that are often counterintuitive yet far-ranging. 
-For example, it has been shown that mass vaccination may be largely ineffective in regions of large transmission rates, but may play a crucial role in areas of low transmittivity. \bg{citation?}
+%Stochastic epidemiology simulators have already proven to be an invaluable tool to policymakers with policy implications that are often counterintuitive yet far-ranging.
 

### 2019-04-26 paper/icml_ai4good/main.tex
finished new intro

@@
 \usepackage{tikz}
+\usepackage{todonotes}
 
@@
 %\icmltitlerunning{Hijacking Simulators for the Greater Good}
-\icmltitlerunning{Automatically Converting Epidemiology Simulations
-to Probabilistic Programs}
+\icmltitlerunning{Hijacking Epidemiology Simulators
+	with Probabilistic Programming}
 
@@
 %\icmltitle{Hijacking Simulators for the Greater Good}
-\icmltitle{Automatically Converting Epidemiology Simulations
-	to Probabilistic Programs}
+%\icmltitle{Automatically Converting Epidemiology Simulations
+%	to Probabilistic Programs}
+\icmltitle{Hijacking Epidemiology Simulators
+	with Probabilistic Programming}
 
@@
 do not require one to construct hand-crafted joint distributions as they implicitly 
-define a joint distribution of the program and encode learnt structures 
+define a joint distribution of the program and encode learned structures 
 directly. This makes simulators powerful tools and much of machine learning (ML) and 
@@
 \section{Introduction}
-Ending the epidemics of AIDS, tuberculosis, malaria and other infectious diseases by 2030 is a key target within the Good Health \& Well-Being section of the UN Sustainable Development Goals \cite{refugees_2030_nodate,un_sustainable_2018}.
-However, despite decades of substantial international efforts, these diseases kill hundreads of million people a year.
+Ending the epidemics of AIDS, tuberculosis, malaria and other infectious diseases by 2030 is a key target within the Good Health \& Well-Being section of the UN Sustainable Development Goals\todo{These citations
+should still have years} \cite{refugees_2030_nodate,un_sustainable_2018}. 
+However, despite decades of substantial international efforts, these diseases kill hundreds of million people a year.
 For example, malaria still annually kills about a quarter of a million children under the age of 5 in Africa alone.
@@
 
-To reach the WHO’s target of reducing malaria incidence and mortality rates by at least 90\% by 2030, policymakers increasingly turn to evidence-based methods and computational simulations CITATIONS. 
+To reach the WHO’s target of reducing malaria incidence and mortality rates by at least 90\% by 2030, policymakers are increasingly turning to evidence-based methods and computational simulations CITATIONS.
 These simulations allow policymakers to infer critical information like
-INSERT and make predictions about the impacts of policies, allowing
+INSERT and make predictions about the impacts of policies
+before they are rolled out, improving understanding and allowing
 more effective decisions to be made in combating the disease at a
-macro-level and ultimately saving lives.
+macro-level, thereby ultimately saving lives.
 For example, it has been shown that mass vaccination may be largely ineffective in regions of large transmission rates, but may play a crucial role in areas of low transition. \bg{citation?}
@@
 However, malaria epidemiology is governed by a complex set of drivers, 
-few of which can be understood in isolation.
+few of which can be understood in isolation CITATION.
 These include within-host dynamics, population-specific traits and even local geography.
-Comprehensive modeling of all of these components remains challenging, particular in a regional-specific context. Therefore simulators 
-need to be both comprehensive and powerful, while it can
+Comprehensive modeling of all of these components remains challenging, particularly in a regional-specific context. Therefore the resulting simulators 
+are both comprehensive and powerful.  Since many of these simulators are stochastic in nature, it can
 be highly non-trivial to both interpret their outputs and draw desired
-inferences when coupling them with observed data.
+inferences when coupling them with observed data CITATION?.
+In this paper, we show how probabilistic programming
+techniques CITATIONS can be used to help automate the required calculations to do this, interpreting the simulators as 
+generative models to perform principled statistical analysis, while
+providing insights and summary statistics on their behavior to users.
+
+Achieving this aim is substantially complicated by the fact that
+a vast array of different epidemiology simulators are currently 
+used in practice.
+Manually reimplementing these systems in existing
+probabilistic programming systems (PPSs) is therefore infeasible, as each
+system would require manual and laborious porting.  Instead, we 
+take the novel approach of \emph{hijacking} existing simulators by
+overriding their internal random number generator, building
+on ideas from~\citep{baydin2018efficient}.  Specifically, by
+replacing the existing low-level random number generator in a simulator
+with a call to a purpose-built ``controller'' PPS, the stochasticity
+of the simulator can be both manipulated and tracked by this controller
+system.  In turn, this allows for a variety of tasks to be performed on
+the hijacked simulator, such as running inference (by conditioning
+the values of certain draws and manipulating others),
+uncovering stochastic structure, and automatically
+producing result summaries, such as establishing the probability
+of different program paths.  By providing a common abstraction 
+framework for different simulators, our approach further allows for
+easy and direct comparison between related or competing
+simulators, a characteristic that is valuable in the context of
+epidemiology simulators CITATIONS.
+
+Critically, this hijacking can be performed on a wide variety of simulators
+with only minimal changes to the simulator.  Our framework already
+supports 13 popular languages for the base simulator, and can easily
+be extended to others if needed.  Moreover, though our focus
+is on hijacking epidemiology simulators, our general approach and
+framework can be applied more widely to a general class of stochastic
+simulators.  In time, we hope our approach will play a critical role in
+bringing recent advancements in probabilistic programming to bear on
+the vast array of existing simulators used throughout the sciences,
+thereby providing wide-ranging impacts across a number of fields.
 
@@
 
-However, since many of these simulators are stochastic in nature and therefore act as black-boxes, understanding the structure of these complex simulators is challenging and poses several challenges for policy makers who want to interpret the decisions being inferred.
-Therefore, sophisticated tools are required that enable us to not only digest the results of the simulation, but provide pathways to perform complex inferences. 
-To this end, we extend the work of \citep{baydin2018efficient} to enable multiple stochastic epidemiology simulators to be transformed into a probabilistic model that can be used as input to a probabilistic programming system, enabling trace paths to be extracted from the simulator on both a per human level and a population level, from which we can extract trace probabilities allowing policy makers to determine what actions have the biggest impact on differing populations dynamics and thus shape policy decisions accordingly.
- We argue that our work is likely to have significant social impact on many livelihoods affected by Malaria and discuss opportunities for future research.
+%However, since many of these simulators are stochastic in nature and therefore act as black-boxes, understanding the structure of these complex simulators is challenging and poses several challenges for policy makers who want to interpret the decisions being inferred.
+%Therefore, sophisticated tools are required that enable us to not only digest the results of the simulation, but provide pathways to perform complex inferences.
+
+%To this end, we extend the work of \citep{baydin2018efficient} to enable multiple stochastic epidemiology simulators to be transformed into a probabilistic model that can be used as input to a probabilistic programming system, enabling trace paths to be extracted from the simulator on both a per human level and a population level, from which we can extract trace probabilities allowing policy makers to determine what actions have the biggest impact on differing populations dynamics and thus shape policy decisions accordingly.
+% We argue that our work is likely to have significant social impact on many livelihoods affected by Malaria and discuss opportunities for future research.
 

### 2019-04-26 paper/icml_ai4good/main.tex
updated

--- 2019-04-26 paper/icml_ai4good/main.tex
@@
 %\usepackage[accepted]{icml2019}
-\usetikzlibrary{arrows,chains,matrix,positioning,scopes}
-%
+\usetikzlibrary{arrows,chains,matrix,positioning,scopes, arrows.meta}
+
+\tikzset{%
+  >={Latex[width=2mm,length=2mm]},
+  % Specifications for style of nodes:
+            base/.style = {rectangle, rounded corners, draw=black,
+                           minimum width=4cm, minimum height=1cm,
+                           text centered, font=\sffamily},
+  activityStarts/.style = {base, fill=blue!30},
+       startstop/.style = {base, fill=red!30},
+    activityRuns/.style = {base, fill=green!30},
+         process/.style = {base, minimum width=2.5cm, fill=orange!15,
+                           font=\ttfamily},
+}
 \makeatletter
@@
 
-%maybe cut some of this paragraph out
-% In contrast to AIDS, Malaria can be routinely cured thanks to the disposability of a variety of medical interventions. 
-% Issues are further aggravated due to the increasing emergence of parasite resistance to multiple drugs, including artemisinin, or even insecticides. 
-% Hence, despite decades of substantial international efforts, initial hopes of regional elimination or even global eradication have not been met.
-%  So difficult is Malaria control that, even today, about a quarter of a million children under the age of 5 succumb to it every year in Africa alone.
-
 To reach the WHO’s target of reducing malaria incidence and mortality rates by at least 90\% by 2030, policymakers increasingly turn to evidence-based methods. 
@@
 However, comprehensive modeling of all of these components in a regional context remains challenging, and is greatly impeded by the lack of available field data for regional calibration, hence the need for comprehensive simulators.
-% Simulation complexity is further aggravated by considering feedback effects from a large arsenal of anti-malaria interventions, ranging from distribution of mosquito nets, indoor residual spraying and mass drug administration to age-specific vaccinations.
+
 Stochastic epidemiology simulators have already proven to be an invaluable tool to policymakers with policy implications that are often counterintuitive yet far-ranging. 
@@
 
-% Recent work by \citep{Baydin} introduces a fundamental paradigm change that allows to tackle the inverse problem directly. 
-% Using a novel software tool named \textit{pyprob}, \citep{Baydin} extract program execution traces from state-of-the-art particle decay simulation engine SHERPA\cite{}.
-%  Using latest techniques from universal probabilistic programming inference, the sampled execution traces are then used in order to characterise the posterior distribution of the simulator output. 
-%  From these, regions in simulator input space can be associated with specified outcomes.
-
-% SHERPA, however, is different from leading epidemiology simulators in that it does not simulate population dynamics, but instead models individual decays with different sampling strategies.
-Thus, the existing work of \cite{baydin2018efficient} must be extended to encapsulate these different strategies.
-%  In this paper, we for the first time demonstrate the applicability of \textit{pyprob}, and hence probabilistic programming inference, to population-based epidemiology simulators. 
-%  In particular, we discuss how the leading malaria epidemiology simulators EMOD\cite{} and OpenMalaria\cite{} can be interfaced with \textit{pyprob}.
-%   We argue that our work is likely to have significant social impact on many livelihoods affected by Malaria and discuss opportunities for future research.
-
-% maybe talk about this framework enables a pathway to inference etc..
-
-% All of the above limitations require contemporary malaria epidemiology simulators to employ stochastic modeling techniques, which negotiate data scarcity and model uncertainty through the use of appropriate prior distributions that can be tuned even from small samples of field data.
-The population dynamics of groups of typically several thousand hosts are then initialised by sampling from the initial distributions, using initial random seeds.
-Forward simulation then proceeds by rolling out population dynamics over time.
-Simulation outcomes of cherry-picked scenarios and interventions of interest are then compared to suitably designed control runs.
-
-% This is great, if we had inference implemented right now.
-% However valuable forward simulations of epidemiology scenarios are, the holy grail of goal-driven policymaking ultimately lies in solving the inverse problem: Given a desired outcome and regional context, identify the series of interventions that will best reach the desired target after a given amount of time.
-Due to computational constraints, even an approximate search over the whole simulator input space usually proves intractable.
-An experience-guided search over the simulation output of a small set of manually-designed scenarios of interest therefore long seemed the only feasible way of investigating implications of policy interventions.
-
-% , and, above all, the lack of available field data for calibration,
-
-% parasite life-cycle, transmission vectors, population dynamics and even local geographic circumstances, with policy measures ranging from distribution of mosquito nets to experimental vaccines, requires such simulators to be stochastic.
-Simulation outcomes thus need to be evaluated on a population of subjects.
-
-% <Section about prob prog inference, with reference to SHERPA>
-
-%- issues of public health are complex and
-
-%To do. 
-%
-%\begin{itemize}
-%
-%\item Outline the objectives
-%\item State the importance of interpretability in the inference of the simulator
-%\item Our system enables one to understand and interpret the most common paths in a program (simulation).
-%\item In highlighting this a practitioner can then go back to the field and explore that parameter/s in more depth.
-%\item This facilitates understanding and the construction of more detailed models and data collection strategies
-%
-%
-%\end{itemize}
-%
-%% taken from Toms thesis. 
-%
-%Stochastic simulators are common across a number of domains, statistical physics~\cite{landau_binder_2014}, financial modeling~\cite{jackel2002monte},
-%weather prediction~\cite{evensen1994sequential}, epidemiology~\cite{smith2008towards} and many others. 
-%An advantage of using stochastic simulators is that they provide a level of interpretability not found in modern deep learning settings, as they directly incorporate model structure from carefully reasoned observations and experiments.
-%Probabilistic programming systems~(PPSs) are purpose built systems for making inferences and probabilistic 
-%modeling accessible. 
-%In particular, PPSs allow probabilistic models to be represented in the form of a generative model through the
-%probabilistic programming language~(PPL), which enables one to write statements that enable conditioning on data~\cite{gordon2014probabilistic,Goodman08church:a}. 
-%Thus, it seems only natural to connect simulators with PPSs, as simulators explicitly define a generative model, in the language in which they are written, and the nature of the PPSs enables us to perform inference in these simulators, by conditioning on observations, which are fed in as input to the simulators.
-%In doing this we can infer things about stochastic input parameters and other variables sampled during the program's forward execution, whilst also providing full interpretability in the inference results, which is absolutely necessary in safety-critical domains.
-%By exploiting the tools that we develop in this paper, one is able to connect epidemiology simulators with \texttt{Pyprob}~\cite{le-2016-inference} to extract fully interpretable posterior structures allowing practitioners to interpret which input parameters have the largest affect on the epidemic, in our case malaria.
-Over five-hundred-thousand people die from malaria each year, mostly children under five years of age, with 90 per cent of malaria cases occurring in Sub-Saharan Africa.
-An estimated 100-300 million people suffer from malaria each year~\cite{world2016world}.
-%Thus, by understanding the prominence of input parameters to epidemiology simulators, guides decision making processes relating to effective treatment and prevention strategies. 
-%\label{sec:related}
-
-\section{Epidemiology Simulators and Probabilistic Programming}
+ \section{Epidemiology Simulators}
  \label{sec:background}
 
-\subsection{Probabilistic Programming}
-Probabilistic programming~\cite{gordon2014probabilistic,staton2016semantics,kozen1979semantics} is the coming together of multiple disciplines; computer science, statistics and machine learning. 
-It attempts to unify the three fields by creating automated frameworks in which one can easily extrapolate their observed data and perform sophisticated statistical analysis with advance inference procedures such as
-Markov Chain Monte Carlo (MCMC), Variational Inference~(VI) and Deep Neural Networks~(DNN) in an accessible way, i.e.
-without the need of expert knowledge in implementing and using such procedures.
-In addition to this, probabilistic programming systems~(PPSs) provide tools that enable the source code of a program to be throughly analysed in an automated fashion.
-Users can then exploit these algorithms and tools on an array of models
-within the PPS, using the language itself to create models that represent complicated distribution objects which are then analysed by the PPS. 
-PPSs are extensions, or reductions, of existing programming languages such as python, c++, go and so forth, which means that users do not have to learn
-a completely new set of operating semantics and can write programs in the same vain as they are accustomed to.  
-The PPS that we use, PyProb~\cite{le-2016-inference,baydin2018efficient}, belongs to the family of universal PPSs 
-that allow the expression of unrestricted probability models in a Turing complete fashion~\cite{wood2014new,goodman2012church,landau_binder_2014,siddharth2017learning,bingham2019pyro}.
-
-\subsection{Stochastic Simulators}
-
-Simulators encode several years, if not decades, of research and development and so by their very nature are structurally complicated. 
-Within the context of epidemiology simulators we can model the treatment, infection and prevention cycle under a variety of different environmental conditions, enabling predictions to be made on a set of field observations \bg{Make sure that I do not state they can already do what PP can do, i.e.
-the ability to update and infer the input parameters continually conditioned on new observations.}.
+Simulators provide users with the ability to model complex environments and problems in a black-box way, enabling new insights to be generated and predictions to be made.
+A simulator typically encodes several years, if not decades, of research and development and so by their very nature are structurally complicated.
+Within the context of epidemiology simulators we can model the treatment, infection and prevention cycle under a variety of different environmental conditions, enabling predictions to be made on a set of field observations.
+This enables practitioners to model a wide array of different scenarios at ease.
 However, since the development happens over several years, or decades, usually without comprehensive documentation, understanding a code base written with legacy libraries and software is incredibly difficult and an almost impossible task for a non-expert, or expert who has not had access to the previous years of development.
@@
 In particular, the framework currently supports 13 popular languages, but there is nothing to stop this being extended to any language of interest. 
-In addition to this, we can understand the structure of simulators as programs for both decoding \emph{black-box structures} and for posterior inference. This enables two things. 
-1) It enables software developers and policy makers to understand complicated code bases and dissect outputs of simulators for guiding policy decision making, as we shall see later in Section~\ref{sec:casestudy}.
+Furthermore, we can understand the structure of simulators as programs for both decoding \emph{black-box structures} and for posterior inference, which enables two things. 
+1) It enables software developers and policy makers to understand complicated code bases and dissect the outputs of simulators for guiding policy decision making.
  2) It provides interpretable inference results that provide policy makers with predictions that are truly interpretable, in the sense that the end-user of the inference results understands what physical events led to the inference outcome.
  This is critical in several domains, especially in epidemiology. 
- We provide examples of former in next Section~\ref{sec:casestudy} and discuss the latter in future work as extensions to the the current framework.
+ We provide examples of (1) in next Section~\ref{sec:casestudy} and discuss the latter in future work as extensions to the the current framework.
 
- \subsubsection*{Epidemiology Simulators}
+Two prominent simulators for modeling Malaria are OpenMalaria~\cite{smith2008towards} and Emod~\cite{bershteyn2018implementation}, which are used by practitioners to model several aspects of the Malaria disease.
+
  \paragraph*{OpenMalaria} 
-Open Malaria is an open-source simulator funded by the Gates Foundation to model different  aspects of the malaria prevention and treatment cycle for varying population dynamics.
- Models in OpenMalaria are specified via a scenario file, in the form of an XML document, which specifies the specifics of a particular treatment, or prevention strategy, climatic factors, the demographics of the population and so forth.
+is an open-source simulator funded by the Gates Foundation to model different aspects of the malaria prevention and treatment cycle for varying population dynamics.
+ Users specify models in OpenMalaria via a scenario file, in the form of an XML document, which determines the specifics of a particular treatment, or prevention strategy, climatic factors, the demographics of the population and so forth.
 In addition to this, field observations can be encoded inside of, or externally to the scenario file.
-The scenario file encodes all input and output parameters, some of which we wish to simulate to try and infer, others remain fixed.
+The scenario file encodes all input and output parameters, some of which we wish to simulate, others remain fixed.
 
-\paragraph{EMOD} Do the same for EMOD as OM
+\paragraph{EMOD} provides the modularity and structural flexibility necessary for modeling multiple infectious diseases at the within-host and population level. 
+EMOD can be used to model HIV, Tuberculosis, Malaria and generic diseases, but for the purposes of the comparison, we just focus on Malaria. 
+To specify models in EMOD a configuration file is written by the user, which is in the form of a JSON and provides similar functionality to OpenMalaria, in terms of what can be specified and the outputs that it produces.
 
 In both these well used simulators users are accustomed to writing their models as either XML, or JSON files and so it is important that any system built around this framework enables the models to be inputted in the same format, without requiring the users to learn how to use a new set of protocols.
+Our system facilitates this as we shall discuss in the proceeding sections.
 
-\section{Case Study}
-\label{sec:casestudy}
-\subsection{Connecting a Epidemiology Simulators to a Probabilistic Programming System}
+\section{Hijacking Simulators}
+\label{sec:hijackingsimulators}
 
-In order to overhaul the stochastic simulators we are only required to override the location of the stochastic primitives, i.e.
+% Probabilistic programming~\cite{gordon2014probabilistic,staton2016semantics,kozen1979semantics} is the intersection of computer science, statistics and machine learning. 
+% It attempts to unify the three fields by creating automated frameworks in which one can easily extrapolate their observed data and perform sophisticated statistical analysis with advance inference procedures such as
+% Markov Chain Monte Carlo (MCMC), Variational Inference~(VI) and Deep Neural Networks~(DNN) in an accessible way, i.e.
+without the need of expert knowledge in implementing and using such procedures.
+In addition to this, probabilistic programming systems~(PPSs) provide tools that enable the source code of a program to be throughly analysed in an automated fashion.
+% Users can then exploit these algorithms and tools on an array of models
+% within the PPS, using the language itself to create models that represent complicated distribution objects which are then analysed by the PPS. 
+% PPSs are extensions, or reductions, of existing programming languages such as python, c++, go and so forth, which means that users do not have to learn
+% a completely new set of operating semantics and can write programs in the same vain as they are accustomed to.  
+% The PPS that we use, PyProb~\cite{le-2016-inference,baydin2018efficient}, belongs to the family of universal PPSs 
+% that allow the expression of unrestricted probability models in a Turing complete fashion~\cite{wood2014new,goodman2012church,landau_binder_2014,siddharth2017learning,bingham2019pyro}.
+
+\begin{itemize}
+\item Add what - What are we trying to do to make peoples lifes easier
+\item Add WHy - Why is this important to do and what does it enable.
+\item Then deal with "How" - How is this done - what is currently there. 
+\end{itemize}
+\begin{tikzpicture}[node distance=1.5cm,
+    every node/.style={fill=white, font=\sffamily}, align=center]
+  % Specification of nodes (position, etc.)
+  % \node (start)             [activityStarts]              {Activity starts};
+  \node (SimulatorSystem)     [activityStarts]          {Simulator System \\ e.g EMOD};
+  \node (onStartBlock)      [process, below of=SimulatorSystem]   {onStart()};
+  \node (onResumeBlock)     [process, below of=onStartBlock]   {onResume()};
+  \node (activityRuns)      [activityRuns, below of=onResumeBlock]
+                                                      {Activity is running};
+  \node (onPauseBlock)      [process, below of=activityRuns, yshift=-1cm]
+                                                                {onPause()};
+  \node (onStopBlock)       [process, below of=onPauseBlock, yshift=-1cm]
+                                                                 {onStop()};
+  \node (onDestroyBlock)    [process, below of=onStopBlock, yshift=-1cm] 
+                                                              {onDestroy()};
+  \node (onRestartBlock)    [process, right of=onStartBlock, xshift=4cm]
+                                                              {onRestart()};
+  \node (ActivityEnds)      [startstop, left of=activityRuns, xshift=-4cm]
+                                                        {Process is killed};
+  \node (ActivityDestroyed) [startstop, below of=onDestroyBlock]
+                                                    {Activity is shut down};     
+  % Specification of lines between nodes specified above
+  % with aditional nodes for description 
+  % \draw[->]             (start) -- (onCreateBlock);
+  \draw[->]     (SimulatorSystem) -- (onStartBlock);
+  \draw[->]      (onStartBlock) -- (onResumeBlock);
+  \draw[->]     (onResumeBlock) -- (activityRuns);
+  \draw[->]      (activityRuns) -- node[text width=4cm]
+                                   {Another activity comes in
+                                    front of the activity} (onPauseBlock);
+  \draw[->]      (onPauseBlock) -- node {The activity is no longer visible}
+                                   (onStopBlock);
+  \draw[->]       (onStopBlock) -- node {The activity is shut down by
+                                   user or system} (onDestroyBlock);
+  \draw[->]    (onRestartBlock) -- (onStartBlock);
+  \draw[->]       (onStopBlock) -| node[yshift=1.25cm, text width=3cm]
+                                   {The activity comes to the foreground}
+                                   (onRestartBlock);
+  \draw[->]    (onDestroyBlock) -- (ActivityDestroyed);
+  \draw[->]      (onPauseBlock) -| node(priorityXMemory)
+                                   {higher priority $\rightarrow$ more memory}
+                                   (ActivityEnds);
+  \draw           (onStopBlock) -| (priorityXMemory);
+  \draw[->]     (ActivityEnds)  |- node [yshift=-2cm, text width=3.1cm]
+                                    {User navigates back to the activity}
+                                    (SimulatorSystem);
+  \draw[->] (onPauseBlock.east) -- ++(2.6,0) -- ++(0,2) -- ++(0,2) --                
+     node[xshift=1.2cm,yshift=-1.5cm, text width=2.5cm]
+     {The activity comes to the foreground}(onResumeBlock.east);
+  \end{tikzpicture}
+As stated previously, Simulators are complex systems and should be viewed as black-boxes, thus deciphering the underlying processes occurring within the simulator facilitates the users understanding of how such a system operates and provides a way of interpreting the outputs in a clear way.
+In addition to this, when using a simulator to make new inferences a user will often want to condition on the output of the simulator, or any new observations gathered.
+In the standard simulator framework this is not possible, but in
+
+Essentially, from the users perspective, our approach can be viewed as follows the user would first select the simulator that they are going to use, for example EMOD, they would then write their model, in the desired simulator, according to the given simulator specification.
+The user would then run the simulator.
+Now comes the hijacking.
+During the forward run of the simulator all sampling calls from the random number generator are directed through the probabilistic programming system~(PPS), which in addition to the standard output of the simulator now generates a series of result summaries and path probabilities, as we demonstrate in Section~\ref{sec:casestudy}.
+
+By connecting with the PPS we track all sampling procedures, which enables us to control the myriad of random variables generated within the simulator.
+
+In order to overhaul a stochastic simulator like EMOD, or OpenMalaria, we are only required to override the location of the stochastic primitives, i.e.
 the base operations that generate the stochasticity and randomness within the simulator.
-We will walk through how this is done for a standard epidemiology simulator.
+We will walk through how this is done for a standard epidemiology simulator, see the flowchart for a pictorial representation of what is happening \bg{add ref to flow chart}.
 
@@
 
-\begin{table}[h!]
-\footnotesize
-\begin{tabular}{p{3.6cm}|p{3.6cm}}
-Before  &  After  \\
-\midrule
-\begin{lstlisting}
-#include <gsl/gsl_cdf.h>
-#include <gsl/gsl_rng.h>
-#include <gsl/gsl_randist.h>
-//preamble
-double random::uniform_01 () {
-    double result =
-# ifdef 
-OM_RANDOM_USE_BOOST
-       rng_uniform01 ();
-# else
- gsl_rng_uniform (rng.gsl_generator);
-# endif
-    return result;
-    }
-\end{lstlisting}&
-\begin{lstlisting}
-#include <pyprob_cpp.h>
-#include "xtensor/xadapt.hpp"
-// preamble
-double random::uniform_01 () {
-auto uniform = pyprob_cpp::distributions::Uniform(0,1);
-return pyprob_cpp::sample(uniform)(0);
-}
-\end{lstlisting}
-\end{tabular} 
-\end{table}
-As can be seen, writing the new sampling call is cleaner and syntactically simpler. 
+% \begin{table}[h!]
+% \footnotesize
+% \begin{tabular}{p{3.6cm}|p{3.6cm}}
+% Before  &  After  \\
+% \midrule
+% \begin{lstlisting}
+% #include <gsl/gsl_cdf.h>
+% #include <gsl/gsl_rng.h>
+% #include <gsl/gsl_randist.h>
+% //preamble
+% double random::uniform_01 () {
+%     double result =
+% # ifdef 
+% OM_RANDOM_USE_BOOST
+%        rng_uniform01 ();
+% # else
+%  gsl_rng_uniform (rng.gsl_generator);
+% # endif
+%     return result;
+%     }
+% \end{lstlisting}&
+% \begin{lstlisting}
+% #include <pyprob_cpp.h>
+% #include "xtensor/xadapt.hpp"
+% // preamble
+% double random::uniform_01 () {
+% auto uniform = pyprob_cpp::distributions::Uniform(0,1);
+% return pyprob_cpp::sample(uniform)(0);
+% }
+% \end{lstlisting}
+% \end{tabular} 
+% \end{table}
+% As can be seen, writing the new sampling call is cleaner and syntactically simpler. 
 The third step is generating the protocols, which enable communication between the simulator and probabilistic programming system. 
@@
 Once these steps are complete, the containerized environment will now contain everything required to pass models through to the simulators, encode the simulator into a probabilistic program, which can then be evaluated in the PPS.
-We provide examples of the generated trace plots from the connection between the simulator and the PPS in Figure~\ref{fig:plotewan}.
-
-\subsection{Extending Probabilistic Programming Protocols}
 
 \bg{Add performance plot showing memory consumptions of pyprob and our new system.}
+\section{Case Study}
+\label{sec:casestudy}
+
+We provide examples of the generated trace plots from the connection between the simulator and the PPS in Figure~\ref{fig:plotewan}. 
+When comparing the two trace plots between EMOD and OpenMalaria we see significant difference, despite both evaluating the same model for the same disease. 
+We can see from the addressing schemes $A1, \ldots, A_N$ \bg{ref table 1 and 2} what physical events are connected to each other and how outputs in EMOD are generated from a different set of procedures as compared to OpenMalaria
+By having access to such diagrams users can internally evaluate and scrutinize the decisions that the simulator is making.
+This is incredibly important for policy makers, or non-experts, as it not only details how we arrive at the given outputs, but it provides an understanding of which processes were most crucial in determining those outputs as can be seen from the path probabilities assigned to each of the vertices.
+
+%
+
 % \begin{tikzpicture}
@@
 	\centering
-	\includegraphics[width=\textwidth]{../plots/ewan_25_pop_10.pdf}
-	\centering
+	\includegraphics[width=\textwidth]{../plots/emod_24_prior_pop_100.pdf}
 	\includegraphics[width=\textwidth]{../plots/ewan_25_pop_100.pdf}
-	\caption{Add something. Population 100}
+	\caption{Here we run two equivalent models, compare the corresponding trace paths and corresponding 
+	path probabilities taken by the thousands of random variables generated internally within the simulators.
+\textit{\textbf{Top:}} The specified model run in EMOD.
+\textit{\textbf{Bottom:}} The specified model in OpenMalaria.}
 	\label{fig:plotewan}
@@
   \setlength{\tabcolsep}{1mm}
-  \caption{Addresses TODO }
+  \caption{A few addresses generated for the OpenMalaria simulator, see the Supplement for the full table.}
   \label{table:addresses}
@@
 
-\bg{Add comments about how this trace structure helps policy makers.}
 \bibliographystyle{icml2019}

### 2019-04-26 paper/icml_ai4good/main.tex

--- 2019-04-26 paper/icml_ai4good/main.tex
@@
 
-\section{Epidemiology Simulators and Probabilistic Programming}
+ \section{Epidemiology Simulators}
  \label{sec:background}
 
-\subsection{Probabilistic Programming}
-Probabilistic programming~\cite{gordon2014probabilistic,staton2016semantics,kozen1979semantics} is the coming together of multiple disciplines; computer science, statistics and machine learning. 
-It attempts to unify the three fields by creating automated frameworks in which one can easily extrapolate their observed data and perform sophisticated statistical analysis with advance inference procedures such as
-Markov Chain Monte Carlo (MCMC), Variational Inference~(VI) and Deep Neural Networks~(DNN) in an accessible way, i.e.
-without the need of expert knowledge in implementing and using such procedures.
-In addition to this, probabilistic programming systems~(PPSs) provide tools that enable the source code of a program to be throughly analysed in an automated fashion.
-Users can then exploit these algorithms and tools on an array of models
-within the PPS, using the language itself to create models that represent complicated distribution objects which are then analysed by the PPS. 
-PPSs are extensions, or reductions, of existing programming languages such as python, c++, go and so forth, which means that users do not have to learn
-a completely new set of operating semantics and can write programs in the same vain as they are accustomed to.  
-The PPS that we use, PyProb~\cite{le-2016-inference,baydin2018efficient}, belongs to the family of universal PPSs 
-that allow the expression of unrestricted probability models in a Turing complete fashion~\cite{wood2014new,goodman2012church,landau_binder_2014,siddharth2017learning,bingham2019pyro}.
-
-\subsection{Stochastic Simulators}
-
-Simulators encode several years, if not decades, of research and development and so by their very nature are structurally complicated. 
-Within the context of epidemiology simulators we can model the treatment, infection and prevention cycle under a variety of different environmental conditions, enabling predictions to be made on a set of field observations \bg{Make sure that I do not state they can already do what PP can do, i.e.
-the ability to update and infer the input parameters continually conditioned on new observations.}.
+Simulators provide users with the ability to model complex environments and problems in a black-box way, enabling new insights to be generated and predictions to be made.
+A simulator typically encodes several years, if not decades, of research and development and so by their very nature are structurally complicated.
+Within the context of epidemiology simulators we can model the treatment, infection and prevention cycle under a variety of different environmental conditions, enabling predictions to be made on a set of field observations.
+This enables practitioners to model a wide array of different scenarios at ease.
 However, since the development happens over several years, or decades, usually without comprehensive documentation, understanding a code base written with legacy libraries and software is incredibly difficult and an almost impossible task for a non-expert, or expert who has not had access to the previous years of development.

### 2019-04-26 paper/icml_ai4good/main.tex
new abstract

--- 2019-04-26 paper/icml_ai4good/main.tex
@@
 \begin{abstract}
-
-A goal of probabilistic programming is to couple simulators, with inference. This is 
-because stochastic simulators are used prominently in many industrial settings,
-do not require one to construct hand-crafted joint distributions as they implicitly 
-define a joint distribution of the program and encode learned structures 
-directly. This makes simulators powerful tools and much of machine learning (ML) and 
-Artificial Intelligence (AI)
-can be seen as trying to emulate such simulators from a purely data-driven approach.
-However, in the 
-ML/AI setting, although we can often infer outcomes, we have little understanding about what 
-in the data led to the outputted inferences. 
-This makes it challenging to deploy ML/AI systems into the wild, especially in health-related and safety-critical domains, such
-as epidemiology, as we lose \emph{interpretability}. 
-In this work, we explain how to design ML/AI systems that combine
-probabilistic programming systems (PPSs) and epidemiology simulators, to extract
-fully interpretable structures, enabling policy makers 
-and practitioners to make decisions from simulator outputs that can be fully reasoned on.
- % interpretable inferences. 
-In particular, we demonstrate this for the Malaria disease in two commonly used simulators; EMOD and OpenMalaria.
+Epidemiology simulations have become a fundamental tool in the fight
+against the epidemics of various infectious diseases like AIDS and
+malaria.
+However, the complicated and stochastic nature of these simulators
+can mean they are difficult to work with and extract the key information
+required by policy makers.
+In this paper, we introduce an approach to \emph{hijack} these
+simulators with an external
+probabilistic programming system (PPS) that controls
+their internal random number generator.
+This allows for various advanced statistical analyses to be performed
+on the simulators in an automated fashion, while also tracking
+the behavior of the simulator to uncover key information such
+as stochastic structure and generate result summaries.
+Critically, this hijacking can be performed with only minimal changes
+to the original simulator, allowing for straightforward application
+of our approach to a wide variety of contexts and existing simulators.
+%
+%A goal of probabilistic programming is to couple simulators, with inference. This is 
+%because stochastic simulators are used prominently in many industrial settings,
+%do not require one to construct hand-crafted joint distributions as they implicitly 
+%define a joint distribution of the program and encode learned structures 
+%directly. This makes simulators powerful tools and much of machine learning (ML) and 
+%Artificial Intelligence (AI)
+%can be seen as trying to emulate such simulators from a purely data-driven approach.
+%However, in the 
+%ML/AI setting, although we can often infer outcomes, we have little understanding about what 
+%in the data led to the outputted inferences. 
+%This makes it challenging to deploy ML/AI systems into the wild, especially in health-related and safety-critical domains, such
+%as epidemiology, as we lose \emph{interpretability}. 
+%In this work, we explain how to design ML/AI systems that combine
+%probabilistic programming systems (PPSs) and epidemiology simulators, to extract
+%fully interpretable structures, enabling policy makers 
+%and practitioners to make decisions from simulator outputs that can be fully reasoned on.
+% % interpretable inferences. 
+%In particular, we demonstrate this for the Malaria disease in two commonly used simulators; EMOD and OpenMalaria.
 \end{abstract}
@@
 Critically, this hijacking can be performed on a wide variety of simulators
-with only minimal changes to the simulator.  Our framework already
+with only minimal changes.  Our framework already
 supports 13 popular languages for the base simulator, and can easily

### 2019-04-26 paper/icml_ai4good/main.tex
title change

@@
 %\icmltitlerunning{Hijacking Simulators for the Greater Good}
-\icmltitlerunning{Hijacking Epidemiology Simulators
+\icmltitlerunning{Hijacking Malaria Simulators
 	with Probabilistic Programming}
@@
 %	to Probabilistic Programs}
-\icmltitle{Hijacking Epidemiology Simulators
+\icmltitle{Hijacking Malaria Simulators
 	with Probabilistic Programming}

### 2019-04-26 paper/icml_ai4good/main.tex
updated

@@
 When comparing the two trace plots between EMOD and OpenMalaria we see significant difference, despite both evaluating the same model for the same disease. 
-We can see from the addressing schemes $A1, \ldots, A_N$ \bg{ref table 1 and 2} what physical events are connected to each other and how outputs in EMOD are generated from a different set of procedures as compared to OpenMalaria
+We can see from the addressing schemes $A_1, \ldots, A_N$ \bg{ref table 1 and 2} what physical events are connected to each other and how outputs in EMOD are generated from a different set of procedures as compared to OpenMalaria
 By having access to such diagrams users can internally evaluate and scrutinize the decisions that the simulator is making.

### 2019-04-26 paper/icml_ai4good/main.tex
updated

@@
 %
-%Stochastic simulators are common across a number of domains, statistical physics~\cite{landau_binder_2014}, financial modeling~\cite{jackel2002monte},
-%weather prediction~\cite{evensen1994sequential}, epidemiology~\cite{smith2008towards} and many others. 
-%An advantage of using stochastic simulators is that they provide a level of interpretability not found in modern deep learning settings, as they directly incorporate model structure from carefully reasoned observations and experiments.
-%Probabilistic programming systems~(PPSs) are purpose built systems for making inferences and probabilistic 
-%modeling accessible. 
+
 %In particular, PPSs allow probabilistic models to be represented in the form of a generative model through the
@@
 
-Simulators provide users with the ability to model complex environments and problems in a black-box way, enabling new insights to be generated and predictions to be made.
+% Stochastic simulators are common across a number of domains, statistical physics~\cite{landau_binder_2014}, financial modeling~\cite{jackel2002monte},
+% weather prediction~\cite{evensen1994sequential}, epidemiology~\cite{smith2008towards} and many others. 
+% An advantage of using stochastic simulators is that they provide a level of interpretability not found in modern deep learning settings, as they directly incorporate model structure from carefully reasoned observations and experiments.
+In addition to this,
+
+In-silico simulators have become a crucial tool in evidence-based decision-making within a large number of disciplines, including statistical physics~\cite{landau_binder_2014}, financial modeling~\cite{jackel2002monte},
+weather prediction~\cite{evensen1994sequential}, epidemiology~\cite{smith2008towards} and many others.
+In many cases, simulation output can augment or even replace real data that may otherwise be costly or even impossible to generate.
+Recent advances in hardware have enabled simulations to model increasingly complex systems.
+Epidemiology studies the prevalence and spreading of diseases across populations.
+Recent advances in hardware have enabled simulations to model the dynamics of infectious diseases, such as Malaria, in ever greater detail.
+Two the most advanced Malaria simulators, namely EMOD~\cite{bershteyn2018implementation} and OpenMalaria\cite{smith2008towards}, have proven to be particularly valuable to policymakers.
+OpenMalaria is based on microsimulations of Plasmodium falciparum in humans and was originally developed to simulate the impacts of malaria vaccines within simple villages or districts
+Compared with OpenMalaria, EMOD is able to simulate a variety of additional drivers, including complex geographies complete with migration and a large number of policy interventions
+Both EMOD and OpenMalaria are open source and implemented in C++.
+
+% Simulators provide users with the ability to model complex environments and problems in a black-box way, enabling new insights to be generated and predictions to be made.
 A simulator typically encodes several years, if not decades, of research and development and so by their very nature are structurally complicated.
-Within the context of epidemiology simulators we can model the treatment, infection and prevention cycle under a variety of different environmental conditions, enabling predictions to be made on a set of field observations.
-This enables practitioners to model a wide array of different scenarios at ease.
-However, since the development happens over several years, or decades, usually without comprehensive documentation, understanding a code base written with legacy libraries and software is incredibly difficult and an almost impossible task for a non-expert, or expert who has not had access to the previous years of development.
+
+% Within the context of epidemiology simulators we can model the treatment, infection and prevention cycle under a variety of different environmental conditions, enabling predictions to be made on a set of field observations.
+Such scenarios in practices would be
+
+% This enables practitioners to model a wide array of different scenarios at ease.
+% However, since the development happens over several years, or decades, usually without comprehensive documentation, understanding a code base written with legacy libraries and software is incredibly difficult and an almost impossible task for a non-expert, or expert who has not had access to the previous years of development.
+%
+
+% Two prominent simulators for modeling Malaria are OpenMalaria~\cite{smith2008towards} and Emod~\cite{bershteyn2018implementation}, which are used by practitioners to model several aspects of the Malaria disease.
+
+%  \paragraph*{OpenMalaria} 
+% is an open-source simulator funded by the Gates Foundation to model different aspects of the malaria prevention and treatment cycle for varying population dynamics.
+%  Users specify models in OpenMalaria via a scenario file, in the form of an XML document, which determines the specifics of a particular treatment, or prevention strategy, climatic factors, the demographics of the population and so forth.
+In addition to this, field observations can be encoded inside of, or externally to the scenario file.
+% The scenario file encodes all input and output parameters, some of which we wish to simulate, others remain fixed.
+
+% \paragraph{EMOD} provides the modularity and structural flexibility necessary for modeling multiple infectious diseases at the within-host and population level. 
+% EMOD can be used to model HIV, Tuberculosis, Malaria and generic diseases, but for the purposes of the comparison, we just focus on Malaria. 
+% To specify models in EMOD a configuration file is written by the user, which is in the form of a JSON and provides similar functionality to OpenMalaria, in terms of what can be specified and the outputs that it produces.
+
+% In both these well used simulators users are accustomed to writing their models as either XML, or JSON files and so it is important that any system built around this framework enables the models to be inputted in the same format, without requiring the users to learn how to use a new set of protocols.
+Our system facilitates this as we shall discuss in the proceeding sections.
+
+\section{Hijacking Simulators}
+\label{sec:hijackingsimulators}
  Our method of overriding simulators builds on the work of~\cite{baydin2018efficient} and extends the framework to encapsulate a more diverse range of simulators, in particular population based simulators such as those found in epidemiology.
@@
  We provide examples of (1) in next Section~\ref{sec:casestudy} and discuss the latter in future work as extensions to the the current framework.
-
-Two prominent simulators for modeling Malaria are OpenMalaria~\cite{smith2008towards} and Emod~\cite{bershteyn2018implementation}, which are used by practitioners to model several aspects of the Malaria disease.
-
- \paragraph*{OpenMalaria} 
-is an open-source simulator funded by the Gates Foundation to model different aspects of the malaria prevention and treatment cycle for varying population dynamics.
- Users specify models in OpenMalaria via a scenario file, in the form of an XML document, which determines the specifics of a particular treatment, or prevention strategy, climatic factors, the demographics of the population and so forth.
-In addition to this, field observations can be encoded inside of, or externally to the scenario file.
-The scenario file encodes all input and output parameters, some of which we wish to simulate, others remain fixed.
-
-\paragraph{EMOD} provides the modularity and structural flexibility necessary for modeling multiple infectious diseases at the within-host and population level. 
-EMOD can be used to model HIV, Tuberculosis, Malaria and generic diseases, but for the purposes of the comparison, we just focus on Malaria. 
-To specify models in EMOD a configuration file is written by the user, which is in the form of a JSON and provides similar functionality to OpenMalaria, in terms of what can be specified and the outputs that it produces.
-
-In both these well used simulators users are accustomed to writing their models as either XML, or JSON files and so it is important that any system built around this framework enables the models to be inputted in the same format, without requiring the users to learn how to use a new set of protocols.
-Our system facilitates this as we shall discuss in the proceeding sections.
-
-\section{Hijacking Simulators}
-\label{sec:hijackingsimulators}
-
 % Probabilistic programming~\cite{gordon2014probabilistic,staton2016semantics,kozen1979semantics} is the intersection of computer science, statistics and machine learning. 
@@
   % \node (start)             [activityStarts]              {Activity starts};
-  \node (SimulatorSystem)     [activityStarts]          {Simulator System \\ e.g EMOD};
-  \node (onStartBlock)      [process, below of=SimulatorSystem]   {onStart()};
-  \node (onResumeBlock)     [process, below of=onStartBlock]   {onResume()};
+  \node (SimulatorSystem)     [activityStarts]          {Simulator System \\ e.g. EMOD};
+  \node (onStartBlock)      [process, below of=SimulatorSystem]   {Specific Model / Specification \\ e.g. XML files};
+  \node (onResumeBlock)     [process, below of=onStartBlock]   {Run Simulator};
   \node (activityRuns)      [activityRuns, below of=onResumeBlock]
@@
   \node (onPauseBlock)      [process, below of=activityRuns, yshift=-1cm]
-                                                                {onPause()};
+                                                                {Posterior Output};
   \node (onStopBlock)       [process, below of=onPauseBlock, yshift=-1cm]
-                                                                 {onStop()};
+                                                                 {};
   \node (onDestroyBlock)    [process, below of=onStopBlock, yshift=-1cm] 
@@
   \draw[->]      (onPauseBlock) -| node(priorityXMemory)
-                                   {higher priority $\rightarrow$ more memory}
+                                   {todo}
                                    (ActivityEnds);
@@
 We provide examples of the generated trace plots from the connection between the simulator and the PPS in Figure~\ref{fig:plotewan}. 
-When comparing the two trace plots between EMOD and OpenMalaria we see significant difference, despite both evaluating the same model for the same disease. 
+When comparing the two trace plots between EMOD and OgpenMalaria we see significant difference, despite both evaluating the same model for the same disease. 
 We can see from the addressing schemes $A_1, \ldots, A_N$ \bg{ref table 1 and 2} what physical events are connected to each other and how outputs in EMOD are generated from a different set of procedures as compared to OpenMalaria

### 2019-04-26 paper/icml_ai4good/main.tex
some intro changes

@@
 
-To reach the WHO’s target of reducing malaria incidence and mortality rates by at least 90\% by 2030, policymakers are increasingly turning to evidence-based methods and computational simulations CITATIONS.
-These simulations allow policymakers to infer critical information like
-INSERT and make predictions about the impacts of policies
-before they are rolled out, improving understanding and allowing
-more effective decisions to be made in combating the disease at a
-macro-level, thereby ultimately saving lives.
-For example, it has been shown that mass vaccination may be largely ineffective in regions of large transmission rates, but may play a crucial role in areas of low transition. \bg{citation?}
+To reach the WHO’s target of reducing malaria incidence and mortality rates by at least 90\% by 2030, policymakers are increasingly turning to evidence-based methods, thus oftentimes relying on computational simulations CITATIONS.
+These simulations allow policymakers to infer critical information on disease dynamics and make predictions about the impacts of policies
+before they are rolled out. This frequently increases the effectiveness of interventions and thus ultimately saves resources, or even lifes.
+For example, it has been shown that mass vaccination may be largely ineffective in regions of large transmission rates, but may play a crucial role in areas of low transmission \bg{citation?}.
 
-However, malaria epidemiology is governed by a complex set of drivers, 
+Malaria epidemiology is governed by a complex set of drivers, 
 few of which can be understood in isolation CITATION.
 These include within-host dynamics, population-specific traits and even local geography.
-Comprehensive modeling of all of these components remains challenging, particularly in a regional-specific context. Therefore the resulting simulators 
-are both comprehensive and powerful.  Since many of these simulators are stochastic in nature, it can
-be highly non-trivial to both interpret their outputs and draw desired
-inferences when coupling them with observed data CITATION?.
+Comprehensive modeling of all of these components remains challenging, particularly in a region-specific context. 
+Computational epidemiology simulators have to reflect these complexities and are usually stochastic in nature.
+This can make simulation output highly non-trivial to interpret, particularly when trying to draw desired inferences coupled with observed data \cite{TODO}.
+
 In this paper, we show how probabilistic programming

### 2019-04-26 paper/icml_ai4good/main.tex
fixed abstract

@@
 However, the complicated and stochastic nature of these simulators
-can mean they are difficult to work with and extract the key information
-required by policy makers.
-In this paper, we introduce an approach to \emph{hijack} these
-simulators with an external
-probabilistic programming system (PPS) that controls
-their internal random number generator.
-This allows for various advanced statistical analyses to be performed
-on the simulators in an automated fashion, while also tracking
-the behavior of the simulator to uncover key information such
-as stochastic structure and generate result summaries.
-Critically, this hijacking can be performed with only minimal changes
-to the original simulator, allowing for straightforward application
-of our approach to a wide variety of contexts and existing simulators.
+can mean their output is difficult to interpret, which reduces their usefulness to policymakers.
+In this paper, we introduce an approach that allows one to treat a large class of population-based epidemiology simulators as probabilistic generative models.
+This is achieved by \emph{hijacking} the internal random number generator calls, using only minimal changes to pre-existing simulators, through the use of an universal probabilistic programming system (PPS).
+We demonstrate that our method can be used for interpretable introspection of and inference on what used to be de facto black box simulators.
+This reinstates much needed trust between policymakers and evidence-based methods.
+
+%This allows for various advanced statistical analyses to be performed
+%on the simulators in an automated fashion, while also tracking
+%the behavior of the simulator to uncover key information such
+%as stochastic structure and generate result summaries.
+%Critically, this hijacking can be performed with only minimal changes
+%to the original simulator, allowing for straightforward application
+%of our approach to a wide variety of contexts and existing simulators.
 %

### 2019-04-26 paper/icml_ai4good/main.tex
abstract v2

@@
 In this paper, we introduce an approach that allows one to treat a large class of population-based epidemiology simulators as probabilistic generative models.
-This is achieved by \emph{hijacking} the internal random number generator calls, using only minimal changes to pre-existing simulators, through the use of an universal probabilistic programming system (PPS).
+This is achieved by \emph{hijacking} the internal random number generator calls, through the use of an universal probabilistic programming system (PPS).
+In contrast to other methods, our approach requires only minimal changes to existing simulator to be made.
 We demonstrate that our method can be used for interpretable introspection of and inference on what used to be de facto black box simulators.

### 2019-04-26 paper/icml_ai4good/main.tex
minor abstract fix
added plot

--- 2019-04-26 paper/icml_ai4good/main.tex
@@
 \usepackage{todonotes}
+\usepackage{tikz-cd}
 
@@
 \usetikzlibrary{arrows,chains,matrix,positioning,scopes, arrows.meta}
-
+\tikzset{
+  shift left/.style ={commutative diagrams/shift left={#1}},
+  shift right/.style={commutative diagrams/shift right={#1}}
+}
 \tikzset{%
@@
                            text centered, font=\sffamily},
-  activityStarts/.style = {base, fill=blue!30},
-       startstop/.style = {base, fill=red!30},
-    activityRuns/.style = {base, fill=green!30},
-         process/.style = {base, minimum width=2.5cm, fill=orange!15,
+  activityStarts/.style = {base, fill=blue!50},
+       startstop/.style = {base, fill=red!50},
+    activityRuns/.style = {base, fill=green!50},
+         process/.style = {base, minimum width=2.5cm, fill=orange!55,
                            font=\ttfamily},
@@
 
-To reach the WHO’s target of reducing malaria incidence and mortality rates by at least 90\% by 2030, policymakers are increasingly turning to evidence-based methods, thus oftentimes relying on computational simulations CITATIONS.
-These simulations allow policymakers to infer critical information on disease dynamics and make predictions about the impacts of policies
-before they are rolled out. This frequently increases the effectiveness of interventions and thus ultimately saves resources, or even lifes.
-For example, it has been shown that mass vaccination may be largely ineffective in regions of large transmission rates, but may play a crucial role in areas of low transmission \bg{citation?}.
+To reach the WHO’s target of reducing malaria incidence and mortality rates by at least 90\% by 2030, policymakers are increasingly turning to evidence-based methods and computational simulations CITATIONS.
+These simulations allow policymakers to infer critical information like
+INSERT and make predictions about the impacts of policies
+before they are rolled out, improving understanding and allowing
+more effective decisions to be made in combating the disease at a
+macro-level, thereby ultimately saving lives.
+For example, it has been shown that mass vaccination may be largely ineffective in regions of large transmission rates, but may play a crucial role in areas of low transition. \bg{citation?}
 
-Malaria epidemiology is governed by a complex set of drivers, 
+However, malaria epidemiology is governed by a complex set of drivers, 
 few of which can be understood in isolation CITATION.
 These include within-host dynamics, population-specific traits and even local geography.
-Comprehensive modeling of all of these components remains challenging, particularly in a region-specific context. 
-Computational epidemiology simulators have to reflect these complexities and are usually stochastic in nature.
-This can make simulation output highly non-trivial to interpret, particularly when trying to draw desired inferences coupled with observed data \cite{TODO}.
-
+Comprehensive modeling of all of these components remains challenging, particularly in a regional-specific context. Therefore the resulting simulators 
+are both comprehensive and powerful.  Since many of these simulators are stochastic in nature, it can
+be highly non-trivial to both interpret their outputs and draw desired
+inferences when coupling them with observed data CITATION?.
 In this paper, we show how probabilistic programming
@@
 
- \section{Epidemiology Simulators}
+ \section{Simulating Diseases}
  \label{sec:background}
@@
 Recent advances in hardware have enabled simulations to model the dynamics of infectious diseases, such as Malaria, in ever greater detail.
+
+\subsection{Epidemiology Simulators}
+
 Two the most advanced Malaria simulators, namely EMOD~\cite{bershteyn2018implementation} and OpenMalaria\cite{smith2008towards}, have proven to be particularly valuable to policymakers.
@@
  We provide examples of (1) in next Section~\ref{sec:casestudy} and discuss the latter in future work as extensions to the the current framework.
+
 % Probabilistic programming~\cite{gordon2014probabilistic,staton2016semantics,kozen1979semantics} is the intersection of computer science, statistics and machine learning. 
@@
 \end{itemize}
+\begin{figure}
 \begin{tikzpicture}[node distance=1.5cm,
@@
   \node (SimulatorSystem)     [activityStarts]          {Simulator System \\ e.g. EMOD};
-  \node (onStartBlock)      [process, below of=SimulatorSystem]   {Specific Model / Specification \\ e.g. XML files};
-  \node (onResumeBlock)     [process, below of=onStartBlock]   {Run Simulator};
-  \node (activityRuns)      [activityRuns, below of=onResumeBlock]
-                                                      {Activity is running};
-  \node (onPauseBlock)      [process, below of=activityRuns, yshift=-1cm]
+   \node (randomprim)      [process, left of=SimulatorSystem, xshift=5.6cm]
+                                                      {Redefine RNG \\ to make \\ externally \\ callable};
+  \node (onStartBlock)      [process, below of=SimulatorSystem]   {Specify Model \\ e.g. XML files};
+  \node (activityRuns)      [activityRuns, below of=onStartBlock]
+                                                      {Run Simulator};
+  \node (onPauseBlock)      [startstop, below of=activityRuns, yshift=-1cm]
                                                                 {Posterior Output};
-  \node (onStopBlock)       [process, below of=onPauseBlock, yshift=-1cm]
-                                                                 {};
-  \node (onDestroyBlock)    [process, below of=onStopBlock, yshift=-1cm] 
-                                                              {onDestroy()};
-  \node (onRestartBlock)    [process, right of=onStartBlock, xshift=4cm]
-                                                              {onRestart()};
-  \node (ActivityEnds)      [startstop, left of=activityRuns, xshift=-4cm]
-                                                        {Process is killed};
-  \node (ActivityDestroyed) [startstop, below of=onDestroyBlock]
-                                                    {Activity is shut down};     
+  % \node (onStopBlock)       [process, below of=onPauseBlock, yshift=-1cm]
+  %                                                                {};
+  % \node (onDestroyBlock)    [process, below of=onStopBlock, yshift=-1cm] 
+  %                                                             {onDestroy()};
+  \node (onRestartBlock)    [process, right of=activityRuns, xshift=3cm]
+                                                              {PPS hijacks \\ random number \\ generator};
+  \node (ActivityEnds)      [startstop, below of=onRestartBlock, yshift=-1cm]
+                                                        {Result summaries \\ path probabilities};
+  % \node (ActivityDestroyed) [startstop, below of=onDestroyBlock]
+  %                                                   {Activity is shut down};     
   % Specification of lines between nodes specified above
@@
   \draw[->]     (SimulatorSystem) -- (onStartBlock);
-  \draw[->]      (onStartBlock) -- (onResumeBlock);
-  \draw[->]     (onResumeBlock) -- (activityRuns);
-  \draw[->]      (activityRuns) -- node[text width=4cm]
-                                   {Another activity comes in
-                                    front of the activity} (onPauseBlock);
-  \draw[->]      (onPauseBlock) -- node {The activity is no longer visible}
-                                   (onStopBlock);
-  \draw[->]       (onStopBlock) -- node {The activity is shut down by
-                                   user or system} (onDestroyBlock);
-  \draw[->]    (onRestartBlock) -- (onStartBlock);
-  \draw[->]       (onStopBlock) -| node[yshift=1.25cm, text width=3cm]
-                                   {The activity comes to the foreground}
-                                   (onRestartBlock);
-  \draw[->]    (onDestroyBlock) -- (ActivityDestroyed);
-  \draw[->]      (onPauseBlock) -| node(priorityXMemory)
-                                   {todo}
-                                   (ActivityEnds);
-  \draw           (onStopBlock) -| (priorityXMemory);
-  \draw[->]     (ActivityEnds)  |- node [yshift=-2cm, text width=3.1cm]
-                                    {User navigates back to the activity}
-                                    (SimulatorSystem);
-  \draw[->] (onPauseBlock.east) -- ++(2.6,0) -- ++(0,2) -- ++(0,2) --                
-     node[xshift=1.2cm,yshift=-1.5cm, text width=2.5cm]
-     {The activity comes to the foreground}(onResumeBlock.east);
+  \draw[dashed, <-]  (SimulatorSystem) -- (randomprim);
+  % \draw[->]      (onStartBlock) -- (onResumeBlock);
+  \draw[->]     (onStartBlock) -- (activityRuns);
+  \draw[->]      (activityRuns) -- (onPauseBlock);
+  % \draw[->]      (onPauseBlock) -- node {The activity is no longer visible}
+  %                                  (onStopBlock);
+  % \draw[->]       (onStopBlock) -- node {The activity is shut down by
+  %                                  user or system} (onDestroyBlock);
+  % \draw[->]    (onRestartBlock) -- (activityRuns);
+  \path[->, shift left=.75ex]
+  (onRestartBlock) edge (activityRuns)
+  (activityRuns) edge (onRestartBlock);
+  % \draw[<-]    (activityRuns) -- (onRestartBlock);
+  % \draw[->]       (onStopBlock) -| node[yshift=1.25cm, text width=3cm]
+  %                                  {The activity comes to the foreground}
+  %                                  (onRestartBlock);
+
+  \draw[->]    (onRestartBlock) -- (ActivityEnds);
+  % \draw[->]      (onPauseBlock) -| node(priorityXMemory)
+  %                                  {todo}
+  %                                  (ActivityEnds);
+  % \draw           (onStopBlock) -| (priorityXMemory);
+  % \draw[->]     (ActivityEnds)  |- node [yshift=-2cm, text width=3.1cm]
+  %                                   {User navigates back to the activity}
+  %                                   (SimulatorSystem);
+  % \draw[->] (onPauseBlock.east) -- ++(2.6,0) -- ++(0,2) -- ++(0,2) --                
+  %    node[xshift=1.2cm,yshift=-1.5cm, text width=2.5cm]
+  %    {The activity comes to the foreground}(onResumeBlock.east);
   \end{tikzpicture}
+  \label{fig:how}
+  \caption{TODO}
+  \end{figure}
 As stated previously, Simulators are complex systems and should be viewed as black-boxes, thus deciphering the underlying processes occurring within the simulator facilitates the users understanding of how such a system operates and provides a way of interpreting the outputs in a clear way.

### 2019-04-26 paper/icml_ai4good/main.tex

@@
 However, the complicated and stochastic nature of these simulators
-can mean they are difficult to work with and extract the key information
-required by policy makers.
-In this paper, we introduce an approach to \emph{hijack} these
-simulators with an external
-probabilistic programming system (PPS) that controls
-their internal random number generator.
-This allows for various advanced statistical analyses to be performed
-on the simulators in an automated fashion, while also tracking
-the behavior of the simulator to uncover key information such
-as stochastic structure and generate result summaries.
-Critically, this hijacking can be performed with only minimal changes
-to the original simulator, allowing for straightforward application
-of our approach to a wide variety of contexts and existing simulators.
+can mean their output is difficult to interpret, which reduces their usefulness to policymakers.
+In this paper, we introduce an approach that allows one to treat a large class of population-based epidemiology simulators as probabilistic generative models.
+This is achieved by \emph{hijacking} the internal random number generator calls, through the use of an universal probabilistic programming system (PPS).
+In contrast to other methods, our approach requires only minimal changes to existing simulators to be made.
+We demonstrate that our method can be used for interpretable introspection of and inference on what used to be de facto black box simulators.
+This reinstates much needed trust between policymakers and evidence-based methods.
+
+%This allows for various advanced statistical analyses to be performed
+%on the simulators in an automated fashion, while also tracking
+%the behavior of the simulator to uncover key information such
+%as stochastic structure and generate result summaries.
+%Critically, this hijacking can be performed with only minimal changes
+%to the original simulator, allowing for straightforward application
+%of our approach to a wide variety of contexts and existing simulators.
 %

### 2019-04-26 paper/icml_ai4good/main.tex
fix after first feedback

--- 2019-04-26 paper/icml_ai4good/main.tex
@@
 This is achieved by \emph{hijacking} the internal random number generator calls, through the use of an universal probabilistic programming system (PPS).
-In contrast to other methods, our approach requires only minimal changes to existing simulator to be made.
-We demonstrate that our method can be used for interpretable introspection of and inference on what used to be de facto black box simulators.
+In contrast to other methods, our approach can be easily retrofitted to simulators written in conventional industrial programming frameworks.
+We demonstrate that our method can be used for interpretable introspection and inference, thus shedding light on black-box simulators  .
 This reinstates much needed trust between policymakers and evidence-based methods.
@@
 
-In this paper, we show how probabilistic programming
-techniques CITATIONS can be used to help automate the required calculations to do this, interpreting the simulators as 
-generative models to perform principled statistical analysis, while
-providing insights and summary statistics on their behavior to users.
+In this paper, we introduce a novel method that allows to shed light on the inner workings of a large class of population-based stochastic simulators.
+We achieve this by interpreting such simulators as probabilistic generative models within the framework of universal probabilistic programming (UPP)\cite{TODO}.
 
-Achieving this aim is substantially complicated by the fact that
-a vast array of different epidemiology simulators are currently 
-used in practice.
-Manually reimplementing these systems in existing
-probabilistic programming systems (PPSs) is therefore infeasible, as each
-system would require manual and laborious porting.  Instead, we 
-take the novel approach of \emph{hijacking} existing simulators by
-overriding their internal random number generator, building
-on ideas from~\citep{baydin2018efficient}.  Specifically, by
-replacing the existing low-level random number generator in a simulator
-with a call to a purpose-built ``controller'' PPS, the stochasticity
-of the simulator can be both manipulated and tracked by this controller
-system.  In turn, this allows for a variety of tasks to be performed on
-the hijacked simulator, such as running inference (by conditioning
-the values of certain draws and manipulating others),
-uncovering stochastic structure, and automatically
-producing result summaries, such as establishing the probability
-of different program paths.  By providing a common abstraction 
-framework for different simulators, our approach further allows for
-easy and direct comparison between related or competing
-simulators, a characteristic that is valuable in the context of
-epidemiology simulators CITATIONS.
+Our approach is based on \cite{TODO} and unique in that it can easily be retrofit to simulators written in common industrial programming frameworks.
+This is crucial as, given the enormous code size and complexity, rewriting epidemiology simulators using a dedicated universal probabilistic programming language, such as Pyro\cite{}, is often infeasible.
 
-Critically, this hijacking can be performed on a wide variety of simulators
-with only minimal changes.  Our framework already
-supports 13 popular languages for the base simulator, and can easily
-be extended to others if needed.  Moreover, though our focus
-is on hijacking epidemiology simulators, our general approach and
-framework can be applied more widely to a general class of stochastic
-simulators.  In time, we hope our approach will play a critical role in
+Our framework already supports application to simulators written in 13 industrial programming frameworks, and is easily extensible. 
+Moreover, though our focus is on hijacking epidemiology simulators, our general approach and framework can be applied more widely to a general class of stochastic simulators.  
+In time, we hope our approach will play a critical role in
 bringing recent advancements in probabilistic programming to bear on
@@
 thereby providing wide-ranging impacts across a number of fields.
+
+This paper first gives an overview of existing Malaria simulators (section \ref{TODO}), and proceed by introducing the necessary background on the \textit{pyprob} framework and the concept of universal probabilistic programming (section \ref{TODO}).
+Our approach is then demonstrated and analysed in the context of a Malaria case study (section \ref{TODO}).
+
+that allows one to treat a large class of population-based epidemiology simulators as probabilistic generative models.
+This is achieved by \emph{hijacking} the internal random number generator calls, through the use of an universal probabilistic programming system (PPS).
+In contrast to other methods, our approach requires only minimal changes to existing simulators to be made.
+We demonstrate that our method can be used for interpretable introspection of and inference on what used to be de facto black box simulators.
+This reinstates much needed trust between policymakers and evidence-based methods.
+
+%In this paper, we show how probabilistic programming
+%techniques CITATIONS can be used to help automate the required calculations to do this, interpreting the simulators as 
+%generative models to perform principled statistical analysis, while
+%providing insights and summary statistics on their behavior to users.
+
+%Achieving this aim is substantially complicated by the fact that
+%a vast array of different epidemiology simulators are currently 
+%used in practice.
+%Manually reimplementing these systems in existing
+%probabilistic programming systems (PPSs) is therefore infeasible, as each
+%system would require manual and laborious porting.  Instead, we 
+%take the novel approach of \emph{hijacking} existing simulators by
+%overriding their internal random number generator, building
+%on ideas from~\citep{baydin2018efficient}.  Specifically, by
+%replacing the existing low-level random number generator in a simulator
+%with a call to a purpose-built ``controller'' PPS, the stochasticity
+%of the simulator can be both manipulated and tracked by this controller
+%system.  In turn, this allows for a variety of tasks to be performed on
+%the hijacked simulator, such as running inference (by conditioning
+%the values of certain draws and manipulating others),
+%uncovering stochastic structure, and automatically
+%producing result summaries, such as establishing the probability
+%of different program paths.  By providing a common abstraction 
+%framework for different simulators, our approach further allows for
+%easy and direct comparison between related or competing
+%simulators, a characteristic that is valuable in the context of
+%epidemiology simulators CITATIONS.
+
+%Critically, this hijacking can be performed on a wide variety of simulators
+%with only minimal changes.  Our framework already
+%supports 13 popular languages for the base simulator, and can easily
+%be extended to others if needed.  Moreover, though our focus
+%is on hijacking epidemiology simulators, our general approach and
+%framework can be applied more widely to a general class of stochastic
+%simulators.  In time, we hope our approach will play a critical role in
+%bringing recent advancements in probabilistic programming to bear on
+%the vast array of existing simulators used throughout the sciences,
+%thereby providing wide-ranging impacts across a number of fields.
 

### 2019-04-26 paper/icml_ai4good/main.tex
finished abstract, intro and section 2

@@
 This is achieved by \emph{hijacking} the internal random number generator calls, through the use of an universal probabilistic programming system (PPS).
-In contrast to other methods, our approach can be easily retrofitted to simulators written in conventional industrial programming frameworks.
-We demonstrate that our method can be used for interpretable introspection and inference, thus shedding light on black-box simulators  .
+In contrast to other methods, our approach can be easily retrofitted to simulators written in popular industrial programming frameworks.
+We demonstrate that our method can be used for interpretable introspection and inference, thus shedding light on black-box simulators.
 This reinstates much needed trust between policymakers and evidence-based methods.
@@
 
-In this paper, we introduce a novel method that allows to shed light on the inner workings of a large class of population-based stochastic simulators.
-We achieve this by interpreting such simulators as probabilistic generative models within the framework of universal probabilistic programming (UPP)\cite{TODO}.
+In this paper, we introduce a novel method that allows one to shed light on the inner workings of a large class of population-based stochastic simulators.
+We achieve this by extending the work of \citep{TODO} by interpreting such population-based simulators as probabilistic generative models within the framework of universal probabilistic programming (UPP)\cite{TODO}.
+To this end, we \emph{hijack} existing simulators by overriding their internal random number generators.
+Specifically, by replacing the existing low-level random number generator in a simulator with a call to a purpose-built UPP ``controller'', which can thus control, track and manipulate the stochasticity of the simulator.
 
-Our approach is based on \cite{TODO} and unique in that it can easily be retrofit to simulators written in common industrial programming frameworks.
-This is crucial as, given the enormous code size and complexity, rewriting epidemiology simulators using a dedicated universal probabilistic programming language, such as Pyro\cite{}, is often infeasible.
+This allows for a variety of tasks to be performed on
+the hijacked simulator, such as running inference (by conditioning
+the values of certain draws and manipulating others),
+uncovering stochastic structure, and automatically
+producing result summaries, such as establishing the probability
+of different program paths/traces.  By providing a common abstraction 
+framework for different simulators, our approach further allows for
+easy and direct comparison between related or competing
+simulators, a characteristic that is valuable in the context of
+epidemiology simulators \cite{TODO}. We provide a case study of the above in Section \ref{sec:casestudy}.
 
 Our framework already supports application to simulators written in 13 industrial programming frameworks, and is easily extensible.
-Moreover, though our focus is on hijacking epidemiology simulators, our general approach and framework can be applied more widely to a general class of stochastic simulators.  
+This is crucial as, given the enormous code size and complexity, rewriting epidemiology simulators using a dedicated universal probabilistic programming language, such as Pyro\cite{}, is often infeasible.
+%Moreover, though our focus is on hijacking epidemiology simulators, our general approach and framework can be applied more widely to a general class of stochastic simulators.
+
 In time, we hope our approach will play a critical role in
@@
 
-This paper first gives an overview of existing Malaria simulators (section \ref{TODO}), and proceed by introducing the necessary background on the \textit{pyprob} framework and the concept of universal probabilistic programming (section \ref{TODO}).
-Our approach is then demonstrated and analysed in the context of a Malaria case study (section \ref{TODO}).
+This paper first gives an overview of existing Malaria simulators (Section \ref{sec:background}), and proceed by introducing the necessary background on the \textit{pyprob} framework and the concept of universal probabilistic programming (Section \ref{sec:hijackingsimulators}).
+Our approach is then demonstrated and analysed in the context of a Malaria case study (Section \ref{sec:casestudy}).
 
-that allows one to treat a large class of population-based epidemiology simulators as probabilistic generative models.
+%Our approach is extending recent work by  and is unique in that it can easily be retrofitted to population-based simulators written in popular industrial programming frameworks.
+This is crucial as, given the enormous code size and complexity, rewriting epidemiology simulators using a dedicated universal probabilistic programming language, such as Pyro\cite{}, is often infeasible.
+
+%that allows one to treat a large class of population-based epidemiology simulators as probabilistic generative models.
 This is achieved by \emph{hijacking} the internal random number generator calls, through the use of an universal probabilistic programming system (PPS).
 In contrast to other methods, our approach requires only minimal changes to existing simulators to be made.
-We demonstrate that our method can be used for interpretable introspection of and inference on what used to be de facto black box simulators.
+%We demonstrate that our method can be used for interpretable introspection of and inference on what used to be de facto black box simulators.
 This reinstates much needed trust between policymakers and evidence-based methods.
@@
 Recent advances in hardware have enabled simulations to model the dynamics of infectious diseases, such as Malaria, in ever greater detail.
+
 Two the most advanced Malaria simulators, namely EMOD~\cite{bershteyn2018implementation} and OpenMalaria\cite{smith2008towards}, have proven to be particularly valuable to policymakers.

### 2019-04-26 paper/icml_ai4good/main.tex

@@
 Recent advances in hardware have enabled simulations to model the dynamics of infectious diseases, such as Malaria, in ever greater detail.
+
+\subsection{Epidemiology Simulators}
 

### 2019-04-26 paper/icml_ai4good/main.tex
dot to svg

@@
 We provide examples of the generated trace plots from the connection between the simulator and the PPS in Figure~\ref{fig:plotewan}. 
-When comparing the two trace plots between EMOD and OgpenMalaria we see significant difference, despite both evaluating the same model for the same disease. 
+When comparing the two trace plots between EMOD and OgpenMalaria we see significant differences, despite both evaluating the same model for the same disease. 
 We can see from the addressing schemes $A_1, \ldots, A_N$ \bg{ref table 1 and 2} what physical events are connected to each other and how outputs in EMOD are generated from a different set of procedures as compared to OpenMalaria
@@
 
-\begin{figure}
-\begin{tikzpicture}
-\begin{axis}[
-    ybar,
-    enlargelimits=0.15,
-    legend style={at={(0.5,-0.15)},
-      anchor=north,legend columns=-1},
-      xlabel = {Address},
-    ylabel={Probabilities},
-    symbolic x coords={Trace 1, Trace 2, Trace 3},
-    xtick=data,
-    nodes near coords,
-    nodes near coords align={vertical},
-    ]
-% \addplot coordinates {(tool8,7) (tool9,9) (tool10,4)};
-\addplot coordinates {(Trace 1,0.2) (Trace 2,0.5) (Trace 3,0.23)};
-% \addplot coordinates {(A1,0.2) (A2,0.5) (A3,0.7) (A4,0.9)};
-% \addplot coordinates {(A1,0.1) (A2,0.1) (A3,0.1) (A4,0.3)};
-% \legend{used,understood,not understood}
-\end{axis}
-\end{tikzpicture}
-\caption{Trace probabilities. Trace 1 corresponds to A1,A1,A4,... 
-Trace 2 corresponds to A1,A1,A1,...A2,... These still need correctly calculating.}
-\end{figure}
+%\begin{figure}
+%\begin{tikzpicture}
+%\begin{axis}[
+%    ybar,
+%    enlargelimits=0.15,
+%    legend style={at={(0.5,-0.15)},
+%      anchor=north,legend columns=-1},
+%      xlabel = {Address},
+%    ylabel={Probabilities},
+%    symbolic x coords={Trace 1, Trace 2, Trace 3},
+%    xtick=data,
+%    nodes near coords,
+%    nodes near coords align={vertical},
+%    ]
+%% \addplot coordinates {(tool8,7) (tool9,9) (tool10,4)};
+%\addplot coordinates {(Trace 1,0.2) (Trace 2,0.5) (Trace 3,0.23)};
+%% \addplot coordinates {(A1,0.2) (A2,0.5) (A3,0.7) (A4,0.9)};
+%% \addplot coordinates {(A1,0.1) (A2,0.1) (A3,0.1) (A4,0.3)};
+%% \legend{used,understood,not understood}
+%\end{axis}
+%\end{tikzpicture}
+%\caption{Trace probabilities. Trace 1 corresponds to A1,A1,A4,... 
+%Trace 2 corresponds to A1,A1,A1,...A2,... These still need correctly calculating.}
+%\end{figure}
 \begin{figure*}[h!]

### 2019-04-26 paper/icml_ai4good/main.tex
fixed emod trace graph
updated

@@
 \section{Introduction}
-Ending the epidemics of AIDS, tuberculosis, malaria and other infectious diseases by 2030 is a key target within the Good Health \& Well-Being section of the UN Sustainable Development Goals\todo{These citations
-should still have years} \cite{refugees_2030_nodate,un_sustainable_2018}. 
+Ending the epidemics of AIDS, tuberculosis, malaria and other infectious diseases by 2030 is a key target within the Good Health \& Well-Being section of the UN Sustainable Development Goals \cite{refugees_2030,un_sustainable2018}.
 However, despite decades of substantial international efforts, these diseases kill hundreds of million people a year.
@@
 before they are rolled out. This frequently increases the effectiveness of interventions and thus ultimately saves resources, or even lifes.
-For example, it has been shown that mass vaccination may be largely ineffective in regions of large transmission rates, but may play a crucial role in areas of low transmission \bg{citation?}.
+For example, it has been shown that mass vaccination may be largely ineffective in regions of large transmission rates, but may play a crucial role in areas of low transmission \cite{cameron2015defining}.
 
 Malaria epidemiology is governed by a complex set of drivers, 
-few of which can be understood in isolation CITATION.
+few of which can be understood in isolation \cite{cameron2015defining,autino_epidemiology_2012,smith2008towards,bershteyn2018implementation}.
 These include within-host dynamics, population-specific traits and even local geography.
@@
 Computational epidemiology simulators have to reflect these complexities and are usually stochastic in nature.
-This can make simulation output highly non-trivial to interpret, particularly when trying to draw desired inferences coupled with observed data \cite{TODO}.
+This can make simulation output highly non-trivial to interpret, particularly when trying to draw desired inferences coupled with observed data \cite{mwendera_challenges_2019,ferris_openmalaria_2015}.
 
 In this paper, we introduce a novel method that allows one to shed light on the inner workings of a large class of population-based stochastic simulators.
-We achieve this by extending the work of \citep{TODO} by interpreting such population-based simulators as probabilistic generative models within the framework of universal probabilistic programming (UPP)\cite{TODO}.
+We achieve this by extending the work of \citep{baydin2018efficient} by interpreting such population-based simulators as probabilistic generative models within the framework of universal probabilistic programming (UPP)\cite{le-2016-inference}.
 To this end, we \emph{hijack} existing simulators by overriding their internal random number generators.
@@
 simulators, a characteristic that is valuable in the context of
-epidemiology simulators \cite{TODO}. We provide a case study of the above in Section \ref{sec:casestudy}.
+epidemiology simulators \cite{ferris_openmalaria_2015}. We provide a case study of the above in Section \ref{sec:casestudy}.
 
 Our framework already supports application to simulators written in 13 industrial programming frameworks, and is easily extensible.
-This is crucial as, given the enormous code size and complexity, rewriting epidemiology simulators using a dedicated universal probabilistic programming language, such as Pyro\cite{}, is often infeasible.
+This is crucial as, given the enormous code size and complexity, rewriting epidemiology simulators using a dedicated universal probabilistic programming language, such as Pyro\cite{bingham2019pyro}, is often infeasible.
 %Moreover, though our focus is on hijacking epidemiology simulators, our general approach and framework can be applied more widely to a general class of stochastic simulators.
@@
   \label{fig:how}
-  \caption{TODO}
+  \caption{This flow-chart provides an overview of the process of how our system hijacks a generic population-based epidemiology simulator, such as EMOD, and how we modify the simulator to be able 
+  to hijack the random number generator (RNG).}
   \end{figure}
@@
 We provide examples of the generated trace plots from the connection between the simulator and the PPS in Figure~\ref{fig:plotewan}. 
-When comparing the two trace plots between EMOD and OgpenMalaria we see significant differences, despite both evaluating the same model for the same disease. 
+When comparing the two trace plots between EMOD and OgpenMalaria we see significant difference, despite both evaluating the same model for the same disease. 
 We can see from the addressing schemes $A_1, \ldots, A_N$ \bg{ref table 1 and 2} what physical events are connected to each other and how outputs in EMOD are generated from a different set of procedures as compared to OpenMalaria
@@
 
-%\begin{figure}
-%\begin{tikzpicture}
-%\begin{axis}[
+% \begin{figure}
+% \begin{tikzpicture}
+% \begin{axis}[
 %     ybar,
@@
 %     ]
-%% \addplot coordinates {(tool8,7) (tool9,9) (tool10,4)};
-%\addplot coordinates {(Trace 1,0.2) (Trace 2,0.5) (Trace 3,0.23)};
-%% \addplot coordinates {(A1,0.2) (A2,0.5) (A3,0.7) (A4,0.9)};
-%% \addplot coordinates {(A1,0.1) (A2,0.1) (A3,0.1) (A4,0.3)};
-%% \legend{used,understood,not understood}
-%\end{axis}
-%\end{tikzpicture}
-%\caption{Trace probabilities. Trace 1 corresponds to A1,A1,A4,... 
-%Trace 2 corresponds to A1,A1,A1,...A2,... These still need correctly calculating.}
-%\end{figure}
+% % \addplot coordinates {(tool8,7) (tool9,9) (tool10,4)};
+% \addplot coordinates {(Trace 1,0.2) (Trace 2,0.5) (Trace 3,0.23)};
+% % \addplot coordinates {(A1,0.2) (A2,0.5) (A3,0.7) (A4,0.9)};
+% % \addplot coordinates {(A1,0.1) (A2,0.1) (A3,0.1) (A4,0.3)};
+% % \legend{used,understood,not understood}
+% \end{axis}
+% \end{tikzpicture}
+% \caption{Trace probabilities. Trace 1 corresponds to A1,A1,A4,... 
+% Trace 2 corresponds to A1,A1,A1,...A2,... These still need correctly calculating.}
+% \end{figure}
 \begin{figure*}[h!]
@@
   \setlength{\tabcolsep}{1mm}
-  \caption{A few addresses generated for the OpenMalaria simulator, see the Supplement for the full table.}
+  \caption{An example of an address generated for the OpenMalaria simulator. We can see that address 1
+  relates to Generating a member of the human population who may or may not be infect with the Malaria disease.
+We get something similar for EMOD, except this relates to Address 7 in the EMOD program execution.}
   \label{table:addresses}
@@
   % \setlength{\tabcolsep}{1mm}
-  \begin{tabularx}{\textwidth}{@{}lX@{}} 
+  \begin{tabularx}{\textwidth}{@{}lX@{}l@{}} 
     \toprule
-    Address ID & Full address \\
+    Address ID & Full address & Interpretation\\
     \midrule
-  A1 & [forward()+0x204; OM::Simulator::start(scnXml::Monitoring const)+0x28a;
+  A1 & [forward()+0x204; OM::Simulator::
+
+  start(scnXml::Monitoring const)+0x28a;
 
@@
 
-  OM::WithinHost::DescriptiveWithinHostModel::DescriptiveWithinHostModel(double)+0x3a;
+  OM::WithinHost::DescriptiveWithinHostModel
+  ::DescriptiveWithinHostModel(double)+0x3a;
 
@@
 
-  OM::util::random::gauss(double, double)+0xb4]\_\_Normal\\
+  OM::util::random::gauss(double, double)+0xb4]\_\_Normal &  Generate a human in the population within host dynamics \\
 
-   $\vdots$ & $\vdots$ \\
-  A5 & [forward()+0x204; OM::Simulator::start(scnXml::Monitoring const\&)+0x468;
+ %   $\vdots$ & $\vdots$ \\
+ %  A5 & [forward()+0x204; OM::Simulator::start(scnXml::Monitoring const\&)+0x468;
 
-   OM::Population::update1(OM::SimTime)+0xff;
+ %   OM::Population::update1(OM::SimTime)+0xff;
 
-    OM::Host::Human::update(bool)+0x2bc;
+ %    OM::Host::Human::update(bool)+0x2bc;
 
-    OM::Clinical::ClinicalModel::update(OM::Host::Human\&, double, bool)+0x96;
+ %    OM::Clinical::ClinicalModel::update(OM::Host::Human\&, double, bool)+0x96;
 
-    OM::Host::NeonatalMortality::eventNeonatalMortality()+0x9;
+ %    OM::Host::NeonatalMortality::eventNeonatalMortality()+0x9;
 
-    OM::util::random::uniform\_01()+0xc0]\_\_Uniform\\
+ %    OM::util::random::uniform\_01()+0xc0]\_\_Uniform\\
 
- $\vdots$ & $\vdots$ \\
+ % $\vdots$ & $\vdots$ \\
 

### 2019-04-26 paper/icml_ai4good/main.tex

@@
 
-% \begin{figure}
-% \begin{tikzpicture}
-% \begin{axis}[
+%\begin{figure}
+%\begin{tikzpicture}
+%\begin{axis}[
 %    ybar,
@@
 %    ]
-% % \addplot coordinates {(tool8,7) (tool9,9) (tool10,4)};
-% \addplot coordinates {(Trace 1,0.2) (Trace 2,0.5) (Trace 3,0.23)};
-% % \addplot coordinates {(A1,0.2) (A2,0.5) (A3,0.7) (A4,0.9)};
-% % \addplot coordinates {(A1,0.1) (A2,0.1) (A3,0.1) (A4,0.3)};
-% % \legend{used,understood,not understood}
-% \end{axis}
-% \end{tikzpicture}
-% \caption{Trace probabilities. Trace 1 corresponds to A1,A1,A4,... 
-% Trace 2 corresponds to A1,A1,A1,...A2,... These still need correctly calculating.}
-% \end{figure}
+%% \addplot coordinates {(tool8,7) (tool9,9) (tool10,4)};
+%\addplot coordinates {(Trace 1,0.2) (Trace 2,0.5) (Trace 3,0.23)};
+%% \addplot coordinates {(A1,0.2) (A2,0.5) (A3,0.7) (A4,0.9)};
+%% \addplot coordinates {(A1,0.1) (A2,0.1) (A3,0.1) (A4,0.3)};
+%% \legend{used,understood,not understood}
+%\end{axis}
+%\end{tikzpicture}
+%\caption{Trace probabilities. Trace 1 corresponds to A1,A1,A4,... 
+%Trace 2 corresponds to A1,A1,A1,...A2,... These still need correctly calculating.}
+%\end{figure}
+
 \begin{figure*}[h!]
 	\centering
-	\includegraphics[width=\textwidth]{../plots/emod_24_prior_pop_100.pdf}
+	\includegraphics[width=\textwidth]{../plots/final_emod_traces__rotated.pdf}
 	\includegraphics[width=\textwidth]{../plots/ewan_25_pop_100.pdf}

### 2019-04-26 paper/icml_ai4good/main.tex
updated

@@
   \label{fig:how}
-  \caption{This flow-chart provides an overview of the process of how our system hijacks a generic population-based epidemiology simulator, such as EMOD, and how we modify the simulator to be able 
+  \caption{This flow-chart provides an overview of the process of how our system hijacks a generic population-based epidemiology simulator, such as EMOD, and how we modify the simulator to 
   to hijack the random number generator (RNG).}

### 2019-04-26 paper/icml_ai4good/main.tex
updated
section 3 bckp

@@
 \label{sec:hijackingsimulators}
- Our method of overriding simulators builds on the work of~\cite{baydin2018efficient} and extends the framework to encapsulate a more diverse range of simulators, in particular population based simulators such as those found in epidemiology.
-Our framework provides a simple solution to do this and makes it easy to transform arbitrary stochastic simulators into probabilistic programs, regardless of the complexity of the simulator and the language that the simulator is written in.
-In particular, the framework currently supports 13 popular languages, but there is nothing to stop this being extended to any language of interest. 
-Furthermore, we can understand the structure of simulators as programs for both decoding \emph{black-box structures} and for posterior inference, which enables two things. 
-1) It enables software developers and policy makers to understand complicated code bases and dissect the outputs of simulators for guiding policy decision making.
- 2) It provides interpretable inference results that provide policy makers with predictions that are truly interpretable, in the sense that the end-user of the inference results understands what physical events led to the inference outcome.
- This is critical in several domains, especially in epidemiology. 
- We provide examples of (1) in next Section~\ref{sec:casestudy} and discuss the latter in future work as extensions to the the current framework.
+
+ Probabilistic programming~\cite{gordon2014probabilistic,staton2016semantics,kozen1979semantics} 
+can be used to express probabilistic models and consequently perform automated inference thereon.
+Once a probabilistic model has been expressed in a probabilistic programming language, a wide range of inference techniques, such as Markov Chain Monte Carlo (MCMC)\cite{}, Variational Inference~(VI)\cite{} and Deep Neural Networks~(DNN)\cite{},  can be used out-of-the-box by non-experts.
+
+\textit{Hijacking a simulator} describes the process by which a simulator's random number generators are replaced by calls to external sampling procedures which are controlled by a probabilistic programming system (PPS).
+In practice, this amounts to performing a small number of surgical incisions into the simulator's source code in order to replace built-in calls to random number generators.
+E.g., given a simulator written in C++/Boost\cite{}, a Gaussian distribution object \textit{boost::normal} is replaced with the corresponding \textit{pyprob\_cpp} distribution object, namely \textit{pyprob\_cpp::distributions::Normal}.
+Sampling from this distribution is then done by requesting the PPS to send a sample back to the simulator.
+
+The PPS and the simulator exchange xTensor objects \cite{} through TCP or ICP using a generic FlatBuffers \cite{TODO} protocol.
+On the PPS side, sampling is done using the deep learning framework PyTorch \cite{}.
+
+After a variable has been sampled, it is sent to the PPS using \textit{pyprob\_cpp::sample}, alongside with its simulator source trace.
+This allows the PPS to construct sample trace probabilities and other summary statistics (cf.
+Section \ref{sec:casestudy}).
+
+%
+%As stated previously, Simulators are complex systems and should be viewed as black-boxes, thus deciphering the underlying processes occurring within the simulator facilitates the users understanding of how such a system operates and provides a way of interpreting the outputs in a clear way.
+In addition to this, when using a simulator to make new inferences a user will often want to condition on the output of the simulator, or any new observations gathered.
+In the standard simulator framework this is not possible, but in
+%
+%Essentially, from the users perspective, our approach can be viewed as follows the user would first select the simulator that they are going to use, for example EMOD, they would then write their model, in the desired simulator, according to the given simulator specification.
+The user would then run the simulator.
+Now comes the hijacking.
+During the forward run of the simulator all sampling calls from the random number generator are directed through the probabilistic programming system~(PPS), which in addition to the standard output of the simulator now generates a series of result summaries and path probabilities, as we demonstrate in Section~\ref{sec:casestudy}.
+%
+%By connecting with the PPS we track all sampling procedures, which enables us to control the myriad of random variables generated within the simulator. 
+%
+%In order to overhaul a stochastic simulator like EMOD, or OpenMalaria, we are only required to override the location of the stochastic primitives, i.e.
+the base operations that generate the stochasticity and randomness within the simulator.
+%We will walk through how this is done for a standard epidemiology simulator, see the flowchart for a pictorial representation of what is happening \bg{add ref to flow chart}.
+%
+%The initial step first requires one to build a containerized environment, such as Docker~\cite{merkel2014docker} and Singularity~\cite{kurtzer2017singularity}, which enables the simulators to be run on any device, independent of different hardware architectures, removing the requirement for hardware specific machines.
+%The second step requires one to override the stochastic primitives, which requires re-writing aspects of the sampling procedures inside the simulator. 
+%This requires zero domain-expertise and only requires one to find where the sampling calls are originated, this can be done with any basic editor search function.
+% Such information is typically saved in $\texttt{RANDOM.<filetype>}$, although it may vary slightly between simulators. We now demonstrate this for a uniform primitive.
+%
+%The third step is generating the protocols, which enable communication between the simulator and probabilistic programming system. 
+%For this we extend the existing bridging framework introduced in~\cite{baydin2018efficient}, which is comprised of the Probabilistic Programming eXecution~(PPX) protocols and a light-weight C ++ interface, however the language of this light-weight interface is dependent on whatever the file type of the $\texttt{RANDOM.<filetype>}$ file is.
+%For both OpenMalaria and EMOD, the file is written in C ++.
+%In using these protocols a larger class of sampling procedures can be overridden.
+The PPX protocols work by making use of the flatbuffer protocols\footnote{\url{https://github.com/google/flatbuffers}}, which require a flatbuffers script to be specified, see here for an example of a flatbuffers script\footnote{https://github.com/probprog/ppx/blob/master/ppx.fbs}.
+Once this is specified the flat buffers script is compiled and automatically constructs the message passing framework between the simulator and the probabilistic programming system.
+The flatbuffers script is simple, clean and intuitive, making it easy for a non-expert to produced.
+%The next steps requires modifying the $\texttt{main}$ function in $\texttt{main.cpp}$ and defining an additional forward function which runs one execution of the simulator to enable message passing between the simulator and PPS\footnote{A link to an example will be provided after the review period to preserve anonymity}.
+%The final step is to create a light-weight interface.
+As both the EMOD and OpenMalaria are solely based on C ++ we can extend an existing interface called pyprob\_cpp\footnote{\url{https://github.com/probprog/pyprob_cpp}} to account for the additional sampling procedures.
+%Once these steps are complete, the containerized environment will now contain everything required to pass models through to the simulators, encode the simulator into a probabilistic program, which can then be evaluated in the PPS.
+
+ %Our method of overriding simulators builds on the work of~\cite{baydin2018efficient} and extends the framework to encapsulate a more diverse range of simulators, in particular population based simulators such as those found in epidemiology.
+%Our framework provides a simple solution to do this and makes it easy to transform arbitrary stochastic simulators into probabilistic programs, regardless of the complexity of the simulator and the language that the simulator is written in.
+%In particular, the framework currently supports 13 popular languages, but there is nothing to stop this being extended to any language of interest. 
+%Furthermore, we can understand the structure of simulators as programs for both decoding \emph{black-box structures} and for posterior inference, which enables two things. 
+%1) It enables software developers and policy makers to understand complicated code bases and dissect the outputs of simulators for guiding policy decision making.
+ %2) It provides interpretable inference results that provide policy makers with predictions that are truly interpretable, in the sense that the end-user of the inference results understands what physical events led to the inference outcome.
+ %This is critical in several domains, especially in epidemiology. 
+ %We provide examples of (1) in next Section~\ref{sec:casestudy} and discuss the latter in future work as extensions to the the current framework.
 
@@
 % that allow the expression of unrestricted probability models in a Turing complete fashion~\cite{wood2014new,goodman2012church,landau_binder_2014,siddharth2017learning,bingham2019pyro}.
-
-\begin{itemize}
-\item Add what - What are we trying to do to make peoples lifes easier
-\item Add WHy - Why is this important to do and what does it enable.
-\item Then deal with "How" - How is this done - what is currently there. 
-\end{itemize}
+%
+%\begin{itemize}
+%\item Add what - What are we trying to do to make peoples lifes easier
+%\item Add WHy - Why is this important to do and what does it enable.
+%\item Then deal with "How" - How is this done - what is currently there. 
+%\end{itemize}
 \begin{figure}
+\centering
+ \scalebox{.5}{
 \begin{tikzpicture}[node distance=1.5cm,
@@
   \end{tikzpicture}
+  }
   \label{fig:how}
@@
   \end{figure}
-As stated previously, Simulators are complex systems and should be viewed as black-boxes, thus deciphering the underlying processes occurring within the simulator facilitates the users understanding of how such a system operates and provides a way of interpreting the outputs in a clear way.
-In addition to this, when using a simulator to make new inferences a user will often want to condition on the output of the simulator, or any new observations gathered.
-In the standard simulator framework this is not possible, but in
-
-Essentially, from the users perspective, our approach can be viewed as follows the user would first select the simulator that they are going to use, for example EMOD, they would then write their model, in the desired simulator, according to the given simulator specification.
-The user would then run the simulator.
-Now comes the hijacking.
-During the forward run of the simulator all sampling calls from the random number generator are directed through the probabilistic programming system~(PPS), which in addition to the standard output of the simulator now generates a series of result summaries and path probabilities, as we demonstrate in Section~\ref{sec:casestudy}.
-
-By connecting with the PPS we track all sampling procedures, which enables us to control the myriad of random variables generated within the simulator.
-
-In order to overhaul a stochastic simulator like EMOD, or OpenMalaria, we are only required to override the location of the stochastic primitives, i.e.
-the base operations that generate the stochasticity and randomness within the simulator.
-We will walk through how this is done for a standard epidemiology simulator, see the flowchart for a pictorial representation of what is happening \bg{add ref to flow chart}.
-
-The initial step first requires one to build a containerized environment, such as Docker~\cite{merkel2014docker} and Singularity~\cite{kurtzer2017singularity}, which enables the simulators to be run on any device, independent of different hardware architectures, removing the requirement for hardware specific machines.
-This greatly improves the scope of applicability of our system and the simulator, as this environment only has to be built once, which then enables the simulator to be run simultaneously on machines of different architectures machine.
-This step will be slightly different for individual simulators, but the overall idea is the same.
-We cannot provide a link to an example of a containerized environment for OpenMalaria and system in this submission due to anonymity\footnote{We shall provide a link in the camera ready submission}.
-The second step requires one to override the stochastic primitives, which requires re-writing aspects of the sampling procedures inside the simulator. 
-This requires zero domain-expertise and only requires one to find where the sampling calls are originated, this can be done with any basic editor search function.
- Such information is typically saved in $\texttt{RANDOM.<filetype>}$, although it may vary slightly between simulators. We now demonstrate this for a uniform primitive.
 
@@
 % As can be seen, writing the new sampling call is cleaner and syntactically simpler. 
-The third step is generating the protocols, which enable communication between the simulator and probabilistic programming system. 
-For this we extend the existing bridging framework introduced in~\cite{baydin2018efficient}, which is comprised of the Probabilistic Programming eXecution~(PPX) protocols and a light-weight C ++ interface, however the language of this light-weight interface is dependent on whatever the file type of the $\texttt{RANDOM.<filetype>}$ file is.
-For both OpenMalaria and EMOD, the file is written in C ++.
-In using these protocols a larger class of sampling procedures can be overridden.
-The PPX protocols work by making use of the flatbuffer protocols\footnote{\url{https://github.com/google/flatbuffers}}, which require a flatbuffers script to be specified, see here for an example of a flatbuffers script\footnote{https://github.com/probprog/ppx/blob/master/ppx.fbs}.
-Once this is specified the flat buffers script is compiled and automatically constructs the message passing framework between the simulator and the probabilistic programming system.
-The flatbuffers script is simple, clean and intuitive, making it easy for a non-expert to produced.
-The next steps requires modifying the $\texttt{main}$ function in $\texttt{main.cpp}$ and defining an additional forward function which runs one execution of the simulator to enable message passing between the simulator and PPS\footnote{A link to an example will be provided after the review period to preserve anonymity}.
-The final step is to create a light-weight interface.
-As both the EMOD and OpenMalaria are solely based on C ++ we can extend an existing interface called pyprob\_cpp\footnote{\url{https://github.com/probprog/pyprob_cpp}} to account for the additional sampling procedures.
-Once these steps are complete, the containerized environment will now contain everything required to pass models through to the simulators, encode the simulator into a probabilistic program, which can then be evaluated in the PPS.
-
-\bg{Add performance plot showing memory consumptions of pyprob and our new system.}
+%\bg{Add performance plot showing memory consumptions of pyprob and our new system.}
 \section{Case Study}

### 2019-04-26 paper/icml_ai4good/main.tex
updated

--- 2019-04-26 paper/icml_ai4good/main.tex
@@
 
-% Probabilistic programming~\cite{gordon2014probabilistic,staton2016semantics,kozen1979semantics} is the intersection of computer science, statistics and machine learning. 
-% It attempts to unify the three fields by creating automated frameworks in which one can easily extrapolate their observed data and perform sophisticated statistical analysis with advance inference procedures such as
-% Markov Chain Monte Carlo (MCMC), Variational Inference~(VI) and Deep Neural Networks~(DNN) in an accessible way, i.e.
-without the need of expert knowledge in implementing and using such procedures.
-In addition to this, probabilistic programming systems~(PPSs) provide tools that enable the source code of a program to be throughly analysed in an automated fashion.
-% Users can then exploit these algorithms and tools on an array of models
-% within the PPS, using the language itself to create models that represent complicated distribution objects which are then analysed by the PPS. 
-% PPSs are extensions, or reductions, of existing programming languages such as python, c++, go and so forth, which means that users do not have to learn
-% a completely new set of operating semantics and can write programs in the same vain as they are accustomed to.  
-% The PPS that we use, PyProb~\cite{le-2016-inference,baydin2018efficient}, belongs to the family of universal PPSs 
-% that allow the expression of unrestricted probability models in a Turing complete fashion~\cite{wood2014new,goodman2012church,landau_binder_2014,siddharth2017learning,bingham2019pyro}.
-
-\begin{itemize}
-\item Add what - What are we trying to do to make peoples lifes easier
-\item Add WHy - Why is this important to do and what does it enable.
-\item Then deal with "How" - How is this done - what is currently there. 
-\end{itemize}
 \begin{figure}
+\centering
+ \scalebox{.5}{
 \begin{tikzpicture}[node distance=1.5cm,
@@
 
-% \begin{table}[h!]
-% \footnotesize
-% \begin{tabular}{p{3.6cm}|p{3.6cm}}
-% Before  &  After  \\
-% \midrule
-% \begin{lstlisting}
-% #include <gsl/gsl_cdf.h>
-% #include <gsl/gsl_rng.h>
-% #include <gsl/gsl_randist.h>
-% //preamble
-% double random::uniform_01 () {
-%     double result =
-% # ifdef 
-% OM_RANDOM_USE_BOOST
-%        rng_uniform01 ();
-% # else
-%  gsl_rng_uniform (rng.gsl_generator);
-% # endif
-%     return result;
-%     }
-% \end{lstlisting}&
-% \begin{lstlisting}
-% #include <pyprob_cpp.h>
-% #include "xtensor/xadapt.hpp"
-% // preamble
-% double random::uniform_01 () {
-% auto uniform = pyprob_cpp::distributions::Uniform(0,1);
-% return pyprob_cpp::sample(uniform)(0);
-% }
-% \end{lstlisting}
-% \end{tabular} 
-% \end{table}
-% As can be seen, writing the new sampling call is cleaner and syntactically simpler. 
-The third step is generating the protocols, which enable communication between the simulator and probabilistic programming system. 
-For this we extend the existing bridging framework introduced in~\cite{baydin2018efficient}, which is comprised of the Probabilistic Programming eXecution~(PPX) protocols and a light-weight C ++ interface, however the language of this light-weight interface is dependent on whatever the file type of the $\texttt{RANDOM.<filetype>}$ file is.
-For both OpenMalaria and EMOD, the file is written in C ++.
-In using these protocols a larger class of sampling procedures can be overridden.
+% The third step is generating the protocols, which enable communication between the simulator and probabilistic programming system. 
+% For this we extend the existing bridging framework introduced in~\cite{baydin2018efficient}, which is comprised of the Probabilistic Programming eXecution~(PPX) protocols and a light-weight C ++ interface, however the language of this light-weight interface is dependent on whatever the file type of the $\texttt{RANDOM.<filetype>}$ file is.
+% For both OpenMalaria and EMOD, the file is written in C ++.
+% In using these protocols a larger class of sampling procedures can be overridden.
 The PPX protocols work by making use of the flatbuffer protocols\footnote{\url{https://github.com/google/flatbuffers}}, which require a flatbuffers script to be specified, see here for an example of a flatbuffers script\footnote{https://github.com/probprog/ppx/blob/master/ppx.fbs}.
@@
 The flatbuffers script is simple, clean and intuitive, making it easy for a non-expert to produced.
-The next steps requires modifying the $\texttt{main}$ function in $\texttt{main.cpp}$ and defining an additional forward function which runs one execution of the simulator to enable message passing between the simulator and PPS\footnote{A link to an example will be provided after the review period to preserve anonymity}.
-The final step is to create a light-weight interface.
+% The next steps requires modifying the $\texttt{main}$ function in $\texttt{main.cpp}$ and defining an additional forward function which runs one execution of the simulator to enable message passing between the simulator and PPS\footnote{A link to an example will be provided after the review period to preserve anonymity}.
+% The final step is to create a light-weight interface.
 As both the EMOD and OpenMalaria are solely based on C ++ we can extend an existing interface called pyprob\_cpp\footnote{\url{https://github.com/probprog/pyprob_cpp}} to account for the additional sampling procedures.
-Once these steps are complete, the containerized environment will now contain everything required to pass models through to the simulators, encode the simulator into a probabilistic program, which can then be evaluated in the PPS.
+% Once these steps are complete, the containerized environment will now contain everything required to pass models through to the simulators, encode the simulator into a probabilistic program, which can then be evaluated in the PPS.
 
@@
 
- \begin{table*}[h!]
+\begin{table}[h!]
 \footnotesize
   \setlength{\tabcolsep}{1mm}
-  \caption{An example of an address generated for the OpenMalaria simulator. We can see that address 1
+\label{table:asingleaddress}
+  \caption{An example of an address generated for the model run in the OpenMalaria simulator. We can see that A1
   relates to Generating a member of the human population who may or may not be infect with the Malaria disease.
-We get something similar for EMOD, except this relates to Address 7 in the EMOD program execution.}
-  \label{table:addresses}
+We get something similar for EMOD, except this relates to A7 in the EMOD program execution.}
   \def\arraystretch{1.25}
   % \setlength{\tabcolsep}{1mm}
-  \begin{tabularx}{\textwidth}{@{}lX@{}l@{}} 
+  \begin{tabularx}{0.46\textwidth}{@{}lX@{}l@{}} 
     \toprule
-    Address ID & Full address & Interpretation\\
+    Address ID & Full address \\
     \midrule
@@
 
-  OM::WithinHost::WHInterface::createWithinHostModel(double)+0x99;
+  OM::WithinHost::WHInterface::
+
+  createWithinHostModel(double)+0x99;
 
@@
 
-  OM::WithinHost::WHFalciparum::WHFalciparum(double)+0xe6;
+  OM::WithinHost::WHFalciparum::
 
-  OM::util::random::gauss(double, double)+0xb4]\_\_Normal &  Generate a human in the population within host dynamics \\
+  WHFalciparum(double)+0xe6;
+
+  OM::util::random::
+
+  gauss(double, double)+0xb4]\_\_Normal \\
+
+  $\vdots$ & $\vdots$ \\
 
@@
   \end{tabularx}
-  \end{table*}
+  \end{table}
 
+\begin{table}[h!]
+  \footnotesize
+  \setlength{\tabcolsep}{1mm}
+  \caption{An interpretation table for each of the address of the 
+  overall trace generated from the corresponding OpenMalaria model.}
+  \label{table:addresses}
+  % \def\arraystretch{1.25}
+  % \setlength{\tabcolsep}{1mm}
+  \begin{tabularx}{0.46\textwidth}{@{}lX@{}} 
+    \toprule
+    Address ID & Interpretation \\
+    \midrule
+  A1 & Generate a human in the
+
+   population within host dynamics\\
+
+  A2 & Generate another human in
+
+   the population within host dynamics\\
+
+  A3 & The population is updated and
+
+  a new human, or humans, may get infected \\
+
+  A4 & Potential children deaths
+
+  within the population are simulated \\
+  A5 & Potential children deaths within
+
+   the population are simulated \\
+
+  A6 & Determines parasite density
+
+   of an individual infection \\
+
+  A7 & Models how the disease is
+
+  progressing within the infected humans \\
+
+	A8 & Models how the disease is
+
+	 progressing within the population \\
+
+  A9 & Models how the disease is
+
+  progressing within the infected humans
+
+   after the population has been updated\\
+
+  A10 & Full clinical update on
+
+  the population for those without severe
+
+  or no Malaria infection. \\
+
+A11 & Full clinical update on the population
+
+for those with severe Malaria infections \\
+\bottomrule
+  \end{tabularx}
+  \end{table}
 \bibliographystyle{icml2019}

### 2019-04-26 paper/icml_ai4good/main.tex

--- 2019-04-26 paper/icml_ai4good/main.tex
@@
 
-% Probabilistic programming~\cite{gordon2014probabilistic,staton2016semantics,kozen1979semantics} is the intersection of computer science, statistics and machine learning. 
-% It attempts to unify the three fields by creating automated frameworks in which one can easily extrapolate their observed data and perform sophisticated statistical analysis with advance inference procedures such as
-% Markov Chain Monte Carlo (MCMC), Variational Inference~(VI) and Deep Neural Networks~(DNN) in an accessible way, i.e.
-without the need of expert knowledge in implementing and using such procedures.
-In addition to this, probabilistic programming systems~(PPSs) provide tools that enable the source code of a program to be throughly analysed in an automated fashion.
-% Users can then exploit these algorithms and tools on an array of models
-% within the PPS, using the language itself to create models that represent complicated distribution objects which are then analysed by the PPS. 
-% PPSs are extensions, or reductions, of existing programming languages such as python, c++, go and so forth, which means that users do not have to learn
-% a completely new set of operating semantics and can write programs in the same vain as they are accustomed to.  
-% The PPS that we use, PyProb~\cite{le-2016-inference,baydin2018efficient}, belongs to the family of universal PPSs 
-% that allow the expression of unrestricted probability models in a Turing complete fashion~\cite{wood2014new,goodman2012church,landau_binder_2014,siddharth2017learning,bingham2019pyro}.
-%
-%\begin{itemize}
-%\item Add what - What are we trying to do to make peoples lifes easier
-%\item Add WHy - Why is this important to do and what does it enable.
-%\item Then deal with "How" - How is this done - what is currently there. 
-%\end{itemize}
 \begin{figure}
@@
 %\bg{Add performance plot showing memory consumptions of pyprob and our new system.}
+
 \section{Case Study}

### 2019-04-26 paper/icml_ai4good/main.tex
section 3 fix

--- 2019-04-26 paper/icml_ai4good/main.tex
@@
  Probabilistic programming~\cite{gordon2014probabilistic,staton2016semantics,kozen1979semantics} 
-can be used to express probabilistic models and consequently perform automated inference thereon.
-Once a probabilistic model has been expressed in a probabilistic programming language, a wide range of inference techniques, such as Markov Chain Monte Carlo (MCMC)\cite{}, Variational Inference~(VI)\cite{} and Deep Neural Networks~(DNN)\cite{},  can be used out-of-the-box by non-experts.
+can be used to express probabilistic models and consequently perform automated inference these.
+Once a probabilistic model has been expressed in a probabilistic programming language, a wide range of inference techniques, such as Markov Chain Monte Carlo (MCMC)\cite{}, Variational Inference~(VI)\cite{} and Deep Neural Networks~(DNN)\cite{},  can be used by non-experts in an automated fashion.
 

### 2019-04-26 paper/icml_ai4good/main.tex
section 3

@@
 \usepackage{tikz-cd}
+\usepackage[binary-units=true]{siunitx}
 
@@
 Section \ref{sec:casestudy}).
+Finally, the entry point to the simulator, i.e. \textit{main} in C++, is replaced by a special \textit{forward} call, in this case \textit{pyprob\_cpp::forward}.
+This allows the PPS to generate roll outs from the simulator remotely.
 
-%
+\paragraph{Population-based simulators} In contrast to the setup used by \cite{Baydin}, who interface a particle physics simulator that creates a single trace per \textit{forward} call, a population-based simulator creates a trace per population member.
+This means that e.g.
+in EMOD, a standard scenario simulation rollout over the span of $3$ years
+with a population of size $n=5000$ \cite{smith} will generate about $ \SI{4}{\tera\byte}$ of raw trace data.
+Unlike with \citep{Baydin}, this amount of data cannot possibly be kept in RAM, which makes \textit{a posteriori} trace analysis very inefficient.
+In order to deal with the shortcomings of \textit{pyprob} in a population-based simulator context, we therefore extend the framework to be able to do trace analysis on the fly.
+
+By extending \textit{pyprob} to handle population-based simulators, the PPS can now track all stochastic random variables that are created within the simulator, which then allows us to generate trace plots and path probabilities associated to the execution paths of the program.
+The corresponding increase of simulator transparency helps reinstate much needed trust between policymakers and evidence-based methods.
+
 %As stated previously, Simulators are complex systems and should be viewed as black-boxes, thus deciphering the underlying processes occurring within the simulator facilitates the users understanding of how such a system operates and provides a way of interpreting the outputs in a clear way.
@@
 
-  A4 & Potential children deaths
+  A4, A5 & Potential child deaths
 
   within the population are simulated \\
-  A5 & Potential children deaths within
-
-   the population are simulated \\
 

### 2019-04-26 paper/icml_ai4good/main.tex
updated

--- 2019-04-26 paper/icml_ai4good/main.tex
@@
 In this paper, we introduce an approach that allows one to treat a large class of population-based epidemiology simulators as probabilistic generative models.
-This is achieved by \emph{hijacking} the internal random number generator calls, through the use of an universal probabilistic programming system (PPS).
+This is achieved by \emph{hijacking} the internal random number generator calls, through the use of a universal probabilistic programming system (PPS).
 In contrast to other methods, our approach can be easily retrofitted to simulators written in popular industrial programming frameworks.
 We demonstrate that our method can be used for interpretable introspection and inference, thus shedding light on black-box simulators.
-This reinstates much needed trust between policymakers and evidence-based methods.
+This reinstates much-needed trust between policymakers and evidence-based methods.
 
@@
 
-To reach the WHO’s target of reducing malaria incidence and mortality rates by at least 90\% by 2030, policymakers are increasingly turning to evidence-based methods, thus oftentimes relying on computational simulations CITATIONS.
+To reach the WHO’s target of reducing malaria incidence and mortality rates by at least 90\% by 2030, policymakers are increasingly turning to evidence-based methods, thus often times relying on computational simulations~\cite{who_global_2015}.
 These simulations allow policymakers to infer critical information on disease dynamics and make predictions about the impacts of policies
@@
 In this paper, we introduce a novel method that allows one to shed light on the inner workings of a large class of population-based stochastic simulators.
-We achieve this by extending the work of \citep{baydin2018efficient} by interpreting such population-based simulators as probabilistic generative models within the framework of universal probabilistic programming (UPP)\cite{le-2016-inference}.
+We achieve this by extending the work of \citep{baydin2018efficient} by interpreting such population-based simulators as probabilistic generative models within the framework of universal probabilistic programming (UPP) \cite{le-2016-inference}.
 To this end, we \emph{hijack} existing simulators by overriding their internal random number generators.
@@
  Probabilistic programming~\cite{gordon2014probabilistic,staton2016semantics,kozen1979semantics} 
-can be used to express probabilistic models and consequently perform automated inference these.
-Once a probabilistic model has been expressed in a probabilistic programming language, a wide range of inference techniques, such as Markov Chain Monte Carlo (MCMC)\cite{}, Variational Inference~(VI)\cite{} and Deep Neural Networks~(DNN)\cite{},  can be used by non-experts in an automated fashion.
+can be used to express probabilistic models and consequently perform automated inference thereon.
+Once a probabilistic model has been expressed in a probabilistic programming language, a wide range of inference techniques, such as Markov Chain Monte Carlo (MCMC)\cite{geyer1992practical}, Variational Inference~(VI)\cite{ranganath2014black} and Deep Neural Networks~(DNN)\cite{Goodfellow:2016:DL:3086952},  can be used out-of-the-box by non-experts.
 
-\textit{Hijacking a simulator} describes the process by which a simulator's random number generators are replaced by calls to external sampling procedures which are controlled by a probabilistic programming system (PPS).
+\paragraph{Hijacking a simulator} describes the process by which a simulator's random number generators are replaced by calls to external sampling procedures, which are controlled by a probabilistic programming system (PPS).
 In practice, this amounts to performing a small number of surgical incisions into the simulator's source code in order to replace built-in calls to random number generators.
-E.g., given a simulator written in C++/Boost\cite{}, a Gaussian distribution object \textit{boost::normal} is replaced with the corresponding \textit{pyprob\_cpp} distribution object, namely \textit{pyprob\_cpp::distributions::Normal}.
+E.g., given a simulator written in C++/Boost\cite{schaling2011boost}, a Gaussian distribution object \textit{boost::normal} is replaced with the corresponding \textit{pyprob\_cpp} distribution object, namely \textit{pyprob\_cpp::distributions::Normal}.
 Sampling from this distribution is then done by requesting the PPS to send a sample back to the simulator.
 
-The PPS and the simulator exchange xTensor objects \cite{} through TCP or ICP using a generic FlatBuffers \cite{TODO} protocol.
-On the PPS side, sampling is done using the deep learning framework PyTorch \cite{}.
+The PPS and the simulator exchange xTensor objects \url{https://xtensor.readthedocs.io/en/latest/} through TCP or ICP using a generic FlatBuffers \url{http://google.github.io/flatbuffers/} protocol.
+On the PPS side, sampling is done using the deep learning framework PyTorch \cite{paszke2017automatic}.
 
@@
 Section \ref{sec:casestudy}).
-
 %

### 2019-04-26 paper/icml_ai4good/main.tex

@@
 \icmlauthor{Fiuea Rrrr}{to}
-\icmlauthor{Tateu H.~Yasehe}{ed,to,goo}
-\icmlauthor{Aaoeu Iasoh}{goo}
-\icmlauthor{Buiui Eueu}{ed}
-\icmlauthor{Aeuia Zzzz}{ed}
-\icmlauthor{Bieea C.~Yyyy}{to,goo}
-\icmlauthor{Teoau Xxxx}{ed}
-\icmlauthor{Eee Pppp}{ed}
 \end{icmlauthorlist}
@@
 Section \ref{sec:casestudy}).
-%
+
+Finally, the entry point to the simulator, i.e. \textit{main} in C++, is replaced by a special \textit{forward} call, in this case \textit{pyprob\_cpp::forward}.
+This allows the PPS to generate roll outs from the simulator remotely.
+
+\paragraph{Population-based simulators} In contrast to the setup used by \cite{baydin2018efficient}, who interface a particle physics simulator that creates a single trace per \textit{forward} call, a population-based simulator creates a trace per population member.
+This means that e.g.
+in EMOD, a standard scenario simulation rollout over the span of $3$ years
+with a population of size $n=5000$ \cite{smith2008towards} will generate about $ \SI{4}{\tera\byte}$ of raw trace data.
+Unlike with \citep{baydin2018efficient}, this amount of data cannot possibly be kept in RAM, which makes \textit{a posteriori} trace analysis very inefficient.
+In order to deal with the shortcomings of \textit{pyprob} in a population-based simulator context, we therefore extend the framework to be able to do trace analysis on the fly.
+
+By extending \textit{pyprob} to handle population-based simulators, the PPS can now track all stochastic random variables that are created within the simulator, which then allows us to generate trace plots and path probabilities associated to the execution paths of the program.
+The corresponding increase of simulator transparency helps reinstate much needed trust between policymakers and evidence-based methods.
+
 %As stated previously, Simulators are complex systems and should be viewed as black-boxes, thus deciphering the underlying processes occurring within the simulator facilitates the users understanding of how such a system operates and provides a way of interpreting the outputs in a clear way.

### 2019-04-27 paper/icml_ai4good/main.tex
finished case section

--- 2019-04-26 paper/icml_ai4good/main.tex
@@
 
+Ensemble methods are commonly used in statistics in order to combine the predictive power of multiple models \cite{}.
+To this end, recent work has attempted to characterise the similarities and differences between two of the most advanced Malaria epidemiology simulators, EMOD \cite{} and OpenMalaria \cite{}.
+Evaluation is usually done by comparing a number of output parameters across a range of hand-crafted standard scenarios reflecting different geographical locations across Africa.
+\cite{Smith}.
+
+In the following, we illustrate how our method introduces a novel introspection paradigm.
+By extracting trace graphs from population-based simulators, policymakers can ask specific questions about intrinsic properties of the model trace flow.
+
+To illustrate the above, we present simulation output generated from a scenario resembling local conditions in the town of Ifakara (Tanzania).
+Both EMOD and OpenMalaria are configured to simulate a single population node of $n=100$ and assume constant climatic conditions and no migration over the simulation period of $3$ years.
+
 We provide examples of the generated trace plots from the connection between the simulator and the PPS in Figure~\ref{fig:plotewan}. 
-When comparing the two trace plots between EMOD and OgpenMalaria we see significant differences, despite both evaluating the same model for the same disease. 
-We can see from the addressing schemes $A_1, \ldots, A_N$ \bg{ref table 1 and 2} what physical events are connected to each other and how outputs in EMOD are generated from a different set of procedures as compared to OpenMalaria
+When comparing the two trace plots between EMOD and OpenMalaria we see significant differences, despite both evaluating the same model for the same disease.
+
+We can see from the addressing schemes $A_1, \ldots, A_N$ \ref{table:addresses}\ref{table:asingleaddress} what physical events are connected to each other and how outputs in EMOD are generated from a different set of procedures as compared to OpenMalaria
 By having access to such diagrams users can internally evaluate and scrutinize the decisions that the simulator is making.
+
 This is incredibly important for policy makers, or non-experts, as it not only details how we arrive at the given outputs, but it provides an understanding of which processes were most crucial in determining those outputs as can be seen from the path probabilities assigned to each of the vertices.
 
-%
+Additionally, by associating nodes in trace graphs representing the same physical processes within different models, the significance of model detail can be evaluated.
+This could help developers better control model complexity, even dynamically during execution.
 

### 2019-04-27 paper/icml_ai4good/main.tex
updated

--- 2019-04-26 paper/icml_ai4good/main.tex
@@
 We achieve this by extending the work of \citep{baydin2018efficient} by interpreting such population-based simulators as probabilistic generative models within the framework of universal probabilistic programming (UPP) \cite{le-2016-inference}.
-To this end, we \emph{hijack} existing simulators by overriding their internal random number generators.
+To this end, we \emph{hijack} existing Bsimulators by overriding their internal random number generators.
 Specifically, by replacing the existing low-level random number generator in a simulator with a call to a purpose-built UPP ``controller'', which can thus control, track and manipulate the stochasticity of the simulator.
@@
 When comparing the two trace plots between EMOD and OgpenMalaria we see significant differences, despite both evaluating the same model for the same disease. 
-We can see from the addressing schemes $A_1, \ldots, A_N$ \bg{ref table 1 and 2} what physical events are connected to each other and how outputs in EMOD are generated from a different set of procedures as compared to OpenMalaria
+We can see from the addressing schemes $A_1, \ldots, A_N$ what physical events are connected to each other and how outputs in EMOD are generated from a different set of procedures as compared to OpenMalaria
 By having access to such diagrams users can internally evaluate and scrutinize the decisions that the simulator is making.
@@
 
+\section{Discussions and Future Work}
+
+\paragraph{Discussion} In this work we have demonstrated a method 
+that enables one to hijack population based simulators, extending 
+the work of \citep{baydin2018efficient}. 
+We applied our method to two Malaria orientated population-based simulators and 
+generated a variety of trace graphs.
+Finally, we have shown how our system
+enables policy makers and non-experts to analyse simulator outputs in a 
+way previously unavailable in the field of epidemiology.
+
+\paragraph{Future Work} To extend our work further we aim to implement additional 
+tools that will facilitate complicated inference procedures that condition on simulator 
+output.
+
 \end{document}

### 2019-04-27 paper/icml_ai4good/main.tex
added EIR Ifakara

--- 2019-04-27 paper/icml_ai4good/main.tex
@@
 Both EMOD and OpenMalaria are configured to simulate a single population node of $n=100$ and assume constant climatic conditions and no migration over the simulation period of $3$ years.
+Please refer to Figure \ref{fig:EIR} for the seasonal Entomological Inoculation Rate (EIR), a measure of infectivity.
+
+\begin{figure}[h!]
+	\centering
+	\includegraphics[width=\textwidth/2]{../plots/EIR_Ifakara.pdf}
+	\caption{Seasonal Entomological Inoculation Rate (EIR) for the Ifakara scenario. Data is averaged over $30$ day periods.}
+	\label{fig:EIR}
+\end{figure}
 
@@
 
-This is incredibly important for policy makers, or non-experts, as it not only details how we arrive at the given outputs, but it provides an understanding of which processes were most crucial in determining those outputs as can be seen from the path probabilities assigned to each of the vertices.
+This is incredibly important for policymakers, or general non-experts, as it not only details how we arrive at the given outputs, but it provides an understanding of which processes were most crucial in determining those outputs as can be seen from the path probabilities assigned to each of the vertices.
 

### 2019-04-27 paper/icml_ai4good/main.tex

--- 2019-04-27 paper/icml_ai4good/main.tex
@@
 
-Ensemble methods are commonly used in statistics in order to combine the predictive power of multiple models \cite{}.
-To this end, recent work has attempted to characterise the similarities and differences between two of the most advanced Malaria epidemiology simulators, EMOD \cite{} and OpenMalaria \cite{}.
-Evaluation is usually done by comparing a number of output parameters across a range of hand-crafted standard scenarios reflecting different geographical locations across Africa.
-\cite{Smith}.
+Ensemble methods are commonly used in statistics in order to combine the predictive power of multiple models \cite{cameron2015defining,smith_ensemble_2012}.
+To this end, recent work has attempted to characterise the similarities and differences between two of the most advanced Malaria epidemiology simulators, EMOD \cite{bershteyn2018implementation} and OpenMalaria \cite{smith2008towards}.
+Evaluation is usually done by comparing a number of output parameters across a range of hand-crafted standard scenarios reflecting different geographical locations across Africa~\cite{smith_ensemble_2012}.
 
@@
 We provide examples of the generated trace plots from the connection between the simulator and the PPS in Figure~\ref{fig:plotewan}. 
-When comparing the two trace plots between EMOD and OpenMalaria we see significant differences, despite both evaluating the same model for the same disease.
-
 We can see from the addressing schemes $A_1, \ldots, A_N$ \ref{table:addresses}\ref{table:asingleaddress} what physical events are connected to each other and how outputs in EMOD are generated from a different set of procedures as compared to OpenMalaria
@@
   \end{table}
+
+\section{Discussions and Future Work}
+
+\paragraph{Discussion} In this work we have demonstrated a method 
+that enables one to hijack population based simulators, extending 
+the work of \citep{baydin2018efficient}. 
+We applied our method to two Malaria orientated population-based simulators and 
+generated a variety of trace graphs.
+Finally, we have shown how our system
+enables policy makers and non-experts to analyse simulator outputs in a 
+way previously unavailable in the field of epidemiology.
+
+\paragraph{Future Work} To extend our work further we aim to implement additional 
+tools that will facilitate complicated inference procedures that condition on simulator 
+output.
+
 \bibliographystyle{icml2019}
 \bibliography{malaria,refs}
-
 \end{document}

### 2019-04-27 paper/icml_ai4good/main.tex
updatedd

@@
 We provide examples of the generated trace plots from the connection between the simulator and the PPS in Figure~\ref{fig:plotewan}. 
-We can see from the addressing schemes $A_1, \ldots, A_N$ \ref{table:addresses}\ref{table:asingleaddress} what physical events are connected to each other and how outputs in EMOD are generated from a different set of procedures as compared to OpenMalaria
+We can see from the addressing schemes $A_1, \ldots, A_N$ Tables \ref{table:addresses} and \ref{table:asingleaddress} what physical events are connected to each other and how outputs in EMOD are generated from a different set of procedures as compared to OpenMalaria
 By having access to such diagrams users can internally evaluate and scrutinize the decisions that the simulator is making.

### 2019-04-27 paper/icml_ai4good/main.tex
almost done

--- 2019-04-27 paper/icml_ai4good/main.tex
@@
 
-\section{Case Study}
+\section{Case Study: Ifakara (Tanzania)}
 \label{sec:casestudy}
@@
 We provide examples of the generated trace plots from the connection between the simulator and the PPS in Figure~\ref{fig:plotewan}. 
-We can see from the addressing schemes $A_1, \ldots, A_N$ \ref{table:addresses}\ref{table:asingleaddress} what physical events are connected to each other and how outputs in EMOD are generated from a different set of procedures as compared to OpenMalaria
+We can see from the addressing schemes $A_1, \ldots, A_N$ \ref{table:addresses}\ref{table:asingleaddress} what physical events are connected to each other and how outputs in EMOD are generated from a different set of procedures as compared to OpenMalaria.
 By having access to such diagrams users can internally evaluate and scrutinize the decisions that the simulator is making.
 
-This is incredibly important for policymakers, or general non-experts, as it not only details how we arrive at the given outputs, but it provides an understanding of which processes were most crucial in determining those outputs as can be seen from the path probabilities assigned to each of the vertices.
+This is important for policymakers, or general non-experts, as it not only details how we arrive at the given outputs, but it provides an understanding of which processes were most crucial in determining those outputs as can be seen from the path probabilities assigned to each of the vertices.
 
 Additionally, by associating nodes in trace graphs representing the same physical processes within different models, the significance of model detail can be evaluated.
-This could help developers better control model complexity, even dynamically during execution.
+For example, the full trace presented in Table \ref{table:asingleaddress} represent a sampling step associated with within-host dynamics of the Malaria parasite \textit{Falciparum} in OpenMalaria node $A1$.
+The same physical process also occurs in EMOD's trace graph at position $A7$ (see Section \ref{table:addresses}).
+Comparisons like these could help developers better control model complexity, and even provide an alternative testing and debugging paradigm.
 
@@
 
+\quad\\
+\quad\\
+\quad\\
+\quad\\
+
 \section{Discussions and Future Work}
@@
 tools that will facilitate complicated inference procedures that condition on simulator 
-output.
+output. We will also evaluate additional scenarios across Africa and South-East Asia to better understand the similarities and differences between EMOD and OpenMalaria.
 

### 2019-04-27 paper/icml_ai4good/main.tex

@@
 We provide examples of the generated trace plots from the connection between the simulator and the PPS in Figure~\ref{fig:plotewan}.
-We can see from the addressing schemes $A_1, \ldots, A_N$ \ref{table:addresses}\ref{table:asingleaddress} what physical events are connected to each other and how outputs in EMOD are generated from a different set of procedures as compared to OpenMalaria.
+
+We can see from the addressing schemes $A_1, \ldots, A_N$ (Tables \ref{table:addresses} and \ref{table:asingleaddress}) what physical events are connected to each other and how outputs in EMOD are generated from a different set of procedures as compared to OpenMalaria
 By having access to such diagrams users can internally evaluate and scrutinize the decisions that the simulator is making.

### 2019-04-27 paper/icml_ai4good/main.tex
updated

--- 2019-04-27 paper/icml_ai4good/main.tex
@@
 
-\textit{Hijacking a simulator} describes the process by which a simulator's random number generators are replaced by calls to external sampling procedures which are controlled by a probabilistic programming system (PPS).
+\paragraph{Hijacking a simulator} describes the process by which a simulator's random number generators are replaced by calls to external sampling procedures, which are controlled by a probabilistic programming system (PPS).
 In practice, this amounts to performing a small number of surgical incisions into the simulator's source code in order to replace built-in calls to random number generators.
-E.g., given a simulator written in C++/Boost\cite{}, a Gaussian distribution object \textit{boost::normal} is replaced with the corresponding \textit{pyprob\_cpp} distribution object, namely \textit{pyprob\_cpp::distributions::Normal}.
+E.g., given a simulator written in C++/Boost\cite{schaling2011boost}, a Gaussian distribution object \textit{boost::normal} is replaced with the corresponding \textit{pyprob\_cpp} distribution object, namely \textit{pyprob\_cpp::distributions::Normal}.
 Sampling from this distribution is then done by requesting the PPS to send a sample back to the simulator.
 
-The PPS and the simulator exchange xTensor objects \cite{} through TCP or ICP using a generic FlatBuffers \cite{TODO} protocol.
-On the PPS side, sampling is done using the deep learning framework PyTorch \cite{}.
+The PPS and the simulator exchange xTensor objects \url{https://xtensor.readthedocs.io/en/latest/} through TCP or ICP using a generic FlatBuffers \url{http://google.github.io/flatbuffers/} protocol.
+On the PPS side, sampling is done using the deep learning framework PyTorch \cite{paszke2017automatic}.
 
@@
 Section \ref{sec:casestudy}).
+
 Finally, the entry point to the simulator, i.e. \textit{main} in C++, is replaced by a special \textit{forward} call, in this case \textit{pyprob\_cpp::forward}.
@@
 
-\paragraph{Population-based simulators} In contrast to the setup used by \cite{Baydin}, who interface a particle physics simulator that creates a single trace per \textit{forward} call, a population-based simulator creates a trace per population member.
+\paragraph{Population-based simulators} In contrast to the setup used by \cite{baydin2018efficient}, who interface a particle physics simulator that creates a single trace per \textit{forward} call, a population-based simulator creates a trace per population member.
 This means that e.g.
 in EMOD, a standard scenario simulation rollout over the span of $3$ years
-with a population of size $n=5000$ \cite{smith} will generate about $ \SI{4}{\tera\byte}$ of raw trace data.
-Unlike with \citep{Baydin}, this amount of data cannot possibly be kept in RAM, which makes \textit{a posteriori} trace analysis very inefficient.
+with a population of size $n=5000$ \cite{smith2008towards} will generate about $ \SI{4}{\tera\byte}$ of raw trace data.
+Unlike with \citep{baydin2018efficient}, this amount of data cannot possibly be kept in RAM, which makes \textit{a posteriori} trace analysis very inefficient.
 In order to deal with the shortcomings of \textit{pyprob} in a population-based simulator context, we therefore extend the framework to be able to do trace analysis on the fly.
+
+By extending \textit{pyprob} to handle population-based simulators, the PPS can now track all stochastic random variables that are created within the simulator, which then allows us to generate trace plots and path probabilities associated to the execution paths of the program.
+The corresponding increase of simulator transparency helps reinstate much needed trust between policymakers and evidence-based methods.
 

### 2019-04-27 paper/icml_ai4good/main.tex

@@
 We provide examples of the generated trace plots from the connection between the simulator and the PPS in Figure~\ref{fig:plotewan}.
-We can see from the addressing schemes $A_1, \ldots, A_N$ Tables \ref{table:addresses} and \ref{table:asingleaddress} what physical events are connected to each other and how outputs in EMOD are generated from a different set of procedures as compared to OpenMalaria
+
+We can see from the addressing schemes $A_1, \ldots, A_N$ (Tables \ref{table:addresses} and \ref{table:asingleaddress}) what physical events are connected to each other and how outputs in EMOD are generated from a different set of procedures as compared to OpenMalaria
 By having access to such diagrams users can internally evaluate and scrutinize the decisions that the simulator is making.
 
-This is incredibly important for policymakers, or general non-experts, as it not only details how we arrive at the given outputs, but it provides an understanding of which processes were most crucial in determining those outputs as can be seen from the path probabilities assigned to each of the vertices.
+This is important for policymakers, or general non-experts, as it not only details how we arrive at the given outputs, but it provides an understanding of which processes were most crucial in determining those outputs as can be seen from the path probabilities assigned to each of the vertices.
 
 Additionally, by associating nodes in trace graphs representing the same physical processes within different models, the significance of model detail can be evaluated.
-This could help developers better control model complexity, even dynamically during execution.
+For example, the full trace presented in Table \ref{table:asingleaddress} represent a sampling step associated with within-host dynamics of the Malaria parasite \textit{Falciparum} in OpenMalaria node $A1$.
+The same physical process also occurs in EMOD's trace graph at position $A7$ (see Section \ref{table:addresses}).
+Comparisons like these could help developers better control model complexity, and even provide an alternative testing and debugging paradigm.
 

### 2019-04-27 paper/icml_ai4good/main.tex
fixed minor shit

--- 2019-04-27 paper/icml_ai4good/main.tex
@@
 Our framework already supports application to simulators written in 13 industrial programming frameworks, and is easily extensible.
-This is crucial as, given the enormous code size and complexity, rewriting epidemiology simulators using a dedicated universal probabilistic programming language, such as Pyro\cite{bingham2019pyro}, is often infeasible.
+This is crucial as, given the enormous code size and complexity, rewriting epidemiology simulators using a dedicated universal probabilistic programming language, such as Pyro \cite{bingham2019pyro}, is often infeasible.
 %Moreover, though our focus is on hijacking epidemiology simulators, our general approach and framework can be applied more widely to a general class of stochastic simulators.
@@
 
-Two the most advanced Malaria simulators, namely EMOD~\cite{bershteyn2018implementation} and OpenMalaria\cite{smith2008towards}, have proven to be particularly valuable to policymakers.
+Two the most advanced Malaria simulators, namely EMOD~\cite{bershteyn2018implementation} and OpenMalaria  \cite{smith2008towards}, have proven to be particularly valuable to policymakers.
 OpenMalaria is based on microsimulations of Plasmodium falciparum in humans and was originally developed to simulate the impacts of malaria vaccines within simple villages or districts
@@
 Finally, the entry point to the simulator, i.e. \textit{main} in C++, is replaced by a special \textit{forward} call, in this case \textit{pyprob\_cpp::forward}.
-This allows the PPS to generate roll outs from the simulator remotely.
+This allows the PPS to generate rollouts from the simulator remotely.
 

### 2019-04-27 paper/icml_ai4good/main.tex
final submission 1

@@
 In the following, we illustrate how our method introduces a novel introspection paradigm.
-By extracting trace graphs from population-based simulators, policymakers can ask specific questions about intrinsic properties of the model trace flow.
+By extracting trace graphs from population-based simulators, policymakers can ask specific questions about properties of the model trace flow.
 
@@
 
-\begin{table}[h!]
+\begin{table}[b!]
 \footnotesize
@@
 
-\quad\\
-\quad\\
-\quad\\
-\quad\\
-
 \section{Discussions and Future Work}
 
-\paragraph{Discussion} In this work we have demonstrated a method 
-that enables one to hijack population based simulators, extending 
+In this work we have demonstrated a method 
+that enables one to hijack populationbased simulators, extending 
 the work of \citep{baydin2018efficient}. 
@@
 
-\paragraph{Future Work} To extend our work further we aim to implement additional 
+To extend our work further we aim to implement additional 
 tools that will facilitate complicated inference procedures that condition on simulator 
 output. We will also evaluate additional scenarios across Africa and South-East Asia to better understand the similarities and differences between EMOD and OpenMalaria.
+
+\quad\\
+\quad\\
 

### 2019-04-27 paper/icml_ai4good/main.tex
updated

@@
 
-To reach the WHO’s target of reducing malaria incidence and mortality rates by at least 90\% by 2030, policymakers are increasingly turning to evidence-based methods, thus oftentimes relying on computational simulations CITATIONS.
+To reach the WHO’s target of reducing malaria incidence and mortality rates by at least 90\% by 2030, policymakers are increasingly turning to evidence-based methods, thus oftentimes relying on computational simulations \cite{who_global_2015}.
 These simulations allow policymakers to infer critical information on disease dynamics and make predictions about the impacts of policies
-before they are rolled out. This frequently increases the effectiveness of interventions and thus ultimately saves resources, or even lifes.
+before they are rolled out. This frequently increases the effectiveness of interventions and thus ultimately saves resources, or even lives.
 For example, it has been shown that mass vaccination may be largely ineffective in regions of large transmission rates, but may play a crucial role in areas of low transmission \cite{cameron2015defining}.
@@
 
-By extending \textit{pyprob} to handle population-based simulators, the PPS can now track all stochastic random variables that are created within the simulator, which then allows us to generate trace plots and path probabilities associated to the execution paths of the program.
-The corresponding increase of simulator transparency helps reinstate much needed trust between policymakers and evidence-based methods.
-
 %As stated previously, Simulators are complex systems and should be viewed as black-boxes, thus deciphering the underlying processes occurring within the simulator facilitates the users understanding of how such a system operates and provides a way of interpreting the outputs in a clear way.
@@
 
-\begin{table}[b!]
+\begin{table}[h!]
 \footnotesize

### 2019-05-15 paper/neurlps2019/main.tex
updated

--- 2019-04-27 paper/icml_ai4good/main.tex
@@
-%%%%%%%% ICML 2019 EXAMPLE LATEX SUBMISSION FILE %%%%%%%%%%%%%%%%%
-
 \documentclass{article}
 
-% Recommended, but optional, packages for figures and better typesetting:
-\usepackage{microtype}
+% if you need to pass options to natbib, use, e.g.:
+%     \PassOptionsToPackage{numbers, compress}{natbib}
+% before loading neurips_2018
+
+% ready for submission
+% \usepackage{neurips_2018}
+
+% to compile a preprint version, e.g., for submission to arXiv, add add the
+% [preprint] option:
+%     \usepackage[preprint]{neurips_2018}
+
+% to compile a camera-ready version, add the [final] option, e.g.:
+     \usepackage[final]{neurips_2019}
+
+% to avoid loading the natbib package, add option nonatbib:
+%     \usepackage[nonatbib]{neurips_2018}
+
+\usepackage[utf8]{inputenc} % allow utf-8 input
+\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
+\usepackage{hyperref}       % hyperlinks
+\usepackage{url}            % simple URL typesetting
 \usepackage{graphicx}
-\usepackage{subfigure}
-\usepackage{booktabs} % for professional tables
+\usepackage{booktabs}       % professional-quality tables
+\usepackage{amsfonts}       % blackboard math symbols
+\usepackage{nicefrac}       % compact symbols for 1/2, etc.
+\usepackage{microtype}      % microtypography
+\usepackage{listings}
 \usepackage{tabularx}
@@
 \usepackage{tikz-cd}
-\usepackage[binary-units=true]{siunitx}
+\usepackage{algorithm}
+\usepackage{algpseudocode}
+% \usepackage[binary-units=true]{siunitx}
+\title{Converting Population-Based Simulators to Probabilistic Programs}
 
-% hyperref makes hyperlinks in the resulting PDF.
-% If your build breaks (sometimes temporarily if a hyperlink spans a page)
-% please comment out the following usepackage line and replace
-% \usepackage{icml2019} with \usepackage[nohyperref]{icml2019} above.
-\usepackage{hyperref}
+% neurlps classification
+% Primary: Applications
+% Secodnary cat:Data, Challenges, Implementations, and Software
+% or Hijacking Malaria Simulators with Probabilistic Programming
 
-% \usepgfplotslibrary{external}
-% Attempt to make hyperref and algorithmic work together better:
-\newcommand{\theHalgorithm}{\arabic{algorithm}}
-
-% Use the following line for the initial blind version submitted for review:
-\usepackage{icml2019}
+% The \author macro works with any number of authors. There are two commands
+% used to separate the names and addresses of multiple authors: \And and \AND.
+%
+% Using \And between authors leaves it to LaTeX to determine where to break the
+% lines. Using \AND forces a line break at that point. So, if LaTeX puts 3 of 4
+% authors names on the first line, and the last on the second line, try using
+% \AND instead of \And before the third author name.
 \newcommand{\bg}[1]{~{{[{\it \textcolor{red}{{\bf BG:} #1}}]}}}
@@
    execute at end node=$]
-% The \icmltitle you define below is probably too long as aer.
-% Therefore, a short form for the running title is supplied here:
-%\icmltitlerunning{Hijacking Simulators for the Greater Good}
-\icmltitlerunning{Hijacking Malaria Simulators
-	with Probabilistic Programming}
 
@@
 \pgfplotsset{width=7cm,compat=1.8}
+\author{%
+Paper ID: }
+  % examples of more authors
+  % \And
+  % Coauthor \\
+  % Affiliation \\
+  % Address \\
+  % \texttt{email} \\
+  % \AND
+  % Coauthor \\
+  % Affiliation \\
+  % Address \\
+  % \texttt{email} \\
+  % \And
+  % Coauthor \\
+  % Affiliation \\
+  % Address \\
+  % \texttt{email} \\
+  % \And
+  % Coauthor \\
+  % Affiliation \\
+  % Address \\
+  % \texttt{email} \\
+%}
 \begin{document}
+% \nipsfinalcopy is no longer used
 
-\twocolumn[
-%\icmltitle{Hijacking Simulators for the Greater Good}
-%\icmltitle{Automatically Converting Epidemiology Simulations
-%	to Probabilistic Programs}
-\icmltitle{Hijacking Malaria Simulators
-	with Probabilistic Programming}
-
-% It is OKAY to include author information, even for blind
-% submissions: the style file will automatically remove it for you
-% unless you've provided the [accepted] option to the icml2019
-% package.
-
-% List of affiliations: The first argument should be a (short)
-% identifier you will use later to specify author affiliations
-% Academic affiliations should list Department, University, City, Region, Country
-% Industry affiliations should list Company, City, Region, Country
-
-% You can specify symbols, otherwise they are numbered in order.
-% Ideally, you should not use this facility. Affiliations will be numbered
-% in order of appearance and this is the preferred way.
-\icmlsetsymbol{equal}{*}
-
-\begin{icmlauthorlist}
-\icmlauthor{Aeiau Zzzz}{equal,to}
-\icmlauthor{Bauiu C.~Yyyy}{equal,to,goo}
-\icmlauthor{Cieua Vvvvv}{goo}
-\icmlauthor{Iaesut Saoeu}{ed}
-\icmlauthor{Fiuea Rrrr}{to}
-\icmlauthor{Tateu H.~Yasehe}{ed,to,goo}
-\icmlauthor{Aaoeu Iasoh}{goo}
-\icmlauthor{Buiui Eueu}{ed}
-\icmlauthor{Aeuia Zzzz}{ed}
-\icmlauthor{Bieea C.~Yyyy}{to,goo}
-\icmlauthor{Teoau Xxxx}{ed}
-\icmlauthor{Eee Pppp}{ed}
-\end{icmlauthorlist}
-
-\icmlaffiliation{to}{Department of Computation, University of Torontoland, Torontoland, Canada}
-\icmlaffiliation{goo}{Googol ShallowMind, New London, Michigan, USA}
-\icmlaffiliation{ed}{School of Computation, University of Edenborrow, Edenborrow, United Kingdom}
-
-\icmlcorrespondingauthor{Cieua Vvvvv}{c.vvvvv@googol.com}
-\icmlcorrespondingauthor{Eee Pppp}{ep@eden.co.uk}
-
-% You may provide any keywords that you
-% find helpful for describing your paper; these are used to populate
-% the "keywords" metadata in the PDF but will not be shown in the document
-\icmlkeywords{Machine Learning, ICML}
-
-\vskip 0.3in
-]
-
-% this must go after the closing bracket ] following \twocolumn[ ...
-
-% This command actually creates the footnote in the first column
-% listing the affiliations and the copyright notice.
-% The command takes one argument, which is text to display at the start of the footnote.
-% The \icmlEqualContribution command is standard text for equal contribution.
-% Remove it (just {}) if you do not need this facility.
-
-%\printAffiliationsAndNotice{}  % leave blank if no need to mention equal contribution
-\printAffiliationsAndNotice{\icmlEqualContribution} % otherwise use the standard text.
+\maketitle
 
 \begin{abstract}
-Epidemiology simulations have become a fundamental tool in the fight
-against the epidemics of various infectious diseases like AIDS and
-malaria.
-However, the complicated and stochastic nature of these simulators
-can mean their output is difficult to interpret, which reduces their usefulness to policymakers.
-In this paper, we introduce an approach that allows one to treat a large class of population-based epidemiology simulators as probabilistic generative models.
-This is achieved by \emph{hijacking} the internal random number generator calls, through the use of an universal probabilistic programming system (PPS).
-In contrast to other methods, our approach can be easily retrofitted to simulators written in popular industrial programming frameworks.
-We demonstrate that our method can be used for interpretable introspection and inference, thus shedding light on black-box simulators.
-This reinstates much needed trust between policymakers and evidence-based methods.
 
-%This allows for various advanced statistical analyses to be performed
-%on the simulators in an automated fashion, while also tracking
-%the behavior of the simulator to uncover key information such
-%as stochastic structure and generate result summaries.
-%Critically, this hijacking can be performed with only minimal changes
-%to the original simulator, allowing for straightforward application
-%of our approach to a wide variety of contexts and existing simulators.
-%
-%A goal of probabilistic programming is to couple simulators, with inference. This is 
-%because stochastic simulators are used prominently in many industrial settings,
-%do not require one to construct hand-crafted joint distributions as they implicitly 
-%define a joint distribution of the program and encode learned structures 
-%directly. This makes simulators powerful tools and much of machine learning (ML) and 
-%Artificial Intelligence (AI)
-%can be seen as trying to emulate such simulators from a purely data-driven approach.
-%However, in the 
-%ML/AI setting, although we can often infer outcomes, we have little understanding about what 
-%in the data led to the outputted inferences. 
-%This makes it challenging to deploy ML/AI systems into the wild, especially in health-related and safety-critical domains, such
-%as epidemiology, as we lose \emph{interpretability}. 
-%In this work, we explain how to design ML/AI systems that combine
-%probabilistic programming systems (PPSs) and epidemiology simulators, to extract
-%fully interpretable structures, enabling policy makers 
-%and practitioners to make decisions from simulator outputs that can be fully reasoned on.
-% % interpretable inferences. 
-%In particular, we demonstrate this for the Malaria disease in two commonly used simulators; EMOD and OpenMalaria.
+% A goal of probabilistic programming is to couple simulators, with inference. This is 
+% because stochastic simulators are used prominently in many industrial settings,
+% do not require one to construct hand-crafted joint distributions as they implicitly 
+% define a joint distribution of the program and encode learnt structures 
+% directly. This makes simulators powerful tools and much of machine learning (ML) and 
+% Artificial Intelligence (AI)
+% can be seen as trying to emulate such simulators from a purely data-driven approach.
+% However, in the 
+% ML/AI setting, although we can often infer outcomes, we have little understanding about what 
+% in the data led to the outputted inferences. 
+% This makes it challenging to deploy ML/AI systems into the wild, especially in health-related and safety-critical domains, such
+% as epidemiology, as we lose \emph{interpretability}. 
+% In this work, we explain how to design ML/AI systems that combine
+% probabilistic programming systems (PPSs) and epidemiology simulators, to extract
+% fully interpretable posterior structures, enabling policy makers 
+% and practitioners to make interpretable inferences. 
+% In particular, we demonstrate this for the Malaria disease and show how we can 
+% perform interpretable inference in such settings.
+
+Probabilistic programming provides a way to perform statistical inference over simulations of events in a programmatic way.
+However, with the myriad of probabilistic programming systems (PPSs) and the complexities involved in running 
+each individual system, such as specifying a model in the given language and $\backslash$ or running inference, which means it can be difficult for users to take advantage of the system. 
+In addition to this, even though there exists a myriad of simulators across many scientific domains 
+and within industry, none can be used within existing PPSs without having to re-implement the simulator in the given PPS, something that is not typically feasible, nor practical.  
+However, recent work
+by \cite{baydin2018efficient} demonstrated a pathway for how to turn a particular
+type of simulator, written in a set of 9 languages, into a probabilistic program via the raw random
+number calls.
+ But, this still meant that a large class of critically important population-based simulators could not be turned into probabilistic programs.
+We, in this work, extend that framework to encompass population-based simulators, a very large class of simulators that are used extensively across epidemiology modeling, multi-agent modeling, financial modeling and so forth.
+In addition to this, we enable simulators written in 13 different languages to be easily converted in probabilistic programs, which we algorithmically outline.
+We demonstrate our PPS on a both a toy population simulator and two important malaria population-based simulators called OpenMalaria and EMOD.
+
+% Epidemiology simulations have become a fundamental tool in the fight
+% against the epidemics of various infectious diseases like AIDS and
+% malaria.
+% However, the complicated and stochastic nature of these simulators
+% can mean their output is difficult to interpret, which reduces their usefulness to policymakers.
+% In this paper, we introduce an approach that allows one to treat a large class of population-based epidemiology simulators as probabilistic generative models.
+%  This is achieved by \emph{hijacking} the internal random number generator calls, through the use of an universal probabilistic programming system (PPS). 
+%  In contrast to other methods, our approach can be easily retrofitted to simulators written in popular industrial programming frameworks. 
+% We demonstrate that our method can be used for interpretable introspection and inference, thus shedding light on black-box simulators. 
+% This reinstates much needed trust between policymakers and evidence-based methods. 
 \end{abstract}
@@
 \section{Introduction}
-Ending the epidemics of AIDS, tuberculosis, malaria and other infectious diseases by 2030 is a key target within the Good Health \& Well-Being section of the UN Sustainable Development Goals \cite{refugees_2030,un_sustainable2018}.
-However, despite decades of substantial international efforts, these diseases kill hundreds of million people a year.
-For example, malaria still annually kills about a quarter of a million children under the age of 5 in Africa alone.
 
-%maybe cut some of this paragraph out
-% In contrast to AIDS, Malaria can be routinely cured thanks to the disposability of a variety of medical interventions. 
-% Issues are further aggravated due to the increasing emergence of parasite resistance to multiple drugs, including artemisinin, or even insecticides. 
-% Hence, despite decades of substantial international efforts, initial hopes of regional elimination or even global eradication have not been met.
-%  So difficult is Malaria control that, even today, about a quarter of a million children under the age of 5 succumb to it every year in Africa alone.
+\begin{itemize}
 
-To reach the WHO’s target of reducing malaria incidence and mortality rates by at least 90\% by 2030, policymakers are increasingly turning to evidence-based methods, thus oftentimes relying on computational simulations \cite{who_global_2015}.
-These simulations allow policymakers to infer critical information on disease dynamics and make predictions about the impacts of policies
-before they are rolled out. This frequently increases the effectiveness of interventions and thus ultimately saves resources, or even lives.
-For example, it has been shown that mass vaccination may be largely ineffective in regions of large transmission rates, but may play a crucial role in areas of low transmission \cite{cameron2015defining}.
+\item Outline the objectives
+\item State the importance of interpretability in the inference of the simulator
+\item Our system enables one to understand and interpret the most common paths in a program (simulation).
+\item In highlighting this a practitioner can then go back to the field and explore that parameter/s in more depth.
+\item This facilitates understanding and the construction of more detailed models and data collection strategies
 
-Malaria epidemiology is governed by a complex set of drivers, 
-few of which can be understood in isolation \cite{cameron2015defining,autino_epidemiology_2012,smith2008towards,bershteyn2018implementation}.
-These include within-host dynamics, population-specific traits and even local geography.
-Comprehensive modeling of all of these components remains challenging, particularly in a region-specific context. 
-Computational epidemiology simulators have to reflect these complexities and are usually stochastic in nature.
-This can make simulation output highly non-trivial to interpret, particularly when trying to draw desired inferences coupled with observed data \cite{mwendera_challenges_2019,ferris_openmalaria_2015}.
+\end{itemize}
+
+\bg{The audience is a probabilistic programming audience - so that is what we will target}
 
@@
 Our approach is then demonstrated and analysed in the context of a Malaria case study (Section \ref{sec:casestudy}).
+\label{sec:related}
 
-%Our approach is extending recent work by  and is unique in that it can easily be retrofitted to population-based simulators written in popular industrial programming frameworks.
-This is crucial as, given the enormous code size and complexity, rewriting epidemiology simulators using a dedicated universal probabilistic programming language, such as Pyro\cite{}, is often infeasible.
+\section{Epidemiology Simulators and Probabilistic Programming}
+\label{sec:background}
+This will kind of be like a background section.
 
-%that allows one to treat a large class of population-based epidemiology simulators as probabilistic generative models.
-This is achieved by \emph{hijacking} the internal random number generator calls, through the use of an universal probabilistic programming system (PPS).
-In contrast to other methods, our approach requires only minimal changes to existing simulators to be made.
-%We demonstrate that our method can be used for interpretable introspection of and inference on what used to be de facto black box simulators.
-This reinstates much needed trust between policymakers and evidence-based methods.
+  \begin{algorithm}
+   \caption{How to perform posterior inference over a stochastic simulator with 
+   }
+    \begin{algorithmic}[1]
+      \Function{Merge}{$A,p,q,r$}\Comment{Where A - array, p - left, q - middle, r - right}
 
-%In this paper, we show how probabilistic programming
-%techniques CITATIONS can be used to help automate the required calculations to do this, interpreting the simulators as 
-%generative models to perform principled statistical analysis, while
-%providing insights and summary statistics on their behavior to users.
+        \State replace raw samples statements 
+        \State ${n_2} = r - q$
+        \State Create Containerized Environment
+        \For{$i = 1$ to ${n_1}$}
+            \State $L[i] = A[p + i - 1]$
+        \EndFor
+        \For{$j = 1$ to ${n_2}$}
+            \State $R[i] = A[q + j]$
+        \EndFor
+        \State $L[{n_1} + 1] =  \infty $
+        \State $R[{n_2} + 1] =  \infty $
+        \State $i = 1$
+       \EndFunction
 
-%Achieving this aim is substantially complicated by the fact that
-%a vast array of different epidemiology simulators are currently 
-%used in practice.
-%Manually reimplementing these systems in existing
-%probabilistic programming systems (PPSs) is therefore infeasible, as each
-%system would require manual and laborious porting.  Instead, we 
-%take the novel approach of \emph{hijacking} existing simulators by
-%overriding their internal random number generator, building
-%on ideas from~\citep{baydin2018efficient}.  Specifically, by
-%replacing the existing low-level random number generator in a simulator
-%with a call to a purpose-built ``controller'' PPS, the stochasticity
-%of the simulator can be both manipulated and tracked by this controller
-%system.  In turn, this allows for a variety of tasks to be performed on
-%the hijacked simulator, such as running inference (by conditioning
-%the values of certain draws and manipulating others),
-%uncovering stochastic structure, and automatically
-%producing result summaries, such as establishing the probability
-%of different program paths.  By providing a common abstraction 
-%framework for different simulators, our approach further allows for
-%easy and direct comparison between related or competing
-%simulators, a characteristic that is valuable in the context of
-%epidemiology simulators CITATIONS.
+\end{algorithmic}
+\end{algorithm}
 
-%Critically, this hijacking can be performed on a wide variety of simulators
-%with only minimal changes.  Our framework already
-%supports 13 popular languages for the base simulator, and can easily
-%be extended to others if needed.  Moreover, though our focus
-%is on hijacking epidemiology simulators, our general approach and
-%framework can be applied more widely to a general class of stochastic
-%simulators.  In time, we hope our approach will play a critical role in
-%bringing recent advancements in probabilistic programming to bear on
-%the vast array of existing simulators used throughout the sciences,
-%thereby providing wide-ranging impacts across a number of fields.
+\section{Types of Models and Probabilistic Programming Languages}
 
-%However, the collection of large quantities of field data is often infeasible due to logistic, economic or political constraints, or outright impossible if novel methods are to be assessed a priori. 
-%Hence in silico simulation of malaria epidemiology plays an increasingly crucial role in evidence-based policymaking.
-%Even in the absence of political, economic and social constraints, 
-%Malaria epidemiology is governed by a complex set of drivers, few of which can be understood in isolation.
-%These include within-host dynamics, population-specific traits and even local geography.
-%Some progress in modeling specific components has been made, such as the parasite life-cycle, transmission vector dynamics and the inter-dependencies between infection rates and vicinity to open water.
-%However, comprehensive modeling of all of these components in a regional context remains challenging, and is greatly impeded by the lack of available field data for regional calibration, hence the need for comprehensive simulators.
-% Simulation complexity is further aggravated by considering feedback effects from a large arsenal of anti-malaria interventions, ranging from distribution of mosquito nets, indoor residual spraying and mass drug administration to age-specific vaccinations.
-%Stochastic epidemiology simulators have already proven to be an invaluable tool to policymakers with policy implications that are often counterintuitive yet far-ranging.
+To bridge the gap between the statistics and machine learning community to the probabilistic programming community we provide a useful overview to expose our ideas more concretely.
 
-%However, since many of these simulators are stochastic in nature and therefore act as black-boxes, understanding the structure of these complex simulators is challenging and poses several challenges for policy makers who want to interpret the decisions being inferred.
-%Therefore, sophisticated tools are required that enable us to not only digest the results of the simulation, but provide pathways to perform complex inferences.
+Probabilistic programming languages, broadly speaking\footnote{For a more comprehensive overview see \cite{rainforth2017automating}}, can be classfied into three categories, each type of langauge extends a particular model space, enabling an increasingly complex set models to be represented as programs, but in doing so the notion of posterior and what it means to do inference over the model, at least in a stastically traditional sense, can change.
+% \begin{figure}
+% \centering
+% \includegraphics[]{../models/types_of_pp.png}
+% \end{figure}
+The first type of langauges are called first-order languages where the priors are well defined and the \textit{joint} program density constructed via the program directly represent the unormalised stastical joint density $p(x,y)$
+\section{Methodology}
+ TODO
 
-%To this end, we extend the work of \citep{baydin2018efficient} to enable multiple stochastic epidemiology simulators to be transformed into a probabilistic model that can be used as input to a probabilistic programming system, enabling trace paths to be extracted from the simulator on both a per human level and a population level, from which we can extract trace probabilities allowing policy makers to determine what actions have the biggest impact on differing populations dynamics and thus shape policy decisions accordingly.
-% We argue that our work is likely to have significant social impact on many livelihoods affected by Malaria and discuss opportunities for future research.
-
-% Recent work by \citep{Baydin} introduces a fundamental paradigm change that allows to tackle the inverse problem directly. 
-% Using a novel software tool named \textit{pyprob}, \citep{Baydin} extract program execution traces from state-of-the-art particle decay simulation engine SHERPA\cite{}.
-%  Using latest techniques from universal probabilistic programming inference, the sampled execution traces are then used in order to characterise the posterior distribution of the simulator output. 
-%  From these, regions in simulator input space can be associated with specified outcomes.
-
-% SHERPA, however, is different from leading epidemiology simulators in that it does not simulate population dynamics, but instead models individual decays with different sampling strategies.
-Thus, the existing work of \cite{baydin2018efficient} must be extended to encapsulate these different strategies.
-%  In this paper, we for the first time demonstrate the applicability of \textit{pyprob}, and hence probabilistic programming inference, to population-based epidemiology simulators. 
-%  In particular, we discuss how the leading malaria epidemiology simulators EMOD\cite{} and OpenMalaria\cite{} can be interfaced with \textit{pyprob}.
-%   We argue that our work is likely to have significant social impact on many livelihoods affected by Malaria and discuss opportunities for future research.
-
-% maybe talk about this framework enables a pathway to inference etc..
-
-% All of the above limitations require contemporary malaria epidemiology simulators to employ stochastic modeling techniques, which negotiate data scarcity and model uncertainty through the use of appropriate prior distributions that can be tuned even from small samples of field data.
-The population dynamics of groups of typically several thousand hosts are then initialised by sampling from the initial distributions, using initial random seeds.
-Forward simulation then proceeds by rolling out population dynamics over time.
-Simulation outcomes of cherry-picked scenarios and interventions of interest are then compared to suitably designed control runs.
-
-% This is great, if we had inference implemented right now.
-% However valuable forward simulations of epidemiology scenarios are, the holy grail of goal-driven policymaking ultimately lies in solving the inverse problem: Given a desired outcome and regional context, identify the series of interventions that will best reach the desired target after a given amount of time.
-Due to computational constraints, even an approximate search over the whole simulator input space usually proves intractable.
-An experience-guided search over the simulation output of a small set of manually-designed scenarios of interest therefore long seemed the only feasible way of investigating implications of policy interventions.
-
-% , and, above all, the lack of available field data for calibration,
-
-% parasite life-cycle, transmission vectors, population dynamics and even local geographic circumstances, with policy measures ranging from distribution of mosquito nets to experimental vaccines, requires such simulators to be stochastic.
-Simulation outcomes thus need to be evaluated on a population of subjects.
-
-% <Section about prob prog inference, with reference to SHERPA>
-
-%- issues of public health are complex and
-
-%To do. 
-%
-%\begin{itemize}
-%
-%\item Outline the objectives
-%\item State the importance of interpretability in the inference of the simulator
-%\item Our system enables one to understand and interpret the most common paths in a program (simulation).
-%\item In highlighting this a practitioner can then go back to the field and explore that parameter/s in more depth.
-%\item This facilitates understanding and the construction of more detailed models and data collection strategies
-%
-%
-%\end{itemize}
-%
-%% taken from Toms thesis. 
-%
-
-%In particular, PPSs allow probabilistic models to be represented in the form of a generative model through the
-%probabilistic programming language~(PPL), which enables one to write statements that enable conditioning on data~\cite{gordon2014probabilistic,Goodman08church:a}. 
-%Thus, it seems only natural to connect simulators with PPSs, as simulators explicitly define a generative model, in the language in which they are written, and the nature of the PPSs enables us to perform inference in these simulators, by conditioning on observations, which are fed in as input to the simulators.
-%In doing this we can infer things about stochastic input parameters and other variables sampled during the program's forward execution, whilst also providing full interpretability in the inference results, which is absolutely necessary in safety-critical domains.
-%By exploiting the tools that we develop in this paper, one is able to connect epidemiology simulators with \texttt{Pyprob}~\cite{le-2016-inference} to extract fully interpretable posterior structures allowing practitioners to interpret which input parameters have the largest affect on the epidemic, in our case malaria.
-Over five-hundred-thousand people die from malaria each year, mostly children under five years of age, with 90 per cent of malaria cases occurring in Sub-Saharan Africa.
-An estimated 100-300 million people suffer from malaria each year~\cite{world2016world}.
-%Thus, by understanding the prominence of input parameters to epidemiology simulators, guides decision making processes relating to effective treatment and prevention strategies. 
-%\label{sec:related}
+Simulators encode several years, if not decades, of research and development and so by their very nature are structurally complicated.
+Since the development happens over several decades, usually without documentation, understanding a code base written with legacy libraries and software is incredibly difficult and an almost impossible task for a non-expert, or expert who has not had access to the previous years of development.
+Our method of hijacking simulators builds on the work of~\cite{baydin2018efficient} and extends the framework to encapsulate a more diverse range of simulators, in particular epidemiology simulators.
+Our framework provides a simple solution and makes it easy to transform arbitrary stochastic simulators into probabilistic programs, regardless of the complexity of the simulator and the language that the simulator is written in\footnote{The framework currently supports 9 popular languages, but there is nothing to stop this being extended to any language of interest.}.
+In addition to this, we can understand the structure of programs in simulators for both decoding \emph{black-box structures} and for posterior inference.
+This enables two things.
+1) It enables software developers and researchers to understand complicated code bases.
+2) It provides interpretable inference results that provide policy makers with predictions that are truly interpretable, in the sense that the end-user of the inference results understands what physical events led to inference outcome.
+This is critical in several domains, and is particularly important in the medical domain.
+We provide examples of both in Section~\textbf{TO ADD REF}.
 
@@
 
+\section{Experiments}
 \subsection{Epidemiology Simulators}
+
+Malaria epidemiology is governed by a complex set of drivers, 
+few of which can be understood in isolation \cite{cameron2015defining,autino_epidemiology_2012,smith2008towards,bershteyn2018implementation}.
+These include within-host dynamics, population-specific traits and even local geography.
+Comprehensive modeling of all of these components remains challenging, particularly in a region-specific context. 
+Computational epidemiology simulators have to reflect these complexities and are usually stochastic in nature.
+This can make simulation output highly non-trivial to interpret, particularly when trying to draw desired inferences coupled with observed data \cite{mwendera_challenges_2019,ferris_openmalaria_2015}.
 
@@
 Both EMOD and OpenMalaria are open source and implemented in C++.
+
+\bg{Add linking stuff now to the toy-example where hopefully we can perform inference and }
 
@@
 in EMOD, a standard scenario simulation rollout over the span of $3$ years
-with a population of size $n=5000$ \cite{smith2008towards} will generate about $ \SI{4}{\tera\byte}$ of raw trace data.
+with a population of size $n=5000$ \cite{smith2008towards} will generate about \textit{four terabytes} of raw trace data.
 Unlike with \citep{baydin2018efficient}, this amount of data cannot possibly be kept in RAM, which makes \textit{a posteriori} trace analysis very inefficient.
@@
 \centering
- \scalebox{.5}{
+ % \scalebox{.5}{
 \begin{tikzpicture}[node distance=1.5cm,
@@
   \end{tikzpicture}
-  }
   \label{fig:how}
@@
   to hijack the random number generator (RNG).}
+
   \end{figure}
@@
   % \setlength{\tabcolsep}{1mm}
-  \begin{tabularx}{0.46\textwidth}{@{}lX@{}l@{}} 
+  \begin{tabularx}{\textwidth}{@{}lX@{}l@{}} 
     \toprule
@@
     \midrule
-  A1 & [forward()+0x204; OM::Simulator::
-
-  start(scnXml::Monitoring const)+0x28a;
-
+  A1 & [forward()+0x204;OM::Simulator::start(scnXml::Monitoringconst)+0x28a;
   OM::Population::createInitialHumans()+0x94;
@@
   OM::Population::newHuman(OM::SimTime)+0x5c;
-
   OM::Host::Human::Human(OM::SimTime)+0x12b;
 
-  OM::WithinHost::WHInterface::
-
-  createWithinHostModel(double)+0x99;
-
-  OM::WithinHost::DescriptiveWithinHostModel
-  ::DescriptiveWithinHostModel(double)+0x3a;
-
-  OM::WithinHost::WHFalciparum::
-
-  WHFalciparum(double)+0xe6;
-
-  OM::util::random::
-
-  gauss(double, double)+0xb4]\_\_Normal \\
+  OM::WithinHost::WHInterface::reateWithinHostModel(double)+0x99;
+  OM::WithinHost::DescriptiveWithinHostModel::DescriptiveWithinHostModel(double)+0x3a;
+  OM::WithinHost::WHFalciparum::WHFalciparum(double)+0xe6;
+  OM::util::random::gauss(double, double)+0xb4]\_\_Normal \\
 
@@
 
- %   $\vdots$ & $\vdots$ \\
- %  A5 & [forward()+0x204; OM::Simulator::start(scnXml::Monitoring const\&)+0x468;
+  A5 & [forward()+0x204;OM::Simulator::start(scnXml::Monitoringconst\&)+0x468;
+  OM::Population::update1(OM::SimTime)+0xff;
 
- %   OM::Population::update1(OM::SimTime)+0xff;
+  OM::Host::Human::update(bool)+0x2bc;
 
- %    OM::Host::Human::update(bool)+0x2bc;
+  OM::Clinical::ClinicalModel::update(OM::Host::Human\&,double, bool)+0x96;
 
- %    OM::Clinical::ClinicalModel::update(OM::Host::Human\&, double, bool)+0x96;
+  OM::Host::NeonatalMortality::eventNeonatalMortality()+0x9;
 
- %    OM::Host::NeonatalMortality::eventNeonatalMortality()+0x9;
-
- %    OM::util::random::uniform\_01()+0xc0]\_\_Uniform\\
+  OM::util::random::uniform\_01()+0xc0]\_\_Uniform\\
+  $\vdots$ & $\vdots$ \\
 
@@
   % \setlength{\tabcolsep}{1mm}
-  \begin{tabularx}{0.46\textwidth}{@{}lX@{}} 
+  \begin{tabularx}{\textwidth}{@{}lX@{}} 
     \toprule
@@
     \midrule
-  A1 & Generate a human in the
+  A1 & Generate a human in the population within host dynamics\\
 
-   population within host dynamics\\
+  A2 & Generate another human in the population within host dynamics\\
 
-  A2 & Generate another human in
+  A3 & The population is updated and a new human, or humans, may get infected \\
 
-   the population within host dynamics\\
+  A4, A5 & Potential child deaths within the population are simulated \\
 
-  A3 & The population is updated and
+  A6 & Determines parasite density of an individual infection \\
 
-  a new human, or humans, may get infected \\
+  A7 & Models how the disease is progressing within the infected humans \\
 
-  A4, A5 & Potential child deaths
+  A8 & Models how the disease is progressing within the population \\
 
-  within the population are simulated \\
+  A9 & Models how the disease is progressing within the infected humans after the population has been updated\\
 
-  A6 & Determines parasite density
+  A10 & Full clinical update on the population for those without severe or no Malaria infection. \\
 
-   of an individual infection \\
-
-  A7 & Models how the disease is
-
-  progressing within the infected humans \\
-
-	A8 & Models how the disease is
-
-	 progressing within the population \\
-
-  A9 & Models how the disease is
-
-  progressing within the infected humans
-
-   after the population has been updated\\
-
-  A10 & Full clinical update on
-
-  the population for those without severe
-
-  or no Malaria infection. \\
-
-A11 & Full clinical update on the population
-
-for those with severe Malaria infections \\
+  A11 & Full clinical update on the population for those with severe Malaria infections \\
 \bottomrule
@@
 
-\quad\\
-\quad\\
+% \subsubsection*{Acknowledgments}
 
-\bibliographystyle{icml2019}
-\bibliography{malaria,refs}
+% Use unnumbered third level headings for the acknowledgments. All acknowledgments
+% go at the end of the paper. Do not include acknowledgments in the anonymized
+% submission, only in the final paper.
+
+% \section*{References}
+
+\bibliographystyle{plain}
+\bibliography{refs,malaria}
+
 \end{document}
-
-% This document was modified from the file originally made available by
-% Pat Langley and Andrea Danyluk for ICML-2K. This version was created
-% by Iain Murray in 2018, and modified by Alexandre Bouchard in
-% 2019. Previous contributors include Dan Roy, Lise Getoor and Tobias
-% Scheffer, which was slightly modified from the 2010 version by
-% Thorsten Joachims & Johannes Fuernkranz, slightly modified from the
-% 2009 version by Kiri Wagstaff and Sam Roweis's 2008 version, which is
-% slightly modified from Prasad Tadepalli's 2007 version which is a
-% lightly changed version of the previous year's version by Andrew
-% Moore, which was in turn edited from those of Kristian Kersting and
-% Codrina Lauth. Alex Smola contributed to the algorithmic style files.

### 2019-05-15 build-docker_host.sh

@@ -1,0 +1,5 @@
+#!/usr/bin/env bash
+docker build OMhost.Dockerfile -t bradleygh/openmalariapp.
+if [ $? -eq 0 ]; then
+  docker push bradleygh/openmalariapp
+fi

### 2019-05-15 run_server.py
auomatically run on server

@@ -1,0 +1,149 @@
+#!/usr/bin/env python3
+# -*- coding: utf-8 -*-
+
+
+import traceback
+import argparse
+import sys
+import os
+from glob import glob
+import pprint
+import uuid
+import time
+import subprocess
+import signal
+import pypdt
+import numpy as np
+import torch
+import pyprob
+from pyprob import RemoteModel, AddressDictionary, PriorInflation, InferenceEngine, InferenceNetwork, ObserveEmbedding, LearningRateScheduler
+import pyprob.diagnostics
+from pyprob.util import get_time_stamp, to_tensor, to_numpy, days_hours_mins_secs_str, progress_bar_init, progress_bar_update, progress_bar_end
+from pyprob.nn import InferenceNetworkLSTM, ProposalUniformTruncatedNormalMixture, ProposalCategoricalCategorical, OfflineDataset
+import csv
+import random
+from collections import defaultdict, OrderedDict
+from spython.main import Client
+
+parser = argparse.ArgumentParser(description='etalumis sherpa tau decay experiments', formatter_class=argparse.ArgumentDefaultsHelpFormatter)
+        parser.add_argument('--device', help='Set the compute device (cpu, cuda:0, cuda:1, etc.)', default='cpu', type=str)
+        parser.add_argument('--seed', help='Random number seed', default=None, type=int)
+        parser.add_argument('--model_executable', help='Model executable to start as a background process', default='/examples/openMalaria')
+        parser.add_argument('--model_address', help='PPX protocol address', default=None, type=str)
+        # parser.add_argument('--mode', help='', choices=['create_truths', 'prior', 'infer', 'plot', 'plot_diagnostics', 'plot_diagnostics_network', 'plot_diagnostics_addresses', 'save_numpy', 'save_dataset', 'train'], nargs='?', default='', type=str)
+        parser.add_argument('--mode', '-m', help='Experiment mode', choices=['ground_truth', 'ground_truth_from_dataset', 'prior', 'posterior', 'plot', 'plot_log_prob', 'plot_autocorrelation', 'plot_gelman_rubin', 'plot_addresses', 'plot_traces', 'plot_graph', 'plot_network', 'diag_address_dict', 'diag_mmd', 'diag_dataset', 'combine', 'save_numpy', 'save_dataset', 'sort_dataset', 'train'], nargs='?', default=None, type=str)
+        parser.add_argument('--num_traces', '-n', help='Number of traces for various tasks', default=None, type=int)
[…]

### 2019-05-16 build-docker_client.sh
remove deleted

@@ -1,5 +1,5 @@
 #!/usr/bin/env bash
-docker build -t bradleygh/openmalariapprun .
+docker build OMhost.Dockerfile -t bradleygh/openmalariapp.
 if [ $? -eq 0 ]; then
-  docker push bradleygh/openmalariapprun
+  docker push bradleygh/openmalariapp
 fi

### 2019-05-16 paper/neurlps2019/main.tex
updated

@@
 \section{Introduction}
+\label{sec:intro}
+\bg{The audience is a probabilistic programming audience - so that is what we will target}
 
@@
 
-\item Outline the objectives
-\item State the importance of interpretability in the inference of the simulator
-\item Our system enables one to understand and interpret the most common paths in a program (simulation).
-\item In highlighting this a practitioner can then go back to the field and explore that parameter/s in more depth.
-\item This facilitates understanding and the construction of more detailed models and data collection strategies
+\item Talk about a goal of probabilistic programming is to simultaneously simulate and perform inference on models conditioned on the observations.
+Although universal ppls have the ability to do this, they are constrained because the simulator has to be re-written as a program inside that system; this is both infeasible for sufficiently large simulators and inefficient.
+It would however be much more efficient to turn simulators into probabilistic programs...
+then talk briefly about the foundations created by baydin et al.
+(2 paragraphs)
+\item Talk about the large body of simulators out there and briefly talk about the types of simulators
+i.e population-based, event-based, single-object based and so fourth. {1 paragraph}
+\item Talk, high level, about the notion of a posterior in terms of traces either add stuff from tom's thesis or directly cite. (1 paragraph)
+\item Talk about the experiments we perform (high- level)
 
@@
 
-\bg{The audience is a probabilistic programming audience - so that is what we will target}
+\bg{This is not an intro - it is a collection of statements - will re-write later}
+
+Simulators encode several years, if not decades, of research and development and so by their very nature are structurally complicated.
+Since the development happens over several decades, usually without documentation, understanding a code base written with legacy libraries and software is incredibly difficult and an almost impossible task for a non-expert, or expert who has not had access to the previous years of development.
+Our method of hijacking simulators builds on the work of~\cite{baydin2018efficient} and extends the framework to encapsulate a more diverse range of simulators, in particular epidemiology simulators.
+Our framework provides a simple solution and makes it easy to transform arbitrary stochastic simulators into probabilistic programs, regardless of the complexity of the simulator and the language that the simulator is written in and for posterior inference.
+This enables two things.
+1) It enables software developers and researchers to understand complicated code bases.
+2) It provides interpretable inference results that provide policy makers with predictions that are truly interpretable, in the sense that the end-user of the inference results understands what physical events led to inference outcome.
+This is critical in several domains, and is particularly important in the medical domain.
+We provide examples of both in Section~\textbf{TO ADD REF}.
 
@@
 Our approach is then demonstrated and analysed in the context of a Malaria case study (Section \ref{sec:casestudy}).
-\label{sec:related}
 
-\section{Epidemiology Simulators and Probabilistic Programming}
-\label{sec:background}
-This will kind of be like a background section.
+\section{Background}
+\label{sec:backgorund}
 
-  \begin{algorithm}
-   \caption{How to perform posterior inference over a stochastic simulator with 
-   }
-    \begin{algorithmic}[1]
-      \Function{Merge}{$A,p,q,r$}\Comment{Where A - array, p - left, q - middle, r - right}
-
-        \State replace raw samples statements 
-        \State ${n_2} = r - q$
-        \State Create Containerized Environment
-        \For{$i = 1$ to ${n_1}$}
-            \State $L[i] = A[p + i - 1]$
-        \EndFor
-        \For{$j = 1$ to ${n_2}$}
-            \State $R[i] = A[q + j]$
-        \EndFor
-        \State $L[{n_1} + 1] =  \infty $
-        \State $R[{n_2} + 1] =  \infty $
-        \State $i = 1$
-       \EndFunction
-
-\end{algorithmic}
-\end{algorithm}
-
-\section{Types of Models and Probabilistic Programming Languages}
-
-To bridge the gap between the statistics and machine learning community to the probabilistic programming community we provide a useful overview to expose our ideas more concretely.
-
-Probabilistic programming languages, broadly speaking\footnote{For a more comprehensive overview see \cite{rainforth2017automating}}, can be classfied into three categories, each type of langauge extends a particular model space, enabling an increasingly complex set models to be represented as programs, but in doing so the notion of posterior and what it means to do inference over the model, at least in a stastically traditional sense, can change.
-% \begin{figure}
-% \centering
-% \includegraphics[]{../models/types_of_pp.png}
-% \end{figure}
-The first type of langauges are called first-order languages where the priors are well defined and the \textit{joint} program density constructed via the program directly represent the unormalised stastical joint density $p(x,y)$
-\section{Methodology}
- TODO
-
-Simulators encode several years, if not decades, of research and development and so by their very nature are structurally complicated.
-Since the development happens over several decades, usually without documentation, understanding a code base written with legacy libraries and software is incredibly difficult and an almost impossible task for a non-expert, or expert who has not had access to the previous years of development.
-Our method of hijacking simulators builds on the work of~\cite{baydin2018efficient} and extends the framework to encapsulate a more diverse range of simulators, in particular epidemiology simulators.
-Our framework provides a simple solution and makes it easy to transform arbitrary stochastic simulators into probabilistic programs, regardless of the complexity of the simulator and the language that the simulator is written in\footnote{The framework currently supports 9 popular languages, but there is nothing to stop this being extended to any language of interest.}.
-In addition to this, we can understand the structure of programs in simulators for both decoding \emph{black-box structures} and for posterior inference.
-This enables two things.
-1) It enables software developers and researchers to understand complicated code bases.
-2) It provides interpretable inference results that provide policy makers with predictions that are truly interpretable, in the sense that the end-user of the inference results understands what physical events led to inference outcome.
-This is critical in several domains, and is particularly important in the medical domain.
-We provide examples of both in Section~\textbf{TO ADD REF}.
-
-\section{Simulating Diseases}
- \label{sec:background}
-
-% Stochastic simulators are common across a number of domains, statistical physics~\cite{landau_binder_2014}, financial modeling~\cite{jackel2002monte},
-% weather prediction~\cite{evensen1994sequential}, epidemiology~\cite{smith2008towards} and many others. 
-% An advantage of using stochastic simulators is that they provide a level of interpretability not found in modern deep learning settings, as they directly incorporate model structure from carefully reasoned observations and experiments.
-In addition to this,
-
+\begin{itemize}
+\item Paragraph on Probabilistic programming
+\item Paragraph on Simulators keep high-level focus on pop sims later
+\item Paragraph on Baydin et al
+\end{itemize}
 In-silico simulators have become a crucial tool in evidence-based decision-making within a large number of disciplines, including statistical physics~\cite{landau_binder_2014}, financial modeling~\cite{jackel2002monte},
@@
 
-\section{Experiments}
-\subsection{Epidemiology Simulators}
-
-Malaria epidemiology is governed by a complex set of drivers, 
-few of which can be understood in isolation \cite{cameron2015defining,autino_epidemiology_2012,smith2008towards,bershteyn2018implementation}.
-These include within-host dynamics, population-specific traits and even local geography.
-Comprehensive modeling of all of these components remains challenging, particularly in a region-specific context. 
-Computational epidemiology simulators have to reflect these complexities and are usually stochastic in nature.
-This can make simulation output highly non-trivial to interpret, particularly when trying to draw desired inferences coupled with observed data \cite{mwendera_challenges_2019,ferris_openmalaria_2015}.
-
-Two the most advanced Malaria simulators, namely EMOD~\cite{bershteyn2018implementation} and OpenMalaria  \cite{smith2008towards}, have proven to be particularly valuable to policymakers.
-OpenMalaria is based on microsimulations of Plasmodium falciparum in humans and was originally developed to simulate the impacts of malaria vaccines within simple villages or districts
-Compared with OpenMalaria, EMOD is able to simulate a variety of additional drivers, including complex geographies complete with migration and a large number of policy interventions
-Both EMOD and OpenMalaria are open source and implemented in C++.
-
-\bg{Add linking stuff now to the toy-example where hopefully we can perform inference and }
-
-% Simulators provide users with the ability to model complex environments and problems in a black-box way, enabling new insights to be generated and predictions to be made.
-A simulator typically encodes several years, if not decades, of research and development and so by their very nature are structurally complicated.
-
-% Within the context of epidemiology simulators we can model the treatment, infection and prevention cycle under a variety of different environmental conditions, enabling predictions to be made on a set of field observations.
-Such scenarios in practices would be
-
-% This enables practitioners to model a wide array of different scenarios at ease.
-% However, since the development happens over several years, or decades, usually without comprehensive documentation, understanding a code base written with legacy libraries and software is incredibly difficult and an almost impossible task for a non-expert, or expert who has not had access to the previous years of development.
-%
-
-% Two prominent simulators for modeling Malaria are OpenMalaria~\cite{smith2008towards} and Emod~\cite{bershteyn2018implementation}, which are used by practitioners to model several aspects of the Malaria disease.
-
-%  \paragraph*{OpenMalaria} 
-% is an open-source simulator funded by the Gates Foundation to model different aspects of the malaria prevention and treatment cycle for varying population dynamics.
-%  Users specify models in OpenMalaria via a scenario file, in the form of an XML document, which determines the specifics of a particular treatment, or prevention strategy, climatic factors, the demographics of the population and so forth.
-In addition to this, field observations can be encoded inside of, or externally to the scenario file.
-% The scenario file encodes all input and output parameters, some of which we wish to simulate, others remain fixed.
-
-% \paragraph{EMOD} provides the modularity and structural flexibility necessary for modeling multiple infectious diseases at the within-host and population level. 
-% EMOD can be used to model HIV, Tuberculosis, Malaria and generic diseases, but for the purposes of the comparison, we just focus on Malaria. 
-% To specify models in EMOD a configuration file is written by the user, which is in the form of a JSON and provides similar functionality to OpenMalaria, in terms of what can be specified and the outputs that it produces.
-
-% In both these well used simulators users are accustomed to writing their models as either XML, or JSON files and so it is important that any system built around this framework enables the models to be inputted in the same format, without requiring the users to learn how to use a new set of protocols.
-Our system facilitates this as we shall discuss in the proceeding sections.
-
-\section{Hijacking Simulators}
-\label{sec:hijackingsimulators}
-
 Probabilistic programming~\cite{gordon2014probabilistic,staton2016semantics,kozen1979semantics} 
@@
 Once a probabilistic model has been expressed in a probabilistic programming language, a wide range of inference techniques, such as Markov Chain Monte Carlo (MCMC)\cite{}, Variational Inference~(VI)\cite{} and Deep Neural Networks~(DNN)\cite{},  can be used by non-experts in an automated fashion.
+
+% \section{Types of Models and Probabilistic Programming Languages}
+
+% To bridge the gap between the statistics and machine learning community to the probabilistic programming community we provide a useful overview to expose our ideas more concretely.
+
+% Probabilistic programming languages, broadly speaking\footnote{For a more comprehensive overview see \cite{rainforth2017automating}}, can be classfied into three categories, each type of langauge extends a particular model space, enabling an increasingly complex set models to be represented as programs, but in doing so the notion of posterior and what it means to do inference over the model, at least in a stastically traditional sense, can change.
+% % \begin{figure}
+% % \centering
+% % \includegraphics[]{../models/types_of_pp.png}
+% % \end{figure}
+% The first type of langauges are called first-order languages where the priors are well defined and the \textit{joint} program density constructed via the program directly represent the unormalised stastical joint density $p(x,y)$
+% \section{Methodology}
+%  TODO
+
+\section{Hijacking Simulators}
 
@@
 Sampling from this distribution is then done by requesting the PPS to send a sample back to the simulator.
+\bg{This will be extended}
+
+ \label{sec:methods}
+
+  \begin{algorithm}
+   \caption{How to perform posterior inference over a stochastic simulator with 
+   }
+    \begin{algorithmic}[1]
+      \Function{Convert Simulator to a Probabilistic Program}{$A$}\Comment{}
+
+        \State replace raw samples statements $f_{t}(x_{t} | \eta_{t}) $
+        \State Create thin wrapper
+        \State Create Containerized Environment $C_{i}$
+        \State Create Connection
+        \State Run Program
+       \EndFunction
+       \Function{Create thin wrapper}{$f_{t}(x_{t} | \eta_{t})$}
+       \State to add
+       \EndFunction
+       \Function{Create Containerized Environment }{$\Delta_{i}$}\Comment{Where $\Delta_{i}$ is one of $i$-th dependencies}
+       \State to add
+       \EndFunction
+
+\end{algorithmic}
+\end{algorithm}
+
+% Stochastic simulators are common across a number of domains, statistical physics~\cite{landau_binder_2014}, financial modeling~\cite{jackel2002monte},
+% weather prediction~\cite{evensen1994sequential}, epidemiology~\cite{smith2008towards} and many others. 
+% An advantage of using stochastic simulators is that they provide a level of interpretability not found in modern deep learning settings, as they directly incorporate model structure from carefully reasoned observations and experiments.
+In addition to this,
 
@@
  %We provide examples of (1) in next Section~\ref{sec:casestudy} and discuss the latter in future work as extensions to the the current framework.
+
+\section{Experiments}
+\subsection{Epidemiology Simulators}
+
+Malaria epidemiology is governed by a complex set of drivers, 
+few of which can be understood in isolation \cite{cameron2015defining,autino_epidemiology_2012,smith2008towards,bershteyn2018implementation}.
+These include within-host dynamics, population-specific traits and even local geography.
+Comprehensive modeling of all of these components remains challenging, particularly in a region-specific context. 
+Computational epidemiology simulators have to reflect these complexities and are usually stochastic in nature.
+This can make simulation output highly non-trivial to interpret, particularly when trying to draw desired inferences coupled with observed data \cite{mwendera_challenges_2019,ferris_openmalaria_2015}.
+
+Two the most advanced Malaria simulators, namely EMOD~\cite{bershteyn2018implementation} and OpenMalaria  \cite{smith2008towards}, have proven to be particularly valuable to policymakers.
+OpenMalaria is based on microsimulations of Plasmodium falciparum in humans and was originally developed to simulate the impacts of malaria vaccines within simple villages or districts
+Compared with OpenMalaria, EMOD is able to simulate a variety of additional drivers, including complex geographies complete with migration and a large number of policy interventions
+Both EMOD and OpenMalaria are open source and implemented in C++.
+
+\bg{Add linking stuff now to the toy-example where hopefully we can perform inference and }
+
+% Simulators provide users with the ability to model complex environments and problems in a black-box way, enabling new insights to be generated and predictions to be made.
+A simulator typically encodes several years, if not decades, of research and development and so by their very nature are structurally complicated.
+
+% Within the context of epidemiology simulators we can model the treatment, infection and prevention cycle under a variety of different environmental conditions, enabling predictions to be made on a set of field observations.
+Such scenarios in practices would be
+
+% This enables practitioners to model a wide array of different scenarios at ease.
+% However, since the development happens over several years, or decades, usually without comprehensive documentation, understanding a code base written with legacy libraries and software is incredibly difficult and an almost impossible task for a non-expert, or expert who has not had access to the previous years of development.
+%
+
+% Two prominent simulators for modeling Malaria are OpenMalaria~\cite{smith2008towards} and Emod~\cite{bershteyn2018implementation}, which are used by practitioners to model several aspects of the Malaria disease.
+
+%  \paragraph*{OpenMalaria} 
+% is an open-source simulator funded by the Gates Foundation to model different aspects of the malaria prevention and treatment cycle for varying population dynamics.
+%  Users specify models in OpenMalaria via a scenario file, in the form of an XML document, which determines the specifics of a particular treatment, or prevention strategy, climatic factors, the demographics of the population and so forth.
+In addition to this, field observations can be encoded inside of, or externally to the scenario file.
+% The scenario file encodes all input and output parameters, some of which we wish to simulate, others remain fixed.
+
+% \paragraph{EMOD} provides the modularity and structural flexibility necessary for modeling multiple infectious diseases at the within-host and population level. 
+% EMOD can be used to model HIV, Tuberculosis, Malaria and generic diseases, but for the purposes of the comparison, we just focus on Malaria. 
+% To specify models in EMOD a configuration file is written by the user, which is in the form of a JSON and provides similar functionality to OpenMalaria, in terms of what can be specified and the outputs that it produces.
+
+% In both these well used simulators users are accustomed to writing their models as either XML, or JSON files and so it is important that any system built around this framework enables the models to be inputted in the same format, without requiring the users to learn how to use a new set of protocols.
+Our system facilitates this as we shall discuss in the proceeding sections.
 
@@
 %\bg{Add performance plot showing memory consumptions of pyprob and our new system.}
-
-\section{Case Study: Ifakara (Tanzania)}
+\subsection{Toy Model}
+\subsection{Real-world Example}
 \label{sec:casestudy}

### 2019-05-16 paper/neurlps2019/main.tex
updated

@@
 However, with the myriad of probabilistic programming systems (PPSs) and the complexities involved in running 
-each individual system, such as specifying a model in the given language and $\backslash$ or running inference, which means it can be difficult for users to take advantage of the system. 
-In addition to this, even though there exists a myriad of simulators across many scientific domains 
-and within industry, none can be used within existing PPSs without having to re-implement the simulator in the given PPS, something that is not typically feasible, nor practical.  
+each individual system, such as specifying a model in the given language, means that it can be difficult to re-implement the myriad of scientific and industrial simulators within an existing PPSs, something that is not typically feasible, nor practical.
 However, recent work
 by \cite{baydin2018efficient} demonstrated a pathway for how to turn a particular
-type of simulator, written in a set of 9 languages, into a probabilistic program via the raw random
-number calls.
+type of event-based simulators into a probabilistic program. 
+% via the raw random number calls of the simulator.
  But, this still meant that a large class of critically important population-based simulators could not be turned into probabilistic programs.
-We, in this work, extend that framework to encompass population-based simulators, a very large class of simulators that are used extensively across epidemiology modeling, multi-agent modeling, financial modeling and so forth.
+We, in this work, extend that framework to encompass population-based simulators, a very large class of simulators that are used extensively across epidemiology modeling, multi-agent modeling and financial modeling.
 In addition to this, we enable simulators written in 13 different languages to be easily converted in probabilistic programs, which we algorithmically outline.
-We demonstrate our PPS on a both a toy population simulator and two important malaria population-based simulators called OpenMalaria and EMOD.
+We demonstrate inference performed with our PPS on a both a toy population simulator and two important malaria population-based simulators called OpenMalaria and EMOD.
 
@@
 \end{itemize}
+
+Probabilistic programming\cite{gordon2014probabilistic} is paradigm that enables users to write models in a programmatic way. 
+The source code is then transformed into a program density, representing a joint density.
+By having direct access to the source code enables Probabilistic Programming Systems~(PPS)
+to analyse the model code in unique way, enabling posterior inference
+to be performed in an automated way, which makes PPSs flexible and adaptive.
 

### 2019-05-16 paper/neurlps2019/main.tex
updated

@@
  But, this still meant that a large class of critically important population-based simulators could not be turned into probabilistic programs.
-We, in this work, extend that framework to encompass population-based simulators, a very large class of simulators that are used extensively across epidemiology modeling, multi-agent modeling and financial modeling.
+In this work, we extend that framework to encompass population-based simulators, a very large class of simulators that are used extensively across epidemiology modeling, multi-agent modeling and financial modeling.
 In addition to this, we enable simulators written in 13 different languages to be easily converted in probabilistic programs, which we algorithmically outline.
-We demonstrate inference performed with our PPS on a both a toy population simulator and two important malaria population-based simulators called OpenMalaria and EMOD.
+We demonstrate inference performed with our PPS on both a toy population simulator and two important malaria population-based simulators called OpenMalaria and EMOD.
 
@@
 
-Probabilistic programming\cite{gordon2014probabilistic} is paradigm that enables users to write models in a programmatic way. 
-The source code is then transformed into a program density, representing a joint density.
-By having direct access to the source code enables Probabilistic Programming Systems~(PPS)
-to analyse the model code in unique way, enabling posterior inference
-to be performed in an automated way, which makes PPSs flexible and adaptive.
+Probabilistic programming\cite{gordon2014probabilistic} is paradigm that enables users to write, and perform inference in, complex probabilistic models in a programmatic, yet efficient and simplified way.
+The source code of the given model, or equivalently \emph{program}, is then transformed into a program density, representing the joint density of the model.
+By having direct access to the source code we can perform interesting 
+operations within a probabilistic programming system~(PPS),
+such as analysing the model code, which enables statistically correct inference \cite{gram2018hamiltonian} in an automated way \bg{Add citations - black box VI etc} to be performed. 
+In addition to this, as models can be directly expressed in program code, we can encode other structures make us of the full structure of a given language.
+
+One powerful \emph{class} of models are simulators.
 

### 2019-05-17 paper/neurlps2019/main.tex
updated

@@
 % \usepackage[binary-units=true]{siunitx}
-\title{Converting Population-Based Simulators to Probabilistic Programs}
+\title{Hijacking Population-Based Simulators}
 
@@
 In addition to this, we enable simulators written in 13 different languages to be easily converted in probabilistic programs, which we algorithmically outline.
-We demonstrate inference performed with our PPS on both a toy population simulator and two important malaria population-based simulators called OpenMalaria and EMOD.
+We demonstrate inference performed with our PPS on both a toy population simulator and two important malaria population-based simulators OpenMalaria and EMOD.
 
@@
 \item Talk about a goal of probabilistic programming is to simultaneously simulate and perform inference on models conditioned on the observations.
-Although universal ppls have the ability to do this, they are constrained because the simulator has to be re-written as a program inside that system; this is both infeasible for sufficiently large simulators and inefficient.
+Although universal probabilistic programming systems have the ability to do this, they are constrained because the simulator has to be re-written as a program inside that system; this is both infeasible for sufficiently large simulators and inefficient.
 It would however be much more efficient to turn simulators into probabilistic programs...
@@
 
-Probabilistic programming\cite{gordon2014probabilistic} is paradigm that enables users to write, and perform inference in, complex probabilistic models in a programmatic, yet efficient and simplified way.
-The source code of the given model, or equivalently \emph{program}, is then transformed into a program density, representing the joint density of the model.
+Probabilistic programming\cite{gordon2014probabilistic,le2016inference} is a machine learning paradigm that enables users to write, and perform inference in, complex probabilistic models in a programmatic, yet efficient and simplified way.
+The source code of the given model, or equivalently the \emph{program}, is then transformed into a program density, which can
+be viewed as the joint density of the model.
 By having direct access to the source code we can perform interesting 
-operations within a probabilistic programming system~(PPS),
-such as analysing the model code, which enables statistically correct inference \cite{gram2018hamiltonian} in an automated way \bg{Add citations - black box VI etc} to be performed. 
-In addition to this, as models can be directly expressed in program code, we can encode other structures make us of the full structure of a given language.
-
+operations within a probabilistic programming system~(PPS) in an automated way,
+such as analysing the model code, to perform statistically correct inference \cite{gram2018hamiltonian,goodman2012church,wood2014new}.
+In addition to this, as models can be directly expressed in program code, we can encode other structures to make use of the full structure of a given language,
+which enables users to integrate PPSs into their standard model development process.
 One powerful \emph{class} of models are simulators.
 
-\bg{This is not an intro - it is a collection of statements - will re-write later}
+Simulators arise in many industrial \cite{di2015simulator,judd2009numerically,kass1990rapid} and scientific applications~\cite{smith2008towards,bershteyn2018implementation,gleisberg_event_2008} encode years, if not decades, of research and development.
+Thus, Simulators by their very nature are
+information rich and naturally describe complicated joint densities.
+However, in order to write a simulator in a standard probabilistic programming set-up one would have to re-write a simulator in the language of the PPS, which 
+in many cases is not feasible, nor efficient due to both the size 
+and complexity of the simulator code base. 
+However, what if one could arbitrary connect a simulator, written in any
+language and connect it directly to a PPS, without having to re-write
+the simulator in the given PPS?
 
-Simulators encode several years, if not decades, of research and development and so by their very nature are structurally complicated.
-Since the development happens over several decades, usually without documentation, understanding a code base written with legacy libraries and software is incredibly difficult and an almost impossible task for a non-expert, or expert who has not had access to the previous years of development.
-Our method of hijacking simulators builds on the work of~\cite{baydin2018efficient} and extends the framework to encapsulate a more diverse range of simulators, in particular epidemiology simulators.
-Our framework provides a simple solution and makes it easy to transform arbitrary stochastic simulators into probabilistic programs, regardless of the complexity of the simulator and the language that the simulator is written in and for posterior inference.
+Our method of hijacking simulators builds on the work of~\cite{baydin2018efficient} and extends the framework to encapsulate a more diverse range of simulators, in particular population-based simulators which arise
+in a myriad of domains \bg{add citations}. 
+Our framework provides a simple solution and makes it easy to transform arbitrary stochastic population-based simulators into probabilistic programs, regardless of the complexity of the simulator. 
+Our system supports simulators written in 13 different programming languages and enables the user to perform for posterior inference over the density of the simulator. 
 This enables two things.
@@
 2) It provides interpretable inference results that provide policy makers with predictions that are truly interpretable, in the sense that the end-user of the inference results understands what physical events led to inference outcome.
-This is critical in several domains, and is particularly important in the medical domain.
-We provide examples of both in Section~\textbf{TO ADD REF}.
+This is critical in several domains, and is particularly important in the medical domain. We provide examples of both in Section~\textbf{TO ADD REF}.
 
@@
 can be used to express probabilistic models and consequently perform automated inference these.
-Once a probabilistic model has been expressed in a probabilistic programming language, a wide range of inference techniques, such as Markov Chain Monte Carlo (MCMC)\cite{}, Variational Inference~(VI)\cite{} and Deep Neural Networks~(DNN)\cite{},  can be used by non-experts in an automated fashion.
+Once a probabilistic model has been expressed in a probabilistic programming language, a wide range of inference techniques, such as Markov Chain Monte Carlo (MCMC)\cite{geyer1992practical}, Variational Inference~(VI)\cite{wainwright2008graphical} and Deep Neural Networks~(DNN)\cite{Goodfellow:2016:DL:3086952},  can be used by non-experts in an automated fashion.
 
@@
 \section{Hijacking Simulators}
+\label{sec:methods}
 
-\paragraph{Hijacking a simulator} describes the process by which a simulator's random number generators are replaced by calls to external sampling procedures, which are controlled by a probabilistic programming system (PPS).
+\subsection{The Engineering}
+Hijacking a simulator a describes the process by which a simulator's random number generators are replaced by calls to external sampling procedures, which are controlled by a probabilistic programming system (PPS).
 In practice, this amounts to performing a small number of surgical incisions into the simulator's source code in order to replace built-in calls to random number generators.
@@
 
+\subsection{The Theory}
+
+\bg{Tom add theory stuff here}
 %As stated previously, Simulators are complex systems and should be viewed as black-boxes, thus deciphering the underlying processes occurring within the simulator facilitates the users understanding of how such a system operates and provides a way of interpreting the outputs in a clear way.

### 2019-05-17 code/openmalaria/model/Monitoring/Continuous.cpp
updated

@@ -231,7 +231,8 @@ namespace OM { namespace Monitoring {
     
     void ContinuousType::registerCallback (string optName, string titles,
             fastdelegate::FastDelegate1<ostream&> outputCb){
-        // This was ccommented out by bradley to allow for multiple successive traces: assert(registered.count(optName) == 0); // name clash/registered twice?
+        // This was ccommented out by bradley to allow for multiple successive traces: 
+        // assert(registered.count(optName) == 0); // name clash/registered twice?
 	registered[optName] = new Callback1( titles, outputCb );
     }
     void ContinuousType::registerCallback (string optName, string titles,

### 2019-05-17 code/openmalaria/model/mon/mon.cpp

@@ -420,7 +420,15 @@ void internal::initReporting( const scnXml::Scenario& scenario ){
     defineOutMeasures();        // set up namedOutMeasures
     // **** commented out by Bradley Gram-Hansen to enable multiple simulator runs***
     // assert(reportedMeasures.empty());
-    
+    if( reportedMeasures.empty())
+    {
+    }
+    else
+    {
+         vector<OutMeasure> reportedMeasures;
+         assert(reportedMeasures.empty());
+
+    }
     // First we put used measures in this list:
     const scnXml::MonitoringOptions& optsElt = scenario.getMonitoring().getSurveyOptions();
     // This should be an upper bound on the number of options we need:

### 2019-05-17 code/openmalaria/model/openMalaria.cpp

@@ -50,7 +50,7 @@ xt::xarray<double> forward() {
 
         // scenarioFile = util::CommandLine::parse (argc, argv);   // parse arguments
 
-        util::BoincWrapper::init();     // BOINC init
+        // util::BoincWrapper::init();     // BOINC init
 
         // Load the scenario document:
         scenarioFile = util::CommandLine::lookupResource (scenarioFile);

### 2019-05-19 paper/neurlps2019/main.tex
minor
updated

@@
 % [preprint] option:
-%     \usepackage[preprint]{neurips_2018}
+\usepackage[preprint]{neurips_2019}
 
 % to compile a camera-ready version, add the [final] option, e.g.:
-     \usepackage[final]{neurips_2019}
+     % \usepackage[final]{neurips_2019}
 
@@
 % \usepackage[binary-units=true]{siunitx}
-\title{Hijacking Population-Based Simulators}
+\title{Hijacking Simulators with Universal Probabilistic Programming}
 
@@
 Probabilistic programming provides a way to perform statistical inference over simulations of events in a programmatic way.
-However, with the myriad of probabilistic programming systems (PPSs) and the complexities involved in running 
-each individual system, such as specifying a model in the given language, means that it can be difficult to re-implement the myriad of scientific and industrial simulators within an existing PPSs, something that is not typically feasible, nor practical.
+Thus, simulators by design are ideal programs for probabilistic programming systems~(PPSs).
+However, within existing PPSs one would have to re-implement the simulator via the PPS language specification, which is inefficient and often not feasible due to the complexity of such scientific and industrial simulators.
 However, recent work
-by \cite{baydin2018efficient} demonstrated a pathway for how to turn a particular
-type of event-based simulators into a probabilistic program. 
+by \cite{baydin2018efficient} demonstrated a pathway to turn a particular
+type of event-based simulator into a probabilistic program. 
 % via the raw random number calls of the simulator.
- But, this still meant that a large class of critically important population-based simulators could not be turned into probabilistic programs.
+ But, this still meant that a large class of critically important population-based simulators could not be turned into probabilistic programs and as such could not be used in a probabilistic programming framework.
 In this work, we extend that framework to encompass population-based simulators, a very large class of simulators that are used extensively across epidemiology modeling, multi-agent modeling and financial modeling.
-In addition to this, we enable simulators written in 13 different languages to be easily converted in probabilistic programs, which we algorithmically outline.
-We demonstrate inference performed with our PPS on both a toy population simulator and two important malaria population-based simulators OpenMalaria and EMOD.
+We enable simulators written in 13 different languages to be easily converted to probabilistic programs, for which we algorithmically outline how to do and provide examples of.
+We demonstrate our method on three models and provide n infected population simulator and demonstrate the hijacking on two important malaria population-based simulators OpenMalaria and EMOD.
 
@@
 
-\begin{itemize}
+% \begin{itemize}
 
-\item Talk about a goal of probabilistic programming is to simultaneously simulate and perform inference on models conditioned on the observations.
+% \item Talk about a goal of probabilistic programming is to simultaneously simulate and perform inference on models conditioned on the observations.
 Although universal probabilistic programming systems have the ability to do this, they are constrained because the simulator has to be re-written as a program inside that system; this is both infeasible for sufficiently large simulators and inefficient.
@@
 (2 paragraphs)
-\item Talk about the large body of simulators out there and briefly talk about the types of simulators
-i.e population-based, event-based, single-object based and so fourth. {1 paragraph}
-\item Talk, high level, about the notion of a posterior in terms of traces either add stuff from tom's thesis or directly cite. (1 paragraph)
-\item Talk about the experiments we perform (high- level)
+% \item Talk about the large body of simulators out there and briefly talk about the types of simulators
+% i.e population-based, event-based, single-object based and so fourth. {1 paragraph}
+% \item Talk, high level, about the notion of a posterior in terms of traces either add stuff from tom's thesis or directly cite. (1 paragraph)
+% \item Talk about the experiments we perform (high- level)
 
-\end{itemize}
+% \end{itemize}
 
@@
 be viewed as the joint density of the model.
-By having direct access to the source code we can perform interesting 
-operations within a probabilistic programming system~(PPS) in an automated way,
-such as analysing the model code, to perform statistically correct inference \cite{gram2018hamiltonian,goodman2012church,wood2014new}.
+Having direct access to the source code enables probabilistic programming system~(PPS)
+to conduct several processes in an automated fashion, such as 
+automating inference \cite{zhou2019lf,goodman2012church,wood2014new}.
 In addition to this, as models can be directly expressed in program code, we can encode other structures to make use of the full structure of a given language,
@@
 
+We first demonstrate our system on a reduced version of the
+
 Two the most advanced Malaria simulators, namely EMOD~\cite{bershteyn2018implementation} and OpenMalaria  \cite{smith2008towards}, have proven to be particularly valuable to policymakers.
@@
 \begin{figure}
-\centering
- % \scalebox{.5}{
-\begin{tikzpicture}[node distance=1.5cm,
-    every node/.style={fill=white, font=\sffamily}, align=center]
-  % Specification of nodes (position, etc.)
-  % \node (start)             [activityStarts]              {Activity starts};
-  \node (SimulatorSystem)     [activityStarts]          {Simulator System \\ e.g. EMOD};
-   \node (randomprim)      [process, left of=SimulatorSystem, xshift=5.6cm]
-                                                      {Redefine RNG \\ to make \\ externally \\ callable};
-  \node (onStartBlock)      [process, below of=SimulatorSystem]   {Specify Model \\ e.g. XML files};
-  \node (activityRuns)      [activityRuns, below of=onStartBlock]
-                                                      {Run Simulator};
-  \node (onPauseBlock)      [startstop, below of=activityRuns, yshift=-1cm]
-                                                                {Posterior Output};
-  % \node (onStopBlock)       [process, below of=onPauseBlock, yshift=-1cm]
-  %                                                                {};
-  % \node (onDestroyBlock)    [process, below of=onStopBlock, yshift=-1cm] 
-  %                                                             {onDestroy()};
-  \node (onRestartBlock)    [process, right of=activityRuns, xshift=3cm]
-                                                              {PPS hijacks \\ random number \\ generator};
-  \node (ActivityEnds)      [startstop, below of=onRestartBlock, yshift=-1cm]
-                                                        {Result summaries \\ path probabilities};
-  % \node (ActivityDestroyed) [startstop, below of=onDestroyBlock]
-  %                                                   {Activity is shut down};     
-  % Specification of lines between nodes specified above
-  % with aditional nodes for description 
-  % \draw[->]             (start) -- (onCreateBlock);
-  \draw[->]     (SimulatorSystem) -- (onStartBlock);
-  \draw[dashed, <-]  (SimulatorSystem) -- (randomprim);
-  % \draw[->]      (onStartBlock) -- (onResumeBlock);
-  \draw[->]     (onStartBlock) -- (activityRuns);
-  \draw[->]      (activityRuns) -- (onPauseBlock);
-  % \draw[->]      (onPauseBlock) -- node {The activity is no longer visible}
-  %                                  (onStopBlock);
-  % \draw[->]       (onStopBlock) -- node {The activity is shut down by
-  %                                  user or system} (onDestroyBlock);
-  % \draw[->]    (onRestartBlock) -- (activityRuns);
-  \path[->, shift left=.75ex]
-  (onRestartBlock) edge (activityRuns)
-  (activityRuns) edge (onRestartBlock);
-  % \draw[<-]    (activityRuns) -- (onRestartBlock);
-  % \draw[->]       (onStopBlock) -| node[yshift=1.25cm, text width=3cm]
-  %                                  {The activity comes to the foreground}
-  %                                  (onRestartBlock);
-
-  \draw[->]    (onRestartBlock) -- (ActivityEnds);
-  % \draw[->]      (onPauseBlock) -| node(priorityXMemory)
-  %                                  {todo}
-  %                                  (ActivityEnds);
-  % \draw           (onStopBlock) -| (priorityXMemory);
-  % \draw[->]     (ActivityEnds)  |- node [yshift=-2cm, text width=3.1cm]
-  %                                   {User navigates back to the activity}
-  %                                   (SimulatorSystem);
-  % \draw[->] (onPauseBlock.east) -- ++(2.6,0) -- ++(0,2) -- ++(0,2) --                
-  %    node[xshift=1.2cm,yshift=-1.5cm, text width=2.5cm]
-  %    {The activity comes to the foreground}(onResumeBlock.east);
-  \end{tikzpicture}
+  \includegraphics[width=\textwidth]{../plots/system.pdf}
   \label{fig:how}
@@
 %\bg{Add performance plot showing memory consumptions of pyprob and our new system.}
-\subsection{Toy Model}
 \subsection{Real-world Example}

### 2019-05-19 paper/main.tex

--- 2019-03-13 paper/main.tex
@@
 \usepackage{microtype}      % microtypography
+\usepackage{listings}
+\usepackage{tabularx}
+\usepackage{listings}
+\usepackage{pgfplots}
+\usepackage{tikz}
+\usepackage{todonotes}
+\usepackage{tikz-cd}
+\usepackage{algorithm}
+\usepackage{algpseudocode}
+% \usepackage[binary-units=true]{siunitx}
+\title{Hijacking Simulators with Universal Probabilistic Programming}
 
-\title{Applied Interpretable Inference: Connecting Probabilistic Programming with Epidemology Simulators}
+% neurlps classification
+% Primary: Applications
+% Secodnary cat:Data, Challenges, Implementations, and Software
+% or Hijacking Malaria Simulators with Probabilistic Programming
 
@@
 % \AND instead of \And before the third author name.
+\newcommand{\bg}[1]{~{{[{\it \textcolor{red}{{\bf BG:} #1}}]}}}
+\newcommand{\cs}[1]{~{{[{\it \textcolor{red}{{\bf CS:} #1}}]}}}
+\newcommand{\ag}[1]{~{{[{\it \textcolor{red}{{\bf AG:} #1}}]}}}
+% If accepted, instead use the following line for the camera-ready submission:
+%\usepackage[accepted]{icml2019}
+\usetikzlibrary{arrows,chains,matrix,positioning,scopes, arrows.meta}
+\tikzset{
+  shift left/.style ={commutative diagrams/shift left={#1}},
+  shift right/.style={commutative diagrams/shift right={#1}}
+}
+\tikzset{%
+  >={Latex[width=2mm,length=2mm]},
+  % Specifications for style of nodes:
+            base/.style = {rectangle, rounded corners, draw=black,
+                           minimum width=4cm, minimum height=1cm,
+                           text centered, font=\sffamily},
+  activityStarts/.style = {base, fill=blue!50},
+       startstop/.style = {base, fill=red!50},
+    activityRuns/.style = {base, fill=green!50},
+         process/.style = {base, minimum width=2.5cm, fill=orange!55,
+                           font=\ttfamily},
+}
+\makeatletter
+\tikzset{join/.code=\tikzset{after node path={%
+\ifx\tikzchainprevious\pgfutil@empty\else(\tikzchainprevious)%
+edge[every join]#1(\tikzchaincurrent)\fi}}}
+\makeatother
+%
+\tikzset{>=stealth',every on chain/.append style={join},
+         every join/.style={->}}
+\tikzstyle{labeled}=[execute at begin node=$\scriptstyle,
+   execute at end node=$]
 
+\lstset{language=C++,
+                basicstyle=\footnotesize,
+                keywordstyle=\color{blue}\ttfamily,
+                stringstyle=\color{red}\ttfamily,
+                commentstyle=\color{green}\ttfamily,
+                morecomment=[l][\color{magenta}]{\#},
+                breaklines = true
+}
+\pgfplotsset{width=7cm,compat=1.8}
 \author{%
-  David S.~Hippocampus\thanks{Use footnote for providing further information
-    about author (webpage, alternative address)---\emph{not} for acknowledging
-    funding agencies.} \\
-  Department of Computer Science\\
-  Cranberry-Lemon University\\
-  Pittsburgh, PA 15213 \\
-  \texttt{hippo@cs.cranberry-lemon.edu} \\
+Paper ID: }
   % examples of more authors
@@
 
-A goal of probabilistic programming is to couple simulators, with inference. This is 
-because stochastic simulators are used prominently in many industrial settings,
-do not require one to construct hand-crafted joint distributions as they implicitly 
-define a joint distribution of the program and encode learnt structures 
-directly. This makes simulators powerful tools and much of machine learning (ML) and 
-Artificial Intelligence (AI)
-can be seen as trying to emulate such simulators from a purely data-driven approach.
-However, in the 
-ML/AI setting, although we can often infer outcomes, we have little understanding about what 
-in the data led to the outputted inferences. 
-This makes it challenging to deploy ML/AI systems into the wild, especially in health-related and safety-critical domains, such
-as epidemiology, as we lose \emph{interpretability}. 
-In this work, we explain how to design ML/AI systems that combine
-probabilistic programming systems (PPSs) and epidemiology simulators, to extract
-fully interpretable posterior structures, enabling policy makers 
-and practitioners to make interpretable inferences. 
-In particular, we demonstrate this for the Malaria disease and show how we can 
-perform interpretable inference in such settings.
+% A goal of probabilistic programming is to couple simulators, with inference. This is 
+% because stochastic simulators are used prominently in many industrial settings,
+% do not require one to construct hand-crafted joint distributions as they implicitly 
+% define a joint distribution of the program and encode learnt structures 
+% directly. This makes simulators powerful tools and much of machine learning (ML) and 
+% Artificial Intelligence (AI)
+% can be seen as trying to emulate such simulators from a purely data-driven approach.
+% However, in the 
+% ML/AI setting, although we can often infer outcomes, we have little understanding about what 
+% in the data led to the outputted inferences. 
+% This makes it challenging to deploy ML/AI systems into the wild, especially in health-related and safety-critical domains, such
+% as epidemiology, as we lose \emph{interpretability}. 
+% In this work, we explain how to design ML/AI systems that combine
+% probabilistic programming systems (PPSs) and epidemiology simulators, to extract
+% fully interpretable posterior structures, enabling policy makers 
+% and practitioners to make interpretable inferences. 
+% In particular, we demonstrate this for the Malaria disease and show how we can 
+% perform interpretable inference in such settings.
+
+Probabilistic programming provides a way to perform statistical inference over simulations of events in a programmatic way.
+Thus, simulators by design are ideal programs for probabilistic programming systems~(PPSs).
+However, within existing PPSs one would have to re-implement the simulator via the PPS language specification, which is inefficient and often not feasible due to the complexity of such scientific and industrial simulators.
+However, recent work
+by \cite{baydin2018efficient} demonstrated a pathway to turn a particular
+type of event-based simulator into a probabilistic program. 
+% via the raw random number calls of the simulator.
+ But, this still meant that a large class of critically important population-based simulators could not be turned into probabilistic programs and as such could not be used in a probabilistic programming framework.
+In this work, we extend that framework to encompass population-based simulators, a very large class of simulators that are used extensively across epidemiology modeling, multi-agent modeling and financial modeling.
+We enable simulators written in 13 different languages to be easily converted to probabilistic programs, for which we algorithmically outline how to do and provide examples of.
+We demonstrate our method on three models and provide n infected population simulator and demonstrate the hijacking on two important malaria population-based simulators OpenMalaria and EMOD.
+
+\todo[inline]{We need to have probabilistic programming in the title to make sure we attract the correct reviewers}
+
+% Epidemiology simulations have become a fundamental tool in the fight
+% against the epidemics of various infectious diseases like AIDS and
+% malaria.
+% However, the complicated and stochastic nature of these simulators
+% can mean their output is difficult to interpret, which reduces their usefulness to policymakers.
+% In this paper, we introduce an approach that allows one to treat a large class of population-based epidemiology simulators as probabilistic generative models.
+%  This is achieved by \emph{hijacking} the internal random number generator calls, through the use of an universal probabilistic programming system (PPS). 
+%  In contrast to other methods, our approach can be easily retrofitted to simulators written in popular industrial programming frameworks. 
+% We demonstrate that our method can be used for interpretable introspection and inference, thus shedding light on black-box simulators. 
+% This reinstates much needed trust between policymakers and evidence-based methods. 
 \end{abstract}
@@
 \section{Introduction}
+\label{sec:intro}
+\bg{The audience is a probabilistic programming audience - so that is what we will target}
 
-To do.
+% \begin{itemize}
+
+% \item Talk about a goal of probabilistic programming is to simultaneously simulate and perform inference on models conditioned on the observations.
+Although universal probabilistic programming systems have the ability to do this, they are constrained because the simulator has to be re-written as a program inside that system; this is both infeasible for sufficiently large simulators and inefficient.
+It would however be much more efficient to turn simulators into probabilistic programs...
+then talk briefly about the foundations created by baydin et al.
+(2 paragraphs)
+% \item Talk about the large body of simulators out there and briefly talk about the types of simulators
+% i.e population-based, event-based, single-object based and so fourth. {1 paragraph}
+% \item Talk, high level, about the notion of a posterior in terms of traces either add stuff from tom's thesis or directly cite. (1 paragraph)
+% \item Talk about the experiments we perform (high- level)
+
+% \end{itemize}
+
+Probabilistic programming\cite{gordon2014probabilistic,le2016inference} is a machine learning paradigm that enables users to write, and perform inference in, complex probabilistic models in a programmatic, yet efficient and simplified way.
+The source code of the given model, or equivalently the \emph{program}, is then transformed into a program density, which can
+be viewed as the joint density of the model.
+Having direct access to the source code enables probabilistic programming system~(PPS)
+to conduct several processes in an automated fashion, such as 
+automating inference \cite{zhou2019lf,goodman2012church,wood2014new}.
+In addition to this, as models can be directly expressed in program code, we can encode other structures to make use of the full structure of a given language,
+which enables users to integrate PPSs into their standard model development process.
+One powerful \emph{class} of models are simulators.
+
+Simulators arise in many industrial \cite{di2015simulator,judd2009numerically,kass1990rapid} and scientific applications~\cite{smith2008towards,bershteyn2018implementation,gleisberg_event_2008} encode years, if not decades, of research and development.
+Thus, Simulators by their very nature are
+information rich and naturally describe complicated joint densities.
+However, in order to write a simulator in a standard probabilistic programming set-up one would have to re-write a simulator in the language of the PPS, which 
+in many cases is not feasible, nor efficient due to both the size 
+and complexity of the simulator code base. 
+However, what if one could arbitrary connect a simulator, written in any
+language and connect it directly to a PPS, without having to re-write
+the simulator in the given PPS?
+
+Our method of hijacking simulators builds on the work of~\cite{baydin2018efficient} and extends the framework to encapsulate a more diverse range of simulators, in particular population-based simulators which arise
+in a myriad of domains \bg{add citations}. 
+Our framework provides a simple solution and makes it easy to transform arbitrary stochastic population-based simulators into probabilistic programs, regardless of the complexity of the simulator. 
+Our system supports simulators written in 13 different programming languages and enables the user to perform for posterior inference over the density of the simulator. 
+This enables two things.
+1) It enables software developers and researchers to understand complicated code bases. 
+2) It provides interpretable inference results that provide policy makers with predictions that are truly interpretable, in the sense that the end-user of the inference results understands what physical events led to inference outcome.
+This is critical in several domains, and is particularly important in the medical domain. We provide examples of both in Section~\textbf{TO ADD REF}.
+
+In this paper, we introduce a novel method that allows one to shed light on the inner workings of a large class of population-based stochastic simulators.
+We achieve this by extending the work of \citep{baydin2018efficient} by interpreting such population-based simulators as probabilistic generative models within the framework of universal probabilistic programming (UPP) \cite{le-2016-inference}.
+To this end, we \emph{hijack} existing simulators by overriding their internal random number generators.
+Specifically, by replacing the existing low-level random number generator in a simulator with a call to a purpose-built UPP ``controller'', which can thus control, track and manipulate the stochasticity of the simulator.
+
+This allows for a variety of tasks to be performed on
+the hijacked simulator, such as running inference (by conditioning
+the values of certain draws and manipulating others),
+uncovering stochastic structure, and automatically
+producing result summaries, such as establishing the probability
+of different program paths/traces.  By providing a common abstraction 
+framework for different simulators, our approach further allows for
+easy and direct comparison between related or competing
+simulators, a characteristic that is valuable in the context of
+epidemiology simulators \cite{ferris_openmalaria_2015}. We provide a case study of the above in Section \ref{sec:casestudy}.
+
+Our framework already supports application to simulators written in 13 industrial programming frameworks, and is easily extensible.
+This is crucial as, given the enormous code size and complexity, rewriting epidemiology simulators using a dedicated universal probabilistic programming language, such as Pyro \cite{bingham2019pyro}, is often infeasible.
+%Moreover, though our focus is on hijacking epidemiology simulators, our general approach and framework can be applied more widely to a general class of stochastic simulators.
+
+In time, we hope our approach will play a critical role in
+bringing recent advancements in probabilistic programming to bear on
+the vast array of existing simulators used throughout the sciences,
+thereby providing wide-ranging impacts across a number of fields.
+
+This paper first gives an overview of existing Malaria simulators (Section \ref{sec:background}), and proceed by introducing the necessary background on the \textit{pyprob} framework and the concept of universal probabilistic programming (Section \ref{sec:hijackingsimulators}).
+Our approach is then demonstrated and analysed in the context of a Malaria case study (Section \ref{sec:casestudy}).
+
+\section{Background}
+\label{sec:backgorund}
 
 \begin{itemize}
+\item Paragraph on Probabilistic programming
+\item Paragraph on Simulators keep high-level focus on pop sims later
+\item Paragraph on Baydin et al
+\end{itemize}
+In-silico simulators have become a crucial tool in evidence-based decision-making within a large number of disciplines, including statistical physics~\cite{landau_binder_2014}, financial modeling~\cite{jackel2002monte},
+weather prediction~\cite{evensen1994sequential}, epidemiology~\cite{smith2008towards} and many others.
+In many cases, simulation output can augment or even replace real data that may otherwise be costly or even impossible to generate.
+Recent advances in hardware have enabled simulations to model increasingly complex systems.
+Epidemiology studies the prevalence and spreading of diseases across populations.
+Recent advances in hardware have enabled simulations to model the dynamics of infectious diseases, such as Malaria, in ever greater detail.
 
-\item Outline the objectives
-\item State the importance of interpretability in the inference of the simulator
-\item Our system enables one to understand and interpret the most common paths in a program (simulation).
-\item In highlighting this a practitioner can then go back to the field and explore that parameter/s in more depth.
-\item This facilitates understanding and the construction of more detailed models and data collection strategies
+Probabilistic programming~\cite{gordon2014probabilistic,staton2016semantics,kozen1979semantics} 
+can be used to express probabilistic models and consequently perform automated inference these.
+Once a probabilistic model has been expressed in a probabilistic programming language, a wide range of inference techniques, such as Markov Chain Monte Carlo (MCMC)\cite{geyer1992practical}, Variational Inference~(VI)\cite{wainwright2008graphical} and Deep Neural Networks~(DNN)\cite{Goodfellow:2016:DL:3086952},  can be used by non-experts in an automated fashion.
 
-\end{itemize}
+% \section{Types of Models and Probabilistic Programming Languages}
 
-% taken from Toms thesis.
+% To bridge the gap between the statistics and machine learning community to the probabilistic programming community we provide a useful overview to expose our ideas more concretely.
 
-Stochastic simulators are common across a number of domains, statistical physics~\cite{landau_binder_2014}, financial modeling~\cite{jackel2002monte},
-weather prediction~\cite{evensen1994sequential}, epidemiology~\cite{smith2008towards} and many others. 
-An advantage of using stochastic simulators is that they provide a level of interpretability not found in modern deep learning settings, as they directly incorporate model structure from carefully reasoned observations and experiments.
-Probabilistic programming systems~(PPSs) are purpose built systems for making inferences and probabilistic 
-modeling accessible. 
-In particular, PPSs allow probabilistic models to be represented in the form of a generative model through the
-probabilistic programming language~(PPL), which enables one to write statements that enable conditioning on data~\cite{gordon2014probabilistic,Goodman08church:a}. 
-Thus, it seems only natural to connect simulators with PPSs, as simulators explicitly define a generative model, in the language in which they are written, and the nature of the PPSs enables us to perform inference in these simulators, by conditioning on observations, which are fed in as input to the simulators.
-In doing this we can infer things about stochastic input parameters and other variables sampled during the program's forward execution, whilst also providing full interpretability in the inference results, which is absolutely necessary in safety-critical domains.
-By exploiting the tools that we develop in this paper, one is able to connect epidemiology simulators with \texttt{Pyprob}~\cite{le-2016-inference} to extract fully interpretable posterior structures allowing practitioners to interpret which input parameters have the largest affect on the epidemic, in our case malaria.
-Over five-hundred-thousand people die from malaria each year, mostly children under five years of age, with 90 per cent of malaria cases occurring in Sub-Saharan Africa.
-An estimated 100-300 million people suffer from malaria each year~\cite{world2016world}.
-Thus, by understanding the prominence of input parameters to epidemiology simulators, guides decision making processes relating to effective treatment and prevention strategies. 
-\label{sec:related}
+% Probabilistic programming languages, broadly speaking\footnote{For a more comprehensive overview see \cite{rainforth2017automating}}, can be classfied into three categories, each type of langauge extends a particular model space, enabling an increasingly complex set models to be represented as programs, but in doing so the notion of posterior and what it means to do inference over the model, at least in a stastically traditional sense, can change.
+% % \begin{figure}
+% % \centering
+% % \includegraphics[]{../models/types_of_pp.png}
+% % \end{figure}
+% The first type of langauges are called first-order languages where the priors are well defined and the \textit{joint} program density constructed via the program directly represent the unormalised stastical joint density $p(x,y)$
+% \section{Methodology}
+%  TODO
 
-\section{Epidemiology Simulators and Probabilistic Programming}
-\label{sec:background}
-This will kind of be like a background section.
+\section{Hijacking Simulators}
+\label{sec:methods}
 
-\section{Methodology}
- TODO
+\subsection{The Engineering}
+Hijacking a simulator a describes the process by which a simulator's random number generators are replaced by calls to external sampling procedures, which are controlled by a probabilistic programming system (PPS).
+In practice, this amounts to performing a small number of surgical incisions into the simulator's source code in order to replace built-in calls to random number generators.
+E.g., given a simulator written in C++/Boost\cite{schaling2011boost}, a Gaussian distribution object \textit{boost::normal} is replaced with the corresponding \textit{pyprob\_cpp} distribution object, namely \textit{pyprob\_cpp::distributions::Normal}.
+Sampling from this distribution is then done by requesting the PPS to send a sample back to the simulator.
+\bg{This will be extended}
 
-\section{Results}
+ \label{sec:methods}
 
-TO ADD
+  \begin{algorithm}
+   \caption{How to perform posterior inference over a stochastic simulator with 
+   }
+    \begin{algorithmic}[1]
+      \Function{Convert Simulator to a Probabilistic Program}{$A$}\Comment{}
+
+        \State replace raw samples statements $f_{t}(x_{t} | \eta_{t}) $
+        \State Create thin wrapper
+        \State Create Containerized Environment $C_{i}$
+        \State Create Connection
+        \State Run Program
+       \EndFunction
+       \Function{Create thin wrapper}{$f_{t}(x_{t} | \eta_{t})$}
+       \State to add
+       \EndFunction
+       \Function{Create Containerized Environment }{$\Delta_{i}$}\Comment{Where $\Delta_{i}$ is one of $i$-th dependencies}
+       \State to add
+       \EndFunction
+
+\end{algorithmic}
+\end{algorithm}
+
+% Stochastic simulators are common across a number of domains, statistical physics~\cite{landau_binder_2014}, financial modeling~\cite{jackel2002monte},
+% weather prediction~\cite{evensen1994sequential}, epidemiology~\cite{smith2008towards} and many others. 
+% An advantage of using stochastic simulators is that they provide a level of interpretability not found in modern deep learning settings, as they directly incorporate model structure from carefully reasoned observations and experiments.
+In addition to this,
+
+The PPS and the simulator exchange xTensor objects \url{https://xtensor.readthedocs.io/en/latest/} through TCP or ICP using a generic FlatBuffers \url{http://google.github.io/flatbuffers/} protocol.
+On the PPS side, sampling is done using the deep learning framework PyTorch \cite{paszke2017automatic}.
+
+After a variable has been sampled, it is sent to the PPS using \textit{pyprob\_cpp::sample}, alongside with its simulator source trace.
+This allows the PPS to construct sample trace probabilities and other summary statistics (cf.
+Section \ref{sec:casestudy}).
+
+Finally, the entry point to the simulator, i.e. \textit{main} in C++, is replaced by a special \textit{forward} call, in this case \textit{pyprob\_cpp::forward}.
+This allows the PPS to generate rollouts from the simulator remotely.
+
+\paragraph{Population-based simulators} In contrast to the setup used by \cite{baydin2018efficient}, who interface a particle physics simulator that creates a single trace per \textit{forward} call, a population-based simulator creates a trace per population member.
+This means that e.g.
+in EMOD, a standard scenario simulation rollout over the span of $3$ years
+with a population of size $n=5000$ \cite{smith2008towards} will generate about \textit{four terabytes} of raw trace data.
+Unlike with \citep{baydin2018efficient}, this amount of data cannot possibly be kept in RAM, which makes \textit{a posteriori} trace analysis very inefficient.
+In order to deal with the shortcomings of \textit{pyprob} in a population-based simulator context, we therefore extend the framework to be able to do trace analysis on the fly.
+
+By extending \textit{pyprob} to handle population-based simulators, the PPS can now track all stochastic random variables that are created within the simulator, which then allows us to generate trace plots and path probabilities associated to the execution paths of the program.
+The corresponding increase of simulator transparency helps reinstate much needed trust between policymakers and evidence-based methods.
+
+\subsection{The Theory}
+
+\bg{Tom add theory stuff here}
+%As stated previously, Simulators are complex systems and should be viewed as black-boxes, thus deciphering the underlying processes occurring within the simulator facilitates the users understanding of how such a system operates and provides a way of interpreting the outputs in a clear way.
+In addition to this, when using a simulator to make new inferences a user will often want to condition on the output of the simulator, or any new observations gathered.
+In the standard simulator framework this is not possible, but in
+%
+%Essentially, from the users perspective, our approach can be viewed as follows the user would first select the simulator that they are going to use, for example EMOD, they would then write their model, in the desired simulator, according to the given simulator specification.
+The user would then run the simulator.
+Now comes the hijacking.
+During the forward run of the simulator all sampling calls from the random number generator are directed through the probabilistic programming system~(PPS), which in addition to the standard output of the simulator now generates a series of result summaries and path probabilities, as we demonstrate in Section~\ref{sec:casestudy}.
+%
+%By connecting with the PPS we track all sampling procedures, which enables us to control the myriad of random variables generated within the simulator. 
+%
+%In order to overhaul a stochastic simulator like EMOD, or OpenMalaria, we are only required to override the location of the stochastic primitives, i.e.
+the base operations that generate the stochasticity and randomness within the simulator.
+%We will walk through how this is done for a standard epidemiology simulator, see the flowchart for a pictorial representation of what is happening \bg{add ref to flow chart}.
+%
+%The initial step first requires one to build a containerized environment, such as Docker~\cite{merkel2014docker} and Singularity~\cite{kurtzer2017singularity}, which enables the simulators to be run on any device, independent of different hardware architectures, removing the requirement for hardware specific machines.
+%The second step requires one to override the stochastic primitives, which requires re-writing aspects of the sampling procedures inside the simulator. 
+%This requires zero domain-expertise and only requires one to find where the sampling calls are originated, this can be done with any basic editor search function.
+% Such information is typically saved in $\texttt{RANDOM.<filetype>}$, although it may vary slightly between simulators. We now demonstrate this for a uniform primitive.
+%
+%The third step is generating the protocols, which enable communication between the simulator and probabilistic programming system. 
+%For this we extend the existing bridging framework introduced in~\cite{baydin2018efficient}, which is comprised of the Probabilistic Programming eXecution~(PPX) protocols and a light-weight C ++ interface, however the language of this light-weight interface is dependent on whatever the file type of the $\texttt{RANDOM.<filetype>}$ file is.
+%For both OpenMalaria and EMOD, the file is written in C ++.
+%In using these protocols a larger class of sampling procedures can be overridden.
+The PPX protocols work by making use of the flatbuffer protocols\footnote{\url{https://github.com/google/flatbuffers}}, which require a flatbuffers script to be specified, see here for an example of a flatbuffers script\footnote{https://github.com/probprog/ppx/blob/master/ppx.fbs}.
+Once this is specified the flat buffers script is compiled and automatically constructs the message passing framework between the simulator and the probabilistic programming system.
+The flatbuffers script is simple, clean and intuitive, making it easy for a non-expert to produced.
+%The next steps requires modifying the $\texttt{main}$ function in $\texttt{main.cpp}$ and defining an additional forward function which runs one execution of the simulator to enable message passing between the simulator and PPS\footnote{A link to an example will be provided after the review period to preserve anonymity}.
+%The final step is to create a light-weight interface.
+As both the EMOD and OpenMalaria are solely based on C ++ we can extend an existing interface called pyprob\_cpp\footnote{\url{https://github.com/probprog/pyprob_cpp}} to account for the additional sampling procedures.
+%Once these steps are complete, the containerized environment will now contain everything required to pass models through to the simulators, encode the simulator into a probabilistic program, which can then be evaluated in the PPS.
+
+ %Our method of overriding simulators builds on the work of~\cite{baydin2018efficient} and extends the framework to encapsulate a more diverse range of simulators, in particular population based simulators such as those found in epidemiology.
+%Our framework provides a simple solution to do this and makes it easy to transform arbitrary stochastic simulators into probabilistic programs, regardless of the complexity of the simulator and the language that the simulator is written in.
+%In particular, the framework currently supports 13 popular languages, but there is nothing to stop this being extended to any language of interest. 
+%Furthermore, we can understand the structure of simulators as programs for both decoding \emph{black-box structures} and for posterior inference, which enables two things. 
+%1) It enables software developers and policy makers to understand complicated code bases and dissect the outputs of simulators for guiding policy decision making.
+ %2) It provides interpretable inference results that provide policy makers with predictions that are truly interpretable, in the sense that the end-user of the inference results understands what physical events led to the inference outcome.
+ %This is critical in several domains, especially in epidemiology. 
+ %We provide examples of (1) in next Section~\ref{sec:casestudy} and discuss the latter in future work as extensions to the the current framework.
+
+\section{Experiments}
+\subsection{Epidemiology Simulators}
+
+Malaria epidemiology is governed by a complex set of drivers, 
+few of which can be understood in isolation \cite{cameron2015defining,autino_epidemiology_2012,smith2008towards,bershteyn2018implementation}.
+These include within-host dynamics, population-specific traits and even local geography.
+Comprehensive modeling of all of these components remains challenging, particularly in a region-specific context. 
+Computational epidemiology simulators have to reflect these complexities and are usually stochastic in nature.
+This can make simulation output highly non-trivial to interpret, particularly when trying to draw desired inferences coupled with observed data \cite{mwendera_challenges_2019,ferris_openmalaria_2015}.
+
+We first demonstrate our system on a reduced version of the
+
+Two the most advanced Malaria simulators, namely EMOD~\cite{bershteyn2018implementation} and OpenMalaria  \cite{smith2008towards}, have proven to be particularly valuable to policymakers.
+OpenMalaria is based on microsimulations of Plasmodium falciparum in humans and was originally developed to simulate the impacts of malaria vaccines within simple villages or districts
+Compared with OpenMalaria, EMOD is able to simulate a variety of additional drivers, including complex geographies complete with migration and a large number of policy interventions
+Both EMOD and OpenMalaria are open source and implemented in C++.
+
+\bg{Add linking stuff now to the toy-example where hopefully we can perform inference and }
+
+% Simulators provide users with the ability to model complex environments and problems in a black-box way, enabling new insights to be generated and predictions to be made.
+A simulator typically encodes several years, if not decades, of research and development and so by their very nature are structurally complicated.
+
+% Within the context of epidemiology simulators we can model the treatment, infection and prevention cycle under a variety of different environmental conditions, enabling predictions to be made on a set of field observations.
+Such scenarios in practices would be
+
+% This enables practitioners to model a wide array of different scenarios at ease.
+% However, since the development happens over several years, or decades, usually without comprehensive documentation, understanding a code base written with legacy libraries and software is incredibly difficult and an almost impossible task for a non-expert, or expert who has not had access to the previous years of development.
+%
+
+% Two prominent simulators for modeling Malaria are OpenMalaria~\cite{smith2008towards} and Emod~\cite{bershteyn2018implementation}, which are used by practitioners to model several aspects of the Malaria disease.
+
+%  \paragraph*{OpenMalaria} 
+% is an open-source simulator funded by the Gates Foundation to model different aspects of the malaria prevention and treatment cycle for varying population dynamics.
+%  Users specify models in OpenMalaria via a scenario file, in the form of an XML document, which determines the specifics of a particular treatment, or prevention strategy, climatic factors, the demographics of the population and so forth.
+In addition to this, field observations can be encoded inside of, or externally to the scenario file.
+% The scenario file encodes all input and output parameters, some of which we wish to simulate, others remain fixed.
+
+% \paragraph{EMOD} provides the modularity and structural flexibility necessary for modeling multiple infectious diseases at the within-host and population level. 
+% EMOD can be used to model HIV, Tuberculosis, Malaria and generic diseases, but for the purposes of the comparison, we just focus on Malaria. 
+% To specify models in EMOD a configuration file is written by the user, which is in the form of a JSON and provides similar functionality to OpenMalaria, in terms of what can be specified and the outputs that it produces.
+
+% In both these well used simulators users are accustomed to writing their models as either XML, or JSON files and so it is important that any system built around this framework enables the models to be inputted in the same format, without requiring the users to learn how to use a new set of protocols.
+Our system facilitates this as we shall discuss in the proceeding sections.
+
+\begin{figure}
+  \includegraphics[width=\textwidth]{../plots/system.pdf}
+  \label{fig:how}
+  \caption{This flow-chart provides an overview of the process of how our system hijacks a generic population-based epidemiology simulator, such as EMOD, and how we modify the simulator to 
+  to hijack the random number generator (RNG).}
+
+  \end{figure}
+
+% \begin{table}[h!]
+% \footnotesize
+% \begin{tabular}{p{3.6cm}|p{3.6cm}}
+% Before  &  After  \\
+% \midrule
+% \begin{lstlisting}
+% #include <gsl/gsl_cdf.h>
+% #include <gsl/gsl_rng.h>
+% #include <gsl/gsl_randist.h>
+% //preamble
+% double random::uniform_01 () {
+%     double result =
+% # ifdef 
+% OM_RANDOM_USE_BOOST
+%        rng_uniform01 ();
+% # else
+%  gsl_rng_uniform (rng.gsl_generator);
+% # endif
+%     return result;
+%     }
+% \end{lstlisting}&
+% \begin{lstlisting}
+% #include <pyprob_cpp.h>
+% #include "xtensor/xadapt.hpp"
+% // preamble
+% double random::uniform_01 () {
+% auto uniform = pyprob_cpp::distributions::Uniform(0,1);
+% return pyprob_cpp::sample(uniform)(0);
+% }
+% \end{lstlisting}
+% \end{tabular} 
+% \end{table}
+% As can be seen, writing the new sampling call is cleaner and syntactically simpler. 
+%\bg{Add performance plot showing memory consumptions of pyprob and our new system.}
+\subsection{Real-world Example}
+\label{sec:casestudy}
+
+Ensemble methods are commonly used in statistics in order to combine the predictive power of multiple models \cite{cameron2015defining,smith_ensemble_2012}.
+To this end, recent work has attempted to characterise the similarities and differences between two of the most advanced Malaria epidemiology simulators, EMOD \cite{bershteyn2018implementation} and OpenMalaria \cite{smith2008towards}.
+Evaluation is usually done by comparing a number of output parameters across a range of hand-crafted standard scenarios reflecting different geographical locations across Africa~\cite{smith_ensemble_2012}.
+
+In the following, we illustrate how our method introduces a novel introspection paradigm.
+By extracting trace graphs from population-based simulators, policymakers can ask specific questions about properties of the model trace flow.
+
+To illustrate the above, we present simulation output generated from a scenario resembling local conditions in the town of Ifakara (Tanzania).
+Both EMOD and OpenMalaria are configured to simulate a single population node of $n=100$ and assume constant climatic conditions and no migration over the simulation period of $3$ years.
+Please refer to Figure \ref{fig:EIR} for the seasonal Entomological Inoculation Rate (EIR), a measure of infectivity.
+
+\begin{figure}[h!]
+  \centering
+  \includegraphics[width=\textwidth/2]{../plots/EIR_Ifakara.pdf}
+  \caption{Seasonal Entomological Inoculation Rate (EIR) for the Ifakara scenario. Data is averaged over $30$ day periods.}
+  \label{fig:EIR}
+\end{figure}
+
+We provide examples of the generated trace plots from the connection between the simulator and the PPS in Figure~\ref{fig:plotewan}.
+
+We can see from the addressing schemes $A_1, \ldots, A_N$ (Tables \ref{table:addresses} and \ref{table:asingleaddress}) what physical events are connected to each other and how outputs in EMOD are generated from a different set of procedures as compared to OpenMalaria
+By having access to such diagrams users can internally evaluate and scrutinize the decisions that the simulator is making.
+
+This is important for policymakers, or general non-experts, as it not only details how we arrive at the given outputs, but it provides an understanding of which processes were most crucial in determining those outputs as can be seen from the path probabilities assigned to each of the vertices.
+
+Additionally, by associating nodes in trace graphs representing the same physical processes within different models, the significance of model detail can be evaluated.
+For example, the full trace presented in Table \ref{table:asingleaddress} represent a sampling step associated with within-host dynamics of the Malaria parasite \textit{Falciparum} in OpenMalaria node $A1$.
+The same physical process also occurs in EMOD's trace graph at position $A7$ (see Section \ref{table:addresses}).
+Comparisons like these could help developers better control model complexity, and even provide an alternative testing and debugging paradigm.
+
+% \begin{tikzpicture}
+% \begin{axis}[xlabel={Population size},ylabel={memory consumption (MB)},
+% legend style={at={(0.5,-0.15)},xmin=0,xmax=700,ymin=0,ymax=70, axis lines=center]
+
+% \addplot[pyprob,domain=0:1000, color=blue,]{x};
+
+% \addplot[improved,domain=0:1000, color=red,]{10};
+
+% \legend{pyprob, improved}
+% \end{axis}
+% \end{tikzpicture}
+
+%\begin{figure}
+%\begin{tikzpicture}
+%\begin{axis}[
+%    ybar,
+%    enlargelimits=0.15,
+%    legend style={at={(0.5,-0.15)},
+%      anchor=north,legend columns=-1},
+%      xlabel = {Address},
+%    ylabel={Probabilities},
+%    symbolic x coords={Trace 1, Trace 2, Trace 3},
+%    xtick=data,
+%    nodes near coords,
+%    nodes near coords align={vertical},
+%    ]
+%% \addplot coordinates {(tool8,7) (tool9,9) (tool10,4)};
+%\addplot coordinates {(Trace 1,0.2) (Trace 2,0.5) (Trace 3,0.23)};
+%% \addplot coordinates {(A1,0.2) (A2,0.5) (A3,0.7) (A4,0.9)};
+%% \addplot coordinates {(A1,0.1) (A2,0.1) (A3,0.1) (A4,0.3)};
+%% \legend{used,understood,not understood}
+%\end{axis}
+%\end{tikzpicture}
+%\caption{Trace probabilities. Trace 1 corresponds to A1,A1,A4,... 
+%Trace 2 corresponds to A1,A1,A1,...A2,... These still need correctly calculating.}
+%\end{figure}
+
+\begin{figure*}[h!]
+  \centering
+  \includegraphics[width=\textwidth]{../plots/final_emod_traces__rotated.pdf}
+  \includegraphics[width=\textwidth]{../plots/ewan_25_pop_100.pdf}
+  \caption{Here we run two equivalent models, compare the corresponding trace paths and corresponding 
+  path probabilities taken by the thousands of random variables generated internally within the simulators.
+\textit{\textbf{Top:}} The specified model run in EMOD.
+\textit{\textbf{Bottom:}} The specified model in OpenMalaria.}
+  \label{fig:plotewan}
+\end{figure*}
+
+\begin{table}[h!]
+\footnotesize
+  \setlength{\tabcolsep}{1mm}
+\label{table:asingleaddress}
+  \caption{An example of an address generated for the model run in the OpenMalaria simulator. We can see that A1
+  relates to Generating a member of the human population who may or may not be infect with the Malaria disease.
+We get something similar for EMOD, except this relates to A7 in the EMOD program execution.}
+  \def\arraystretch{1.25}
+  % \setlength{\tabcolsep}{1mm}
+  \begin{tabularx}{\textwidth}{@{}lX@{}l@{}} 
+    \toprule
+    Address ID & Full address \\
+    \midrule
+  A1 & [forward()+0x204;OM::Simulator::start(scnXml::Monitoringconst)+0x28a;
+  OM::Population::createInitialHumans()+0x94;
+
+  OM::Population::newHuman(OM::SimTime)+0x5c;
+  OM::Host::Human::Human(OM::SimTime)+0x12b;
+
+  OM::WithinHost::WHInterface::reateWithinHostModel(double)+0x99;
+  OM::WithinHost::DescriptiveWithinHostModel::DescriptiveWithinHostModel(double)+0x3a;
+  OM::WithinHost::WHFalciparum::WHFalciparum(double)+0xe6;
+  OM::util::random::gauss(double, double)+0xb4]\_\_Normal \\
+
+  $\vdots$ & $\vdots$ \\
+
+  A5 & [forward()+0x204;OM::Simulator::start(scnXml::Monitoringconst\&)+0x468;
+  OM::Population::update1(OM::SimTime)+0xff;
+
+  OM::Host::Human::update(bool)+0x2bc;
+
+  OM::Clinical::ClinicalModel::update(OM::Host::Human\&,double, bool)+0x96;
+
+  OM::Host::NeonatalMortality::eventNeonatalMortality()+0x9;
+
+  OM::util::random::uniform\_01()+0xc0]\_\_Uniform\\
+  $\vdots$ & $\vdots$ \\
+
+ % $\vdots$ & $\vdots$ \\
+
+\bottomrule
+  \end{tabularx}
+  \end{table}
+
+\begin{table}[h!]
+  \footnotesize
+  \setlength{\tabcolsep}{1mm}
+  \caption{An interpretation table for each of the address of the 
+  overall trace generated from the corresponding OpenMalaria model.}
+  \label{table:addresses}
+  % \def\arraystretch{1.25}
+  % \setlength{\tabcolsep}{1mm}
+  \begin{tabularx}{\textwidth}{@{}lX@{}} 
+    \toprule
+    Address ID & Interpretation \\
+    \midrule
+  A1 & Generate a human in the population within host dynamics\\
+
+  A2 & Generate another human in the population within host dynamics\\
+
+  A3 & The population is updated and a new human, or humans, may get infected \\
+
+  A4, A5 & Potential child deaths within the population are simulated \\
+
+  A6 & Determines parasite density of an individual infection \\
+
+  A7 & Models how the disease is progressing within the infected humans \\
+
+  A8 & Models how the disease is progressing within the population \\
+
+  A9 & Models how the disease is progressing within the infected humans after the population has been updated\\
+
+  A10 & Full clinical update on the population for those without severe or no Malaria infection. \\
+
+  A11 & Full clinical update on the population for those with severe Malaria infections \\
+\bottomrule
+  \end{tabularx}
+  \end{table}
+
+\section{Discussions and Future Work}
+
+In this work we have demonstrated a method 
+that enables one to hijack populationbased simulators, extending 
+the work of \citep{baydin2018efficient}. 
+We applied our method to two Malaria orientated population-based simulators and 
+generated a variety of trace graphs.
+Finally, we have shown how our system
+enables policy makers and non-experts to analyse simulator outputs in a 
+way previously unavailable in the field of epidemiology.
+
+To extend our work further we aim to implement additional 
+tools that will facilitate complicated inference procedures that condition on simulator 
+output. We will also evaluate additional scenarios across Africa and South-East Asia to better understand the similarities and differences between EMOD and OpenMalaria.
 
@@
 
-\section*{References}
+% \section*{References}
 
 \bibliographystyle{plain}
-\bibliography{refs}
+\bibliography{refs,malaria}
 

### 2019-05-19 paper/neurlps2019/main.tex
updated

@@
 We enable simulators written in 13 different languages to be easily converted to probabilistic programs, for which we algorithmically outline how to do and provide examples of.
-We demonstrate our method on three models and provide n infected population simulator and demonstrate the hijacking on two important malaria population-based simulators OpenMalaria and EMOD.
+We demonstrate our method on three real-world simulators for epidemiology.
+
+\todo[inline]{We need to have probabilistic programming in the title to make sure we attract the correct reviewers}
 

### 2019-05-20 paper/neurlps2019/main.tex
updated

@@
 \usepackage{algpseudocode}
+\usepackage{wrapfig}
 % \usepackage[binary-units=true]{siunitx}
@@
 Probabilistic programming provides a way to perform statistical inference over simulations of events in a programmatic way.
-Thus, simulators by design are ideal programs for probabilistic programming systems~(PPSs).
+Thus, by design, simulators are ideal programs for probabilistic programming systems~(PPSs).
+Simulators arise in a number of industrial scientific and industrial domains and encode sophisticated generative models.
 However, within existing PPSs one would have to re-implement the simulator via the PPS language specification, which is inefficient and often not feasible due to the complexity of such scientific and industrial simulators.
-However, recent work
-by \cite{baydin2018efficient} demonstrated a pathway to turn a particular
-type of event-based simulator into a probabilistic program. 
+Recent work by \cite{baydin2018efficient} demonstrated a pathway to turn a particular
+type of event-based simulator into a probabilistic program, without having to re-implement the simulator in the PPS. 
 % via the raw random number calls of the simulator.
- But, this still meant that a large class of critically important population-based simulators could not be turned into probabilistic programs and as such could not be used in a probabilistic programming framework.
-In this work, we extend that framework to encompass population-based simulators, a very large class of simulators that are used extensively across epidemiology modeling, multi-agent modeling and financial modeling.
-We enable simulators written in 13 different languages to be easily converted to probabilistic programs, for which we algorithmically outline how to do and provide examples of.
-We demonstrate our method on three real-world simulators for epidemiology.
+ However, this still meant that a large class of critically important population-based simulators could not be turned into probabilistic programs and as such could not be used in a probabilistic programming framework.
+In this work, we extend that framework to encompass population-based simulators, a very large class of simulators that are used extensively across epidemiology , multi-agent and financial modeling. 
+We enable simulators written in 13 different languages to be easily converted to probabilistic programs, for which we algorithmically outline and provide comprehensive examples of our system applied to a number of real-world population-based simulators.
 
@@
 % \end{itemize}
+\begin{wrapfigure}{R}{0.45\textwidth}
+\centering
+% \includegraphics[width=0.35\textwidth]{../plots/mpenny-pop10/mpenny_10.pdf}
+\includegraphics[width=0.40\textwidth]{../plots/model_lambda/model_posterior.pdf}
+\includegraphics[width=0.40\textwidth]{../plots/model_lambda/hist_posteriors.png}
+\caption{Hijacking the OpenMalaria simulator}
+\label{fig:hijacking} 
+\end{wrapfigure}
 
@@
 The source code of the given model, or equivalently the \emph{program}, is then transformed into a program density, which can
-be viewed as the joint density of the model.
-Having direct access to the source code enables probabilistic programming system~(PPS)
+be viewed as the joint density of the user-specified model.
+Having direct access to the source code enables probabilistic programming systems~(PPSs)
 to conduct several processes in an automated fashion, such as 
-automating inference \cite{zhou2019lf,goodman2012church,wood2014new}.
-In addition to this, as models can be directly expressed in program code, we can encode other structures to make use of the full structure of a given language,
-which enables users to integrate PPSs into their standard model development process.
+the automation of complex inference procedures\cite{zhou2019lf,goodman2012church,wood2014new}.
+In addition to this, as models can be directly expressed in program code we can encode other structures to make use of the full structure of a given language,
+which enables users to integrate PPSs into their standard model development process, increasing efficiency and reducing model iteration time.
+
 One powerful \emph{class} of models are simulators. 
-
 Simulators arise in many industrial \cite{di2015simulator,judd2009numerically,kass1990rapid} and scientific applications~\cite{smith2008towards,bershteyn2018implementation,gleisberg_event_2008} encode years, if not decades, of research and development.
@@
 
+\bg{Chris will write up}
 Ensemble methods are commonly used in statistics in order to combine the predictive power of multiple models \cite{cameron2015defining,smith_ensemble_2012}.
@@
 
-% \begin{tikzpicture}
-% \begin{axis}[xlabel={Population size},ylabel={memory consumption (MB)},
-% legend style={at={(0.5,-0.15)},xmin=0,xmax=700,ymin=0,ymax=70, axis lines=center]
-
-% \addplot[pyprob,domain=0:1000, color=blue,]{x};
-
-% \addplot[improved,domain=0:1000, color=red,]{10};
-
-% \legend{pyprob, improved}
-% \end{axis}
-% \end{tikzpicture}
-
-%\begin{figure}
-%\begin{tikzpicture}
-%\begin{axis}[
-%    ybar,
-%    enlargelimits=0.15,
-%    legend style={at={(0.5,-0.15)},
-%      anchor=north,legend columns=-1},
-%      xlabel = {Address},
-%    ylabel={Probabilities},
-%    symbolic x coords={Trace 1, Trace 2, Trace 3},
-%    xtick=data,
-%    nodes near coords,
-%    nodes near coords align={vertical},
-%    ]
-%% \addplot coordinates {(tool8,7) (tool9,9) (tool10,4)};
-%\addplot coordinates {(Trace 1,0.2) (Trace 2,0.5) (Trace 3,0.23)};
-%% \addplot coordinates {(A1,0.2) (A2,0.5) (A3,0.7) (A4,0.9)};
-%% \addplot coordinates {(A1,0.1) (A2,0.1) (A3,0.1) (A4,0.3)};
-%% \legend{used,understood,not understood}
-%\end{axis}
-%\end{tikzpicture}
-%\caption{Trace probabilities. Trace 1 corresponds to A1,A1,A4,... 
-%Trace 2 corresponds to A1,A1,A1,...A2,... These still need correctly calculating.}
-%\end{figure}
+\subsection{Larger Scale-Simulators: OpenMalaria and EMOD}
 

### 2019-05-20 paper/neurlps2019/main.tex
updated

@@
 
-Probabilistic programming\cite{gordon2014probabilistic,le2016inference} is a machine learning paradigm that enables users to write, and perform inference in, complex probabilistic models in a programmatic, yet efficient and simplified way.
-The source code of the given model, or equivalently the \emph{program}, is then transformed into a program density, which can
-be viewed as the joint density of the user-specified model.
-Having direct access to the source code enables probabilistic programming systems~(PPSs)
-to conduct several processes in an automated fashion, such as 
-the automation of complex inference procedures\cite{zhou2019lf,goodman2012church,wood2014new}.
-In addition to this, as models can be directly expressed in program code we can encode other structures to make use of the full structure of a given language,
-which enables users to integrate PPSs into their standard model development process, increasing efficiency and reducing model iteration time.
+% One powerful \emph{class} of models are simulators. 
+Simulators are useful tools when conducting experiments in the real system 
+is impossible, or impractical and they arise in numerous industrial \cite{di2015simulator,judd2009numerically,kass1990rapid} and scientific applications~\cite{smith2008towards,bershteyn2018implementation,gleisberg_event_2008} .
+They are rich models and encode years, if not decades, of research and development.
+The \emph{model}, or equivalently \emph{program}, specified by the simulator defines a joint density $p(x,y)$ and the inputs into the simulator denote priors, $p(x)$.
 
-One powerful \emph{class} of models are simulators. 
-Simulators arise in many industrial \cite{di2015simulator,judd2009numerically,kass1990rapid} and scientific applications~\cite{smith2008towards,bershteyn2018implementation,gleisberg_event_2008} encode years, if not decades, of research and development.
+ def that are difficult to generate represent a large class of generative models, where the simulator itself represents a joint density
+$p(x|y)$  arise in many industrial
 Thus, Simulators by their very nature are
 information rich and naturally describe complicated joint densities.
-However, in order to write a simulator in a standard probabilistic programming set-up one would have to re-write a simulator in the language of the PPS, which 
+However, in order to write a simulator in a standard probabilistic programming set-up one would have to re-write a simulator in the language of the probabilistic programming system (PPS), which 
 in many cases is not feasible, nor efficient due to both the size 
@@
 the simulator in the given PPS?
+
+\paragraph{Probabilistic programming}\cite{gordon2014probabilistic,le2016inference} is a machine learning paradigm that enables users to write, and perform inference in, complex probabilistic models in a programmatic, yet efficient and simplified way.
+The source code of the given model, or equivalently the \emph{program}, is then transformed into a program density, which can
+be viewed as the joint density of the user-specified model.
+Having direct access to the source code enables PPSs
+to conduct several processes in an automated fashion, such as 
+the automation of complex inference procedures\cite{zhou2019lf,goodman2012church,wood2014new}.
+In addition to this, as models can be directly expressed in program code we can encode other structures to make use of the full structure of a given language,
+which enables users to integrate PPSs into their standard model development process, increasing efficiency and reducing model iteration time.
 

### 2019-05-20 paper/neurlps2019/main.tex
updated

@@
 % \end{itemize}
-\begin{wrapfigure}{R}{0.45\textwidth}
+\begin{wrapfigure}{R}{0.44\textwidth}
 \centering
@@
 They are rich models and encode years, if not decades, of research and development.
-The \emph{model}, or equivalently \emph{program}, specified by the simulator defines a joint density $p(x,y)$ and the inputs into the simulator denote priors, $p(x)$.
-
- def that are difficult to generate represent a large class of generative models, where the simulator itself represents a joint density
-$p(x|y)$  arise in many industrial
 Thus, Simulators by their very nature are
 information rich and naturally describe complicated joint densities.
+That is the \emph{model}, or equivalently \emph{program}, specified by the simulator defines a joint density $p(x,y)$ and the inputs into the simulator denote priors, $p(x)$.
+
 However, in order to write a simulator in a standard probabilistic programming set-up one would have to re-write a simulator in the language of the probabilistic programming system (PPS), which 

### 2019-05-21 paper/neurlps2019/main.tex
updated

--- 2019-05-20 paper/neurlps2019/main.tex
@@
 Recent work by \cite{baydin2018efficient} demonstrated a pathway to turn a particular
-type of event-based simulator into a probabilistic program, without having to re-implement the simulator in the PPS. 
+type of event-based simulator into a probabilistic program, without having to re-implement the simulator in the existing PPS. 
 % via the raw random number calls of the simulator.
- However, this still meant that a large class of critically important population-based simulators could not be turned into probabilistic programs and as such could not be used in a probabilistic programming framework.
-In this work, we extend that framework to encompass population-based simulators, a very large class of simulators that are used extensively across epidemiology , multi-agent and financial modeling. 
+ However, this still meant that a large class of critically important population-based simulators could not be turned into probabilistic programs and as such could not be used within a probabilistic programming framework.
+In this work, we extend that framework to encompass population-based simulators, a very large class of simulators that are used extensively across epidemiology, multi-agent and financial modeling and provide the relevant
+theoretical underpinnings. 
 We enable simulators written in 13 different languages to be easily converted to probabilistic programs, for which we algorithmically outline and provide comprehensive examples of our system applied to a number of real-world population-based simulators.
 
-\todo[inline]{We need to have probabilistic programming in the title to make sure we attract the correct reviewers}
+% \todo[inline]{We need to have probabilistic programming in the title to make sure we attract the correct reviewers}
 
@@
 \label{sec:intro}
-\bg{The audience is a probabilistic programming audience - so that is what we will target}
+% \bg{The audience is a probabilistic programming audience - so that is what we will target}
 
@@
 % \end{itemize}
-\begin{wrapfigure}{R}{0.45\textwidth}
+
+% One powerful \emph{class} of models are simulators. 
+Simulators are useful tools when conducting experiments in the real system 
+is impossible, or impractical and they arise in numerous industrial \cite{di2015simulator,judd2009numerically,kass1990rapid} and scientific applications~\cite{smith2008towards,bershteyn2018implementation,gleisberg_event_2008}.
+They are rich models and provide a level of interpretability not typically found in modern deep learning methodologies due to the several years, if not decades, of research and development that has led to their construction.
+Thus, simulators by their very nature are information rich and naturally describe complicated joint densities.
+That is the \emph{model}, or equivalently \emph{program}, specified by the simulator defines a joint density $p(x,y)$ and the inputs into the simulator denote priors $p(x)$. 
+But, for all the benefits of using simulations there exist two challenging problems that are currently faced:
+\begin{itemize}
+\item It is difficult to calibrate the internal simulator parameters, as the events they are simulating are expensive to reproduce, or are very rare, so data is scarce. 
+\item As the development process takes place over several years the code base becomes convoluted, making it increasingly challenging to understand the structure of the joint density and the internal operations within the simulator, diminishing the interpretability that the simulator once provided.
+\end{itemize}
+
+\begin{wrapfigure}{R}{0.44\textwidth}
 \centering
@@
 \includegraphics[width=0.40\textwidth]{../plots/model_lambda/hist_posteriors.png}
-\caption{Hijacking the OpenMalaria simulator}
+\caption{An example of the posterior output of population-based simulator after hijacking.
+\textit{Top}: Represents the posterior execution traces of the simulator, inside the the probabilistic programming
+system. Each address $A_{i}$ correspond to a physical process within the simulator, we describe these addresses and their structures in more details in Section~\ref{sec:casestudy}.
+\textit{Bottom}: Represents the histograms generated after posterior inference for the parameters of interest, in this case those parameters are represented by addresses $A_{2}$ to $A_{6}$, conditioned on some observations.
+The grey dashed lines represent the ground truth.}
 \label{fig:hijacking} 
 \end{wrapfigure}
-
-Probabilistic programming\cite{gordon2014probabilistic,le2016inference} is a machine learning paradigm that enables users to write, and perform inference in, complex probabilistic models in a programmatic, yet efficient and simplified way.
-The source code of the given model, or equivalently the \emph{program}, is then transformed into a program density, which can
-be viewed as the joint density of the user-specified model.
-Having direct access to the source code enables probabilistic programming systems~(PPSs)
-to conduct several processes in an automated fashion, such as 
-the automation of complex inference procedures\cite{zhou2019lf,goodman2012church,wood2014new}.
-In addition to this, as models can be directly expressed in program code we can encode other structures to make use of the full structure of a given language,
+One such machine learning paradigm that has the ability to resolve these issues is \emph{Probabilistic programming}\cite{gordon2014probabilistic,le2016inference}. 
+Probabilistic programming enables users to write, condition on and perform inference in, complex probabilistic models in a programmatic, yet efficient and simplified way. 
+The source code of the given model, or equivalently the \emph{program}, is transformed into a program density, which can
+be viewed as the joint density of the user-specified model, \emph{the simulator} in this instance.
+Having direct access to the source code enables PPSs to conduct several processes automatically, such as 
+the automation of complex statistical inference procedures\cite{zhou2019lf,goodman2012church,wood2014new}, which enables us to efficiently calibrate
+model parameters, and we can also track how the internals of the program change during
+run time, enabling a full analysis of the program structure and traces generated, which provides greater interpretability, see Figure~\ref{fig:hijacking}.
+In addition to this, as models are directly expressed in program code we can encode other structures to make use of the full structure of a given language, whether that is \texttt{python, C++} or 
+any other language,
 which enables users to integrate PPSs into their standard model development process, increasing efficiency and reducing model iteration time.
 
-One powerful \emph{class} of models are simulators. 
-Simulators arise in many industrial \cite{di2015simulator,judd2009numerically,kass1990rapid} and scientific applications~\cite{smith2008towards,bershteyn2018implementation,gleisberg_event_2008} encode years, if not decades, of research and development.
-Thus, Simulators by their very nature are
-information rich and naturally describe complicated joint densities.
-However, in order to write a simulator in a standard probabilistic programming set-up one would have to re-write a simulator in the language of the PPS, which 
-in many cases is not feasible, nor efficient due to both the size 
-and complexity of the simulator code base. 
-However, what if one could arbitrary connect a simulator, written in any
-language and connect it directly to a PPS, without having to re-write
+For all the brilliance of probabilistic programming, if one did want to run an existing simulator in a standard probabilistic programming set-up one would have to entirely re-write the simulator in the language of the PPS.
+This in many cases is not feasible, nor efficient due to both the size 
+and complexity of the existing simulator code base. 
+However, what if one could connect arbitrary a simulator, written in any
+language and directly connect it to a PPS, without having to re-write
 the simulator in the given PPS?
 
-Our method of hijacking simulators builds on the work of~\cite{baydin2018efficient} and extends the framework to encapsulate a more diverse range of simulators, in particular population-based simulators which arise
-in a myriad of domains \bg{add citations}. 
-Our framework provides a simple solution and makes it easy to transform arbitrary stochastic population-based simulators into probabilistic programs, regardless of the complexity of the simulator. 
-Our system supports simulators written in 13 different programming languages and enables the user to perform for posterior inference over the density of the simulator. 
-This enables two things.
-1) It enables software developers and researchers to understand complicated code bases. 
-2) It provides interpretable inference results that provide policy makers with predictions that are truly interpretable, in the sense that the end-user of the inference results understands what physical events led to inference outcome.
-This is critical in several domains, and is particularly important in the medical domain. We provide examples of both in Section~\textbf{TO ADD REF}.
+\paragraph{Related work} Our method of hijacking simulators builds on the work of~\cite{baydin2018efficient}. 
+They demonstrated how this could be done for event-based simulators, but there methods neglected a large class of population-based simulators that model interactions 
+between populations, that are utilized for modeling epidemics, multi-agent systems and financial economics  \bg{add citations}. 
+Our framework provides a solution and makes it easy to transform arbitrary stochastic population-based simulator, in addition to the event-based simulators, into probabilistic programs, regardless of the complexity of the simulator.
+Our system supports simulators written in 13 different programming languages and enables the user to perform posterior inference over the density of the simulator, conditioned on what data the user has access too, including
+simulator output. We provide a theoretical justification for our approach and many examples.
+% This enables two things.
+% 1) It enables software developers and researchers to understand complicated code bases. 
+% 2) It provides interpretable inference results that provide policy makers with predictions that are truly interpretable, in the sense that the end-user of the inference results understands what physical events led to inference outcome.
+% This is critical in several domains, and is particularly important in the medical domain. We provide examples of both in Section~\textbf{TO ADD REF}.
 
-In this paper, we introduce a novel method that allows one to shed light on the inner workings of a large class of population-based stochastic simulators.
+% In this paper, we introduce a novel method that allows one to shed light on the inner workings of a large class of population-based stochastic simulators.
 We achieve this by extending the work of \citep{baydin2018efficient} by interpreting such population-based simulators as probabilistic generative models within the framework of universal probabilistic programming (UPP) \cite{le-2016-inference}.
@@
 
-This allows for a variety of tasks to be performed on
-the hijacked simulator, such as running inference (by conditioning
-the values of certain draws and manipulating others),
-uncovering stochastic structure, and automatically
-producing result summaries, such as establishing the probability
-of different program paths/traces.  By providing a common abstraction 
-framework for different simulators, our approach further allows for
-easy and direct comparison between related or competing
-simulators, a characteristic that is valuable in the context of
-epidemiology simulators \cite{ferris_openmalaria_2015}. We provide a case study of the above in Section \ref{sec:casestudy}.
-
-Our framework already supports application to simulators written in 13 industrial programming frameworks, and is easily extensible.
-This is crucial as, given the enormous code size and complexity, rewriting epidemiology simulators using a dedicated universal probabilistic programming language, such as Pyro \cite{bingham2019pyro}, is often infeasible.
 %Moreover, though our focus is on hijacking epidemiology simulators, our general approach and framework can be applied more widely to a general class of stochastic simulators.
 
-In time, we hope our approach will play a critical role in
-bringing recent advancements in probabilistic programming to bear on
-the vast array of existing simulators used throughout the sciences,
-thereby providing wide-ranging impacts across a number of fields.
+% In time, we hope our approach will play a critical role in
+% bringing recent advancements in probabilistic programming to bear on
+% the vast array of existing simulators used throughout the sciences,
+% thereby providing wide-ranging impacts across a number of fields.
 
@@
 
-\section{Background}
-\label{sec:backgorund}
+% \section{Background}
+% \label{sec:backgorund}
 
-\begin{itemize}
-\item Paragraph on Probabilistic programming
-\item Paragraph on Simulators keep high-level focus on pop sims later
-\item Paragraph on Baydin et al
-\end{itemize}
-In-silico simulators have become a crucial tool in evidence-based decision-making within a large number of disciplines, including statistical physics~\cite{landau_binder_2014}, financial modeling~\cite{jackel2002monte},
-weather prediction~\cite{evensen1994sequential}, epidemiology~\cite{smith2008towards} and many others.
-In many cases, simulation output can augment or even replace real data that may otherwise be costly or even impossible to generate.
-Recent advances in hardware have enabled simulations to model increasingly complex systems.
-Epidemiology studies the prevalence and spreading of diseases across populations.
+% \begin{itemize}
+% \item Paragraph on Probabilistic programming
+% \item Paragraph on Simulators keep high-level focus on pop sims later
+% \item Paragraph on Baydin et al
+% \end{itemize}
+% In-silico simulators have become a crucial tool in evidence-based decision-making within a large number of disciplines, including statistical physics~\cite{landau_binder_2014}, financial modeling~\cite{jackel2002monte},
+% weather prediction~\cite{evensen1994sequential}, epidemiology~\cite{smith2008towards} and many others.
+% In many cases, simulation output can augment or even replace real data that may otherwise be costly or even impossible to generate.
+% Recent advances in hardware have enabled simulations to model increasingly complex systems.
+% Epidemiology studies the prevalence and spreading of diseases across populations.
 Recent advances in hardware have enabled simulations to model the dynamics of infectious diseases, such as Malaria, in ever greater detail.
 
-Probabilistic programming~\cite{gordon2014probabilistic,staton2016semantics,kozen1979semantics} 
-can be used to express probabilistic models and consequently perform automated inference these.
+% Probabilistic programming~\cite{gordon2014probabilistic,staton2016semantics,kozen1979semantics} 
+% can be used to express probabilistic models and consequently perform automated inference these.
 Once a probabilistic model has been expressed in a probabilistic programming language, a wide range of inference techniques, such as Markov Chain Monte Carlo (MCMC)\cite{geyer1992practical}, Variational Inference~(VI)\cite{wainwright2008graphical} and Deep Neural Networks~(DNN)\cite{Goodfellow:2016:DL:3086952},  can be used by non-experts in an automated fashion.
@@
 \label{sec:methods}
+In-silico simulators have become a crucial tool in evidence-based decision-making within a large number of disciplines, including statistical physics~\cite{landau_binder_2014}, financial modeling~\cite{jackel2002monte},
+weather prediction~\cite{evensen1994sequential}, epidemiology~\cite{smith2008towards} and many others.
+In many cases, simulation output can augment or even replace real data that may otherwise be costly or even impossible to generate.
+Recent advances in hardware have enabled simulations to model increasingly complex systems.
+Epidemiology studies the prevalence and spreading of diseases across populations.
+Recent advances in hardware have enabled simulations to model the dynamics of infectious diseases, such as Malaria, in ever greater detail.
 
@@
 
-We first demonstrate our system on a reduced version of the
+We first demonstrate our system on a reduced version of the OpenMalria simulator, proposed by Smith et .al~\cite{smith2006relationship}.
+The following model directly models the relationship between the entomological inoculation rate and the force of infection for the \emph{plasmodium flaciprum} Malaria.
 

### 2019-05-22 code/openmalaria/model/WithinHost/Diagnostic.cpp
bckp

@@ -123,9 +123,15 @@ void diagnostics::init( const Parameters& parameters, const scnXml::Scenario& sc
     if(scenario.getDiagnostics().present()){
         foreach( const scnXml::Diagnostic& diagnostic, scenario.getDiagnostics().get().getDiagnostic() ){
             string name = diagnostic.getName();     // conversion fails without this extra line
-            bool inserted = diagnostic_set.insert( name, new Diagnostic(parameters, diagnostic) ).second;
-            if( !inserted ){
-                throw util::xml_scenario_error( string("diagnostic with this name already set: ").append(diagnostic.getName()) );
+            //diagnostic[name] = new Diagnostic(parameters, diagnostic);
+            auto inserted = diagnostic_set.insert( name, new Diagnostic(parameters, diagnostic) );
+            if( !inserted.second ){
+                //throw util::xml_scenario_error( string("diagnostic with this name already set: ").append(diagnostic.getName()) );
+                 diagnostic_set.erase(inserted.first);
+                 bool inserted = diagnostic_set.insert( name, new Diagnostic(parameters, diagnostic) ).second;
+                 if( !inserted ){
+                      throw util::xml_scenario_error( string("diagnostic with this name already set: ").append(diagnostic.getName()) );
+                 } 
             }
         }
     }

### 2019-05-22 paper/neurlps2019/main.tex

@@
 \usepackage{wrapfig}
+
+\newcommand{\n}[1]{\ensuremath{n_{#1}}}
+\newcommand{\rv}[1]{\ensuremath{{#1}_{1:n_{#1}}}}
+\newcommand{\rvv}[1]{\ensuremath{{#1}_{1:n_{#1'}}}}
+\newcommand{\model}[1]{\ensuremath{\mathcal{M}{(#1)}}}
+\newcommand{\simu}[1]{\ensuremath{\Omega(\model{#1})}}
+\newcommand{\micro}[2]{\ensuremath{\phi_{#1}(#2)}}
+\newcommand{\pdf}[1]{\ensuremath{\mathcal{P}_{d}(#1)}}
+\newcommand{\pmf}[1]{\ensuremath{\mathcal{P}_{m}(#1)}}
+\newcommand{\add}[1]{\ensuremath{A_{#1}}}
+\newcommand{\adds}[2]{\ensuremath{A_{#1}_{#2}}}
+\newcommand{\expect}[1]{\ensuremath{\mathbb{E}[#1]}}
+\newcommand{\trace}[1]{\ensuremath{\mathcal{T}_{#1}}}
+\newcommand{\traces}[2]{\ensuremath{\textit{tr}^{#1}_{#2}}}
+\newcommand{\stoc}[2]{\ensuremath{f^{#1}_{#2}(\rvv{x})}}
 % \usepackage[binary-units=true]{siunitx}
@@
 
+One such machine learning paradigm that has the ability to resolve these issues is \emph{Probabilistic programming}\cite{gordon2014probabilistic,milch2005blog,van2018introduction,tran2017deep}. 
+Probabilistic programming enables users to write, condition on and perform inference in, complex probabilistic models in a programmatic, yet efficient and simplified way. 
+The source code of the given model, or equivalently the \emph{program}, is transformed into a program density, which can
+be viewed as the joint density of the user-specified model, \emph{the simulator} in this instance.
+Having direct access to the source code enables PPSs to conduct several processes automatically, such as 
+the automation of complex statistical inference procedures\cite{zhou2019lf,goodman2012church,wood2014new,rainforth2017automating}, which enables us to efficiently calibrate
+model parameters, and we can also track how the internals of the program change during
+run time, enabling a full analysis of the program structure and traces generated\cite{milch2005blog}, which provides greater interpretability, see Figure~\ref{fig:hijacking}.
+In addition to this, as models are directly expressed in program code we can encode other structures to make use of the full structure of a given language, whether that is \texttt{python, C++} or 
+any other language,
+which enables users to integrate PPSs into their standard model development process, increasing efficiency and reducing model iteration time.
+
 \begin{wrapfigure}{R}{0.44\textwidth}
+\vspace{-20pt}
 \centering
@@
 \includegraphics[width=0.40\textwidth]{../plots/model_lambda/hist_posteriors.png}
+% \vspace{-10pt}
 \caption{An example of the posterior output of population-based simulator after hijacking.
@@
 The grey dashed lines represent the ground truth.}
+\vspace{-25pt}
 \label{fig:hijacking} 
 \end{wrapfigure}
-One such machine learning paradigm that has the ability to resolve these issues is \emph{Probabilistic programming}\cite{gordon2014probabilistic,le2016inference}. 
-Probabilistic programming enables users to write, condition on and perform inference in, complex probabilistic models in a programmatic, yet efficient and simplified way. 
-The source code of the given model, or equivalently the \emph{program}, is transformed into a program density, which can
-be viewed as the joint density of the user-specified model, \emph{the simulator} in this instance.
-Having direct access to the source code enables PPSs to conduct several processes automatically, such as 
-the automation of complex statistical inference procedures\cite{zhou2019lf,goodman2012church,wood2014new}, which enables us to efficiently calibrate
-model parameters, and we can also track how the internals of the program change during
-run time, enabling a full analysis of the program structure and traces generated, which provides greater interpretability, see Figure~\ref{fig:hijacking}.
-In addition to this, as models are directly expressed in program code we can encode other structures to make use of the full structure of a given language, whether that is \texttt{python, C++} or 
-any other language,
-which enables users to integrate PPSs into their standard model development process, increasing efficiency and reducing model iteration time.
-
 For all the brilliance of probabilistic programming, if one did want to run an existing simulator in a standard probabilistic programming set-up one would have to entirely re-write the simulator in the language of the PPS.
@@
 
-\paragraph{Related work} Our method of hijacking simulators builds on the work of~\cite{baydin2018efficient}. 
+\paragraph{Related work} Our method of hijacking simulators builds on the work of~\cite{baydin2018efficient,casado2017improvements}. 
 They demonstrated how this could be done for event-based simulators, but there methods neglected a large class of population-based simulators that model interactions
-between populations, that are utilized for modeling epidemics, multi-agent systems and financial economics  \bg{add citations}. 
+
+between populations, that are utilized for modeling epidemics, multi-agent systems and financial economics \cite{raberto2001agent,landau_binder_2014,hartig2011statistical,smith2008towards,bershteyn2018implementation}.
 Our framework provides a solution and makes it easy to transform arbitrary stochastic population-based simulator, in addition to the event-based simulators, into probabilistic programs, regardless of the complexity of the simulator.
 Our system supports simulators written in 13 different programming languages and enables the user to perform posterior inference over the density of the simulator, conditioned on what data the user has access too, including
-simulator output. We provide a theoretical justification for our approach and many examples.
+simulator output. 
+We provide a theoretical justification for our approach and many examples.
 % This enables two things.
 % 1) It enables software developers and researchers to understand complicated code bases. 
-% 2) It provides interpretable inference results that provide policy makers with predictions that are truly interpretable, in the sense that the end-user of the inference results understands what physical events led to inference outcome.
+% 2) It provides interpretable inference results that provide policy makers with predictions that are truly interpretable, in the sense that the end-user, of the inference results understands what physical events led to inference outcome.
 % This is critical in several domains, and is particularly important in the medical domain. We provide examples of both in Section~\textbf{TO ADD REF}.
@@
 
-%Moreover, though our focus is on hijacking epidemiology simulators, our general approach and framework can be applied more widely to a general class of stochastic simulators.
+% Moreover, though our focus is on hijacking epidemiology simulators, our general approach and framework can be applied more widely to a general class of stochastic simulators.
 
@@
 % thereby providing wide-ranging impacts across a number of fields.
-
-This paper first gives an overview of existing Malaria simulators (Section \ref{sec:background}), and proceed by introducing the necessary background on the \textit{pyprob} framework and the concept of universal probabilistic programming (Section \ref{sec:hijackingsimulators}).
-Our approach is then demonstrated and analysed in the context of a Malaria case study (Section \ref{sec:casestudy}).
 
@@
 
- \label{sec:methods}
+\begin{center}
+\begin{tabular}{ |c|c|c| }
+\hline
+Symbol & Meaning  \\
+\hline
+\n{x} & Number of latent variables \\
+\n{y} & Number of input observations \\
+\n{y’} & Number of output observations, generated simulator values \\
+\rv{x} & The latent parameters \\
+\rv{y} & The input observed parameters \\
+\rv{y’} & The output observed parameters \\
+\model{\cdot} & Model - i.e .xml file \\
+\simu{\cdot} & Simulator \\
+\micro{k}{\cdot} & $k$-th Micro-simulation \\
+\pdf{\cdot} & probability density function \\
+\pmf{\cdot} & probability mass function \\
+\add{i} & Address $i$ \\
+\expect{\cdot} & Expectation (first moment) \\
+\trace{j} & The $j$-th collapsed Trace \\
+\traces{i}{j} & A single, un-collapsed Trace for the $i$th trace. ``A program trace’' \\
+\stoc{h}{i} & A stochastic process at step $h$ for \add{i} \\
+\hline
+\end{tabular}
+\end{center}
+  \begin{algorithm}
+  \small
+   \caption{Hijacking a stochastic simulator for an arbitrary machine}
+    \begin{algorithmic}\small
+      \Function{Simulator to a Probabilistic Program}{Simulator $\Omega(\cdot)$, Model input to simulator \model{\cdot}, PPS sampling primitive $f_t()$, Message passing protocol (PPX) $h_t()$, Random variable $x_{sim}$, PPS $\mathcal{P}$ }
+       \State  $ x_{sim} \rightarrow f_{t}(x_{sim}) $\Comment{Replace raw sample statements}
+       \State $\simu{f_{t}(x^{t}_{sim})} \rightarrow  h_{t}(\simu{f_{t}(x^{t}_{sim})}) \rightarrow \mathcal{P}(f_{t}(x^{t}_{sim}))$ \Comment{An API between the simulator enables raw RNG to be passed through the PPS}
 
-  \begin{algorithm}
-   \caption{How to perform posterior inference over a stochastic simulator with 
-   }
-    \begin{algorithmic}[1]
-      \Function{Convert Simulator to a Probabilistic Program}{$A$}\Comment{}
-
-        \State replace raw samples statements $f_{t}(x_{t} | \eta_{t}) $
-        \State Create thin wrapper
-        \State Create Containerized Environment $C_{i}$
-        \State Create Connection
+        \Function{Create Containerized Environment}{$C(\Delta)$}
+        \For{$\delta$ in $\Delta$}
+        \State Add dependency $\delta$
+        \EndFor
         \State Run Program
@@
        \State to add
+       \EndFunction
        \EndFunction
@@
 
+\begin{figure*}
+  \centering
+  \includegraphics[width=\textwidth]{../plots/mastsari/mastsari_prior_trace_structure.pdf}
+  \caption{Prior trace probabilities of the Mastsari model.}
+  \label{fig:priortracemastsari}
+\end{figure*}
+
 \begin{table}[h!]

### 2019-05-22 paper/toypaper/main.tex
new abstract

--- 2019-05-21 paper/neurlps2019/main.tex
@@
 
-Probabilistic programming provides a way to perform statistical inference over simulations of events in a programmatic way.
+%Probabilistic programming provides a way to perform statistical inference over simulations of events in a programmatic way.
 Thus, by design, simulators are ideal programs for probabilistic programming systems~(PPSs).
-Simulators arise in a number of industrial scientific and industrial domains and encode sophisticated generative models.
-However, within existing PPSs one would have to re-implement the simulator via the PPS language specification, which is inefficient and often not feasible due to the complexity of such scientific and industrial simulators.
-Recent work by \cite{baydin2018efficient} demonstrated a pathway to turn a particular
-type of event-based simulator into a probabilistic program, without having to re-implement the simulator in the existing PPS. 
+%Simulators arise in a number of industrial scientific and industrial domains and encode sophisticated generative models.
+%However, within existing PPSs one would have to re-implement the simulator via the PPS language specification, which is inefficient and often not feasible due to the complexity of such scientific and industrial simulators.
+%Recent work by \cite{baydin2018efficient} demonstrated a pathway to turn a particular
+%type of event-based simulator into a probabilistic program, without having to re-implement the simulator in the existing PPS. 
 % via the raw random number calls of the simulator.
- However, this still meant that a large class of critically important population-based simulators could not be turned into probabilistic programs and as such could not be used within a probabilistic programming framework.
-In this work, we extend that framework to encompass population-based simulators, a very large class of simulators that are used extensively across epidemiology, multi-agent and financial modeling and provide the relevant
+% However, this still meant that a large class of critically important population-based simulators could not be turned into probabilistic programs and as such could not be used within a probabilistic programming framework.
+%In this work, we extend that framework to encompass population-based simulators, a very large class of simulators that are used extensively across epidemiology, multi-agent and financial modeling and provide the relevant
 theoretical underpinnings. 
-We enable simulators written in 13 different languages to be easily converted to probabilistic programs, for which we algorithmically outline and provide comprehensive examples of our system applied to a number of real-world population-based simulators.
+%We enable simulators written in 13 different languages to be easily converted to probabilistic programs, for which we algorithmically outline and provide comprehensive examples of our system applied to a number of real-world population-based simulators.
+
+Stochastic population-based simulators play an important role in many industrial and scientific settings.
+Many problems surrounding such simulators involve solving the “inverse problem”, which consists of finding the simulator parameters that generate output that is closest to a given set of field observations.
+The framework of probabilistic programming naturally allows to express and solve such inverse problems.
+Unfortunately, application of existing tools requires refactoring of existing simulators within specialised probabilistic programming frameworks.
+This is usually infeasible given the size and complexity of existing simulators.
+Recent work by Baydin et al now enables existing simulators to be treated as a probabilistic program without extensive modification.
+We, in this paper, extend these exciting developments in two ways: First, we develop a mathematical formalism that allows us to succinctly define posterior densities of programs generated from stochastic population-based simulators.
+We then illustrate the applicability of Baydin et al’s approach by hijacking a series of important stochastic population-based Malaria epidemiology models.
+Our approach is capable of finding suitable model parameters much more efficiently than currently used approaches in the literature and refines .
 

### 2019-05-22 examples/matsari_intervention/plot.py
updated

@@ -1,0 +1,31 @@
+from matplotlib import pyplot as plt
+import numpy as np
+
+age_pop = None
+age_inf = None
+with open("output_copy.txt", "r") as f:
+    tmp = list(f.readlines())
+    vals = list(map(lambda x:float(x.split("	")[-1]), tmp))
+    age_pop = vals[:12]
+    age_inf = vals[12:]
+
+om = np.array(age_inf) / np.array(age_pop)
+toy1 = np.array([0.1228, 0.1234, 0.1272, 0.1428, 0.1760, 0.2267, 0.1642, 0.1047, 0.0928,
+        0.0892, 0.0882, 0.0881])
+obs = np.array([0.001156069364162,
+       0.018208092485549,
+       0.050289017341041,
+       0.135260115606936,
+       0.038150289017341,
+       0.165606936416185,
+       0.172254335260116,
+       0.053757225433526,
+       0.038439306358382,
+       0.032947976878613,
+       0.039595375722543,
+       0.047109826589595])
+plt.plot(om)
+plt.plot(obs)
+plt.plot(toy1)
+plt.show()
+plt.savefig("out.png")

### 2019-05-26 paper/icml_ai4good/main.tex
updated;

@@
 % Use the following line for the initial blind version submitted for review:
-\usepackage{icml2019}
+% \usepackage{icml2019}
 \newcommand{\bg}[1]{~{{[{\it \textcolor{red}{{\bf BG:} #1}}]}}}
@@
 % If accepted, instead use the following line for the camera-ready submission:
-%\usepackage[accepted]{icml2019}
+\usepackage[accepted]{icml2019}
 \usetikzlibrary{arrows,chains,matrix,positioning,scopes, arrows.meta}
@@
 % in order of appearance and this is the preferred way.
+% \icmlsetsymbol{equal}{*}
+
+% \begin{icmlauthorlist}
+% \icmlauthor{Bradley.~J Gram-Hansen}{equal,en}
+% \icmlauthor{Christian Schroeder de Witt}{equal,en}
+% \icmlauthor{Tom Rainforth}{st}
+% \icmlauthor{Phil H.S. Torr}{eng}
+% \icmlauthor{Yee Whye Teh}{st}
+% \icmlauthor{At\i{}l\i{}m G\"{u}ne\c{s} Baydin}{eng}
+% \end{icmlauthorlist}
+
+% \icmlcorrespondingauthor{Bradley.~J Gram-Hansen}{bradley@robots.ox.ac.uk}
+% \icmlcorrespondingauthor{Eee Pppp}{ep@eden.co.uk}
 \icmlsetsymbol{equal}{*}
@@
 \begin{icmlauthorlist}
-\icmlauthor{Aeiau Zzzz}{equal,to}
-\icmlauthor{Bauiu C.~Yyyy}{equal,to,goo}
-\icmlauthor{Cieua Vvvvv}{goo}
-\icmlauthor{Iaesut Saoeu}{ed}
-\icmlauthor{Fiuea Rrrr}{to}
-\icmlauthor{Tateu H.~Yasehe}{ed,to,goo}
-\icmlauthor{Aaoeu Iasoh}{goo}
-\icmlauthor{Buiui Eueu}{ed}
-\icmlauthor{Aeuia Zzzz}{ed}
-\icmlauthor{Bieea C.~Yyyy}{to,goo}
-\icmlauthor{Teoau Xxxx}{ed}
-\icmlauthor{Eee Pppp}{ed}
+\icmlauthor{Bradley.~J Gram-Hansen}{equal,en}
+\icmlauthor{Christian Schr\"{o}der de Witt}{equal,en}
+\icmlauthor{Tom Rainforth}{st}
+\icmlauthor{Phil H.S. Torr}{en}
+\icmlauthor{Yee Whye Teh}{st}
+\icmlauthor{At\i{}l\i{}m G\"{u}ne\c{s} Baydin}{en}
 \end{icmlauthorlist}
 
-\icmlaffiliation{to}{Department of Computation, University of Torontoland, Torontoland, Canada}
-\icmlaffiliation{goo}{Googol ShallowMind, New London, Michigan, USA}
-\icmlaffiliation{ed}{School of Computation, University of Edenborrow, Edenborrow, United Kingdom}
+\icmlcorrespondingauthor{Bradley.~J Gram-Hansen}{bradley@robots.ox.ac.uk}
 
-\icmlcorrespondingauthor{Cieua Vvvvv}{c.vvvvv@googol.com}
-\icmlcorrespondingauthor{Eee Pppp}{ep@eden.co.uk}
+\icmlaffiliation{en}{Department of Engineering, University of Oxford, UK}
+\icmlaffiliation{st}{Department of Statistics, University of Oxford, UK}
 
@@
 \section{Introduction}
-Ending the epidemics of AIDS, tuberculosis, malaria and other infectious diseases by 2030 is a key target within the Good Health \& Well-Being section of the UN Sustainable Development Goals \cite{refugees_2030,un_sustainable2018}.
+Ending the epidemics of AIDS, tuberculosis, malaria and other infectious diseases by 2030 is a key target within the Good Health \& Well-Being section of the UN Sustainable Development Goals \cite{unitednations,un_sustainable2018}.
 However, despite decades of substantial international efforts, these diseases kill hundreds of million people a year.
@@
 
-This paper first gives an overview of existing Malaria simulators (Section \ref{sec:background}), and proceed by introducing the necessary background on the \textit{pyprob} framework and the concept of universal probabilistic programming (Section \ref{sec:hijackingsimulators}).
+This paper first gives an overview of existing malaria simulators (Section \ref{sec:background}), and proceed by introducing the necessary background on the \textit{pyprob} framework and the concept of universal probabilistic programming (Section \ref{sec:hijackingsimulators}).
 Our approach is then demonstrated and analysed in the context of a Malaria case study (Section \ref{sec:casestudy}).
@@
 Epidemiology studies the prevalence and spreading of diseases across populations.
-Recent advances in hardware have enabled simulations to model the dynamics of infectious diseases, such as Malaria, in ever greater detail.
+Recent advances in hardware have enabled simulations to model the dynamics of infectious diseases, such as malaria, in ever greater detail.
 
@@
 
-Two the most advanced Malaria simulators, namely EMOD~\cite{bershteyn2018implementation} and OpenMalaria  \cite{smith2008towards}, have proven to be particularly valuable to policymakers.
-OpenMalaria is based on microsimulations of Plasmodium falciparum in humans and was originally developed to simulate the impacts of malaria vaccines within simple villages or districts
+Two the most advanced malaria simulators, namely EMOD~\cite{bershteyn2018implementation} and OpenMalaria  \cite{smith2008towards}, have proven to be particularly valuable to policymakers.
+OpenMalaria is based on microsimulations of \textit{Plasmodium falciparum} in humans and was originally developed to simulate the impacts of malaria vaccines within simple villages or districts
 Compared with OpenMalaria, EMOD is able to simulate a variety of additional drivers, including complex geographies complete with migration and a large number of policy interventions
@@
 can be used to express probabilistic models and consequently perform automated inference these.
-Once a probabilistic model has been expressed in a probabilistic programming language, a wide range of inference techniques, such as Markov Chain Monte Carlo (MCMC)\cite{}, Variational Inference~(VI)\cite{} and Deep Neural Networks~(DNN)\cite{},  can be used by non-experts in an automated fashion.
+Once a probabilistic model has been expressed in a probabilistic programming language, a wide range of inference techniques, such as Markov Chain Monte Carlo (MCMC)\cite{geyer1992practical}, Black-box Variational Inference~(VI)\cite{ranganath2014black} and amortized inference\cite{le2016inference},  can be used by non-experts in an automated fashion.
 
@@
 Finally, the entry point to the simulator, i.e. \textit{main} in C++, is replaced by a special \textit{forward} call, in this case \textit{pyprob\_cpp::forward}.
-This allows the PPS to generate rollouts from the simulator remotely.
+This allows the PPS to generate rollouts from the simulator remotely. A pictorial overview of this process is depicted in Figure~\ref{fig:how}.
 
@@
 \centering
- \scalebox{.5}{
-\begin{tikzpicture}[node distance=1.5cm,
-    every node/.style={fill=white, font=\sffamily}, align=center]
-  % Specification of nodes (position, etc.)
-  % \node (start)             [activityStarts]              {Activity starts};
-  \node (SimulatorSystem)     [activityStarts]          {Simulator System \\ e.g. EMOD};
-   \node (randomprim)      [process, left of=SimulatorSystem, xshift=5.6cm]
-                                                      {Redefine RNG \\ to make \\ externally \\ callable};
-  \node (onStartBlock)      [process, below of=SimulatorSystem]   {Specify Model \\ e.g. XML files};
-  \node (activityRuns)      [activityRuns, below of=onStartBlock]
-                                                      {Run Simulator};
-  \node (onPauseBlock)      [startstop, below of=activityRuns, yshift=-1cm]
-                                                                {Posterior Output};
-  % \node (onStopBlock)       [process, below of=onPauseBlock, yshift=-1cm]
-  %                                                                {};
-  % \node (onDestroyBlock)    [process, below of=onStopBlock, yshift=-1cm] 
-  %                                                             {onDestroy()};
-  \node (onRestartBlock)    [process, right of=activityRuns, xshift=3cm]
-                                                              {PPS hijacks \\ random number \\ generator};
-  \node (ActivityEnds)      [startstop, below of=onRestartBlock, yshift=-1cm]
-                                                        {Result summaries \\ path probabilities};
-  % \node (ActivityDestroyed) [startstop, below of=onDestroyBlock]
-  %                                                   {Activity is shut down};     
-  % Specification of lines between nodes specified above
-  % with aditional nodes for description 
-  % \draw[->]             (start) -- (onCreateBlock);
-  \draw[->]     (SimulatorSystem) -- (onStartBlock);
-  \draw[dashed, <-]  (SimulatorSystem) -- (randomprim);
-  % \draw[->]      (onStartBlock) -- (onResumeBlock);
-  \draw[->]     (onStartBlock) -- (activityRuns);
-  \draw[->]      (activityRuns) -- (onPauseBlock);
-  % \draw[->]      (onPauseBlock) -- node {The activity is no longer visible}
-  %                                  (onStopBlock);
-  % \draw[->]       (onStopBlock) -- node {The activity is shut down by
-  %                                  user or system} (onDestroyBlock);
-  % \draw[->]    (onRestartBlock) -- (activityRuns);
-  \path[->, shift left=.75ex]
-  (onRestartBlock) edge (activityRuns)
-  (activityRuns) edge (onRestartBlock);
-  % \draw[<-]    (activityRuns) -- (onRestartBlock);
-  % \draw[->]       (onStopBlock) -| node[yshift=1.25cm, text width=3cm]
-  %                                  {The activity comes to the foreground}
-  %                                  (onRestartBlock);
+%  \scalebox{.5}{
+% \begin{tikzpicture}[node distance=1.5cm,
+%     every node/.style={fill=white, font=\sffamily}, align=center]
+%   % Specification of nodes (position, etc.)
+%   % \node (start)             [activityStarts]              {Activity starts};
+%   \node (SimulatorSystem)     [activityStarts]          {Simulator System \\ e.g. EMOD};
+%    \node (randomprim)      [process, left of=SimulatorSystem, xshift=5.6cm]
+%                                                       {Redefine RNG \\ to make \\ externally \\ callable};
+%   \node (onStartBlock)      [process, below of=SimulatorSystem]   {Specify Model \\ e.g. XML files};
+%   \node (activityRuns)      [activityRuns, below of=onStartBlock]
+%                                                       {Run Simulator};
+%   \node (onPauseBlock)      [startstop, below of=activityRuns, yshift=-1cm]
+%                                                                 {Posterior Output};
+%   % \node (onStopBlock)       [process, below of=onPauseBlock, yshift=-1cm]
+%   %                                                                {};
+%   % \node (onDestroyBlock)    [process, below of=onStopBlock, yshift=-1cm] 
+%   %                                                             {onDestroy()};
+%   \node (onRestartBlock)    [process, right of=activityRuns, xshift=3cm]
+%                                                               {PPS hijacks \\ random number \\ generator};
+%   \node (ActivityEnds)      [startstop, below of=onRestartBlock, yshift=-1cm]
+%                                                         {Result summaries \\ path probabilities};
+%   % \node (ActivityDestroyed) [startstop, below of=onDestroyBlock]
+%   %                                                   {Activity is shut down};     
+%   % Specification of lines between nodes specified above
+%   % with aditional nodes for description 
+%   % \draw[->]             (start) -- (onCreateBlock);
+%   \draw[->]     (SimulatorSystem) -- (onStartBlock);
+%   \draw[dashed, <-]  (SimulatorSystem) -- (randomprim);
+%   % \draw[->]      (onStartBlock) -- (onResumeBlock);
+%   \draw[->]     (onStartBlock) -- (activityRuns);
+%   \draw[->]      (activityRuns) -- (onPauseBlock);
+%   % \draw[->]      (onPauseBlock) -- node {The activity is no longer visible}
+%   %                                  (onStopBlock);
+%   % \draw[->]       (onStopBlock) -- node {The activity is shut down by
+%   %                                  user or system} (onDestroyBlock);
+%   % \draw[->]    (onRestartBlock) -- (activityRuns);
+%   \path[->, shift left=.75ex]
+%   (onRestartBlock) edge (activityRuns)
+%   (activityRuns) edge (onRestartBlock);
+%   % \draw[<-]    (activityRuns) -- (onRestartBlock);
+%   % \draw[->]       (onStopBlock) -| node[yshift=1.25cm, text width=3cm]
+%   %                                  {The activity comes to the foreground}
+%   %                                  (onRestartBlock);
 
-  \draw[->]    (onRestartBlock) -- (ActivityEnds);
-  % \draw[->]      (onPauseBlock) -| node(priorityXMemory)
-  %                                  {todo}
-  %                                  (ActivityEnds);
-  % \draw           (onStopBlock) -| (priorityXMemory);
-  % \draw[->]     (ActivityEnds)  |- node [yshift=-2cm, text width=3.1cm]
-  %                                   {User navigates back to the activity}
-  %                                   (SimulatorSystem);
-  % \draw[->] (onPauseBlock.east) -- ++(2.6,0) -- ++(0,2) -- ++(0,2) --                
-  %    node[xshift=1.2cm,yshift=-1.5cm, text width=2.5cm]
-  %    {The activity comes to the foreground}(onResumeBlock.east);
-  \end{tikzpicture}
-  }
+%   \draw[->]    (onRestartBlock) -- (ActivityEnds);
+%   % \draw[->]      (onPauseBlock) -| node(priorityXMemory)
+%   %                                  {todo}
+%   %                                  (ActivityEnds);
+%   % \draw           (onStopBlock) -| (priorityXMemory);
+%   % \draw[->]     (ActivityEnds)  |- node [yshift=-2cm, text width=3.1cm]
+%   %                                   {User navigates back to the activity}
+%   %                                   (SimulatorSystem);
+%   % \draw[->] (onPauseBlock.east) -- ++(2.6,0) -- ++(0,2) -- ++(0,2) --                
+%   %    node[xshift=1.2cm,yshift=-1.5cm, text width=2.5cm]
+%   %    {The activity comes to the foreground}(onResumeBlock.east);
+%   \end{tikzpicture}
+% }
+   \includegraphics[width=0.45\textwidth]{../plots/system.pdf}
   \label{fig:how}
-  \caption{This flow-chart provides an overview of the process of how our system hijacks a generic population-based epidemiology simulator, such as EMOD, and how we modify the simulator to 
-  to hijack the random number generator (RNG).}
+  \caption{This flow-chart provides an overview of the process of how our we hijack a generic population-based epidemiology simulator, such as OpenMalaria, and how we modify the simulator to 
+  to hijack the random number generator (RNG). It demonstrates how information is exchanged between the simulator and the PPS. Green represents events linked to the simulator, purple corresponds to 
+  events occurring in the PPS and brown represents an external processes.}
 \end{figure}
@@
 Ensemble methods are commonly used in statistics in order to combine the predictive power of multiple models \cite{cameron2015defining,smith_ensemble_2012}.
-To this end, recent work has attempted to characterise the similarities and differences between two of the most advanced Malaria epidemiology simulators, EMOD \cite{bershteyn2018implementation} and OpenMalaria \cite{smith2008towards}.
+To this end, recent work has attempted to characterise the similarities and differences between two of the most advanced malaria epidemiology simulators, EMOD \cite{bershteyn2018implementation} and OpenMalaria \cite{smith2008towards}.
 Evaluation is usually done by comparing a number of output parameters across a range of hand-crafted standard scenarios reflecting different geographical locations across Africa~\cite{smith_ensemble_2012}.
@@
 In the following, we illustrate how our method introduces a novel introspection paradigm.
-By extracting trace graphs from population-based simulators, policymakers can ask specific questions about properties of the model trace flow.
+By extracting trace graphs from population-based simulators, policymakers can ask specific questions about properties of the model trace flow
+and not just the outcomes of the model, thus providing additional interpretability to the decision making process.
 
@@
 	\includegraphics[width=\textwidth/2]{../plots/EIR_Ifakara.pdf}
+  \vspace{-20pt}
 	\caption{Seasonal Entomological Inoculation Rate (EIR) for the Ifakara scenario. Data is averaged over $30$ day periods.}
 	\label{fig:EIR}
+  \vspace{-10pt}
 \end{figure}
@@
 Additionally, by associating nodes in trace graphs representing the same physical processes within different models, the significance of model detail can be evaluated.
-For example, the full trace presented in Table \ref{table:asingleaddress} represent a sampling step associated with within-host dynamics of the Malaria parasite \textit{Falciparum} in OpenMalaria node $A1$.
+For example, the full trace presented in Table \ref{table:asingleaddress} represent a sampling step associated with within-host dynamics of the malaria parasite \textit{Falciparum} in OpenMalaria node $A1$.
 The same physical process also occurs in EMOD's trace graph at position $A7$ (see Section \ref{table:addresses}).
@@
   \caption{An example of an address generated for the model run in the OpenMalaria simulator. We can see that A1
-  relates to Generating a member of the human population who may or may not be infect with the Malaria disease.
+  relates to Generating a member of the human population who may or may not be infect with the malaria disease.
 We get something similar for EMOD, except this relates to A7 in the EMOD program execution.}
@@
 In this work we have demonstrated a method 
-that enables one to hijack populationbased simulators, extending 
+that enables one to hijack population-based simulators, extending 
 the work of \citep{baydin2018efficient}. 
-We applied our method to two Malaria orientated population-based simulators and 
+We applied our method to two malaria orientated population-based simulators and 
 generated a variety of trace graphs.
@@
 tools that will facilitate complicated inference procedures that condition on simulator 
-output. We will also evaluate additional scenarios across Africa and South-East Asia to better understand the similarities and differences between EMOD and OpenMalaria.
+output. We will also evaluate additional scenarios across Africa and Southeast Asia to better understand the similarities and differences between EMOD and OpenMalaria.
 
-\quad\\
-\quad\\
+\section{Acknowledgments}
+We thank Ewan Cameron of MAP at the Big Data Institute and the OpenMalaria team at the Swiss Tropical Health Institute for their time and help.
+BGH is supported by the EPRSC Autonomous  Intelligent Machines and Systems grant.
+CSW  is supported by the project Free the Drones (FreeD) under the Innovation Fund Denmark and Microsoft.
+YWT’s and TR’s research leading to these results has received funding from the European Research Council under the European Union’s Seventh Framework Programme (FP7/2007-2013) ERC grant agreement no.
+617071.
+AGB and PH are supported by EPSRC/MURI grant EP/N019474/1 and AGB is also supported by Lawrence Berkeley National Lab.
 

### 2019-05-26 paper/icml_ai4good/abstract.tex
cr abstract

--- 2019-04-27 paper/icml_ai4good/main.tex
@@
-%%%%%%%% ICML 2019 EXAMPLE LATEX SUBMISSION FILE %%%%%%%%%%%%%%%%%
-
-\documentclass{article}
-
-% Recommended, but optional, packages for figures and better typesetting:
-\usepackage{microtype}
-\usepackage{graphicx}
-\usepackage{subfigure}
-\usepackage{booktabs} % for professional tables
-\usepackage{tabularx}
-\usepackage{listings}
-\usepackage{pgfplots}
-\usepackage{tikz}
-\usepackage{todonotes}
-\usepackage{tikz-cd}
-\usepackage[binary-units=true]{siunitx}
-
-% hyperref makes hyperlinks in the resulting PDF.
-% If your build breaks (sometimes temporarily if a hyperlink spans a page)
-% please comment out the following usepackage line and replace
-% \usepackage{icml2019} with \usepackage[nohyperref]{icml2019} above.
-\usepackage{hyperref}
-
-% \usepgfplotslibrary{external}
-% Attempt to make hyperref and algorithmic work together better:
-\newcommand{\theHalgorithm}{\arabic{algorithm}}
-
-% Use the following line for the initial blind version submitted for review:
-\usepackage{icml2019}
-\newcommand{\bg}[1]{~{{[{\it \textcolor{red}{{\bf BG:} #1}}]}}}
-\newcommand{\cs}[1]{~{{[{\it \textcolor{red}{{\bf CS:} #1}}]}}}
-\newcommand{\ag}[1]{~{{[{\it \textcolor{red}{{\bf AG:} #1}}]}}}
-% If accepted, instead use the following line for the camera-ready submission:
-%\usepackage[accepted]{icml2019}
-\usetikzlibrary{arrows,chains,matrix,positioning,scopes, arrows.meta}
-\tikzset{
-  shift left/.style ={commutative diagrams/shift left={#1}},
-  shift right/.style={commutative diagrams/shift right={#1}}
-}
-\tikzset{%
-  >={Latex[width=2mm,length=2mm]},
-  % Specifications for style of nodes:
-            base/.style = {rectangle, rounded corners, draw=black,
-                           minimum width=4cm, minimum height=1cm,
-                           text centered, font=\sffamily},
-  activityStarts/.style = {base, fill=blue!50},
-       startstop/.style = {base, fill=red!50},
-    activityRuns/.style = {base, fill=green!50},
-         process/.style = {base, minimum width=2.5cm, fill=orange!55,
-                           font=\ttfamily},
-}
-\makeatletter
-\tikzset{join/.code=\tikzset{after node path={%
-\ifx\tikzchainprevious\pgfutil@empty\else(\tikzchainprevious)%
-edge[every join]#1(\tikzchaincurrent)\fi}}}
-\makeatother
-%
-\tikzset{>=stealth',every on chain/.append style={join},
-         every join/.style={->}}
-\tikzstyle{labeled}=[execute at begin node=$\scriptstyle,
-   execute at end node=$]
-% The \icmltitle you define below is probably too long as aer.
-% Therefore, a short form for the running title is supplied here:
-%\icmltitlerunning{Hijacking Simulators for the Greater Good}
-\icmltitlerunning{Hijacking Malaria Simulators
-	with Probabilistic Programming}
-
-\lstset{language=C++,
-                basicstyle=\footnotesize,
-                keywordstyle=\color{blue}\ttfamily,
-                stringstyle=\color{red}\ttfamily,
-                commentstyle=\color{green}\ttfamily,
-                morecomment=[l][\color{magenta}]{\#},
-                breaklines = true
-}
-\pgfplotsset{width=7cm,compat=1.8}
-\begin{document}
-
-\twocolumn[
-%\icmltitle{Hijacking Simulators for the Greater Good}
-%\icmltitle{Automatically Converting Epidemiology Simulations
-%	to Probabilistic Programs}
-\icmltitle{Hijacking Malaria Simulators
-	with Probabilistic Programming}
-
-% It is OKAY to include author information, even for blind
-% submissions: the style file will automatically remove it for you
-% unless you've provided the [accepted] option to the icml2019
-% package.
-
-% List of affiliations: The first argument should be a (short)
-% identifier you will use later to specify author affiliations
-% Academic affiliations should list Department, University, City, Region, Country
-% Industry affiliations should list Company, City, Region, Country
-
-% You can specify symbols, otherwise they are numbered in order.
-% Ideally, you should not use this facility. Affiliations will be numbered
-% in order of appearance and this is the preferred way.
-\icmlsetsymbol{equal}{*}
-
-\begin{icmlauthorlist}
-\icmlauthor{Aeiau Zzzz}{equal,to}
-\icmlauthor{Bauiu C.~Yyyy}{equal,to,goo}
-\icmlauthor{Cieua Vvvvv}{goo}
-\icmlauthor{Iaesut Saoeu}{ed}
-\icmlauthor{Fiuea Rrrr}{to}
-\icmlauthor{Tateu H.~Yasehe}{ed,to,goo}
-\icmlauthor{Aaoeu Iasoh}{goo}
-\icmlauthor{Buiui Eueu}{ed}
-\icmlauthor{Aeuia Zzzz}{ed}
-\icmlauthor{Bieea C.~Yyyy}{to,goo}
-\icmlauthor{Teoau Xxxx}{ed}
-\icmlauthor{Eee Pppp}{ed}
-\end{icmlauthorlist}
-
-\icmlaffiliation{to}{Department of Computation, University of Torontoland, Torontoland, Canada}
-\icmlaffiliation{goo}{Googol ShallowMind, New London, Michigan, USA}
-\icmlaffiliation{ed}{School of Computation, University of Edenborrow, Edenborrow, United Kingdom}
-
-\icmlcorrespondingauthor{Cieua Vvvvv}{c.vvvvv@googol.com}
-\icmlcorrespondingauthor{Eee Pppp}{ep@eden.co.uk}
-
-% You may provide any keywords that you
-% find helpful for describing your paper; these are used to populate
-% the "keywords" metadata in the PDF but will not be shown in the document
-\icmlkeywords{Machine Learning, ICML}
-
-\vskip 0.3in
-]
-
-% this must go after the closing bracket ] following \twocolumn[ ...
-
-% This command actually creates the footnote in the first column
-% listing the affiliations and the copyright notice.
-% The command takes one argument, which is text to display at the start of the footnote.
-% The \icmlEqualContribution command is standard text for equal contribution.
-% Remove it (just {}) if you do not need this facility.
-
-%\printAffiliationsAndNotice{}  % leave blank if no need to mention equal contribution
-\printAffiliationsAndNotice{\icmlEqualContribution} % otherwise use the standard text.
-
-\begin{abstract}
 Epidemiology simulations have become a fundamental tool in the fight
 against the epidemics of various infectious diseases like AIDS and
-malaria.
-However, the complicated and stochastic nature of these simulators
+malaria. However, the complicated and stochastic nature of these simulators
 can mean their output is difficult to interpret, which reduces their usefulness to policymakers.
 In this paper, we introduce an approach that allows one to treat a large class of population-based epidemiology simulators as probabilistic generative models.
-This is achieved by \emph{hijacking} the internal random number generator calls, through the use of an universal probabilistic programming system (PPS).
+This is achieved by hijacking the internal random number generator calls, through the use of a universal probabilistic programming system (PPS).
 In contrast to other methods, our approach can be easily retrofitted to simulators written in popular industrial programming frameworks.
 We demonstrate that our method can be used for interpretable introspection and inference, thus shedding light on black-box simulators.
-This reinstates much needed trust between policymakers and evidence-based methods.
-
-%This allows for various advanced statistical analyses to be performed
-%on the simulators in an automated fashion, while also tracking
-%the behavior of the simulator to uncover key information such
-%as stochastic structure and generate result summaries.
-%Critically, this hijacking can be performed with only minimal changes
-%to the original simulator, allowing for straightforward application
-%of our approach to a wide variety of contexts and existing simulators.
-%
-%A goal of probabilistic programming is to couple simulators, with inference. This is 
-%because stochastic simulators are used prominently in many industrial settings,
-%do not require one to construct hand-crafted joint distributions as they implicitly 
-%define a joint distribution of the program and encode learned structures 
-%directly. This makes simulators powerful tools and much of machine learning (ML) and 
-%Artificial Intelligence (AI)
-%can be seen as trying to emulate such simulators from a purely data-driven approach.
-%However, in the 
-%ML/AI setting, although we can often infer outcomes, we have little understanding about what 
-%in the data led to the outputted inferences. 
-%This makes it challenging to deploy ML/AI systems into the wild, especially in health-related and safety-critical domains, such
-%as epidemiology, as we lose \emph{interpretability}. 
-%In this work, we explain how to design ML/AI systems that combine
-%probabilistic programming systems (PPSs) and epidemiology simulators, to extract
-%fully interpretable structures, enabling policy makers 
-%and practitioners to make decisions from simulator outputs that can be fully reasoned on.
-% % interpretable inferences. 
-%In particular, we demonstrate this for the Malaria disease in two commonly used simulators; EMOD and OpenMalaria.
-\end{abstract}
-
-\section{Introduction}
-Ending the epidemics of AIDS, tuberculosis, malaria and other infectious diseases by 2030 is a key target within the Good Health \& Well-Being section of the UN Sustainable Development Goals \cite{refugees_2030,un_sustainable2018}.
-However, despite decades of substantial international efforts, these diseases kill hundreds of million people a year.
-For example, malaria still annually kills about a quarter of a million children under the age of 5 in Africa alone.
-
-%maybe cut some of this paragraph out
-% In contrast to AIDS, Malaria can be routinely cured thanks to the disposability of a variety of medical interventions. 
-% Issues are further aggravated due to the increasing emergence of parasite resistance to multiple drugs, including artemisinin, or even insecticides. 
-% Hence, despite decades of substantial international efforts, initial hopes of regional elimination or even global eradication have not been met.
-%  So difficult is Malaria control that, even today, about a quarter of a million children under the age of 5 succumb to it every year in Africa alone.
-
-To reach the WHO’s target of reducing malaria incidence and mortality rates by at least 90\% by 2030, policymakers are increasingly turning to evidence-based methods, thus oftentimes relying on computational simulations \cite{who_global_2015}.
-These simulations allow policymakers to infer critical information on disease dynamics and make predictions about the impacts of policies
-before they are rolled out. This frequently increases the effectiveness of interventions and thus ultimately saves resources, or even lives.
-For example, it has been shown that mass vaccination may be largely ineffective in regions of large transmission rates, but may play a crucial role in areas of low transmission \cite{cameron2015defining}.
-
-Malaria epidemiology is governed by a complex set of drivers, 
-few of which can be understood in isolation \cite{cameron2015defining,autino_epidemiology_2012,smith2008towards,bershteyn2018implementation}.
-These include within-host dynamics, population-specific traits and even local geography.
-Comprehensive modeling of all of these components remains challenging, particularly in a region-specific context. 
-Computational epidemiology simulators have to reflect these complexities and are usually stochastic in nature.
-This can make simulation output highly non-trivial to interpret, particularly when trying to draw desired inferences coupled with observed data \cite{mwendera_challenges_2019,ferris_openmalaria_2015}.
-
-In this paper, we introduce a novel method that allows one to shed light on the inner workings of a large class of population-based stochastic simulators.
-We achieve this by extending the work of \citep{baydin2018efficient} by interpreting such population-based simulators as probabilistic generative models within the framework of universal probabilistic programming (UPP) \cite{le-2016-inference}.
-To this end, we \emph{hijack} existing simulators by overriding their internal random number generators.
-Specifically, by replacing the existing low-level random number generator in a simulator with a call to a purpose-built UPP ``controller'', which can thus control, track and manipulate the stochasticity of the simulator.
-
-This allows for a variety of tasks to be performed on
-the hijacked simulator, such as running inference (by conditioning
-the values of certain draws and manipulating others),
-uncovering stochastic structure, and automatically
-producing result summaries, such as establishing the probability
-of different program paths/traces.  By providing a common abstraction 
-framework for different simulators, our approach further allows for
-easy and direct comparison between related or competing
-simulators, a characteristic that is valuable in the context of
-epidemiology simulators \cite{ferris_openmalaria_2015}. We provide a case study of the above in Section \ref{sec:casestudy}.
-
-Our framework already supports application to simulators written in 13 industrial programming frameworks, and is easily extensible.
-This is crucial as, given the enormous code size and complexity, rewriting epidemiology simulators using a dedicated universal probabilistic programming language, such as Pyro \cite{bingham2019pyro}, is often infeasible.
-%Moreover, though our focus is on hijacking epidemiology simulators, our general approach and framework can be applied more widely to a general class of stochastic simulators.
-
-In time, we hope our approach will play a critical role in
-bringing recent advancements in probabilistic programming to bear on
-the vast array of existing simulators used throughout the sciences,
-thereby providing wide-ranging impacts across a number of fields.
-
-This paper first gives an overview of existing Malaria simulators (Section \ref{sec:background}), and proceed by introducing the necessary background on the \textit{pyprob} framework and the concept of universal probabilistic programming (Section \ref{sec:hijackingsimulators}).
-Our approach is then demonstrated and analysed in the context of a Malaria case study (Section \ref{sec:casestudy}).
-
-%Our approach is extending recent work by  and is unique in that it can easily be retrofitted to population-based simulators written in popular industrial programming frameworks.
-This is crucial as, given the enormous code size and complexity, rewriting epidemiology simulators using a dedicated universal probabilistic programming language, such as Pyro\cite{}, is often infeasible.
-
-%that allows one to treat a large class of population-based epidemiology simulators as probabilistic generative models.
-This is achieved by \emph{hijacking} the internal random number generator calls, through the use of an universal probabilistic programming system (PPS).
-In contrast to other methods, our approach requires only minimal changes to existing simulators to be made.
-%We demonstrate that our method can be used for interpretable introspection of and inference on what used to be de facto black box simulators.
-This reinstates much needed trust between policymakers and evidence-based methods.
-
-%In this paper, we show how probabilistic programming
-%techniques CITATIONS can be used to help automate the required calculations to do this, interpreting the simulators as 
-%generative models to perform principled statistical analysis, while
-%providing insights and summary statistics on their behavior to users.
-
-%Achieving this aim is substantially complicated by the fact that
-%a vast array of different epidemiology simulators are currently 
-%used in practice.
-%Manually reimplementing these systems in existing
-%probabilistic programming systems (PPSs) is therefore infeasible, as each
-%system would require manual and laborious porting.  Instead, we 
-%take the novel approach of \emph{hijacking} existing simulators by
-%overriding their internal random number generator, building
-%on ideas from~\citep{baydin2018efficient}.  Specifically, by
-%replacing the existing low-level random number generator in a simulator
-%with a call to a purpose-built ``controller'' PPS, the stochasticity
-%of the simulator can be both manipulated and tracked by this controller
-%system.  In turn, this allows for a variety of tasks to be performed on
-%the hijacked simulator, such as running inference (by conditioning
-%the values of certain draws and manipulating others),
-%uncovering stochastic structure, and automatically
-%producing result summaries, such as establishing the probability
-%of different program paths.  By providing a common abstraction 
-%framework for different simulators, our approach further allows for
-%easy and direct comparison between related or competing
-%simulators, a characteristic that is valuable in the context of
-%epidemiology simulators CITATIONS.
-
-%Critically, this hijacking can be performed on a wide variety of simulators
-%with only minimal changes.  Our framework already
-%supports 13 popular languages for the base simulator, and can easily
-%be extended to others if needed.  Moreover, though our focus
-%is on hijacking epidemiology simulators, our general approach and
-%framework can be applied more widely to a general class of stochastic
-%simulators.  In time, we hope our approach will play a critical role in
-%bringing recent advancements in probabilistic programming to bear on
-%the vast array of existing simulators used throughout the sciences,
-%thereby providing wide-ranging impacts across a number of fields.
-
-%However, the collection of large quantities of field data is often infeasible due to logistic, economic or political constraints, or outright impossible if novel methods are to be assessed a priori. 
-%Hence in silico simulation of malaria epidemiology plays an increasingly crucial role in evidence-based policymaking.
-%Even in the absence of political, economic and social constraints, 
-%Malaria epidemiology is governed by a complex set of drivers, few of which can be understood in isolation.
-%These include within-host dynamics, population-specific traits and even local geography.
-%Some progress in modeling specific components has been made, such as the parasite life-cycle, transmission vector dynamics and the inter-dependencies between infection rates and vicinity to open water.
-%However, comprehensive modeling of all of these components in a regional context remains challenging, and is greatly impeded by the lack of available field data for regional calibration, hence the need for comprehensive simulators.
-% Simulation complexity is further aggravated by considering feedback effects from a large arsenal of anti-malaria interventions, ranging from distribution of mosquito nets, indoor residual spraying and mass drug administration to age-specific vaccinations.
-%Stochastic epidemiology simulators have already proven to be an invaluable tool to policymakers with policy implications that are often counterintuitive yet far-ranging.
-
-%However, since many of these simulators are stochastic in nature and therefore act as black-boxes, understanding the structure of these complex simulators is challenging and poses several challenges for policy makers who want to interpret the decisions being inferred.
-%Therefore, sophisticated tools are required that enable us to not only digest the results of the simulation, but provide pathways to perform complex inferences.
-
-%To this end, we extend the work of \citep{baydin2018efficient} to enable multiple stochastic epidemiology simulators to be transformed into a probabilistic model that can be used as input to a probabilistic programming system, enabling trace paths to be extracted from the simulator on both a per human level and a population level, from which we can extract trace probabilities allowing policy makers to determine what actions have the biggest impact on differing populations dynamics and thus shape policy decisions accordingly.
-% We argue that our work is likely to have significant social impact on many livelihoods affected by Malaria and discuss opportunities for future research.
-
-% Recent work by \citep{Baydin} introduces a fundamental paradigm change that allows to tackle the inverse problem directly. 
-% Using a novel software tool named \textit{pyprob}, \citep{Baydin} extract program execution traces from state-of-the-art particle decay simulation engine SHERPA\cite{}.
-%  Using latest techniques from universal probabilistic programming inference, the sampled execution traces are then used in order to characterise the posterior distribution of the simulator output. 
-%  From these, regions in simulator input space can be associated with specified outcomes.
-
-% SHERPA, however, is different from leading epidemiology simulators in that it does not simulate population dynamics, but instead models individual decays with different sampling strategies.
-Thus, the existing work of \cite{baydin2018efficient} must be extended to encapsulate these different strategies.
-%  In this paper, we for the first time demonstrate the applicability of \textit{pyprob}, and hence probabilistic programming inference, to population-based epidemiology simulators. 
-%  In particular, we discuss how the leading malaria epidemiology simulators EMOD\cite{} and OpenMalaria\cite{} can be interfaced with \textit{pyprob}.
-%   We argue that our work is likely to have significant social impact on many livelihoods affected by Malaria and discuss opportunities for future research.
-
-% maybe talk about this framework enables a pathway to inference etc..
-
-% All of the above limitations require contemporary malaria epidemiology simulators to employ stochastic modeling techniques, which negotiate data scarcity and model uncertainty through the use of appropriate prior distributions that can be tuned even from small samples of field data.
-The population dynamics of groups of typically several thousand hosts are then initialised by sampling from the initial distributions, using initial random seeds.
-Forward simulation then proceeds by rolling out population dynamics over time.
-Simulation outcomes of cherry-picked scenarios and interventions of interest are then compared to suitably designed control runs.
-
-% This is great, if we had inference implemented right now.
-% However valuable forward simulations of epidemiology scenarios are, the holy grail of goal-driven policymaking ultimately lies in solving the inverse problem: Given a desired outcome and regional context, identify the series of interventions that will best reach the desired target after a given amount of time.
-Due to computational constraints, even an approximate search over the whole simulator input space usually proves intractable.
-An experience-guided search over the simulation output of a small set of manually-designed scenarios of interest therefore long seemed the only feasible way of investigating implications of policy interventions.
-
-% , and, above all, the lack of available field data for calibration,
-
-% parasite life-cycle, transmission vectors, population dynamics and even local geographic circumstances, with policy measures ranging from distribution of mosquito nets to experimental vaccines, requires such simulators to be stochastic.
-Simulation outcomes thus need to be evaluated on a population of subjects.
-
-% <Section about prob prog inference, with reference to SHERPA>
-
-%- issues of public health are complex and
-
-%To do. 
-%
-%\begin{itemize}
-%
-%\item Outline the objectives
-%\item State the importance of interpretability in the inference of the simulator
-%\item Our system enables one to understand and interpret the most common paths in a program (simulation).
-%\item In highlighting this a practitioner can then go back to the field and explore that parameter/s in more depth.
-%\item This facilitates understanding and the construction of more detailed models and data collection strategies
-%
-%
-%\end{itemize}
-%
-%% taken from Toms thesis. 
-%
-
-%In particular, PPSs allow probabilistic models to be represented in the form of a generative model through the
-%probabilistic programming language~(PPL), which enables one to write statements that enable conditioning on data~\cite{gordon2014probabilistic,Goodman08church:a}. 
-%Thus, it seems only natural to connect simulators with PPSs, as simulators explicitly define a generative model, in the language in which they are written, and the nature of the PPSs enables us to perform inference in these simulators, by conditioning on observations, which are fed in as input to the simulators.
-%In doing this we can infer things about stochastic input parameters and other variables sampled during the program's forward execution, whilst also providing full interpretability in the inference results, which is absolutely necessary in safety-critical domains.
-%By exploiting the tools that we develop in this paper, one is able to connect epidemiology simulators with \texttt{Pyprob}~\cite{le-2016-inference} to extract fully interpretable posterior structures allowing practitioners to interpret which input parameters have the largest affect on the epidemic, in our case malaria.
-Over five-hundred-thousand people die from malaria each year, mostly children under five years of age, with 90 per cent of malaria cases occurring in Sub-Saharan Africa.
-An estimated 100-300 million people suffer from malaria each year~\cite{world2016world}.
-%Thus, by understanding the prominence of input parameters to epidemiology simulators, guides decision making processes relating to effective treatment and prevention strategies. 
-%\label{sec:related}
-
- \section{Simulating Diseases}
- \label{sec:background}
-
-% Stochastic simulators are common across a number of domains, statistical physics~\cite{landau_binder_2014}, financial modeling~\cite{jackel2002monte},
-% weather prediction~\cite{evensen1994sequential}, epidemiology~\cite{smith2008towards} and many others. 
-% An advantage of using stochastic simulators is that they provide a level of interpretability not found in modern deep learning settings, as they directly incorporate model structure from carefully reasoned observations and experiments.
-In addition to this,
-
-In-silico simulators have become a crucial tool in evidence-based decision-making within a large number of disciplines, including statistical physics~\cite{landau_binder_2014}, financial modeling~\cite{jackel2002monte},
-weather prediction~\cite{evensen1994sequential}, epidemiology~\cite{smith2008towards} and many others.
-In many cases, simulation output can augment or even replace real data that may otherwise be costly or even impossible to generate.
-Recent advances in hardware have enabled simulations to model increasingly complex systems.
-Epidemiology studies the prevalence and spreading of diseases across populations.
-Recent advances in hardware have enabled simulations to model the dynamics of infectious diseases, such as Malaria, in ever greater detail.
-
-\subsection{Epidemiology Simulators}
-
-Two the most advanced Malaria simulators, namely EMOD~\cite{bershteyn2018implementation} and OpenMalaria  \cite{smith2008towards}, have proven to be particularly valuable to policymakers.
-OpenMalaria is based on microsimulations of Plasmodium falciparum in humans and was originally developed to simulate the impacts of malaria vaccines within simple villages or districts
-Compared with OpenMalaria, EMOD is able to simulate a variety of additional drivers, including complex geographies complete with migration and a large number of policy interventions
-Both EMOD and OpenMalaria are open source and implemented in C++.
-
-% Simulators provide users with the ability to model complex environments and problems in a black-box way, enabling new insights to be generated and predictions to be made.
-A simulator typically encodes several years, if not decades, of research and development and so by their very nature are structurally complicated.
-
-% Within the context of epidemiology simulators we can model the treatment, infection and prevention cycle under a variety of different environmental conditions, enabling predictions to be made on a set of field observations.
-Such scenarios in practices would be
-
-% This enables practitioners to model a wide array of different scenarios at ease.
-% However, since the development happens over several years, or decades, usually without comprehensive documentation, understanding a code base written with legacy libraries and software is incredibly difficult and an almost impossible task for a non-expert, or expert who has not had access to the previous years of development.
-%
-
-% Two prominent simulators for modeling Malaria are OpenMalaria~\cite{smith2008towards} and Emod~\cite{bershteyn2018implementation}, which are used by practitioners to model several aspects of the Malaria disease.
-
-%  \paragraph*{OpenMalaria} 
-% is an open-source simulator funded by the Gates Foundation to model different aspects of the malaria prevention and treatment cycle for varying population dynamics.
-%  Users specify models in OpenMalaria via a scenario file, in the form of an XML document, which determines the specifics of a particular treatment, or prevention strategy, climatic factors, the demographics of the population and so forth.
-In addition to this, field observations can be encoded inside of, or externally to the scenario file.
-% The scenario file encodes all input and output parameters, some of which we wish to simulate, others remain fixed.
-
-% \paragraph{EMOD} provides the modularity and structural flexibility necessary for modeling multiple infectious diseases at the within-host and population level. 
-% EMOD can be used to model HIV, Tuberculosis, Malaria and generic diseases, but for the purposes of the comparison, we just focus on Malaria. 
-% To specify models in EMOD a configuration file is written by the user, which is in the form of a JSON and provides similar functionality to OpenMalaria, in terms of what can be specified and the outputs that it produces.
-
-% In both these well used simulators users are accustomed to writing their models as either XML, or JSON files and so it is important that any system built around this framework enables the models to be inputted in the same format, without requiring the users to learn how to use a new set of protocols.
-Our system facilitates this as we shall discuss in the proceeding sections.
-
-\section{Hijacking Simulators}
-\label{sec:hijackingsimulators}
-
- Probabilistic programming~\cite{gordon2014probabilistic,staton2016semantics,kozen1979semantics} 
-can be used to express probabilistic models and consequently perform automated inference these.
-Once a probabilistic model has been expressed in a probabilistic programming language, a wide range of inference techniques, such as Markov Chain Monte Carlo (MCMC)\cite{}, Variational Inference~(VI)\cite{} and Deep Neural Networks~(DNN)\cite{},  can be used by non-experts in an automated fashion.
-
-\paragraph{Hijacking a simulator} describes the process by which a simulator's random number generators are replaced by calls to external sampling procedures, which are controlled by a probabilistic programming system (PPS).
-In practice, this amounts to performing a small number of surgical incisions into the simulator's source code in order to replace built-in calls to random number generators.
-E.g., given a simulator written in C++/Boost\cite{schaling2011boost}, a Gaussian distribution object \textit{boost::normal} is replaced with the corresponding \textit{pyprob\_cpp} distribution object, namely \textit{pyprob\_cpp::distributions::Normal}.
-Sampling from this distribution is then done by requesting the PPS to send a sample back to the simulator.
-
-The PPS and the simulator exchange xTensor objects \url{https://xtensor.readthedocs.io/en/latest/} through TCP or ICP using a generic FlatBuffers \url{http://google.github.io/flatbuffers/} protocol.
-On the PPS side, sampling is done using the deep learning framework PyTorch \cite{paszke2017automatic}.
-
-After a variable has been sampled, it is sent to the PPS using \textit{pyprob\_cpp::sample}, alongside with its simulator source trace.
-This allows the PPS to construct sample trace probabilities and other summary statistics (cf.
-Section \ref{sec:casestudy}).
-
-Finally, the entry point to the simulator, i.e. \textit{main} in C++, is replaced by a special \textit{forward} call, in this case \textit{pyprob\_cpp::forward}.
-This allows the PPS to generate rollouts from the simulator remotely.
-
-\paragraph{Population-based simulators} In contrast to the setup used by \cite{baydin2018efficient}, who interface a particle physics simulator that creates a single trace per \textit{forward} call, a population-based simulator creates a trace per population member.
-This means that e.g.
-in EMOD, a standard scenario simulation rollout over the span of $3$ years
-with a population of size $n=5000$ \cite{smith2008towards} will generate about $ \SI{4}{\tera\byte}$ of raw trace data.
-Unlike with \citep{baydin2018efficient}, this amount of data cannot possibly be kept in RAM, which makes \textit{a posteriori} trace analysis very inefficient.
-In order to deal with the shortcomings of \textit{pyprob} in a population-based simulator context, we therefore extend the framework to be able to do trace analysis on the fly.
-
-By extending \textit{pyprob} to handle population-based simulators, the PPS can now track all stochastic random variables that are created within the simulator, which then allows us to generate trace plots and path probabilities associated to the execution paths of the program.
-The corresponding increase of simulator transparency helps reinstate much needed trust between policymakers and evidence-based methods.
-
-%As stated previously, Simulators are complex systems and should be viewed as black-boxes, thus deciphering the underlying processes occurring within the simulator facilitates the users understanding of how such a system operates and provides a way of interpreting the outputs in a clear way.
-In addition to this, when using a simulator to make new inferences a user will often want to condition on the output of the simulator, or any new observations gathered.
-In the standard simulator framework this is not possible, but in
-%
-%Essentially, from the users perspective, our approach can be viewed as follows the user would first select the simulator that they are going to use, for example EMOD, they would then write their model, in the desired simulator, according to the given simulator specification.
-The user would then run the simulator.
-Now comes the hijacking.
-During the forward run of the simulator all sampling calls from the random number generator are directed through the probabilistic programming system~(PPS), which in addition to the standard output of the simulator now generates a series of result summaries and path probabilities, as we demonstrate in Section~\ref{sec:casestudy}.
-%
-%By connecting with the PPS we track all sampling procedures, which enables us to control the myriad of random variables generated within the simulator. 
-%
-%In order to overhaul a stochastic simulator like EMOD, or OpenMalaria, we are only required to override the location of the stochastic primitives, i.e.
-the base operations that generate the stochasticity and randomness within the simulator.
-%We will walk through how this is done for a standard epidemiology simulator, see the flowchart for a pictorial representation of what is happening \bg{add ref to flow chart}.
-%
-%The initial step first requires one to build a containerized environment, such as Docker~\cite{merkel2014docker} and Singularity~\cite{kurtzer2017singularity}, which enables the simulators to be run on any device, independent of different hardware architectures, removing the requirement for hardware specific machines.
-%The second step requires one to override the stochastic primitives, which requires re-writing aspects of the sampling procedures inside the simulator. 
-%This requires zero domain-expertise and only requires one to find where the sampling calls are originated, this can be done with any basic editor search function.
-% Such information is typically saved in $\texttt{RANDOM.<filetype>}$, although it may vary slightly between simulators. We now demonstrate this for a uniform primitive.
-%
-%The third step is generating the protocols, which enable communication between the simulator and probabilistic programming system. 
-%For this we extend the existing bridging framework introduced in~\cite{baydin2018efficient}, which is comprised of the Probabilistic Programming eXecution~(PPX) protocols and a light-weight C ++ interface, however the language of this light-weight interface is dependent on whatever the file type of the $\texttt{RANDOM.<filetype>}$ file is.
-%For both OpenMalaria and EMOD, the file is written in C ++.
-%In using these protocols a larger class of sampling procedures can be overridden.
-The PPX protocols work by making use of the flatbuffer protocols\footnote{\url{https://github.com/google/flatbuffers}}, which require a flatbuffers script to be specified, see here for an example of a flatbuffers script\footnote{https://github.com/probprog/ppx/blob/master/ppx.fbs}.
-Once this is specified the flat buffers script is compiled and automatically constructs the message passing framework between the simulator and the probabilistic programming system.
-The flatbuffers script is simple, clean and intuitive, making it easy for a non-expert to produced.
-%The next steps requires modifying the $\texttt{main}$ function in $\texttt{main.cpp}$ and defining an additional forward function which runs one execution of the simulator to enable message passing between the simulator and PPS\footnote{A link to an example will be provided after the review period to preserve anonymity}.
-%The final step is to create a light-weight interface.
-As both the EMOD and OpenMalaria are solely based on C ++ we can extend an existing interface called pyprob\_cpp\footnote{\url{https://github.com/probprog/pyprob_cpp}} to account for the additional sampling procedures.
-%Once these steps are complete, the containerized environment will now contain everything required to pass models through to the simulators, encode the simulator into a probabilistic program, which can then be evaluated in the PPS.
-
- %Our method of overriding simulators builds on the work of~\cite{baydin2018efficient} and extends the framework to encapsulate a more diverse range of simulators, in particular population based simulators such as those found in epidemiology.
-%Our framework provides a simple solution to do this and makes it easy to transform arbitrary stochastic simulators into probabilistic programs, regardless of the complexity of the simulator and the language that the simulator is written in.
-%In particular, the framework currently supports 13 popular languages, but there is nothing to stop this being extended to any language of interest. 
-%Furthermore, we can understand the structure of simulators as programs for both decoding \emph{black-box structures} and for posterior inference, which enables two things. 
-%1) It enables software developers and policy makers to understand complicated code bases and dissect the outputs of simulators for guiding policy decision making.
- %2) It provides interpretable inference results that provide policy makers with predictions that are truly interpretable, in the sense that the end-user of the inference results understands what physical events led to the inference outcome.
- %This is critical in several domains, especially in epidemiology. 
- %We provide examples of (1) in next Section~\ref{sec:casestudy} and discuss the latter in future work as extensions to the the current framework.
-
-\begin{figure}
-\centering
- \scalebox{.5}{
-\begin{tikzpicture}[node distance=1.5cm,
-    every node/.style={fill=white, font=\sffamily}, align=center]
-  % Specification of nodes (position, etc.)
-  % \node (start)             [activityStarts]              {Activity starts};
-  \node (SimulatorSystem)     [activityStarts]          {Simulator System \\ e.g. EMOD};
-   \node (randomprim)      [process, left of=SimulatorSystem, xshift=5.6cm]
-                                                      {Redefine RNG \\ to make \\ externally \\ callable};
-  \node (onStartBlock)      [process, below of=SimulatorSystem]   {Specify Model \\ e.g. XML files};
-  \node (activityRuns)      [activityRuns, below of=onStartBlock]
-                                                      {Run Simulator};
-  \node (onPauseBlock)      [startstop, below of=activityRuns, yshift=-1cm]
-                                                                {Posterior Output};
-  % \node (onStopBlock)       [process, below of=onPauseBlock, yshift=-1cm]
-  %                                                                {};
-  % \node (onDestroyBlock)    [process, below of=onStopBlock, yshift=-1cm] 
-  %                                                             {onDestroy()};
-  \node (onRestartBlock)    [process, right of=activityRuns, xshift=3cm]
-                                                              {PPS hijacks \\ random number \\ generator};
-  \node (ActivityEnds)      [startstop, below of=onRestartBlock, yshift=-1cm]
-                                                        {Result summaries \\ path probabilities};
-  % \node (ActivityDestroyed) [startstop, below of=onDestroyBlock]
-  %                                                   {Activity is shut down};     
-  % Specification of lines between nodes specified above
-  % with aditional nodes for description 
-  % \draw[->]             (start) -- (onCreateBlock);
-  \draw[->]     (SimulatorSystem) -- (onStartBlock);
-  \draw[dashed, <-]  (SimulatorSystem) -- (randomprim);
-  % \draw[->]      (onStartBlock) -- (onResumeBlock);
-  \draw[->]     (onStartBlock) -- (activityRuns);
-  \draw[->]      (activityRuns) -- (onPauseBlock);
-  % \draw[->]      (onPauseBlock) -- node {The activity is no longer visible}
-  %                                  (onStopBlock);
-  % \draw[->]       (onStopBlock) -- node {The activity is shut down by
-  %                                  user or system} (onDestroyBlock);
-  % \draw[->]    (onRestartBlock) -- (activityRuns);
-  \path[->, shift left=.75ex]
-  (onRestartBlock) edge (activityRuns)
-  (activityRuns) edge (onRestartBlock);
-  % \draw[<-]    (activityRuns) -- (onRestartBlock);
-  % \draw[->]       (onStopBlock) -| node[yshift=1.25cm, text width=3cm]
-  %                                  {The activity comes to the foreground}
-  %                                  (onRestartBlock);
-
-  \draw[->]    (onRestartBlock) -- (ActivityEnds);
-  % \draw[->]      (onPauseBlock) -| node(priorityXMemory)
-  %                                  {todo}
-  %                                  (ActivityEnds);
-  % \draw           (onStopBlock) -| (priorityXMemory);
-  % \draw[->]     (ActivityEnds)  |- node [yshift=-2cm, text width=3.1cm]
-  %                                   {User navigates back to the activity}
-  %                                   (SimulatorSystem);
-  % \draw[->] (onPauseBlock.east) -- ++(2.6,0) -- ++(0,2) -- ++(0,2) --                
-  %    node[xshift=1.2cm,yshift=-1.5cm, text width=2.5cm]
-  %    {The activity comes to the foreground}(onResumeBlock.east);
-  \end{tikzpicture}
-  }
-  \label{fig:how}
-  \caption{This flow-chart provides an overview of the process of how our system hijacks a generic population-based epidemiology simulator, such as EMOD, and how we modify the simulator to 
-  to hijack the random number generator (RNG).}
-  \end{figure}
-
-% \begin{table}[h!]
-% \footnotesize
-% \begin{tabular}{p{3.6cm}|p{3.6cm}}
-% Before  &  After  \\
-% \midrule
-% \begin{lstlisting}
-% #include <gsl/gsl_cdf.h>
-% #include <gsl/gsl_rng.h>
-% #include <gsl/gsl_randist.h>
-% //preamble
-% double random::uniform_01 () {
-%     double result =
-% # ifdef 
-% OM_RANDOM_USE_BOOST
-%        rng_uniform01 ();
-% # else
-%  gsl_rng_uniform (rng.gsl_generator);
-% # endif
-%     return result;
-%     }
-% \end{lstlisting}&
-% \begin{lstlisting}
-% #include <pyprob_cpp.h>
-% #include "xtensor/xadapt.hpp"
-% // preamble
-% double random::uniform_01 () {
-% auto uniform = pyprob_cpp::distributions::Uniform(0,1);
-% return pyprob_cpp::sample(uniform)(0);
-% }
-% \end{lstlisting}
-% \end{tabular} 
-% \end{table}
-% As can be seen, writing the new sampling call is cleaner and syntactically simpler. 
-%\bg{Add performance plot showing memory consumptions of pyprob and our new system.}
-
-\section{Case Study: Ifakara (Tanzania)}
-\label{sec:casestudy}
-
-Ensemble methods are commonly used in statistics in order to combine the predictive power of multiple models \cite{cameron2015defining,smith_ensemble_2012}.
-To this end, recent work has attempted to characterise the similarities and differences between two of the most advanced Malaria epidemiology simulators, EMOD \cite{bershteyn2018implementation} and OpenMalaria \cite{smith2008towards}.
-Evaluation is usually done by comparing a number of output parameters across a range of hand-crafted standard scenarios reflecting different geographical locations across Africa~\cite{smith_ensemble_2012}.
-
-In the following, we illustrate how our method introduces a novel introspection paradigm.
-By extracting trace graphs from population-based simulators, policymakers can ask specific questions about properties of the model trace flow.
-
-To illustrate the above, we present simulation output generated from a scenario resembling local conditions in the town of Ifakara (Tanzania).
-Both EMOD and OpenMalaria are configured to simulate a single population node of $n=100$ and assume constant climatic conditions and no migration over the simulation period of $3$ years.
-Please refer to Figure \ref{fig:EIR} for the seasonal Entomological Inoculation Rate (EIR), a measure of infectivity.
-
-\begin{figure}[h!]
-	\centering
-	\includegraphics[width=\textwidth/2]{../plots/EIR_Ifakara.pdf}
-	\caption{Seasonal Entomological Inoculation Rate (EIR) for the Ifakara scenario. Data is averaged over $30$ day periods.}
-	\label{fig:EIR}
-\end{figure}
-
-We provide examples of the generated trace plots from the connection between the simulator and the PPS in Figure~\ref{fig:plotewan}.
-
-We can see from the addressing schemes $A_1, \ldots, A_N$ (Tables \ref{table:addresses} and \ref{table:asingleaddress}) what physical events are connected to each other and how outputs in EMOD are generated from a different set of procedures as compared to OpenMalaria
-By having access to such diagrams users can internally evaluate and scrutinize the decisions that the simulator is making.
-
-This is important for policymakers, or general non-experts, as it not only details how we arrive at the given outputs, but it provides an understanding of which processes were most crucial in determining those outputs as can be seen from the path probabilities assigned to each of the vertices.
-
-Additionally, by associating nodes in trace graphs representing the same physical processes within different models, the significance of model detail can be evaluated.
-For example, the full trace presented in Table \ref{table:asingleaddress} represent a sampling step associated with within-host dynamics of the Malaria parasite \textit{Falciparum} in OpenMalaria node $A1$.
-The same physical process also occurs in EMOD's trace graph at position $A7$ (see Section \ref{table:addresses}).
-Comparisons like these could help developers better control model complexity, and even provide an alternative testing and debugging paradigm.
-
-% \begin{tikzpicture}
-% \begin{axis}[xlabel={Population size},ylabel={memory consumption (MB)},
-% legend style={at={(0.5,-0.15)},xmin=0,xmax=700,ymin=0,ymax=70, axis lines=center]
-
-% \addplot[pyprob,domain=0:1000, color=blue,]{x};
-
-% \addplot[improved,domain=0:1000, color=red,]{10};
-
-% \legend{pyprob, improved}
-% \end{axis}
-% \end{tikzpicture}
-
-%\begin{figure}
-%\begin{tikzpicture}
-%\begin{axis}[
-%    ybar,
-%    enlargelimits=0.15,
-%    legend style={at={(0.5,-0.15)},
-%      anchor=north,legend columns=-1},
-%      xlabel = {Address},
-%    ylabel={Probabilities},
-%    symbolic x coords={Trace 1, Trace 2, Trace 3},
-%    xtick=data,
-%    nodes near coords,
-%    nodes near coords align={vertical},
-%    ]
-%% \addplot coordinates {(tool8,7) (tool9,9) (tool10,4)};
-%\addplot coordinates {(Trace 1,0.2) (Trace 2,0.5) (Trace 3,0.23)};
-%% \addplot coordinates {(A1,0.2) (A2,0.5) (A3,0.7) (A4,0.9)};
-%% \addplot coordinates {(A1,0.1) (A2,0.1) (A3,0.1) (A4,0.3)};
-%% \legend{used,understood,not understood}
-%\end{axis}
-%\end{tikzpicture}
-%\caption{Trace probabilities. Trace 1 corresponds to A1,A1,A4,... 
-%Trace 2 corresponds to A1,A1,A1,...A2,... These still need correctly calculating.}
-%\end{figure}
-
-\begin{figure*}[h!]
-	\centering
-	\includegraphics[width=\textwidth]{../plots/final_emod_traces__rotated.pdf}
-	\includegraphics[width=\textwidth]{../plots/ewan_25_pop_100.pdf}
-	\caption{Here we run two equivalent models, compare the corresponding trace paths and corresponding 
-	path probabilities taken by the thousands of random variables generated internally within the simulators.
-\textit{\textbf{Top:}} The specified model run in EMOD.
-\textit{\textbf{Bottom:}} The specified model in OpenMalaria.}
-	\label{fig:plotewan}
-\end{figure*}
-
-\begin{table}[h!]
-\footnotesize
-  \setlength{\tabcolsep}{1mm}
-\label{table:asingleaddress}
-  \caption{An example of an address generated for the model run in the OpenMalaria simulator. We can see that A1
-  relates to Generating a member of the human population who may or may not be infect with the Malaria disease.
-We get something similar for EMOD, except this relates to A7 in the EMOD program execution.}
-  \def\arraystretch{1.25}
-  % \setlength{\tabcolsep}{1mm}
-  \begin{tabularx}{0.46\textwidth}{@{}lX@{}l@{}} 
-    \toprule
-    Address ID & Full address \\
-    \midrule
-  A1 & [forward()+0x204; OM::Simulator::
-
-  start(scnXml::Monitoring const)+0x28a;
-
-  OM::Population::createInitialHumans()+0x94;
-
-  OM::Population::newHuman(OM::SimTime)+0x5c;
-
-  OM::Host::Human::Human(OM::SimTime)+0x12b;
-
-  OM::WithinHost::WHInterface::
-
-  createWithinHostModel(double)+0x99;
-
-  OM::WithinHost::DescriptiveWithinHostModel
-  ::DescriptiveWithinHostModel(double)+0x3a;
-
-  OM::WithinHost::WHFalciparum::
-
-  WHFalciparum(double)+0xe6;
-
-  OM::util::random::
-
-  gauss(double, double)+0xb4]\_\_Normal \\
-
-  $\vdots$ & $\vdots$ \\
-
- %   $\vdots$ & $\vdots$ \\
- %  A5 & [forward()+0x204; OM::Simulator::start(scnXml::Monitoring const\&)+0x468;
-
- %   OM::Population::update1(OM::SimTime)+0xff;
-
- %    OM::Host::Human::update(bool)+0x2bc;
-
- %    OM::Clinical::ClinicalModel::update(OM::Host::Human\&, double, bool)+0x96;
-
- %    OM::Host::NeonatalMortality::eventNeonatalMortality()+0x9;
-
- %    OM::util::random::uniform\_01()+0xc0]\_\_Uniform\\
-
- % $\vdots$ & $\vdots$ \\
-
-\bottomrule
-  \end{tabularx}
-  \end{table}
-
-\begin{table}[h!]
-  \footnotesize
-  \setlength{\tabcolsep}{1mm}
-  \caption{An interpretation table for each of the address of the 
-  overall trace generated from the corresponding OpenMalaria model.}
-  \label{table:addresses}
-  % \def\arraystretch{1.25}
-  % \setlength{\tabcolsep}{1mm}
-  \begin{tabularx}{0.46\textwidth}{@{}lX@{}} 
-    \toprule
-    Address ID & Interpretation \\
-    \midrule
-  A1 & Generate a human in the
-
-   population within host dynamics\\
-
-  A2 & Generate another human in
-
-   the population within host dynamics\\
-
-  A3 & The population is updated and
-
-  a new human, or humans, may get infected \\
-
-  A4, A5 & Potential child deaths
-
-  within the population are simulated \\
-
-  A6 & Determines parasite density
-
-   of an individual infection \\
-
-  A7 & Models how the disease is
-
-  progressing within the infected humans \\
-
-	A8 & Models how the disease is
-
-	 progressing within the population \\
-
-  A9 & Models how the disease is
-
-  progressing within the infected humans
-
-   after the population has been updated\\
-
-  A10 & Full clinical update on
-
-  the population for those without severe
-
-  or no Malaria infection. \\
-
-A11 & Full clinical update on the population
-
-for those with severe Malaria infections \\
-\bottomrule
-  \end{tabularx}
-  \end{table}
-
-\section{Discussions and Future Work}
-
-In this work we have demonstrated a method 
-that enables one to hijack populationbased simulators, extending 
-the work of \citep{baydin2018efficient}. 
-We applied our method to two Malaria orientated population-based simulators and 
-generated a variety of trace graphs.
-Finally, we have shown how our system
-enables policy makers and non-experts to analyse simulator outputs in a 
-way previously unavailable in the field of epidemiology.
-
-To extend our work further we aim to implement additional 
-tools that will facilitate complicated inference procedures that condition on simulator 
-output. We will also evaluate additional scenarios across Africa and South-East Asia to better understand the similarities and differences between EMOD and OpenMalaria.
-
-\quad\\
-\quad\\
-
-\bibliographystyle{icml2019}
-\bibliography{malaria,refs}
-\end{document}
-
-% This document was modified from the file originally made available by
-% Pat Langley and Andrea Danyluk for ICML-2K. This version was created
-% by Iain Murray in 2018, and modified by Alexandre Bouchard in
-% 2019. Previous contributors include Dan Roy, Lise Getoor and Tobias
-% Scheffer, which was slightly modified from the 2010 version by
-% Thorsten Joachims & Johannes Fuernkranz, slightly modified from the
-% 2009 version by Kiri Wagstaff and Sam Roweis's 2008 version, which is
-% slightly modified from Prasad Tadepalli's 2007 version which is a
-% lightly changed version of the previous year's version by Andrew
-% Moore, which was in turn edited from those of Kristian Kersting and
-% Codrina Lauth. Alex Smola contributed to the algorithmic style files.
+This reinstates much-needed trust between policymakers and evidence-based methods.

### 2019-05-27 paper/icml_ai4good/main.tex
minor

@@
 
-\paragraph{Population-based simulators} In contrast to the setup used by \cite{baydin2018efficient}, who interface a particle physics simulator that creates a single trace per \textit{forward} call, a population-based simulator creates a trace per population member.
+\paragraph{Population-based simulators} create a trace per population member, in contrast to event-based simulators that create only a single trace per \textit{forward} call~\cite{baydin2018efficient}.
 This means that e.g.
-in EMOD, a standard scenario simulation rollout over the span of $3$ years
-with a population of size $n=5000$ \cite{smith2008towards} will generate about $ \SI{4}{\tera\byte}$ of raw trace data.
+in OpenMalaria, a standard scenario simulation rollout over the span of $3$ years
+with a population of size $n=5000$ \cite{smith2008towards} will generate about \textit{four terabytes} of raw trace data.
 Unlike with \citep{baydin2018efficient}, this amount of data cannot possibly be kept in RAM, which makes \textit{a posteriori} trace analysis very inefficient.

### 2019-07-04 paper/icml_ai4good/paper.tex
added tom notes

--- 2019-05-27 paper/icml_ai4good/main.tex
@@
 \begin{icmlauthorlist}
-\icmlauthor{Bradley.~J Gram-Hansen}{equal,en}
-\icmlauthor{Christian Schr\"{o}der de Witt}{equal,en}
+\icmlauthor{Bradley~J. Gram-Hansen}{equal,en}
+\icmlauthor{Christian Schr\"{o}der de Witt}{equal,en}\\
 \icmlauthor{Tom Rainforth}{st}
-\icmlauthor{Phil H.S. Torr}{en}
+\icmlauthor{Philip H.S. Torr}{en}
 \icmlauthor{Yee Whye Teh}{st}
@@
 
-\icmlcorrespondingauthor{Bradley.~J Gram-Hansen}{bradley@robots.ox.ac.uk}
+\icmlcorrespondingauthor{Bradley~J. Gram-Hansen}{bradley@robots.ox.ac.uk}
 
-\icmlaffiliation{en}{Department of Engineering, University of Oxford, UK}
+\icmlaffiliation{en}{Department of Engineering Science, University of Oxford, UK}
 \icmlaffiliation{st}{Department of Statistics, University of Oxford, UK}
@@
 
-To reach the WHO’s target of reducing malaria incidence and mortality rates by at least 90\% by 2030, policymakers are increasingly turning to evidence-based methods, thus oftentimes relying on computational simulations \cite{who_global_2015}.
+To reach the WHO's target of reducing malaria incidence and mortality rates by at least 90\% by 2030, policymakers are increasingly turning to evidence-based methods, thus oftentimes relying on computational simulations \cite{who_global_2015}.
 These simulations allow policymakers to infer critical information on disease dynamics and make predictions about the impacts of policies
@@
 In this paper, we introduce a novel method that allows one to shed light on the inner workings of a large class of population-based stochastic simulators.
-We achieve this by extending the work of \citep{baydin2018efficient} by interpreting such population-based simulators as probabilistic generative models within the framework of universal probabilistic programming (UPP) \cite{le-2016-inference}.
+We achieve this by extending the work of \citet{baydin2018efficient} by interpreting such population-based simulators as probabilistic generative models within the framework of universal probabilistic programming (UPP) \cite{le-2016-inference}.
 To this end, we \emph{hijack} existing simulators by overriding their internal random number generators.
@@
 
-Our framework already supports application to simulators written in 13 industrial programming frameworks, and is easily extensible.
+Our framework already supports application to simulators written in 13 general-purpose programming languages, and is easily extensible.
 This is crucial as, given the enormous code size and complexity, rewriting epidemiology simulators using a dedicated universal probabilistic programming language, such as Pyro \cite{bingham2019pyro}, is often infeasible.
@@
 This paper first gives an overview of existing malaria simulators (Section \ref{sec:background}), and proceed by introducing the necessary background on the \textit{pyprob} framework and the concept of universal probabilistic programming (Section \ref{sec:hijackingsimulators}).
-Our approach is then demonstrated and analysed in the context of a Malaria case study (Section \ref{sec:casestudy}).
+Our approach is then demonstrated and analysed in the context of a malaria case study (Section \ref{sec:casestudy}).
 
@@
 Two the most advanced malaria simulators, namely EMOD~\cite{bershteyn2018implementation} and OpenMalaria  \cite{smith2008towards}, have proven to be particularly valuable to policymakers.
-OpenMalaria is based on microsimulations of \textit{Plasmodium falciparum} in humans and was originally developed to simulate the impacts of malaria vaccines within simple villages or districts
-Compared with OpenMalaria, EMOD is able to simulate a variety of additional drivers, including complex geographies complete with migration and a large number of policy interventions
+OpenMalaria is based on microsimulations of \textit{Plasmodium falciparum} in humans and was originally developed to simulate the impacts of malaria vaccines within simple villages or districts.
+Compared with OpenMalaria, EMOD is able to simulate a variety of additional drivers, including complex geographies complete with migration and a large number of policy interventions.
 Both EMOD and OpenMalaria are open source and implemented in C++.
@@
  Probabilistic programming~\cite{gordon2014probabilistic,staton2016semantics,kozen1979semantics} 
-can be used to express probabilistic models and consequently perform automated inference these.
-Once a probabilistic model has been expressed in a probabilistic programming language, a wide range of inference techniques, such as Markov Chain Monte Carlo (MCMC)\cite{geyer1992practical}, Black-box Variational Inference~(VI)\cite{ranganath2014black} and amortized inference\cite{le2016inference},  can be used by non-experts in an automated fashion.
+can be used to express probabilistic models and consequently perform automated inference in these.
+Once a probabilistic model has been expressed in a probabilistic programming language, a wide range of inference techniques, such as Markov chain Monte Carlo (MCMC) \cite{geyer1992practical}, black-box variational inference~(VI) \cite{ranganath2014black} and amortized inference\cite{le2016inference}, can be used by non-experts in an automated fashion.
 
@@
 In practice, this amounts to performing a small number of surgical incisions into the simulator's source code in order to replace built-in calls to random number generators.
-E.g., given a simulator written in C++/Boost\cite{schaling2011boost}, a Gaussian distribution object \textit{boost::normal} is replaced with the corresponding \textit{pyprob\_cpp} distribution object, namely \textit{pyprob\_cpp::distributions::Normal}.
+E.g., given a simulator written in C++/Boost \cite{schaling2011boost}, a Gaussian distribution object \textit{boost::normal} is replaced with the corresponding \textit{pyprob\_cpp} distribution object, namely \textit{pyprob\_cpp::distributions::Normal}.
 Sampling from this distribution is then done by requesting the PPS to send a sample back to the simulator.
 
-The PPS and the simulator exchange xTensor objects \url{https://xtensor.readthedocs.io/en/latest/} through TCP or ICP using a generic FlatBuffers \url{http://google.github.io/flatbuffers/} protocol.
+The PPS and the simulator exchange xTensor objects\footnote{\url{https://xtensor.readthedocs.io/en/latest/}} through TCP or IPC using a generic FlatBuffers\footnote{\url{http://google.github.io/flatbuffers/}} protocol.
 On the PPS side, sampling is done using the deep learning framework PyTorch \cite{paszke2017automatic}.
 
-After a variable has been sampled, it is sent to the PPS using \textit{pyprob\_cpp::sample}, alongside with its simulator source trace.
+After a variable has been sampled, using \textit{pyprob\_cpp::sample}, it is sent by the PPS to the simulator, at the same recording the simulator execution trace as a side effect.
 This allows the PPS to construct sample trace probabilities and other summary statistics (cf.
@@
 
-Finally, the entry point to the simulator, i.e. \textit{main} in C++, is replaced by a special \textit{forward} call, in this case \textit{pyprob\_cpp::forward}.
+Finally, the entry point to the simulator, i.e., \textit{main} in C++, is replaced by a special \textit{forward} call, in this case \textit{pyprob\_cpp::forward}.
 This allows the PPS to generate rollouts from the simulator remotely. A pictorial overview of this process is depicted in Figure~\ref{fig:how}.
@@
 \paragraph{Population-based simulators} create a trace per population member, in contrast to event-based simulators that create only a single trace per \textit{forward} call~\cite{baydin2018efficient}.
-This means that e.g.
-in OpenMalaria, a standard scenario simulation rollout over the span of $3$ years
+This means that, e.g., in OpenMalaria, a standard scenario simulation rollout over the span of $3$ years
 with a population of size $n=5000$ \cite{smith2008towards} will generate about \textit{four terabytes} of raw trace data.
-Unlike with \citep{baydin2018efficient}, this amount of data cannot possibly be kept in RAM, which makes \textit{a posteriori} trace analysis very inefficient.
+Unlike the simulator used by \citet{baydin2018efficient}, this amount of data cannot feasibly be kept in RAM, which makes \textit{a posteriori} trace analysis very inefficient.
 In order to deal with the shortcomings of \textit{pyprob} in a population-based simulator context, we therefore extend the framework to be able to do trace analysis on the fly.
@@
 
-\section{Case Study: Ifakara (Tanzania)}
+\section{Case Study: Ifakara, Tanzania}
 \label{sec:casestudy}
@@
 By extracting trace graphs from population-based simulators, policymakers can ask specific questions about properties of the model trace flow
-and not just the outcomes of the model, thus providing additional interpretability to the decision making process.
+and not just the outcomes of the model, thus providing additional interpretability to the decision-making process.
 
-To illustrate the above, we present simulation output generated from a scenario resembling local conditions in the town of Ifakara (Tanzania).
-Both EMOD and OpenMalaria are configured to simulate a single population node of $n=100$ and assume constant climatic conditions and no migration over the simulation period of $3$ years.
+To illustrate the above, we present simulation output generated from a scenario resembling local conditions in the town of Ifakara, Tanzania.
+Both EMOD and OpenMalaria are configured to simulate a single population node of $n=100$ and assume constant climatic conditions and no migration over the simulation period of three years.
 Please refer to Figure \ref{fig:EIR} for the seasonal Entomological Inoculation Rate (EIR), a measure of infectivity.
 
-\begin{figure}[h!]
+\begin{figure}
 	\centering
@@
 
-We can see from the addressing schemes $A_1, \ldots, A_N$ (Tables \ref{table:addresses} and \ref{table:asingleaddress}) what physical events are connected to each other and how outputs in EMOD are generated from a different set of procedures as compared to OpenMalaria
-By having access to such diagrams users can internally evaluate and scrutinize the decisions that the simulator is making.
+We can see from the addressing schemes $A_1, \ldots, A_N$ (Tables \ref{table:addresses} and \ref{table:asingleaddress}) what physical events are connected to each other and how outputs in EMOD are generated from a different set of procedures as compared to OpenMalaria.
+By having access to such diagrams, users can internally evaluate and scrutinize the decisions that the simulator is making.
 
@@
 	\includegraphics[width=\textwidth]{../plots/final_emod_traces__rotated.pdf}
-	\includegraphics[width=\textwidth]{../plots/ewan_25_pop_100.pdf}
+	\includegraphics[width=0.89\textwidth]{../plots/ewan_25_pop_100.pdf}
 	\caption{Here we run two equivalent models, compare the corresponding trace paths and corresponding 
@@
 that enables one to hijack population-based simulators, extending 
-the work of \citep{baydin2018efficient}. 
-We applied our method to two malaria orientated population-based simulators and 
+the work of \citet{baydin2018efficient}. 
+We applied our method to two malaria-orientated population-based simulators and 
 generated a variety of trace graphs.
@@
 CSW  is supported by the project Free the Drones (FreeD) under the Innovation Fund Denmark and Microsoft.
-YWT’s and TR’s research leading to these results has received funding from the European Research Council under the European Union’s Seventh Framework Programme (FP7/2007-2013) ERC grant agreement no.
+YWT's and TR's research leading to these results has received funding from the European Research Council under the European Union's Seventh Framework Programme (FP7/2007-2013) ERC grant agreement no.
 617071.

### 2019-07-04 paper/learning/notes.tex
added tom notes

--- 2019-05-22 paper/neurlps2019/main.tex
@@
 \usepackage{pgfplots}
-\usepackage{tikz}
+\usepackage{tikz,pgf}
 \usepackage{todonotes}
@@
 
+\usetikzlibrary{arrows,automata}
+\usetikzlibrary{positioning}
 \newcommand{\n}[1]{\ensuremath{n_{#1}}}
@@
 %\usepackage[accepted]{icml2019}
-\usetikzlibrary{arrows,chains,matrix,positioning,scopes, arrows.meta}
-\tikzset{
-  shift left/.style ={commutative diagrams/shift left={#1}},
-  shift right/.style={commutative diagrams/shift right={#1}}
-}
-\tikzset{%
-  >={Latex[width=2mm,length=2mm]},
-  % Specifications for style of nodes:
-            base/.style = {rectangle, rounded corners, draw=black,
-                           minimum width=4cm, minimum height=1cm,
-                           text centered, font=\sffamily},
-  activityStarts/.style = {base, fill=blue!50},
-       startstop/.style = {base, fill=red!50},
-    activityRuns/.style = {base, fill=green!50},
-         process/.style = {base, minimum width=2.5cm, fill=orange!55,
-                           font=\ttfamily},
-}
-\makeatletter
-\tikzset{join/.code=\tikzset{after node path={%
-\ifx\tikzchainprevious\pgfutil@empty\else(\tikzchainprevious)%
-edge[every join]#1(\tikzchaincurrent)\fi}}}
-\makeatother
-%
-\tikzset{>=stealth',every on chain/.append style={join},
-         every join/.style={->}}
-\tikzstyle{labeled}=[execute at begin node=$\scriptstyle,
-   execute at end node=$]
 
@@
 \author{%
-Paper ID: }
+Paper ID: 1010 }
   % examples of more authors
@@
 
-% A goal of probabilistic programming is to couple simulators, with inference. This is 
-% because stochastic simulators are used prominently in many industrial settings,
-% do not require one to construct hand-crafted joint distributions as they implicitly 
-% define a joint distribution of the program and encode learnt structures 
-% directly. This makes simulators powerful tools and much of machine learning (ML) and 
-% Artificial Intelligence (AI)
-% can be seen as trying to emulate such simulators from a purely data-driven approach.
-% However, in the 
-% ML/AI setting, although we can often infer outcomes, we have little understanding about what 
-% in the data led to the outputted inferences. 
-% This makes it challenging to deploy ML/AI systems into the wild, especially in health-related and safety-critical domains, such
-% as epidemiology, as we lose \emph{interpretability}. 
-% In this work, we explain how to design ML/AI systems that combine
-% probabilistic programming systems (PPSs) and epidemiology simulators, to extract
-% fully interpretable posterior structures, enabling policy makers 
-% and practitioners to make interpretable inferences. 
-% In particular, we demonstrate this for the Malaria disease and show how we can 
-% perform interpretable inference in such settings.
-
-Probabilistic programming provides a way to perform statistical inference over simulations of events in a programmatic way.
-Thus, by design, simulators are ideal programs for probabilistic programming systems~(PPSs).
-Simulators arise in a number of industrial scientific and industrial domains and encode sophisticated generative models.
-However, within existing PPSs one would have to re-implement the simulator via the PPS language specification, which is inefficient and often not feasible due to the complexity of such scientific and industrial simulators.
-Recent work by \cite{baydin2018efficient} demonstrated a pathway to turn a particular
-type of event-based simulator into a probabilistic program, without having to re-implement the simulator in the existing PPS. 
-% via the raw random number calls of the simulator.
- However, this still meant that a large class of critically important population-based simulators could not be turned into probabilistic programs and as such could not be used within a probabilistic programming framework.
-In this work, we extend that framework to encompass population-based simulators, a very large class of simulators that are used extensively across epidemiology, multi-agent and financial modeling and provide the relevant
-theoretical underpinnings. 
-We enable simulators written in 13 different languages to be easily converted to probabilistic programs, for which we algorithmically outline and provide comprehensive examples of our system applied to a number of real-world population-based simulators.
-
-% \todo[inline]{We need to have probabilistic programming in the title to make sure we attract the correct reviewers}
-
-% Epidemiology simulations have become a fundamental tool in the fight
-% against the epidemics of various infectious diseases like AIDS and
-% malaria.
-% However, the complicated and stochastic nature of these simulators
-% can mean their output is difficult to interpret, which reduces their usefulness to policymakers.
-% In this paper, we introduce an approach that allows one to treat a large class of population-based epidemiology simulators as probabilistic generative models.
-%  This is achieved by \emph{hijacking} the internal random number generator calls, through the use of an universal probabilistic programming system (PPS). 
-%  In contrast to other methods, our approach can be easily retrofitted to simulators written in popular industrial programming frameworks. 
-% We demonstrate that our method can be used for interpretable introspection and inference, thus shedding light on black-box simulators. 
-% This reinstates much needed trust between policymakers and evidence-based methods. 
+    Notes on how to develop a formalism to perform inference in population based simulators.
 \end{abstract}
@@
 \label{sec:intro}
-% \bg{The audience is a probabilistic programming audience - so that is what we will target}
 
-% \begin{itemize}
+One challenge in performing inference in such models is the combinatorial increase in traces, \emph{paths}
+that can be taken within the given probabilistic model. This is because each member of the population has 
+a bounded, but large number of decisions to make. In addition to this, each member of the population can interact
+with other members of the population. To model a single population member particle-based methods such as 
+sequential monte carlo (SMC) provide an ability to explore multiple paths
 
-% \item Talk about a goal of probabilistic programming is to simultaneously simulate and perform inference on models conditioned on the observations.
-Although universal probabilistic programming systems have the ability to do this, they are constrained because the simulator has to be re-written as a program inside that system; this is both infeasible for sufficiently large simulators and inefficient.
-It would however be much more efficient to turn simulators into probabilistic programs...
-then talk briefly about the foundations created by baydin et al.
-(2 paragraphs)
-% \item Talk about the large body of simulators out there and briefly talk about the types of simulators
-% i.e population-based, event-based, single-object based and so fourth. {1 paragraph}
-% \item Talk, high level, about the notion of a posterior in terms of traces either add stuff from tom's thesis or directly cite. (1 paragraph)
-% \item Talk about the experiments we perform (high- level)
+As updates are made sequentially MCMC methods typically are not-well suited this type of recusrive
+estimation problem. As each time we get a new data point $y_{t+1}$ we have to run new MCMC simulations 
+for $P(x_{0:t_1}| y_{0:t+1})$, we also cannot directly reuse the samples $\{x^{i}_{0:t}\}$. 
+Likewise, importance sampling is not well suited either, as we cannot reuse the samples and weights for a time $t$ $\{x^{i}_{0:t}, \tilde{w}^{i}_{t}\}^{N}_{i=1}$ 
+to sample from $(x_{0:t_1}| y_{0:t+1})$.
 
-% \end{itemize}
+\input{thesystem.tex}
+\section{Limitations with different inference engines}
 
-% One powerful \emph{class} of models are simulators. 
-Simulators are useful tools when conducting experiments in the real system 
-is impossible, or impractical and they arise in numerous industrial \cite{di2015simulator,judd2009numerically,kass1990rapid} and scientific applications~\cite{smith2008towards,bershteyn2018implementation,gleisberg_event_2008}.
-They are rich models and provide a level of interpretability not typically found in modern deep learning methodologies due to the several years, if not decades, of research and development that has led to their construction.
-Thus, simulators by their very nature are information rich and naturally describe complicated joint densities.
-That is the \emph{model}, or equivalently \emph{program}, specified by the simulator defines a joint density $p(x,y)$ and the inputs into the simulator denote priors $p(x)$. 
-But, for all the benefits of using simulations there exist two challenging problems that are currently faced:
-\begin{itemize}
-\item It is difficult to calibrate the internal simulator parameters, as the events they are simulating are expensive to reproduce, or are very rare, so data is scarce. 
-\item As the development process takes place over several years the code base becomes convoluted, making it increasingly challenging to understand the structure of the joint density and the internal operations within the simulator, diminishing the interpretability that the simulator once provided.
-\end{itemize}
+\subsection{Prior based sampling}
 
-One such machine learning paradigm that has the ability to resolve these issues is \emph{Probabilistic programming}\cite{gordon2014probabilistic,milch2005blog,van2018introduction,tran2017deep}. 
-Probabilistic programming enables users to write, condition on and perform inference in, complex probabilistic models in a programmatic, yet efficient and simplified way. 
-The source code of the given model, or equivalently the \emph{program}, is transformed into a program density, which can
-be viewed as the joint density of the user-specified model, \emph{the simulator} in this instance.
-Having direct access to the source code enables PPSs to conduct several processes automatically, such as 
-the automation of complex statistical inference procedures\cite{zhou2019lf,goodman2012church,wood2014new,rainforth2017automating}, which enables us to efficiently calibrate
-model parameters, and we can also track how the internals of the program change during
-run time, enabling a full analysis of the program structure and traces generated\cite{milch2005blog}, which provides greater interpretability, see Figure~\ref{fig:hijacking}.
-In addition to this, as models are directly expressed in program code we can encode other structures to make use of the full structure of a given language, whether that is \texttt{python, C++} or 
-any other language,
-which enables users to integrate PPSs into their standard model development process, increasing efficiency and reducing model iteration time.
+\subsection{Importance Sampling}
 
-\begin{wrapfigure}{R}{0.44\textwidth}
-\vspace{-20pt}
-\centering
-% \includegraphics[width=0.35\textwidth]{../plots/mpenny-pop10/mpenny_10.pdf}
-\includegraphics[width=0.40\textwidth]{../plots/model_lambda/model_posterior.pdf}
-\includegraphics[width=0.40\textwidth]{../plots/model_lambda/hist_posteriors.png}
-% \vspace{-10pt}
-\caption{An example of the posterior output of population-based simulator after hijacking.
-\textit{Top}: Represents the posterior execution traces of the simulator, inside the the probabilistic programming
-system. Each address $A_{i}$ correspond to a physical process within the simulator, we describe these addresses and their structures in more details in Section~\ref{sec:casestudy}.
-\textit{Bottom}: Represents the histograms generated after posterior inference for the parameters of interest, in this case those parameters are represented by addresses $A_{2}$ to $A_{6}$, conditioned on some observations.
-The grey dashed lines represent the ground truth.}
-\vspace{-25pt}
-\label{fig:hijacking} 
-\end{wrapfigure}
-For all the brilliance of probabilistic programming, if one did want to run an existing simulator in a standard probabilistic programming set-up one would have to entirely re-write the simulator in the language of the PPS.
-This in many cases is not feasible, nor efficient due to both the size 
-and complexity of the existing simulator code base. 
-However, what if one could connect arbitrary a simulator, written in any
-language and directly connect it to a PPS, without having to re-write
-the simulator in the given PPS?
+\subsection{Sequential Monte Carlo}
 
-\paragraph{Related work} Our method of hijacking simulators builds on the work of~\cite{baydin2018efficient,casado2017improvements}. 
-They demonstrated how this could be done for event-based simulators, but there methods neglected a large class of population-based simulators that model interactions
-
-between populations, that are utilized for modeling epidemics, multi-agent systems and financial economics \cite{raberto2001agent,landau_binder_2014,hartig2011statistical,smith2008towards,bershteyn2018implementation}.
-Our framework provides a solution and makes it easy to transform arbitrary stochastic population-based simulator, in addition to the event-based simulators, into probabilistic programs, regardless of the complexity of the simulator.
-Our system supports simulators written in 13 different programming languages and enables the user to perform posterior inference over the density of the simulator, conditioned on what data the user has access too, including
-simulator output. 
-We provide a theoretical justification for our approach and many examples.
-% This enables two things.
-% 1) It enables software developers and researchers to understand complicated code bases. 
-% 2) It provides interpretable inference results that provide policy makers with predictions that are truly interpretable, in the sense that the end-user, of the inference results understands what physical events led to inference outcome.
-% This is critical in several domains, and is particularly important in the medical domain. We provide examples of both in Section~\textbf{TO ADD REF}.
-
-% In this paper, we introduce a novel method that allows one to shed light on the inner workings of a large class of population-based stochastic simulators.
-We achieve this by extending the work of \citep{baydin2018efficient} by interpreting such population-based simulators as probabilistic generative models within the framework of universal probabilistic programming (UPP) \cite{le-2016-inference}.
-To this end, we \emph{hijack} existing simulators by overriding their internal random number generators.
-Specifically, by replacing the existing low-level random number generator in a simulator with a call to a purpose-built UPP ``controller'', which can thus control, track and manipulate the stochasticity of the simulator.
-
-% Moreover, though our focus is on hijacking epidemiology simulators, our general approach and framework can be applied more widely to a general class of stochastic simulators.
-
-% In time, we hope our approach will play a critical role in
-% bringing recent advancements in probabilistic programming to bear on
-% the vast array of existing simulators used throughout the sciences,
-% thereby providing wide-ranging impacts across a number of fields.
-
- % \section{Background}
-% \label{sec:backgorund}
-
-% \begin{itemize}
-% \item Paragraph on Probabilistic programming
-% \item Paragraph on Simulators keep high-level focus on pop sims later
-% \item Paragraph on Baydin et al
-% \end{itemize}
-% In-silico simulators have become a crucial tool in evidence-based decision-making within a large number of disciplines, including statistical physics~\cite{landau_binder_2014}, financial modeling~\cite{jackel2002monte},
-% weather prediction~\cite{evensen1994sequential}, epidemiology~\cite{smith2008towards} and many others.
-% In many cases, simulation output can augment or even replace real data that may otherwise be costly or even impossible to generate.
-% Recent advances in hardware have enabled simulations to model increasingly complex systems.
-% Epidemiology studies the prevalence and spreading of diseases across populations.
-Recent advances in hardware have enabled simulations to model the dynamics of infectious diseases, such as Malaria, in ever greater detail.
-
-% Probabilistic programming~\cite{gordon2014probabilistic,staton2016semantics,kozen1979semantics} 
-% can be used to express probabilistic models and consequently perform automated inference these.
-Once a probabilistic model has been expressed in a probabilistic programming language, a wide range of inference techniques, such as Markov Chain Monte Carlo (MCMC)\cite{geyer1992practical}, Variational Inference~(VI)\cite{wainwright2008graphical} and Deep Neural Networks~(DNN)\cite{Goodfellow:2016:DL:3086952},  can be used by non-experts in an automated fashion.
-
-% \section{Types of Models and Probabilistic Programming Languages}
-
-% To bridge the gap between the statistics and machine learning community to the probabilistic programming community we provide a useful overview to expose our ideas more concretely.
-
-% Probabilistic programming languages, broadly speaking\footnote{For a more comprehensive overview see \cite{rainforth2017automating}}, can be classfied into three categories, each type of langauge extends a particular model space, enabling an increasingly complex set models to be represented as programs, but in doing so the notion of posterior and what it means to do inference over the model, at least in a stastically traditional sense, can change.
-% % \begin{figure}
-% % \centering
-% % \includegraphics[]{../models/types_of_pp.png}
-% % \end{figure}
-% The first type of langauges are called first-order languages where the priors are well defined and the \textit{joint} program density constructed via the program directly represent the unormalised stastical joint density $p(x,y)$
-% \section{Methodology}
-%  TODO
-
-\section{Hijacking Simulators}
-\label{sec:methods}
-In-silico simulators have become a crucial tool in evidence-based decision-making within a large number of disciplines, including statistical physics~\cite{landau_binder_2014}, financial modeling~\cite{jackel2002monte},
-weather prediction~\cite{evensen1994sequential}, epidemiology~\cite{smith2008towards} and many others.
-In many cases, simulation output can augment or even replace real data that may otherwise be costly or even impossible to generate.
-Recent advances in hardware have enabled simulations to model increasingly complex systems.
-Epidemiology studies the prevalence and spreading of diseases across populations.
-Recent advances in hardware have enabled simulations to model the dynamics of infectious diseases, such as Malaria, in ever greater detail.
-
-\subsection{The Engineering}
-Hijacking a simulator a describes the process by which a simulator's random number generators are replaced by calls to external sampling procedures, which are controlled by a probabilistic programming system (PPS).
-In practice, this amounts to performing a small number of surgical incisions into the simulator's source code in order to replace built-in calls to random number generators.
-E.g., given a simulator written in C++/Boost\cite{schaling2011boost}, a Gaussian distribution object \textit{boost::normal} is replaced with the corresponding \textit{pyprob\_cpp} distribution object, namely \textit{pyprob\_cpp::distributions::Normal}.
-Sampling from this distribution is then done by requesting the PPS to send a sample back to the simulator.
-\bg{This will be extended}
-
-\begin{center}
-\begin{tabular}{ |c|c|c| }
-\hline
-Symbol & Meaning  \\
-\hline
-\n{x} & Number of latent variables \\
-\n{y} & Number of input observations \\
-\n{y’} & Number of output observations, generated simulator values \\
-\rv{x} & The latent parameters \\
-\rv{y} & The input observed parameters \\
-\rv{y’} & The output observed parameters \\
-\model{\cdot} & Model - i.e .xml file \\
-\simu{\cdot} & Simulator \\
-\micro{k}{\cdot} & $k$-th Micro-simulation \\
-\pdf{\cdot} & probability density function \\
-\pmf{\cdot} & probability mass function \\
-\add{i} & Address $i$ \\
-\expect{\cdot} & Expectation (first moment) \\
-\trace{j} & The $j$-th collapsed Trace \\
-\traces{i}{j} & A single, un-collapsed Trace for the $i$th trace. ``A program trace’' \\
-\stoc{h}{i} & A stochastic process at step $h$ for \add{i} \\
-\hline
-\end{tabular}
-\end{center}
-  \begin{algorithm}
-  \small
-   \caption{Hijacking a stochastic simulator for an arbitrary machine}
-    \begin{algorithmic}\small
-      \Function{Simulator to a Probabilistic Program}{Simulator $\Omega(\cdot)$, Model input to simulator \model{\cdot}, PPS sampling primitive $f_t()$, Message passing protocol (PPX) $h_t()$, Random variable $x_{sim}$, PPS $\mathcal{P}$ }
-       \State  $ x_{sim} \rightarrow f_{t}(x_{sim}) $\Comment{Replace raw sample statements}
-       \State $\simu{f_{t}(x^{t}_{sim})} \rightarrow  h_{t}(\simu{f_{t}(x^{t}_{sim})}) \rightarrow \mathcal{P}(f_{t}(x^{t}_{sim}))$ \Comment{An API between the simulator enables raw RNG to be passed through the PPS}
-
-        \Function{Create Containerized Environment}{$C(\Delta)$}
-        \For{$\delta$ in $\Delta$}
-        \State Add dependency $\delta$
-        \EndFor
-        \State Run Program
-       \EndFunction
-       \Function{Create thin wrapper}{$f_{t}(x_{t} | \eta_{t})$}
-       \State to add
-       \EndFunction
-       \Function{Create Containerized Environment }{$\Delta_{i}$}\Comment{Where $\Delta_{i}$ is one of $i$-th dependencies}
-       \State to add
-       \EndFunction
-       \EndFunction
-
-\end{algorithmic}
-\end{algorithm}
-
-% Stochastic simulators are common across a number of domains, statistical physics~\cite{landau_binder_2014}, financial modeling~\cite{jackel2002monte},
-% weather prediction~\cite{evensen1994sequential}, epidemiology~\cite{smith2008towards} and many others. 
-% An advantage of using stochastic simulators is that they provide a level of interpretability not found in modern deep learning settings, as they directly incorporate model structure from carefully reasoned observations and experiments.
-In addition to this,
-
-The PPS and the simulator exchange xTensor objects \url{https://xtensor.readthedocs.io/en/latest/} through TCP or ICP using a generic FlatBuffers \url{http://google.github.io/flatbuffers/} protocol.
-On the PPS side, sampling is done using the deep learning framework PyTorch \cite{paszke2017automatic}.
-
-After a variable has been sampled, it is sent to the PPS using \textit{pyprob\_cpp::sample}, alongside with its simulator source trace.
-This allows the PPS to construct sample trace probabilities and other summary statistics (cf.
-Section \ref{sec:casestudy}).
-
-Finally, the entry point to the simulator, i.e. \textit{main} in C++, is replaced by a special \textit{forward} call, in this case \textit{pyprob\_cpp::forward}.
-This allows the PPS to generate rollouts from the simulator remotely.
-
-\paragraph{Population-based simulators} In contrast to the setup used by \cite{baydin2018efficient}, who interface a particle physics simulator that creates a single trace per \textit{forward} call, a population-based simulator creates a trace per population member.
-This means that e.g.
-in EMOD, a standard scenario simulation rollout over the span of $3$ years
-with a population of size $n=5000$ \cite{smith2008towards} will generate about \textit{four terabytes} of raw trace data.
-Unlike with \citep{baydin2018efficient}, this amount of data cannot possibly be kept in RAM, which makes \textit{a posteriori} trace analysis very inefficient.
-In order to deal with the shortcomings of \textit{pyprob} in a population-based simulator context, we therefore extend the framework to be able to do trace analysis on the fly.
-
-By extending \textit{pyprob} to handle population-based simulators, the PPS can now track all stochastic random variables that are created within the simulator, which then allows us to generate trace plots and path probabilities associated to the execution paths of the program.
-The corresponding increase of simulator transparency helps reinstate much needed trust between policymakers and evidence-based methods.
-
-\subsection{The Theory}
-
-\bg{Tom add theory stuff here}
-%As stated previously, Simulators are complex systems and should be viewed as black-boxes, thus deciphering the underlying processes occurring within the simulator facilitates the users understanding of how such a system operates and provides a way of interpreting the outputs in a clear way.
-In addition to this, when using a simulator to make new inferences a user will often want to condition on the output of the simulator, or any new observations gathered.
-In the standard simulator framework this is not possible, but in
-%
-%Essentially, from the users perspective, our approach can be viewed as follows the user would first select the simulator that they are going to use, for example EMOD, they would then write their model, in the desired simulator, according to the given simulator specification.
-The user would then run the simulator.
-Now comes the hijacking.
-During the forward run of the simulator all sampling calls from the random number generator are directed through the probabilistic programming system~(PPS), which in addition to the standard output of the simulator now generates a series of result summaries and path probabilities, as we demonstrate in Section~\ref{sec:casestudy}.
-%
-%By connecting with the PPS we track all sampling procedures, which enables us to control the myriad of random variables generated within the simulator. 
-%
-%In order to overhaul a stochastic simulator like EMOD, or OpenMalaria, we are only required to override the location of the stochastic primitives, i.e.
-the base operations that generate the stochasticity and randomness within the simulator.
-%We will walk through how this is done for a standard epidemiology simulator, see the flowchart for a pictorial representation of what is happening \bg{add ref to flow chart}.
-%
-%The initial step first requires one to build a containerized environment, such as Docker~\cite{merkel2014docker} and Singularity~\cite{kurtzer2017singularity}, which enables the simulators to be run on any device, independent of different hardware architectures, removing the requirement for hardware specific machines.
-%The second step requires one to override the stochastic primitives, which requires re-writing aspects of the sampling procedures inside the simulator. 
-%This requires zero domain-expertise and only requires one to find where the sampling calls are originated, this can be done with any basic editor search function.
-% Such information is typically saved in $\texttt{RANDOM.<filetype>}$, although it may vary slightly between simulators. We now demonstrate this for a uniform primitive.
-%
-%The third step is generating the protocols, which enable communication between the simulator and probabilistic programming system. 
-%For this we extend the existing bridging framework introduced in~\cite{baydin2018efficient}, which is comprised of the Probabilistic Programming eXecution~(PPX) protocols and a light-weight C ++ interface, however the language of this light-weight interface is dependent on whatever the file type of the $\texttt{RANDOM.<filetype>}$ file is.
-%For both OpenMalaria and EMOD, the file is written in C ++.
-%In using these protocols a larger class of sampling procedures can be overridden.
-The PPX protocols work by making use of the flatbuffer protocols\footnote{\url{https://github.com/google/flatbuffers}}, which require a flatbuffers script to be specified, see here for an example of a flatbuffers script\footnote{https://github.com/probprog/ppx/blob/master/ppx.fbs}.
-Once this is specified the flat buffers script is compiled and automatically constructs the message passing framework between the simulator and the probabilistic programming system.
-The flatbuffers script is simple, clean and intuitive, making it easy for a non-expert to produced.
-%The next steps requires modifying the $\texttt{main}$ function in $\texttt{main.cpp}$ and defining an additional forward function which runs one execution of the simulator to enable message passing between the simulator and PPS\footnote{A link to an example will be provided after the review period to preserve anonymity}.
-%The final step is to create a light-weight interface.
-As both the EMOD and OpenMalaria are solely based on C ++ we can extend an existing interface called pyprob\_cpp\footnote{\url{https://github.com/probprog/pyprob_cpp}} to account for the additional sampling procedures.
-%Once these steps are complete, the containerized environment will now contain everything required to pass models through to the simulators, encode the simulator into a probabilistic program, which can then be evaluated in the PPS.
-
- %Our method of overriding simulators builds on the work of~\cite{baydin2018efficient} and extends the framework to encapsulate a more diverse range of simulators, in particular population based simulators such as those found in epidemiology.
-%Our framework provides a simple solution to do this and makes it easy to transform arbitrary stochastic simulators into probabilistic programs, regardless of the complexity of the simulator and the language that the simulator is written in.
-%In particular, the framework currently supports 13 popular languages, but there is nothing to stop this being extended to any language of interest. 
-%Furthermore, we can understand the structure of simulators as programs for both decoding \emph{black-box structures} and for posterior inference, which enables two things. 
-%1) It enables software developers and policy makers to understand complicated code bases and dissect the outputs of simulators for guiding policy decision making.
- %2) It provides interpretable inference results that provide policy makers with predictions that are truly interpretable, in the sense that the end-user of the inference results understands what physical events led to the inference outcome.
- %This is critical in several domains, especially in epidemiology. 
- %We provide examples of (1) in next Section~\ref{sec:casestudy} and discuss the latter in future work as extensions to the the current framework.
-
-\section{Experiments}
-\subsection{Epidemiology Simulators}
-
-Malaria epidemiology is governed by a complex set of drivers, 
-few of which can be understood in isolation \cite{cameron2015defining,autino_epidemiology_2012,smith2008towards,bershteyn2018implementation}.
-These include within-host dynamics, population-specific traits and even local geography.
-Comprehensive modeling of all of these components remains challenging, particularly in a region-specific context. 
-Computational epidemiology simulators have to reflect these complexities and are usually stochastic in nature.
-This can make simulation output highly non-trivial to interpret, particularly when trying to draw desired inferences coupled with observed data \cite{mwendera_challenges_2019,ferris_openmalaria_2015}.
-
-We first demonstrate our system on a reduced version of the OpenMalria simulator, proposed by Smith et .al~\cite{smith2006relationship}.
-The following model directly models the relationship between the entomological inoculation rate and the force of infection for the \emph{plasmodium flaciprum} Malaria.
-
-Two the most advanced Malaria simulators, namely EMOD~\cite{bershteyn2018implementation} and OpenMalaria  \cite{smith2008towards}, have proven to be particularly valuable to policymakers.
-OpenMalaria is based on microsimulations of Plasmodium falciparum in humans and was originally developed to simulate the impacts of malaria vaccines within simple villages or districts
-Compared with OpenMalaria, EMOD is able to simulate a variety of additional drivers, including complex geographies complete with migration and a large number of policy interventions
-Both EMOD and OpenMalaria are open source and implemented in C++.
-
-\bg{Add linking stuff now to the toy-example where hopefully we can perform inference and }
-
-% Simulators provide users with the ability to model complex environments and problems in a black-box way, enabling new insights to be generated and predictions to be made.
-A simulator typically encodes several years, if not decades, of research and development and so by their very nature are structurally complicated.
-
-% Within the context of epidemiology simulators we can model the treatment, infection and prevention cycle under a variety of different environmental conditions, enabling predictions to be made on a set of field observations.
-Such scenarios in practices would be
-
-% This enables practitioners to model a wide array of different scenarios at ease.
-% However, since the development happens over several years, or decades, usually without comprehensive documentation, understanding a code base written with legacy libraries and software is incredibly difficult and an almost impossible task for a non-expert, or expert who has not had access to the previous years of development.
-%
-
-% Two prominent simulators for modeling Malaria are OpenMalaria~\cite{smith2008towards} and Emod~\cite{bershteyn2018implementation}, which are used by practitioners to model several aspects of the Malaria disease.
-
-%  \paragraph*{OpenMalaria} 
-% is an open-source simulator funded by the Gates Foundation to model different aspects of the malaria prevention and treatment cycle for varying population dynamics.
-%  Users specify models in OpenMalaria via a scenario file, in the form of an XML document, which determines the specifics of a particular treatment, or prevention strategy, climatic factors, the demographics of the population and so forth.
-In addition to this, field observations can be encoded inside of, or externally to the scenario file.
-% The scenario file encodes all input and output parameters, some of which we wish to simulate, others remain fixed.
-
-% \paragraph{EMOD} provides the modularity and structural flexibility necessary for modeling multiple infectious diseases at the within-host and population level. 
-% EMOD can be used to model HIV, Tuberculosis, Malaria and generic diseases, but for the purposes of the comparison, we just focus on Malaria. 
-% To specify models in EMOD a configuration file is written by the user, which is in the form of a JSON and provides similar functionality to OpenMalaria, in terms of what can be specified and the outputs that it produces.
-
-% In both these well used simulators users are accustomed to writing their models as either XML, or JSON files and so it is important that any system built around this framework enables the models to be inputted in the same format, without requiring the users to learn how to use a new set of protocols.
-Our system facilitates this as we shall discuss in the proceeding sections.
-
-\begin{figure}
-  \includegraphics[width=\textwidth]{../plots/system.pdf}
-  \label{fig:how}
-  \caption{This flow-chart provides an overview of the process of how our system hijacks a generic population-based epidemiology simulator, such as EMOD, and how we modify the simulator to 
-  to hijack the random number generator (RNG).}
-
-  \end{figure}
-
-% \begin{table}[h!]
-% \footnotesize
-% \begin{tabular}{p{3.6cm}|p{3.6cm}}
-% Before  &  After  \\
-% \midrule
-% \begin{lstlisting}
-% #include <gsl/gsl_cdf.h>
-% #include <gsl/gsl_rng.h>
-% #include <gsl/gsl_randist.h>
-% //preamble
-% double random::uniform_01 () {
-%     double result =
-% # ifdef 
-% OM_RANDOM_USE_BOOST
-%        rng_uniform01 ();
-% # else
-%  gsl_rng_uniform (rng.gsl_generator);
-% # endif
-%     return result;
-%     }
-% \end{lstlisting}&
-% \begin{lstlisting}
-% #include <pyprob_cpp.h>
-% #include "xtensor/xadapt.hpp"
-% // preamble
-% double random::uniform_01 () {
-% auto uniform = pyprob_cpp::distributions::Uniform(0,1);
-% return pyprob_cpp::sample(uniform)(0);
-% }
-% \end{lstlisting}
-% \end{tabular} 
-% \end{table}
-% As can be seen, writing the new sampling call is cleaner and syntactically simpler. 
-%\bg{Add performance plot showing memory consumptions of pyprob and our new system.}
-\subsection{Real-world Example}
-\label{sec:casestudy}
-
-\bg{Chris will write up}
-Ensemble methods are commonly used in statistics in order to combine the predictive power of multiple models \cite{cameron2015defining,smith_ensemble_2012}.
-To this end, recent work has attempted to characterise the similarities and differences between two of the most advanced Malaria epidemiology simulators, EMOD \cite{bershteyn2018implementation} and OpenMalaria \cite{smith2008towards}.
-Evaluation is usually done by comparing a number of output parameters across a range of hand-crafted standard scenarios reflecting different geographical locations across Africa~\cite{smith_ensemble_2012}.
-
-In the following, we illustrate how our method introduces a novel introspection paradigm.
-By extracting trace graphs from population-based simulators, policymakers can ask specific questions about properties of the model trace flow.
-
-To illustrate the above, we present simulation output generated from a scenario resembling local conditions in the town of Ifakara (Tanzania).
-Both EMOD and OpenMalaria are configured to simulate a single population node of $n=100$ and assume constant climatic conditions and no migration over the simulation period of $3$ years.
-Please refer to Figure \ref{fig:EIR} for the seasonal Entomological Inoculation Rate (EIR), a measure of infectivity.
-
-\begin{figure}[h!]
-  \centering
-  \includegraphics[width=\textwidth/2]{../plots/EIR_Ifakara.pdf}
-  \caption{Seasonal Entomological Inoculation Rate (EIR) for the Ifakara scenario. Data is averaged over $30$ day periods.}
-  \label{fig:EIR}
-\end{figure}
-
-We provide examples of the generated trace plots from the connection between the simulator and the PPS in Figure~\ref{fig:plotewan}.
-
-We can see from the addressing schemes $A_1, \ldots, A_N$ (Tables \ref{table:addresses} and \ref{table:asingleaddress}) what physical events are connected to each other and how outputs in EMOD are generated from a different set of procedures as compared to OpenMalaria
-By having access to such diagrams users can internally evaluate and scrutinize the decisions that the simulator is making.
-
-This is important for policymakers, or general non-experts, as it not only details how we arrive at the given outputs, but it provides an understanding of which processes were most crucial in determining those outputs as can be seen from the path probabilities assigned to each of the vertices.
-
-Additionally, by associating nodes in trace graphs representing the same physical processes within different models, the significance of model detail can be evaluated.
-For example, the full trace presented in Table \ref{table:asingleaddress} represent a sampling step associated with within-host dynamics of the Malaria parasite \textit{Falciparum} in OpenMalaria node $A1$.
-The same physical process also occurs in EMOD's trace graph at position $A7$ (see Section \ref{table:addresses}).
-Comparisons like these could help developers better control model complexity, and even provide an alternative testing and debugging paradigm.
-
-\subsection{Larger Scale-Simulators: OpenMalaria and EMOD}
-
-\begin{figure*}[h!]
-  \centering
-  \includegraphics[width=\textwidth]{../plots/final_emod_traces__rotated.pdf}
-  \includegraphics[width=\textwidth]{../plots/ewan_25_pop_100.pdf}
-  \caption{Here we run two equivalent models, compare the corresponding trace paths and corresponding 
-  path probabilities taken by the thousands of random variables generated internally within the simulators.
-\textit{\textbf{Top:}} The specified model run in EMOD.
-\textit{\textbf{Bottom:}} The specified model in OpenMalaria.}
-  \label{fig:plotewan}
-\end{figure*}
-
-\begin{figure*}
-  \centering
-  \includegraphics[width=\textwidth]{../plots/mastsari/mastsari_prior_trace_structure.pdf}
-  \caption{Prior trace probabilities of the Mastsari model.}
-  \label{fig:priortracemastsari}
-\end{figure*}
-
-\begin{table}[h!]
-\footnotesize
-  \setlength{\tabcolsep}{1mm}
-\label{table:asingleaddress}
-  \caption{An example of an address generated for the model run in the OpenMalaria simulator. We can see that A1
-  relates to Generating a member of the human population who may or may not be infect with the Malaria disease.
-We get something similar for EMOD, except this relates to A7 in the EMOD program execution.}
-  \def\arraystretch{1.25}
-  % \setlength{\tabcolsep}{1mm}
-  \begin{tabularx}{\textwidth}{@{}lX@{}l@{}} 
-    \toprule
-    Address ID & Full address \\
-    \midrule
-  A1 & [forward()+0x204;OM::Simulator::start(scnXml::Monitoringconst)+0x28a;
-  OM::Population::createInitialHumans()+0x94;
-
-  OM::Population::newHuman(OM::SimTime)+0x5c;
-  OM::Host::Human::Human(OM::SimTime)+0x12b;
-
-  OM::WithinHost::WHInterface::reateWithinHostModel(double)+0x99;
-  OM::WithinHost::DescriptiveWithinHostModel::DescriptiveWithinHostModel(double)+0x3a;
-  OM::WithinHost::WHFalciparum::WHFalciparum(double)+0xe6;
-  OM::util::random::gauss(double, double)+0xb4]\_\_Normal \\
-
-  $\vdots$ & $\vdots$ \\
-
-  A5 & [forward()+0x204;OM::Simulator::start(scnXml::Monitoringconst\&)+0x468;
-  OM::Population::update1(OM::SimTime)+0xff;
-
-  OM::Host::Human::update(bool)+0x2bc;
-
-  OM::Clinical::ClinicalModel::update(OM::Host::Human\&,double, bool)+0x96;
-
-  OM::Host::NeonatalMortality::eventNeonatalMortality()+0x9;
-
-  OM::util::random::uniform\_01()+0xc0]\_\_Uniform\\
-  $\vdots$ & $\vdots$ \\
-
- % $\vdots$ & $\vdots$ \\
-
-\bottomrule
-  \end{tabularx}
-  \end{table}
-
-\begin{table}[h!]
-  \footnotesize
-  \setlength{\tabcolsep}{1mm}
-  \caption{An interpretation table for each of the address of the 
-  overall trace generated from the corresponding OpenMalaria model.}
-  \label{table:addresses}
-  % \def\arraystretch{1.25}
-  % \setlength{\tabcolsep}{1mm}
-  \begin{tabularx}{\textwidth}{@{}lX@{}} 
-    \toprule
-    Address ID & Interpretation \\
-    \midrule
-  A1 & Generate a human in the population within host dynamics\\
-
-  A2 & Generate another human in the population within host dynamics\\
-
-  A3 & The population is updated and a new human, or humans, may get infected \\
-
-  A4, A5 & Potential child deaths within the population are simulated \\
-
-  A6 & Determines parasite density of an individual infection \\
-
-  A7 & Models how the disease is progressing within the infected humans \\
-
-  A8 & Models how the disease is progressing within the population \\
-
-  A9 & Models how the disease is progressing within the infected humans after the population has been updated\\
-
-  A10 & Full clinical update on the population for those without severe or no Malaria infection. \\
-
-  A11 & Full clinical update on the population for those with severe Malaria infections \\
-\bottomrule
-  \end{tabularx}
-  \end{table}
-
-\section{Discussions and Future Work}
-
-In this work we have demonstrated a method 
-that enables one to hijack populationbased simulators, extending 
-the work of \citep{baydin2018efficient}. 
-We applied our method to two Malaria orientated population-based simulators and 
-generated a variety of trace graphs.
-Finally, we have shown how our system
-enables policy makers and non-experts to analyse simulator outputs in a 
-way previously unavailable in the field of epidemiology.
-
-To extend our work further we aim to implement additional 
-tools that will facilitate complicated inference procedures that condition on simulator 
-output. We will also evaluate additional scenarios across Africa and South-East Asia to better understand the similarities and differences between EMOD and OpenMalaria.
-
-% \subsubsection*{Acknowledgments}
-
-% Use unnumbered third level headings for the acknowledgments. All acknowledgments
-% go at the end of the paper. Do not include acknowledgments in the anonymized
-% submission, only in the final paper.
-
-% \section*{References}
+\subsection{Random-walk Metropolis Hastings}
 
 \bibliographystyle{plain}
-\bibliography{refs,malaria}
+\bibliography{refs}
 

### 2019-07-04 code/openmalaria/model/WithinHost/Diagnostic.h

@@ -1,4 +1,4 @@
-/* This file is part of OpenMalaria.
+cd ../* This file is part of OpenMalaria.
  * 
  * Copyright (C) 2005-2015 Swiss Tropical and Public Health Institute
  * Copyright (C) 2005-2015 Liverpool School Of Tropical Medicine

### 2019-07-04 code/toy_model/toy/toy_tempy2.py

@@ -1,0 +1,283 @@
+# %%
+import pyprob
+from pyprob import Model
+from pyprob.distributions import Poisson, Categorical, Normal, TruncatedNormal, Beta, Uniform
+import torch as th
+import torch
+
+import matplotlib.pyplot as plt
+# % matplotlib
+# inline
+
+
+
+
+class ToyMalaria(Model):
+    def __init__(self, pop_size=1050, likelihood_stddev=0.1, EIR=None):
+        super().__init__(name='Toy Malaria')
+        self.pop_size = pop_size
+        self.likelihood_stddev = likelihood_stddev
+        self.t_max = 70
+        self.days_per_step = 5
+        #self.EIR = th.FloatTensor(self.t_max * self.days_per_step).zero_() + 67.0/365.0
+        self.EIR = th.FloatTensor(self.t_max * self.days_per_step).zero_() + 5.5 / 365.0 # Matsari, intervention phase
+        self.EIR_background = th.FloatTensor([67.0/365.0])
+        self.A_max = 1.63
+        self.prevalence_bins_upper = torch.FloatTensor([0.246746962983009,
+                                             0.727755459545533,
+                                             1.44479521753634,
+                                             2.88435310594495,
+                                             4.81735572357366,
+                                             7.82460584536638,
+                                             12.087165127225,
+                                             16.8889599400757,
+                                             24.2654245103456,
+                                             33.7166980261767,
+                                             48.1735572357366,
+                                             67.3111742713562]) * 365
+        self._eval = False
+
+    def forward(self, _eval=False):
+        '''
+        Parameters:
+            i - Individual
+            t - Individual at time t
+            a - age group of individual at time t
+            A(a(i,t)) - Mean body suface area
+
+        Return:
+            E_a(i,t) - The estimate at each time, for an individual i of
+            number of infected bites recieved, adjusted for age.
+
+        Surivival function
+            S_p(i,t) - Probability that the progeny of each incoculation survive
+            to give rise to a patent blood stage infection
+
+        Force of infection
+            lambda(i,t) = S_p(i,t) * E_a(i.t)
+
+        Pre-erythocytic challenge
+            X_p = \int_{t-a)^{t} E_a(i,t) d\tau
+
+        Latent variables:
+            h
+            gamma_p
[…]

### 2019-07-04 paper/learning/thesystem.tex
added tom notes

% !TEX root = notes.tex

\tikzset{
    state/.style={
           rectangle,
           rounded corners,
           draw=black, very thick,
           minimum height=2em,
           inner sep=2pt,
           text centered,
           },
}

\begin{tikzpicture}[->,>=stealth']

 % Position of Sample 
 % Use previously defined 'state' as layout (see above)
 % use tabular for content to get columns/rows
 % parbox to limit width of the listing
 \node[state,
 text width=3cm, 	% max text width
 yshift=0cm, 		% move 2cm in y
 node distance=0cm, 	% distance to Sample
 anchor=center] (Sample) 
 {\begin{tabular}{l}{
  \textbf{Sample}\\
  \parbox{4cm}{
    \small{Program call:\\
    \emph{Program}.\texttt{forward()}}
  }
 \end{tabular}};

 \node[state,    	% layout (defined above)
 text width=3cm, 	% max text width
 yshift=-2cm, 		% move 2cm in y
 below of=Sample, 	% Position is to the right of Sample
 node distance=0cm, 	% distance to Sample
 anchor=center] (OBSERVE) 	% posistion relative to the center of the 'box'
{%
\begin{tabular}{l} 	% content
 \textbf{Observe}\\
 \parbox{4cm}{
  \small{Program call:\\
  \emph{Program}.\texttt{forward()}}
}
\end{tabular}};

 % State: PyProb with different content
 \node[state,    	% layout (defined above)
  text width=3cm, 	% max text width
  yshift=0cm, 		% move 2cm in y
  right of=Sample, 	% Position is to the right of Sample
  node distance=6.5cm, 	% distance to Sample
  anchor=center] (PyProb) 	% posistion relative to the center of the 'box'
 {%
 \begin{tabular}{l} 	% content
  \textbf{System}\\ %system sample
  \parbox{4cm}{Pyprob processes\\ \texttt{sample}}
 \end{tabular}
 };

 % STATE System
 \node[state,
  below of=PyProb,
  yshift=-1cm,
  anchor=center,
  text width=3cm] (System) 
 {%
 \begin{tabular}{l}
  \textbf{System}\\
  \parbox{4cm}{Pyprob processes \\\texttt{observe}}
 \end{tabular}
 };

 % STATE EPC
 \node[state,
  right of=PyProb,
  yshift = -0cm,
  node distance=5cm,
  anchor=center] (EPC) 
 {%
 \begin{tabular}{l}
  \textbf{Program density}\\
  \parbox{4cm}{P(\mathcal{T}= $x_{1:n_x}| \lambda)$}
 \end{tabular}
 };

  % STATE OUTPUT
  \node[state,
  right of=System,
  yshift = 0cm,
  node distance=5cm,
  anchor=center] (OUTPUT) 
 {%
 \begin{tabular}{l}
  \textbf{Program Output}\\
  \parbox{4cm}{$\Omega = \{P(x_{1:n_{x}}), w\}$}
 \end{tabular}
 };

 % draw the paths and and print some Text below/above the graph
 \path (Sample) 	edge[bend left=20]  node[anchor=south,above]{$f_{a_{j}} ; \eta_{j}$}(PyProb)
 (OBSERVE)     	edge[bend right=20] node[anchor=south,above]{$g_{b_{i}},\phi_i , y_i $} (System)
 (PyProb)       	edge [bend left=20] node[anchor=south,above]{$g_{b_{i}}(y_i, \phi_i)$} (EPC)
 (PyProb)       edge [bend left=9] node[anchor=west, above]{$x_{i}$ \emph{returned}} (OBSERVE)
 (System)       	edge [bend right=10]                                          (EPC)
 (EPC)              edge                                                    (OUTPUT)
 (System)  	edge   node[anchor=north,above]{\emph{Control}} (OBSERVE);
%  (System)  	edge[loop below]    node[anchor=north,below]{$SC_n\neq 0$} (System)
%  (System)  	edge                node[anchor=left,right]{$SC_n = 0$} (PyProb);

\end{tikzpicture}

### 2019-07-04 code/openmalaria/model/Simulator.cpp

@@ -39,7 +39,7 @@
 #include "util/random.h"
 #include "util/StreamValidator.h"
 #include "schema/scenario.h"
-
+#include <iostream>
 #include <fstream>
 #include <gzstream/gzstream.h>
 
@@ -152,11 +152,9 @@ void Simulator::start(const scnXml::Monitoring& monitoring){
     // length required by vector warmup, and is a whole number of years.
 
 
-    bool local = false;
-    pyprob_cpp::setLocal(local);
+
     SimTime humanWarmupLength = sim::maxHumanAge();
     if( humanWarmupLength < sim::transmission().minPreinitDuration() ){
-        // added by Bradley - non OM team - Turns pyprob tracking off
         cerr << "Warning: human life-span (" << humanWarmupLength.inYears();
         cerr << ") shorter than length of warm-up requested by" << endl;
         cerr << "transmission model ("
@@ -167,14 +165,8 @@ void Simulator::start(const scnXml::Monitoring& monitoring){
     }
     humanWarmupLength = SimTime::fromYearsI( static_cast<int>(ceil(humanWarmupLength.inYears())) );
 
-    /* Bradley: We may need to an additional if statement in the above condition
-       to look for the last run of warm up as we will require the value of that sample
-     */
-    // added by Bradley - non OM team - Turns pyprob tracking back on
-    bool new_local = false;
-    pyprob_cpp::setLocal(new_local);
 
-    totalSimDuration = humanWarmupLength  // ONE_LIFE_SPAN
+     totalSimDuration = humanWarmupLength  // ONE_LIFE_SPAN
         + sim::transmission().expectedInitDuration()
         // plus MAIN_PHASE: survey period plus one TS for last survey
         + mon::finalSurveyTime() + SimTime::oneTS();
@@ -192,9 +184,11 @@ void Simulator::start(const scnXml::Monitoring& monitoring){
     // set once, since we exit after a checkpoint triggered this way.
     SimTime testCheckpointTime = util::CommandLine::getNextCheckpointTime( sim::now() );
     SimTime testCheckpointDieTime = testCheckpointTime;        // kill program at same time
-    
+
     // phase loop
     while (true){
+
+
         // loop for steps within a phase
         while (sim::now() < simPeriodEnd){
             util::BoincWrapper::reportProgress(sim::now().raw(), totalSimDuration.raw());
@@ -225,9 +219,18 @@ void Simulator::start(const scnXml::Monitoring& monitoring){
[…]

### 2019-07-04 code/toy_model/toy/toy.py

@@ -1,0 +1,107 @@
+import os
+
+from envs.multiagentenv import MultiAgentEnv
+from utils.dict2namedtuple import convert
+import torch as th
+
+
+class ToyMalaria(Model):
+    def __init__(self, pop_size=10, likelihood_stddev=0.1, EIR=None):
+        super().__init__(name='Toy Malaria')
+        self.pop_size = pop_size
+        self.likelihood_stddev = likelihood_stddev
+        self.t_max = 10
+        self.days_per_step = 1
+        self.EIR = EIR
+        self.A_max = 1.63 # average surface area of population >= 20 years
+        # Note: This is currently crudely estimated as 160cm, 60kg
+        # by Mosteller: 1.63 m^2
+
+    def forward(self):
+        # Generate initial population age distribution
+        pop_age_days = th.LongTensor(self.pop_size).random_(0, 80*365)
+        E_a = None # equiv. to E_a
+        E_max = self.EIR
+
+        # Initialize X_p
+        X_p = th.FloatTensor(self.pop_size).zero_() # equiv. to X_p, Equ. 6
+        for t in range(max(pop_age_days)):
+            X_p[pop_age_days >= t] += self.get_EIR_age_adjusted(pop_age_days[pop_age_days >= t]*0+t)
+
+        # now evolve over time
+        for t in range(self.t_max):
+            # update age-adjusted EIR
+            E_a = ( E_max[t] / self.A_max ) * self.get_mean_body_surface_area(pop_age_days) #self.get_EIR_age_adjusted(pop_age)
+            X_p += E_a # update X_p
+
+            # construct S_1
+            S_1 = S_infinity + (1 - S_infinity) / (1 + E_a / E_star)
+
+            # construct S_2
+            S_2 = S_imm + (1 - S_imm) / (1 + (X_p / X_p_star) ** gamma_P)
+
+            # construct S_p
+            S_p = S_1 * S_2
+
+            # construct lambda
+            _lambda = S_p * E_a
+
+            # increase population age
+            pop_age_days += 1
+
+        return
+
+    def get_mean_body_surface_area(pop_age_days):
+        """
+        We use a crude estimate based on a quintic regression
+        based on a fit on data points:
+        1. 3kg, 50cm=0.20m^2 at birth (age=0)
+        2. 24kg, 127cm=0.92m^2 at (age=10years)
+        3. 60kg, 160cm=1.63m^2 at (age=20years, limit)
+        4. 0.45m^2 at age=1
[…]

### 2019-07-04 code/toy_model/toy/toy_parasite.py

@@ -1,0 +1,498 @@
+# %%
+import pyprob
+from pyprob import Model
+from pyprob.distributions import Poisson, Categorical, Normal, TruncatedNormal, Beta, Uniform, LogNormal
+import torch as th
+import torch
+import math
+
+import matplotlib.pyplot as plt
+# % matplotlib
+# inline
+
+# http://www.ajtmh.org/docserver/fulltext/14761645/75/2_suppl/0750019.pdf?expires=1559041639&id=id&accname=guest&checksum=63084BE5CD7723A6025084787A900053
+
+
+class ToyMalaria(Model):
+    def __init__(self, pop_size=1050, likelihood_stddev=0.1, EIR=None):
+        super().__init__(name='Toy Malaria')
+        self.pop_size = pop_size
+        self.likelihood_stddev = likelihood_stddev
+        self.t_max = 365 * 2 # not totally clear
+        self.days_per_step = 1
+        #self.EIR = th.FloatTensor(self.t_max * self.days_per_step).zero_() + 67.0/365.0
+        self.EIR = th.FloatTensor(self.t_max * self.days_per_step).zero_() + 5.5 / 365.0 # Matsari, intervention phase
+        self.EIR_background = th.FloatTensor([67.0/365.0])
+        self.A_max = 1.63
+        self.prevalence_bins_upper = torch.FloatTensor([0.246746962983009,
+                                             0.727755459545533,
+                                             1.44479521753634,
+                                             2.88435310594495,
+                                             4.81735572357366,
+                                             7.82460584536638,
+                                             12.087165127225,
+                                             16.8889599400757,
+                                             24.2654245103456,
+                                             33.7166980261767,
+                                             48.1735572357366,
+                                             67.3111742713562]) * 365
+        self._eval = False
+        self.n_age_bands = 7
+        self.only_warmup = False
+
+    def forward(self, _eval=False, use_pyprob=True):
+        '''
+        Parameters:
+            i - Individual
+            t - Individual at time t
+            a - age group of individual at time t
+            A(a(i,t)) - Mean body suface area
+
+        Return:
+            E_a(i,t) - The estimate at each time, for an individual i of
+            number of infected bites recieved, adjusted for age.
+
+        Latent variables:
+
+
+
+        '''
+
+        age_bands_lower = [0,1,5,9,19,29,44][:self.n_age_bands]
+        age_bands_upper = [1,4,8,18,28,43,67][:self.n_age_bands]
+
+        if use_pyprob:
+            # Generate initial population age distribution
[…]

### 2019-07-04 code/toy_model/toy/toy_tempy.py

@@ -1,0 +1,311 @@
+# %%
+import pyprob
+from pyprob import Model
+from pyprob.distributions import Poisson, Categorical, Normal, TruncatedNormal, Beta, Uniform
+import torch as th
+import torch
+
+import matplotlib.pyplot as plt
+# % matplotlib
+# inline
+
+
+
+
+class ToyMalaria(Model):
+    def __init__(self, pop_size=1050, likelihood_stddev=0.1, EIR=None):
+        super().__init__(name='Toy Malaria')
+        self.pop_size = pop_size
+        self.likelihood_stddev = likelihood_stddev
+        self.t_max = 70
+        self.days_per_step = 1
+        #self.EIR = th.FloatTensor(self.t_max * self.days_per_step).zero_() + 67.0/365.0
+        self.EIR = th.FloatTensor(self.t_max * self.days_per_step).zero_() + 5.5 / 365.0 # Matsari, intervention phase
+        self.EIR_background = th.FloatTensor([67.0/365.0])
+        self.A_max = 1.63
+        self.prevalence_bins_upper = torch.FloatTensor([0.246746962983009,
+                                             0.727755459545533,
+                                             1.44479521753634,
+                                             2.88435310594495,
+                                             4.81735572357366,
+                                             7.82460584536638,
+                                             12.087165127225,
+                                             16.8889599400757,
+                                             24.2654245103456,
+                                             33.7166980261767,
+                                             48.1735572357366,
+                                             67.3111742713562]) * 365
+        self._eval = False
+
+    def forward(self,
+                _eval=False,
+                use_pyprob=True):
+        '''
+        Parameters:
+            i - Individual
+            t - Individual at time t
+            a - age group of individual at time t
+            A(a(i,t)) - Mean body suface area
+
+        Return:
+            E_a(i,t) - The estimate at each time, for an individual i of
+            number of infected bites recieved, adjusted for age.
+
+        Surivival function
+            S_p(i,t) - Probability that the progeny of each incoculation survive
+            to give rise to a patent blood stage infection
+
+        Force of infection
+            lambda(i,t) = S_p(i,t) * E_a(i.t)
+
+        Pre-erythocytic challenge
+            X_p = \int_{t-a)^{t} E_a(i,t) d\tau
+
[…]

### 2019-07-04 openmalaria_nochanges/openmalaria/model/mon/reporting.h

@@ -238,7 +238,7 @@ void reportStatMSF( Measure measure, size_t species, double val );
 /// Report some value (floating point) by genotype to the current survey.
 void reportStatMSGF( Measure measure, size_t species, size_t genotype, double val );
 
-/// Query whether an output measure is used.
+/// Sample whether an output measure is used.
 /// This function is not fast, so it is recommended to cache the result.
 bool isUsedM( Measure measure );
 

### 2019-07-04 paper/models/simple_fc_nn_files/analytics.js

@@ -1,16 +1,16 @@
 (function(){var k=this,l=function(a,b){a=a.split(".");var c=k;a[0]in c||"undefined"==typeof c.execScript||c.execScript("var "+a[0]);for(var d;a.length&&(d=a.shift());)a.length||void 0===b?c=c[d]&&c[d]!==Object.prototype[d]?c[d]:c[d]={}:c[d]=b};var m=function(a,b){for(var c in b)b.hasOwnProperty(c)&&(a[c]=b[c])},n=function(a){for(var b in a)if(a.hasOwnProperty(b))return!0;return!1};var q=/^(?:(?:https?|mailto|ftp):|[^:/?#]*(?:[/?#]|$))/i;var r=window,t=document,u=function(a,b){t.addEventListener?t.addEventListener(a,b,!1):t.attachEvent&&t.attachEvent("on"+a,b)};var v=/:[0-9]+$/,x=function(a,b){b&&(b=String(b).toLowerCase());if("protocol"===b||"port"===b)a.protocol=w(a.protocol)||w(r.location.protocol);"port"===b?a.port=String(Number(a.hostname?a.port:r.location.port)||("http"==a.protocol?80:"https"==a.protocol?443:"")):"host"===b&&(a.hostname=(a.hostname||r.location.hostname).replace(v,"").toLowerCase());var c=w(a.protocol);b&&(b=String(b).toLowerCase());switch(b){case "url_no_fragment":b="";a&&a.href&&(b=a.href.indexOf("#"),b=0>b?a.href:a.href.substr(0,
-b));a=b;break;case "protocol":a=c;break;case "host":a=a.hostname.replace(v,"").toLowerCase();break;case "port":a=String(Number(a.port)||("http"==c?80:"https"==c?443:""));break;case "path":a="/"==a.pathname.substr(0,1)?a.pathname:"/"+a.pathname;a=a.split("/");a:if(b=a[a.length-1],c=[],Array.prototype.indexOf)b=c.indexOf(b),b="number"==typeof b?b:-1;else{for(var d=0;d<c.length;d++)if(c[d]===b){b=d;break a}b=-1}0<=b&&(a[a.length-1]="");a=a.join("/");break;case "query":a=a.search.replace("?","");break;
[…]

### 2019-07-05 paper/learning/notes.tex
added more notes

@@
 \usepackage{wrapfig}
+\usepackage{amsmath}
 
@@
 
-One challenge in performing inference in such models is the combinatorial increase in traces, \emph{paths}
-that can be taken within the given probabilistic model. This is because each member of the population has 
-a bounded, but large number of decisions to make. In addition to this, each member of the population can interact
-with other members of the population. To model a single population member particle-based methods such as 
-sequential monte carlo (SMC) provide an ability to explore multiple paths
+One challenge in performing inference in such models is the combinatorial increase in \emph{paths}
+that can be taken within the given probabilistic model.
+This is because each member of the population has 
+a bounded, but large number of decisions to make. 
+In addition to this, each member of the population can interact
+with other members of the population.
+To model a single population member quantum-based methods leveraging 
+linear superposition and entanglement provide an ability to explore multiple paths simultaneously maintaining 
+the whole program, \emph{state}, in one iteration.
+ Unfortunately, analytically writing the functional 
+form of the state is, in this instance, impossible.   
+The usability of current classical inference methods in this context is also problematic. 
+However, by combining hijacking and amortizing parts of the simulator, in particular rejection sampling 
+loops, we can not only correctly perform inference in such models, as the rejection sampling loops are 
+now replaced with learnt functions, but we can also perform inference much more efficiently. 
+This is because the rejection sampling loops are now replaced with learnt functions which enables us to correctly
+construct the density and as they are amortized evaluation is cheap and only has to be performed once per call.
+This both reduces memory consumption and running time.
 
-As updates are made sequentially MCMC methods typically are not-well suited this type of recusrive
-estimation problem. As each time we get a new data point $y_{t+1}$ we have to run new MCMC simulations 
+As updates are made sequentially MCMC methods typically are not-well suited this type of recursive
+estimation problem. 
+As each time we get a new data point $y_{t+1}$ we have to run new MCMC simulations 
 for $P(x_{0:t_1}| y_{0:t+1})$, we also cannot directly reuse the samples $\{x^{i}_{0:t}\}$. 
@@
 to sample from $(x_{0:t_1}| y_{0:t+1})$.
+ However, as we can only run the simulator $\texttt{forward()}$ we cannot return back to a previous state and re-run. So adapting existing inference techniques 
+ is the only viable option.
 
@@
 \subsection{Prior based sampling}
+In prior based sampling we take our existing program and look directly at the product of all the raw sample calls, denoted $f_{a_i}(x_i | \eta)$ and 
+construct a proposal that is is dependent on those. In this sampling scheme our proposal is defined as:
+\begin{equation}
+  q(x_{1:n_{x}}) = 
+  \begin{cases}
+    \prod_{j=1}^{n_x} f_{a_i}(x_i | \eta) \hspace{0.1cm} \texttt{if} \hspace{0.1cm} \mathcal{B}(x_{1:n_{x})=1 \\
+    0
+  \end{cases}
+\end{equation}
+However, this makes use of no conditioning and is not sufficient for performing inference in
+complex, population-based simulations.
+
+\section{Non-prior Based Sampling}
 

### 2019-07-05 paper/learning/thesystem.tex
added more notes

@@
  \node[state,
-  right of=PyProb,
-  yshift = -0cm,
+  right of=System,
+  yshift = 1cm,
   node distance=5cm,
@@
   \textbf{Program density}\\
-  \parbox{4cm}{P(\mathcal{T}= $x_{1:n_x}| \lambda)$}
+  \parbox{5cm}{P(\mathcal{T}= $x_{1:n_x}| \lambda)$\\
+  \{$x_{1:n_{x}}\}$,$w= \prod^{n_{y}}_{i=1}g_b_i (y_i | \psi_i)}$}
  \end{tabular}
@@
   \node[state,
-  right of=System,
-  yshift = 0cm,
+  above of=EPC,
+  yshift = -1cm,
   node distance=5cm,
@@
   \textbf{Program Output}\\
-  \parbox{4cm}{$\Omega = \{P(x_{1:n_{x}}), w\}$}
+  \parbox{5cm}{$\Omega = \{P_{r}(x_{1:n_{x}}), w\}$}
  \end{tabular}
@@
  (OBSERVE)     	edge[bend right=20] node[anchor=south,above]{$g_{b_{i}},\phi_i , y_i $} (System)
- (PyProb)       	edge [bend left=20] node[anchor=south,above]{$g_{b_{i}}(y_i, \phi_i)$} (EPC)
- (PyProb)       edge [bend left=9] node[anchor=west, above]{$x_{i}$ \emph{returned}} (OBSERVE)
- (System)       	edge [bend right=10]                                          (EPC)
+ (System)       	edge [bend right=20] node[anchor=south,below]{$g_{b_{i}}(y_i, \phi_i)$} (EPC)
+ (PyProb)       edge [bend left=9] node[anchor=west, above]{$x_{i}$} (Sample)
+ (PyProb)       edge [bend left=9] node[anchor=west, above]{$x_{i}$} (EPC)
+%  (System)       	edge [bend left=10]                                          (EPC)
  (EPC)              edge                                                    (OUTPUT)
- (System)  	edge   node[anchor=north,above]{\emph{Control}} (OBSERVE);
+ (System)  	edge [bend right=10]  node[anchor=north,above]{\emph{Control}} (OBSERVE)
+ (Sample)   edge [bend left= 20] (OUTPUT) ;
 %  (System)  	edge[loop below]    node[anchor=north,below]{$SC_n\neq 0$} (System)

### 2019-07-06 paper/learning/notes.tex
added more notes

@@
   \begin{cases}
-    \prod_{j=1}^{n_x} f_{a_i}(x_i | \eta) \hspace{0.1cm} \texttt{if} \hspace{0.1cm} \mathcal{B}(x_{1:n_{x})=1 \\
+    \prod_{i=1}^{n_x} f_{a_i}(x_i | \eta) \hspace{0.1cm} \texttt{if} \hspace{0.1cm} \mathcal{B}(x_{1:n_{x}})=1 \\
     0
@@
 \end{equation}
+where $\mathcal{B}(\cdot)$ is satisfied by the simulator and $f_a_i(\cdot)$ represents a
+raw random sample, $\eta$ is a subset of the variables in-scope at the point of sampling
+and $n_{x}$ is the number of latent variables.
+ This is true in all inference engines used in this hijacking framework.
 However, this makes use of no conditioning and is not sufficient for performing inference in
@@
 \section{Non-prior Based Sampling}
+In non-prior based sampling we utilise conditioning too, which enables 
+us to explore spaces more efficiently. 
+In the current hijacking set-up, we
+can only do conditioning at the end of each \texttt{forward()} run of the simulator. 
+This does inhibit certain inference engines and makes aspects of performing inference 
+within this setting more complicated. 
+Nonetheless, there are several inference schemes 
+that we can exploit and utilise. 
+We shall now give an Introduction to each of these schemes.
 
 \subsection{Importance Sampling}
+In the generalised probabilistic programming setting an importance sampling scheme over the 
+program raw samples and observations can be defined as follows:
+\begin{equation}
+  \label{eq:posterior}
+  p(\mathcal{T}=x_{1:n_x}| \lambda)=
+  \begin{cases}
+  \prod_{i=1}^{n_x} f_{a_i}(x_i | \eta) \prod_{j=1}^{n_y} g_{b_j}(y_j | \phi) \texttt{if} \mathbal{B(x_{1:n_x})} = 1\\
+  0
+  \end{cases}
+\end{equation}
+
+where the observe statements are defined by $g_{b_j}(.)$, $\phi$ plays the same role as 
+$\eta$ and $n_y$ represents the number of observations.
+The expectation over the whole program can be defined as: 
+\begin{equation}
+\mathbb{E}_{p(\Omega | \lambda)}[h(\Omega)] = \mathbb{E}_{p(x_{1:n_x} | \lambda)}[h(P_r(x_{1:n_x}))]
+\end{equation}
+\textbf{Note: It is a little weird how we define this, as the the program output $\Omega$ should
+also include the product of the weights}
+
+where $r$ represents a run of the program, $P_{r}(\cdot)$ is the entirety of the program output,
+$h(\cdot)$ is just a deterministic function.
+
+As the importance sampler updates the weights according to the following formula:
+\begin{equation}
+  w( x_{1:n_x}) = \prod^{n_x}_{i=1}\frac{f_a_i(x_i | \eta_i)}{q_i(x_i |f_a_i, \eta_i)} \prod^{n_y}_{j=1} g_b_j(y_j | \phi_j)
+\end{equation}
+
+When the number of observations is fixed, the combined weight is equal to $w(x_{1:n_x}) = \prod^{n_y}_{j=1} w_j $. 
+To generate the weight up to each observe we can define $K(j)$ to be a function that returns the least $i$ before hitting an observe $j$.
+This enables us to write the $j$-th weight as follows:
+\begin{equation}
+  w_j(x1:K(j)) = g_b_j(y_j | \psi_j) \prod^{K(j)}_{K(j-1)+1} v_i
+\end{equation}
+where we define $v_j = \frac{f_a_j(x_j | \eta_j)}{q_j(x_j |f_a_j, \eta_j)}$ \textbf{Note: Ask Tom why the proposal is also dependent
+on $f_a_i$. }
+
+Taking these definitions we can construct a target density over the 
+latents that is proportional to the products of \texttt{sample} and \texttt{observe}
+statements \textbf{Note: Add update posterior here}
+Thus, given the proposal $q(x_{1:K(t)})$ we define the importance weight as:
+\begin{multline}
+  w(x_{1:t}) = \prod^{t}_{j=1} w_j (x_{1:K(j)}) = \\
+   \frac{p_t(x_{1:K(t)})}{q(x_{1:K(t)})} =
+ \begin{cases}
+  \prod^{t}_{j=1}g_{b_j}(y_j | w_i)\prod^K(j)_{K(j-1)+1} v_i \hspace{0.1cm} \texttt{if} \hspace{0.1cm} B_t(x_1:K(t))=1 \\
+  0
+ \end{cases}
+\end{multline}
 
 \subsection{Sequential Monte Carlo}
+Given multiple observations, we can ake the inference more computationally efficient 
+by employing sequential monte carlo (SMC), although as we can only condition at the end 
+of one full run of the simulation, computational speed increases are minimal.
+SMC allows for conditioning on multiple observations, but the number of observations must
+be fixed. \textbf{Note: If we could condition on observations within the simulator, we 
+could get dramatic speed increases over standard importance sampling.}
+
+\begin{algorithm}
+    \caption{SMC for Population Based Simulators - well any general program - not new}
+    \label{alg:euclid}
+    \begin{algorithmic}[1] % The number tells where the line numbering should start
+        \Procedure{SMC}{$T,M$,\textbf{x}, \textbf{w},\textbf{v}, \textbf{g}}
+          \For{$t = 1:T$}
+            \For{$m = 1:M$}
+              \State $w^{m} = 1  $
+              \For{$k = K(t-1)+1:K(t)$}
+                \State $x^m_k \sim q_k(x^m_k) f^m_{a_k,\eta^m_{j}}$
+                \State $w^m_t \gets w^t_m v_j^m(x_k^m)$
+                \EndFor
+            \EndFor
+          \State $w^m_t \gets w^m_t g_bt^m(y_t^m | \psi^m_t)$
+          \State $\{x^{1:m}_{1:t}\}$ $\gets$ \texttt{resample}( $\{x^{1:m}_{1:t}\}$,  $\{w^{1:m}_{1:t}\}$ )
+        \EndFor
+        \EndProcedure
+    \end{algorithmic}
+\end{algorithm}
 
@@
 
+Due to the current set-up performing Random-walk Metropolis Hastings (RMH) is not feasible 
+as the rejection sampling blocks within the simulator mean that we cannot correctly perform RMH
+as we do not have a functional form for all parts of the simulator \textbf{Note: Re Word}
 \bibliographystyle{plain}

### 2019-07-08 paper/learning/notes.tex
generated some nice figures

@@
 \usepackage{wrapfig}
-\usepackage{amsmath}
-
-\usetikzlibrary{arrows,automata}
+\usepackage{amsmath,amsthm}
+\usepackage{amssymb}
+\tikzset{>=latex}
+\usetikzlibrary{bayesnet}
+\usetikzlibrary{arrows,shapes,automata}
 \usetikzlibrary{positioning}
@@
 as we do not have a functional form for all parts of the simulator \textbf{Note: Re Word}
+
+\section{An Example: Modelling Parasite Densities in Simulation}
+
+\subsection{The Model}
+
+\begin{figure}[ht!]
+  \centering
+  \label{fig:parasite}
+  \includegraphics[width=0.5\textwidth]{parasite_model.pdf}
+  \caption{The DAG of the Parasite model.}
+\end{figure}
+
+\begin{figure}[ht!]
+  \label{fig:amortized_sampling}
+  \includegraphics[width=0.5\textwidth,height=0.5\textheight]{amoritized_rejection.pdf}
+  \caption{Using PyProb we can locate all regions of inefficient sampling 
+  and replace those points in the code with learnt functions that represent
+  the distributions of the rejection sampling statements. }
+\end{figure}
 \bibliographystyle{plain}

### 2019-07-09 paper/learning/notes.tex
updated

@@
 \end{equation}
-where $\mathcal{B}(\cdot)$ is satisfied by the simulator and $f_a_i(\cdot)$ represents a
+where $\mathcal{B}(\cdot)$ is satisfied by the simulator and $f_{a_i}(\cdot)$ represents a
 raw random sample, $\eta$ is a subset of the variables in-scope at the point of sampling
@@
   \begin{cases}
-  \prod_{i=1}^{n_x} f_{a_i}(x_i | \eta) \prod_{j=1}^{n_y} g_{b_j}(y_j | \phi) \texttt{if} \mathbal{B(x_{1:n_x})} = 1\\
+  \prod_{i=1}^{n_x} f_{a_i}(x_i | \eta) \prod_{j=1}^{n_y} g_{b_j}(y_j | \phi) \hspace{0.1cm} \texttt{if} \mathcal{B}(x_{1:n_x}) = 1\\
   0
@@
 \begin{equation}
-  w( x_{1:n_x}) = \prod^{n_x}_{i=1}\frac{f_a_i(x_i | \eta_i)}{q_i(x_i |f_a_i, \eta_i)} \prod^{n_y}_{j=1} g_b_j(y_j | \phi_j)
+  w(x_{1:n_x}) = \prod^{n_x}_{i=1}\frac{f_{a_i}(x_i | \eta_i)}{q_i(x_i |f_{a_i}, \eta_i)} \prod^{n_y}_{j=1} g_{b_j}(y_j | \phi_j)
 \end{equation}
@@
 \begin{equation}
-  w_j(x1:K(j)) = g_b_j(y_j | \psi_j) \prod^{K(j)}_{K(j-1)+1} v_i
+  w_j(x1:K(j)) = g_{b_j}(y_j | \psi_j) \prod^{K(j)}_{K(j-1)+1} v_i
 \end{equation}
-where we define $v_j = \frac{f_a_j(x_j | \eta_j)}{q_j(x_j |f_a_j, \eta_j)}$ \textbf{Note: Ask Tom why the proposal is also dependent
-on $f_a_i$. }
+where we define $v_j = \frac{f_{a_j}(x_j | \eta_j)}{q_j(x_j |f_{a_j}, \eta_j)}$ \textbf{Note: Ask Tom why the proposal is also dependent
+on $f_{a_i}$. }
 
@@
               \For{$k = K(t-1)+1:K(t)$}
-                \State $x^m_k \sim q_k(x^m_k) f^m_{a_k,\eta^m_{j}}$
+                \State $x^m_k \sim q_k(x^m_k | f_{a^{m}_{k}},\eta^{m}_{j})$
                 \State $w^m_t \gets w^t_m v_j^m(x_k^m)$
@@
             \EndFor
-          \State $w^m_t \gets w^m_t g_bt^m(y_t^m | \psi^m_t)$
+          \State $w^m_t \gets w^m_t g_{b_{t}^m}(y_t^m | \psi^m_t)$
           \State $\{x^{1:m}_{1:t}\}$ $\gets$ \texttt{resample}( $\{x^{1:m}_{1:t}\}$,  $\{w^{1:m}_{1:t}\}$ )
@@
 
+The parasite model by Smith el al.~\cite{smith2006relationship} is a simple stochastic 
+model for determining the number of infections within a given population demographic.
+The models hypothesised in ~\cite{smith2006relationship} were found, 
+after extensive testing and development, to not characterise well the number 
+of infections among different demographics, especially those under 15 years old. 
+By using probabilistic programming we can immediately understand 
+the expressiveness of the proposed model and can exact important insights for 
+extending the proposed models, which increases ones ability to develop 
+improved models to estimate parasite densities, given 
+our observations of the seasonal entomological inoculation rate (EIR).
+In addition to this we can greatly reduce the computational costs, whilst
+improving model comprehension and expressivity.  
+Although a statistical model is not formally defined within~\cite{smith2006relationship}
+we define the unknown \emph{disease dynamics} parameters with a set of latent variables$\{x_{i}\}^{N}$
+and construct the model as follows:
+
+\begin{align}
+  \mu_i &= N \\ % pop size
+  % disease dynamics
+  S_{\infty} &\sim \mathcal{N}(\mu_\infty, \sigma^{2}_\infty) \\
+  E_{crit} &\sim \mathcal{N}(\mu_{crit}, \sigma^{2}_{crit}) \\
+  S_{imm} &\sim \mathcal{N}(\mu_{imm}, \sigma^{2}_{imm}) \\
+  X_{p_{crit}} &\sim \mathcal{N}(\mu_{p_{crit}}, \sigma^{2}_{p_{crit}}) \\
+  \gamma_p &\sim \mathcal{N}(\mu_{p}, \sigma^{2}_{p}) \\
+  \rho_i & \sim \mathcal{U}(\mathbb{L}_i , \mathbb{U}_i) \\
+  % integration for parasite densities in different demographics
+  for \hspace{0.1cm} t &= 1:t_{max}: \\
+  E_a &= \frac{E_{{max}_t}}{A_{max}} \times h_{\mu_{SA}}() \\
+  X_p &= E_a + X_p \\
+  S_1 &= S_{\infty} + \frac{(1 - S_\infty) }{1 + \frac{E_a}{E_{crit}}} \\
+  S_2 &= S_{imm} + \frac{1 - S_{imm}}{1 + (\frac{X_p}{X_p_{crit}})^{\gamma_p}}\\
+  %Probability of survival function
+  S_p &= S_1 \times S_2 \\
+  \lambda &= S_p \times E_a \\
+  % Number of infections generated in unit time per individual sampled from poisson
+  h &\sim Pos(\lambda) \\
+  \texttt{if} \hspace{0.1cm} h > 0: \\
+  \hspace{2.3cm} n_{infect} = 1
+  % % end for - add prevelance construction below
+  % \mathcal{P}rev &= 
+\end{align}
+
 \begin{figure}[ht!]
@@
 
-\begin{figure}[ht!]
+\begin{figure}[hb!]
   \label{fig:amortized_sampling}
@@
 \end{figure}
+
 \bibliographystyle{plain}

### 2019-07-11 paper/learning/notes.tex
updated

@@
 This is because the rejection sampling loops are now replaced with learnt functions which enables us to correctly
-construct the density and as they are amortized evaluation is cheap and only has to be performed once per call.
-This both reduces memory consumption and running time.
+construct the density and as they are amortized, evaluation is cheap and only has to be performed once per call.
+This significantly reduces memory consumption and running time.
 
-As updates are made sequentially MCMC methods typically are not-well suited this type of recursive
+Without the amortization procedure, as updates are made sequentially MCMC methods are typically not-well suited this type of recursive
 estimation problem. 
@@
 for $P(x_{0:t_1}| y_{0:t+1})$, we also cannot directly reuse the samples $\{x^{i}_{0:t}\}$. 
-Likewise, importance sampling is not well suited either, as we cannot reuse the samples and weights for a time $t$ $\{x^{i}_{0:t}, \tilde{w}^{i}_{t}\}^{N}_{i=1}$ 
-to sample from $(x_{0:t_1}| y_{0:t+1})$.
- However, as we can only run the simulator $\texttt{forward()}$ we cannot return back to a previous state and re-run. So adapting existing inference techniques 
- is the only viable option.
+However, as we can only run the simulator $\texttt{forward()}$ we cannot return back to a previous state and re-run. 
+So adapting existing inference techniques is the only viable option. 
+MCMC methods can be effective with a well designed amortized inference procedure (\textbf{Note: This is not guaranteed}).
+Non-MCMC based methods such as importance sampling (IS) and sequential monte carlo (SMC), may also be
+sensible inference engines, although we shall discuss the limitations within the simulator context in the proceeding sections.
+
+However, to implement this in practice we are required to do the following:
+ \begin{itemize}
+    \item Create a sensible training procedure
+    \item Have 2 additional functions implemented within pyprob
+    \begin{enumerate} \item 1) That stores only the necessary raw samples from the desired rejection sampling addresses to be used for training.
+                       \item 2) A function at \emph{test time} that replaces those given addresses with a learnt surrogate.
+    \end{enumerate}
+    \item Finer grained control over the actual latent variables of interest, to avoid unnecessary memory consumption and ensure that inference problems remain tractable.
+  \end{itemize}
+There is no doubt that implementing some fo these features will be more invasive to the underlying simulator code than previously
+just hijacking the raw-random number draws, but, it would enable us to attack difficulties in the inference procedures that are
+currently not amenable to the current approach.
 
@@
 \end{equation}
-where we define $v_j = \frac{f_{a_j}(x_j | \eta_j)}{q_j(x_j |f_{a_j}, \eta_j)}$ \textbf{Note: Ask Tom why the proposal is also dependent
-on $f_{a_i}$. }
+where we define $v_j = \frac{f_{a_j}(x_j | \eta_j)}{q_j(x_j |f_{a_j}, \eta_j)}$
 
@@
 latents that is proportional to the products of \texttt{sample} and \texttt{observe}
-statements \textbf{Note: Add update posterior here}
+statements \textbf{Note: Add update posterior here}.
 Thus, given the proposal $q(x_{1:K(t)})$ we define the importance weight as:
@@
 \end{multline}
+
+One problem with importance sampling is in its ability to scale and condition upon multiple observations. |
+Choosing a \emph{good} prior remains challenging.
 
@@
 
+To implement SMC in the current set-up we round require a forking method and the
+
 \subsection{Random-walk Metropolis Hastings}
@@
 as the rejection sampling blocks within the simulator mean that we cannot correctly perform RMH
-as we do not have a functional form for all parts of the simulator \textbf{Note: Re Word}
+as we do not have a functional form for all parts of the simulator to construct a target density
+that we can evaluate. \textbf{Note:
+Ask tom what:  one can sometimes achieve improvements by instead using the type of the distribution object 
+associated with $x_m$ (what does this first part mean?) 
+ to automatically construct a valid random walk proposal that allows for improved
+ hill climbing behavior on the individual updates. }.
 
 \section{An Example: Modelling Parasite Densities in Simulation}
-
-\subsection{The Model}
-
+% \begin{wrapfigure}{r}{0.4\textwidth}
+  \begin{figure}[h!]
+    \centering
+    \label{fig:parasite}
+    \includegraphics[width=0.4\textwidth]{parasite_model.pdf}
+    \caption{The DAG of the Parasite model.}
+  % \end{wrapfigure}
+  \end{figure}
 The parasite model by Smith el al.~\cite{smith2006relationship} is a simple stochastic 
@@
   h &\sim Pos(\lambda) \\
-  \texttt{if} \hspace{0.1cm} h > 0: \\
-  \hspace{2.3cm} n_{infect} = 1
+  \texttt{if} &\hspace{0.1cm} h > 0: \\
+  \hspace{2.3cm} &n_{infect} = 1 \\
+  \text{TODO: add prevelance construction below}
   % % end for - add prevelance construction below
@@
 
-\begin{figure}[ht!]
-  \centering
-  \label{fig:parasite}
-  \includegraphics[width=0.5\textwidth]{parasite_model.pdf}
-  \caption{The DAG of the Parasite model.}
-\end{figure}
+\textbf{TODO: Add some of the plots generated from this model and an accompanying analysis to them.}
 
-\begin{figure}[hb!]
+\begin{wrapfigure}{l}{0.5\textwidth}
   \label{fig:amortized_sampling}
   \includegraphics[width=0.5\textwidth,height=0.5\textheight]{amoritized_rejection.pdf}
+  \vspace{-50pt}
   \caption{Using PyProb we can locate all regions of inefficient sampling 
   and replace those points in the code with learnt functions that represent
-  the distributions of the rejection sampling statements. }
-\end{figure}
+  the distributions of the rejection sampling statements. $Z_{i}$ represents the marginal and 
+  by learning surrogate functions for each of the rejection sampling loops.
+  In doing so we can learn approximate raw sample distributions $f(x_{i} | \eta)$ for each
+  rejection sampling loop. This means that we can evaluate the function at each part in the simulator, enabling
+  us to construct a target density that can be evaluated when using MCMC inference schemes such as RMH.
+    }
+  \end{wrapfigure}
 
+\section{Additional TODOs}
+
+\begin{itemize}
+\item Before implementing such tools into the simulator, spend time looking for simpler stochastic simulators 
+that can be hijacked but have target densities that can be constructed analytically. 
+\item If we are not utilizing the additional information stored in the trace, which is only used during the inference compilation
+training procedure, then we do not need it all, as for importance sampling we only require the weights after each \texttt{forward()}
+update. 
+\item If we cannot find any pre-existing simulators that are amenable to this approach then, we need to create our own, with
+a rejection sampling loop, that can be exploited by the new machinery. This would enable us to debug, provide a simple example 
+for the paper and understand the feasibility of the entire approach. 
+\end{itemize}
 \bibliographystyle{plain}

### 2019-07-11 paper/learning/thesystem.tex
updated

@@
 
+\begin{figure}
 \begin{tikzpicture}[->,>=stealth']
@@
  anchor=center] (Sample) 
- {\begin{tabular}{l}{
+{%
+\begin{tabular}{l} 	% content
   \textbf{Sample}\\
@@
   \textbf{Program density}\\
-  \parbox{5cm}{P(\mathcal{T}= $x_{1:n_x}| \lambda)$\\
-  \{$x_{1:n_{x}}\}$,$w= \prod^{n_{y}}_{i=1}g_b_i (y_i | \psi_i)}$}
+  \parbox{5cm}{P($\mathcal{T}= x_{1:n_x}| \lambda)$\\
+  \{$x_{1:n_{x}}\}$,$w= \prod^{n_{y}}_{i=1} g_{b_i}(y_i | \psi_i)$}
  \end{tabular}
@@
 \end{tikzpicture}
+\caption{An overview of the hijacking system. $x_i$ represent latent variables. $f_{a_i}$ represent raw sample calls,
+$g_{b_j}$ represent conditioning statements. $\mathcal{T}$ corresponds to the program trace. $w$ represents the collective weights. 
+$r$ represents run number. $n_x$ represents the number of latent variables and $n_y$ represents the number of
+observations, which will always be fixed, but is model dependent. $\lambda$ represents all the other variables generated 
+by running the simulator \texttt{.forward()}.
+}
+\label{fig:systemoverview}
+\end{figure}

### 2019-07-19 paper/learning/notes.tex
updated with new tom stuff

@@
 \usepackage{amssymb}
+\usepackage{cancel}
 \tikzset{>=latex}
@@
 model for determining the number of infections within a given population demographic.
-The models hypothesised in ~\cite{smith2006relationship} were found, 
+The models hypothesized in ~\cite{smith2006relationship} were found, 
 after extensive testing and development, to not characterise well the number 
@@
 improving model comprehension and expressivity.
+
+\paragraph{Common features of the models}
+The models used are all discrete time micro-simulations of malaria in humans, 
+based on the published model used in[1], which we refer to as the base model.
+The model for infection of the vector as a function of parasite densities[2] 
+(and hence for the effect of vaccination on onward transmission), the case-management model[3], 
+the model of vaccination by pre-erythrocytic vaccines[4], the models for clinical outcomes and mortality[5,6], 
+and the basic design of the scenarios used for predicting impacts[3,7] all correspond to the previous descriptions,
+ which are accompanied by detailed rationales for the formulations used.
+
+For all models, the number of infections  introduced in individual $i$ in five-day time step $t$, is distributed as:
+\begin{equation}
+  h(i,t) \sim \textit{Pos}(\lambda(i,t))
+\end{equation}
+where the expected number $\lambda(i,t) = S(i,t)E_a(i,t)$ and $S(i,t)$ is the suseptability
+(proportion of inoculations that result in infections) and and $E_{a}(i,t)$ is the expected number of entomological inoculations, adjusted for age and individual factors.
+
+\paragraph{Base model}
 Although a statistical model is not formally defined within~\cite{smith2006relationship}
@@
 
-\begin{align}
-  \mu_i &= N \\ % pop size
-  % disease dynamics
-  S_{\infty} &\sim \mathcal{N}(\mu_\infty, \sigma^{2}_\infty) \\
-  E_{crit} &\sim \mathcal{N}(\mu_{crit}, \sigma^{2}_{crit}) \\
-  S_{imm} &\sim \mathcal{N}(\mu_{imm}, \sigma^{2}_{imm}) \\
-  X_{p_{crit}} &\sim \mathcal{N}(\mu_{p_{crit}}, \sigma^{2}_{p_{crit}}) \\
-  \gamma_p &\sim \mathcal{N}(\mu_{p}, \sigma^{2}_{p}) \\
-  \rho_i & \sim \mathcal{U}(\mathbb{L}_i , \mathbb{U}_i) \\
-  % integration for parasite densities in different demographics
-  for \hspace{0.1cm} t &= 1:t_{max}: \\
-  E_a &= \frac{E_{{max}_t}}{A_{max}} \times h_{\mu_{SA}}() \\
-  X_p &= E_a + X_p \\
-  S_1 &= S_{\infty} + \frac{(1 - S_\infty) }{1 + \frac{E_a}{E_{crit}}} \\
-  S_2 &= S_{imm} + \frac{1 - S_{imm}}{1 + (\frac{X_p}{X_p_{crit}})^{\gamma_p}}\\
-  %Probability of survival function
-  S_p &= S_1 \times S_2 \\
-  \lambda &= S_p \times E_a \\
-  % Number of infections generated in unit time per individual sampled from poisson
-  h &\sim Pos(\lambda) \\
-  \texttt{if} &\hspace{0.1cm} h > 0: \\
-  \hspace{2.3cm} &n_{infect} = 1 \\
-  \text{TODO: add prevelance construction below}
-  % % end for - add prevelance construction below
-  % \mathcal{P}rev &= 
-\end{align}
+In the base model $E_{a}(i,t) = E_{max}(t) \frac{A(a(i,t))}{A_{max}}$ where $E_{max}(t)$ refers to the usual measure of the EIR computed from human bait collections on adults $A(\cdot)$ and , the availability to mosquitoes,
+ is assumed to be proportional to average body surface area and to depend only on age $a(i,t)$.
+The function used to compute $A(a(i,t))$ increases with age up to age 20 years where it reaches a value of $A_{max}$.
+
+To capture observed relationships between infection and measured entomological exposure, the susceptibility is then:
+\begin{equation}
+  S(i,t) = (S_{\infty} + \frac{1 - S_{\infty}}{1 + \frac{E_a(i,t)}{E^{*}}})(S_{imm} + \frac{1 - S_{imm}}{1 + (\frac{X_p(i,t)}{X_{p}^{*}})^{\gamma_p}})
+\end{equation}
+where $S_{imm}, X^{*}_{p} E^*, \gamma_p, S_\infty$ are latent variables
+and $X_p(i,t) = \int^{t}_{-a(i,t)} E_{a}(i,\tau) d\tau$
+Fitting this model to data of~\cite{beier1994plasmodium} using their method learns the values $S_{\infty} = 0.049$ and $E^* = 0.032$ 
+inoculations/person-night. 
+The model for each individual infection $j$ in host $i$ comprises a time series of parasite densities.
+  In a naïve host the expectation of the logarithm of the density of the infection at time point $\tau$
+in the course of the infection, $\ln(y_0(i,j,\tau))$, is taken from a statistical description of parasite densities
+in malaria therapy patients~\cite{maire2006model}. 
+In the presence of previous exposure and co-infection, the expected log density for each concurrent infection is: 
+\begin{equation}
+  E(\ln(y(i,j,\tau))) = D_{y}(i,t)D_{m}(i,t)D_{h}(i,t)\ln(y_0(i,j,\tau)) + \ln(\frac{D_x}{M(i,t)} + 1 - D_x))
+\end{equation}
+where $M(i,t)$ is the total multiplicity of infection and 
+\begin{equation}
+  D_{y}(i,t) = \frac{1}{1 + \frac{X_y(i,j,t)}{X^{*}_{y}}}
+\end{equation}
+where $X_y(i,j,t) = \sum^{t}_{t-a}Y(i,t) - \sum^{t}_{t_{0,j}} y(i,j,\tau)$ (note that a continuous time approximation to this
+is given in the original publications~\cite{smith2006mathematical,maire2006model}
+and hence measures the cumulative parasite load, and  \textbf{TODO: To finish}
+  % \begin{align}
+%   \mu_i &= N \\ % pop size
+%   % disease dynamics
+%   S_{\infty} &\sim \mathcal{N}(\mu_\infty, \sigma^{2}_\infty) \\
+%   E_{crit} &\sim \mathcal{N}(\mu_{crit}, \sigma^{2}_{crit}) \\
+%   S_{imm} &\sim \mathcal{N}(\mu_{imm}, \sigma^{2}_{imm}) \\
+%   X_{p_{crit}} &\sim \mathcal{N}(\mu_{p_{crit}}, \sigma^{2}_{p_{crit}}) \\
+%   \gamma_p &\sim \mathcal{N}(\mu_{p}, \sigma^{2}_{p}) \\
+%   \rho_i & \sim \mathcal{U}(\mathbb{L}_i , \mathbb{U}_i) \\
+%   % integration for parasite densities in different demographics
+%   for \hspace{0.1cm} t &= 1:t_{max}: \\
+%   E_a &= \frac{E_{{max}_t}}{A_{max}} \times h_{\mu_{SA}}() \\
+%   X_p &= E_a + X_p \\
+%   S_1 &= S_{\infty} + \frac{(1 - S_\infty) }{1 + \frac{E_a}{E_{crit}}} \\
+%   S_2 &= S_{imm} + \frac{1 - S_{imm}}{1 + (\frac{X_p}{X_p_{crit}})^{\gamma_p}}\\
+%   %Probability of survival function
+%   S_p &= S_1 \times S_2 \\
+%   \lambda &= S_p \times E_a \\
+%   % Number of infections generated in unit time per individual sampled from poisson
+%   h &\sim Pos(\lambda) \\
+%   \texttt{if} &\hspace{0.1cm} h > 0: \\
+%   \hspace{2.3cm} &n_{infect} = 1 \\
+%   \text{TODO: add prevelance construction below}
+%   % % end for - add prevelance construction below
+%   % \mathcal{P}rev &= 
+% \end{align}
 
@@
 
-\begin{wrapfigure}{l}{0.5\textwidth}
-  \label{fig:amortized_sampling}
-  \includegraphics[width=0.5\textwidth,height=0.5\textheight]{amoritized_rejection.pdf}
-  \vspace{-50pt}
+\section{Amortized Rejection Sampling}
+
+One common problem in traditional simulators, \emph{programs}, outside of the standard probabilistic
+programming framework is that they rely heavily on rejection sampling 
+to generate samples from the target density. Not only is rejection sampling 
+computational inefficient, [especially if criteria in which samples are generated
+has been poorly calibrated] (may leave out this statement as it may make a reviewer
+question why would amortization work any better.), but it creates complications for inference.
+As the rejection sampling loops provide no explicit form of the target density,
+we can not simply evaluate the density and leverage statistically and computationally
+efficient algorithms such as MCMC sampling. This inhibits our ability to transform an arbitrary
+stochastic simulator into a probabilistic programming framework~\cite{baydin2018efficient}. 
+To this end, we derive both a mathematical and engineering based framework to facilitate 
+the amortization of rejection sampling loops in arbitrary simulators, which provides a path
+way to both leverage a larger, more efficient class of inference methods and convert arbitrary
+stochastic simulators into probabilistic programs systems.
+
+\begin{figure}
+  % \begin{minipage}[c]{0.67\textwidth}
+  \includegraphics[]{amoritized_rejection.pdf}
+  % \end{minipage}\hfill
+  % \begin{minipage}[c]{0.3\textwidth}
     \caption{Using PyProb we can locate all regions of inefficient sampling 
@@
     rejection sampling loop. This means that we can evaluate the function at each part in the simulator, enabling
-  us to construct a target density that can be evaluated when using MCMC inference schemes such as RMH.
-    }
-  \end{wrapfigure}
+    us to construct a target density that can be evaluated when using MCMC inference schemes such as RMH.}
+    \label{fig:amortized_sampling}
+  % \end{minipage}
+\end{figure}
+
+\section{Details}
+
+We do this by  Replacing the rejection sampling loops with 
+learnt functions from an amortized inference procedure Figure~\ref{fig:amortized_sampling}.
+This then enables us to replace rejection sampling loops with amortized functions that 
+are quick to evaluate, provide a functional form to the density and significantly reduce 
+memory over-heads, as compared to~\cite{baydin2018efficient}, due to not having to track the 
+thousands of raw sample calls $f_{a_i}$ generated in standard simulators.
+
+Consider the following program, that is representative of a simple stochastic simulator
+and two types of rejection sampler:
+
+\noindent
+\begin{tabular}{|p{3.6cm}|p{3.6cm}|p{3.6cm}|}
+\hline
+Stochastic Simulator  & Rejection Sampler 1 & Rejection Sampler 2 \\
+\hline
+\begin{lstlisting}[mathescape=true]
+    def f():
+      $z_{1} \sim \mathcal{N}(0,1)$
+      $z_{2} \gets R_{1}(z_1, 1)$
+      $z_{3} \sim \mathcal{U}(0,2)$
+      $z_{4} \gets R_{2}(z_2,z_3, 1)$
+    return $g(z_{1:4})$
+\end{lstlisting}&
+\begin{lstlisting}[mathescape=true]
+def $R_1(z_1)$:
+  $x \sim \mathcal{N}(z_1,1)$
+  if $x > 0$:
+    return $x$
+  else:
+    return $R_{1}(z_1)$
+\end{lstlisting}&
+\begin{lstlisting}[mathescape=true]
+def $R_2(z_2, z_3)$:
+  $y \gets 0$
+  while $y < z_2$:
+    $y \sim \mathcal{N}(z_3,1)$
+  end
+  return $y$
+\end{lstlisting}\\
+\hline
+$P(\mathcal{T}= x_{1:n_x} | \lambda)$ & $q_{1}(z_{2} | z_{1})$ & $q_{2}(z_{4} |z_{2},z_{3})$ \\
+\hline
+\end{tabular}
+
+Here the rejection sampling functions take an additional argument $1$, or $0$,
+to specify whether or not we want to collect samples to train, $1$, else replace with 
+the learnt function, $0$. $P$ represents the program density and $q_{1}$ and $q_{2}$ represents 
+the unknown densities that we wish to learn through an amortized approach.
+
+However, in order todo this, we must correctly specify our variational objective.
+
+\section{Constructing the Objective}
+
+The framework we present in this section is generally applicable to stochastic simulator 
+environemnt, where there exists rejection sampling loops.
+
+We start by noting that the prior, defined in the context of the outter samples is 
+defined as: 
+\begin{equation}
+  p(x) = 
+  \begin{cases}
+    \prod^{n_{x}}_{j=1} f_{a_j}(x_j ; \psi_j) if \mathcal{B}(x) = 1 \\
+    0
+  \end{cases}
+\end{equation}
+We start with the information projection we construct the variational objective as follows:
+\begin{align}
+  J(\eta) = \mathbb{E}_{p(x)}[KL(p(x) || q(x; \eta))] \\
+   = \int p(x') [\int p(x)\ln\left(\frac{p(x)}{q(x;\eta)}\right) dx]dx' \\
+  %  \int p(x')dx' \gets 1 \\
+  %  &= \mathbb{E}_{p(x)}[\ln(p(x))]  - \mathbb{E}_{p(x)}[\ln(q(x;\eta))]\\
+  %  &= \ln(1) -  \mathbb{E}_{p(x)}[\ln(q(x;\eta))]\\ % I do not agree with this line
+  %  &= -\mathbb{E}_{p(x)}[\ln(q(x;\eta))] \\
+  %  &= -\mathbb{E}_{p(x)}[\sum^{n_x}_{j=1}\ln(q_{a_j}(x_j;\eta))] \\
+   \text{Let } a_{j} \in \{1, \ldots, L\} , S_{R} \subset \{1, \ldots, L\} \\
+   = \cancelto{\text{no dependence on } \eta}{\mathbb{E}_{p(x)}[\sum^{n_x}_{j=1,a_j \notin S_{R}}\ln(p(x))]}-\mathbb{E}_{p(x)}[\sum^{n_x}_{j=1,a_j \in S_{R}}\ln(q_{a_j}(x_j;\eta))]\\
+\end{align}
+
+Thus, to define the objective for each subproblem $r  \in S_{R}$ as:
+\begin{equation}
+  J_{r}(\eta_r) = -\mathbb{E}_{p(x)}[\sum^{n_x}_{j=1}\mathbb{I}(a_j = r)\ln(q_r(x_j;\eta_r, \phi_j))]
+\end{equation}
+
+where during training we extract samples from each forward run and train each rejection sampler
+separately.
 
@@
 \end{itemize}
+
+\bibliography{refs}
 \bibliographystyle{plain}
-\bibliography{refs}
 

### 2019-07-23 code/amortized_rs/nsf-master/experiments/vae.py
updated

@@ -1,0 +1,562 @@
+import argparse
+import json
+import numpy as np
+import os
+import torch
+
+import data as data_
+import nn as nn_
+import utils
+import vae
+
+from functools import partial
+from matplotlib import pyplot as plt
+from tensorboardX import SummaryWriter
+from time import sleep
+from torch import nn, optim
+from torch.nn import functional as F
+from torch.utils import data
+from torch.utils.data.sampler import SubsetRandomSampler
+from torchvision import datasets, transforms as tvtransforms
+from torchvision.transforms import functional as tvF
+from torchvision.utils import make_grid
+from tqdm import tqdm
+
+from experiments import cutils
+from nde import distributions as distributions_, flows, transforms
+
+parser = argparse.ArgumentParser()
+
+parser.add_argument('--num_runs', type=int, default=1,
+                    help='Number of runs with different seeds to do.')
+# data
+parser.add_argument('--dataset_name', type=str, default='emnist',
+                    choices=['mnist', 'emnist', 'omniglot', 'fashion-mnist'],
+                    help='Which dataset to use.')
+parser.add_argument('--batch_size', type=int, default=256,
+                    help='Batch size for training.')
+
+# optimization
+parser.add_argument('--num_training_steps', type=int, default=150000,
+                    help='Total number of training steps.')
+parser.add_argument('--learning_rate', type=float, default=5e-4,
+                    help='Learning rate.')
+parser.add_argument('--kl_multiplier_schedule', type=str, default='linear',
+                    choices=['constant', 'linear'],
+                    help='Kind of schedule to use for KL multiplier.')
+parser.add_argument('--kl_warmup_fraction', type=float, default=0.1,
+                    help='Length of training to do KL warmup.')
+parser.add_argument('--kl_multiplier_initial', type=float, default=0.5,
+                    help='Initial value for KL multiplier.')
+parser.add_argument('--kl_multiplier_max', type=float, default=1,
+                    help='Max value for KL multiplier.')
+
+# features size
+parser.add_argument('--latent_features', type=int, default=32,
+                    help='Number of latent features to use.')
+parser.add_argument('--context_features', type=int, default=64,
+                    help='Number of context features computed by encoder.')
+
+# prior choices
+parser.add_argument('--prior_type', type=str, default='rq-autoregressive',
+                    choices=['standard-normal', 'affine-coupling',
+                             'rq-coupling', 'affine-autoregressive',
+                             'rq-autoregressive'],
[…]

### 2019-07-23 paper/documents/paper/main.tex
updated

\documentclass[]{scrartcl}
\usepackage[round]{natbib}
\usepackage{proof}
\usepackage{listings}
\usepackage{color}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{amssymb}
\usepackage{dsfont}
\usepackage{pgf}
\usepackage{subcaption}
\usepackage{hyperref}

\usetikzlibrary{fit,positioning}
%opening
\title{Interpretable Inference: Probabilistic Programming for Policy Making}
\author{Bradley Gram-hansen}

\begin{document}

	\maketitle

	\begin{abstract}

		Machine learning has stated many successes in the realms of computer vision, data analysis and automation, however, it has not had the same successes in the
		social, or the physical sciences. This, in part, is due to the lack of interpretability that is provided by present day models. Interpretability here represents the ability to understand the decision making process, not to simply understand how uncertain we are in an outcome. 
		In the presented work, we demonstrate an interpretable machine learning framework for generic-stochastic-forward-simulator and demonstrate how we can arrive optimal decisions that can be \emph{scrutinized}. In particular, in expending the work of [Gunes et al.], we show how we can hijack a Malaria epidemiology simulator, and interlace both inference and interpretability, to understand how Malaria will spread given a series of field observations. In addition to this, we combine together 
	\end{abstract}

	\section{Outline}

	The goal is to combine the work of \citet{2017vanden} on neural discrete representation, Generalized Hamiltonian Monte Carlo ( and look at the two pre-requisites \citep{hoffman2017learning} and \citep{sohl2014hamiltonian}) and VAE-s to construct the most advance MCMC sampler of all time. Look at Deep NN as Gaussian Processes \citet{lee2017deep}

	Look into Deep Generative models see \href{http://www.cs.toronto.edu/~slwang/generative_model.pdf}{here}, \href{http://www.mit.edu/~9.520/spring10/Classes/class19_DBN_2010.pdf}{here}, \href{https://www.slideshare.net/mlreview/tutorial-on-deep-generative-models}{here} and \href{https://www.cs.cmu.edu/~rsalakhu/papers/annrev.pdf}{here} for tutorials.

	\section{Variational Autoencoders}
	\input{vae.tex}

	\section{Transformations of Constrained Variables}
	\input{trans.tex}

	\newpage
	\bibliographystyle{apalike}
	\bibliography{refs}
\end{document}

### 2019-07-24 paper/learning/notes.tex
updated

@@
 is given in the original publications~\cite{smith2006mathematical,maire2006model}
-and hence measures the cumulative parasite load, and  \textbf{TODO: To finish}
+and hence measures the cumulative parasite load, where
+\begin{equation}
+  D_{y}(i,t) = \frac{1}{1 + \frac{X_{h}(i,t)}{X^{*}_h}}
+\end{equation}
+where $X_h(i,t) = \sum^{t}_{t-a}h(i,\tau) - 1$, the number of inoculations since birth, excluding
+the one under consideration, which measures the inocula experienced by the host
+up to the time point under consideration. 
+\begin{equation}
+  D_{m}(i,t) = 1 - \alpha_m \exp(-\frac{0.693a(i,t)}{a^{*}_m})
+\end{equation}
+which measures the effect of maternal immunity. 
+$X^{*}_{y}$, $X^{*}_h$, $D_x$, $a^*_{m}$ and $\alpha_m$ are essentially 
+latent variables to be learnt given the field data.
+
+\paragraph{Mass action model of the force of infection}
+The base model assumes that susceptibility is a sigmoidal function of the exposure, which conflicts
+with the usual mass action assumption for infectious disease transmission.
+However, if entomological exposure is over-dispersed, 
+infection-exposure relationships similar to those observed in the field 
+can be reconciled with mass action principles, thus providing explanation
+for the observed non-linear relationship between infection and entomological exposure
+and hence a more justifiable representation of the infection process than that used
+in the base model~\cite{smith2008estimation}. 
+We simulate this by including random variation in the availability of the human host to mosquitoes.
+This variation comprises an individual-specific component which is captured by defining
+for each simulated human a baseline availability, $A_b(i)$, which is sampled at birth using:
+\begin{equation}
+  \log(A_b(i)) \sim \mathcal{N}(-\frac{\sigma_b^{2}}{2}, \sigma^{2}_b)
+\end{equation}
+
+so that $A_b(i)$ is log-normally distributed with arithmetic mean 1. 
+The age adhusted EIR at time $t$, adjusted to the individual is then:
+\begin{equation}
+  E_{b}(i,t) = E_{max}(t)A_b(t)\frac{A(a(i,t))}{A_{max}}
+\end{equation}
+Additional log normal variation is introduced at each time step in the simulation 
+to make the expected number of entomological inoculations a function
+both of the individual and of log-normal noise measured by the
+variance parameter, $\sigma^2_n$ :
+\begin{equation}
+  \ln(E_{a}(i,t)) \sim Normal(\ln(E_{b}(i,t) - \frac{\sigma^2_n, \sigma^2_n}))
+\end{equation}
+The susceptibility in these models is independent of EIR, i.e.:
+\begin{equation}
+  S(i,t( = S_{\infty}(S_{imm} + \frac{1 - S_{imm}}{1 + (\frac{X_p(i,t)}{X_{p}^{*}})^{\gamma_p}})
+\end{equation}
+
+Three different parameterisations, XML files, of this model were considered in [8] (E0063,R0065 and R0068).
+In each case, the overall variability in $E_{a}(i,t)$ was constrained to the same total,
+estimated by fitting to the data of~\cite{beier1994plasmodium}.
+A probabilistic programming approach to determining these paramteters would
+have provided a way to leverage state of the art inference algorithms that 
+enable conditioning on arbitrary models. 
+The parameters learnt from the original approach taken by the Malaria community 
+generated values of $\sigma_b^2 = 0.5, 0.3$ and $0.05$ for each model respectively, corresponding
+to $\sigma^2_n = 0.187, 0.567$ and $0.704$. 
+\textbf{TODO: To finish - Add prob prog example + plots}
+
+\paragraph{Models of decay of natural immunity}
+\textbf{To complete}
   % \begin{align}
@@
 
+\section{Density estimation using Normalising flows}
+Notes from models that define an explicit,  
+tractable density are highly effective, 
+because they permit the use of an optimization algorithm directly on the log-likelihood of the training data.
+However,  the  family  of  models  that provide  a  tractable  density  is  limited,
+with  different  families  having  different disadvantage.
+The rejection sampling loops, if proposed correctly, provide a way of generating 
+samples from some density, but do not provide a way of constructing the target density (\textbf{Bradley: Ask Tom, is this true?}). 
+Within event and population-based simulators, the rejection sampling loops have been designed with care,
+to ensure that they are representative to the events that they are simulating. So we can, with confidence,
+assume that they are representative of some density. 
+However, to utilise powerful inference engines, we must be able to construct the 
+full density of the program.
+
+One recent technique for density estimation, in the amortized inference setting,
+is normalising flows combined with variation inference~\cite{rezende2015variational}.
+Flow-based methods work by leveraging a simple property of differential geometry applied 
+to random variables. 
+That being the change of variables.
+Let $f(\epsilon)$ be a differentiable transform of noise $\epsilon \sim \pi(\epsilon)$, which models 
+some process to generate some data $y = f(\epsilon)$. 
+Then the probability density of $y$ under the flow is
+obtained by a change of variables: 
+\begin{equation}
+p(y) = \pi(f^{-1}(y)) \left|\det \frac{\partial f^{-1}}{\partial y}\right|
+\end{equation}
+
 \section{Constructing the Objective}
@@
 We start with the information projection we construct the variational objective as follows:
-\begin{align}
-  J(\eta) = \mathbb{E}_{p(x)}[KL(p(x) || q(x; \eta))] \\
-   = \int p(x') [\int p(x)\ln\left(\frac{p(x)}{q(x;\eta)}\right) dx]dx' \\
+\begin{align*}
+  J(\eta) = KL(p(x) || q(x; \eta)) \\
+   = \int p(x)\ln\left(\frac{p(x)}{q(x;\eta)}\right) dx \\
   %  \int p(x')dx' \gets 1 \\
@@
    = \cancelto{\text{no dependence on } \eta}{\mathbb{E}_{p(x)}[\sum^{n_x}_{j=1,a_j \notin S_{R}}\ln(p(x))]}-\mathbb{E}_{p(x)}[\sum^{n_x}_{j=1,a_j \in S_{R}}\ln(q_{a_j}(x_j;\eta))]\\
-\end{align}
+\end{align*}
 

### 2019-07-24 paper/learning/notes.tex
updated

@@
 
+\begin{figure}
+  \centering
+  \includegraphics{rs_plots.png}
+  \caption{ Simulated densities for rejection sampling block 1, \emph{blue} and rejection
+  sampling block 2, \emph{orange}.}
+  \label{fig:rs_simulated}
+\end{figure}
 However, in order todo this, we must correctly specify our variational objective.
+\section{ An overview of Amortized Inference schemes}
 
-\section{Density estimation using Normalising flows}
+\subsection{Density estimation using Normalising flows}
 Notes from models that define an explicit,  
@@
 
+where the $y$'s are generated from the rejection sampling loops.
+
+\begin{itemize}
+  \item Do we need to know the base distribution that the RS block is sampling from? i.e there maybe be multiple $\pi(\epsilon)$ in each 
+  RS block. 
+  \item Is it right to say here that $y$ is the output of the RS block, we do not know $f$ here. 
+  \item We need to learn $f(.)$
+\end{itemize}
+
 \section{Constructing the Objective}
@@
 The framework we present in this section is generally applicable to stochastic simulator 
-environemnt, where there exists rejection sampling loops.
+environment, where there exists rejection sampling loops.
 
-We start by noting that the prior, defined in the context of the outter samples is 
+We start by noting that the prior, defined in the context of the outer samples is 
 defined as: 

### 2019-07-25 paper/learning/notes.tex
updated

@@
 
+\begin{itemize}
+  \item Whilst the variational posterior we choose as our proposal has a functional form 
+  and can be sampled, the functional form of the rejection sampling loops is not explicitly known,
+  which leads to problems when calculating the $KL$-divergence. 
+  \item There are two approaches that we can take to calculating some divergence metric. 
+  \item 1) Using something like a non-parametric divergence estimator that makes use of Renyi and $L_{2}$ divergences (see \cite{basseville2013divergence} for details)
+  \item 2) Use some form of kernel density to estimation to approximate the functional form of the rejection sampling 
+  block, which can be used in place. From which we learn a variational approximation characterized by a trained network that we can sample from quickly. 
+  \item An issue with approach 2) is that if I can approximate the density in functional form aren't there more efficient methods that we can use? (\textbf{Ask tom})
+  \item Or, if assuming that the rejection sampling block can be arbitrary dimension, are we first learning a flow to represent that density and then learn a variational approximation of something that we can easily sample from.
+  \item But, this is very much the idea behind normalising flows, take a simple distribution that you can sample and then map it to the ``complicated'' density. 
+  \item Stein Variational inference may be quite nice to use here~\cite{feng2017learning}, but it does make explicit assumptions in requiring the densities to be differentiable. 
+  \item  For an overview of all variational methods, including the use of different divergence metrics see here ~\cite{8588399}
+\end{itemize}
+  \subsection{Using variants of VAEs}
+
 \subsection{Density estimation using Normalising flows}
@@
 That being the change of variables.
-Let $f(\epsilon)$ be a differentiable transform of noise $\epsilon \sim \pi(\epsilon)$, which models 
-some process to generate some data $y = f(\epsilon)$. 
-Then the probability density of $y$ under the flow is
-obtained by a change of variables: 
+Let $z \sim r(z)$ where $r(z)$ represents the posterior defined by the rejection
+sampling loop, and define $\pi$ to be a smooth and invertible function.
+When $\pi$ acts $z$ we transform the random variable from one domain, to another. 
+Thus, if $z\prime = \pi(z)$ then we can construct the density of  
+the probability density of the new random variable $z\prime$ under the flow initiated by $\pi$: 
 \begin{equation}
-p(y) = \pi(f^{-1}(y)) \left|\det \frac{\partial f^{-1}}{\partial y}\right|
+r(z\prime) = r(\pi^{-1}(z\prime)) \left|\det \frac{\partial \pi^{-1}}{\partial z'}\right| = r(\pi^{-1}(z\prime)) \left|\det \frac{\partial \pi}{\partial z\prime}\right|^{-1} 
 \end{equation}
-
+However, since 
 where the $y$'s are generated from the rejection sampling loops.

### 2019-07-25 paper/learning/notes.tex
updated

@@
 \end{equation}
-However, since 
-where the $y$'s are generated from the rejection sampling loops.
 
 \begin{itemize}
-  \item Do we need to know the base distribution that the RS block is sampling from? i.e there maybe be multiple $\pi(\epsilon)$ in each 
+  \item Do we need to know the base distribution that the RS block is sampling from? i.e there maybe be multiple $r(z)$ in each 
   RS block. 
-  \item Is it right to say here that $y$ is the output of the RS block, we do not know $f$ here. 
-  \item We need to learn $f(.)$
+  \item Is it right to say here that $z\prime$ is the output of the RS block, we do not know $\pi$ here, or is $r(z)$ the raw sample $f_{a_i}$ and $\pi$ the distribution specified by the rejection sampling loop?
+  \item We need to learn $\pi(.)$
 \end{itemize}

### 2019-07-30 paper/learning/notes.tex
updated

@@
 \textbf{To complete}
-  % \begin{align}
-%   \mu_i &= N \\ % pop size
-%   % disease dynamics
-%   S_{\infty} &\sim \mathcal{N}(\mu_\infty, \sigma^{2}_\infty) \\
-%   E_{crit} &\sim \mathcal{N}(\mu_{crit}, \sigma^{2}_{crit}) \\
-%   S_{imm} &\sim \mathcal{N}(\mu_{imm}, \sigma^{2}_{imm}) \\
-%   X_{p_{crit}} &\sim \mathcal{N}(\mu_{p_{crit}}, \sigma^{2}_{p_{crit}}) \\
-%   \gamma_p &\sim \mathcal{N}(\mu_{p}, \sigma^{2}_{p}) \\
-%   \rho_i & \sim \mathcal{U}(\mathbb{L}_i , \mathbb{U}_i) \\
-%   % integration for parasite densities in different demographics
-%   for \hspace{0.1cm} t &= 1:t_{max}: \\
-%   E_a &= \frac{E_{{max}_t}}{A_{max}} \times h_{\mu_{SA}}() \\
-%   X_p &= E_a + X_p \\
-%   S_1 &= S_{\infty} + \frac{(1 - S_\infty) }{1 + \frac{E_a}{E_{crit}}} \\
-%   S_2 &= S_{imm} + \frac{1 - S_{imm}}{1 + (\frac{X_p}{X_p_{crit}})^{\gamma_p}}\\
-%   %Probability of survival function
-%   S_p &= S_1 \times S_2 \\
-%   \lambda &= S_p \times E_a \\
-%   % Number of infections generated in unit time per individual sampled from poisson
-%   h &\sim Pos(\lambda) \\
-%   \texttt{if} &\hspace{0.1cm} h > 0: \\
-%   \hspace{2.3cm} &n_{infect} = 1 \\
-%   \text{TODO: add prevelance construction below}
-%   % % end for - add prevelance construction below
-%   % \mathcal{P}rev &= 
-% \end{align}
 
@@
 
+\section{Rejection Sampling}
+
+Rejection samplers are heavily used when designing simulators, due to their ease of use and ability to model 
+any arbitrary distribution on $\mathbb{R}^{D}$, with a sufficiently well designed rejection sampling scheme~\cite{MEYER20083408,martino2012improved,casella} and have 
+their origins in 1940 - 1950's, during the development of other prominent Monte carlo techniques at Los Alamos.
+
+In the standard setting we let $X$ be a random variable whose density is $p(x)$ that we cannot easily 
+sample from, but can evaluate. Using a random variable $Z$, whose density is $g(z)$  that we can efficiently
+sample from, then we can generate a sample $X$ by sampling $Z$ and accept $Z$ with probability
+$p_{accept} = \frac{p(x)}{Mg(x)}$, which we can do until acceptance. 
+M is constant that has a finite bound, $M \in (1, \infty)$, on the likelihood ratio $\frac{p(x)}{g(x)}$ over the 
+support of $X$. $M$ must also satify the following inequality $p(x) \leq M g(x)  \forall x$. 
+This implies that the support of $Z$ must also contained the support of $X$, that is
+$g(x) > 0$ whenever $p(x)>0$. 
+The standard algorithm is defined as follows:
+\begin{algorithm}
+  \small
+   \caption{Rejection Sampling}
+    \begin{algorithmic}\small
+       \State {At iteration $i$ s.t $i \geq 1$}
+       \State $X_{i} \sim g(X_{i})$ 
+       \State $U_{i} \sim \mathcal{U}(0,1)$
+       \If{$U_{i} \leq M \frac{f(X_i)}{g(X_{i})}$}
+       \State{ accept $X_i \sim f$}
+       \Else
+       \State{$i \gets i + 1$, repeat}
+       \EndIf
+\end{algorithmic}
+\end{algorithm}
+
+As useful rejection sampling is, it has many short comings. 
+In particular, to simulate proposals for certain distributions,
+such as truncated normal distributions, regardless of the easy to sample from proposal
+chosen the computational running time of such rejection statements can be shown to have an
+expectation equal to infinity. Thus, the ability to learn and replace such rejection
+sampling blocks with learnt surrogate functions, that are expressive enough to 
+reciprocate the behavior of the rejection sampling block buys the user two things. 
+A) It reduces the computational complexity of the simulator. 
+B) As we now have a direct functional form of the density, we can correctly leverage
+more computationally and sample efficient inference schemes. 
 \section{Details}
 
-We do this by  Replacing the rejection sampling loops with 
+We do this by  Replacing the rejection sampling blocks with 
 learnt functions from an amortized inference procedure Figure~\ref{fig:amortized_sampling}.
@@
 \end{figure}
-However, in order todo this, we must correctly specify our variational objective.
+However, in order to do this, we must correctly specify our variational objective.
+
 \section{ An overview of Amortized Inference schemes}
-
+\textbf{Bradley: Todo: Tom has provided me with all the information to conduct this approach}
 \begin{itemize}
@@
 \end{itemize}
+
+\subsection{Using a simple Neural Network regressor}
+
+As a first approach, we use a regressor to learn a function to map
+between the input / output pairs that are outputted from the rejection
+sampling blocks with in the simulator.
+
+\paragraph{Architecture} For each rejection sampling block we learn a proposal $q(z_1, \ldots, z_i | \eta)$
 \subsection{Using variants of VAEs}
@@
 \bibliographystyle{plain}
+\appendix
+
+\section{The unconditional acceptance probability}
+
+Given a collection of i.i.d random variables $\{X_{i}\}_{i = 1:n}$ with 
+cumulative density function (CDF) $F_{X}$. Then by the Glivenko-Cantelli theorem~\cite{tucker1959generalization} 
+we can define the empirical distribution function for $x \in \mathbb{R}$ as:
+\begin{equation}
+  F_{n}(x) = \frac{1}{n}\sum^{n}_{i=1} \mathbb{I}_{[X_{i}, \infty)}(x)
+\end{equation}
+And for any fixed $x \in \mathbb{R}$, the Strong law of large numbers ensures that 
+$F_{n}(x) \rightarrow F_{X}(x)$ as $n \rightarrow \infty$ as 
+\begin{equation}
+\mathbb{E}[\mathbb{I}_{[X_{i},\infty)}(x)] = 
+ P[\mathbb{I}_{X_{i},\infty)}(x) = 1] = P[X_{i}\leq x] = F_(X)(x)
+\end{equation}
+where $\mathbb{I}_{A}(x)$  is the indicator function for a set $A$.
+
+\begin{align}
+  \text{Prob}(U \leq \frac{p_{X}(z)}{Mg_{Z}(z)} = \mathbb{E}_()
+\end{align}
+\section{Law of total expectations}
+
+If $X$ is a random variable whose expected value is $\mathbb{E}[X]$ and $Z$ is a random variable on the same probability space
+then: \begin{equation}
+  \label{eq:lawoftotalexp}
+  \mathbb{E}[X] = \mathbb{E}_{p(z)}[\mathbb{E}_{p(x|z)}[X | Z]]
+\end{equation}
+That is, the expected value of $X$ given $Z$ is the same as the expectation of $X$.
 

### 2019-07-30 code/amortized_rs/src/main.py
updated

@@ -30,4 +30,16 @@ class DensityEstimatorTrainer():
         self.optmizer = optmizer
         self.steps = 0
 
-    def train(self, ):
+    def train(self, epochs, data):
+        """
+        Method to train the density estimator. 
+
+        :param: epochs :type: int :descrp: Number of epochs to train for.
+        :param: data :type: torch.Tensor  :descrp: The data to pass through the network. 
+
+        :return: Trained model
+        """
+
+        for epoch in range(epochs):
+            epoch_loss = 0
+            for 

### 2019-07-30 code/amortized_rs/src/model.py
notes by jimmy

@@ -6,10 +6,16 @@ from torch.nn import functional as F
 class density_estimator(nn.Module):
     def __init__(self, inputSize, outputSize):
         super(density_estimator, self).__init__()
-        self.linear = torch.nn.Linear(inputSize, outputSize)
+        self.hidden = nn.Sequential(
+            nn.Linear(inputSize, inputSize//2),
+            nn.ReLU(),
+
+        )
+        self.linear_mu = nn.Linear(inputSize, outputSize)
+        self.linear_std = nn.Linear(inputSize, outputSize)
 
     def forward(self, x):
-        out = self.linear(x)
-        return out
+
+        return self.linear_mu(x), nn.Softplus(self.linear_std(x))+1e-6
 
 

### 2019-07-30 code/amortized_rs/src/utils.py

@@ -59,7 +59,7 @@ def gen_samples(total_samples=1000, n_outputs=4, simulator=sim):
     n_samples = int(total_samples/10)
     samples_1 = torch.zeros(n_samples,n_outputs)
     samples_2 = torch.zeros(n_samples,n_outputs)
-    samples_3 =torch.zeros(n_samples,n_outputs)
+    samples_3 = torch.zeros(n_samples,n_outputs)
     samples_4 = torch.zeros(n_samples,n_outputs)
     samples_5 = torch.zeros(n_samples,n_outputs)
     samples_6 = torch.zeros(n_samples,n_outputs)
@@ -82,7 +82,7 @@ def gen_samples(total_samples=1000, n_outputs=4, simulator=sim):
             torch.save(samples_2, 'samples_2_{}.pt'.format(i))
             torch.save(samples_3, 'samples_3_{}.pt'.format(i))
             torch.save(samples_4, 'samples_4_{}.pt'.format(i))
-            torch.save(samples_5,'samples_5_{}.pt'.format(i))
+            torch.save(samples_5, 'samples_5_{}.pt'.format(i))
             torch.save(samples_6, 'samples_6_{}.pt'.format(i))
             torch.save(samples_7, 'samples_7_{}.pt'.format(i))
             torch.save(samples_8, 'samples_8_{}.pt'.format(i))
@@ -103,3 +103,55 @@ def gen_samples(total_samples=1000, n_outputs=4, simulator=sim):
     end = time.time()
     total_time = end - start
     print(' Time taken is {} for {} samples'.format(total_time, n_samples))
+
+from torch.utils.data import Dataset, DataLoader
+class RejectionDataset(Dataset):
+    def __init__(self, split):
+        self.a = torch.load("patha/..")[:100] if split == 'train' else torch.load("path/..")[100:]
+        self.b = torch.load("pathb/..")[:100] if split == 'train' else torch.load("path/..")[100:]
+
+    def __getitem__(self, idx):
+        return self.a[idx], self.b[idx]
+
+from torch import optim
+train_loader = DataLoader(RejectionDataset(split='train'), batch_size=128, shuffle=True)
+test_loader = DataLoader(RejectionDataset(split='test'), batch_size=128, shuffle=True)
+
+epochs = 10
+loss_fn = torch.nn.MSELoss()
+optimizer = optim.Adam(filter(lambda p: p.requires_grad, model.parameters()), lr=args.lr, amsgrad=True)
+
+model.to(device)
+import torch.distributions as dist
+
+
[…]

### 2019-07-30 code/amortized_rs/src/utils.py
loader for constructing n-D data

@@ -104,18 +104,29 @@ def gen_samples(total_samples=1000, n_outputs=4, simulator=sim):
     total_time = end - start
     print(' Time taken is {} for {} samples'.format(total_time, n_samples))
 
+
+
 from torch.utils.data import Dataset, DataLoader
 class RejectionDataset(Dataset):
-    def __init__(self, split):
-        self.a = torch.load("patha/..")[:100] if split == 'train' else torch.load("path/..")[100:]
-        self.b = torch.load("pathb/..")[:100] if split == 'train' else torch.load("path/..")[100:]
+    def __init__(self, split, l_data, train_percentage):
+
+        _a = torch.load("patha/..")
+        n_split = (len(_a)*train_percentage//l_data)*l_data
+
+        self.a = torch.load("patha/..")[:n_split] if split == 'train' else torch.load("path/..")[n_split:]
+        self.b = torch.load("pathb/..")[:n_split] if split == 'train' else torch.load("path/..")[n_split:]
+        perm = torch.randperm(len(self.a))
+        self.a, self.b = self.a[perm], self.b[perm]
+        self.l_data = l_data
 
     def __getitem__(self, idx):
-        return self.a[idx], self.b[idx]
+        start_idx = idx*self.l_data
+        end_idx = start_idx + self.l_data
+        return self.a[start_idx:end_idx], self.b[start_idx:end_idx]
 
 from torch import optim
-train_loader = DataLoader(RejectionDataset(split='train'), batch_size=128, shuffle=True)
-test_loader = DataLoader(RejectionDataset(split='test'), batch_size=128, shuffle=True)
+train_loader = DataLoader(RejectionDataset(split='train', l_data=128, train_percentage=0.8), batch_size=128, shuffle=True)
+test_loader = DataLoader(RejectionDataset(split='test', l_data=128, train_percentage=0.8), batch_size=128, shuffle=True)
 
 epochs = 10
 loss_fn = torch.nn.MSELoss()

### 2019-08-01 paper/learning/notes.tex
minor additions to the text

@@
 In addition to this, each member of the population can interact
-with other members of the population.
-To model a single population member quantum-based methods leveraging 
-linear superposition and entanglement provide an ability to explore multiple paths simultaneously maintaining 
-the whole program, \emph{state}, in one iteration.
- Unfortunately, analytically writing the functional 
-form of the state is, in this instance, impossible.   
-The usability of current classical inference methods in this context is also problematic. 
-However, by combining hijacking and amortizing parts of the simulator, in particular rejection sampling 
-loops, we can not only correctly perform inference in such models, as the rejection sampling loops are 
-now replaced with learnt functions, but we can also perform inference much more efficiently. 
-This is because the rejection sampling loops are now replaced with learnt functions which enables us to correctly
-construct the density and as they are amortized, evaluation is cheap and only has to be performed once per call.
-This significantly reduces memory consumption and running time.
+with other members of the population, which presents challenges for current inference methods.
+Because in many simulated settings we cannot correctly construct the functional form of the density of the 
+desired program. 
+This has spawned a large body of research in recent years to understand how to perform 
+inference in likelihood free methods~\cite{greenberg2019automatic} \textbf{Add normalising flow citations}
+for conditional density estimation. However, whilst many of these methods look to generate a surragote 
+for the entire simulator, we propose amortizing only computational expensive sub-components 
+of the simulator, which we can do in an automated way with probabilistic programming. 
+This enables three things. 
+1) It reduces the loss of information, as when simulators are amortized with in current
+settings we require, potentially infinite samples, to provide a useful learnt representation of the simulator. 
+2) It increases computational efficiency, whilst maintaining expressivity. 
+3) It converts expensive rejection sampling loops into to learnt functions that can be 
+easily sampled from and used to construct a functional form of the likelihood, that enables the 
+target density to be fully constructed, enabling one to leverage a large class of sample effient inference schemes, 
+namely MCMC.
+
+% To model a single population member quantum-based methods leveraging 
+% linear superposition and entanglement provide an ability to explore multiple paths simultaneously maintaining 
+% the whole program, \emph{state}, in one iteration.
+%  Unfortunately, analytically writing the functional 
+% form of the state is, in this instance, impossible.   
+% 
+In particular, we focus on combining the hijacking of simulators via probabilistic programming~\cite{baydin2018efficient} and 
+state-of-the-art amortization schemes for density estimation \textbf{ Do home work on current papers}.
+This enables such a process to happen in automatic way, simplifying existing set-ups, whilst providing
+more expressive modelling capabilities. 
+We are able to demonstrate this by replacing rejection sampling blocks, which are utilised within 
+many prominent simulators to simulate densities that have no well defined cumulative density 
+function, nor no known process to efficiently sample from such a density exists. 
+By amortizing the rejection sampling blocks, during the inference stage when a simulator is run \emph{forword} we 
+can replace such blocks with their leant surrogate, from which we can then construct the entire 
+density of the simulator, which can then be utilized, \emph{correctly}, contrary to ~\cite{baydin2018efficient},
+by sample efficient Markov chain Monte Carlo inference schemes, such as Hamiltonian Monte Carlo~(HMC),
+random-walk Metropolis Hastings~(RMH), Light-weight Metropolis Hastings~(LMH) and many other schemes.
+Whilst also significantly reducing memory consumption and computational running time of 
+both the simulator and of the probabilistic programming system, as evaluating amortized functions 
+is cheap and the reduction in rejection sampling blocks significantly reduces the number of 
+unnecessary latent variables tracked~\cite{baydin2018efficient}.
 
@@
 
-However, to implement this in practice we are required to do the following:
+However, to construct this hybrid procedure requires both substantial engineering contributions 
+and a description of the mathematical framework for these non-standard probabilistic programs. 
+We provide both of these in this paper. As to implement this in practice we are required to do the following:
  \begin{itemize}
-    \item Create a sensible training procedure
-    \item Have 2 additional functions implemented within pyprob
-    \begin{enumerate} \item 1) That stores only the necessary raw samples from the desired rejection sampling addresses to be used for training.
-                       \item 2) A function at \emph{test time} that replaces those given addresses with a learnt surrogate.
+    \item An automated training procedure for any given simulator.
+    \item Extend the protocols of existing probabilistic programming systems that hijack simulators, to enable:
+    \begin{enumerate} \item 1) Only the necessary raw samples from the desired rejection sampling addresses to be collected for training.
+                       \item 2) A function at inference time that replaces those given addresses with a learnt surrogate.
     \end{enumerate}
@@
   \end{itemize}
-There is no doubt that implementing some fo these features will be more invasive to the underlying simulator code than previously
+
+\textbf{There is no doubt that implementing some fo these features will be more invasive to the underlying simulator code than previously
 just hijacking the raw-random number draws, but, it would enable us to attack difficulties in the inference procedures that are
-currently not amenable to the current approach.
+currently not amenable to the current approach.}
 
- \input{thesystem.tex}
 \section{Limitations with different inference engines}
-
-\subsection{Prior based sampling}
-In prior based sampling we take our existing program and look directly at the product of all the raw sample calls, denoted $f_{a_i}(x_i | \eta)$ and 
-construct a proposal that is is dependent on those. In this sampling scheme our proposal is defined as:
-\begin{equation}
-  q(x_{1:n_{x}}) = 
-  \begin{cases}
-    \prod_{i=1}^{n_x} f_{a_i}(x_i | \eta) \hspace{0.1cm} \texttt{if} \hspace{0.1cm} \mathcal{B}(x_{1:n_{x}})=1 \\
-    0
-  \end{cases}
-\end{equation}
-where $\mathcal{B}(\cdot)$ is satisfied by the simulator and $f_{a_i}(\cdot)$ represents a
-raw random sample, $\eta$ is a subset of the variables in-scope at the point of sampling
-and $n_{x}$ is the number of latent variables.
- This is true in all inference engines used in this hijacking framework.
-However, this makes use of no conditioning and is not sufficient for performing inference in
-complex, population-based simulations.
 
@@
 be fixed. \textbf{Note: If we could condition on observations within the simulator, we 
-could get dramatic speed increases over standard importance sampling.}
+could get dramatic speed increases over standard importance sampling.We can actually do this 
+in pyprob, but it requires knowing where to condition. That can be quite challenging 
+for arbitrary code. }
 
@@
  hill climbing behavior on the individual updates. }.
+
+\textbf{TODO: Add some of the plots generated from this model and an accompanying analysis to them.}
+
+\section{Amortized Rejection Sampling}
+
+One common problem in traditional simulators, \emph{programs}, outside of the standard probabilistic
+programming framework is that they rely heavily on rejection sampling 
+to generate samples from the target density. Not only is rejection sampling 
+computational inefficient, [especially if criteria in which samples are generated
+has been poorly calibrated] (may leave out this statement as it may make a reviewer
+question why would amortization work any better.), but it creates complications for inference.
+As the rejection sampling loops provide no explicit form of the target density,
+we can not simply evaluate the density and leverage statistically and computationally
+efficient algorithms such as MCMC sampling. This inhibits our ability to transform an arbitrary
+stochastic simulator into a probabilistic programming framework~\cite{baydin2018efficient}. 
+To this end, we derive both a mathematical and engineering based framework to facilitate 
+the amortization of rejection sampling loops in arbitrary simulators, which provides a path
+way to both leverage a larger, more efficient class of inference methods and convert arbitrary
+stochastic simulators into probabilistic programs systems.
+
+\begin{figure}
+  % \begin{minipage}[c]{0.67\textwidth}
+  \includegraphics[]{amoritized_rejection.pdf}
+  % \end{minipage}\hfill
+  % \begin{minipage}[c]{0.3\textwidth}
+    \caption{Using PyProb we can locate all regions of inefficient sampling 
+    and replace those points in the code with learnt functions that represent
+    the distributions of the rejection sampling statements. $Z_{i}$ represents the marginal and 
+    by learning surrogate functions for each of the rejection sampling loops.
+    In doing so we can learn approximate raw sample distributions $f(x_{i} | \eta)$ for each
+    rejection sampling loop. This means that we can evaluate the function at each part in the simulator, enabling
+    us to construct a target density that can be evaluated when using MCMC inference schemes such as RMH.}
+    \label{fig:amortized_sampling}
+  % \end{minipage}
+\end{figure}
+
+\section{Rejection Sampling}
+
+Rejection samplers are heavily used when designing simulators, due to their ease of use and ability to model 
+any arbitrary distribution on $\mathbb{R}^{D}$, with a sufficiently well designed rejection sampling scheme~\cite{MEYER20083408,martino2012improved,casella} and have 
+their origins in 1940 - 1950's, during the development of other prominent Monte carlo techniques at Los Alamos.
+
+In the standard setting we let $X$ be a random variable whose density is $f(x)$ that we cannot easily 
+sample from, but can evaluate. Using a random variable $Z$, whose density is $g(z)$  that we can efficiently
+sample from, then we can generate a sample $X$ by sampling $Z$ and accept $Z$ with probability
+$p_{accept} = \frac{f(x)}{Mg(x)}$, which we can do until acceptance. 
+M is constant that has a finite bound, $M \in (1, \infty)$, on the likelihood ratio $\frac{f(x)}{g(x)}$ over the 
+support of $X$. $M$ must also satify the following inequality $f(x) \leq M g(x)  \forall x$. 
+This implies that the support of $Z$ must also contained the support of $X$, that is
+$g(x) > 0$ whenever $f(x)>0$. 
+The standard algorithm is defined as follows:
+\begin{algorithm}
+  \small
+   \caption{Rejection Sampling}
+    \begin{algorithmic}\small
+       \State {At iteration $i$ s.t $i \geq 1$}
+       \State $X_{i} \sim g(X_{i})$ 
+       \State $U_{i} \sim \mathcal{U}(0,1)$
+       \If{$U_{i} \leq M \frac{f(X_i)}{g(X_{i})}$}
+       \State{ accept $X_i \sim f$}
+       \Else
+       \State{$i \gets i + 1$, repeat}
+       \EndIf
+\end{algorithmic}
+\end{algorithm}
+
+As useful rejection sampling is, it has many short comings. 
+In particular, to simulate proposals for certain distributions,
+such as truncated normal distributions, regardless of the easy to sample from proposal
+chosen the computational running time of such rejection statements can be shown to have an
+expectation equal to infinity. Thus, the ability to learn and replace such rejection
+sampling blocks with learnt surrogate functions, that are expressive enough to 
+reciprocate the behavior of the rejection sampling block buys the user two things. 
+A) It reduces the computational complexity of the simulator. 
+B) As we now have a direct functional form of the density, we can correctly leverage
+more computationally and sample efficient inference schemes.
+
+\section{Defining the Density and Sub-Densities within Simulators}
+
+To construct the mathematical framework in this setting we employ the notation of~\cite{rainforth2017automating}, 
+where \emph{raw samples} are represented by $ f_{a_i}(x_i | \eta)$ and can
+
+\subsection{Prior based sampling}
+In prior based sampling we take our existing program and look directly at the product of all the raw sample calls, denoted $f_{a_i}(x_i | \eta)$ and 
+construct a proposal that is is dependent on those. In this sampling scheme our proposal is defined as:
+\begin{equation}
+  q(x_{1:n_{x}}) = 
+  \begin{cases}
+    \prod_{i=1}^{n_x} f_{a_i}(x_i | \eta) \hspace{0.1cm} \texttt{if} \hspace{0.1cm} \mathcal{B}(x_{1:n_{x}})=1 \\
+    0
+  \end{cases}
+\end{equation}
+where $\mathcal{B}(\cdot)$ is satisfied by the simulator and $f_{a_i}(\cdot)$ represents a
+raw random sample, $\eta$ is a subset of the variables in-scope at the point of sampling
+and $n_{x}$ is the number of latent variables.
+ This is true in all inference engines used in this hijacking framework.
+However, this makes use of no conditioning and is not sufficient for performing inference in
+complex, population-based simulations.
+
+In order to correctly characterize this problem we first provide a useful diagram demonstrating 
+the density of the system.
+
+ \input{thesystem.tex}
+
+We do this by  Replacing the rejection sampling blocks with 
+learnt functions from an amortized inference procedure Figure~\ref{fig:amortized_sampling}.
+This then enables us to replace rejection sampling loops with amortized functions that 
+are quick to evaluate, provide a functional form to the density and significantly reduce 
+memory over-heads, as compared to~\cite{baydin2018efficient}, due to not having to track the 
+thousands of raw sample calls $f_{a_i}$ generated in standard simulators.
+
+Consider the following program, that is representative of a simple stochastic simulator
+and two types of rejection sampler:
+
+\noindent
+\begin{tabular}{|p{3.6cm}|p{3.6cm}|p{3.6cm}|}
+\hline
+Stochastic Simulator  & Rejection Sampler 1 & Rejection Sampler 2 \\
+\hline
+\begin{lstlisting}[mathescape=true]
+    def f():
+      $z_{1} \sim \mathcal{N}(0,1)$
+      $z_{2} \gets R_{1}(z_1, 1)$
+      $z_{3} \sim \mathcal{U}(0,2)$
+      $z_{4} \gets R_{2}(z_2,z_3, 1)$
+    return $g(z_{1:4})$
+\end{lstlisting}&
+\begin{lstlisting}[mathescape=true]
+def $R_1(z_1)$:
+  $x \sim \mathcal{N}(z_1,1)$
+  if $x > 0$:
+    return $x$
+  else:
+    return $R_{1}(z_1)$
+\end{lstlisting}&
+\begin{lstlisting}[mathescape=true]
+def $R_2(z_2, z_3)$:
+  $y \gets 0$
+  while $y < z_2$:
+    $y \sim \mathcal{N}(z_3,1)$
+  end
+  return $y$
+\end{lstlisting}\\
+\hline
+$P(\mathcal{T}= x_{1:n_x} | \lambda)$ & $q_{1}(z_{2} | z_{1})$ & $q_{2}(z_{4} |z_{2},z_{3})$ \\
+\hline
+\end{tabular}
+
+Here the rejection sampling functions take an additional argument $1$, or $0$,
+to specify whether or not we want to collect samples to train, $1$, else replace with 
+the learnt function, $0$. $P$ represents the program density and $q_{1}$ and $q_{2}$ represents 
+the unknown densities that we wish to learn through an amortized approach.
+
+\begin{figure}
+  \centering
+  \includegraphics{rs_plots.png}
+  \caption{ Simulated densities for rejection sampling block 1, \emph{blue} and rejection
+  sampling block 2, \emph{orange}.}
+  \label{fig:rs_simulated}
+\end{figure}
+However, in order to do this, we must correctly specify our variational objective.
+
+\section{ An overview of Amortized Inference schemes}
+\begin{itemize}
+  \item Whilst the variational posterior we choose as our proposal has a functional form 
+  and can be sampled, the functional form of the rejection sampling loops is not explicitly known,
+  which leads to problems when calculating the $KL$-divergence. 
+  \item There are two approaches that we can take to calculating some divergence metric. 
+  \item 1) Using something like a non-parametric divergence estimator that makes use of Renyi and $L_{2}$ divergences (see \cite{basseville2013divergence} for details)
+  \item 2) Use some form of kernel density to estimation to approximate the functional form of the rejection sampling 
+  block, which can be used in place. From which we learn a variational approximation characterized by a trained network that we can sample from quickly. 
+  \item An issue with approach 2) is that if I can approximate the density in functional form aren't there more efficient methods that we can use? (\textbf{Ask tom})
+  \item Or, if assuming that the rejection sampling block can be arbitrary dimension, are we first learning a flow to represent that density and then learn a variational approximation of something that we can easily sample from.
+  \item But, this is very much the idea behind normalising flows, take a simple distribution that you can sample and then map it to the ``complicated'' density. 
+  \item Stein Variational inference may be quite nice to use here~\cite{feng2017learning}, but it does make explicit assumptions in requiring the densities to be differentiable. 
+  \item  For an overview of all variational methods, including the use of different divergence metrics see here ~\cite{8588399}
+\end{itemize}
+
+New questions to ask Tom
+
+\begin{itemize}
+  \item Isn't what we are doing just conditional density estimation? It seems
+  like there is a lot of prior work here. I.e see this paper here for a good overview ~\cite{greenberg2019automatic}
+  \item It does seem like this field has a large extensive past.
+\end{itemize}
+
+\subsection{Using a simple Neural Network regressor}
+
+As a first approach, we use a regressor to learn a function to map
+between the input / output pairs that are outputted from the rejection
+sampling blocks with in the simulator.
+
+\paragraph{Architecture} For each rejection sampling block we learn a proposal $q(z_1, \ldots, z_i | \eta)$
+\subsection{Using variants of VAEs}
+
+\subsection{Density estimation using Normalising flows}
+Notes from models that define an explicit,  
+tractable density are highly effective, 
+because they permit the use of an optimization algorithm directly on the log-likelihood of the training data.
+However,  the  family  of  models  that provide  a  tractable  density  is  limited,
+with  different  families  having  different disadvantage.
+The rejection sampling loops, if proposed correctly, provide a way of generating 
+samples from some density, but do not provide a way of constructing the target density (\textbf{Bradley: Ask Tom, is this true?}). 
+Within event and population-based simulators, the rejection sampling loops have been designed with care,
+to ensure that they are representative to the events that they are simulating. So we can, with confidence,
+assume that they are representative of some density. 
+However, to utilise powerful inference engines, we must be able to construct the 
+full density of the program.
+
+One recent technique for density estimation, in the amortized inference setting,
+is normalising flows combined with variation inference~\cite{rezende2015variational}.
+Flow-based methods work by leveraging a simple property of differential geometry applied 
+to random variables. 
+That being the change of variables.
+Let $z \sim r(z)$ where $r(z)$ represents the posterior defined by the rejection
+sampling loop, and define $\pi$ to be a smooth and invertible function.
+When $\pi$ acts $z$ we transform the random variable from one domain, to another. 
+Thus, if $z\prime = \pi(z)$ then we can construct the density of  
+the probability density of the new random variable $z\prime$ under the flow initiated by $\pi$: 
+\begin{equation}
+r(z\prime) = r(\pi^{-1}(z\prime)) \left|\det \frac{\partial \pi^{-1}}{\partial z'}\right| = r(\pi^{-1}(z\prime)) \left|\det \frac{\partial \pi}{\partial z\prime}\right|^{-1} 
+\end{equation}
+
+\begin{itemize}
+  \item Do we need to know the base distribution that the RS block is sampling from? i.e there maybe be multiple $r(z)$ in each 
+  RS block. 
+  \item Is it right to say here that $z\prime$ is the output of the RS block, we do not know $\pi$ here, or is $r(z)$ the raw sample $f_{a_i}$ and $\pi$ the distribution specified by the rejection sampling loop?
+  \item We need to learn $\pi(.)$
+\end{itemize}
+
+\section{Constructing the Objective}
+
+The framework we present in this section is generally applicable to stochastic simulator 
+environment, where there exists rejection sampling loops.
+
+We start by noting that the prior, defined in the context of the outer samples is 
+defined as: 
+\begin{equation}
+  p(x) = 
+  \begin{cases}
+    \prod^{n_{x}}_{j=1} f_{a_j}(x_j ; \psi_j) if \mathcal{B}(x) = 1 \\
+    0
+  \end{cases}
+\end{equation}
+We start with the information projection we construct the variational objective as follows:
+\begin{align*}
+  J(\eta) = KL(p(x) || q(x; \eta)) \\
+   = \int p(x)\ln\left(\frac{p(x)}{q(x;\eta)}\right) dx \\
+  %  \int p(x')dx' \gets 1 \\
+  %  &= \mathbb{E}_{p(x)}[\ln(p(x))]  - \mathbb{E}_{p(x)}[\ln(q(x;\eta))]\\
+  %  &= \ln(1) -  \mathbb{E}_{p(x)}[\ln(q(x;\eta))]\\ % I do not agree with this line
+  %  &= -\mathbb{E}_{p(x)}[\ln(q(x;\eta))] \\
+  %  &= -\mathbb{E}_{p(x)}[\sum^{n_x}_{j=1}\ln(q_{a_j}(x_j;\eta))] \\
+   \text{Let } a_{j} \in \{1, \ldots, L\} , S_{R} \subset \{1, \ldots, L\} \\
+   = \cancelto{\text{no dependence on } \eta}{\mathbb{E}_{p(x)}[\sum^{n_x}_{j=1,a_j \notin S_{R}}\ln(p(x))]}-\mathbb{E}_{p(x)}[\sum^{n_x}_{j=1,a_j \in S_{R}}\ln(q_{a_j}(x_j;\eta))]\\
+\end{align*}
+
+Thus, to define the objective for each subproblem $r  \in S_{R}$ as:
+\begin{equation}
+  J_{r}(\eta_r) = -\mathbb{E}_{p(x)}[\sum^{n_x}_{j=1}\mathbb{I}(a_j = r)\ln(q_r(x_j;\eta_r, \phi_j))]
+\end{equation}
+
+where during training we extract samples from each forward run and train each rejection sampler
+separately.
+
+\section{Additional TODOs}
+
+\begin{itemize}
+\item Before implementing such tools into the simulator, spend time looking for simpler stochastic simulators 
+that can be hijacked but have target densities that can be constructed analytically. 
+\item If we are not utilizing the additional information stored in the trace, which is only used during the inference compilation
+training procedure, then we do not need it all, as for importance sampling we only require the weights after each \texttt{forward()}
+update. 
+\item If we cannot find any pre-existing simulators that are amenable to this approach then, we need to create our own, with
+a rejection sampling loop, that can be exploited by the new machinery. This would enable us to debug, provide a simple example 
+for the paper and understand the feasibility of the entire approach. 
+\end{itemize}
+
+\bibliography{refs}
+\bibliographystyle{plain}
+\appendix
+
+\section{The unconditional acceptance probability}
+
+Given a collection of i.i.d random variables $\{X_{i}\}_{i = 1:n}$ with 
+cumulative density function (CDF) $F_{X}$. Then by the Glivenko-Cantelli theorem~\cite{tucker1959generalization} 
+we can define the empirical distribution function for $x \in \mathbb{R}$ as:
+\begin{equation}
+  F_{n}(x) = \frac{1}{n}\sum^{n}_{i=1} \mathbb{I}_{[X_{i}, \infty)}(x)
+\end{equation}
+And for any fixed $x \in \mathbb{R}$, the Strong law of large numbers ensures that 
+$F_{n}(x) \rightarrow F_{X}(x)$ as $n \rightarrow \infty$ as 
+\begin{equation}
+\mathbb{E}[\mathbb{I}_{[X_{i},\infty)}(x)] = 
+ P[\mathbb{I}_{X_{i},\infty)}(x) = 1] = P[X_{i}\leq x] = F_(X)(x)
+\end{equation}
+where $\mathbb{I}_{A}(x)$  is the indicator function for a set $A$.
+
+\begin{align}
+  \text{Prob}(U \leq \frac{p_{X}(z)}{Mg_{Z}(z)} = \mathbb{E}_()
+\end{align}
+\section{Law of total expectations}
+
+If $X$ is a random variable whose expected value is $\mathbb{E}[X]$ and $Z$ is a random variable on the same probability space
+then: \begin{equation}
+  \label{eq:lawoftotalexp}
+  \mathbb{E}[X] = \mathbb{E}_{p(z)}[\mathbb{E}_{p(x|z)}[X | Z]]
+\end{equation}
+That is, the expected value of $X$ given $Z$ is the same as the expectation of $X$.
 
@@
 
-\textbf{TODO: Add some of the plots generated from this model and an accompanying analysis to them.}
-
-\section{Amortized Rejection Sampling}
-
-One common problem in traditional simulators, \emph{programs}, outside of the standard probabilistic
-programming framework is that they rely heavily on rejection sampling 
-to generate samples from the target density. Not only is rejection sampling 
-computational inefficient, [especially if criteria in which samples are generated
-has been poorly calibrated] (may leave out this statement as it may make a reviewer
-question why would amortization work any better.), but it creates complications for inference.
-As the rejection sampling loops provide no explicit form of the target density,
-we can not simply evaluate the density and leverage statistically and computationally
-efficient algorithms such as MCMC sampling. This inhibits our ability to transform an arbitrary
-stochastic simulator into a probabilistic programming framework~\cite{baydin2018efficient}. 
-To this end, we derive both a mathematical and engineering based framework to facilitate 
-the amortization of rejection sampling loops in arbitrary simulators, which provides a path
-way to both leverage a larger, more efficient class of inference methods and convert arbitrary
-stochastic simulators into probabilistic programs systems.
-
-\begin{figure}
-  % \begin{minipage}[c]{0.67\textwidth}
-  \includegraphics[]{amoritized_rejection.pdf}
-  % \end{minipage}\hfill
-  % \begin{minipage}[c]{0.3\textwidth}
-    \caption{Using PyProb we can locate all regions of inefficient sampling 
-    and replace those points in the code with learnt functions that represent
-    the distributions of the rejection sampling statements. $Z_{i}$ represents the marginal and 
-    by learning surrogate functions for each of the rejection sampling loops.
-    In doing so we can learn approximate raw sample distributions $f(x_{i} | \eta)$ for each
-    rejection sampling loop. This means that we can evaluate the function at each part in the simulator, enabling
-    us to construct a target density that can be evaluated when using MCMC inference schemes such as RMH.}
-    \label{fig:amortized_sampling}
-  % \end{minipage}
-\end{figure}
-
-\section{Rejection Sampling}
-
-Rejection samplers are heavily used when designing simulators, due to their ease of use and ability to model 
-any arbitrary distribution on $\mathbb{R}^{D}$, with a sufficiently well designed rejection sampling scheme~\cite{MEYER20083408,martino2012improved,casella} and have 
-their origins in 1940 - 1950's, during the development of other prominent Monte carlo techniques at Los Alamos.
-
-In the standard setting we let $X$ be a random variable whose density is $p(x)$ that we cannot easily 
-sample from, but can evaluate. Using a random variable $Z$, whose density is $g(z)$  that we can efficiently
-sample from, then we can generate a sample $X$ by sampling $Z$ and accept $Z$ with probability
-$p_{accept} = \frac{p(x)}{Mg(x)}$, which we can do until acceptance. 
-M is constant that has a finite bound, $M \in (1, \infty)$, on the likelihood ratio $\frac{p(x)}{g(x)}$ over the 
-support of $X$. $M$ must also satify the following inequality $p(x) \leq M g(x)  \forall x$. 
-This implies that the support of $Z$ must also contained the support of $X$, that is
-$g(x) > 0$ whenever $p(x)>0$. 
-The standard algorithm is defined as follows:
-\begin{algorithm}
-  \small
-   \caption{Rejection Sampling}
-    \begin{algorithmic}\small
-       \State {At iteration $i$ s.t $i \geq 1$}
-       \State $X_{i} \sim g(X_{i})$ 
-       \State $U_{i} \sim \mathcal{U}(0,1)$
-       \If{$U_{i} \leq M \frac{f(X_i)}{g(X_{i})}$}
-       \State{ accept $X_i \sim f$}
-       \Else
-       \State{$i \gets i + 1$, repeat}
-       \EndIf
-\end{algorithmic}
-\end{algorithm}
-
-As useful rejection sampling is, it has many short comings. 
-In particular, to simulate proposals for certain distributions,
-such as truncated normal distributions, regardless of the easy to sample from proposal
-chosen the computational running time of such rejection statements can be shown to have an
-expectation equal to infinity. Thus, the ability to learn and replace such rejection
-sampling blocks with learnt surrogate functions, that are expressive enough to 
-reciprocate the behavior of the rejection sampling block buys the user two things. 
-A) It reduces the computational complexity of the simulator. 
-B) As we now have a direct functional form of the density, we can correctly leverage
-more computationally and sample efficient inference schemes. 
-\section{Details}
-
-We do this by  Replacing the rejection sampling blocks with 
-learnt functions from an amortized inference procedure Figure~\ref{fig:amortized_sampling}.
-This then enables us to replace rejection sampling loops with amortized functions that 
-are quick to evaluate, provide a functional form to the density and significantly reduce 
-memory over-heads, as compared to~\cite{baydin2018efficient}, due to not having to track the 
-thousands of raw sample calls $f_{a_i}$ generated in standard simulators.
-
-Consider the following program, that is representative of a simple stochastic simulator
-and two types of rejection sampler:
-
-\noindent
-\begin{tabular}{|p{3.6cm}|p{3.6cm}|p{3.6cm}|}
-\hline
-Stochastic Simulator  & Rejection Sampler 1 & Rejection Sampler 2 \\
-\hline
-\begin{lstlisting}[mathescape=true]
-    def f():
-      $z_{1} \sim \mathcal{N}(0,1)$
-      $z_{2} \gets R_{1}(z_1, 1)$
-      $z_{3} \sim \mathcal{U}(0,2)$
-      $z_{4} \gets R_{2}(z_2,z_3, 1)$
-    return $g(z_{1:4})$
-\end{lstlisting}&
-\begin{lstlisting}[mathescape=true]
-def $R_1(z_1)$:
-  $x \sim \mathcal{N}(z_1,1)$
-  if $x > 0$:
-    return $x$
-  else:
-    return $R_{1}(z_1)$
-\end{lstlisting}&
-\begin{lstlisting}[mathescape=true]
-def $R_2(z_2, z_3)$:
-  $y \gets 0$
-  while $y < z_2$:
-    $y \sim \mathcal{N}(z_3,1)$
-  end
-  return $y$
-\end{lstlisting}\\
-\hline
-$P(\mathcal{T}= x_{1:n_x} | \lambda)$ & $q_{1}(z_{2} | z_{1})$ & $q_{2}(z_{4} |z_{2},z_{3})$ \\
-\hline
-\end{tabular}
-
-Here the rejection sampling functions take an additional argument $1$, or $0$,
-to specify whether or not we want to collect samples to train, $1$, else replace with 
-the learnt function, $0$. $P$ represents the program density and $q_{1}$ and $q_{2}$ represents 
-the unknown densities that we wish to learn through an amortized approach.
-
-\begin{figure}
-  \centering
-  \includegraphics{rs_plots.png}
-  \caption{ Simulated densities for rejection sampling block 1, \emph{blue} and rejection
-  sampling block 2, \emph{orange}.}
-  \label{fig:rs_simulated}
-\end{figure}
-However, in order to do this, we must correctly specify our variational objective.
-
-\section{ An overview of Amortized Inference schemes}
-\textbf{Bradley: Todo: Tom has provided me with all the information to conduct this approach}
-\begin{itemize}
-  \item Whilst the variational posterior we choose as our proposal has a functional form 
-  and can be sampled, the functional form of the rejection sampling loops is not explicitly known,
-  which leads to problems when calculating the $KL$-divergence. 
-  \item There are two approaches that we can take to calculating some divergence metric. 
-  \item 1) Using something like a non-parametric divergence estimator that makes use of Renyi and $L_{2}$ divergences (see \cite{basseville2013divergence} for details)
-  \item 2) Use some form of kernel density to estimation to approximate the functional form of the rejection sampling 
-  block, which can be used in place. From which we learn a variational approximation characterized by a trained network that we can sample from quickly. 
-  \item An issue with approach 2) is that if I can approximate the density in functional form aren't there more efficient methods that we can use? (\textbf{Ask tom})
-  \item Or, if assuming that the rejection sampling block can be arbitrary dimension, are we first learning a flow to represent that density and then learn a variational approximation of something that we can easily sample from.
-  \item But, this is very much the idea behind normalising flows, take a simple distribution that you can sample and then map it to the ``complicated'' density. 
-  \item Stein Variational inference may be quite nice to use here~\cite{feng2017learning}, but it does make explicit assumptions in requiring the densities to be differentiable. 
-  \item  For an overview of all variational methods, including the use of different divergence metrics see here ~\cite{8588399}
-\end{itemize}
-
-\subsection{Using a simple Neural Network regressor}
-
-As a first approach, we use a regressor to learn a function to map
-between the input / output pairs that are outputted from the rejection
-sampling blocks with in the simulator.
-
-\paragraph{Architecture} For each rejection sampling block we learn a proposal $q(z_1, \ldots, z_i | \eta)$
-\subsection{Using variants of VAEs}
-
-\subsection{Density estimation using Normalising flows}
-Notes from models that define an explicit,  
-tractable density are highly effective, 
-because they permit the use of an optimization algorithm directly on the log-likelihood of the training data.
-However,  the  family  of  models  that provide  a  tractable  density  is  limited,
-with  different  families  having  different disadvantage.
-The rejection sampling loops, if proposed correctly, provide a way of generating 
-samples from some density, but do not provide a way of constructing the target density (\textbf{Bradley: Ask Tom, is this true?}). 
-Within event and population-based simulators, the rejection sampling loops have been designed with care,
-to ensure that they are representative to the events that they are simulating. So we can, with confidence,
-assume that they are representative of some density. 
-However, to utilise powerful inference engines, we must be able to construct the 
-full density of the program.
-
-One recent technique for density estimation, in the amortized inference setting,
-is normalising flows combined with variation inference~\cite{rezende2015variational}.
-Flow-based methods work by leveraging a simple property of differential geometry applied 
-to random variables. 
-That being the change of variables.
-Let $z \sim r(z)$ where $r(z)$ represents the posterior defined by the rejection
-sampling loop, and define $\pi$ to be a smooth and invertible function.
-When $\pi$ acts $z$ we transform the random variable from one domain, to another. 
-Thus, if $z\prime = \pi(z)$ then we can construct the density of  
-the probability density of the new random variable $z\prime$ under the flow initiated by $\pi$: 
-\begin{equation}
-r(z\prime) = r(\pi^{-1}(z\prime)) \left|\det \frac{\partial \pi^{-1}}{\partial z'}\right| = r(\pi^{-1}(z\prime)) \left|\det \frac{\partial \pi}{\partial z\prime}\right|^{-1} 
-\end{equation}
-
-\begin{itemize}
-  \item Do we need to know the base distribution that the RS block is sampling from? i.e there maybe be multiple $r(z)$ in each 
-  RS block. 
-  \item Is it right to say here that $z\prime$ is the output of the RS block, we do not know $\pi$ here, or is $r(z)$ the raw sample $f_{a_i}$ and $\pi$ the distribution specified by the rejection sampling loop?
-  \item We need to learn $\pi(.)$
-\end{itemize}
-
-\section{Constructing the Objective}
-
-The framework we present in this section is generally applicable to stochastic simulator 
-environment, where there exists rejection sampling loops.
-
-We start by noting that the prior, defined in the context of the outer samples is 
-defined as: 
-\begin{equation}
-  p(x) = 
-  \begin{cases}
-    \prod^{n_{x}}_{j=1} f_{a_j}(x_j ; \psi_j) if \mathcal{B}(x) = 1 \\
-    0
-  \end{cases}
-\end{equation}
-We start with the information projection we construct the variational objective as follows:
-\begin{align*}
-  J(\eta) = KL(p(x) || q(x; \eta)) \\
-   = \int p(x)\ln\left(\frac{p(x)}{q(x;\eta)}\right) dx \\
-  %  \int p(x')dx' \gets 1 \\
-  %  &= \mathbb{E}_{p(x)}[\ln(p(x))]  - \mathbb{E}_{p(x)}[\ln(q(x;\eta))]\\
-  %  &= \ln(1) -  \mathbb{E}_{p(x)}[\ln(q(x;\eta))]\\ % I do not agree with this line
-  %  &= -\mathbb{E}_{p(x)}[\ln(q(x;\eta))] \\
-  %  &= -\mathbb{E}_{p(x)}[\sum^{n_x}_{j=1}\ln(q_{a_j}(x_j;\eta))] \\
-   \text{Let } a_{j} \in \{1, \ldots, L\} , S_{R} \subset \{1, \ldots, L\} \\
-   = \cancelto{\text{no dependence on } \eta}{\mathbb{E}_{p(x)}[\sum^{n_x}_{j=1,a_j \notin S_{R}}\ln(p(x))]}-\mathbb{E}_{p(x)}[\sum^{n_x}_{j=1,a_j \in S_{R}}\ln(q_{a_j}(x_j;\eta))]\\
-\end{align*}
-
-Thus, to define the objective for each subproblem $r  \in S_{R}$ as:
-\begin{equation}
-  J_{r}(\eta_r) = -\mathbb{E}_{p(x)}[\sum^{n_x}_{j=1}\mathbb{I}(a_j = r)\ln(q_r(x_j;\eta_r, \phi_j))]
-\end{equation}
-
-where during training we extract samples from each forward run and train each rejection sampler
-separately.
-
-\section{Additional TODOs}
-
-\begin{itemize}
-\item Before implementing such tools into the simulator, spend time looking for simpler stochastic simulators 
-that can be hijacked but have target densities that can be constructed analytically. 
-\item If we are not utilizing the additional information stored in the trace, which is only used during the inference compilation
-training procedure, then we do not need it all, as for importance sampling we only require the weights after each \texttt{forward()}
-update. 
-\item If we cannot find any pre-existing simulators that are amenable to this approach then, we need to create our own, with
-a rejection sampling loop, that can be exploited by the new machinery. This would enable us to debug, provide a simple example 
-for the paper and understand the feasibility of the entire approach. 
-\end{itemize}
-
-\bibliography{refs}
-\bibliographystyle{plain}
-\appendix
-
-\section{The unconditional acceptance probability}
-
-Given a collection of i.i.d random variables $\{X_{i}\}_{i = 1:n}$ with 
-cumulative density function (CDF) $F_{X}$. Then by the Glivenko-Cantelli theorem~\cite{tucker1959generalization} 
-we can define the empirical distribution function for $x \in \mathbb{R}$ as:
-\begin{equation}
-  F_{n}(x) = \frac{1}{n}\sum^{n}_{i=1} \mathbb{I}_{[X_{i}, \infty)}(x)
-\end{equation}
-And for any fixed $x \in \mathbb{R}$, the Strong law of large numbers ensures that 
-$F_{n}(x) \rightarrow F_{X}(x)$ as $n \rightarrow \infty$ as 
-\begin{equation}
-\mathbb{E}[\mathbb{I}_{[X_{i},\infty)}(x)] = 
- P[\mathbb{I}_{X_{i},\infty)}(x) = 1] = P[X_{i}\leq x] = F_(X)(x)
-\end{equation}
-where $\mathbb{I}_{A}(x)$  is the indicator function for a set $A$.
-
-\begin{align}
-  \text{Prob}(U \leq \frac{p_{X}(z)}{Mg_{Z}(z)} = \mathbb{E}_()
-\end{align}
-\section{Law of total expectations}
-
-If $X$ is a random variable whose expected value is $\mathbb{E}[X]$ and $Z$ is a random variable on the same probability space
-then: \begin{equation}
-  \label{eq:lawoftotalexp}
-  \mathbb{E}[X] = \mathbb{E}_{p(z)}[\mathbb{E}_{p(x|z)}[X | Z]]
-\end{equation}
-That is, the expected value of $X$ given $Z$ is the same as the expectation of $X$.
-
 \end{document}

### 2019-08-06 code/amortized_rs/src/main.py
updated

@@ -9,7 +9,7 @@ import torch.distributions as dist
 from torch.multiprocessing import cpu_count
 from utils import RejectionDataset
 from torch.utils.data import DataLoader
-
+from torch import distributions as dist
 # Open config file
 with open(config_path) as config_file:
     config = json.load(config_file)
@@ -18,7 +18,7 @@ with open(config_path) as config_file:
 with open(directory + '/config.json', 'w') as config_file:
     json.dump(config, config_file)
 
-inputs = DataLoader(RejectionDataset(split='train', l_data=128, train_percentage=0.8,fname_test, fname_train, InIndx, OutIndx), batch_size=128, shuffle=True, norma num_workers=cpu_count-2)
+inputs = DataLoader(RejectionDataset(split='train', l_data=128, train_percentage=0.8,fname_test, fname_train, InIndx, OutIndx), batch_size=128, shuffle=True, num_workers=cpu_count-2)
 outputs = DataLoader(RejectionDataset(split='test', l_data=128, test_percentage=0.8,fname_test, fname_train, InIndx, OutIndx), batch_size=128, shuffle=True,num_workers=cpu_count-2)
 
 epochs = 10

### 2019-08-06 code/amortized_rs/src/rejection_samplers.py
updated

@@ -53,7 +53,7 @@ class frepeat():
         while z2 == math.inf:
             z2 = self.R1(self.z1)
         z4 = self.R2(self.z2Prev,self.z3)
-        return torch.tensor([self.z1,z2,self.z3,z4])
+        return torch.tensor([self.z1,z2,self.z2Prev,self.z3,z4])
 
     def R1(z1, i=0):
         i = 0

### 2019-08-06 code/amortized_rs/src/utils.py

@@ -181,13 +181,14 @@ def create_dataset(data=None, PATH=None, batch_size=5, totalSamples=1000, nOutpu
 class RejectionDataset(Dataset):
     def __init__(self, split, l_data, train_percentage, fname_test, fname_train, InIndx, OutIndx):
         
-        _samples = torch.load('/data/'+fname_test)
+        _samples = torch.load('/data/'+fname_train)
         n_split = (len(_samples)*train_percentage//l_data)*l_data
         
             
         self.zinput = _samples[:n_split,InIndx] if split == 'train' else torch.load("data/"+fname_test)[n_split:]
-        self.zoutput = samples[:n_split,OutIndx] if split == 'train' else torch.load("data/"+fname_test)[n_split:]
-        # using shuffle= True does the same job
+        self.zoutput = _samples[:n_split,OutIndx] if split == 'train' else torch.load("data/"+fname_test)[n_split:]
+        # center inputs 
+        self.zinput = (self.zinput - torch.mean(self.zinput, 0).mul(torch.ones(self.zinput.shape))) / torch.std(self.zinput, dim=0).mul(torch.ones(x.shape))
         # perm = torch.randperm(len(self.zinput))
         # self.zinput, self.zoutput = self.zinput[perm], self.zoutput[perm]
         # self.l_data = l_data

### 2019-08-06 code/amortized_rs/memory_src/pytorch_cpp_ext/hogwild.py
better name

@@ -1,0 +1,100 @@
+import torch.multiprocessing as mp
+from model import density_estimator
+from torch.utils.cpp_extension import load
+import torch as th
+from torch import optim
+from torch.distributions import Normal
+from utils import RejectionDataset
+from torch.utils.data import DataLoader
+import os
+import time
+
+amortized_rs = load(name="amortized_rs",
+                    sources=["amortized_rs.cpp"])
+
+def train(model,  N, data_flag, batch_size, load_data=False):
+    # Construct data_loader, optimizer, etc.
+    # for it in range(max_it):
+    model.train()
+    n_samples = batch_size*N
+    # data = th.stack([amortized_rs.f() for _ in range(batch_size)]) # hmm, shouldn't this be the number of training samples, rather than batch_size?
+    # data = th.stack([amortized_rs.f() for _ in range(n_samples)]) # actually this loop needs to create multiple runs of the simulator of each z1, z2,z3 with
+    if load_data = True:
+        data = torch.load('../data/all_batch_samples.pt')
+        data = data[0:n_samples,:]
+    optimizer.zero_grad()
+    count  = 0
+    # no longer need write the batch loop in c - we can do that later
+    def get_batch(data_flag, batch_size, count, load_data=True):
+        if not load_data:
+            data = th.stack([amortized_rs.f() for _ in range(batch_size)]) 
+        if data_flag == 'R1':
+            inR1 = data[count:batch_size+count*batch_size,0]
+            outR1 = data[count:batch_size+count*batch_size,1]
+            count = count + 1
+            return inR1.to(device),outR1.to(device), count
+        elif data_flag == 'R2':
+            inR2 = th.stack([data[count:batch_size+count*batch_size,1], data[count:batch_size+count*batch_size],2], dim=1))
+            outR2 = data[count:batch_size+count*batch_size,3]
+            count = count + 1
+        return inR2.to(device),outR2.to(device), count
+
+    # The network needs to learn z1 -> z2 and z2,z3 -> z4
+    for i in range(N):
+            inData, outData, count =get_batch(data_flag, batch_size, count)
+            proposal = dist.Normal(*model(inData))
+            pred = proposal.rsample()
+
+            optimizer.zero_grad()
+            _loss = loss_fn(outData, pred)
+            _loss.backward()
+            optimizer.step()
+
+            outLoss += loss.item()
+
[…]

### 2019-08-06 code/amortized_rs/memory_src/pytorch_cpp_ext/amortized_rs.py
better name

@@ -28,24 +28,11 @@ def f():
     # the following is used to avoid tail recursion  - simple solution and stops stack overflow
     # You can not use this trick straight forwardly if you wish to access multiple core / threads.
     while z2 == math.inf:
-#         print(' while loop triggered')
         z2 = R1(z1)
         ctr += 1
-    #print("CTR: ", ctr)
     z3 = Uniform(0,2).sample()
-    #print(type(z3))
     z4 = R2(z2,z3)
     return torch.tensor([z1,z2,z3,z4])
-
-def R1(z1, i=0):
-    temp = Normal(z1,1).sample()
-    if temp.data > 0:
-        return temp
-    elif i == 100:
-        return math.inf
-    else:
-        i = i+1
-        return R1(z1,i)
 
 def R2(z2,z3):
     ''' Need to ensure that we don't return a
@@ -57,22 +44,12 @@ def R2(z2,z3):
         temp = Normal(z3,1).sample()
     return temp
 
-import timeit
-torch.manual_seed(7)
-print(f())
-print(timeit.timeit("f()", setup="from __main__ import f", number=2000))
-
-
 # recursion-free
 def R1(z1, i=0):
     i = 0
     bs = 1
     while True:
-        #temp = Normal(z1, 1).sample((bs,))
         temp = Normal(z1, 1).sample()
-        #mask = temp>0
-        #if torch.sum(mask) > 0:
-        #    return temp[mask][0]
         if temp > 0:
             return temp
         if i >= 10000:

### 2019-08-06 code/amortized_rs/pytorch_cpp_ext/hogwild.py

@@ -1,88 +1,0 @@
-import torch.multiprocessing as mp
-from model import MyModel
-from torch.utils.cpp_extension import load
-import torch as th
-from torch import optim
-from torch.distributions import Normal
-from utils import RejectionDataset
-from torch.utils.data import DataLoader
-
-amortized_rs = load(name="amortized_rs",
-                    sources=["amortized_rs.cpp"])
-loss_fn = th.nn.MSELoss()
-device = th.device("cuda" if th.cuda.is_available() else "cpu")
-inputs = DataLoader(RejectionDataset(split='train', l_data=128, train_percentage=0.8,fname_test, fname_train, InIndx, OutIndx), batch_size=128, shuffle=True, norma num_workers=cpu_count-2)
-outputs = DataLoader(RejectionDataset(split='test', l_data=128, test_percentage=0.8,fname_test, fname_train, InIndx, OutIndx), batch_size=128, shuffle=True,num_workers=cpu_count-2)
-optimizer = optim.Adam(filter(lambda p: p.requires_grad, model.parameters()), lr=config['lr'], amsgrad=True)
-
-def train(model, max_it, data_flag):
-    # Construct data_loader, optimizer, etc.
-    # optimizer = optim.SGD(model.parameters(), lr=0.01, momentum=0.9)
-    for it in range(max_it):
-        model.train()
-        batch_size = 128
-        N = 50000
-        n_samples = batch_size*N
-        # data = th.stack([amortized_rs.f() for _ in range(batch_size)]) # hmm, shouldn't this be the number of training samples, rather than batch_size?
-        # data = th.stack([amortized_rs.f() for _ in range(n_samples)]) # actually this loop needs to create multiple runs of the simulator of each z1, z2,z3 with
-        data = torch.load('/data/all_batch_samples.pt')
-        data = data[0:n_samples,:]
-        optimizer.zero_grad()
-        # TODO: Extract data, labels and define loss_fn etc!
-        count  = 0
-        def get_batch(data_flag, batch_size, count):
-            if data_flag == 'R1':
-                inR1 = data[count:batch_size+count*batch_size,0]
-                outR1 = data[count:batch_size+count*batch_size,1]
-                count = count + 1
-                return inR1.to(device),outR1.to(device), count
-            elif data_flag == 'R2':
-                inR2 = th.stack([data[count:batch_size+count*batch_size,1], data[count:batch_size+count*batch_size],2], dim=1))
-                outR2 = data[count:batch_size+count*batch_size,3]
-                count = count + 1
[…]

### 2019-08-06 code/amortized_rs/memory_src/hogwild.py
minor

@@ -1,0 +1,100 @@
+import torch.multiprocessing as mp
+from model import density_estimator
+from torch.utils.cpp_extension import load
+import torch as th
+from torch import optim
+from torch.distributions import Normal
+from utils import RejectionDataset
+from torch.utils.data import DataLoader
+import os
+import time
+
+amortized_rs = load(name="amortized_rs",
+                    sources=["amortized_rs.cpp"])
+
+def train(model, optimizer,  N, data_flag, batch_size, load_data=False):
+    # Construct data_loader, optimizer, etc.
+    # for it in range(max_it):
+    model.train()
+    n_samples = batch_size*N
+    # data = th.stack([amortized_rs.f() for _ in range(batch_size)]) # hmm, shouldn't this be the number of training samples, rather than batch_size?
+    # data = th.stack([amortized_rs.f() for _ in range(n_samples)]) # actually this loop needs to create multiple runs of the simulator of each z1, z2,z3 with
+    if load_data = True:
+        data = torch.load('../data/all_batch_samples.pt')
+        data = data[0:n_samples,:]
+    optimizer.zero_grad()
+    count  = 0
+    # no longer need write the batch loop in c - we can do that later
+    def get_batch(data_flag, batch_size, count, load_data=True):
+        if not load_data:
+            data = th.stack([amortized_rs.f() for _ in range(batch_size)]) 
+        if data_flag == 'R1':
+            inR1 = data[count:batch_size+count*batch_size,0]
+            outR1 = data[count:batch_size+count*batch_size,1]
+            count = count + 1
+            return inR1.to(device),outR1.to(device), count
+        elif data_flag == 'R2':
+            inR2 = th.stack([data[count:batch_size+count*batch_size,1], data[count:batch_size+count*batch_size],2], dim=1))
+            outR2 = data[count:batch_size+count*batch_size,3]
+            count = count + 1
+        return inR2.to(device),outR2.to(device), count
+
+    # The network needs to learn z1 -> z2 and z2,z3 -> z4
+    for i in range(N):
+            inData, outData, count =get_batch(data_flag, batch_size, count)
+            proposal = dist.Normal(*model(inData))
+            pred = proposal.rsample()
+
+            optimizer.zero_grad()
+            _loss = loss_fn(outData, pred)
+            _loss.backward()
+            optimizer.step()
+
+            outLoss += loss.item()
+
[…]

### 2019-08-06 code/amortized_rs/memory_src/hogwild.py
updated

@@ -86,7 +86,7 @@ if __name__ == '__main__':
     inputSize = batch_size
     outputSize = 2
     model = density_estimator(inputSize, outputSize)
-    num_processes = mp.cpu_count()
+    num_processes = mp.cpu_count() - 1
     N = 10000
     data_flag= 'R1'
     # NOTE: this is required for the ``fork`` method to work

### 2019-08-06 code/amortized_rs/memory_src/pytorch_cpp_ext/hogwild.py
updated

@@ -1,100 +1,0 @@
-import torch.multiprocessing as mp
-from model import density_estimator
-from torch.utils.cpp_extension import load
-import torch as th
-from torch import optim
-from torch.distributions import Normal
-from utils import RejectionDataset
-from torch.utils.data import DataLoader
-import os
-import time
-
-amortized_rs = load(name="amortized_rs",
-                    sources=["amortized_rs.cpp"])
-
-def train(model,  N, data_flag, batch_size, load_data=False):
-    # Construct data_loader, optimizer, etc.
-    # for it in range(max_it):
-    model.train()
-    n_samples = batch_size*N
-    # data = th.stack([amortized_rs.f() for _ in range(batch_size)]) # hmm, shouldn't this be the number of training samples, rather than batch_size?
-    # data = th.stack([amortized_rs.f() for _ in range(n_samples)]) # actually this loop needs to create multiple runs of the simulator of each z1, z2,z3 with
-    if load_data = True:
-        data = torch.load('../data/all_batch_samples.pt')
-        data = data[0:n_samples,:]
-    optimizer.zero_grad()
-    count  = 0
-    # no longer need write the batch loop in c - we can do that later
-    def get_batch(data_flag, batch_size, count, load_data=True):
-        if not load_data:
-            data = th.stack([amortized_rs.f() for _ in range(batch_size)]) 
-        if data_flag == 'R1':
-            inR1 = data[count:batch_size+count*batch_size,0]
-            outR1 = data[count:batch_size+count*batch_size,1]
-            count = count + 1
-            return inR1.to(device),outR1.to(device), count
-        elif data_flag == 'R2':
-            inR2 = th.stack([data[count:batch_size+count*batch_size,1], data[count:batch_size+count*batch_size],2], dim=1))
-            outR2 = data[count:batch_size+count*batch_size,3]
-            count = count + 1
-        return inR2.to(device),outR2.to(device), count
-
-    # The network needs to learn z1 -> z2 and z2,z3 -> z4
-    for i in range(N):
-            inData, outData, count =get_batch(data_flag, batch_size, count)
-            proposal = dist.Normal(*model(inData))
-            pred = proposal.rsample()
-
-            optimizer.zero_grad()
-            _loss = loss_fn(outData, pred)
-            _loss.backward()
-            optimizer.step()
-
-            outLoss += loss.item()
-
[…]

### 2019-08-06 code/amortized_rs/memory_src/hogwild.py
updated

@@ -4,8 +4,8 @@ from torch.utils.cpp_extension import load
 import torch as th
 from torch import optim
 from torch.distributions import Normal
-from utils import RejectionDataset
-from torch.utils.data import DataLoader
+# from utils import RejectionDataset
+# from torch.utils.data import DataLoader
 import os
 import time
 
@@ -19,7 +19,7 @@ def train(model, optimizer,  N, data_flag, batch_size, load_data=False):
     n_samples = batch_size*N
     # data = th.stack([amortized_rs.f() for _ in range(batch_size)]) # hmm, shouldn't this be the number of training samples, rather than batch_size?
     # data = th.stack([amortized_rs.f() for _ in range(n_samples)]) # actually this loop needs to create multiple runs of the simulator of each z1, z2,z3 with
-    if load_data = True:
+    if load_data == True:
         data = torch.load('../data/all_batch_samples.pt')
         data = data[0:n_samples,:]
     optimizer.zero_grad()
@@ -29,12 +29,13 @@ def train(model, optimizer,  N, data_flag, batch_size, load_data=False):
         if not load_data:
             data = th.stack([amortized_rs.f() for _ in range(batch_size)]) 
         if data_flag == 'R1':
+            print(data)
             inR1 = data[count:batch_size+count*batch_size,0]
             outR1 = data[count:batch_size+count*batch_size,1]
             count = count + 1
             return inR1.to(device),outR1.to(device), count
         elif data_flag == 'R2':
-            inR2 = th.stack([data[count:batch_size+count*batch_size,1], data[count:batch_size+count*batch_size],2], dim=1))
+            inR2 = th.stack([data[count:batch_size+count*batch_size,1], data[count:batch_size+count*batch_size,2]], dim=1)
             outR2 = data[count:batch_size+count*batch_size,3]
             count = count + 1
         return inR2.to(device),outR2.to(device), count
@@ -56,11 +57,11 @@ def train(model, optimizer,  N, data_flag, batch_size, load_data=False):
                 print("iteration {:04d}/{:d}: loss: {:6.3f}".format(i, iterations,
                                                                     loss.item() / args.batch_size))
             print('====> Epoch: {:03d} Train loss: {:.4f}'.format(epoch, ...))
-        if not os.path.exists('../model/'):
[…]

### 2019-08-06 code/amortized_rs/memory_src/model.py

@@ -1,0 +1,23 @@
+import torch
+from torch import nn
+from torch.nn import functional as F
+
+class density_estimator(nn.Module):
+    def __init__(self, inputSize, outputSize):
+        # super(density_estimator, self).__init__()
+        self.hidden = nn.Sequential(
+            nn.Linear(inputSize, inputSize//2),
+            nn.ReLU(inputSize //2, inputSize//2),
+            nn.Linear(inputSize//2, inputSize//(inputSize//2)) # ensures that the output would be 2, assuming inputsize = 2^n
+
+
+        )
+        self.linear_mu = nn.Linear( inputSize//(inputSize//2), outputSize)
+        self.linear_std = nn.Linear( inputSize//(inputSize//2), outputSize)
+
+    def forward(self, x):
+         # x is of shape [batch_size, input_dim]
+        hidden = self.hidden(x) # return [batch_size//2, 2]
+        return self.linear_mu(hidden), nn.Softplus(self.linear_std(hidden))+1e-6
+
+

### 2019-08-06 code/amortized_rs/cython_code/amortized_rs.pyx
updated

@@ -1,35 +1,0 @@
-import numpy as np
-cimport numpy as cnp
-
-def  f():
-    DTYPE = cnp.float64_t
-    retarr = np.zeros((4,), dtype=DTYPE)
-    cdef double z_1 = cnp.random.normal(0, 1, 1)
-    cdef double z_2
-    while z_2 == cnp.inf:
-        z_2 = R1(z_1, 1)
-    cdef double z_3 = cnp.random.uniform(0, 2)
-    cdef double z_4 = R2(z_2, z_3, 1)
-    retarr[0] = z_1
-    retarr[1] = z_2
-    retarr[2] = z_3
-    retarr[3] = z_4
-    return retarr
-
-cdef double R1(double z_1, int i) nogil:
-    cdef double temp = cnp.random.normal(z_1, 1, 1)
-    if temp > 0:
-        return temp
-    elif i == 1000:
-        return cnp.inf
-    else:
-        i += 1
-        return R1(z_1,i)
-
-cdef double R2(double z_2, double z_3) nogil:
-    cdef double temp = 0
-    while temp < z_2:
-        temp = cnp.random.normal(z_3, 1, 1)
-    return temp
-
-# cdef f_nonrec():

### 2019-08-06 code/amortized_rs/cython_code/build_amortized_rs.sh

@@ -1,5 +1,0 @@
-#!/bin/bash
-python3 setup_amortized_rs.py build_ext --inplace
-yes | cp -rf cython_code/amortized*.so amortized_rs.so
-rm -rf build
-rm -rf cython_code

### 2019-08-06 code/amortized_rs/memory_src/hogwild.py
minor

@@ -86,11 +86,11 @@ if __name__ == '__main__':
     # optimizer = optim.SGD(model.parameters(), lr=0.01, momentum=0.9)
     batch_size = 128
     inputSize = batch_size
-    outputSize = 2
+    data_flag= 'R1'
+    outputSize = 1 # for R1
     model = density_estimator(inputSize, outputSize)
     num_processes = mp.cpu_count() - 1
     N = 10000
-    data_flag= 'R1'
     # NOTE: this is required for the ``fork`` method to work
     model.share_memory()
     processes = []

### 2019-08-06 paper/learning/notes.tex
updated

@@
 
-\section{Limitations with different inference engines}
-
-\section{Non-prior Based Sampling}
-In non-prior based sampling we utilise conditioning too, which enables 
-us to explore spaces more efficiently. 
-In the current hijacking set-up, we
-can only do conditioning at the end of each \texttt{forward()} run of the simulator. 
-This does inhibit certain inference engines and makes aspects of performing inference 
-within this setting more complicated. 
-Nonetheless, there are several inference schemes 
-that we can exploit and utilise. 
-We shall now give an Introduction to each of these schemes.
-
-\subsection{Importance Sampling}
-In the generalised probabilistic programming setting an importance sampling scheme over the 
-program raw samples and observations can be defined as follows:
-\begin{equation}
-  \label{eq:posterior}
-  p(\mathcal{T}=x_{1:n_x}| \lambda)=
-  \begin{cases}
-  \prod_{i=1}^{n_x} f_{a_i}(x_i | \eta) \prod_{j=1}^{n_y} g_{b_j}(y_j | \phi) \hspace{0.1cm} \texttt{if} \mathcal{B}(x_{1:n_x}) = 1\\
-  0
-  \end{cases}
-\end{equation}
-
-where the observe statements are defined by $g_{b_j}(.)$, $\phi$ plays the same role as 
-$\eta$ and $n_y$ represents the number of observations.
-The expectation over the whole program can be defined as: 
-\begin{equation}
-\mathbb{E}_{p(\Omega | \lambda)}[h(\Omega)] = \mathbb{E}_{p(x_{1:n_x} | \lambda)}[h(P_r(x_{1:n_x}))]
-\end{equation}
-\textbf{Note: It is a little weird how we define this, as the the program output $\Omega$ should
-also include the product of the weights}
-
-where $r$ represents a run of the program, $P_{r}(\cdot)$ is the entirety of the program output,
-$h(\cdot)$ is just a deterministic function.
-
-As the importance sampler updates the weights according to the following formula:
-\begin{equation}
-  w(x_{1:n_x}) = \prod^{n_x}_{i=1}\frac{f_{a_i}(x_i | \eta_i)}{q_i(x_i |f_{a_i}, \eta_i)} \prod^{n_y}_{j=1} g_{b_j}(y_j | \phi_j)
-\end{equation}
-
-When the number of observations is fixed, the combined weight is equal to $w(x_{1:n_x}) = \prod^{n_y}_{j=1} w_j $. 
-To generate the weight up to each observe we can define $K(j)$ to be a function that returns the least $i$ before hitting an observe $j$.
-This enables us to write the $j$-th weight as follows:
-\begin{equation}
-  w_j(x1:K(j)) = g_{b_j}(y_j | \psi_j) \prod^{K(j)}_{K(j-1)+1} v_i
-\end{equation}
-where we define $v_j = \frac{f_{a_j}(x_j | \eta_j)}{q_j(x_j |f_{a_j}, \eta_j)}$
-
-Taking these definitions we can construct a target density over the 
-latents that is proportional to the products of \texttt{sample} and \texttt{observe}
-statements \textbf{Note: Add update posterior here}.
-Thus, given the proposal $q(x_{1:K(t)})$ we define the importance weight as:
-\begin{multline}
-  w(x_{1:t}) = \prod^{t}_{j=1} w_j (x_{1:K(j)}) = \\
-   \frac{p_t(x_{1:K(t)})}{q(x_{1:K(t)})} =
- \begin{cases}
-  \prod^{t}_{j=1}g_{b_j}(y_j | w_i)\prod^K(j)_{K(j-1)+1} v_i \hspace{0.1cm} \texttt{if} \hspace{0.1cm} B_t(x_1:K(t))=1 \\
-  0
- \end{cases}
-\end{multline}
-
-One problem with importance sampling is in its ability to scale and condition upon multiple observations. |
-Choosing a \emph{good} prior remains challenging.
-
-\subsection{Sequential Monte Carlo}
-Given multiple observations, we can ake the inference more computationally efficient 
-by employing sequential monte carlo (SMC), although as we can only condition at the end 
-of one full run of the simulation, computational speed increases are minimal.
-SMC allows for conditioning on multiple observations, but the number of observations must
-be fixed. \textbf{Note: If we could condition on observations within the simulator, we 
-could get dramatic speed increases over standard importance sampling.We can actually do this 
-in pyprob, but it requires knowing where to condition. That can be quite challenging 
-for arbitrary code. }
-
-\begin{algorithm}
-    \caption{SMC for Population Based Simulators - well any general program - not new}
-    \label{alg:euclid}
-    \begin{algorithmic}[1] % The number tells where the line numbering should start
-        \Procedure{SMC}{$T,M$,\textbf{x}, \textbf{w},\textbf{v}, \textbf{g}}
-          \For{$t = 1:T$}
-            \For{$m = 1:M$}
-              \State $w^{m} = 1  $
-              \For{$k = K(t-1)+1:K(t)$}
-                \State $x^m_k \sim q_k(x^m_k | f_{a^{m}_{k}},\eta^{m}_{j})$
-                \State $w^m_t \gets w^t_m v_j^m(x_k^m)$
-                \EndFor
-            \EndFor
-          \State $w^m_t \gets w^m_t g_{b_{t}^m}(y_t^m | \psi^m_t)$
-          \State $\{x^{1:m}_{1:t}\}$ $\gets$ \texttt{resample}( $\{x^{1:m}_{1:t}\}$,  $\{w^{1:m}_{1:t}\}$ )
-        \EndFor
-        \EndProcedure
-    \end{algorithmic}
-\end{algorithm}
-
-To implement SMC in the current set-up we round require a forking method and the
-
-\subsection{Random-walk Metropolis Hastings}
-
-Due to the current set-up performing Random-walk Metropolis Hastings (RMH) is not feasible 
-as the rejection sampling blocks within the simulator mean that we cannot correctly perform RMH
-as we do not have a functional form for all parts of the simulator to construct a target density
-that we can evaluate. \textbf{Note:
-Ask tom what:  one can sometimes achieve improvements by instead using the type of the distribution object 
-associated with $x_m$ (what does this first part mean?) 
- to automatically construct a valid random walk proposal that allows for improved
- hill climbing behavior on the individual updates. }.
-
-\textbf{TODO: Add some of the plots generated from this model and an accompanying analysis to them.}
-
 \section{Amortized Rejection Sampling}
@@
 
-Rejection samplers are heavily used when designing simulators, due to their ease of use and ability to model 
-any arbitrary distribution on $\mathbb{R}^{D}$, with a sufficiently well designed rejection sampling scheme~\cite{MEYER20083408,martino2012improved,casella} and have 
-their origins in 1940 - 1950's, during the development of other prominent Monte carlo techniques at Los Alamos.
+Rejection samplers are heavily used when designing simulators [ add SHERPA etc], due to their ease of use and ability to model 
+any arbitrary distribution on $\mathbb{R}^{D}$, with a sufficiently well designed rejection sampling scheme~\cite{MEYER20083408,martino2012improved,casella}
 
@@
 
-\subsection{Prior based sampling}
-In prior based sampling we take our existing program and look directly at the product of all the raw sample calls, denoted $f_{a_i}(x_i | \eta)$ and 
-construct a proposal that is is dependent on those. In this sampling scheme our proposal is defined as:
-\begin{equation}
-  q(x_{1:n_{x}}) = 
-  \begin{cases}
-    \prod_{i=1}^{n_x} f_{a_i}(x_i | \eta) \hspace{0.1cm} \texttt{if} \hspace{0.1cm} \mathcal{B}(x_{1:n_{x}})=1 \\
-    0
-  \end{cases}
-\end{equation}
-where $\mathcal{B}(\cdot)$ is satisfied by the simulator and $f_{a_i}(\cdot)$ represents a
-raw random sample, $\eta$ is a subset of the variables in-scope at the point of sampling
-and $n_{x}$ is the number of latent variables.
- This is true in all inference engines used in this hijacking framework.
-However, this makes use of no conditioning and is not sufficient for performing inference in
-complex, population-based simulations.
-
 In order to correctly characterize this problem we first provide a useful diagram demonstrating 
@@
 
-We do this by  Replacing the rejection sampling blocks with 
+% In order to correctly characterize this problem we first provide a useful diagram demonstrating 
+% the density of the system.
+
+%  \input{thesystem.tex}
+
+We do this by replacing the rejection sampling blocks with 
 learnt functions from an amortized inference procedure Figure~\ref{fig:amortized_sampling}.
@@
 
+\section{Limitations with different inference engines}
+
+\subsection{Prior based sampling}
+In prior based sampling we take our existing program and look directly at the product of all the raw sample calls, denoted $f_{a_i}(x_i | \eta)$ and 
+construct a proposal that is is dependent on those. In this sampling scheme our proposal is defined as:
+\begin{equation}
+  q(x_{1:n_{x}}) = 
+  \begin{cases}
+    \prod_{i=1}^{n_x} f_{a_i}(x_i | \eta) \hspace{0.1cm} \texttt{if} \hspace{0.1cm} \mathcal{B}(x_{1:n_{x}})=1 \\
+    0
+  \end{cases}
+\end{equation}
+where $\mathcal{B}(\cdot)$ is satisfied by the simulator and $f_{a_i}(\cdot)$ represents a
+raw random sample, $\eta$ is a subset of the variables in-scope at the point of sampling
+and $n_{x}$ is the number of latent variables.
+ This is true in all inference engines used in this hijacking framework.
+However, this makes use of no conditioning and is not sufficient for performing inference in
+complex, population-based simulations.
+
+\subsection{Non-prior Based Sampling}
+In non-prior based sampling we utilise conditioning too, which enables 
+us to explore spaces more efficiently. 
+In the current hijacking set-up, we
+can only do conditioning at the end of each \texttt{forward()} run of the simulator. 
+This does inhibit certain inference engines and makes aspects of performing inference 
+within this setting more complicated. 
+Nonetheless, there are several inference schemes 
+that we can exploit and utilise. 
+We shall now give an Introduction to each of these schemes.
+
+\subsection{Importance Sampling}
+In the generalised probabilistic programming setting an importance sampling scheme over the 
+program raw samples and observations can be defined as follows:
+\begin{equation}
+  \label{eq:posterior}
+  p(\mathcal{T}=x_{1:n_x}| \lambda)=
+  \begin{cases}
+  \prod_{i=1}^{n_x} f_{a_i}(x_i | \eta) \prod_{j=1}^{n_y} g_{b_j}(y_j | \phi) \hspace{0.1cm} \texttt{if} \mathcal{B}(x_{1:n_x}) = 1\\
+  0
+  \end{cases}
+\end{equation}
+
+where the observe statements are defined by $g_{b_j}(.)$, $\phi$ plays the same role as 
+$\eta$ and $n_y$ represents the number of observations.
+The expectation over the whole program can be defined as: 
+\begin{equation}
+\mathbb{E}_{p(\Omega | \lambda)}[h(\Omega)] = \mathbb{E}_{p(x_{1:n_x} | \lambda)}[h(P_r(x_{1:n_x}))]
+\end{equation}
+\textbf{Note: It is a little weird how we define this, as the the program output $\Omega$ should
+also include the product of the weights}
+
+where $r$ represents a run of the program, $P_{r}(\cdot)$ is the entirety of the program output,
+$h(\cdot)$ is just a deterministic function.
+
+As the importance sampler updates the weights according to the following formula:
+\begin{equation}
+  w(x_{1:n_x}) = \prod^{n_x}_{i=1}\frac{f_{a_i}(x_i | \eta_i)}{q_i(x_i |f_{a_i}, \eta_i)} \prod^{n_y}_{j=1} g_{b_j}(y_j | \phi_j)
+\end{equation}
+
+When the number of observations is fixed, the combined weight is equal to $w(x_{1:n_x}) = \prod^{n_y}_{j=1} w_j $. 
+To generate the weight up to each observe we can define $K(j)$ to be a function that returns the least $i$ before hitting an observe $j$.
+This enables us to write the $j$-th weight as follows:
+\begin{equation}
+  w_j(x1:K(j)) = g_{b_j}(y_j | \psi_j) \prod^{K(j)}_{K(j-1)+1} v_i
+\end{equation}
+where we define $v_j = \frac{f_{a_j}(x_j | \eta_j)}{q_j(x_j |f_{a_j}, \eta_j)}$
+
+Taking these definitions we can construct a target density over the 
+latents that is proportional to the products of \texttt{sample} and \texttt{observe}
+statements \textbf{Note: Add update posterior here}.
+Thus, given the proposal $q(x_{1:K(t)})$ we define the importance weight as:
+\begin{multline}
+  w(x_{1:t}) = \prod^{t}_{j=1} w_j (x_{1:K(j)}) = \\
+   \frac{p_t(x_{1:K(t)})}{q(x_{1:K(t)})} =
+ \begin{cases}
+  \prod^{t}_{j=1}g_{b_j}(y_j | w_i)\prod^K(j)_{K(j-1)+1} v_i \hspace{0.1cm} \texttt{if} \hspace{0.1cm} B_t(x_1:K(t))=1 \\
+  0
+ \end{cases}
+\end{multline}
+
+One problem with importance sampling is in its ability to scale and condition upon multiple observations. |
+Choosing a \emph{good} prior remains challenging.
+
+\subsection{Sequential Monte Carlo}
+Given multiple observations, we can ake the inference more computationally efficient 
+by employing sequential monte carlo (SMC), although as we can only condition at the end 
+of one full run of the simulation, computational speed increases are minimal.
+SMC allows for conditioning on multiple observations, but the number of observations must
+be fixed. \textbf{Note: If we could condition on observations within the simulator, we 
+could get dramatic speed increases over standard importance sampling.We can actually do this 
+in pyprob, but it requires knowing where to condition. That can be quite challenging 
+for arbitrary code. }
+
+\begin{algorithm}
+    \caption{SMC for Population Based Simulators - well any general program - not new}
+    \label{alg:euclid}
+    \begin{algorithmic}[1] % The number tells where the line numbering should start
+        \Procedure{SMC}{$T,M$,\textbf{x}, \textbf{w},\textbf{v}, \textbf{g}}
+          \For{$t = 1:T$}
+            \For{$m = 1:M$}
+              \State $w^{m} = 1  $
+              \For{$k = K(t-1)+1:K(t)$}
+                \State $x^m_k \sim q_k(x^m_k | f_{a^{m}_{k}},\eta^{m}_{j})$
+                \State $w^m_t \gets w^t_m v_j^m(x_k^m)$
+                \EndFor
+            \EndFor
+          \State $w^m_t \gets w^m_t g_{b_{t}^m}(y_t^m | \psi^m_t)$
+          \State $\{x^{1:m}_{1:t}\}$ $\gets$ \texttt{resample}( $\{x^{1:m}_{1:t}\}$,  $\{w^{1:m}_{1:t}\}$ )
+        \EndFor
+        \EndProcedure
+    \end{algorithmic}
+\end{algorithm}
+
+To implement SMC in the current set-up we round require a forking method and the
+
+\subsection{Random-walk Metropolis Hastings}
+
+Due to the current set-up performing Random-walk Metropolis Hastings (RMH) is not feasible 
+as the rejection sampling blocks within the simulator mean that we cannot correctly perform RMH
+as we do not have a functional form for all parts of the simulator to construct a target density
+that we can evaluate. \textbf{Note:
+Ask tom what:  one can sometimes achieve improvements by instead using the type of the distribution object 
+associated with $x_m$ (what does this first part mean?) 
+ to automatically construct a valid random walk proposal that allows for improved
+ hill climbing behavior on the individual updates. }.
+
+\textbf{TODO: Add some of the plots generated from this model and an accompanying analysis to them.}
+
 \section{An Example: Modelling Parasite Densities in Simulation}

### 2019-08-06 paper/learning/notes.tex
updated

@@
 \usepackage{wrapfig}
+\usepackage{bookmark}
 \usepackage{amsmath,amsthm}
@@
 
-\section{Amortized Rejection Sampling}
-
-One common problem in traditional simulators, \emph{programs}, outside of the standard probabilistic
-programming framework is that they rely heavily on rejection sampling 
-to generate samples from the target density. Not only is rejection sampling 
-computational inefficient, [especially if criteria in which samples are generated
-has been poorly calibrated] (may leave out this statement as it may make a reviewer
-question why would amortization work any better.), but it creates complications for inference.
-As the rejection sampling loops provide no explicit form of the target density,
-we can not simply evaluate the density and leverage statistically and computationally
-efficient algorithms such as MCMC sampling. This inhibits our ability to transform an arbitrary
-stochastic simulator into a probabilistic programming framework~\cite{baydin2018efficient}. 
-To this end, we derive both a mathematical and engineering based framework to facilitate 
-the amortization of rejection sampling loops in arbitrary simulators, which provides a path
-way to both leverage a larger, more efficient class of inference methods and convert arbitrary
-stochastic simulators into probabilistic programs systems.
-
-\begin{figure}
-  % \begin{minipage}[c]{0.67\textwidth}
-  \includegraphics[]{amoritized_rejection.pdf}
-  % \end{minipage}\hfill
-  % \begin{minipage}[c]{0.3\textwidth}
-    \caption{Using PyProb we can locate all regions of inefficient sampling 
-    and replace those points in the code with learnt functions that represent
-    the distributions of the rejection sampling statements. $Z_{i}$ represents the marginal and 
-    by learning surrogate functions for each of the rejection sampling loops.
-    In doing so we can learn approximate raw sample distributions $f(x_{i} | \eta)$ for each
-    rejection sampling loop. This means that we can evaluate the function at each part in the simulator, enabling
-    us to construct a target density that can be evaluated when using MCMC inference schemes such as RMH.}
-    \label{fig:amortized_sampling}
-  % \end{minipage}
-\end{figure}
-
 \section{Rejection Sampling}
@@
 
-\section{Defining the Density and Sub-Densities within Simulators}
+\section{Amortized Rejection Sampling}
+
+One common problem in traditional simulators, \emph{programs}, outside of the standard probabilistic
+programming framework is that they rely heavily on rejection sampling 
+to generate samples from the target density. Not only is rejection sampling 
+computational inefficient, [especially if criteria in which samples are generated
+has been poorly calibrated] (may leave out this statement as it may make a reviewer
+question why would amortization work any better.), but it creates complications for inference.
+As the rejection sampling loops provide no explicit form of the target density,
+we can not simply evaluate the density and leverage statistically and computationally
+efficient algorithms such as MCMC sampling. This inhibits our ability to transform an arbitrary
+stochastic simulator into a probabilistic programming framework~\cite{baydin2018efficient}. 
+To this end, we derive both a mathematical and engineering based framework to facilitate 
+the amortization of rejection sampling loops in arbitrary simulators, which provides a path
+way to both leverage a larger, more efficient class of inference methods and convert arbitrary
+stochastic simulators into probabilistic programs systems.
+
+\begin{figure}
+  % \begin{minipage}[c]{0.67\textwidth}
+  \includegraphics[]{amoritized_rejection.pdf}
+  % \end{minipage}\hfill
+  % \begin{minipage}[c]{0.3\textwidth}
+    \caption{Using PyProb we can locate all regions of inefficient sampling 
+    and replace those points in the code with learnt functions that represent
+    the distributions of the rejection sampling statements. $Z_{i}$ represents the marginal and 
+    by learning surrogate functions for each of the rejection sampling loops.
+    In doing so we can learn approximate raw sample distributions $f(x_{i} | \eta)$ for each
+    rejection sampling loop. This means that we can evaluate the function at each part in the simulator, enabling
+    us to construct a target density that can be evaluated when using MCMC inference schemes such as RMH.}
+    \label{fig:amortized_sampling}
+  % \end{minipage}
+\end{figure}
+
+\ssubection{Defining the Density and Sub-Densities within Simulators}
 
@@
 
-\section{ An overview of Amortized Inference schemes}
-\begin{itemize}
-  \item Whilst the variational posterior we choose as our proposal has a functional form 
-  and can be sampled, the functional form of the rejection sampling loops is not explicitly known,
-  which leads to problems when calculating the $KL$-divergence. 
-  \item There are two approaches that we can take to calculating some divergence metric. 
-  \item 1) Using something like a non-parametric divergence estimator that makes use of Renyi and $L_{2}$ divergences (see \cite{basseville2013divergence} for details)
-  \item 2) Use some form of kernel density to estimation to approximate the functional form of the rejection sampling 
-  block, which can be used in place. From which we learn a variational approximation characterized by a trained network that we can sample from quickly. 
-  \item An issue with approach 2) is that if I can approximate the density in functional form aren't there more efficient methods that we can use? (\textbf{Ask tom})
-  \item Or, if assuming that the rejection sampling block can be arbitrary dimension, are we first learning a flow to represent that density and then learn a variational approximation of something that we can easily sample from.
-  \item But, this is very much the idea behind normalising flows, take a simple distribution that you can sample and then map it to the ``complicated'' density. 
-  \item Stein Variational inference may be quite nice to use here~\cite{feng2017learning}, but it does make explicit assumptions in requiring the densities to be differentiable. 
-  \item  For an overview of all variational methods, including the use of different divergence metrics see here ~\cite{8588399}
-\end{itemize}
-
-New questions to ask Tom
-
-\begin{itemize}
-  \item Isn't what we are doing just conditional density estimation? It seems
-  like there is a lot of prior work here. I.e see this paper here for a good overview ~\cite{greenberg2019automatic}
-  \item It does seem like this field has a large extensive past.
-\end{itemize}
-
-\subsection{Using a simple Neural Network regressor}
-
-As a first approach, we use a regressor to learn a function to map
-between the input / output pairs that are outputted from the rejection
-sampling blocks with in the simulator.
-
-\paragraph{Architecture} For each rejection sampling block we learn a proposal $q(z_1, \ldots, z_i | \eta)$
-\subsection{Using variants of VAEs}
-
-\subsection{Density estimation using Normalising flows}
-Notes from models that define an explicit,  
-tractable density are highly effective, 
-because they permit the use of an optimization algorithm directly on the log-likelihood of the training data.
-However,  the  family  of  models  that provide  a  tractable  density  is  limited,
-with  different  families  having  different disadvantage.
-The rejection sampling loops, if proposed correctly, provide a way of generating 
-samples from some density, but do not provide a way of constructing the target density (\textbf{Bradley: Ask Tom, is this true?}). 
-Within event and population-based simulators, the rejection sampling loops have been designed with care,
-to ensure that they are representative to the events that they are simulating. So we can, with confidence,
-assume that they are representative of some density. 
-However, to utilise powerful inference engines, we must be able to construct the 
-full density of the program.
-
-One recent technique for density estimation, in the amortized inference setting,
-is normalising flows combined with variation inference~\cite{rezende2015variational}.
-Flow-based methods work by leveraging a simple property of differential geometry applied 
-to random variables. 
-That being the change of variables.
-Let $z \sim r(z)$ where $r(z)$ represents the posterior defined by the rejection
-sampling loop, and define $\pi$ to be a smooth and invertible function.
-When $\pi$ acts $z$ we transform the random variable from one domain, to another. 
-Thus, if $z\prime = \pi(z)$ then we can construct the density of  
-the probability density of the new random variable $z\prime$ under the flow initiated by $\pi$: 
-\begin{equation}
-r(z\prime) = r(\pi^{-1}(z\prime)) \left|\det \frac{\partial \pi^{-1}}{\partial z'}\right| = r(\pi^{-1}(z\prime)) \left|\det \frac{\partial \pi}{\partial z\prime}\right|^{-1} 
-\end{equation}
-
-\begin{itemize}
-  \item Do we need to know the base distribution that the RS block is sampling from? i.e there maybe be multiple $r(z)$ in each 
-  RS block. 
-  \item Is it right to say here that $z\prime$ is the output of the RS block, we do not know $\pi$ here, or is $r(z)$ the raw sample $f_{a_i}$ and $\pi$ the distribution specified by the rejection sampling loop?
-  \item We need to learn $\pi(.)$
-\end{itemize}
-
-\section{Constructing the Objective}
+\subsection{}{Constructing the Objective}
 
@@
 separately. 
+% \section{ An overview of Amortized Inference schemes}
 
+\section{Experiments: Amortizing Stochastic Processes}
+
+\subsection{Baseliner}
+As a first approach, we train a regressor to learn a function to map
+between the input / output pairs that are outputted from the rejection
+sampling blocks with in the simulator.
+
+For $R1$ we have one input and one output, for $R2$ we have 2 inputs and 1 output. 
+For the loss function we use a simple mean squared error loss and use a the regressor the learn 
+and unknown mean $\mu$ and variance $\sigma^{2}$. This means that we are trying to fit a simple Normal distribution
+to each rejection sampling block. Once learned we can then efficiently sample from the parameterized normal distribution, which 
+represents our learned proposal $q(z_i, \ldots, z_{n} | \eta)$. 
+\paragraph{Training scheme} For each rejection sampling block we learn a proposal $q(z_1, \ldots, z_i | \eta)$, in this case 
+\textbf{Note: Add results here. Code almost complete}
+
+\subsection{More complicated: Conditional Density estimation}
+
+\paragraph{Normalizing flows} One recent technique for density estimation, in the amortized inference setting,
+is normalising flows combined with variation inference~\cite{rezende2015variational}.
+Flow-based methods work by leveraging a simple property of differential geometry applied 
+to random variables. 
+That being the change of variables. 
+Using flow based methods we can construct our rejection sampling block density, by learning a flow to map from
+a simpler distribution to more complicated one.
+Let $z \sim r(z)$ where $r(z)$ represents the posterior defined by the rejection
+sampling loop, and define $\pi$ to be a smooth and invertible function.
+When $\pi$ acts $z$ we transform the random variable from one domain, to another. 
+Thus, if $z\prime = \pi(z)$ then we can construct the density of  
+the probability density of the new random variable $z\prime$ under the flow initiated by $\pi$: 
+\begin{equation}
+r(z\prime) = r(\pi^{-1}(z\prime)) \left|\det \frac{\partial \pi^{-1}}{\partial z'}\right| = r(\pi^{-1}(z\prime)) \left|\det \frac{\partial \pi}{\partial z\prime}\right|^{-1} 
+\end{equation}
+
+Flows availabile are MVP and Neural Spline flows [add citations].
+
+for other techniques ask Guens.
+
+\begin{itemize}
+  \item Do we need to know the base distribution that the RS block is sampling from? i.e there maybe be multiple $r(z)$ in each 
+  RS block. 
+  \item Is it right to say here that $z\prime$ is the output of the RS block, we do not know $\pi$ here, or is $r(z)$ the raw sample $f_{a_i}$ and $\pi$ the distribution specified by the rejection sampling loop?
+  \item We need to learn $\pi(.)$
+\end{itemize}
+\subsection{Using variants of VAEs}
+
+\subsection{Density estimation using Normalising flows}
+Notes from models that define an explicit,  
+tractable density are highly effective, 
+because they permit the use of an optimization algorithm directly on the log-likelihood of the training data.
+However,  the  family  of  models  that provide  a  tractable  density  is  limited,
+with  different  families  having  different disadvantage.
+The rejection sampling loops, if proposed correctly, provide a way of generating 
+samples from some density, but do not provide a way of constructing the target density (\textbf{Bradley: Ask Tom, is this true?}). 
+Within event and population-based simulators, the rejection sampling loops have been designed with care,
+to ensure that they are representative to the events that they are simulating. So we can, with confidence,
+assume that they are representative of some density. 
+However, to utilise powerful inference engines, we must be able to construct the 
+full density of the program.
+
+\section{ Questions for Tom}
+\begin{itemize}
+  \item Whilst the variational posterior we choose as our proposal has a functional form  
+  and can be sampled, the functional form of the rejection sampling loops is not explicitly known,
+  which leads to problems when calculating the $KL$-divergence - reparametrize a with a function from a family of easy to sample distribution. 
+  \item There are two approaches that we can take to calculating some divergence metric. 
+  \item 1) Using something like a non-parametric divergence estimator that makes use of Renyi and $L_{2}$ divergences (see \cite{basseville2013divergence} for details) - No point, more difficult to implement
+  \item 2) Use some form of kernel density to estimation to approximate the functional form of the rejection sampling 
+  block, which can be used in place. From which we learn a variational approximation characterized by a trained network that we can sample from quickly. 
+  \item An issue with approach 2) is that if I can approximate the density in functional form aren't there more efficient methods that we can use?
+- There are other methods out there and it would maybe make sense to compare some of those methods.
+However, it may still be difficult to sample.
+  \item Or, if assuming that the rejection sampling block can be arbitrary dimension, are we first learning a flow to represent that density and then learn a variational approximation of something that we can easily sample from.
+  \item But, this is very much the idea behind normalising flows, take a simple distribution that you can sample and then map it to the ``complicated'' density - Question not well defined. 
+  \item Stein Variational inference may be quite nice to use here~\cite{feng2017learning}, but it does make explicit assumptions in requiring the densities to be differentiable.
+- Yes this metod is good, but only works in the continuous domain.
+  \item  For an overview of all variational methods, including the use of different divergence metrics see here ~\cite{8588399} - Answered - Quite finickity to set up. 
+\end{itemize}
+
+New questions to ask Tom
+
+\begin{itemize}
+  \item Isn't what we are doing just conditional density estimation? It seems
+  like there is a lot of prior work here. I.e see this paper here for a good overview ~\cite{greenberg2019automatic}
+  \item It does seem like this field has a large extensive past, but we still have a small neesh .
+\end{itemize}
 \section{Additional TODOs}

### 2019-08-06 paper/learning/notes.tex

@@
 \usepackage{wrapfig}
-\usepackage{bookmark}
+
 \usepackage{amsmath,amsthm}
@@
 easily sampled from and used to construct a functional form of the likelihood, that enables the 
-target density to be fully constructed, enabling one to leverage a large class of sample effient inference schemes, 
-namely MCMC.
+target density to be fully constructed, enabling one to leverage a large class of sample effient inference schemes, namely MCMC.
 
@@
 
-\ssubection{Defining the Density and Sub-Densities within Simulators}
+\subsection{Defining the Density and Sub-Densities within Simulators}
 
@@
 
-\subsection{}{Constructing the Objective}
+\subsection{Constructing the Objective}
 
@@
 separately.
-% \section{ An overview of Amortized Inference schemes}
 
@@
 
-\subsection{Baseliner}
+\subsection{Baseline}
 As a first approach, we train a regressor to learn a function to map
@@
 
+\subsection{2nd Baseline}
+Instead of using a single guassain, use a proposal that is a mixture of gaussians. 
+\texbf{TODO: Add math tomorrow}
 \subsection{More complicated: Conditional Density estimation}
@@
 
-\subsection{Density estimation using Normalising flows}
 Notes from models that define an explicit,  
@@
 \end{itemize}
+=======
+
+>>>>>>> e2910a3683b657af5fa91292c678419ec3b427c7
 \section{Additional TODOs}

### 2019-08-07 code/amortized_rs/memory_src/hogwild.py
minor

@@ -60,14 +60,14 @@ def train(model, optimizer,  N, data_flag, batch_size, load_data=False):
     if not os.path.exists('../model/'):
         os.makedirs('../model/')
     fname = '../model/model_{}_rejectionBlock_{}'.format(strftime("%Y-%m-%d_%H-%M"), data_flag)
-    torch.save(model.state_dict(), fname)
+    th.save(model.state_dict(), fname)
     print(' Model is saved at : {}'.format(fname))
 
 def test(test_iterations,model_name):
-    model = torch.load(model_name)
+    model = th.load(model_name)
     model.eval()
     out_loss = 0
-    with torch.no_grad():
+    with th.no_grad():
         for i in range(test_iterations):
             inData, outData = get_batch(data_flag, batch_size, count)
             proposal = dist.Normal(*model(inData))

### 2019-08-07 code/amortized_rs/memory_src/hogwild.py
minor

@@ -29,15 +29,14 @@ def train(model, optimizer, loss_fn,  N, data_flag, batch_size, load_data=False)
         if not load_data:
             data = th.stack([amortized_rs.f() for _ in range(batch_size)]) 
         if data_flag == 'R1':
-            print(data)
-            inR1 = data[count:batch_size+count*batch_size,0].view(batch_size,1)
-            outR1 = data[count:batch_size+count*batch_size,1].view(batch_size,1)
+            inR1 = data[count:batch_size+count*batch_size,0].view(128,1)
+            outR1 = data[count:batch_size+count*batch_size,1].view(128,1)
             # print(inR1.shape)
             # print(outR1.shape)
             count = count + 1
             return inR1.to(device),outR1.to(device), count
         elif data_flag == 'R2':
-            inR2 = th.stack([data[count:batch_size+count*batch_size,1], data[count:batch_size+count*batch_size,2]], dim=1)
+            inR2 = th.stack([data[count:batch_size+count*batch_size,1], data[count:batch_size+count*batch_size,2]], dim=1).t()
             outR2 = data[count:batch_size+count*batch_size,3].view(batch_size,1)
             count = count + 1
         return inR2.to(device),outR2.to(device), count
@@ -77,6 +76,8 @@ def test(test_iterations,model_name):
             _outloss += loss.item()
         print('Test loss: {:.4f}\n'.format(_outloss.item()/test_iterations))
 
+def objective(zlearn, *args, **kwargs):
+    ''' This has to be representative of the objective, equation 2'''
 if __name__ == '__main__':
     loss_fn = th.nn.MSELoss()
     # device = th.device("cuda" if th.cuda.is_available() else "cpu")

### 2019-08-07 code/amortized_rs/memory_src/model.py

@@ -16,6 +16,7 @@ class density_estimator(nn.Module):
     def forward(self, x):
          # x is of shape [batch_size, input_dim]
         hidden = self.hidden(x) # return [batch_size//2, 2]
+        print(' Debug in model.py. Forward pass completed')
         return self.linear_mu(hidden), nn.Softplus(self.linear_std(hidden))
 
 

### 2019-08-07 code/amortized_rs/memory_src/hogwild.py
add import

@@ -4,6 +4,7 @@ from torch.utils.cpp_extension import load
 import torch as th
 from torch import optim
 import torch.distributions as dist 
+from time import strftime
 # from utils import RejectionDataset
 # from torch.utils.data import DataLoader
 import os
@@ -78,6 +79,7 @@ def test(test_iterations,model_name):
 
 def objective(zlearn, *args, **kwargs):
     ''' This has to be representative of the objective, equation 2'''
+
 if __name__ == '__main__':
     loss_fn = th.nn.MSELoss()
     # device = th.device("cuda" if th.cuda.is_available() else "cpu")

### 2019-08-07 code/amortized_rs/memory_src/hogwild.py
clean

@@ -1,22 +1,32 @@
 import torch.multiprocessing as mp
-from ..model import density_estimator
+from model import density_estimator
 from torch.utils.cpp_extension import load
 import torch as th
 from torch import optim
 import torch.distributions as dist 
 from time import strftime
-# from utils import RejectionDataset
-from time import strftime
-# from torch.utils.data import DataLoader
 import os
 import time
 
 amortized_rs = load(name="amortized_rs",
                     sources=["amortized_rs.cpp"])
 
+# no longer need write the batch loop in c - we can do that later
+def get_batch(data_flag, batch_size, count, load_data=False):
+    if not load_data:
+        data = th.stack([amortized_rs.f() for _ in range(batch_size)]) 
+    if data_flag == 'R1':
+        inR1 = data[count:batch_size+count*batch_size,0].view(128,1)
+        outR1 = data[count:batch_size+count*batch_size,1].view(128,1)
+        count = count + 1
+        return inR1.to(device),outR1.to(device), count
+    elif data_flag == 'R2':
+        inR2 = th.stack([data[count:batch_size+count*batch_size,1], data[count:batch_size+count*batch_size,2]], dim=1).t()
+        outR2 = data[count:batch_size+count*batch_size,3].view(batch_size,1)
+        count = count + 1
+    return inR2.to(device),outR2.to(device), count
+    
 def train(model, optimizer, loss_fn,  N, data_flag, batch_size, load_data=False):
-    # Construct data_loader, optimizer, etc.
-    # for it in range(max_it):
     model.train()
     n_samples = batch_size*N
     # data = th.stack([amortized_rs.f() for _ in range(batch_size)]) # hmm, shouldn't this be the number of training samples, rather than batch_size?
@@ -26,22 +36,6 @@ def train(model, optimizer, loss_fn,  N, data_flag, batch_size, load_data=False)
         data = data[0:n_samples,:]
     optimizer.zero_grad()
     count  = 0
-    # no longer need write the batch loop in c - we can do that later
-    def get_batch(data_flag, batch_size, count, load_data=False):
-        if not load_data:
-            data = th.stack([amortized_rs.f() for _ in range(batch_size)]) 
-        if data_flag == 'R1':
-            inR1 = data[count:batch_size+count*batch_size,0].view(128,1)
-            outR1 = data[count:batch_size+count*batch_size,1].view(128,1)
-            # print(inR1.shape)
-            # print(outR1.shape)
[…]

### 2019-08-07 paper/learning/notes.tex
updated

@@
 
-\section{Rejection Sampling}
+\section{Nested-Programs within Simulators}
+
+\begin{itemize}
+  \item Talk about how within simulators we have nested-structures which are \emph{nested-programs}
+  \item Define the notations used to define the processes within the simulator, in terms of the PPL framework
+  \item Talk about the problems that such nested-programs cause to inference procedures
+  \item Define the density of the simulator in terms of these sub-programs. 
+\end{itemize}
+
+To construct the mathematical framework in this setting we employ the notation of~\cite{rainforth2017automating}, 
+where \emph{raw samples} are represented by $ f_{a_i}(x_i | \eta)$ and can be thought of as prior 
+\textbf{To finish - Linked back to figure 1}
+In order to correctly characterize this problem we first provide a useful diagram demonstrating 
+the density of the system~\ref{fig:systemoverview}. \textbf{Lots to do}
+
+ \input{thesystem.tex}
+
+% In order to correctly characterize this problem we first provide a useful diagram demonstrating 
+% the density of the system.
+
+%  \input{thesystem.tex}
+
+\section{Amortized Rejection Sampling}
+
+One common problem in traditional simulators, \emph{programs}, outside of the standard probabilistic
+programming framework is that they rely heavily on rejection sampling 
+to generate samples from the target density. Not only is rejection sampling 
+computational inefficient, [especially if criteria in which samples are generated
+has been poorly calibrated] (may leave out this statement as it may make a reviewer
+question why would amortization work any better.), but it creates complications for inference.
+As the rejection sampling loops provide no explicit form of the target density,
+we can not simply evaluate the density and leverage statistically and computationally
+efficient algorithms such as MCMC sampling. This inhibits our ability to transform an arbitrary
+stochastic simulator into a probabilistic programming framework~\cite{baydin2018efficient}. 
+To this end, we derive both a mathematical and engineering based framework to facilitate 
+the amortization of rejection sampling loops in arbitrary simulators, which provides a path
+way to both leverage a larger, more efficient class of inference methods and convert arbitrary
+stochastic simulators into probabilistic programs systems.
+
+\subsection{Rejection Sampling}
 
@@
 
-\section{Amortized Rejection Sampling}
-
-One common problem in traditional simulators, \emph{programs}, outside of the standard probabilistic
-programming framework is that they rely heavily on rejection sampling 
-to generate samples from the target density. Not only is rejection sampling 
-computational inefficient, [especially if criteria in which samples are generated
-has been poorly calibrated] (may leave out this statement as it may make a reviewer
-question why would amortization work any better.), but it creates complications for inference.
-As the rejection sampling loops provide no explicit form of the target density,
-we can not simply evaluate the density and leverage statistically and computationally
-efficient algorithms such as MCMC sampling. This inhibits our ability to transform an arbitrary
-stochastic simulator into a probabilistic programming framework~\cite{baydin2018efficient}. 
-To this end, we derive both a mathematical and engineering based framework to facilitate 
-the amortization of rejection sampling loops in arbitrary simulators, which provides a path
-way to both leverage a larger, more efficient class of inference methods and convert arbitrary
-stochastic simulators into probabilistic programs systems.
-
 \begin{figure}
@@
 \end{figure}
-
-\subsection{Defining the Density and Sub-Densities within Simulators}
-
-To construct the mathematical framework in this setting we employ the notation of~\cite{rainforth2017automating}, 
-where \emph{raw samples} are represented by $ f_{a_i}(x_i | \eta)$ and can
-
-In order to correctly characterize this problem we first provide a useful diagram demonstrating 
-the density of the system.
-
- \input{thesystem.tex}
-
-% In order to correctly characterize this problem we first provide a useful diagram demonstrating 
-% the density of the system.
-
-%  \input{thesystem.tex}
 
@@
 the unknown densities that we wish to learn through an amortized approach.
-
+It is important to note that the expected running time of rejection sampler one, $R1$, 
+has an expected running time $\mathbb{E}(\mathcal{O}(R1)) = \infty$. 
+The simulated densities for both each rejection sampling block can be seen in Figure~\ref{fig:rs_simulated}.
 \begin{figure}
@@
 \end{figure}
-However, in order to do this, we must correctly specify our variational objective.
+
+In order to replace such rejection sampling blocks with learnt functions that we 
+can efficiently sample from, we must correctly specify our variational objective.
 
@@
 
-For $R1$ we have one input and one output, for $R2$ we have 2 inputs and 1 output. 
-For the loss function we use a simple mean squared error loss and use a the regressor the learn 
-and unknown mean $\mu$ and variance $\sigma^{2}$. This means that we are trying to fit a simple Normal distribution
-to each rejection sampling block. Once learned we can then efficiently sample from the parameterized normal distribution, which 
-represents our learned proposal $q(z_i, \ldots, z_{n} | \eta)$. 
-\paragraph{Training scheme} For each rejection sampling block we learn a proposal $q(z_1, \ldots, z_i | \eta)$, in this case 
-\textbf{Note: Add results here. Code almost complete}
+For $R1$ we have one input $z^{r1} \in \mathbb{R}^{1 \times 1}$ and one output $y^{r1} \in \mathbb{R}^{1 \times 1}$.
+For $R2$ we have two inputs $z^{r2} \in \mathbb{R}^{2 \times 1}$ and one output $y^{r2} \in \mathbb{R}^{1 \times 1}$.
+For training purposes we generate $n_{b}$ batches of $n_{t}$ samples for training. 
+This means that $z^{r1}_{batch}\mathbb{R}^{n_{b} \times 1}$. 
+For the loss function we use a simple mean squared error loss \textbf{We can try different loss 
+functions and optimizers} and use the
+neural network, $\Gamma(z, y)$, to learn a $\hat{\mu}$ and $\hat{\sigma}^{2}$ to parametize 
+a normal proposal. For $R1$, this means $\Gamma_{r1}(z,y) \rightarrow \hat{\mu_{r1}},  \hat{\sigma_{r1}}^{2}$  
+and so the proposal $q(z_{1} | z_{2}) = \mathcal{N}(\hat{\mu_{r1}},  \hat{\sigma_{r1}}^{2})$. 
+Once learned we can then efficiently sample from the parameterized normal distribution, which
+represents the density of the sub-program represented by $R1$. 
+This is then replaced in the simulator, during inference time.
+
+\textbf{Add results from current experiments}
 
 \subsection{2nd Baseline}
-Instead of using a single guassain, use a proposal that is a mixture of gaussians. 
-\texbf{TODO: Add math tomorrow}
+Instead of using a single Gaussian, use a proposal that is a mixture of Gaussians. 
+That is, for $R1$,   $q(z_{1} | z_{2}) = \sum^{N}_{i=1}K_{i}\mathcal{N}(\hat{\mu^{i}_{r1}},  \hat{\sigma^{i}_{r1}}^{2})$
+under the constraint that $\sum^{N}_{i=1}K_{i} = 1$. Maybe $N$ can be treated as a hyper-parameter. 
+\textbf{TODO: Add math tomorrow}
 \subsection{More complicated: Conditional Density estimation}
@@
 
-Flows availabile are MVP and Neural Spline flows [add citations].
-
-for other techniques ask Guens.
+Flows available are Real-NVP~\cite{dinh2016density} and Neural Spline flows~\cite{durkan2019neural}.
+There are several others, for other techniques ask Gunes.
 
@@
 \section{ Questions for Tom}
+
+\textbf{New questions to ask Tom}
+
+\begin{itemize}
+  \item Isn't what we are doing just conditional density estimation? It seems
+  like there is a lot of prior work here. I.e see this paper here for a good overview ~\cite{greenberg2019automatic}
+  \item It does seem like this field has a large extensive past, but we still have a small neesh .
+\end{itemize}
+
+\textbf{Old questions for Tom}
 \begin{itemize}
@@
 
-New questions to ask Tom
-
-\begin{itemize}
-  \item Isn't what we are doing just conditional density estimation? It seems
-  like there is a lot of prior work here. I.e see this paper here for a good overview ~\cite{greenberg2019automatic}
-  \item It does seem like this field has a large extensive past, but we still have a small neesh .
-\end{itemize}
-=======
-
->>>>>>> e2910a3683b657af5fa91292c678419ec3b427c7
 \section{Additional TODOs}

### 2019-08-07 code/amortized_rs/memory_src/amortized_rs_std.cpp
added amortized cpp std code

@@ -2,18 +2,30 @@
 
 #include <iostream>
 #include <limits>
+#include <random>
 
 using namespace at;
 
 //https://pytorch.org/cppdocs/api/namespace_at.html#namespace-at
-Scalar R1(Scalar z_1, unsigned int i);
-Scalar R2(Scalar z_2, Scalar z_3);
+Tensor f(void);
+Tensor R1(double z_1);
+Tensor R2(double z_2, double z_3);
 
 
-at::Tensor f(void) {
-  Scalar z_1 = at::empty({1}, kFloat).normal_(0.0, 1.0).item();
-  Scalar z_2;
+Tensor batch_f(unsigned int bs){
+    auto ret = torch::empty({batch_size, 4}, kFloat);
+    for(int i=0;i<bs;i++){
+        ret_view = ret.narrow(i, 0, -1)
+        ret_view = f()
+    }
+    return ret
+}
 
+inline double f(void) {
+  std::default_random_engine generator;
+  std::normal_distribution<double> distribution(0.0,1.0);
+  double z_1 = distribution(generator);
+  double z_2;
   int ctr = 0;
   do {
     z_2 = R1(z_1, 0);
@@ -32,16 +44,8 @@ at::Tensor f(void) {
   return ret; //torch::from_blob(data, {4});
 }
 
-Scalar R1(Scalar z_1, unsigned int i) {
+inline double R1(double z_1) {
     Scalar temp;
-//    if (temp.to<float>() > 0){
-//        return temp;
-//    } else if (i == 100){
-//        return CPU(kFloat).scalarTensor(std::numeric_limits<float>::infinity()).item();
-//    } else {
-//        i += 1;
-//        return R1(z_1, i);
-//    }
     long j = 0;
     while (true){
         temp = at::empty({1}, kFloat).normal_(z_1.to<float>(), 1.0).item();
@@ -55,21 +59,12 @@ Scalar R1(Scalar z_1, unsigned int i) {
     }
 }
 
-Scalar R2(Scalar z_2, Scalar z_3) {
+inline double R2(Scalar z_2, Scalar z_3) {
     Scalar temp = at::empty({1}, kFloat).normal_(z_3.to<float>(), 1.0).item();
     while (temp.to<float>()  < z_2.to<float>() ) {
         temp = at::empty({1}, kFloat).normal_(z_3.to<float>(), 1.0).item();
     }
     return temp;
-}
-
[…]

### 2019-08-07 paper/learning/notes.tex

@@
 \label{sec:intro}
-
-One challenge in performing inference in such models is the combinatorial increase in \emph{paths}
-that can be taken within the given probabilistic model.
-This is because each member of the population has 
-a bounded, but large number of decisions to make. 
-In addition to this, each member of the population can interact
-with other members of the population, which presents challenges for current inference methods.
-Because in many simulated settings we cannot correctly construct the functional form of the density of the 
-desired program. 
-This has spawned a large body of research in recent years to understand how to perform 
-inference in likelihood free methods~\cite{greenberg2019automatic} \textbf{Add normalising flow citations}
+Simulators arise in a myriad of scientific and industrial settings; epidemiology, physics, engineering, climate modelling and so forth [add citations].
+One interesting class of simulators are population-based simulators, using in epidemiology, reinforcement learning
+and financial modelling [add citations].
+Performing inference in such population-based simulators is challenging, because we have nested-programs, 
+which come in the form of rejection sampling blocks. 
+In addition to this, each population member can traverse a number of different paths and 
+can interact with other members, in certain instances, which can lead to a combinatorial number of paths 
+to explore. 
+Equally, in many simulated settings we cannot correctly construct the functional form of the density of the desired program,
+since we do not have explicit access to the density of the \emph{program}.
+Where program density refers to the joint density constructed by the program, which in this case is represented by the simulator.
+This inability to evaluate the joint density has spawned research into how to perform 
+inference in this likelihood free setting~\cite{greenberg2019automatic} \textbf{Add normalising flow citations}
 for conditional density estimation. However, whilst many of these methods look to generate a surragote 
@@
   % \begin{minipage}[c]{0.67\textwidth}
-  \includegraphics[]{amoritized_rejection.pdf}
+  \includegraphics[width=0.6\textwidth, height=0.6\textheight,keepaspectratio]{amoritized_rejection.pdf}
   % \end{minipage}\hfill
@@
 the unknown densities that we wish to learn through an amortized approach.
-It is important to note that the expected running time of rejection sampler one, $R1$, 
-has an expected running time $\mathbb{E}(\mathcal{O}(R1)) = \infty$. 
+It is important to note that the expected running time of the rejection sampler is expected to geometric mean, we can show this as follows:
+\begin{equation}
+\text{derivation}
+\end{equation}
+
+In addition to this, if we have direct access to a rejection sampling loop then it is indeed 
+possible to construct the density, up to a constant of proportionality. 
+That is:
+\begin{align*}
+y \sim q(\theta) \\
+do something \\
+\alpha = p_{accept}(y,\theta) \\
+\mathtt{if} Bern(\alpha) = 1: \\
+return y
+\end{align*}
+thus the density of $y,\theta \propto q(\theta, y) \alpha$.
+However, in order to do construct this we need the input $\theta$, $y$, $\alpha$ and the distributions utilised for the rejection sampling loop. 
+For any arbitrary simulator, written in a generic code base, this is an incredibly hard 
+task to do, because rejection sampling blocks within simulators are not necessarily written in a standard template
+\textbf{Add examples in a supplement.  Need to demonstrate to YW that this is challenging to do, even if you know where 
+the rejection sampling loops are.
+Hypothesis, RS loops are not written in a templated form}.
+With the PPX protocols previously developed in \cite{baydin2018efficient,gram2019hijacking} we can however,
+extract all variables, other than $\alpha$ in an automated way.
+However, without being able to easily extract $\alpha$ this means that we are unable to 
+correctly build the form of the density and thus we limited in our ability to leverage 
+powerful inference methods such as MCMC and Variational inference.
+Thus, we propose another approach, automating the amortization of rejection sampling loops. 
+To learn functions that we can easily sample from and use to construct the density of the entire
+simulator.
+
 The simulated densities for both each rejection sampling block can be seen in Figure~\ref{fig:rs_simulated}.
@@
   \centering
-  \includegraphics{rs_plots.png}
+  \includegraphics[width=0.2\textwidth, height=0.2\textheight, keepaspectratio]{rs_plots.png}
   \caption{ Simulated densities for rejection sampling block 1, \emph{blue} and rejection
@@
 
+\subsection{Ground truth}
+
+Because in this simple case we have direct access to $\alpha$, the rejection sampler acceptance probability,
+we can calculate the true density up to a constant of proportionality. 
+\textbf{I will add this tomorrow}. I had no idea about this, until my meeting with 
+YW today.
+
 \subsection{Baseline}
@@
 under the constraint that $\sum^{N}_{i=1}K_{i} = 1$. Maybe $N$ can be treated as a hyper-parameter. 
-\textbf{TODO: Add math tomorrow}
+Or choose particular values.
 \subsection{More complicated: Conditional Density estimation}
+
+\subsection{3rd Baseline}
+
+Use the variational objective constructed above.
 

### 2019-08-07 code/amortized_rs/memory_src/amortized_rs.cpp
reworked pytorch cpp

@@ -6,9 +6,17 @@
 using namespace at;
 
 //https://pytorch.org/cppdocs/api/namespace_at.html#namespace-at
+at::Tensor f(void);
 Scalar R1(Scalar z_1, unsigned int i);
 Scalar R2(Scalar z_2, Scalar z_3);
 
+torch::Tensor batch_f(unsigned int bs){
+    auto ret = torch::empty({bs, 4}, kFloat);
+    for(int i=0;i<bs;i++){
+        ret[i] = f();
+    }
+    return ret;
+}
 
 at::Tensor f(void) {
   Scalar z_1 = at::empty({1}, kFloat).normal_(0.0, 1.0).item();
@@ -48,7 +56,9 @@ Scalar R1(Scalar z_1, unsigned int i) {
         if (temp.to<float>() > 0){
             return temp;
         } else if (j == 10000){
-            return CPU(kFloat).scalarType(std::numeric_limits<float>::infinity()).item();
+            // return at::CPU(kFloat).scalarType(std::numeric_limits<float>::infinity()).item();
+            //return at::from_blob({std::numeric_limits<float>::infinity()}, {1}).item();
+            return Scalar(std::numeric_limits<float>::infinity());
         } else {
             j += 1;
         }
@@ -63,14 +73,6 @@ Scalar R2(Scalar z_2, Scalar z_3) {
     return temp;
 }
 
-Tensor batch_f(unsigned int bs){
-    auto ret = torch::empty({batch_size, 4}, kFloat);
-    for(int i=0;i<bs;i++){
-        ret_view = ret.narrow(i, 0, -1)
-        ret_view = f()
-    }
-    return ret
-}
 
 PYBIND11_MODULE(TORCH_EXTENSION_NAME, m) {
   m.def("f", &f, "f");

### 2019-08-07 code/amortized_rs/memory_src/amortized_rs.py

@@ -10,10 +10,18 @@ amortized_rs = load(name="amortized_rs",
 #     print(a)
 
 f_cpp = amortized_rs.f
+batch_f = amortized_rs.batch_f
 import timeit
 torch.manual_seed(7)
 print(f_cpp())
+torch.manual_seed(7)
+print(batch_f(1))
+torch.manual_seed(7)
 print(timeit.timeit("f_cpp()", setup="from __main__ import f_cpp", number=2000))
+torch.manual_seed(7)
+print(timeit.timeit("[f_cpp() for _ in range(2000)]", setup="from __main__ import f_cpp", number=1))
+torch.manual_seed(7)
+print(timeit.timeit("batch_f(2000)", setup="from __main__ import batch_f", number=1))
 
 # pure python
 import math
@@ -60,3 +68,14 @@ import timeit
 torch.manual_seed(7)
 print(f())
 print(timeit.timeit("f()", setup="from __main__ import f", number=2000))
+
+amortized_rs_std = load(name="amortized_rs_std",
+                    sources=["amortized_rs_std.cpp"])
+
+f_cpp = amortized_rs_std.f
+batch_f = amortized_rs_std.batch_f
+import timeit
+torch.manual_seed(7)
+print(f_cpp())
+print(timeit.timeit("f_cpp()", setup="from __main__ import f_cpp", number=2000))
+print(timeit.timeit("batch_f(2000)", setup="from __main__ import batch_f", number=1))

### 2019-08-07 code/amortized_rs/memory_src/amortized_rs_std.cpp

@@ -7,62 +7,65 @@
 using namespace at;
 
 //https://pytorch.org/cppdocs/api/namespace_at.html#namespace-at
-Tensor f(void);
-Tensor R1(double z_1);
-Tensor R2(double z_2, double z_3);
+torch::Tensor f(void);
+double R1(double z_1, std::default_random_engine &generator);
+double R2(double z_2, double z_3, std::default_random_engine &generator);
 
 
-Tensor batch_f(unsigned int bs){
-    auto ret = torch::empty({batch_size, 4}, kFloat);
+torch::Tensor batch_f(unsigned int bs){
+    auto ret = torch::empty({bs, 4}, kFloat);
     for(int i=0;i<bs;i++){
-        ret_view = ret.narrow(i, 0, -1)
-        ret_view = f()
+        auto ret_view = ret.narrow(0, i, 1);
+        ret_view = f();
     }
-    return ret
+    return ret;
 }
 
-inline double f(void) {
-  std::default_random_engine generator;
-  std::normal_distribution<double> distribution(0.0,1.0);
+inline torch::Tensor f(void) {
+  std::random_device rd;
+  std::default_random_engine generator( rd() );
+  std::normal_distribution<double> distribution(0.0, 1.0);
   double z_1 = distribution(generator);
   double z_2;
   int ctr = 0;
   do {
-    z_2 = R1(z_1, 0);
+    z_2 = R1(z_1, generator);
     ctr += 1;
-  } while (z_2.to<float>() == std::numeric_limits<float>::infinity());
-  //std::cout << ctr;
-  Scalar z_3 = at::empty({1}, kFloat).uniform_(0,2).item();
-  Scalar z_4 = R2(z_2, z_3);
-  //std::cout << z_1 << "|" << z_2 << "|" << z_3 << "|" << z_4;
-  //float data[] = {z_1.to<float>(), z_2.to<float>(), z_3.to<float>(), z_4.to<float>()};
+  } while (z_2 == std::numeric_limits<float>::infinity());
+  // double z_3 = at::empty({1}, kFloat).uniform_(0,2).item();
+  std::uniform_real_distribution<double> uni_dist(0.0, 2.0);
+  double z_3 = uni_dist(generator);
+  double z_4 = R2(z_2, z_3, generator);
   auto ret = torch::empty({4}, kFloat);
   ret[0] = z_1;
   ret[1] = z_2;
   ret[2] = z_3;
   ret[3] = z_4;
[…]

### 2019-08-08 code/amortized_rs/memory_src/hogwild.py
updated

@@ -15,16 +15,28 @@ amortized_rs = load(name="amortized_rs",
 def get_batch(data_flag, batch_size, count, load_data=False):
     if not load_data:
         data = th.stack([amortized_rs.f() for _ in range(batch_size)]) 
-    if data_flag == 'R1':
-        inR1 = data[count:batch_size+count*batch_size,0].view(128,1)
-        outR1 = data[count:batch_size+count*batch_size,1].view(128,1)
+    if data_flag == 'R1' and not load_data:
+        inR1 = data[0:batch_size,0]
+        outR1 = data[0:batch_size,1]
+        count = count + 1 # not actually need if load_data == True
+        return inR1.to(device),outR1.to(device), count
+    elif data_flag == 'R1' and load_data:
+        inR1 = data[count:batch_size+count*batch_size,0]
+        outR1 = data[count:batch_size+count*batch_size,1]
         count = count + 1
         return inR1.to(device),outR1.to(device), count
-    elif data_flag == 'R2':
-        inR2 = th.stack([data[count:batch_size+count*batch_size,1], data[count:batch_size+count*batch_size,2]], dim=1).t()
-        outR2 = data[count:batch_size+count*batch_size,3].view(batch_size,1)
+    elif data_flag == 'R2'and not load_data:
+        inR2 = th.stack([data[0:batch_size,1], data[0:batch_size,2]], dim=1)
+        outR2 = data[0:batch_size,3]
         count = count + 1
-    return inR2.to(device),outR2.to(device), count
+        return inR2.to(device),outR2.to(device), count
+    elif data_flag == 'R2' and load_data:
+        inR2 = th.stack([data[count:batch_size+count*batch_size,1], data[count:batch_size+count*batch_size,2]], dim=1)
+        outR2 = data[count:batch_size+count*batch_size,1]
+        count = count + 1
+        return inR2.to(device), outR2.to(device), count
+
+
     
 def train(model, optimizer, loss_fn,  N, data_flag, batch_size, load_data=False):
     model.train()
@@ -36,6 +48,7 @@ def train(model, optimizer, loss_fn,  N, data_flag, batch_size, load_data=False)
         data = data[0:n_samples,:]
     optimizer.zero_grad()
     count  = 0
+    _outLoss = 0
 
     # The network needs to learn z1 -> z2 and z2,z3 -> z4
     for i in range(N):
[…]

### 2019-08-08 code/amortized_rs/memory_src/model.py

@@ -6,9 +6,9 @@ class density_estimator(nn.Module):
     def __init__(self, inputSize, outputSize):
         super(density_estimator, self).__init__()
         self.hidden = nn.Sequential(
-            nn.Linear(inputSize, inputSize//2),
+            nn.Linear(inputSize, inputSize//2,bias=True),
             nn.ReLU(inputSize//2),
-            nn.Linear(inputSize//2, inputSize//(inputSize//2)) # ensures that the output would be 2, assuming inputsize = 2^n
+            nn.Linear(inputSize//2, inputSize//(inputSize//2),bias=True) # ensures that the output would be 2, assuming inputsize = 2^n
             )
         self.linear_mu = nn.Linear( inputSize//(inputSize//2), outputSize)
         self.linear_std = nn.Linear( inputSize//(inputSize//2), outputSize)
@@ -16,7 +16,6 @@ class density_estimator(nn.Module):
     def forward(self, x):
          # x is of shape [batch_size, input_dim]
         hidden = self.hidden(x) # return [batch_size//2, 2]
-        print(' Debug in model.py. Forward pass completed')
-        return self.linear_mu(hidden), nn.Softplus(self.linear_std(hidden))
+        return self.linear_mu(hidden), F.softplus(self.linear_std(hidden))
 
 

### 2019-08-08 code/amortized_rs/memory_src/hogwild.py
updated

@@ -38,7 +38,7 @@ def get_batch(data_flag, batch_size, count, load_data=False):
 
 
     
-def train(model, optimizer, loss_fn,  N, data_flag, batch_size, load_data=False):
+def train(model, optimizer, loss_fn,  N, data_flag, batch_size, rank, load_data=False):
     model.train()
     n_samples = batch_size*N
     # data = th.stack([amortized_rs.f() for _ in range(batch_size)]) # hmm, shouldn't this be the number of training samples, rather than batch_size?
@@ -63,12 +63,12 @@ def train(model, optimizer, loss_fn,  N, data_flag, batch_size, load_data=False)
 
             _outLoss += _loss.item()
 
-            if i % 100 == 0:
+            if i % 10 == 0:
                 print("iteration {}: loss: {:6.3f}".format(i,_outLoss))
             # print('====> Epoch: {:03d} Train loss: {:.4f}'.format(epoch, outloss))
     if not os.path.exists('../model/'):
         os.makedirs('../model/')
-    fname = '../model/model_{}_rejectionBlock_{}'.format(strftime("%Y-%m-%d_%H-%M"), data_flag)
+    fname = '../model/model_{}_rejectionBlock_{}_process_{}'.format(strftime("%Y-%m-%d_%H-%M"), data_flag, rank)
     th.save(model.state_dict(), fname)
     print(' Model is saved at : {}'.format(fname))
 
@@ -93,6 +93,7 @@ if __name__ == '__main__':
     # device = th.device("cuda" if th.cuda.is_available() else "cpu")
     device = th.device("cpu")
     lr = 4e-5
+    load_data=False
     # inputs = DataLoader(RejectionDataset(split='train', l_data=128, train_percentage=0.8,fname_test, fname_train, InIndx, OutIndx), batch_size=128, shuffle=True, norma num_workers=cpu_count-2)
     # outputs = DataLoader(RejectionDataset(split='test', l_data=128, test_percentage=0.8,fname_test, fname_train, InIndx, OutIndx), batch_size=128, shuffle=True,num_workers=cpu_count-2)
     batch_size = 128
@@ -101,16 +102,15 @@ if __name__ == '__main__':
     outputSize = 1 # for R1
     model = density_estimator(inputSize, outputSize)
     num_processes = mp.cpu_count() - 1
-    N = 10000
+    N = 100
     # optimizer = optim.Adam(filter(lambda p: p.requires_grad, model.parameters()), lr=lr, amsgrad=True)
[…]

### 2019-08-08 code/amortized_rs/memory_src/hogwild.py
resolve loss issue

@@ -64,7 +64,8 @@ def train(model, optimizer, loss_fn,  N, data_flag, batch_size, rank, load_data=
             _outLoss += _loss.item()
 
             if i % 10 == 0:
-                print("iteration {}: loss: {:6.3f}".format(i,_outLoss))
+                avgLoss = _outLoss / (i+1)
+                print("iteration {}: Average loss: {:6.3f}".format(i,avgLoss))
             # print('====> Epoch: {:03d} Train loss: {:.4f}'.format(epoch, outloss))
     if not os.path.exists('../model/'):
         os.makedirs('../model/')

### 2019-08-08 paper/learning/notes.tex
added todo list in section 4

@@
 
+\paragraph{TODO list}
+\begin{itemize}
+  \item Implement baselines
+  \item Implement plotting scripts and evaluation scripts
+  \item testing code and infrastructure
+  \item litrature review on any methods to learn rejection sampling methods, 
+  and conditional density estimation. See this paper for kernel approximation 
+  methods~\cite{erraqabi2016pliable}. In addition to the methods listed in the 
+  rejection sampling subsection. 
+\end{itemize}
 \subsection{Ground truth}

### 2019-08-12 paper/learning/notes.tex
updated

@@
 
+\paragraph{observations} As the functions are relatively simple the loss functions it gets t
+to a loss of around 0.6 and stops decreasing. Although, this seems low, I don't think that it is
+actually good, because that would mean on average, for each data point there is a loss of $\pm$ 0.6
+which per data point is quite bad. That may be for a number of reasons, using MSE loss, using a simple
+model architecture. It really struggles to model the tails. It may require a much larger batch input size. 
+. Increasing the batch size helped to correct for the peaked density, however the behavior is still odd.
+The peaked behaviour still does not disappear, this, I believe is due to the loss function averaging 
+over all points,
+
  \textbf{Add results from current experiments}

### 2019-08-12 code/amortized_rs/memory_src/hogwild.py

@@ -6,6 +6,8 @@ from torch import optim
 import torch.distributions as dist 
 from time import strftime
 import os
+import seaborn as sns
+import matplotlib.pyplot as plt
 import time
 
 amortized_rs = load(name="amortized_rs",
@@ -26,7 +28,7 @@ def get_batch(data_flag, batch_size, count, load_data=False):
         count = count + 1
         return inR1.to(device),outR1.to(device), count
     elif data_flag == 'R2'and not load_data:
-        inR2 = th.cat([data[0:batch_size,1], data[0:batch_size,2]], dim=0).view(1,256)
+        inR2 = th.cat([data[0:batch_size,1], data[0:batch_size,2]], dim=0).view(1,batch_size*2)
         outR2 = data[0:batch_size,3].view([1,batch_size])
         count = count + 1
         return inR2.to(device),outR2.to(device), count
@@ -36,6 +38,11 @@ def get_batch(data_flag, batch_size, count, load_data=False):
         count = count + 1
         return inR2.to(device), outR2.to(device), count
 
+def center_data(data):
+    mean = th.mean(data,dim =0)
+    zeroed = data - mean*th.ones(data.shape)
+    stddata  = th.std(data, dim=0)
+    return zeroed / stddata
 def train(model, optimizer, loss_fn,  N, data_flag, batch_size, rank, load_data=False):
     model.train()
     n_samples = batch_size*N
@@ -46,12 +53,16 @@ def train(model, optimizer, loss_fn,  N, data_flag, batch_size, rank, load_data=
     count  = 0
     _outLoss = 0
 
+    if not os.path.exists('../plots/'):
+        os.makedirs('../plots/')
+
     # The network needs to learn z1 -> z2 and z2,z3 -> z4
     for i in range(N):
             inData, outData, count =get_batch(data_flag, batch_size, count)
             proposal = dist.Normal(*model(inData))
-            pred = proposal.rsample(sample_shape=[batch_size]).view(1,128)
+            pred = proposal.rsample(sample_shape=[batch_size]).view(1,batch_size)
             optimizer.zero_grad()
+            plt.show()
             _loss = loss_fn(outData, pred)
             _loss.backward()
             optimizer.step()
@@ -59,7 +70,15 @@ def train(model, optimizer, loss_fn,  N, data_flag, batch_size, rank, load_data=
             _outLoss += _loss.item()
             # if _outLoss == nan:
             #     break
-            if i % 100 == 0:
+            if i % 1 == 0:
[…]

### 2019-08-12 code/amortized_rs/memory_src/inference.py

@@ -9,7 +9,7 @@ import os
 import time
 import shutil
 import logging
-
+import seaborn as sns
 amortized_rs = load(name="amortized_rs",
                     sources=["amortized_rs.cpp"])
 
@@ -206,6 +206,7 @@ class Inference():
                 inData, outData, count = self.get_batch(self.address, self.batchSize, count)
                 proposal = dist.Normal(*model(inData))
                 pred = proposal.rsample(sample_shape=[self.batchSize]).view(1, 128)
+
                 # learns a \hat{y} for the whole batch and then generates n_batch_size samples to predict the output.
                 _loss = self.loss_fn(outData, pred)
                 _outloss += _loss.item()

### 2019-08-14 paper/learning/notes.tex
stuff to ask tom

@@
 \end{align*}
-thus the density of $y,\theta \propto q(\theta, y) \alpha$.
+thus the density of $y,\theta \propto q(\theta) \alpha$.
 However, in order to do construct this we need the input $\theta$, $y$, $\alpha$ and the distributions utilised for the rejection sampling loop. 
@@
 \begin{equation}
+  \label{eq:objective}
   J_{r}(\eta_r) = -\mathbb{E}_{p(x)}[\sum^{n_x}_{j=1}\mathbb{I}(a_j = r)\ln(q_r(x_j;\eta_r, \phi_j))]
@@
   rejection sampling subsection. 
+\end{itemize}
+
+\paragraph{Tom Questions}
+
+\begin{itemize}
+  \item I don't understand how the objective function is comparing the generated samples from the density 
+  estimator to the ground truth, at least in the way it is currently defined? 
+  \item All RS references I look at state that we must sample from a uniform distribution and
+  then check if our proposed value is less than that uniformly sampled point, yet our RS do not have this,
+  nor what is the acceptance probability within our rejection sampling statements? The proof for rejection sampling seems to rely on the fact that you are drawing from a uniform distribution.
+
   \end{itemize}
@@
 The peaked behaviour still does not disappear, this, I believe is due to the loss function averaging 
-over all points,
+over all points.
 
@@
 \begin{itemize}
+  \item How is Equation~\ref{eq:}
   \item Isn't what we are doing just conditional density estimation? It seems
   like there is a lot of prior work here. I.e see this paper here for a good overview ~\cite{greenberg2019automatic}
-  \item It does seem like this field has a large extensive past, but we still have a small neesh .
+  \item It does seem like this field has a large extensive past, but we still have a small niche.
 \end{itemize}

### 2019-08-14 code/amortized_rs/memory_src/hogwild.py
save plots

@@ -71,9 +71,15 @@ def train(model, optimizer, loss_fn,  N, data_flag, batch_size, rank, load_data=
             # if _outLoss == nan:
             #     break
             if i % 1 == 0:
-                plt.show()
-                sns.distplot(pred[0, :].detach().numpy(), kde=True, color='r')
-                sns.distplot(outData[0, :].detach().numpy(), kde=True, color='b')
+                # plt.show()
+                ax= sns.distplot(pred[0, :].detach().numpy(), kde=True, color='r',label='pred')
+                ax = sns.distplot(outData[0, :].detach().numpy(), kde=True, color='b', label='ground truth')
+                ax.legend()
+                ax.set_title('Iteration_{}_Rejection_block_{}'.format(i,data_flag))
+                fig = ax.get_figure()
+                fname= '../plots/{}_compare_pred_vs_gt_iteration_rejection_block_{}.png'.format(strftime("%M:%S"),data_flag)
+                fig.savefig(fname=fname)
+
                 # plot_pred= sns.distplot(pred[0, :].detach().numpy(), kde=True, color='r')
                 # plot_true =sns.distplot(outData[0, :].detach().numpy(), kde=True, color='b')
                 # fnamePred = '../plot/' + 'predicted_iteration_{}_process_{}_rejectionBlock_{}'.format(i, rank, data_flag)
@@ -119,7 +125,7 @@ if __name__ == '__main__':
     # device = th.device("cuda" if th.cuda.is_available() else "cpu")
     device = th.device("cpu")
     lr = 0.001
-    momentum= 0.75
+    momentum= 0.9
     load_data=False
     batch_size = 2**9
     data_flag= 'R1'
@@ -131,12 +137,12 @@ if __name__ == '__main__':
         inputSize = batch_size
     model = density_estimator(inputSize, outputSize)
     num_processes = mp.cpu_count()
-    N = 1000
+    N =  200
     trainOn = True
     # loss_fn = th.nn.CosineEmbeddingLoss()
     loss_fn = th.nn.MSELoss()
-    optimizer = optim.Adam(filter(lambda p: p.requires_grad, model.parameters()), lr=lr, amsgrad=True)
-    # optimizer = optim.SGD(model.parameters(), lr=lr, momentum=momentum)
+    # optimizer = optim.Adam(filter(lambda p: p.requires_grad, model.parameters()), lr=lr, amsgrad=True)
[…]

### 2019-08-15 paper/learning/notes.tex
save plots
adding SVI notes and example - still to finish

@@
 It is important to note that the expected running time of the rejection sampler is expected to geometric mean, we can show this as follows:
-\begin{equation}
-\text{derivation}
-\end{equation}
+\textbf{Bradley:  You only arrive at this, if the density is uniform}
+\begin{align*}
+P( X= x) = \sum^{\infty}_{n=1} P(\text{reject $n-1$ times}, \text{draw} Y = x) \\
+\text{todo} 
+\end{align*}
 
@@
 
+\section{Stochastic Variational Inference}
+
+Using the standard notations we define the joint as follows:
+\begin{equation}
+  p(x, z ; \theta) = p_{\theta}(x,z) = p_{\theta}(x| z)p_{\theta}(z)
+\end{equation}
+
+We assume that the distributions that create the \emph{model} $p_{\theta}(x, z)$ can be:
+\begin{enumerate}
+  \item We can sample from each $p_i$
+  \item We can compute the pointwise log pdf of $p_i$
+  \item $p_i$ is differentiable w.r.t the parameters $\theta$
+  \end{enumerate}
+
+\subsection{Learning the model}
+
+Our objective will either want to maximize, or minimize, depending on what we want to discover.
+
+\begin{equation}
+  \theta_{\text{min} \backslash \text{max} } = \text{argmax$\backslash$min}_{\theta}\log p_\theta(x)
+\end{equation}
+
+where the log evidence $p_\theta(x)$ is given by:
+\begin{equation}
+  \log p_{\theta}(x) = \log \int dz p_\theta(x,z)
+\end{equation}
+
+This is a doubly hard problem as even if $\theta$ is fixed the integral over the set
+of latent variables defined by $z$ is intractable. 
+Even if we know how to calculate the log evidence for all $\theta$, maximising or minimising, 
+the log evidence as a function of $\theta$ is in general a difficult non-convex problem.
+
+In addition to finding the parameters $\theta$ we also require the posterior over
+the latent variables $z$:
+\begin{equation}
+  p_{\theta_{\text{max$\backslash$min}}}(z|x) \frac{p_{\theta}(x,z)}{\int dz p_{\theta_{\text{max$\backslash$min}}}(x,z)} 
+\end{equation}
+
+To this in the variational inference setting we learn an approximation to $p_{\theta_{\text{max$\backslash$min}}}(z|x)$
+denoted by $q_\phi(z)$ where $\phi$ are the variational parameters.
+
+\section{The Objective}
+
+In many of these settings the ELBO loss characterises the problem that we 
+aim to solve.
+
+\begin{equation}
+  ELBO = \mathbb{E}_{q_{\phi}(z)} [ \log p_{\theta}(x,z) - \log q_{\phi}(z) ]
+\end{equation}
+and we have for all choices of $\theta$ and $\phi$ that:
+\begin{equation}
+  \log p_{\theta}(x) \geq ELBO 
+\end{equation}
 \section{An Example: Modelling Parasite Densities in Simulation}

### 2019-08-15 paper/learning/notes.tex
updated

@@
 
-In the standard setting we let $X$ be a random variable whose density is $f(x)$ that we cannot easily 
-sample from, but can evaluate. Using a random variable $Z$, whose density is $g(z)$  that we can efficiently
-sample from, then we can generate a sample $X$ by sampling $Z$ and accept $Z$ with probability
+There are two different forms of rejection samplers found with in simulators and program code. 
+In both instances we we have $X$, a random variable whose density is $f(x)$ that we cannot easily 
+sample from, but can evaluate.
+We then define a proposal using a random variable $Z$, whose density is $g(z)$  that we can efficiently
+sample from.
+Where the two methods diverge is in how the acceptance probability is defined and how the samples are generated.
+In the standard statistical setting, see algorithm~\ref{alg:rsstats}, we generate a sample $X$ by sampling $Z$ and accept $Z$ with probability
 $p_{accept} = \frac{f(x)}{Mg(x)}$, which we can do until acceptance. 
-M is constant that has a finite bound, $M \in (1, \infty)$, on the likelihood ratio $\frac{f(x)}{g(x)}$ over the 
-support of $X$. $M$ must also satify the following inequality $f(x) \leq M g(x)  \forall x$. 
+Here $M$ is a constant that has a finite bound, $M \in (1, \infty)$ and the likelihood ratio $\frac{f(x)}{g(x)}$ is defined over the 
+support of $X$. 
+$M$ must also satisfy the following inequality $f(x) \leq M g(x)  \forall x$. 
 This implies that the support of $Z$ must also contained the support of $X$, that is
 $g(x) > 0$ whenever $f(x)>0$. 
+We construct this acceptance probability by sampling from a uniform distribution defined on the interval 
+$(0,1)$, which the proof ultimately relies on to ensure that a sample generated from the rejection sampler directly 
+corresponds to a sample from the true, normalized, target density $f(x)$. 
+But, provided we can calculate $P_{\text{accept}}(\mathcal{D} \leq \mathcal{C})$ for the given density, or mass, function
+ $\mathcal{D}$ and condition $\mathcal{C}$, we can also satisfy the standard proof with a little more work.
+ However, as we will demonstrate, rejection samplers 
+ need not take this form and arise in many nested structures within probabilistic programs~\cite{rainforth2018nesting} and simulators~\cite{smith2008towards,bershteyn2018implementation,gleisberg2009event}.
+
 The standard algorithm is defined as follows:
 \begin{algorithm}
+  \label{alg:rsstats}
   \small
@@
 chosen the computational running time of such rejection statements can be shown to have an
-expectation equal to infinity. Thus, the ability to learn and replace such rejection
+expectation equal to infinity. \textbf{Bradley: Is this true?}
+
+Thus, the ability to learn and replace such rejection
 sampling blocks with learnt surrogate functions, that are expressive enough to 
@@
 
-\paragraph{Tom Questions}
-
-\begin{itemize}
-  \item I don't understand how the objective function is comparing the generated samples from the density 
-  estimator to the ground truth, at least in the way it is currently defined? 
-  \item All RS references I look at state that we must sample from a uniform distribution and
-  then check if our proposed value is less than that uniformly sampled point, yet our RS do not have this,
-  nor what is the acceptance probability within our rejection sampling statements? The proof for rejection sampling seems to rely on the fact that you are drawing from a uniform distribution.
-
-  \end{itemize}
 \subsection{Ground truth}
@@
 over all points.
+
+After my meeting with Tom, by using a loss function the problem of the network collapsing on a particular value was always going to 
+happen as we should be optimizing our function via some objective, rather than looking at point-wise differences.
+
+We had defined an objective above but it was not clear how this was supposed to be implemented in practice with the given notations.
+
+We discussed at length how to correctly define the objective, I have now updated the objective with the one derivied in the revised discussion,
+along with additional notes on rejection sampling.
 
@@
 
-\begin{itemize}
-  \item How is Equation~\ref{eq:}
-  \item Isn't what we are doing just conditional density estimation? It seems
-  like there is a lot of prior work here. I.e see this paper here for a good overview ~\cite{greenberg2019automatic}
-  \item It does seem like this field has a large extensive past, but we still have a small niche.
-\end{itemize}
-
 \textbf{Old questions for Tom}
 \begin{itemize}
+  \item I don't understand how the objective function is comparing the generated samples from the density 
+  estimator to the ground truth, at least in the way it is currently defined? Answered - we updated the objective defined and added more comments
+  \item All RS references I look at state that we must sample from a uniform distribution and
+  then check if our proposed value is less than that uniformly sampled point, yet our RS do not have this.
+Answered - Rejection samplers we have are valid rejection samplers, this just might not define the density we expect them to.
+..?
+  nor what is the acceptance probability within our rejection sampling statements? The proof for rejection sampling seems to rely on the fact that you are drawing from a uniform distribution.
+
   \item Whilst the variational posterior we choose as our proposal has a functional form  
@@
   \item Stein Variational inference may be quite nice to use here~\cite{feng2017learning}, but it does make explicit assumptions in requiring the densities to be differentiable.
-- Yes this metod is good, but only works in the continuous domain.
-  \item  For an overview of all variational methods, including the use of different divergence metrics see here ~\cite{8588399} - Answered - Quite finickity to set up. 
+- Yes this method is good, but only works in the continuous domain.
+  \item  For an overview of all variational methods, including the use of different divergence metrics see here ~\cite{8588399} - Answered - Quite finicky to set up. 
 \end{itemize}

### 2019-08-16 paper/learning/notes.tex
add tom notes

--- 2019-08-14 paper/learning/notes.tex
@@
 
+\section{Friday morning}
+The subprograms within the simulator should be treated as nested probabilistic programs.
+
+There are three ways we can do these things, rather than just an amortization
+
+\subsection{Method 1: Learn the $\gamma_{a_{j}}(\cdot ; \eta_{a_{j}}, \cdot)$}
+
+As per we are currently planning. Relies on generating exact samples, which 
+we have access due to the rejection sampling loops.
+
+In this setting we can approximate the output
+\subsection{Method 2: Proper amortized inference of sub-programs}
+
+Learn a $q_{a_{j}}(z_{1:j}; \eta_{a_j}, \phi_j)$ and uses this as a proposal for the 
+inner inference. 
+I.e. if we learn that $q$ we could develop a more efficient rejection sampler by replacing the proposal.
+
+I.e we are calibrating the proposal to make it more amenable to accepting samples. 
+We can do this by using self normalized importance sampling on the inner program.
+
+The likelihood is just a $\{0,1\}$ likelihood, ie Bernoulli.
+
+\subsection{Method 3: Learn marginal likelihood $p(\cdot)$ approximation}
+
+\begin{equation}
+  p(x) = 
+  \begin{cases}
+  \prod^{n_x}_{j=1,{a_j}\in S_{R}} f_{a_j}(x_j ; \psi_j) \prod^{n_x}_{j=1,{a_j}\in S_{R}} \frac{P^{in}_{a_j}(x_j,\psi_j)}{P^{in}_{a_j}(\psi_j)} \\
+  0
+  \end{cases}
+\end{equation}
+The normalising constant inside (in) the rejection sampler is intractable.
+
+The things inside of the rejection sampler are proportional to the following:
+% \begin{equation}
+%   \begin{cases}
+%   \end{cases}
+% \end{equation}
+
+It is really important to ensure that these pairs of samples are unbiased.
+
+If you minimise an $L_2$ norm the optimal 
 \section{ Questions for Tom}
@@
 \begin{itemize}
-  \item How is Equation~\ref{eq:}
+  \item How is Equation 
   \item Isn't what we are doing just conditional density estimation? It seems

### 2019-08-16 paper/learning/notes.tex

--- 2019-08-15 paper/learning/notes.tex
@@
 
+\section{Friday morning}
+The subprograms within the simulator should be treated as nested probabilistic programs.
+
+There are three ways we can do these things, rather than just an amortization
+
+\subsection{Method 1: Learn the}
+$\gamma_{a_{j}}(\cdot ; \eta_{a_{j}}, \cdot)$
+
+As per we are currently planning. Relies on generating exact samples, which 
+we have access due to the rejection sampling loops.
+
+In this setting we can approximate the output
+\subsection{Method 2: Proper amortized inference of sub-programs}
+
+Learn a $q_{a_{j}}(z_{1:j}; \eta_{a_j}, \phi_j)$ and uses this as a proposal for the 
+inner inference. 
+I.e. if we learn that $q$ we could develop a more efficient rejection sampler by replacing the proposal.
+
+I.e we are calibrating the proposal to make it more amenable to accepting samples. 
+We can do this by using self normalized importance sampling on the inner program.
+
+The likelihood is just a $\{0,1\}$ likelihood, ie Bernoulli.
+
+\subsection{Method 3: Learn marginal likelihood $p(\cdot)$ approximation}
+
+\begin{equation}
+  p(x) = 
+  \begin{cases}
+  \prod^{n_x}_{j=1,{a_j}\in S_{R}} f_{a_j}(x_j ; \psi_j) \prod^{n_x}_{j=1,{a_j}\in S_{R}} \frac{P^{in}_{a_j}(x_j,\psi_j)}{P^{in}_{a_j}(\psi_j)} \\
+  0
+  \end{cases}
+\end{equation}
+The normalising constant inside (in) the rejection sampler is intractable.
+
+The things inside of the rejection sampler are proportional to the following:
+% \begin{equation}
+%   \begin{cases}
+%   \end{cases}
+% \end{equation}
+
+It is really important to ensure that these pairs of samples are unbiased.
+
+If you minimise an $L_2$ norm the optimal 
 \section{ Questions for Tom}

### 2019-08-17 paper/learning/notes.tex
updated

@@
 \usepackage{cancel}
+\setlength{\abovedisplayskip}{3pt}
+\setlength{\belowdisplayskip}{3pt}
 \tikzset{>=latex}
@@
 % \usepackage[binary-units=true]{siunitx}
-\title{Hijacking Simulators with Universal Probabilistic Programming}
+\title{AIRS: Amortized Inference for Rejection Samplers for Fast Simulator Inference}
 
@@
   \small
-   \caption{Rejection Sampling}
+   \caption{ABC: Rejection Sampling}
     \begin{algorithmic}\small
@@
 \end{algorithm}
+
+but more generally speaking we can write a rejection sampler as follows:
+\begin{algorithm}
+  \label{alg:rsstats}
+  \small
+   \caption{Simulator: Rejection Sampling}
+    \begin{algorithmic}\small
+       \State {At iteration $i$ s.t $i \geq 1$}
+       \State $X_{i} \sim p(x)$ 
+       \State $\mathbb{I}(g(x) > 0)$
+       \State $ Y \sim p(X=x | \mathbb{I}(g(x) > 0))$
+       \State accept $X_i$ with $p(\mathbb{I}(g(x) > 0))$
+\end{algorithmic}
+\end{algorithm}
+Thus, the probability of generating a sample $X = x$ from the true distribution can be defined, using Bayes rule, 
+as: 
+\begin{equation}
+  p(X=x| \mathbb{I}) = \frac{p(\mathbb{I} | X=x)p(X=x)}{p(\mathbb{I})} 
+\end{equation}
+where $p(\mathbb{I} | X=x) = \{0,1\}$.
 
@@
 \end{lstlisting}\\
+% \begin{lstlisting}[mathescape=true]
+%   def $R_3(z_4)$:
+%     $y \sim \mathcal{U}(0,1)$
+%     $x \sim \mathcal{U}(0,1.5)$
+%     while $y < z_2$:
+%       $y \sim \mathcal{N}(z_3,1)$
+%     end
+%     return $y$
+%   \end{lstlisting}\\
 \hline
@@
 
-The framework we present in this section is generally applicable to stochastic simulator 
-environment, where there exists rejection sampling loops.
-
-We start by noting that the prior, defined in the context of the outer samples is 
-defined as: 
+The framework we present in this section is generally applicable to any stochastic simulator 
+environment, where there exists any form of rejection sampling blocks.
+To perform the inferences that we require, we construct a variational approximation of the rejection sampling 
+blocks via an amortization procedure
+and note that the variational approximation in the context of our setting is specified by three things:
+\begin{enumerate}
+  \itemsep0em 
+  \item Makes use of existing parts of the outer program including all deterministic operations.
+  \item Form of the $\gamma_{a_j}(x_j ; \eta_{a_j}, \psi_j) \forall a_j \in S_r$. 
+  \item There exists variational parameters $\eta_{a_j} \forall a_j \in S_r$.
+\end{enumerate}
+where $a_j \in \{1, \ldots, L\}$ corresponds to the set of all addresses, $S_r$ denotes the set of addresses $a_j$ that correspond directly to the locations 
+of the rejection sampling blocks, that is $a_j \cup S_r = \{1, \ldots, L\}$, $\gamma_{a_j}$ denotes the variational approximation at address $a_j$, 
+for example, $\gamma_{a_j} = \mathcal{N}(x_j ; \mu_{a_j}(\psi_j; \eta_{a_j}), \Sigma_{a_j}(\psi_j ; \eta_{a_j}))$, with $\eta_{a_j}$ representing the variational
+parameters and $\psi_j$ representing all variables that input into the given program.
+
+We start by noting that the prior, defined in the context of the outer samples is defined as: 
 \begin{equation}
@@
 \end{equation}
-We start with the information projection we construct the variational objective as follows:
+next, we state for a given program $\mathcal{P}$ we denote the proposal for the program as:
+\begin{equation}
+  q(\mathbf{x}; \kappa) = 
+  \begin{cases}
+  \prod^{n_x}_{j=1, j\notin S_{r}} f_{a_j}(x_j ; \psi_j)(\prod^{n_x}_{j=1, j \in S_{r}} \gamma_{a_j}(x_j; \eta_{a_j}, \psi_{j})) \texttt{ if } \mathcal{B}(x_{1:n_{x}} = 1)\\
+  0
+  \end{cases}
+\end{equation}
+where $\kappa = \{\eta_{a_j} ; a_j \in S_{r}\}$ are the variational parameters.
+As we now have the density of the proposal and the density of the program we can construct a variational objective as follows:
+
+using the information projection we construct the variational objective as follows:
 \begin{align*}
-  J(\eta) = KL(p(x) || q(x; \eta)) \\
+  J(\eta) = KL(p(x) || q(x; \kappa)) \\
    = \int p(x)\ln\left(\frac{p(x)}{q(x;\eta)}\right) dx \\
@@
    \text{Let } a_{j} \in \{1, \ldots, L\} , S_{R} \subset \{1, \ldots, L\} \\
-   = \cancelto{\text{no dependence on } \eta}{\mathbb{E}_{p(x)}[\sum^{n_x}_{j=1,a_j \notin S_{R}}\ln(p(x))]}-\mathbb{E}_{p(x)}[\sum^{n_x}_{j=1,a_j \in S_{R}}\ln(q_{a_j}(x_j;\eta))]\\
+   \kappa^{*} = \operatorname*{argmin}_{\kappa} \mathbb{E}_{X \sim p(x)}[-\log(q(X=x; \kappa))] \\
+   = \operatorname*{argmax}_{\kappa} \mathbb{E}_{X \sim p(x)}[\sum^{n_x}_{j=1, j \in S_r} \log( \gamma_{a_j}(X_j = x_j ; \eta_{a_j}, \psi_j))] \\
+   \forall r \in S_r, \eta^{*}_r = \operatorname*{argmax}_{r} \mathbb{E}_{x \sim p(x)}[\sum^{n_x}_{j=1}\mathbb{I}(a_j =r) \log(\gamma_r (X_j = x_j ; \eta_r, \psi_j))] \\
+   \nabla_{\kappa} J(\kappa) = \nabla_{\kappa} \mathbb{E}_{X \sim p(x)}[\sum^{n_x}_{j=1}\mathbb{I}(a_j =r) \log(\gamma_r (X_j = x_j ; \eta_r, \psi_j))] \\
+   = \mathbb{E}_{X \sim p(x)}[ \sum^{n_x}_{j=1, j \in S_r} \nabla_\kappa \log(\gamma_r (X_j = x_j ; \eta_r, \psi_j)]\\
+   \approx \frac{1}{N} \sum^{N}_{n=1} \sum^{n_x}_{j=1, j\in S_r} \nabla_{\kappa}\log(\gamma_r (x^{n}_j ; \eta_r, \psi_j))
+  %  = \cancelto{\text{no dependence on } \eta}{\mathbb{E}_{p(x)}[\sum^{n_x}_{j=1,a_j \notin S_{R}}\ln(p(x))]}-\mathbb{E}_{p(x)}[\sum^{n_x}_{j=1,a_j \in S_{R}}\ln(q_{a_j}(x_j;\eta))]\\
 \end{align*}
 
-Thus, to define the objective for each subproblem $r  \in S_{R}$ as:
+Thus, to define the objective for each subproblem $r  \in S_{r}$ as:
 \begin{equation}
   \label{eq:objective}
-  J_{r}(\eta_r) = -\mathbb{E}_{p(x)}[\sum^{n_x}_{j=1}\mathbb{I}(a_j = r)\ln(q_r(x_j;\eta_r, \phi_j))]
-\end{equation}
+  \nabla_{\eta_r} J(\kappa) \approx \frac{1}{N} \sum^{N}_{n=1} \sum^{n_x}_{j=1} \mathbb{I}(r=a_j) \nabla_{\eta_r}\log(\gamma_r (x^{n}_j ; \eta_r, \psi_j))
+\end{equation}
+
+where $x^{n}_{1:n_x} \stackrel{iid}\sim p(x)$.
 
@@
 
-\subsection{Method 3: Learn marginal likelihood $p(\cdot)$ approximation}
-
+\subsection{Method 3: Learn marginal likelihood approximation}
+With $ p(\cdot)$ representing the marginal likelihood. 
 \begin{equation}

### 2019-08-18 code/amortized_rs/memory_src/amortized_rs.py
added R3

@@ -28,8 +28,9 @@ import math
 import numpy as np
 import torch
 from torch.distributions import Normal, Uniform
+import scipy.stats as ss
 
-def f():
+def s():
     z1 = Normal(0,1).sample()
     ctr = 1
     z2 = R1(z1)
@@ -64,9 +65,43 @@ def R1(z1, i=0):
             return math.inf
         i += bs
 
+
+def f(x):
+    return ss.norm(30, 10).pdf(x) + ss.norm(80,20).pdf(x)
+
+
+def g(x):
+    return ss.norm(50, 30).pdf(x)
+
+
+x = np.arange(-50, 151)
+k = max(f(x) / g(x))
+
+
+def R3(iter=1000):
+    samples = []
+
+    for i in range(iter):
+        y = np.random.normal(50, 30)
+        u = np.random.uniform(0, k*g(y))
+
+        if u <= f(y):
+            samples.append(y)
+
+    return np.array(samples)
+
+
+# def plotR3(x)
+#     plt.plot(x, f(x))
+#     plt.plot(x, k*g(x))
+#     plt.show()
+#
+#     s = R3(iter=100000)
+#     sns.distplot(s)
+
 import timeit
 torch.manual_seed(7)
-print(f())
+print(s())
 print(timeit.timeit("f()", setup="from __main__ import f", number=2000))
 
 amortized_rs_std = load(name="amortized_rs_std",

### 2019-08-18 paper/learning/notes.tex
updated

@@
  need not take this form and arise in many nested structures within probabilistic programs~\cite{rainforth2018nesting} and simulators~\cite{smith2008towards,bershteyn2018implementation,gleisberg2009event}.
-
-The standard algorithm is defined as follows:
+ More generally speaking we can write a rejection sampler in the form of Algorithm~\ref{alg:rssim}
 \begin{algorithm}
@@
 
-but more generally speaking we can write a rejection sampler as follows:
 \begin{algorithm}
-  \label{alg:rsstats}
+  \label{alg:rssim}
   \small
@@
 \noindent
-\begin{tabular}{|p{3.6cm}|p{3.6cm}|p{3.6cm}|}
+\begin{tabular}{|p{3.6cm}|p{3.6cm}|p{3.6cm}|p{3.6cm}|}
 \hline
-Stochastic Simulator  & Rejection Sampler 1 & Rejection Sampler 2 \\
+Stochastic Simulator  &  R1 & R2 & R3 \\
 \hline
@@
       $z_{4} \gets R_{2}(z_2,z_3, 1)$
+      $z_{5} \sim R_{3}(z_4, 1)$
     return $g(z_{1:4})$
@@
   return $y$
+\end{lstlisting}&
+\begin{lstlisting}[mathescape=true]
+def $R_3(z_4)$:
+  $f(x) = \mathcal{N}(30,10^2) +$
+  $\mathcal{N}(80,20^2)$
+  $g(x) = \mathcal{N}(50,30^2)$
+  $y \sim \mathcal{N}(50,30^2)$
+  $u \sim \mathcal{U}(0,1.5)$
+  if $u \leq f(y)$:
+    return $y$
+  else:
+    return $R_3(z_4)$
 \end{lstlisting}\\
-% \begin{lstlisting}[mathescape=true]
-%   def $R_3(z_4)$:
-%     $y \sim \mathcal{U}(0,1)$
-%     $x \sim \mathcal{U}(0,1.5)$
-%     while $y < z_2$:
-%       $y \sim \mathcal{N}(z_3,1)$
-%     end
-%     return $y$
-%   \end{lstlisting}\\
 \hline
-$P(\mathcal{T}= x_{1:n_x} | \lambda)$ & $q_{1}(z_{2} | z_{1})$ & $q_{2}(z_{4} |z_{2},z_{3})$ \\
+$P(\mathcal{T}= x_{1:n_x} | \lambda)$ & $q_{1}(z_{2} | z_{1})$ & $q_{2}(z_{4} |z_{2},z_{3})$ & $q(z_5 |z_4 )$ \\
 \hline
@@
 However, in order to do construct this we need the input $\theta$, $y$, $\alpha$ and the distributions utilised for the rejection sampling loop. 
-For any arbitrary simulator, written in a generic code base, this is an incredibly hard 
+For any rejection sampling block, written in a generic code base, not in the form of R3, this is an incredibly hard 
 task to do, because rejection sampling blocks within simulators are not necessarily written in a standard template
@@
 \begin{equation}
-  p_{\theta_{\text{max$\backslash$min}}}(z|x) \frac{p_{\theta}(x,z)}{\int dz p_{\theta_{\text{max$\backslash$min}}}(x,z)} 
+  p_{\theta_{\text{max$\backslash$min}}}(z|x) =  \frac{p_{\theta}(x,z)}{\int dz p_{\theta_{\text{max$\backslash$min}}}(x,z)} 
 \end{equation}

### 2019-08-19 code/amortized_rs/memory_src/amortized_rs.py
add ABC style rs + test

@@ -28,20 +28,64 @@ import math
 import numpy as np
 import torch
 from torch.distributions import Normal, Uniform
-import scipy.stats as ss
 
 def s():
+    ''' In this simple simulator we care about the input output sets:
+    {z1, z2}
+    {(z2,z3), z4}
+    {z5
+    '''
     z1 = Normal(0,1).sample()
-    ctr = 1
     z2 = R1(z1)
     # the following is used to avoid tail recursion  - simple solution and stops stack overflow
     # You can not use this trick straight forwardly if you wish to access multiple core / threads.
     while z2 == math.inf:
+        # we resample again to increase chance of acceptance.
+        # This does not change acceptance probabilities. This is just for
+        # computational efficiency we generating samples.
+        z1 = Normal(0, 1).sample()
         z2 = R1(z1)
-        ctr += 1
     z3 = Uniform(0,2).sample()
     z4 = R2(z2,z3)
-    return torch.tensor([z1,z2,z3,z4])
+    z5 = Normal(50,30).sample()
+    M = torch.tensor(3.0337) # determined via x = torch.linsapce(-50.0,151.0, nsamples); torch.max(f(x)/g(x))
+    z6, gz, fz = R3(z5,M)
+    while z6 == math.inf:
+        # we resample again to increase chance of acceptance.
+        # This does not change acceptance probabilities. This is just for
+        # computational efficiency we generating samples.
+        z5 = Normal(50,30).sample()
+        z6 = R3(z5, M)
+    return torch.tensor([z1,z2,z3,z4,z5,z6])
+
+def f(x, mu1, sigma1, mu2, sigma2):
+    ' PDF of Mixture of gaussian that we want to sample from: N(mu1,sigma1) + N(mu2,sigma2)'
+    const1 =  1 / (2 * np.pi * sigma1 ** 2 *torch.ones(x.shape)) ** 0.5
+    const2 =  1 / (2 * np.pi * sigma2 ** 2 *torch.ones(x.shape)) ** 0.5
+    body1 =  torch.exp(-(x - mu1*torch.ones(x.shape)) ** 2 / (2 * sigma1 ** 2 *torch.ones(x.shape)))
[…]

### 2019-08-19 paper/learning/notes.tex
updated

@@
       $z_{4} \gets R_{2}(z_2,z_3, 1)$
-      $z_{5} \sim R_{3}(z_4, 1)$
-    return $g(z_{1:4})$
+      $z_{5} \sim \mathcal{U}(50,30^2)$
+      $z_{6} \sim R_{3}(z_5, 1)$
+    return $g(z_{1:6})$
 \end{lstlisting}&
@@
 \begin{lstlisting}[mathescape=true]
-def $R_3(z_4)$:
+def $R_3(z_5)$:
   $f(x) = \mathcal{N}(30,10^2) +$
@@
   $g(x) = \mathcal{N}(50,30^2)$
-  $y \sim \mathcal{N}(50,30^2)$
-  $u \sim \mathcal{U}(0,1.5)$
-  if $u \leq f(y)$:
-    return $y$
+  $u \sim \mathcal{U}(0,M*g(z_5))$
+  if $u \leq f(z_5)$:
+    return $z_5$
   else:
-    return $R_3(z_4)$
+    return $R_3(z_5)$
 \end{lstlisting}\\
 \hline
-$P(\mathcal{T}= x_{1:n_x} | \lambda)$ & $q_{1}(z_{2} | z_{1})$ & $q_{2}(z_{4} |z_{2},z_{3})$ & $q(z_5 |z_4 )$ \\
+$P(\mathcal{T}= x_{1:n_x} | \lambda)$ & $q_{1}(z_{2} | z_{1})$ & $q_{2}(z_{4} |z_{2},z_{3})$ & $q(z_5)$ \\
 \hline
@@
 the unknown densities that we wish to learn through an amortized approach.
-It is important to note that the expected running time of the rejection sampler is expected to geometric mean, we can show this as follows:
-\textbf{Bradley:  You only arrive at this, if the density is uniform}
-\begin{align*}
-P( X= x) = \sum^{\infty}_{n=1} P(\text{reject $n-1$ times}, \text{draw} Y = x) \\
-\text{todo} 
-\end{align*}
-
-In addition to this, if we have direct access to a rejection sampling loop then it is indeed 
+% It is important to note that the expected running time of the rejection sampler is expected to geometric mean, we can show this as follows:
+% \textbf{Bradley:  You only arrive at this, if the density is uniform} i.e R3
+% \begin{align*}
+% P( X= x) = \sum^{\infty}_{n=1} P(\text{reject $n-1$ times}, \text{draw} Y = x) \\
+% \text{todo} 
+% \end{align*}
+
+If we have direct access to a rejection sampling loop then it is indeed 
 possible to construct the density, up to a constant of proportionality. 
@@
 \begin{align*}
-y \sim q(\theta) \\
+z \sim g(\theta) \\
 do something \\
-\alpha = p_{accept}(y,\theta) \\
+\alpha = p_{accept}(z,\theta) \\
 \mathtt{if} Bern(\alpha) = 1: \\
-return y
+return z
 \end{align*}
@@
 For any rejection sampling block, written in a generic code base, not in the form of R3, this is an incredibly hard 
-task to do, because rejection sampling blocks within simulators are not necessarily written in a standard template
+task to do, because rejection sampling blocks within simulators are not necessarily written in a standard template such as R1 and R2. 
 \textbf{Add examples in a supplement.  Need to demonstrate to YW that this is challenging to do, even if you know where 
-the rejection sampling loops are.
-Hypothesis, RS loops are not written in a templated form}.
-With the PPX protocols previously developed in \cite{baydin2018efficient,gram2019hijacking} we can however,
+the rejection sampling loops are. Hypothesis, RS loops are not written in a templated form}.
+y leveraging the PPX protocols previously developed in \cite{baydin2018efficient,gram2019hijacking} we can however, 
 extract all variables, other than $\alpha$ in an automated way.
 However, without being able to easily extract $\alpha$ this means that we are unable to 
-correctly build the form of the density and thus we limited in our ability to leverage 
+correctly build the form of the density and thus are limited in our ability to leverage 
 powerful inference methods such as MCMC and Variational inference.
-Thus, we propose another approach, automating the amortization of rejection sampling loops. 
-To learn functions that we can easily sample from and use to construct the density of the entire
+Thus, we propose another approach, automating the amortization of rejection sampling loops of type R1, R2 and R3. 
+This enables us to learn functions that we can easily sample from and via the amortization process construct the density of the entire
 simulator.
 
-The simulated densities for both each rejection sampling block can be seen in Figure~\ref{fig:rs_simulated}.
+The simulated densities for the rejection sampling block can be seen in Figure~\ref{fig:rs_simulated}.
 \begin{figure}

### 2019-08-20 code/amortized_rs/memory_src/amortized_rs.cpp
remove redundent

@@ -30,6 +30,7 @@ at::Tensor f(void) {
   //std::cout << ctr;
   Scalar z_3 = at::empty({1}, kFloat).uniform_(0,2).item();
   Scalar z_4 = R2(z_2, z_3);
+  Scalar z_5 = R3(z_4);
   //std::cout << z_1 << "|" << z_2 << "|" << z_3 << "|" << z_4;
   //float data[] = {z_1.to<float>(), z_2.to<float>(), z_3.to<float>(), z_4.to<float>()};
   auto ret = torch::empty({4}, kFloat);
@@ -72,7 +73,6 @@ Scalar R2(Scalar z_2, Scalar z_3) {
     }
     return temp;
 }
-
 
 PYBIND11_MODULE(TORCH_EXTENSION_NAME, m) {
   m.def("f", &f, "f");

### 2019-08-20 code/amortized_rs/memory_src/amortized_rs.py

@@ -49,7 +49,7 @@ def s():
     z4 = R2(z2,z3)
     z5 = Normal(50,30).sample()
     M = torch.tensor(3.0337) # determined via x = torch.linsapce(-50.0,151.0, nsamples); torch.max(f(x)/g(x))
-    z6, gz, fz = R3(z5,M)
+    z6 = R3(z5,M)
     while z6 == math.inf:
         # we resample again to increase chance of acceptance.
         # This does not change acceptance probabilities. This is just for
@@ -109,23 +109,6 @@ def R1(z1, i=0):
             return math.inf
         i += bs
 
-
-def f(x, mu1, sigma1, mu2, sigma2):
-    return 1 / (2*np.pi*sigma1**2)**0.5 * torch.exp(-(x-mu1)**2 / 2*sigma1**2)  + 1 / (2*np.pi*sigma2**2)**0.5 * torch.exp(-(x-mu2)**2 / 2*sigma2**2)
-
-def g(x,mu1, sigma1):
-    return 1 / (2 * np.pi * sigma1 ** 2) ** 0.5 * torch.exp(-(x - mu1) ** 2 / 2 * sigma1 ** 2)
-
-def R3(z5):
-    k = max(f(z5,30,10,80,20)/g(z5,50,30))
-    u = Uniform(0, k*g(z5, 50,30)).item()
-    x = True
-    while x:
-        if u <= f(z5, 30, 10, 80, 20).item():
-            x = False
-        else:
-            continue
-    return z5
 
 
 # def plotR3(x)

### 2019-08-20 code/amortized_rs/memory_src/hogwild.py

@@ -60,7 +60,7 @@ def train(model, optimizer, loss_fn,  N, data_flag, batch_size, rank, load_data=
     for i in range(N):
             inData, outData, count =get_batch(data_flag, batch_size, count)
             proposal = dist.Normal(*model(inData))
-            pred = proposal.rsample(sample_shape=[batch_size]).view(1,batch_size)
+            pred = proposal.rsample(sample_shape=[batch_size]).view(1,batch_size))
             optimizer.zero_grad()
             plt.show()
             _loss = loss_fn(outData, pred)
@@ -127,7 +127,7 @@ if __name__ == '__main__':
     lr = 0.0001
     momentum= 0.9
     load_data=False
-    batch_size = 2**13
+    batch_size = 2**10
     data_flag= 'R1'
     outputSize = 1
     # outputSize = 128 # for R1 and R2
@@ -143,18 +143,18 @@ if __name__ == '__main__':
     loss_fn = th.nn.MSELoss()
     # optimizer = optim.Adam(filter(lambda p: p.requires_grad, model.parameters()), lr=lr, amsgrad=True)
     optimizer = optim.SGD(model.parameters(), lr=lr, momentum=momentum, weight_decay=0.01)
-    # train(model, optimizer, loss_fn, N, data_flag, batch_size, rank=0, load_data=False)
+    train(model, optimizer, loss_fn, N, data_flag, batch_size, rank=0, load_data=False)
     # NOTE: this is required for the ``fork`` method to work
-    if trainOn:
-        model.share_memory()
-        processes = []
-        for rank in range(num_processes):
-            p = mp.Process(target=train, args=(model,optimizer,loss_fn, N,data_flag, batch_size, rank))
-            p.start()
-            processes.append(p)
-        for p in processes:
-            p.join()
-    testOn = False
+    # if trainOn:
+    #     model.share_memory()
+    #     processes = []
+    #     for rank in range(num_processes):
+    #         p = mp.Process(target=train, args=(model,optimizer,loss_fn, N,data_flag, batch_size, rank))
+    #         p.start()
+    #         processes.append(p)
+    #     for p in processes:
+    #         p.join()
+    # testOn = False
     if testOn:
         model_name = 'model_2019-08-12_09-35_rejectionBlock_R2_process_8'
         n_test = 1000

### 2019-08-20 paper/learning/notes.tex
added plots - still to update text

@@
 Where the two methods diverge is in how the acceptance probability is defined and how the samples are generated.
-In the standard statistical setting, see algorithm~\ref{alg:rsstats}, we generate a sample $X$ by sampling $Z$ and accept $Z$ with probability
+In the standard statistical setting, see Algorithm~\ref{alg:rsstats}, we generate a sample $X$ by sampling $Z$ and accept $Z$ with probability
 $p_{accept} = \frac{f(x)}{Mg(x)}$, which we can do until acceptance. 
@@
 \begin{algorithm}
-  \label{alg:rsstats}
   \small
    \caption{ABC: Rejection Sampling}
+   \label{alg:rsstats}
     \begin{algorithmic}\small
@@
 \begin{algorithm}
-  \label{alg:rssim}
   \small
    \caption{Simulator: Rejection Sampling}
+   \label{alg:rssim}
     \begin{algorithmic}\small
@@
 For any rejection sampling block, written in a generic code base, not in the form of R3, this is an incredibly hard 
-task to do, because rejection sampling blocks within simulators are not necessarily written in a standard template such as R1 and R2. 
+task to do, because rejection sampling blocks within simulators are not necessarily written in a standard template. 
 \textbf{Add examples in a supplement.  Need to demonstrate to YW that this is challenging to do, even if you know where 
@@
 \begin{figure}
+  \label{fig:rs_simulated}
   \centering
-  \includegraphics[width=0.2\textwidth, height=0.2\textheight, keepaspectratio]{rs_plots.png}
+  \begin{minipage}{.5\textwidth}
+    \centering
+    \includegraphics[width=0.6\textwidth, height=0.6\textheight, keepaspectratio]{rs_plots.png}
+  \end{minipage}%
+  \begin{minipage}{.5\textwidth}
+    \centering
+    \includegraphics[width=0.6\textwidth, height=0.6\textheight, keepaspectratio]{rs3_plot.png}
+  \end{minipage}
   \caption{ Simulated densities for rejection sampling block 1, \emph{blue} and rejection
-  sampling block 2, \emph{orange}.}
-  \label{fig:rs_simulated}
+  sampling block 2, \emph{orange} and rejection sampling block 3, \emph{red}.}
  \end{figure}
@@
 
-\paragraph{observations} As the functions are relatively simple the loss functions it gets t
+\paragraph{observations} 
+Using a standard $L2$-loss function $L_{2} = (y_{pred} - y_{true})^{2}$
+As the functions are relatively simple the loss functions it gets t
 to a loss of around 0.6 and stops decreasing. Although, this seems low, I don't think that it is

### 2019-08-20 paper/learning/notes.tex
updated - to add nested stuff tomorrow morning

@@
 
-  As useful rejection sampling is, it has many short comings. 
-In particular, to simulate proposals for certain distributions,
-such as truncated normal distributions, regardless of the easy to sample from proposal
-chosen the computational running time of such rejection statements can be shown to have an
-expectation equal to infinity. \textbf{Bradley: Is this true?}
+However, in many of the non-standard ABC forms of the rejection samplers, the acceptance 
+probability is completely intractable, or is very difficult to calculate. 
+Thus, as useful rejection sampling is for generating samples from densities that we 
+would not be able to sample from, it has many short comings. 
+These are of course well known and it is the reason for developing more sophisticated 
+sampling schemes, such as importance sampling based methods.
+However, in the setting of such complex simulators, written over many years and
+in convoluted code bases, there is no easy way to re-implement such sampling 
+schemes.
+Hence, to leverage existing frameworks we can make use of the probabilistic 
+programming executation (PPX) protocols~\cite{baydin2018efficient}. 
+These protocols allow one to exact all raw sample calls $f_{a_j}$ from the
+simulator. This then enables one to easily construct a prior for the full program
+$P(\mathcal{T} = x_{1:n_{x}}| \lambda)$. 
+In addition to this, once such rejection schemes, \emph{nested programs}, are located, we have a number of 
+primatives that we can leverage to isolate calls to those addresses during training, to collect the samples generated during 
+\texttt{.forward()} runs of the simulator, 
+and during the inference procedure replace calls to those addresses $a_j$ with a
+learnt proposal $\gamma_{a_j}(x_j ; \eta_{a_j}, \phi_j ) $.
+
+% In particular, to simulate proposals for certain distributions,
+% such as truncated normal distributions, it can be very difficult to
+% hand-craft a proposal such that the acceptance rate.
 
@@
 sampling blocks with learnt surrogate functions, that are expressive enough to 
-reciprocate the behavior of the rejection sampling block buys the user two things. 
-A) It reduces the computational complexity of the simulator. 
-B) As we now have a direct functional form of the density, we can correctly leverage
-more computationally and sample efficient inference schemes.
+reciprocate the behavior of the rejection sampling block provides the user with three things. 
+\begin{itemize}
+  \item It reduces the computational complexity of the simulator.  
+  \item As we now have a direct functional form of the density, we can correctly leverage more computationally and sample efficient inference schemes. 
+  \item By only amortizing sub-programs and not amortizing the whole simulator, the larger structure of the simulator remains interpretable, in the sense that we can 
+  still decode the decisions that it is making.  
+\end{itemize}
 
@@
 
-We do this by replacing the rejection sampling blocks with 
-learnt functions from an amortized inference procedure Figure~\ref{fig:amortized_sampling}.
-This then enables us to replace rejection sampling loops with amortized functions that 
-are quick to evaluate, provide a functional form to the density and significantly reduce 
-memory over-heads, as compared to~\cite{baydin2018efficient}, due to not having to track the 
-thousands of raw sample calls $f_{a_i}$ generated in standard simulators.
-
+% We do this by replacing the rejection sampling blocks with 
+% learnt functions from an amortized inference procedure Figure~\ref{fig:amortized_sampling}.
+% This then enables us to replace rejection sampling loops with amortized functions that 
+% are quick to evaluate, provide a functional form to the density and significantly reduce 
+% memory over-heads, as compared to~\cite{baydin2018efficient}, due to not having to track the 
+% thousands of raw sample calls $f_{a_i}$ generated in standard simulators.
+
+\section{A Stochastic Simulator}
+
+In this section, we present a simple stochastic simulator and povide 
+examples of the rejection blocks found within standard stochastic
+simulators~\cite{smith2008towards,bershteyn2018implementation, gleisberg2009event}. 
+This simulator will form the basis for our experiemnts in Section~\ref{sec:experiments}.
 Consider the following program, that is representative of a simple stochastic simulator
@@
 \begin{tabular}{|p{3.6cm}|p{3.6cm}|p{3.6cm}|p{3.6cm}|}
+
 \hline
@@
 \begin{lstlisting}[mathescape=true]
-    def f():
+    def s():
       $z_{1} \sim \mathcal{N}(0,1)$
@@
 possible to construct the density, up to a constant of proportionality. 
-That is:
-\begin{align*}
-z \sim g(\theta) \\
-do something \\
-\alpha = p_{accept}(z,\theta) \\
-\mathtt{if} Bern(\alpha) = 1: \\
-return z
-\end{align*}
-thus the density of $y,\theta \propto q(\theta) \alpha$.
-However, in order to do construct this we need the input $\theta$, $y$, $\alpha$ and the distributions utilised for the rejection sampling loop. 
+% That is:
+% \begin{align*}
+% z \sim g(\theta) \\
+% do something \\
+% \alpha = p_{accept}(z,\theta) \\
+% \mathtt{if} Bern(\alpha) = 1: \\
+% return z
+% \end{align*}
+ $x,z \propto f(\theta) p(accept)$.
+However, in order to do construct this we need the input $z$, the sample from the proposal $x$ and $p(accept)$ (should use the notation above).
 For any rejection sampling block, written in a generic code base, not in the form of R3, this is an incredibly hard 
@@
 y leveraging the PPX protocols previously developed in \cite{baydin2018efficient,gram2019hijacking} we can however, 
-extract all variables, other than $\alpha$ in an automated way.
-However, without being able to easily extract $\alpha$ this means that we are unable to 
+extract all variables, other than $P(accept)$ in an automated way.
+As without being able to easily extract $p(accept)$ this means that we are unable to 
 correctly build the form of the density and thus are limited in our ability to leverage 
-powerful inference methods such as MCMC and Variational inference.
+powerful inference methods such as MCMC and Variational methods.
 Thus, we propose another approach, automating the amortization of rejection sampling loops of type R1, R2 and R3. 
@@
   \begin{cases}
-    \prod^{n_{x}}_{j=1} f_{a_j}(x_j ; \psi_j) if \mathcal{B}(x) = 1 \\
+    \prod^{n_{x}}_{j=1} f_{a_j}(x_j ; \psi_j) \texttt{ if } \mathcal{B}(x) = 1 \\
     0
@@
 \section{Experiments: Amortizing Stochastic Processes}
-
+\label{sec:experiments}
 \paragraph{TODO list}
@@
 \subsection{Baseline}
+For gif plots see the plots folder in code$\backslash$amortized\_rs$\backslash$plots
+
 As a first approach, we train a regressor to learn a function to map
@@
 
+\section{The difference between this paper and Inference compilation paper}
+
+The main fundemental difference is that IC cares about the full global program, 
+we only care about the local (sub / nested programs).
+
+We are not interesting in calculating the weights $w^k$ since the method that we 
+develop here is applicable to any inference scheme, in theory.
+
+In the IC paper you are approximating $p(x|y)$ the \emph{the global} program, where 
+as you can view this work as learning an approximation for the \emph{local}, sub-programs
+$P^{in}(x|\nu)$ where the nuissince parameters in this paper are represented by $\psi$.
+
+In the notation of the inference compilation paper we essentailly learn $q_{a_{t},i_{t}}(x_t| x_{1:t-1}, y)$
+where the indicies are part of the set $S_r$.
+ The difference here though, is that we only care about the input at that address, your proposals 
+ are condiitoning on everything upto that point of the executation trace $\mathcal{T}$
+ in the simulator. 
+ The training data represented by (11) in the IC paper is not the training data we use in this paper,
+ we only care about the $x^{m}_t$ or $z$'s and $x$'s in this paper (notation is not entirely consitant throuhg out this paper yet)
+ But these are the parameters that are generated from raw sample calls within side the RS loop $f_{a_i \in S_r}(\cdot))$.
+
+Thus the density of the full joint of the program in this setting is given as:
+\begin{equation}
+  p(x,y) = \prod_{t=1, t \notin S_r} f_{a_t}(x_t | \psi_t) \prod_{n=1} g_{b_n}(y_n | \phi) \prod_{h=1}\gamma_{a_j \in S_r}(\cdot)
+\end{equation}
 \section{Friday morning}
@@
 
-\subsection{Method 1: Learn the}
-$\gamma_{a_{j}}(\cdot ; \eta_{a_{j}}, \cdot)$
+\subsection{Method 1: Learn the maoritzation of the sub-programs}
+That is learn $\gamma_{a_{j}}(\cdot ; \eta_{a_{j}}, \cdot)$ for each program in the set $S_r$.
 
@@
 The things inside of the rejection sampler are proportional to the following:
-% \begin{equation}
-%   \begin{cases}
-%   \end{cases}
-% \end{equation}
+\begin{equation}
+  P_{a_j}^{in}(x_j ; \phi_j) \propto
+  \begin{cases}
+  \prod^{n^{z}_{a_j}} f^{in}_{i}(z_{ij} | \theta_i) \prod_{k=1}^{n^{y}_{b_k}}g^{in}_{k}(y_i ; \phi_{i}) \texttt{ if } \mathcal{B}(z_{1:n^{z}_{a_j}}) = 1 \\
+  0
+  \end{cases}
+\end{equation}
+
+Thus, we need a regression from $\phi_j \rightarrow P^{in}){a_j}(\phi_j) $.
+We have access to pairs $\{\phi_j, \hat{P}^{in}_{a_j}(\phi_j)\}$ for which 
+the expectation $\mathbb{E}[\hat{P}^{in}_{a_j}( \phi_j)| \phi_j] = P^{in}_{a_j} (\phi_j)$ where
+$\hat{P}$ is \textbf{Play back convo to find out}
 
@@
 
-If you minimise an $L_2$ norm the optimal 
+If you minimise an $\mathcal{L}_2$ norm for this given problem then the optimal proposal natural arises, see~\ref{fig:tomfig} 
+The training regime would still need to be non-bias.
+
+\begin{figure}[h!]
+  \centering
+  \includegraphics[width=\textwidth]{tom_derivation.jpg}
+  \caption{Tom's derivation proving that the $\mathcal{L}_2$ norm in this problem set-up enables us to learn a proposal that exactly
+  matches the true distribution.  }
+  \label{fig:tomfig}
+\end{figure}
 \section{ Questions for Tom}

### 2019-08-21 paper/learning/notes.tex
updated

--- 2019-08-20 paper/learning/notes.tex
@@
 % \usepackage[binary-units=true]{siunitx}
-\title{AIRS: Amortized Inference for Rejection Samplers for Fast Simulator Inference}
+\title{Automating the Amortization of Sub-programs \\ for Fast Simulator Inference}
 
@@
 Simulators arise in a myriad of scientific and industrial settings; epidemiology, physics, engineering, climate modelling and so forth [add citations].
-One interesting class of simulators are population-based simulators, using in epidemiology, reinforcement learning
+One interesting class of simulators are population-based simulators, used in epidemiology, reinforcement learning
 and financial modelling [add citations].
@@
 since we do not have explicit access to the density of the \emph{program}.
-Where program density refers to the joint density constructed by the program, which in this case is represented by the simulator.
+Where density refers to the joint density constructed by the program, which in this case is represented by the simulator.
 This inability to evaluate the joint density has spawned research into how to perform 
-inference in this likelihood free setting~\cite{greenberg2019automatic} \textbf{Add normalising flow citations}
+inference in likelihood free settings~\cite{greenberg2019automatic, mohamed2016learning, papamakarios2018sequential,brehmer2018mining}
 for conditional density estimation. However, whilst many of these methods look to generate a surragote 
 for the entire simulator, we propose amortizing only computational expensive sub-components 
-of the simulator, which we can do in an automated way with probabilistic programming. 
+of the simulator, which we can automate with probabilistic programming. 
 This enables three things. 
 1) It reduces the loss of information, as when simulators are amortized with in current
-settings we require, potentially infinite samples, to provide a useful learnt representation of the simulator. 
+settings we require, a potentially infinite number of samples, to provide a useful learnt representation of the simulator. 
 2) It increases computational efficiency, whilst maintaining expressivity. 
@@
 easily sampled from and used to construct a functional form of the likelihood, that enables the 
-target density to be fully constructed, enabling one to leverage a large class of sample effient inference schemes, namely MCMC.
+target density to be fully constructed, enabling one to leverage a large class of sample efficient inference schemes, namely MCMC.
 
@@
 more expressive modelling capabilities. 
-We are able to demonstrate this by replacing rejection sampling blocks, which are utilised within 
-many prominent simulators to simulate densities that have no well defined cumulative density 
+We are able to demonstrate this by replacing sub-programs, in particular, rejection sampling blocks.
+Rejection sampling blocks arise in many non-standard forms within simulator to simulate densities that have no well defined cumulative density 
 function, nor no known process to efficiently sample from such a density exists. 
@@
 can replace such blocks with their leant surrogate, from which we can then construct the entire 
-density of the simulator, which can then be utilized, \emph{correctly}, contrary to ~\cite{baydin2018efficient},
+density of the simulator.
+This can then be utilized, \emph{correctly}, contrary to ~\cite{baydin2018efficient},
 by sample efficient Markov chain Monte Carlo inference schemes, such as Hamiltonian Monte Carlo~(HMC),
-random-walk Metropolis Hastings~(RMH), Light-weight Metropolis Hastings~(LMH) and many other schemes.
+random-walk Metropolis Hastings~(RMH), Light-weight Metropolis Hastings~(LMH) and other schemes.
 Whilst also significantly reducing memory consumption and computational running time of 
-both the simulator and of the probabilistic programming system, as evaluating amortized functions 
+both the \texttt{forward()} runs of the simulator and the inferences performed via a probabilistic programming system.
+This is because evaluating amortized functions 
 is cheap and the reduction in rejection sampling blocks significantly reduces the number of 
-unnecessary latent variables tracked~\cite{baydin2018efficient}.
-
-Without the amortization procedure, as updates are made sequentially MCMC methods are typically not-well suited this type of recursive
-estimation problem. 
-As each time we get a new data point $y_{t+1}$ we have to run new MCMC simulations 
-for $P(x_{0:t_1}| y_{0:t+1})$, we also cannot directly reuse the samples $\{x^{i}_{0:t}\}$. 
-However, as we can only run the simulator $\texttt{forward()}$ we cannot return back to a previous state and re-run. 
-So adapting existing inference techniques is the only viable option. 
-MCMC methods can be effective with a well designed amortized inference procedure (\textbf{Note: This is not guaranteed}).
-Non-MCMC based methods such as importance sampling (IS) and sequential monte carlo (SMC), may also be
-sensible inference engines, although we shall discuss the limitations within the simulator context in the proceeding sections.
+unnecessary latent variables tracked by the system developed in \cite{baydin2018efficient}.
+
+% Without the amortization procedure, as updates are made sequentially, MCMC methods are typically not-well suited this type of recursive
+% estimation problem. 
+% As each time we get a new data point $y_{t+1}$ we have to run new MCMC simulations 
+% for $P(x_{0:t_1}| y_{0:t+1})$, we also cannot directly reuse the samples $\{x^{i}_{0:t}\}$. 
+% However, as we can only run the simulator $\texttt{forward()}$ we cannot return back to a previous state and re-run. 
+% So adapting existing inference techniques is the only viable option. 
+% MCMC methods can be effective with a well designed amortized inference procedure (\textbf{Note: This is not guaranteed}).
+% Non-MCMC based methods such as importance sampling (IS) and sequential monte carlo (SMC), may also be
+% sensible inference engines, although we shall discuss the limitations within the simulator context in the proceeding sections.
 
@@
     \begin{enumerate} \item 1) Only the necessary raw samples from the desired rejection sampling addresses to be collected for training.
-                       \item 2) A function at inference time that replaces those given addresses with a learnt surrogate.
+                       \item 2) A process at inference time that replaces those given addresses with a learnt surrogate.
     \end{enumerate}
-    \item Finer grained control over the actual latent variables of interest, to avoid unnecessary memory consumption and ensure that inference problems remain tractable.
+    \item Finer grained control over the actual latent variables of interest, to avoid unnecessary memory consumption and ensure that inference problem encountered remain tractable.
   \end{itemize}
 
-\textbf{There is no doubt that implementing some fo these features will be more invasive to the underlying simulator code than previously
-just hijacking the raw-random number draws, but, it would enable us to attack difficulties in the inference procedures that are
-currently not amenable to the current approach.}
+% \textbf{There is no doubt that implementing some fo these features will be more invasive to the underlying simulator code than previously
+% just hijacking the raw-random number draws, but, it would enable us to attack difficulties in the inference procedures that are
+% currently not amenable to the current approach.}
 

### 2019-08-21 paper/learning/notes.tex
updated

@@
 \usepackage{wrapfig}
-
 \usepackage{amsmath,amsthm}
@@
 \usepackage{cancel}
+\usepackage{wrapfig}
 \setlength{\abovedisplayskip}{3pt}
@@
 \usetikzlibrary{positioning}
+
 \newcommand{\n}[1]{\ensuremath{n_{#1}}}
@@
 To construct the mathematical framework in this setting we employ the notation of~\cite{rainforth2017automating}, 
-where \emph{raw samples} are represented by $ f_{a_i}(x_i | \eta)$ and can be thought of as prior 
+where \emph{raw samples} are represented by $ f_{a_i}(x_i | \eta)$ and can be thought of as prior terms.
+
 \textbf{To finish - Linked back to figure 1}
@@
  More generally speaking we can write a rejection sampler in the form of Algorithm~\ref{alg:rssim}
-\begin{algorithm}
-  \small
+\begin{wrapfigure}{L}{0.40\textwidth}
+ \begin{minipage}{0.40\textwidth}
+  \null 
+   \begin{algorithm}[H]
    \caption{ABC: Rejection Sampling}
    \label{alg:rsstats}
-    \begin{algorithmic}\small
+      \begin{algorithmic}
        \State {At iteration $i$ s.t $i \geq 1$}
@@
     \end{algorithm}
-
-\begin{algorithm}
-  \small
+  \end{minipage}%
+\end{wrapfigure}
+
+\begin{wrapfigure}{L}{0.40\textwidth}
+  \begin{minipage}{0.40\textwidth}
+   \begin{algorithm}[H]
     \caption{Simulator: Rejection Sampling}
     \label{alg:rssim}
-    \begin{algorithmic}\small
+      \begin{algorithmic}
         \State {At iteration $i$ s.t $i \geq 1$}
@@
     \end{algorithm}
+  \end{minipage}
+\end{wrapfigure}
+
 Thus, the probability of generating a sample $X = x$ from the true distribution can be defined, using Bayes rule, 

### 2019-08-27 paper/learning/notes.tex
add iclr template

@@
 % [preprint] option:
-\usepackage[preprint]{neurips_2019}
-
+\usepackage{iclr2020_conference,times}
 % to compile a camera-ready version, add the [final] option, e.g.:
@@
 \usepackage{wrapfig}
+
 \setlength{\abovedisplayskip}{3pt}
@@
 % \usepackage[binary-units=true]{siunitx}
-\title{Automating the Amortization of Sub-programs \\ for Fast Simulator Inference}
+\title{Amortization of Sub-programs in Simulators }%\\ for Fast Simulator Inference}
 
@@
 
-    Notes on how to develop a formalism to perform inference in population based simulators.
+  In this work we devleop both a formalism and framework for amortizing sub-programs
+  in simulators leveraging state-of-the-art variational inference techniques.
+  In many real world scenarios, simulators are deployed to model computationally expensive
+  and hard, or expensive, to observe scenarios.
+   These may represent a set of events, or a population of interacting agents.
+
    \end{abstract}
@@
 \bibliography{refs}
-\bibliographystyle{plain}
+\bibliographystyle{iclr2020_conference}
 \appendix

### 2019-08-27 paper/learning/thesystem.tex
updated

@@
  node distance=0cm, 	% distance to Sample
- anchor=center] (Sample) 
+ anchor=center,
+ fill=purple] (Sample) 
 {%
 \begin{tabular}{l} 	% content
-  \textbf{Sample}\\
+  \textcolor{white}{\textbf{Sample}}\\
   \parbox{4cm}{
-   \small{Program call:\\
-   \emph{Program}.\texttt{forward()}}
+  \textcolor{white}{\small{Program call:\\
+  \emph{Program}.\texttt{forward()}}}
  }
@@
  node distance=0cm, 	% distance to Sample
- anchor=center] (OBSERVE) 	% posistion relative to the center of the 'box'
+ anchor=center,
+ fill= purple] (OBSERVE) 	% posistion relative to the center of the 'box'
 {%
 \begin{tabular}{l} 	% content
- \textbf{Observe}\\
+  \textcolor{white}{\textbf{Observe}}\\
  \parbox{4cm}{
-  \small{Program call:\\
-  \emph{Program}.\texttt{forward()}}
+  \textcolor{white}{\small{Program call:\\
+  \emph{Program}.\texttt{forward()}}}
 }
@@
   node distance=6.5cm, 	% distance to Sample
-  anchor=center] (PyProb) 	% posistion relative to the center of the 'box'
+  anchor=center,
+  fill = black!50!green] (PyProb) 	% posistion relative to the center of the 'box'
  {%
  \begin{tabular}{l} 	% content
-  \textbf{System}\\ %system sample
-  \parbox{4cm}{Pyprob processes\\ \texttt{sample}}
+  \textcolor{white}{\textbf{System}}\\ %system sample
+  \parbox{4cm}{\textcolor{white}{Pyprob processes\\ \texttt{sample}}}
  \end{tabular}
@@
   anchor=center,
-  text width=3cm] (System) 
+  text width=3cm,
+  fill = black!50!green] (System) 
  {%
  \begin{tabular}{l}
-  \textbf{System}\\
-  \parbox{4cm}{Pyprob processes \\\texttt{observe}}
+  \textcolor{white}{\textbf{System}}\\
+  \parbox{4cm}{\textcolor{white}{Pyprob processes \\\texttt{observe}}}
  \end{tabular}

### 2019-08-27 paper/learning/notes.tex
updated
add basic abstract

@@
 
-  In this work we devleop both a formalism and framework for amortizing sub-programs
-  in simulators leveraging state-of-the-art variational inference techniques.
+  In this work we develop both a formalism and framework for amortizing sub-programs
+  in simulators leveraging state-of-the-art probabilistic programming and variational inference techniques.
   In many real world scenarios, simulators are deployed to model computationally expensive
   and hard, or expensive, to observe scenarios.
-   These may represent a set of events, or a population of interacting agents.
-
+  The simulators may represent a set of events, or a population of interacting agents. 
+  Recent work by \citet{baydin2018efficient} provided a framework for converting event-based simulators 
+  to probabilistic programs, without having to re-implement the simulator within a given probabilistic
+  programming system (PPS).
+  However, issues arise with both memory and inference when \emph{nested}, or sub-programs within the simulator are simulating 
+  some random process. 
+  We demonstrate that by amortizing such sub-programs not only do we significantly reduce memory consumption, but
+  we also enable inference to be performed in a statistically valid framework. 
 \end{abstract}
@@
  \begin{itemize}
+  \item A mathematical framework to amortize sub-programs within simulators. 
     \item An automated training procedure for any given simulator.
-    \item Extend the protocols of existing probabilistic programming systems that hijack simulators, to enable:
-    \begin{enumerate} \item 1) Only the necessary raw samples from the desired rejection sampling addresses to be collected for training.
-                       \item 2) A process at inference time that replaces those given addresses with a learnt surrogate.
-    \end{enumerate}
-    \item Finer grained control over the actual latent variables of interest, to avoid unnecessary memory consumption and ensure that inference problem encountered remain tractable.
+    \item Extend the protocols of existing probabilistic programming systems that hijack simulators, 
+    to enable only the necessary raw samples from the desired rejection sampling addresses to be collected for training.
+    \item A process at inference time that replaces those given addresses with a learnt surrogate.
+        % \item Finer grained control over the actual latent variables of interest, to avoid unnecessary memory consumption and ensure that inference problem encountered remain tractable.
   \end{itemize}
@@
 
-Thus, we need a regression from $\phi_j \rightarrow P^{in}){a_j}(\phi_j) $.
+Thus, we need a regression from $\phi_j \rightarrow P^{in}{a_j}(\phi_j) $.
 We have access to pairs $\{\phi_j, \hat{P}^{in}_{a_j}(\phi_j)\}$ for which 

### 2019-08-27 code/amortized_rs/memory_src/hogwild.py
added paper objective for given proposal. Tidy up code base tomorrow

@@ -59,26 +59,30 @@ def train(model, optimizer, loss_fn,  N, data_flag, batch_size, rank, load_data=
     # The network needs to learn z1 -> z2 and z2,z3 -> z4
     for i in range(N):
             inData, outData, count =get_batch(data_flag, batch_size, count)
-            proposal = dist.Normal(*model(inData))
-            pred = proposal.rsample(sample_shape=[batch_size]).view(1,batch_size))
+            mu, std = model(inData)
+            proposal = dist.Normal(mu, std)
+            # pred = proposal.rsample(sample_shape=[batch_size]).view(1,batch_size))
             optimizer.zero_grad()
-            plt.show()
-            _loss = loss_fn(outData, pred)
-            _loss.backward()
+            # plt.show()
+            # _loss = loss_fn(outData, pred)
+            _loss = -proposal.log_prob(outData)
+            total_loss = _loss.sum() / len(inData)
+            total_loss.backward()
             optimizer.step()
 
-            _outLoss += _loss.item()
+            _outLoss += total_loss.item()
+            iterationLoss = total_loss.item()
             # if _outLoss == nan:
             #     break
-            if i % 1 == 0:
+            if i % 20 == 0:
                 # plt.show()
-                ax= sns.distplot(pred[0, :].detach().numpy(), kde=True, color='r',label='pred')
-                ax = sns.distplot(outData[0, :].detach().numpy(), kde=True, color='b', label='ground truth')
-                ax.legend()
-                ax.set_title('Iteration_{}_Rejection_block_{}'.format(i,data_flag))
-                fig = ax.get_figure()
-                fname= '../plots/{}_compare_pred_vs_gt_iteration_rejection_block_{}.png'.format(strftime("%M:%S"),data_flag)
-                fig.savefig(fname=fname)
+                # ax= sns.distplot(pred[0, :].detach().numpy(), kde=True, color='r',label='pred')
+                # ax = sns.distplot(outData[0, :].detach().numpy(), kde=True, color='b', label='ground truth')
+                # ax.legend()
+                # ax.set_title('Iteration_{}_Rejection_block_{}'.format(i,data_flag))
+                # fig = ax.get_figure()
[…]

### 2019-08-27 code/amortized_rs/memory_src/model.py

@@ -2,17 +2,36 @@ import torch
 from torch import nn
 from torch.nn import functional as F
 
+# class density_estimator(nn.Module):
+#     def __init__(self, inputSize, outputSize):
+#         super(density_estimator, self).__init__()
+#         self.hidden = nn.Sequential(
+#             nn.Linear(inputSize, inputSize // 2, bias=True),
+#             nn.ReLU(),
+#             nn.Linear(inputSize // 2, inputSize // (inputSize // 2), bias=True)
+#             # ensures that the output would be 2, assuming inputsize = 2^n
+#         )
+#         self.linear_mu = nn.Linear(inputSize // (inputSize // 2), outputSize)
+#         self.linear_std = nn.Linear(inputSize // (inputSize // 2), outputSize)
+#
+#     def forward(self, x):
+#          # x is of shape [batch_size, input_dim]
+#         hidden = self.hidden(x) # return [batch_size//2, 2]
+#         mu = self.linear_mu(hidden)
+#         sigma = F.softplus(self.linear_std(hidden)) + 1e-6
+#         return mu, sigma
+#
 class density_estimator(nn.Module):
     def __init__(self, inputSize, outputSize):
         super(density_estimator, self).__init__()
         self.hidden = nn.Sequential(
-            nn.Linear(inputSize, inputSize // 2, bias=True),
+            nn.Linear(inputSize, inputSize, bias=True),
             nn.ReLU(),
-            nn.Linear(inputSize // 2, inputSize // (inputSize // 2), bias=True)
+            nn.Linear(inputSize , inputSize, bias=True)
             # ensures that the output would be 2, assuming inputsize = 2^n
         )
-        self.linear_mu = nn.Linear(inputSize // (inputSize // 2), outputSize)
-        self.linear_std = nn.Linear(inputSize // (inputSize // 2), outputSize)
+        self.linear_mu = nn.Linear(inputSize, inputSize)
+        self.linear_std = nn.Linear(inputSize, inputSize)
 
     def forward(self, x):
          # x is of shape [batch_size, input_dim]

### 2019-08-28 code/amortized_rs/memory_src/hogwild.py
adding input/output fix to resolve dimensonality issues

@@ -131,33 +131,26 @@ if __name__ == '__main__':
     lr = 0.00001
     momentum= 0.60
     load_data=False
-    batch_size = 128
-    data_flag= 'R1'
+    batchSize = 2**7
+    data_flag= 'R2'
     outputSize = 1
     # outputSize = 128 # for R1 and R2
     if data_flag =='R2':
-        inputSize = batch_size*2
+        inputSize = batchSize*2
     if data_flag == 'R1':
-        inputSize = batch_size
-    model = density_estimator(inputSize, outputSize)
+        inputSize = batchSize
+    model = density_estimator(inputSize, outputSize,batchSize)
     num_processes = mp.cpu_count() - 2
-    N =  1000
+    N =  10000
     trainOn = True
+    testOn=False
     # loss_fn = th.nn.CosineEmbeddingLoss()
     loss_fn = th.nn.MSELoss()
     optimizer = optim.Adam(filter(lambda p: p.requires_grad, model.parameters()), lr=lr, amsgrad=True)
     # optimizer = optim.SGD(model.parameters(), lr=lr, momentum=momentum, weight_decay=0.01)
-    train(model, optimizer, loss_fn, N, data_flag, batch_size, rank=0, load_data=False)
+    train(model, optimizer, loss_fn, N, data_flag, batchSize, rank=0, load_data=False)
     # NOTE: this is required for the ``fork`` method to work
-    # if trainOn:
-    #     model.share_memory()
-    #     processes = []
-    #     for rank in range(num_processes):
-    #         p = mp.Process(target=train, args=(model,optimizer,loss_fn, N,data_flag, batch_size, rank))
-    #         p.start()
-    #         processes.append(p)
-    #     for p in processes:
-    #         p.join()
+
     # testOn = False
     if testOn:
         model_name = 'model_2019-08-12_09-35_rejectionBlock_R2_process_8'

### 2019-08-28 code/amortized_rs/memory_src/inference.py

@@ -251,7 +251,7 @@ if __name__ == '__main__':
     lr = 0.001
     momentum = 0.9
     self.loadData = False
-    batch_size = 128
+    self.batchSize = 128
     self.address = 'R2'
     outputSize = 1
     # outputSize = 128 # for R1 and R2
@@ -259,7 +259,7 @@ if __name__ == '__main__':
         inputSize = batch_size * 2
     if self.address == 'R1':
         inputSize = batch_size
-    model = density_estimator(inputSize, outputSize)
+    model = density_estimator(inputSize, outputSize, batchSize)
     num_processes = mp.cpu_count()
     N = 2000
     trainOn = True

### 2019-08-28 code/amortized_rs/memory_src/model.py

@@ -22,16 +22,20 @@ from torch.nn import functional as F
 #         return mu, sigma
 #
 class density_estimator(nn.Module):
-    def __init__(self, inputSize, outputSize):
+    def __init__(self, inputSize, outputSize, batchSize):
         super(density_estimator, self).__init__()
+        # if batchSize % inputSize == 1:
+        #     outputSize = inputSize
+        # else:
+        #     outputSize = batchSize
         self.hidden = nn.Sequential(
             nn.Linear(inputSize, inputSize, bias=True),
             nn.ReLU(),
             nn.Linear(inputSize , inputSize, bias=True)
             # ensures that the output would be 2, assuming inputsize = 2^n
         )
-        self.linear_mu = nn.Linear(inputSize, inputSize)
-        self.linear_std = nn.Linear(inputSize, inputSize)
+        self.linear_mu = nn.Linear(inputSize, batchSize)
+        self.linear_std = nn.Linear(inputSize, batchSize)
 
     def forward(self, x):
          # x is of shape [batch_size, input_dim]

### 2019-08-28 paper/learning/notes.tex
updated

@@
 \usepackage{wrapfig}
-
+\usepackage{amssymb}
 \setlength{\abovedisplayskip}{3pt}
@@
 \usetikzlibrary{positioning}
-
+\lstnewenvironment{cpp}
+{\lstset{language=C++,
+  basicstyle=\footnotesize,
+  keywordstyle=\color{blue}\ttfamily,
+  stringstyle=\color{red}\ttfamily,
+  commentstyle=\color{green}\ttfamily,
+  morecomment=[l][\color{magenta}]{\#},
+  breaklines = true,
+  stringstyle=\color{blue}\ttfamily, % typewriter type for strings, blue 
+  showstringspaces=false,
+  postbreak=\raisebox{0ex}[0ex][0ex]{\color{red}$\hookrightarrow$\space},
+}}{}
 \newcommand{\n}[1]{\ensuremath{n_{#1}}}
@@
 
-\lstset{language=C++,
-                basicstyle=\footnotesize,
-                keywordstyle=\color{blue}\ttfamily,
-                stringstyle=\color{red}\ttfamily,
-                commentstyle=\color{green}\ttfamily,
-                morecomment=[l][\color{magenta}]{\#},
-                breaklines = true
-}
+% \lstset{language=C++,
+%                 basicstyle=\footnotesize,
+%                 keywordstyle=\color{blue}\ttfamily,
+%                 stringstyle=\color{red}\ttfamily,
+%                 commentstyle=\color{green}\ttfamily,
+%                 morecomment=[l][\color{magenta}]{\#},
+%                 breaklines = true
+% }
 \pgfplotsset{width=7cm,compat=1.8}
@@
 
-\section{Nested-Programs within Simulators}
+\section{Sub-Programs within Simulators}
 
@@
 
+Sub-programs, all called nested programs~\cite{}
 To construct the mathematical framework in this setting we employ the notation of~\cite{rainforth2017automating}, 
@@
 
-\section{Amortized Rejection Sampling}
+\section{Amortized Sub-programs}
 
@@
 \bibliographystyle{iclr2020_conference}
+
+%%%%%%% Appendix %%%%%%%%
 \appendix
 
+\section{Code Examples of Stochastic Processes in Simulators }
+
+The following examples are taken from SHERPA~\cite{gleisberg2009event}
+a million-line plus code base used for the modelling of complex physics
+simulators at CERN.
+
+\small\begin{table}[]
+  \begin{tabular}{|l|l|l|l|}
+  \hline
+  \textbf{Code} & \textbf{Base} & \textbf{Timed} & \textbf{Stochastic Control Flow} \\ \hline
+  \begin{cpp}
+  double Splitter_Base::SelectZ(const double & delta,
+  const bool & lead)
+   {double zmin(0.5*(1.-sqrt(1.-delta))),
+   zmax(0.5*(1.+sqrt(1.-delta))), z;
+    do {
+      z = zmin+ran->Get()*sqrt(1.-delta);
+        } 
+    while (1.-2.*z*(1.-z) < ran->Get());
+  return z;
+   }
+  \end{cpp}&Yes&No&Yes\\ 
+  \hline
+  \begin{cpp}
+double Decay_Channel::GenerateMass(const double& max,
+const double& width) const
+{
+ double mass=-1.0;
+ double decaymin = MinimalMass();
+ if(decaymin>max) mass=-1.0;
+ else if (decaymin==0.0) {
+   mass=m_flavours[0].RelBWMass(decaymin, max,
+                p_ms->Mass(m_flavours[0]), width);
+      }
+ else {
+   double s=sqr(p_ms->Mass(GetDecaying()));
+   double mb(0.0), mc(0.0);
+   for (int i=0; i<NOut(); ++i) {
+     mc+=p_ms->Mass(GetDecayProduct(i));
+     if(p_ms->Mass(GetDecayProduct(i))>mb)
+       mb=p_ms->Mass(GetDecayProduct(i));
+   }
+   mc-=mb;
+   double b=sqr(mb);
+   double c=sqr(mc);
+   double spmax=2.0*b+2.0*c+sqrt(sqr(b)+14.0*b*c+sqr(c));
+   double wmax=MassWeight(s,spmax,b,c);
+   double w=0.0;
+   int trials(0);
+   do {
+     mass=m_flavours[0].RelBWMass(decaymin, max,
+                                  p_ms->Mass(m_flavours[0]), width);
+     double sp=sqr(mass);
+     w=MassWeight(s,sp,b,c);
+     ++trials;
+   } while (w<ran->Get()*wmax && trials<1000);
+ }
+ return mass;
+}
+\end{cpp}
+        &  No             & Yes                     & No  \\ \hline
+                &               &                     &                                  \\ \hline
+  \end{tabular}
+  \end{table}
 \section{The unconditional acceptance probability}

### 2019-08-31 code/amortized_rs/memory_src/amortized_rs.cpp
minor updates to normal parameters of R1 to increase computation speed

@@ -19,7 +19,7 @@ torch::Tensor batch_f(unsigned int bs){
 }
 
 at::Tensor f(void) {
-  Scalar z_1 = at::empty({1}, kFloat).normal_(0.0, 1.0).item();
+  Scalar z_1 = at::empty({1}, kFloat).normal_(0.0, 0.1).item();
   Scalar z_2;
 
   int ctr = 0;
@@ -53,7 +53,7 @@ Scalar R1(Scalar z_1, unsigned int i) {
 //    }
     long j = 0;
     while (true){
-        temp = at::empty({1}, kFloat).normal_(z_1.to<float>(), 1.0).item();
+        temp = at::empty({1}, kFloat).normal_(z_1.to<float>(), 5).item();
         if (temp.to<float>() > 0){
             return temp;
         } else if (j == 10000){

### 2019-08-31 paper/learning/notes.tex

@@
 double Decay_Channel::GenerateMass(const double& max,
-const double& width) const
+const double& width) 
+const
 {
@@
    double c=sqr(mc);
-   double spmax=2.0*b+2.0*c+sqrt(sqr(b)+14.0*b*c+sqr(c));
+   double spmax=2.0*b+2.0*c+\\
+   sqrt(sqr(b)+14.0*b*c+sqr(c));
    double wmax=MassWeight(s,spmax,b,c);

### 2019-09-02 paper/learning/notes.tex
updated

@@
 % Primary: Applications
-% Secodnary cat:Data, Challenges, Implementations, and Software
+% Secondary cat:Data, Challenges, Implementations, and Software
 % or Hijacking Malaria Simulators with Probabilistic Programming
@@
 inference in likelihood free settings~\cite{greenberg2019automatic, mohamed2016learning, papamakarios2018sequential,brehmer2018mining}
-for conditional density estimation. However, whilst many of these methods look to generate a surragote 
+for conditional density estimation. However, whilst many of these methods look to generate a surrogate 
 for the entire simulator, we propose amortizing only computational expensive sub-components 
@@
 
-Sub-programs, all called nested programs~\cite{}
-To construct the mathematical framework in this setting we employ the notation of~\cite{rainforth2017automating}, 
-where \emph{raw samples} are represented by $ f_{a_i}(x_i | \eta)$ and can be thought of as prior terms.
-
-\textbf{To finish - Linked back to figure 1}
-In order to correctly characterize this problem we first provide a useful diagram demonstrating 
-the density of the system~\ref{fig:systemoverview}. \textbf{Lots to do}
+Sub-programs, also called nested programs~\cite{rainforth2018nesting}, are an integral part of stochastic simulators.
+Within stochastic simulators they are used to simulate densities that cannot be sampled from in any known, or efficient way.
+Such nested-programs may be multi-tiered, in the sense that there exists several nested structures within a given sub-program.
+For example, given an outer sub-process $S_{out}$, that contains another sub-process $S_{in_{1}}$,
+which itself contains another sub-program, $S_{in_{2}}$, that is $S_{in_2} \in S_{in_1} \in S_{out}$. 
+Due to these recurrent dependency structures within simulators performing statistical inference becomes increasingly challenging, part of this
+difficulty is due to the exploration of the domain of the density $R^{d}$. 
+Because of the nested structures the inference strategy see a subset of the domain. 
+This happens due to stochastic control flow within the sub-processes, which means only a subset of paths are available 
+during the evaluation of the density. 
+\textbf{Talk about why this causes trouble for the inference procedures etc}
+
+Due to these complexities and the increase in computation time for both inference, \texttt{.backward()}, and the simulation, \texttt{.forward()},
+we leverage the use of amortization schemes to, as we show in Section~\ref{sec:amortizesubp}, correctly utilise MCMC inference procedures and reduce
+computation time.
+
+Each nested process representing a different density, each simulating some statistical process.
+
+% To construct the mathematical framework in this setting we employ the notation of~\cite{rainforth2017automating}, 
+% where \emph{raw samples} are represented by $ f_{a_i}(x_i | \eta)$ and can be thought of as prior terms.
+
+% \textbf{To finish - Linked back to figure 1}
+% In order to correctly characterize this problem we first provide a useful diagram demonstrating 
+% the density of the system~\ref{fig:systemoverview}. \textbf{Lots to do}
 
@@
 
+\label{sec:amortizesubp}
 One common problem in traditional simulators, \emph{programs}, outside of the standard probabilistic
 programming framework is that they rely heavily on rejection sampling 
-to generate samples from the target density. Not only is rejection sampling 
-computational inefficient, [especially if criteria in which samples are generated
+to generate samples from the target density. 
+Not only is rejection sampling 
+computational inefficient, [especially if the criteria in which samples are generated
 has been poorly calibrated] (may leave out this statement as it may make a reviewer
@@
 Hence, to leverage existing frameworks we can make use of the probabilistic 
-programming executation (PPX) protocols~\cite{baydin2018efficient}. 
+programming execution (PPX) protocols~\cite{baydin2018efficient}. 
 These protocols allow one to exact all raw sample calls $f_{a_j}$ from the
@@
 In addition to this, once such rejection schemes, \emph{nested programs}, are located, we have a number of 
-primatives that we can leverage to isolate calls to those addresses during training, to collect the samples generated during 
+primitives that we can leverage to isolate calls to those addresses during training, to collect the samples generated during 
 \texttt{.forward()} runs of the simulator, 
@@
   \item testing code and infrastructure
-  \item litrature review on any methods to learn rejection sampling methods, 
+  \item literature review on any methods to learn rejection sampling methods, 
   and conditional density estimation. See this paper for kernel approximation 
@@
 functions and optimizers} and use the
-neural network, $\Gamma(z, y)$, to learn a $\hat{\mu}$ and $\hat{\sigma}^{2}$ to parametize 
+neural network, $\Gamma(z, y)$, to learn a $\hat{\mu}$ and $\hat{\sigma}^{2}$ to parameterize 
 a normal proposal. For $R1$, this means $\Gamma_{r1}(z,y) \rightarrow \hat{\mu_{r1}},  \hat{\sigma_{r1}}^{2}$  
@@
 
-We discussed at length how to correctly define the objective, I have now updated the objective with the one derivied in the revised discussion,
+We discussed at length how to correctly define the objective, I have now updated the objective with the one derived in the revised discussion,
 along with additional notes on rejection sampling.
@@
 
-The main fundemental difference is that IC cares about the full global program, 
+The main fundamental difference is that IC cares about the full global program, 
 we only care about the local (sub / nested programs).
@@
 as you can view this work as learning an approximation for the \emph{local}, sub-programs
-$P^{in}(x|\nu)$ where the nuissince parameters in this paper are represented by $\psi$.
-
-In the notation of the inference compilation paper we essentailly learn $q_{a_{t},i_{t}}(x_t| x_{1:t-1}, y)$
-where the indicies are part of the set $S_r$.
+$P^{in}(x|\nu)$ where the nuisance parameters in this paper are represented by $\psi$.
+
+In the notation of the inference compilation paper we essentially learn $q_{a_{t},i_{t}}(x_t| x_{1:t-1}, y)$
+where the indices are part of the set $S_r$.
  The difference here though, is that we only care about the input at that address, your proposals 
- are condiitoning on everything upto that point of the executation trace $\mathcal{T}$
+ are conditioning on everything up to that point of the execution trace $\mathcal{T}$
  in the simulator. 
  The training data represented by (11) in the IC paper is not the training data we use in this paper,
- we only care about the $x^{m}_t$ or $z$'s and $x$'s in this paper (notation is not entirely consitant throuhg out this paper yet)
+ we only care about the $x^{m}_t$ or $z$'s and $x$'s in this paper (notation is not entirely consistent through out this paper yet)
  But these are the parameters that are generated from raw sample calls within side the RS loop $f_{a_i \in S_r}(\cdot))$.
@@
 
-\subsection{Method 1: Learn the maoritzation of the sub-programs}
+\subsection{Method 1: Learn the amortization of the sub-programs}
 That is learn $\gamma_{a_{j}}(\cdot ; \eta_{a_{j}}, \cdot)$ for each program in the set $S_r$.

### 2019-09-03 code/amortized_rs/memory_src/inference.py
finished writing code, now to test

@@ -49,6 +49,8 @@ class Inference():
         self.batchSize = parameters.batchSize
         self.model = parameters.model
         self.optimizerParams = parameters.optimizerParams
+        # generate optimizer
+        self.optimizer_Fn()
         self.testOn = parameters.teston
         if self.testOn:
             try:
@@ -92,7 +94,7 @@ class Inference():
 
 
 
-
+    @classmethod
     def optimizer_Fn(self):
         '''
         Sets-up the optimizer
@@ -245,7 +247,7 @@ class Inference():
         for i in range(self.nIterations):
             inData, outData, count = self.get_batch(count)
 
-            proposal = self.proposalClass #TODO call proposal method,whcih relates to the proposal being used.d It requires using self.proposalName
+            proposal = self.proposalClass.__call__(inData=inData, batchSize=self.batchSize, model=self.model) #TODO call proposal method,whcih relates to the proposal being used.d It requires using self.proposalName
             self.optimizer.zero_grad()
             _loss = -proposal.log_prob(outData)
 

### 2019-09-03 code/amortized_rs/memory_src/proposal.py

@@ -1,0 +1,35 @@
+#!/usr/bin/env python3
+# -*- coding: utf-8 -*-
+'''
+Author: Bradley Gram-Hansen
+Time created:  17:30
+Date created:  02/09/2019
+
+License: MIT
+'''
+
+import torch
+import torch.distributions as dist
+import warnings
+
+class Proposal():
+
+    def __init__(self, *args, **kwargs):
+        args = args
+        kwargs = kwargs
+        self.proposalMethod = kwargs['proposalMethod'] if kwargs['proposalMethod'] else warnings.warn('{} A valid proposal is required {}'.format(5*'*'))
+
+    def __call__(self, *args, **kwargs):
+        return exec(self.proposalMethod+'({},{})'.format(*args, **kwargs))
+
+    def Normalapproximator(self, *args, **kwargs):
+        '''
+
+        :param inData: :type torch.tensor The data that you want to regress over to learn amortized function
+        :param batchSize: :type int batch size
+        :param model: neural network
+        :return: prediction to the regression problem specified by inData
+        '''
+
+        proposal = dist.Normal(*kwargs['model'](kwargs['inData']))
+        return proposal.rsample(sample_shape=[kwargs['batchSize']]).view(1, kwargs['batchSize'])

### 2019-09-03 paper/learning/notes.tex
minor

@@
 
+\subsection{Theory of Nested-programs}
+
 Each nested process representing a different density, each simulating some statistical process.
@@
 For any rejection sampling block, written in a generic code base, not in the form of R3, this is an incredibly hard 
-task to do, because rejection sampling blocks within simulators are not necessarily written in a standard template. 
-\textbf{Add examples in a supplement.  Need to demonstrate to YW that this is challenging to do, even if you know where 
-the rejection sampling loops are. Hypothesis, RS loops are not written in a templated form}.
-y leveraging the PPX protocols previously developed in \cite{baydin2018efficient,gram2019hijacking} we can however, 
+task to do, because rejection sampling blocks within simulators are not necessarily written in a standard template, see appendix~\ref{app:exampleprograms}.
+
+We leveraging the PPX protocols previously developed in \cite{baydin2018efficient,gram2019hijacking} we can however, 
 extract all variables, other than $P(accept)$ in an automated way.
@@
 \section{Code Examples of Stochastic Processes in Simulators }
-
+\label{app:exampleprograms}
 The following examples are taken from SHERPA~\cite{gleisberg2009event}

### 2019-09-07 code/amortized_rs/memory_src/inference.py
fixed param error

@@ -490,7 +490,7 @@ def main(test=False):
                                 type=bool)
             parser.add_argument('--checkpoint', '--chk', help='Ifmain you want to save checkpoints. Default is True.  ', default= True, type=bool)
             parser.add_argument('--savedata', '--sd', help='True if you want to save data, else False. Default True',default=True, type=bool )
-            parser.add_argument('--datastr', help='The date str of the experiment to load',
+            parser.add_argument('--datestr', help='The date str of the experiment to load',
                                 default=None, type=str)
 
 
@@ -506,7 +506,7 @@ def main(test=False):
 
 
 if __name__ == '__main__':
-    # time_start = time.time()
+    time_start = time.time()
     main(test=False)
-    # print('\nTotal duration: {}'.format((time.time() - time_start)))
-    # sys.exit(0)
+    print('\nTotal duration: {}'.format((time.time() - time_start)))
+    sys.exit(0)

### 2019-09-07 code/amortized_rs/memory_src/amortize_cpp/__init__.py
fixed multi dim issue

@@ -1,0 +1,9 @@
+#!/usr/bin/env python3
+# -*- coding: utf-8 -*-
+'''
+Author: Bradley Gram-Hansen
+Time created:  06:14
+Date created:  07/09/2019
+
+License: MIT
+'''

### 2019-09-07 code/amortized_rs/memory_src/amortize_cpp/hogwild.py

@@ -1,0 +1,166 @@
+import torch.multiprocessing as mp
+from model import density_estimator
+from torch.utils.cpp_extension import load
+import torch as th
+from torch import optim
+import torch.distributions as dist 
+from time import strftime
+import os
+import seaborn as sns
+import matplotlib.pyplot as plt
+import time
+
+amortized_rs = load(name="amortized_rs",
+                    sources=["amortized_rs.cpp"])
+
+# no longer need write the batch loop in c - we can do that later
+def get_batch(data_flag, batch_size, count, load_data=False):
+    if not load_data:
+        data = th.stack([amortized_rs.f() for _ in range(batch_size)]) 
+    if data_flag == 'R1' and not load_data:
+        inR1 = data[0:batch_size,0].view([batch_size,1]).view([1,batch_size])
+        outR1 = data[0:batch_size,1].view([batch_size,1]).view([1,batch_size])
+        count = count + 1 # not actually need if load_data == True
+        return inR1.to(device),outR1.to(device), count
+    elif data_flag == 'R1' and load_data:
+        inR1 = data[count:batch_size+count*batch_size,0]
+        outR1 = data[count:batch_size+count*batch_size,1]
+        count = count + 1
+        return inR1.to(device),outR1.to(device), count
+    elif data_flag == 'R2'and not load_data:
+        inR2 = th.cat([data[0:batch_size,1], data[0:batch_size,2]], dim=0).view(1,batch_size*2)
+        outR2 = data[0:batch_size,3].view([1,batch_size])
+        count = count + 1
+        return inR2.to(device),outR2.to(device), count
+    elif data_flag == 'R2' and load_data:
+        inR2 = th.stack([data[count:batch_size+count*batch_size,1], data[count:batch_size+count*batch_size,2]], dim=1)
+        outR2 = data[count:batch_size+count*batch_size,1]
+        count = count + 1
+        return inR2.to(device), outR2.to(device), count
+
+def center_data(data):
+    mean = th.mean(data,dim =0)
+    zeroed = data - mean*th.ones(data.shape)
+    stddata  = th.std(data, dim=0)
+    return zeroed / stddata
+def train(model, optimizer, loss_fn,  N, data_flag, batch_size, rank, load_data=False):
+    model.train()
+    n_samples = batch_size*N
+    if load_data == True:
+        data = th.load('../data/all_batch_samples.pt')
+        data = data[0:n_samples,:]
[…]

### 2019-09-07 code/amortized_rs/memory_src/inference.py

@@ -58,6 +58,7 @@ class Inference():
         if not os.path.exists(self.BASEPATH):
             os.makedirs(self.BASEPATH)
         self.address = parameters.address
+        print('The address is {}'.format(self.address))
         self.batchSize = parameters.batchsize
         self.model = parameters.model
         self.optimizerParams = parameters.optimparams
@@ -93,9 +94,9 @@ class Inference():
         else:
             self.processes = parameters.ncores
         if self.address == 'R2':
-            self.inputsize = 2*self.batchSize
+            self.inputsize = 2 * self.batchSize
             self.outputsize = self.batchSize
-        if self.address == 'R1' or 'R3':
+        if self.address == 'R1' or self.address == 'R3':
             self.inputsize = self.batchSize
             self.outputsize = self.batchSize
         if parameters.model:
@@ -221,7 +222,7 @@ class Inference():
             count = count + 1
             return inR1.to(self.device), outR1.to(self.device), count
         elif self.address == 'R2' and not self.loadData:
-            inR2 = th.cat([data[0:self.batchSize, 1], data[0:self.batchSize, 2]], dim=0).view(1, 256)
+            inR2 = th.cat([data[0:self.batchSize, 1], data[0:self.batchSize, 2]], dim=0).view(1, self.inputsize)
             outR2 = data[0:self.batchSize, 3].view([1, self.batchSize])
             count = count + 1
             return inR2.to(self.device), outR2.to(self.device), count
@@ -431,7 +432,7 @@ def main(test=False):
     if test:
         class Params():
             device = 'cpu'
-            address = 'R1'
+            address = 'R2'
             batchsize = 1024
             optimparams= '{"name": "Adam", "lr": 0.001, "betas": (0.9, 0.999)}'
             model = 'density_estimator'
@@ -441,7 +442,7 @@ def main(test=False):
             testiterations =2
             loadcheckpoint = False
             modelname = None
-            ncores = 4
+            ncores = 1
             proposal = 'Normalapproximator'
             loaddata =None
             logpath = None
@@ -507,6 +508,6 @@ def main(test=False):
 
 if __name__ == '__main__':
     time_start = time.time()
-    main(test=False)
+    main(test=True)
     print('\nTotal duration: {}'.format((time.time() - time_start)))
     sys.exit(0)

### 2019-09-07 code/amortized_rs/memory_src/run.sh

@@ -24,7 +24,7 @@ python $script --address $address \
 --trainiterations 1000 \
 --testiterations 1000 \
 --ncores 1 \
---proposal 'NormalApproximator' \
+--proposal 'Normalapproximator' \
 --savemodel True \
 --op "{'name': 'Adam', 'lr': 0.001, 'betas': (0.9, 0.999)}"
 

### 2019-09-07 paper/learning/notes.tex
add tom notes

@@
 
-\subsection{3rd Baseline}
+% \subsection{3rd Baseline}
 
@@
 \end{figure}
-\section{ Questions for Tom}
+
+\subsection{Are the samples in Method 3 Unbiased?}
+The short answer is that we seem to be able to do something that will give better per-iteration performance than vanilla LMH, 
+but that we have an extra term to calculate that means we don't also get the speedups in how much each iteration costs that we might have hoped for.
+
+The expected number of cycles through the rejection sampler $\mathbb{E}[N_a | \lambda]$,
+is equal to 1 over the acceptance ratio of the sampler $Z(\lambda)$ as per the proof we did earlier. \textbf{Bradley add proof}
+This means that we can use the $N_a$ to learn a direct approximator for $\frac{1}{Z(\lambda)}$ for method 3.
+
+Let $P^{in}(z)$ denote the density of the rejection sampler with the conditioning statement removed 
+(i.e. the density on the outputs if we just accepted all samples).  
+Let $a(z)=1$ denote the event of acceptance. 
+Let $p(z)$ denote the density of the outputs from running the rejection sampler.
+
+We now have the following results
+\begin{align}
+   p(z ; \lambda) =& P^{in}(z ; \lambda) \frac{\mathbb{I}(a(z ; \lambda)=1)}{Z(\lambda)} \\
+   \mathbb{E}[P^{in}(z ; \lambda) \mathbb{I}(a(z ; \lambda)=1) N_a | z] =& P^{in}(z ; \lambda) \mathbb{I}(a(z ; \lambda)=1) \mathbb{E}[N_a | z, \lambda]\\
+=& P^{in}(z ; \lambda) \mathbb{I}(a(z ; \lambda)=1) \mathbb{E}[N_a | \lambda] \\ %(see explanation below) 
+=& p(z ; \lambda) = P^{in}(z ; \lambda) \frac{\mathbb{I}(a(z ; \lambda)=1)}{Z(\lambda)}
+\end{align}
+
+The key here, that I was worrying about before, is the second step
+that is does $\mathbb{E}[N_a | z, \lambda] = \mathbb{E}[N_a | \lambda]$?
+The answer is that it does if we think about things from the perspective as $z$
+as the output from the rejection sampler. 
+The exact value of this has no impact on $N_a$, only the fact that it was accepted.
+But, by definition, we know that $z$ has been accepted, so  $\mathbb{E}[N_a | z, \lambda]$
+is the same for all possible $z$ and thus equal to  $\mathbb{E}[N_a | \lambda]$.
+
+This argument is perhaps easiest to see in reverse: we can think of first sampling $N_a \sim P(N_a)$
+to figure out how long we sample for before acceptance, then we can sample an output using $z \sim p(z ; \lambda)$, which clearly does not depend on $N_a$ so the two must be independent.
+In other words, though the full set of samples that are generated during running $x_{1:N_a}$ is clearly not independent from $N_a$, the final sample we return $z$ definitely is.
+Note that this does not cause any complications with the fact that we never actually evaluate $P^{in}(z)$, but its preimage $P^{in}(x_{N_a})$
+instead, because the result further implies that $x_{N_a}$ and $N_a$ are independent, even though $x_{1:N_a}$ and $N_a$ are not.
+
+In general, it is best to think about pseudo-marginal schemes (i.e. MCMC chains based on unbiased density estimates) as estimating the density by drawing a
+random number seed $\epsilon \sim p(\epsilon)$, then producing a deterministic density for a given parameter value and this random number seed. 
+We can thus think of the density estimate as being constructed through the process
+
+\begin{align}
+  \epsilon &\sim p(\epsilon)\text{(draw a seed)} \\
+  \hat{p}(z) &= p(z;\epsilon)\text{(i.e. the estimate is a deterministic function of $z$ and epsilon)}
+\end{align}
+
+If we now do an MCMC step, our implied proposal on $\{\epsilon,z\}$ space becomes $q(\epsilon',z'|\epsilon,z) = p(\epsilon')q(z'|z)$ for some $q(z'|z)$ that we choose.
+Using our density estimates now gives the pseudo-marginal MH acceptance ratio
+\begin{equation}
+  P_a = \frac{ \hat{p}(z') q(z | z') }{ \hat{p}(z') p(z;\epsilon) q(z' | z) }
+\end{equation}
+and from the previous results we have
+\begin{align}
+&=  \frac{ p(z';\epsilon') q(z | z') }{ p(z;\epsilon) q(z' | z) } \\
+&= \frac{ p(z';\epsilon') q(\epsilon,z|\epsilon',z')/p(\epsilon)  }{ p(z;\epsilon) q(\epsilon',z'|\epsilon,z)/ p(\epsilon')  } \\
+&=  \frac{ p(z';\epsilon') p(\epsilon') q(\epsilon,z|\epsilon',z') }{ p(z;\epsilon) p(\epsilon)  q(\epsilon',z'|\epsilon,z) }
+\end{align}
+which is a valid MH sampler on the unnormalized density $p(\epsilon) p(z;\epsilon)$.
+If our $\hat{p}(z)$ are unbiased, this implies that $\mathbb{E}[ \hat{p}(z) | z] = \int p(epsilon) p(z;\epsilon)  d\epsilon = p(z)$,
+hence $p(z)$ is the marginal distribution of $ p(\epsilon) p(z;\epsilon)$, so drawing samples from  $p(\epsilon) p(z;\epsilon)$  produces the desired distribution on z and the pseudo-marginal MCMC sampler is valid.
+
+We can now view our case in the general pseudo-marginal setting above.  
+Effectively we have $\epsilon \sim p(\epsilon)$ (draw a random number seed for the local rejection sampler)
+$N_a = N_a(\epsilon; \lambda)$, (i.e. $N_a$ is a deterministic function of $\epsilon$ and $\lambda$)
+$\hat{p}(z) = p_in(z ; \lambda) \mathbb{I}(a(z ; \lambda)=1) N_a(\epsilon; \lambda)$
+
+This is perfectly fine to run a pseudo-marginal scheme on, but the problem is that if $\lambda$ changes, so does $N_a$.  
+Thus even though we could theoretically use our trick about $N_a$ to run things in a pseudo marginal manner with, e.g., 
+LMH updates where we keep the $\epsilon$ and $z$ fixed for rejection samplers downstream from the one we are updating, 
+the change in $\lambda$ means not only do we need to recalculate $P^{in}(z ; \lambda) \mathbb{I}(a(z ; \lambda)$ 
+as we would normally do but which only requires a single pass through the rejection sampler, we also have to recalculate $N_a (\epsilon; \lambda)$,
+which requires us to pass through the rejection sampler multiple times, such that our updates remain extremely expensive. 
+Alternatively, we could also just redraw $N_a$ for all downstream rejection samplers (with a bit of care that I haven't not fully gone through),
+which effectively equates to redrawing the epsilons and is thus simply a different MH update.
+However, again this means we need to repass through the rejection sampling loops.
+
+Interestingly, there is still potentially some gain to be had here as we have circumvented the need to redraw the rejection sampler outputs (i.e.
+$z$) of future draws in the way that we would need to for vanilla LMH.
+As such, this would give us "method 0b" and it should be more efficient than vanilla LMH as even though the updates remain expensive, 
+it does avoid the curse of dimensionality issues we get from having to redraw all the future $z$'s from the prior each time we update an earlier one: 
+we only have to recalculate the number of cycles to acceptance and then use these in the acceptance ratio (and next iteration if they are accepted).
+
+The per-iteration cost of this will be similar to an iteration of training a regressor to predict $N_a$, but much higher than running inference once we have learned such a regressor.  
+There is also potential to combine the two methods: running pseudo-marginal LMH will generate $(\lambda,N_a)$ sample pairs that can be used for the training so you could initially run 
+pseudo-marginal LMH and at some point switch when you're confident you have learned a good enough regressor switch to relying on that instead.
 
@@
 
+\section{ Questions for Tom}
 \section{Additional TODOs}

### 2019-09-08 paper/learning/notes.tex
updated nores and changed structure - cpp load is not working still...

@@
 \section{Sub-Programs within Simulators}
-
+\label{sec:nested}
 \begin{itemize}
@@
 
-Rejection samplers are heavily used when designing simulators [ add SHERPA etc], due to their ease of use and ability to model 
-any arbitrary distribution on $\mathbb{R}^{D}$, with a sufficiently well designed rejection sampling scheme~\cite{MEYER20083408,martino2012improved,casella}
-
-There are two different forms of rejection samplers found with in simulators and program code. 
+Rejection samplers allow from sampling from densities $p(X=x)$ by constructing an upper bound to $p(X=x)$
+specified by some predicate expression that determines whether we accept or reject samples from a density proportional to the bounding envelope. 
+The functions we choose to define the envelope are much simpler than $p(X=x)$ and easy to sample from. 
+When designing rejection samplers we aim to reduce the rate of rejections and increase the rate of acceptance, for computational purposes.
+However, designing efficient and expressive rejection samplers is challenging, leading to complex and inefficient constructions of 
+rejection samplers in simulators, see appendix~\ref{app:exampleprograms} for examples.
+Due to these inefficiencies, and additional complications that arise in nested structures, as described in Section~\ref{sec:nested}
+we propose a scheme to amortize such processes and learn accurate surragotes for the individual sub-programs, which in many cases are complex rejection
+sampling schemes.
+
+% are heavily used when designing simulators,
+%  due to their ease of use and ability to model 
+% any arbitrary distribution on $\mathbb{R}^{D}$, with a sufficiently well designed rejection sampling scheme~\cite{MEYER20083408,martino2012improved,casella} 
+\subsection{Types of rejection samplers}
+
+There are essentially two different forms of rejection samplers found with in simulators and program code. 
 In both instances we we have $X$, a random variable whose density is $f(x)$ that we cannot easily 

### 2019-09-08 code/amortized_rs/memory_src/amortize_cpp/__init__.py

@@ -1,5 +1,3 @@
-#!/usr/bin/env python3
-# -*- coding: utf-8 -*-
 '''
 Author: Bradley Gram-Hansen
 Time created:  06:14

### 2019-09-08 code/amortized_rs/memory_src/cython_code/setup_amortized_rs.py

@@ -1,6 +1,0 @@
-from distutils.core import setup
-from Cython.Build import cythonize
-
-setup(
-    ext_modules = cythonize("amortized_rs.pyx")
-)

### 2019-09-08 code/amortized_rs/memory_src/inference.py

@@ -18,10 +18,7 @@ import sys
 import datetime
 import ast
 import seaborn as sns
-import simulator as sim
-
-# amortized_rs = load(name="amortized_rs",
-#                             sources=["amortized_rs.cpp"])
+# import simulator as sim
 
 class Inference():
     """
@@ -58,7 +55,6 @@ class Inference():
         if not os.path.exists(self.BASEPATH):
             os.makedirs(self.BASEPATH)
         self.address = parameters.address
-        print('The address is {}'.format(self.address))
         self.batchSize = parameters.batchsize
         self.model = parameters.model
         self.optimizerParams = parameters.optimparams
@@ -116,6 +112,8 @@ class Inference():
             if not os.path.exists(self.dataPath):
                 os.makedirs(self.dataPath)
 
+        self.amortized_rs = load(name="amortized_rs",
+                            sources=["amortized_rs_std.cpp"])
 
 
     def optimizer_Fn(self,optimizerParams):
@@ -206,7 +204,8 @@ class Inference():
                 count :type int Updated count
         '''
         if self.loadData is None:
-            data = th.stack([sim.s() for _ in range(self.batchSize)])
+            # data = th.stack([sim.s() for _ in range(self.batchSize)])
+            data = th.stack([self.amortized_rs.f() for _ in range(self.batchSize)])
             if self.saveData:
                 fname = self.dataPath + str(float(iteration))+'_samples_.th'
                 th.save(data,fname)
@@ -260,7 +259,7 @@ class Inference():
             _loss = -proposal.log_prob(outData)
 
             # calculate the SVI update sum of log(prob..) / 'batchsize'
-            totalLoss = _loss.sum() / len(inData)
+            totalLoss = _loss.sum() / inData.size()[1]
 
             totalLoss.backward()
             self.optimizer.step()
@@ -438,7 +437,7 @@ def main(test=False):
             model = 'density_estimator'
             trainon = True
             teston= True
-            trainiterations = 1000
+            trainiterations = 2
             testiterations =2
             loadcheckpoint = False
             modelname = None
